### 2018-02-28 paper/compute.tex
Add first draft of the paper

\section{Compute}\label{sec-compute}

\begin{figure}
\begin{minted}{haskell}
type IdentityCompute    k v = forall f.                  (k -> f v) -> k -> f v
type FunctorialCompute  k v = forall f. Functor     f => (k -> f v) -> k -> f v
type ApplicativeCompute k v = forall f. Applicative f => (k -> f v) -> k -> f v
type AlternativeCompute k v = forall f. Alternative f => (k -> f v) -> k -> f v
type MonadicCompute     k v = forall m. Monad       m => (k -> m v) -> k -> m v
type MonadPlusedCompute k v = forall m. MonadPlus   m => (k -> m v) -> k -> m v
\end{minted}
\caption{Compute abstractions}\label{fig-compute}
\end{figure}

Consider abstractions in Fig.~\ref{fig-compute}.

### 2018-02-28 paper/intro.tex
Add first draft of the paper

\section{Introduction}\label{sec-intro}

Build systems are awesome. Build systems are terrifying.

We separate build systems into \emph{compute} and \emph{build} components,
see~\S\ref{sec-compute}.

### 2018-03-02 paper/background.tex
Add draft of the background section

\section{Background}\label{sec-background}

Below we define basic notions used in build systems and other similar domains,
for example, spreadsheets.

\subsection{Keys, values, hashes and store}

\emph{Keys} are used to distinguish \emph{values}. In build systems keys are
typically filenames, e.g. \textsf{src/file.c}, whereas values are file contents
(a C program source code in this case). In spreadsheets keys are cell names,
e.g. \textsf{A1}, and values are numbers, text, etc. that are typically displayed
inside cells. We will use type variables \hs{k} and \hs{v} to denote keys and
values, respectively.

A \emph{store} associates keys to values. It is convenient to assume that a store
is total, i.e. it contains a value for every possible key. We therefore also
assume that the type of values is capable of encoding values corresponding to
non-existent keys (missing files or empty cells).

We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash v} for
efficient tracking and sharing of build results.

\subsection{Input, intermediate and output values}

Some values must be provided by the user as \emph{input}. For example,
\textsf{src/file.c} can be edited by the user who relies on the build system to
compile it into \textsf{obj/file.o}. Similarly, the user can input \textsf{A1 = 5}
and \textsf{B1 = 9} expecting the spreadsheet to compute their sum in \textsf{C1},
i.e. \textsf{C1 = 14}.

In the above examples, \textsf{obj/file.o} and \textsf{C1} are \emph{output} values.

In some situations we might also need the notion of \emph{intermediate} values,
which are not interesting for the user but are produced in the process of turning
inputs into outputs. For example, the user might only be interested in the
executable \textsf{bin/file.exe} obtained by linking \textsf{obj/file.o} with
standard libraries, in which case \textsf{obj/file.o} can be considered an
intermediate value.

\subsection{Non-deterministic computations}

Build systems and spreadsheets compute output values from input and intermediate
values. In the most typical case, these \emph{computations} are \emph{functions},
such as \textsf{C1 = A1 + B1}, i.e. their result is uniquely determined by the
input values. However, in general they can be \emph{relations}, i.e. have
multiple valid results. A spreadsheet example: \textsf{A2 = A1 + RANDOM(1,6)}.
This computation has six valid results for each input value \textsf{A1}. In
build systems, the object file \textsf{obj/file.o} is sometimes not uniquely
determined by the source \textsf{src/file.c} -- different compiler runs may
produce different valid results.

\subsection{Applicative build systems}

Make ...

Excel ...

\subsection{Monadic build systems}

Shake ...

\subsection{Caching build results}

XXX ...

### 2018-03-02 paper/background.tex
Expand the draft

@@
+\clearpage
+
 \section{Background}\label{sec-background}
@@
 
-\subsection{Applicative build systems}
+\subsection{Dynamic dependencies}
 
-Make ...
+Many existing build systems need to know all the dependencies between files or
+cells, i.e. the dependency graph, at the start of the build process. This makes
+it possible to analyse the graph ahead of time, but is fundamentally limited:
+parts of the dependency graph can often be discovered only during the build
+process, i.e. the dependency graph is dynamic, not static. This is illustrated
+by the following example.
+\todo{AM}{Add two examples: include files and cyclic spreadsheet computations.}
 
-Excel ...
+\subsection{Requirements for build systems}
 
-\subsection{Monadic build systems}
-
-Shake ...
-
-\subsection{Caching build results}
-
-XXX ...
+\begin{itemize}
+    \item Correctness
+    \item Minimality
+    \item Support for sharing and skipping intermediate values
+\end{itemize}
+...

### 2018-03-02 paper/intro.tex
Expand the draft

@@
 
-Build systems are awesome. Build systems are terrifying.
+A build system is a critical component of most software projects, responsible
+for compiling the source code written in various programming languages and
+producing executable programs -- end software products. Build systems sound
+simple, but they are not; in large software projects the build system grows from
+simple beginnings into a huge, complex engineering artefact. Why? Because it
+evolves every day as new features are continuously being added by software
+developers; because it builds programs for a huge variety of target hardware
+configurations (from mobile to cloud); and because it operates under strict
+correctness and performance requirements, standing on the critical path between
+the development of a new feature and its deployment into production.
+
+It is known that build systems can take up to 27\% of software development
+effort, and that improvements to build systems rapidly pay off~\cite{build_maintenance}.
+Despite its importance, this subject is severely under-researched, which prompts
+major companies, such as Microsoft, Facebook and Google, to invest significant
+internal resources to make their own bespoke build system frameworks.
+
+...
+
+Some build systems do not look like build systems but they are. A good example
+is spreadsheets, where cells play the role of files, and formulas play the role
+of build rules.
+
+...
 
 We separate build systems into \emph{compute} and \emph{build} components,
-see~\S\ref{sec-compute}.
+see~\S\ref{sec-build-abstractions}.
+
+...

### 2018-03-02 paper/build-examples.tex
Add build examples section

\clearpage
\section{Build systems, concretely}\label{sec-examples}

This section present several concrete examples of build systems, providing
simple implementations that use the previously introduced abstractions.

\subsection{Applicative build systems}

Applicative build system can only accept an applicative compute, because they
rely on building the full dependency graph upfront, which is impossible with
dynamic dependencies.

\vspace{4mm}
\subsubsection{An incorrect applicative build system}~\\

\todo{AM}{Describe the \hs{dumbBuild} that builds outputs in the given order.}

\vspace{4mm}
\subsubsection{\Make: a correct but non-minimal applicative build system}~\\

We start by applicative build systems, such as the classic vanilla
\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple but reasonable
approximation to a real \Make that you might find on your machine.} and a more
recent \Ninja.

Make is correct but non-minimal. It is unusual from other build systems in that
it relies on \emph{timed values}, i.e. on a map \hs{age :: v -> Time}, as a
heuristic to decide whether a value is up-to-date.

\todo{NM}{Add a description of \Make.}

\vspace{4mm}
\subsubsection{\Ninja: a correct and minimal applicative build system}~\\

\Ninja is a modern alternative to \Make ...

\todo{NM}{Add a description of \Ninja.}
\todo{AM}{What about Nix? John Ericson suggested in a blog comment that it may
be somewhat monadic, see:
\url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.}

\subsection{Monadic build systems}

...
\subsubsection{An incorrect monadic build system}~\\

See \hs{dumbBuild}.

\subsubsection{Correct but not minimal monadic build system}~\\

As of writing, XXX is a correct but non-minimal monadic build system. Here is
an example where it does unnecessary computation:

\begin{minted}[frame=single]{text}
A1 = 10
A2 = MIN(A1, 4)
A3 = SQRT(A2)
\end{minted}

After performing the build, XXX computes \textsf{A2 = 4} and \textsf{A3 = 2}
from the input cell \textsf{A1}. Now if the user changes the \textsf{A1} to 5,
XXX will recompute both \textsf{A2 = 4}, which is necessary, but also
\textsf{A3 = 2}, which is unnecessary since its only dependency (\textsf{A1})
has not changed since the previous build.

One possible implementation of a spreadsheet build
system\footnote{For example, see \url{http://www.decisionmodels.com/calcsecretsc.htm}.}
is to maintain a
sequence of cells in which they should be evaluated. The build system takes this
sequence as input, marks all cells as \emph{unevaluated}, and then attempts to
evaluate all cells in the order specified by the sequence. If the cell to be
evaluated depends on an unevaluated cell, this means the provided sequence is
not a correct topological order and is updated by moving the current cell to the
back of the sequence. The build then proceeds with the next cell. Otherwise, if
all dependencies of the current cell have been marked as \emph{evaluated}, the
build system computes the correct value for the cell by evaluating the formula,
writes it into the spreadsheet and proceeds to the next cell in the sequence. If
there are no cyclic dependencies, this process eventually terminates with
correct results. The resulting sequence, which is guaranteed to respect the
topological order of dependencies is stored to be reused during the next build.

...

Interesting note: XXX is unusual in that it tracks changes in the compute,
i.e. if a user edits a formula, XXX will correctly rebuild results. Most build
systems, including \Shake, do not meet this requirement and a clean rebuild may
be necessary if the user edits build rules.

...

\subsubsection{Correct and minimal monadic build system}~\\

\Shake is a correct and minimal monadic build system.
\todo{NM}{Describe \Shake using our abstractions.}

\subsection{Cloud build systems}

Codebase of large software projects comprises millions of lines of code spread
across thousands of files, each built independently by thousands of developers
on thousands of machines. A distributed cloud build system speeds up builds
dramatically (and saves energy) by transparently sharing build products among
developers.

\todo{AM}{Also mention YYY (developed by ZZZ), Buck (Facebook) and Pants
(Twitter et al). Buck is essentially the same. Are they essentially the same?}

\Bazel is a cloud build system developed by Google, which supports caching of
build results. To achieve that, it maintains two partial maps:

\begin{minted}{haskell}
type Cache =  Hash  -> Maybe v
type Known = [Hash] -> Maybe Hash
\end{minted}

\hs{Cache} is a conventional \emph{content-addressable store}, that can be used
to fetch a previously computed value given its hash.

\hs{Known} records a known outcome of a computation that took a list of values
as input (represented by their hashes) and produced a resulting value (also
represented by its hash). Now, a build system has the following two options to
recompute a value:

\begin{itemize}
    \item Call the compute function on the store containing all up-to-date
    dependencies and put the computed value to the store.
    \item Collect hashes of all up-to-date dependencies \hs{hs} and query the
    \hs{Known} map. If this computation has been performed before, the map will
    contain the hash \hs{h} of one valid result (recall that the compute function
    may be non-deterministic). It is now possible to lookup \hs{h} in the
    \hs{Cache} and if it contains a value \hs{v} it can be downloaded to the
    local store instead of running the compute.
\end{itemize}

...

\todo{AM}{Also describe the XYZ algorithm.}

### 2018-03-04 src/Neil/Build.hs
Work on a new formulation of build systems

@@ -1,0 +1,89 @@
+{-# LANGUAGE Rank2Types, FlexibleContexts #-}
+
+module Neil.Build(
+    dumb,
+    dumbOnce,
+    make,
+    Shake, shake,
+    ) where
+
+import Neil.Constraints
+import Control.Monad.Extra
+import qualified Data.Set as Set
+import qualified Data.Map as Map
+
+
+-- | Dumbest build system possible, always compute everything from scratch multiple times
+dumb :: (Build m k v) => [k] -> m ()
+dumb = mapM_ f
+    where f k = maybe (getStore k) (putStore k) =<< run f k
+
+
+newtype Once k = Once (Set.Set k)
+
+once :: (Build m k v, Temp m (Once k)) => k -> m v -> m v
+once k build = do
+    Once done <- getTemp
+    if k `Set.member` done then
+        getStore k
+    else do
+        r <- build
+        updateTemp $ \(Once set) -> Once $ Set.insert k set
+        return r
+
+
+-- | Refinement of dumb, compute everything but at most once per execution
+dumbOnce :: (Build m k v, Temp m (Once k)) => [k] -> m ()
+dumbOnce = mapM_ f
+    where f k = once k $ maybe (getStore k) (putStore k) =<< run f k
+
+
+-- | The simplified Make approach where we build a dependency graph and topological sort it
+make :: (Build m k v, PreDepends m k, HasTime v) => [k] -> m ()
+make ks = do
+    deps <- depends
+    forM_ (topSort $ unroll deps ks) $ \k -> do
+        kt <- fmap getTime <$> getStoreMaybe k
+        ds <- mapM (fmap getTime . getStore) $ deps k
+        case kt of
+            Just xt | all (< xt) ds -> return ()
+            _ -> maybe (return ()) (void . putStore k) =<< run getStore k
+
+unroll :: Ord k => (k -> [k]) -> [k] -> Map.Map k [k]
+unroll deps = f Map.empty
+    where
+        f mp [] = mp
+        f mp (t:odo)
+            | t `Map.member` mp = f mp odo
+            | otherwise = let ds = deps t in f (Map.insert t ds mp) (ds ++ odo)
+
+topSort :: Ord k => Map.Map k [k] -> [k]
+topSort mp
+    | Map.null mp = []
+    | Map.null leaf = error "cycles!"
+    | otherwise = Map.keys leaf ++ topSort (Map.map (filter (`Map.notMember` leaf)) rest)
[…]

### 2018-03-04 src/Neil/Constraints.hs

@@ -1,0 +1,46 @@
+{-# LANGUAGE ConstraintKinds, MultiParamTypeClasses, FunctionalDependencies #-}
+
+module Neil.Constraints where
+
+class Monad m => Store m k v | m -> k v where
+    getStoreMaybe :: k -> m (Maybe v)
+    getStore :: k -> m v
+    putStore :: k -> v -> m v
+
+class Monad m => Run m k v | m -> k v where
+    runTrace :: (k -> m v) -> k -> m ([k], Maybe v)
+
+    run :: (k -> m v) -> k -> m (Maybe v)
+    run ask k = snd <$> runTrace ask k
+
+type Build m k v = (Store m k v, Run m k v, Ord k)
+
+class PreDepends m k | m -> k where
+    depends :: m (k -> [k])
+
+class Monad m => Temp m a where
+    getTemp :: m a
+    putTemp :: a -> m ()
+
+    updateTemp :: (a -> a) -> m ()
+    updateTemp f = putTemp . f =<< getTemp
+
+class Monad m => Info m a where
+    getInfo :: m a
+    putInfo :: a -> m ()
+
+    updateInfo :: (a -> a) -> m ()
+    updateInfo f = putInfo . f =<< getInfo
+
+
+newtype Time = Time Int
+    deriving (Eq,Ord)
+
+class HasTime v where
+    getTime :: v -> Time
+
+newtype Hash v = Hash Int
+    deriving Eq
+
+class HasHash v where
+    getHash :: v -> Hash v

### 2018-03-04 src/Neil/DynamicMap.hs

@@ -1,0 +1,24 @@
+{-# LANGUAGE Rank2Types, FlexibleContexts, ScopedTypeVariables, GeneralizedNewtypeDeriving #-}
+
+module Neil.DynamicMap(
+    DynamicMap, empty, lookup, insert
+    ) where
+
+import Prelude hiding (lookup)
+import Data.Dynamic
+import Data.Typeable
+import qualified Data.Map as Map
+
+
+newtype DynamicMap = DynamicMap (Map.Map TypeRep Dynamic)
+    deriving (Monoid, Show)
+
+
+empty :: DynamicMap
+empty = DynamicMap Map.empty
+
+insert :: Typeable a => a -> DynamicMap -> DynamicMap
+insert x (DynamicMap mp) = DynamicMap $ Map.insert (typeOf x) (toDyn x) mp
+
+lookup :: forall a . Typeable a => DynamicMap -> Maybe a
+lookup (DynamicMap mp) = flip fromDyn (error "can't happen") <$> Map.lookup (typeOf (undefined :: a)) mp

### 2018-03-04 src/Neil/Example.hs

@@ -1,10 +1,11 @@
-{-# LANGUAGE Rank2Types, ConstraintKinds, DeriveFunctor, GADTs #-}
+{-# LANGUAGE Rank2Types, ConstraintKinds, DeriveFunctor, GADTs, ScopedTypeVariables #-}
 
 module Neil.Example where
 
+import Neil.Build
 import Neil.Compute
-import Neil.Types
-import Neil.Implementation
+import Neil.Execute
+import qualified Data.Map as Map
 
 data Add k v = Add k k
              | Source
@@ -20,24 +21,21 @@ example "b" = Source
 example "c" = Add "a" "b"
 example "d" = Add "c" "c"
 
-files0 "a" = Just 1
-files0 "b" = Just 2
-files0 _ = Nothing
+store0 :: Map.Map String Int
+store0 = Map.fromList [("a",1),("b",2)]
+store' = Map.insert "a" 3
 
-files' d "a" = Just 3
-files' d x = d x
+disp (Disk a b) = "Disk " ++ show a ++ " " ++ show b
 
-disp (Disk a b) = "Disk " ++ show a ++ " " ++ show [(a, b [a]) | a <- ['a'..'d']]
-
-test :: (Show i, k ~ String, v ~ Int) => (Compute Monad k v -> [k] -> Disk i k v -> Disk i k v) -> i -> IO ()
-test build start = do
-    let d1 = build (runAdd example) ["d"] $ Disk start files0
-        d2 = build (runAdd example) ["d"] $ d1{diskFiles = files' $ diskFiles d1}
+test :: forall c k v . (k ~ String, v ~ Int) => ([k] -> M Applicative k v ()) -> IO ()
+test build = do
+    let d1 = execute (build ["d"]) (runAdd example) (Disk store0 mempty)
+        d2 = execute (build ["d"]) (runAdd example) d1{diskStore = store' $ diskStore d1}
     print $ disp d1
     print $ disp d2    
 
 main = do
-    test dumb ()
-    test dumbOnce ()
---    test make ()
-    test shake mempty
+    test dumb
+    test dumbOnce
+    -- test make
+    test shake

### 2018-03-04 src/Neil/Execute.hs

@@ -1,0 +1,47 @@
+{-# LANGUAGE ConstraintKinds, Rank2Types, FlexibleInstances, MultiParamTypeClasses, GeneralizedNewtypeDeriving #-}
+
+module Neil.Execute(
+    Disk(..),
+    M, execute,
+    ) where
+
+import Neil.Compute
+import Neil.Constraints
+import qualified Neil.DynamicMap as DM
+import Control.Monad.Trans.State
+import qualified Data.Map as Map
+
+data Disk k v = Disk
+    {diskStore :: Map.Map k v
+    ,diskInfo :: DM.DynamicMap
+    }
+
+instance Ord k => Monoid (Disk k v) where
+    mempty = Disk mempty mempty
+    mappend (Disk x1 x2) (Disk y1 y2) = Disk (mappend x1 y1) (mappend x2 y2)
+
+execute :: M c k v () -> Compute c k v -> Disk k v -> Disk k v
+execute (M m) c d = disk $ execState m $ S c d DM.empty
+
+data S c k v = S
+    {compute :: Compute c k v
+    ,disk :: Disk k v
+    ,temp :: DM.DynamicMap
+    }
+
+newtype M c k v r = M (State (S c k v) r)
+    deriving (Functor, Applicative, Monad)
+
+instance Run (M c k v) k v where
+
+instance Store (M c k v) k v where
+
+instance Temp (M c k v) t where
+
+instance Info (M c k v) i where
+
+instance HasHash Int where
+
+instance HasTime (Timed v) where
+
+data Timed v = Timed Time v

### 2018-03-04 paper/build-abstractions.tex

\clearpage

\section{Build systems, abstractly}\label{sec-build-abstractions}

In this section we present the \emph{compute} and \emph{build} abstractions:
\begin{itemize}
    \item Compute represents build rules (in build systems) or formulas (in
    spreadsheets), and is completely isolated from the world of compilers, file
    systems, memories, caches, and all other complexities of real build systems.
    \item Build corresponds to the algorithm used for bringing a key-value store
    to a coherent state by running compute.
\end{itemize}

\begin{figure}
\begin{minted}{haskell}
type Compute f k v = (k -> f v) -> k -> f (Maybe v)

type FunctorialCompute  f k v = Functor     f => Compute f k v
type ApplicativeCompute f k v = Applicative f => Compute f k v
type AlternativeCompute f k v = Alternative f => Compute f k v
type MonadicCompute     f k v = Monad       f => Compute f k v
type MonadPlusedCompute f k v = MonadPlus   f => Compute f k v
\end{minted}
\caption{Compute abstractions}\label{fig-compute}
\end{figure}

Consider abstractions in Fig.~\ref{fig-compute}.

...

We now describe several types of build systems according to their constraint
on the functor \hs{f}.

...

### 2018-03-05 paper/background.tex
Expand introduction

@@
 
-We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash v} for
+We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for
 efficient tracking and sharing of build results.
@@
 
-Many existing build systems need to know all the dependencies between files or
-cells, i.e. the dependency graph, at the start of the build process. This makes
-it possible to analyse the graph ahead of time, but is fundamentally limited:
-parts of the dependency graph can often be discovered only during the build
-process, i.e. the dependency graph is dynamic, not static. This is illustrated
-by the following example.
+...
+
 \todo{AM}{Add two examples: include files and cyclic spreadsheet computations.}

### 2018-03-05 paper/intro.tex
Expand introduction

@@
 \section{Introduction}\label{sec-intro}
+
+Build systems automate the execution of simple repeatable tasks and are used by
+many individuals and organisations. There are software build systems such as
+\Make, \Ninja, \Shake, \Buck, \Bazel and many others, as well as various
+incremental calculation engines, such as \Excel. Existing build systems vary in
+several different ways:
+
+\begin{itemize}
+    \item Many build systems, such as the venerable \Make, need to know the
+    complete \emph{dependency graph} between tasks at the start of the build
+    process. This makes it possible to analyse the graph ahead of time, which
+    helps with efficient task scheduling but is fundamentally limited: parts of
+    the dependency graph can often be discovered only during the build process,
+    i.e. the dependency graph is \emph{dynamic}, not \emph{static}.
+
+    \item When build systems are used by large teams, different team members
+    often end up executing exactly the same tasks on their local machines.
+    A \emph{cloud build system} can speed up builds dramatically by
+    transparently sharing build results among team members. Furthermore, cloud
+    build systems allow one to perform \emph{shallow builds} which materialise
+    only end build products on a local machine, leaving all the intermediates
+    in the cloud. This is a significant optimisation compared to \emph{deep
+    builds} that require all transitive dependencies of an end build product to
+    be locally available.
+
+    \item Some build systems, e.g. \Buck, require all tasks to be
+    \emph{deterministic}, i.e. produce exactly the same output when run on the
+    same inputs. However, not all tasks are deterministic, and there are build
+    systems that support \emph{non-determinism}. A simple example is \Excel's
+    \textsf{RANDBETWEEN(low, high)} that returns a random integer in the
+    interval \textsf{[low, high]}.
+
+    \item If a build system executes a task and the resulting value is unchanged
+    from the previous run, it can skip the execution of all tasks that depend on
+    the value. We call this \emph{early cut-off}. Not all build systems support
+    early cut-off: \Make and \Excel do not, whereas \Shake and \Buck do.
+
+    \item Most build systems are designed to track changes of values and rerun
+    dependent tasks when they change, but some build systems can also track
+    changes in the tasks themselves: if a task is changed, the build system
+    will rerun it. For example, when a cell's formula is changed, \Excel will
+    recompute its value and propagate the changes. This is uncommon in software
+    build systems, where one typically needs to do a full rebuild when changing
+    a task.
+
+    \item Most build systems store auxiliary information between builds for
+    profiling and optimisation purposes: \Make stores file modification times,
+    \Shake stores the discovered dynamic dependency graph, etc.
+
+    % \item Fine-grain dependencies
+\end{itemize}
+
+% Build systems solve a complex
+% problem and have complex implementations. Different build systems take
+
+...
+
+\todo{AM}{The contribution of this paper is...}
+
+...
+
+Some text copied from the fellowship proposal, not sure whether anything is useful:
 

### 2018-03-06 paper/intro.tex
Work on intro

--- 2018-02-28 paper/intro.tex
@@
 
-Build systems are awesome. Build systems are terrifying.
+Build systems automate the execution of simple repeatable tasks for individual
+users, as well as for large organisations. There are software build systems such
+as \Make, \Ninja, \Shake, \Bazel, \Buck and many others, as well as various
+incremental calculation engines, such as \Excel and \Calc. Existing build
+systems vary in several different ways:
 
-We separate build systems into \emph{compute} and \emph{build} components,
-see~\S\ref{sec-compute}.
+\begin{itemize}
+    \item Many build systems, such as the venerable \Make, need to know the
+    complete \emph{dependency graph} between tasks at the start of the build
+    process. This makes it possible to analyse the graph ahead of time, which
+    helps with efficient task scheduling but is fundamentally limited: parts of
+    the dependency graph can often be discovered only during the build process,
+    i.e. the dependency graph is \emph{dynamic}, not \emph{static}.
+
+    \item When build systems are used by large teams, different team members
+    often end up executing exactly the same tasks on their local machines.
+    A \emph{cloud build system} can speed up builds dramatically by
+    transparently sharing build results among team members. Furthermore, cloud
+    build systems allow one to perform \emph{shallow builds} that materialise
+    only end build products on a local machine, leaving all intermediates in the
+    cloud. This is a significant optimisation compared to \emph{deep builds}
+    that require all transitive dependencies of an end build product to be
+    locally available. Non-cloud build systems cannot support shallow builds.
+
+    \item Some build systems, e.g. \Buck, require all tasks to be
+    \emph{deterministic}, i.e. produce exactly the same output when run on the
+    same inputs. However, not all tasks are deterministic, and there are build
+    systems that support \emph{non-determinism}. A simple example is \Excel's
+    \textsf{RANDBETWEEN(low, high)} that returns a random integer in the
+    interval \textsf{[low, high]}.
+
+    \item If a build system executes a task and the result is unchanged from the
+    previous build, it is unnecessary to execute the dependent tasks. We call
+    this optimisation \emph{early cut-off}. Not all build systems support early
+    cut-off: \Make and \Excel do not, whereas \Shake and \Buck do.
+
+    \item Most build systems track changes of inputs and intermediate results,
+    executing dependent tasks whenever they change, but some build systems can
+    also track changes in the tasks themselves: if a task has changed, the build
+    system will execute it. For example, when a cell's formula has changed,
+    \Excel will recompute its value and propagate the changes. We call this
+    \emph{self-tracking}. Self-tracking is uncommon in software build systems,
+    where one often needs to manually initiate a rebuild of all tasks even if
+    just a single task has changed.
+
+    \item Most build systems persistently store auxiliary \emph{build
+    information} for profiling and optimisation purposes: \Make stores file
+    modification times, \Shake stores the discovered dynamic dependency graph,
+    \Bazel and other cloud build systems store information about inputs and
+    outputs of previously executed tasks, etc.
+\end{itemize}
+
+This paper presents a purely functional abstraction for build systems that
+allows us to express all the above intricacies of build systems and design
+complex build systems from simple primitives. The presented abstraction fits in
+just two lines of Haskell code, which are explained
+in~\S\ref{sec-build-abstractions}:
+
+\begin{minted}{haskell}
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> f (Maybe v)
+type Build c i k v = Compute c k v -> [k] -> Maybe i -> Map k v -> (i, Map k v)
+\end{minted}
+
+% It is known that build systems can take up to 27\% of software development
+% effort, and that improvements to build systems rapidly pay off~\cite{build_maintenance}.
+% Despite its importance, this subject is severely under-researched, which prompts
+% major companies, such as Microsoft, Facebook and Google, to invest significant
+% internal resources to make their own bespoke build system frameworks.
+
+% Some build systems do not look like build systems but they are. A good example
+% is spreadsheets, where cells play the role of files, and formulas play the role
+% of build rules.
+
+% We separate build systems into \emph{compute} and \emph{build} components,
+% see~\S\ref{sec-build-abstractions}.

### 2018-03-06 src/Neil/Builder.hs
Add a recursive combinator

@@ -56,15 +56,17 @@ newtype Recursive k = Recursive (Set.Set k)
     deriving Default
 
 -- | Build a rule at most once in a single execution
-recursive :: (Show k, Ord k, Typeable k) => k -> M i k v v -> M i k v v
-recursive k build = do
-    Recursive done <- getTemp
-    if k `Set.member` done then
-        getStore k
-    else do
-        r <- build
-        modifyTemp $ \(Recursive set) -> Recursive $ Set.insert k set
-        return r
+recursive :: Default i => (k -> (k -> M i k v v) -> M i k v ([k], v) -> M i k v ()) -> Build Monad i k v
+recursive step compute = runM . ensure
+    where
+        ensure k = do
+            let ask x = ensure x >> getStore x
+            Recursive done <- getTemp
+            when (k `Set.notMember` done) $ do
+                modifyTemp $ \(Recursive set) -> Recursive $ Set.insert k set
+                case trackDependencies compute ask k of
+                    Nothing -> return ()
+                    Just act -> step k ask act
 
 
 -- | Figure out when files change, like a modtime
@@ -98,8 +100,7 @@ dumb compute = runM . f
 
 -- | Refinement of dumb, compute everything but at most once per execution
 dumbOnce :: Build Monad () k v
-dumbOnce compute = runM . f
-    where f k = recursive k $ maybe (getStore k) (putStore k =<<) $ compute f k
+dumbOnce = recursive $ \k _ act -> putStore_ k . snd =<< act
 
 
 -- | The simplified Make approach where we build a dependency graph and topological sort it
@@ -130,23 +131,18 @@ type Shake k v = Map.Map k (Hash v, [(k, Hash v)])
 
 -- | The simplified Shake approach of recording previous traces
 shake :: Hashable v => Build Monad (Shake k v) k v
-shake compute = runM . f
-    where
-        f k = recursive k $ do
-            info <- getInfo
-            valid <- case Map.lookup k info of
-                Nothing -> return False
-                Just (me, deps) ->
-                    (maybe False (== me) <$> getStoreHashMaybe k) &&^
-                    allM (\(d,h) -> (== h) . getHash <$> f d) deps
-            case trackDependencies compute f k of
-                Just res | not valid -> do
-                    (ds, v) <- res
-                    putStore k v
-                    dsh <- mapM (fmap getHash . getStore) ds
-                    modifyInfo $ Map.insert k (getHash v, zip ds dsh)
[…]

### 2018-03-07 paper/intro.tex
Switch to new types
Add Simon's text

@@
 \section{Introduction}\label{sec-intro}
+
+Build systems (such as make) are big, complicated, but unloved part of
+the software ecosystem.  Every developer on the planet uses one, but
+they are very much a means to an end, and seldom the focus of
+attention.  But the challenges of scale have driven large software firms
+like Microsoft, Facebook, and Google to develop their own build
+systems, each with its own choices and idiosyncracies.
+
+Seldom do people ask questions like "What does it mean for my build
+system to be correct?" or "What are the trade-offs between different
+approaches?".  Complex build systems use subtle algorithms, but they
+are often hidden away, and not the object of study.
+
+In this paper we offer a general framework in which to discuss these
+questions in a way that is both absract (omitting incidental detail)
+and yet precise (implemented as Haskell code).  Specifically we make
+these contributions:
+\begin{itemize}
+\item We describe some simple but novel abstractions that
+crisply encapsulate what a build system is.
+\item We show that we can instantiate
+these abstractions to describe the essence of a variety of different
+build systems, including make, Shake, Buck, Bazel, and Excel, each in
+a dozen lines of code or so.
+Doing this in a single setting allows
+the differences and similarities between these huge systems to be
+brought out clearly.
+\item Build systems vary on many axes;
+for example: static vs dynamic dependencies; cloud-build, including
+shallow vs deep; deterministic vs non-deterministic build rules;
+early cut-off; self-tracking build systems; and persistent metadata.
+These properties (which we define in Section~\ref{xxx}) are often
+deeply-built-in assumptions of a particular build system.
+In contrast, our framework allows them to be distinguished,
+reasoned about, and varied.
+\item Two particularly desirable properties are dynamic dependencies
+and cloud-build; yet no currently-available build system supports
+both.  Our framework makes it rather easy to do so, for the first
+time.
+\end{itemize}
+Thus equipped, instead of seeing build systems as unrelated
+points in space, we can re-envisage them as locations in a landscape,
+leading to better understanding of what they do and how they compare,
+and suggesting exploration of other (as yet unoccupied points) in the
+landscape.
+
+Papers about "frameworks" are often fuzzy.  This one is not: all our
+abstractions are defined in Haskell, and we have (freely-available)
+executable models of all the build systems we describe.  An unusual
+feature is that we include Excel in our line-up because, looked at
+in the right way, it certainly is a build system.
+
+... Old text ...
 
@@
 \begin{minted}{haskell}
-type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> f (Maybe v)
-type Build c i k v = Compute c k v -> [k] -> Maybe i -> Map k v -> (i, Map k v)
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
 \end{minted}

### 2018-03-07 paper/engineering.tex
Restructure paper

\clearpage
\section{Engineering aspects}\label{sec-engineering}

\begin{itemize}
    \item Sandboxing
    \item Non-determinism, specifically, it's useless form. This doesn't
          include \Excel's \textsf{RANDBETWEEN} function, which is useful.
          To define correctness of non-deterministic build systems, one needs to
          switch from \hs{Applicative} and \hs{Monad} abstractions to
          \hs{Alternative} and \hs{MonadPlus}, respectively, but this is not
          required for actual implementations, most of which can happily accept
          non-deterministic results (with a notable exception of \Buck).
    \item Concurrency
    \item Cloud
\end{itemize}

### 2018-03-07 paper/related.tex
Restructure paper

\clearpage
\section{Related work}\label{sec-related}

\begin{itemize}
    \item Memoization
    \item Self-adjusting computation
    \item Lens-like formulation
\end{itemize}

### 2018-03-07 paper/abstractions.tex
Restructure paper
Move a part of intro to background

--- 2018-03-04 paper/build-abstractions.tex
@@
 
-\section{Build systems, abstractly}\label{sec-build-abstractions}
+\section{Build systems, abstractly}\label{sec-abstractions}
 
@@
 \begin{minted}{haskell}
-type Compute f k v = (k -> f v) -> k -> f (Maybe v)
-
-type FunctorialCompute  f k v = Functor     f => Compute f k v
-type ApplicativeCompute f k v = Applicative f => Compute f k v
-type AlternativeCompute f k v = Alternative f => Compute f k v
-type MonadicCompute     f k v = Monad       f => Compute f k v
-type MonadPlusedCompute f k v = MonadPlus   f => Compute f k v
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
 \end{minted}

### 2018-03-07 paper/background.tex
Move a part of intro to background

@@
 \clearpage
+\section{Background}\label{sec-background}
 
-\section{Background}\label{sec-background}
+Build systems automate the execution of simple repeatable tasks for individual
+users, as well as for large organisations. There are software build systems such
+as \Make, \Ninja, \Shake, \Bazel, \Buck and many others, as well as various
+incremental calculation engines, such as \Excel and \Calc. Existing build
+systems vary in several different ways:
+
+\begin{itemize}
+    \item Many build systems, such as the venerable \Make, need to know the
+    complete \emph{dependency graph} between tasks at the start of the build
+    process. This makes it possible to analyse the graph ahead of time, which
+    helps with efficient task scheduling but is fundamentally limited: parts of
+    the dependency graph can often be discovered only during the build process,
+    i.e. the dependency graph is \emph{dynamic}, not \emph{static}.
+
+    \item When build systems are used by large teams, different team members
+    often end up executing exactly the same tasks on their local machines.
+    A \emph{cloud build system} can speed up builds dramatically by
+    transparently sharing build results among team members. Furthermore, cloud
+    build systems allow one to perform \emph{shallow builds} that materialise
+    only end build products on a local machine, leaving all intermediates in the
+    cloud. This is a significant optimisation compared to \emph{deep builds}
+    that require all transitive dependencies of an end build product to be
+    locally available. Non-cloud build systems cannot support shallow builds.
+
+    \item Some build systems, e.g. \Buck, require all tasks to be
+    \emph{deterministic}, i.e. produce exactly the same output when run on the
+    same inputs. However, not all tasks are deterministic, and there are build
+    systems that support \emph{non-determinism}. A simple example is \Excel's
+    \textsf{RANDBETWEEN(low, high)} that returns a random integer in the
+    interval \textsf{[low, high]}.
+
+    \item If a build system executes a task and the result is unchanged from the
+    previous build, it is unnecessary to execute the dependent tasks. We call
+    this optimisation \emph{early cut-off}. Not all build systems support early
+    cut-off: \Make and \Excel do not, whereas \Shake and \Buck do.
+
+    \item Most build systems track changes of inputs and intermediate results,
+    executing dependent tasks whenever they change, but some build systems can
+    also track changes in the tasks themselves: if a task has changed, the build
+    system will execute it. For example, when a cell's formula has changed,
+    \Excel will recompute its value and propagate the changes. We call this
+    \emph{self-tracking}. Self-tracking is uncommon in software build systems,
+    where one often needs to manually initiate a rebuild of all tasks even if
+    just a single task has changed.
+
+    \item Most build systems persistently store auxiliary \emph{build
+    information} for profiling and optimisation purposes: \Make stores file
+    modification times, \Shake stores the discovered dynamic dependency graph,
+    \Bazel and other cloud build systems store information about inputs and
+    outputs of previously executed tasks, etc.
+\end{itemize}
+
+This paper presents a purely functional abstraction for build systems that
+allows us to express all the above intricacies of build systems and design
+complex build systems from simple primitives. The presented abstraction fits in
+just two lines of Haskell code, which are explained
+in~\S\ref{sec-abstractions}:
+
+\begin{minted}{haskell}
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
+\end{minted}
 

### 2018-03-07 paper/intro.tex
Move a part of intro to background

@@
 like Microsoft, Facebook, and Google to develop their own build
-systems, each with its own choices and idiosyncracies.
+systems, each with its own choices and idiosyncrasies.
 
@@
 In this paper we offer a general framework in which to discuss these
-questions in a way that is both absract (omitting incidental detail)
+questions in a way that is both abstract (omitting incidental detail)
 and yet precise (implemented as Haskell code).  Specifically we make
@@
 these abstractions to describe the essence of a variety of different
-build systems, including make, Shake, Buck, Bazel, and Excel, each in
+build systems, including \Make, \Shake, \Bazel, and \Excel, each in
 a dozen lines of code or so.
@@
 early cut-off; self-tracking build systems; and persistent metadata.
-These properties (which we define in Section~\ref{xxx}) are often
+These properties (which we define in Section~\ref{sec-background}) are often
 deeply-built-in assumptions of a particular build system.
@@
 
-... Old text ...
-
-Build systems automate the execution of simple repeatable tasks for individual
-users, as well as for large organisations. There are software build systems such
-as \Make, \Ninja, \Shake, \Bazel, \Buck and many others, as well as various
-incremental calculation engines, such as \Excel and \Calc. Existing build
-systems vary in several different ways:
-
-\begin{itemize}
-    \item Many build systems, such as the venerable \Make, need to know the
-    complete \emph{dependency graph} between tasks at the start of the build
-    process. This makes it possible to analyse the graph ahead of time, which
-    helps with efficient task scheduling but is fundamentally limited: parts of
-    the dependency graph can often be discovered only during the build process,
-    i.e. the dependency graph is \emph{dynamic}, not \emph{static}.
-
-    \item When build systems are used by large teams, different team members
-    often end up executing exactly the same tasks on their local machines.
-    A \emph{cloud build system} can speed up builds dramatically by
-    transparently sharing build results among team members. Furthermore, cloud
-    build systems allow one to perform \emph{shallow builds} that materialise
-    only end build products on a local machine, leaving all intermediates in the
-    cloud. This is a significant optimisation compared to \emph{deep builds}
-    that require all transitive dependencies of an end build product to be
-    locally available. Non-cloud build systems cannot support shallow builds.
-
-    \item Some build systems, e.g. \Buck, require all tasks to be
-    \emph{deterministic}, i.e. produce exactly the same output when run on the
-    same inputs. However, not all tasks are deterministic, and there are build
-    systems that support \emph{non-determinism}. A simple example is \Excel's
-    \textsf{RANDBETWEEN(low, high)} that returns a random integer in the
-    interval \textsf{[low, high]}.
-
-    \item If a build system executes a task and the result is unchanged from the
-    previous build, it is unnecessary to execute the dependent tasks. We call
-    this optimisation \emph{early cut-off}. Not all build systems support early
-    cut-off: \Make and \Excel do not, whereas \Shake and \Buck do.
-
-    \item Most build systems track changes of inputs and intermediate results,
-    executing dependent tasks whenever they change, but some build systems can
-    also track changes in the tasks themselves: if a task has changed, the build
-    system will execute it. For example, when a cell's formula has changed,
-    \Excel will recompute its value and propagate the changes. We call this
-    \emph{self-tracking}. Self-tracking is uncommon in software build systems,
-    where one often needs to manually initiate a rebuild of all tasks even if
-    just a single task has changed.
-
-    \item Most build systems persistently store auxiliary \emph{build
-    information} for profiling and optimisation purposes: \Make stores file
-    modification times, \Shake stores the discovered dynamic dependency graph,
-    \Bazel and other cloud build systems store information about inputs and
-    outputs of previously executed tasks, etc.
-\end{itemize}
-
-This paper presents a purely functional abstraction for build systems that
-allows us to express all the above intricacies of build systems and design
-complex build systems from simple primitives. The presented abstraction fits in
-just two lines of Haskell code, which are explained
-in~\S\ref{sec-build-abstractions}:
-
-\begin{minted}{haskell}
-type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
-\end{minted}
-
-% It is known that build systems can take up to 27\% of software development
-% effort, and that improvements to build systems rapidly pay off~\cite{build_maintenance}.
-% Despite its importance, this subject is severely under-researched, which prompts
-% major companies, such as Microsoft, Facebook and Google, to invest significant
-% internal resources to make their own bespoke build system frameworks.
-
-% Some build systems do not look like build systems but they are. A good example
-% is spreadsheets, where cells play the role of files, and formulas play the role
-% of build rules.
-
-% We separate build systems into \emph{compute} and \emph{build} components,
-% see~\S\ref{sec-build-abstractions}.
+\todo{AM/NM}{Add a paragraph on the decomposition of build systems into
+\emph{compute} and \emph{build} components, with the latter split into
+\emph{when} and \emph{how} parts.}

### 2018-03-07 src/Neil/Build.hs
A few tweaks to the implementation

@@ -26,9 +26,9 @@ import qualified Data.Map as Map
 type Build c i k v = (Ord k, Show k, Typeable k) => Compute c k v -> k -> Maybe i -> Map.Map k v -> (i, Map.Map k v)
 
 
-runM :: (Ord k, Default i) => M i k v a -> Maybe i -> Map.Map k v -> (i, Map.Map k v)
+runM :: (Default i) => M i k v a -> Maybe i -> Map.Map k v -> (i, Map.Map k v)
 runM (M m) i s = (info res, store res)
-    where res = execState m $ S s (fromMaybe def i) mempty mempty
+    where res = execState m $ S s (fromMaybe def i) mempty Set.empty
 
 data S i k v = S
     {store :: Map.Map k v

### 2018-03-07 src/Neil/Builder.hs

@@ -7,6 +7,7 @@ module Neil.Builder(
     dumbRecursive,
     make,
     makeTrace,
+    makeDirtyBit,
     shake,
     shakeDirtyBit,
     spreadsheet,
@@ -111,6 +112,13 @@ make = withChangedApplicative $ topological $ \k ds act -> do
     case kt of
         Just xt | all (<= xt) ds -> return ()
         _ -> putStore_ k =<< act
+
+makeDirtyBit :: Eq v => Build Applicative (Changed k v, ()) k v
+makeDirtyBit = withChangedApplicative $ topological $ \k ds act -> do
+    dirty <- fmap isNothing (getStoreMaybe k) ||^ getChanged k ||^ anyM getChanged ds
+    when dirty $
+        putStore_ k =<< act
+
 
 type MakeHash k v = Map.Map (k, [Hash v]) (Hash v)
 

### 2018-03-07 src/Neil/Example.hs

@@ -43,6 +43,7 @@ main = do
     test dumbRecursive
     test make
     test makeTrace
+    test makeDirtyBit
     test shake
     test shakeDirtyBit
     test spreadsheet

### 2018-03-07 paper/implementations.tex
Start working on the implementation.tex file

--- 2018-03-02 paper/build-examples.tex
@@
 \section{Build systems, concretely}\label{sec-examples}
+
+In the previous sections we've discussed the types of build system in text, but the way we came to our conclusions was by developing a framework to fit the builds in.
+In this section we explain the framework, some of the abstractions we introduced (which model the properties from \S\ref{todo}), and how we can make a meaningful implementation of the build systems discussed in \S\ref{sec-background}.
+
+\subsection{Framework}
+
+As we saw in the introduction, a build system can be defined as:
+
+\begin{minted}{haskell}
+type Build c i k v = Compute c k v -> k -> Maybe i -> Map.Map k v -> (i, Map.Map k v)
+\end{minted}
+
+Using the framework we have built we can express the simplest and stupidest build system as:
+
+\begin{minted}{haskell}
+dumb :: Build Monad () k v
+dumb compute = runM . f
+    where f k = maybe (getStore k) (putStore k =<<) $ compute f k
+\end{minted}
+
+\begin{figure}
+\begin{minted}{haskell}
+data M i k v r
+runM :: Default i => M i k v a -> Maybe i -> Map.Map k v -> (i, Map.Map k v)
+getStore :: k -> M i k v v
+putStore :: k -> v -> M i k v v
+\end{minted}
+Note: we have omitted the context on \textt{k}, things like \texttt{Ord} (so the implementation can use a finite map) or \texttt{Show} (so the implementation can produce nice error messages).
+\caption{API with which to implement build systems}
+\label{fig-M-api}
+\end{figure}
+
+The types of important definitions are given in Figure \ref{fig-M-api}
+
+but
 
@@
 \todo{NM}{Add a description of \Ninja.}
+
 \todo{AM}{What about Nix? John Ericson suggested in a blog comment that it may

### 2018-03-07 src/Development/Build/Compute.hs
Simplify

@@ -15,8 +15,8 @@ type Compute c k v = forall f. c f => (k -> f v) -> k -> Maybe (f v)
 inputCompute :: (k -> f v) -> k -> Maybe (f v)
 inputCompute _ _ = Nothing
 
--- TODO: How do we say it works for any possible constraint?
-isInput :: Compute MonadPlus k v -> k -> Bool
+-- TODO: How do we express this in terms of Compute? Drop forall in Compute?
+isInput :: ((k -> Proxy v) -> k -> Maybe (Proxy v)) -> k -> Bool
 isInput compute = isNothing . compute (const Proxy)
 
 --------------------------- Compute Functor: Collatz ---------------------------

### 2018-03-07 src/Development/Build/Compute/Applicative.hs

@@ -1,9 +1,11 @@
 {-# LANGUAGE RankNTypes #-}
 module Development.Build.Compute.Applicative (
-    pureCompute, dependencies, transitiveDependencies, acyclic, runPartial
+    pureCompute, dependencies, transitiveDependencies, acyclic, debugPartial,
+    partial, exceptional
     ) where
 
 import Control.Applicative
+import Data.Functor.Compose
 import Data.Maybe
 
 import Development.Build.Compute
@@ -27,8 +29,24 @@ acyclic compute = isJust . transitiveDependencies compute
 -- that the compute failed due to a missing dependency @k@. Otherwise, the
 -- result @Right (Just v)@ yields the computed value, and @Right Nothing@ is
 -- returned if the given key is an input.
-runPartial :: Applicative f => Compute Applicative k v
+debugPartial :: Applicative f => Compute Applicative k v
                             -> (k -> f (Maybe v)) -> k -> Maybe (f (Either k v))
-runPartial compute partialGet = fmap runEitherT . compute get
+debugPartial compute partialGet = fmap getCompose . compute (Compose . get)
   where
-    get k = EitherT $ maybe (Left k) Right <$> partialGet k
+    get k = maybe (Left k) Right <$> partialGet k
+
+-- | Convert a compute with a total lookup function @k -> m v@ into a compute
+-- with a partial lookup function @k -> m (Maybe v)@. This essentially lifts the
+-- compute from the type of values @v@ to @Maybe v@, where the result @Nothing@
+-- indicates that the compute failed because of a missing dependency.
+-- Use 'debugPartial' if you need to know which dependency was missing.
+partial :: Compute Applicative k v -> Compute Applicative k (Maybe v)
+partial compute get = fmap getCompose . compute (Compose . get)
+
+-- | Convert a compute with a total lookup function @k -> m v@ into a compute
+-- with a lookup function that can throw exceptions @k -> m (Either e v)@. This
+-- essentially lifts the compute from the type of values @v@ to @Either e v@,
+-- where the result @Left e@ indicates that the compute failed because of a
+-- failed dependency lookup, and @Right v@ yeilds the value otherwise.
+exceptional :: Compute Applicative k v -> Compute Applicative k (Either e v)
+exceptional compute get = fmap getCompose . compute (Compose . get)

### 2018-03-07 src/Neil/Build.hs
Simplify the code a bit for the paper

@@ -3,7 +3,7 @@
 module Neil.Build(
     Build,
     M, runM,
-    getStoreMaybe, getStore, putStore, putStore_,
+    getStoreMaybe, getStore, putStore,
     Changed, getChanged, withChangedApplicative, withChangedMonad,
     Time, getStoreTime, getStoreTimeMaybe,
     getInfo, putInfo, modifyInfo,
@@ -74,11 +74,8 @@ getStoreMaybe k = Map.lookup k . store <$> M get
 getStore :: (Ord k, Show k) => k -> M i k v v
 getStore k = fromMaybe (error $ "getStore failed on " ++ show k) <$> getStoreMaybe k
 
-putStore :: Ord k => k -> v -> M i k v v
-putStore k v = do putStore_ k v; return v
-
-putStore_ :: Ord k => k -> v -> M i k v ()
-putStore_ k v = M $ modify $ \x -> x{store = Map.insert k v $ store x, changed = Set.insert k $ changed x}
+putStore :: Ord k => k -> v -> M i k v ()
+putStore k v = M $ modify $ \x -> x{store = Map.insert k v $ store x, changed = Set.insert k $ changed x}
 
 getTemp :: (Typeable t, Default t) => M i k v t
 getTemp = fromMaybe def . DM.lookup . temp <$> M get

### 2018-03-07 src/Neil/Builder.hs

@@ -1,4 +1,4 @@
-{-# LANGUAGE Rank2Types, FlexibleContexts, GeneralizedNewtypeDeriving, ScopedTypeVariables, RecordWildCards, ConstraintKinds #-}
+{-# LANGUAGE Rank2Types, FlexibleContexts, GeneralizedNewtypeDeriving, ScopedTypeVariables, RecordWildCards, ConstraintKinds, TypeFamilies #-}
 
 module Neil.Builder(
     dumb,
@@ -60,7 +60,9 @@ recursive step compute = runM . ensure
                     Just act -> step k (getDependenciesMaybe compute k) ask act
 
 
-dynamic :: Default i => (k -> Maybe [k] -> (k -> M (i, [k]) k v (Maybe v)) -> M (i, [k]) k v (Either k ([k], v)) -> M (i, [k]) k v (Maybe k)) -> Build Monad (i, [k]) k v
+dynamic
+    :: (m ~ M (i, [k]) k v, Default i)
+    => (k -> Maybe [k] -> (k -> m (Maybe v)) -> m (Either k ([k], v)) -> m (Maybe k)) -> Build Monad (i, [k]) k v
 dynamic step compute k = runM $ do
     order <- snd <$> getInfo
     order <- f Set.empty $ order ++ [k | k `notElem` order]
@@ -84,15 +86,15 @@ dynamic step compute k = runM $ do
 
 -- | Dumbest build system possible, always compute everything from scratch multiple times
 dumb :: Build Monad () k v
-dumb compute = runM . f
-    where f k = maybe (getStore k) (putStore k =<<) $ compute f k
+dumb compute k = runM (f k)
+    where f k = maybe (getStore k) (\act -> do v <- act; putStore k v; return v) $ compute f k
 
 -- | Refinement of dumb, compute everything but at most once per execution
 dumbRecursive :: Build Monad () k v
-dumbRecursive = recursive $ \k _ _ act -> putStore_ k . snd =<< act
+dumbRecursive = recursive $ \k _ _ act -> putStore k . snd =<< act
 
 dumbTopological :: Build Applicative () k v
-dumbTopological = topological $ \k _ act -> putStore_ k =<< act
+dumbTopological = topological $ \k _ act -> putStore k =<< act
 
 dumbDynamic :: Build Monad ((), [k]) k v
 dumbDynamic = dynamic $ \k _ _ act -> do
[…]

### 2018-03-07 paper/implementations.tex
Initial stab at the implementations section

@@
 dumb :: Build Monad () k v
-dumb compute = runM . f
-    where f k = maybe (getStore k) (putStore k =<<) $ compute f k
+dumb compute k = runM (f k)
+    where f k = maybe (getStore k) (\act -> do v <- act; putStore k v; return v) $ compute f k
 \end{minted}
@@
 getStore :: k -> M i k v v
-putStore :: k -> v -> M i k v v
+putStore :: k -> v -> M i k v ()
 \end{minted}
-Note: we have omitted the context on \textt{k}, things like \texttt{Ord} (so the implementation can use a finite map) or \texttt{Show} (so the implementation can produce nice error messages).
+Note: we have omitted the context on \hs{k}, things like \hs{Ord} (so the implementation can use a finite map) or \hs{Show} (so the implementation can produce nice error messages).
 \caption{API with which to implement build systems}
@@
 
-The types of important definitions are given in Figure \ref{fig-M-api}
+The types of important definitions are given in Figure \ref{fig-M-api}.
+In essence we take the pure \hs{Map} to \hs{Map} interface of \hs{Build} and instead provide a Monad \hs{M} which keeps track of the \hs{i} info and \hs{Map k v} as a store.
+We can then provide functions like \hs{getStore} to load from the store and \hs{putStore} to write to the store.
 
-but
+With that the \hs{dumb} build from above should start to make sense.
+The function \hs{dumb} takes \hs{compute} and \hs{k}, then calls \hs{runM} on \hs{f x}, where \hs{f x} has type \hs{M () k v v}.
+The \hs{runM} function takes care of putting the information into a state monad, running the computation, then extracting it afterwards - to match the signature of \hs{build}.
 
-This section present several concrete examples of build systems, providing
-simple implementations that use the previously introduced abstractions.
+We'll introduce the other functions in in the monad \hs{M} as we need them while exploring the build systems.
 
-\subsection{Applicative build systems}
+\subsection{\Make}
 
-Applicative build system can only accept an applicative compute, because they
-rely on building the full dependency graph upfront, which is impossible with
-dynamic dependencies.
+The \Make build system builds files in a linear order based on a topological sort.
+For each file, it builds it if the output file does not exist or is older than any of the inputs.
+We can express this in our framework with:
 
-\vspace{4mm}
-\subsubsection{An incorrect applicative build system}~\\
-
-\todo{AM}{Describe the \hs{dumbBuild} that builds outputs in the given order.}
-
-\vspace{4mm}
-\subsubsection{\Make: a correct but non-minimal applicative build system}~\\
-
-We start by applicative build systems, such as the classic vanilla
-\Make\footnote{There are numerous implementations of \Make and none comes with a
-formal specification. In this paper we therefore use a simple but reasonable
-approximation to a real \Make that you might find on your machine.} and a more
-recent \Ninja.
-
-Make is correct but non-minimal. It is unusual from other build systems in that
-it relies on \emph{timed values}, i.e. on a map \hs{age :: v -> Time}, as a
-heuristic to decide whether a value is up-to-date.
-
-\todo{NM}{Add a description of \Make.}
-
-\vspace{4mm}
-\subsubsection{\Ninja: a correct and minimal applicative build system}~\\
-
-\Ninja is a modern alternative to \Make ...
-
-\todo{NM}{Add a description of \Ninja.}
-
-\todo{AM}{What about Nix? John Ericson suggested in a blog comment that it may
-be somewhat monadic, see:
-\url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.}
-
-\subsection{Monadic build systems}
-
-...
-\subsubsection{An incorrect monadic build system}~\\
-
-See \hs{dumbBuild}.
-
-\subsubsection{Correct but not minimal monadic build system}~\\
-
-As of writing, XXX is a correct but non-minimal monadic build system. Here is
-an example where it does unnecessary computation:
-
-\begin{minted}[frame=single]{text}
-A1 = 10
-A2 = MIN(A1, 4)
-A3 = SQRT(A2)
+\begin{minted}{haskell}
+make :: Eq v => Build Applicative (Changed k v, ()) k v
+make = withChangedApplicative $ topological $ \k ds act -> do
+    kt <- getStoreTimeMaybe k
+    ds <- mapM getStoreTime ds
+    case kt of
+        Just xt | all (<= xt) ds -> return ()
+        _ -> putStore k =<< act
 \end{minted}
 
-After performing the build, XXX computes \textsf{A2 = 4} and \textsf{A3 = 2}
-from the input cell \textsf{A1}. Now if the user changes the \textsf{A1} to 5,
-XXX will recompute both \textsf{A2 = 4}, which is necessary, but also
-\textsf{A3 = 2}, which is unnecessary since its only dependency (\textsf{A1})
-has not changed since the previous build.
+We'll come back to \hs{withChangedApplicative} afterwards, but ignoring that, reading the rest of the implementation you can see the underying essence of \Make.
+We use the helper function \hs{topological} which gets called with each file in a dependency-respecting order given the file (\hs{k}), the dependencies (\hs{ds}) and an action to generate the result (\hs{act}).
+With those three things in hand, the implementation first tries to get the timestamp of the target, then all the dependencies (which must have been rebuilt if necessary, thanks to the guarantees on \hs{topological}), and if everything is fine does nothing, otherwise rebuilds and stores the result.
 
-One possible implementation of a spreadsheet build
-system\footnote{For example, see \url{http://www.decisionmodels.com/calcsecretsc.htm}.}
-is to maintain a
-sequence of cells in which they should be evaluated. The build system takes this
-sequence as input, marks all cells as \emph{unevaluated}, and then attempts to
-evaluate all cells in the order specified by the sequence. If the cell to be
-evaluated depends on an unevaluated cell, this means the provided sequence is
-not a correct topological order and is updated by moving the current cell to the
-back of the sequence. The build then proceeds with the next cell. Otherwise, if
-all dependencies of the current cell have been marked as \emph{evaluated}, the
-build system computes the correct value for the cell by evaluating the formula,
-writes it into the spreadsheet and proceeds to the next cell in the sequence. If
-there are no cyclic dependencies, this process eventually terminates with
-correct results. The resulting sequence, which is guaranteed to respect the
-topological order of dependencies is stored to be reused during the next build.
-
-...
-
-Interesting note: XXX is unusual in that it tracks changes in the compute,
-i.e. if a user edits a formula, XXX will correctly rebuild results. Most build
-systems, including \Shake, do not meet this requirement and a clean rebuild may
-be necessary if the user edits build rules.
-
-...
-
-\subsubsection{Correct and minimal monadic build system}~\\
-
-\Shake is a correct and minimal monadic build system.
-\todo{NM}{Describe \Shake using our abstractions.}
-
-\subsection{Cloud build systems}
-
-Codebase of large software projects comprises millions of lines of code spread
-across thousands of files, each built independently by thousands of developers
-on thousands of machines. A distributed cloud build system speeds up builds
-dramatically (and saves energy) by transparently sharing build products among
-developers.
-
-\todo{AM}{Also mention YYY (developed by ZZZ), Buck (Facebook) and Pants
-(Twitter et al). Buck is essentially the same. Are they essentially the same?}
-
-\Bazel is a cloud build system developed by Google, which supports caching of
-build results. To achieve that, it maintains two partial maps:
+The implementation of \hs{topological} encodes the dependency strategy that \Make has chosen to use. The function itself is defined as:
 
 \begin{minted}{haskell}
-type Cache =  Hash  -> Maybe v
-type Known = [Hash] -> Maybe Hash
+topological :: Default i => (k -> [k] -> M i k v v -> M i k v ()) -> Build Applicative i k v
+topological step compute k = runM $ do
+    let depends = getDependencies compute
+    forM_ (topSort depends $ transitiveClosure depends k) $ \k ->
+        case compute getStore k of
+            Nothing -> return ()
+            Just act -> step k (depends k) act
 \end{minted}
 
-\hs{Cache} is a conventional \emph{content-addressable store}, that can be used
-to fetch a previously computed value given its hash.
+It uses a \hs{getDependencies} on the computation function, takes the transitive-closure and topological-sort to provide an order in which to build the keys.
+For each key, if it is an input it does nothing, otherwise it calls the supplied \hs{step} function.
+Note that \hs{getDependencies} is only defined for \hs{Applicative} computes, which is what restricts \Make to Applicative build systems.
+Moreover, anyone following the same topological-sort approach will also have the \hs{Applicative} restriction.
 
-\hs{Known} records a known outcome of a computation that took a list of values
-as input (represented by their hashes) and produced a resulting value (also
-represented by its hash). Now, a build system has the following two options to
-recompute a value:
+The ``nasty'' element of this implementation is \hs{getStoreTime}, which is supported with a few somewhat abstract functions on \hs{M}:
+
+\begin{minted}{haskell}
+getChanged :: Eq v => k -> M (Changed k v, i) k v Bool
+withChangedApplicative :: Build Applicative (Changed k v, i) k v -> Build Applicative (Changed k v, i) k v
+
+data Time = LastBuild | AfterLastBuild
+getStoreTimeMaybe :: Eq v => k -> M (Changed k v, i) k v (Maybe Time)
+getStoreTime :: Eq v => k -> M (Changed k v, i) k v Time
+\end{minted}
+
+Recall that our \hs{Build} operation only gets a \hs{Map} - not a set of timestamps at which the build was updated.
+To model a timed store on top the build system implementation needs to store as info (the \hs{i}) field the old file store at the end, and then it can consider anything that has changed to be dirty.
+With this implementation \hs{getChanged k} will return \hs{True} if either the value associated with \hs{k} changed between the last execution and the start of this execution (tested for using the \hs{Eq} class), \textit{or} if \hs{putStore} has been called on it in this run.
+That essentially encodes the dirty bit required for Excel.
+We use \hs{withChangedApplicative} to do the tedious step of capturing the state at the end of the build and storing it as the info.
+
+On top of a dirty bit we can encode a time with a very small domain, concretely if the file was last changed before the last build ended, or afterwards.
+That domain is sufficient for the \Make system.
+At first the fact that was sufficient surprised us, but it eventually became clear that \Make is using modification times as a dirty bit!
+Moreover, that explains why that a \hs{putStore} of an equal value doesn't avoid setting the changed bit - it can't because modification times can't encode that.
+Now we can give an equivalent but simpler formulation:
+
+\begin{minted}{haskell}
+makeDirtyBit :: Eq v => Build Applicative (Changed k v, ()) k v
+makeDirtyBit = withChangedApplicative $ topological $ \k ds act -> do
+    dirty <- fmap isNothing (getStoreMaybe k) ||^ anyM getChanged ds
+    when dirty $
+        putStore k =<< act
+\end{minted}
+
+Here we directly test for changes, which is simpler.
+
+\subsection{Spreadsheet}
+
+The spreadsheet is actually one of the most challenging to implement, and one of the most challenging to into pieces. The code is:
+
+\begin{minted}{haskell}
+spreadsheet :: Eq v => Build Monad (Changed k v, [k]) k v
+spreadsheet = withChangedMonad $ dynamic $ \k ds _ act -> do
+    dirty <- fmap isNothing (getStoreMaybe k) ||^ getChanged k ||^ maybe (return True) (anyM getChanged) ds
+    if not dirty then
+        return Nothing
+    else do
+        res <- act
+        case res of
+            Left e -> return $ Just e
+            Right (_, v) -> do
+                putStore k v
+                return Nothing
+\end{minted}
+
+We use basically the same change computation to test for dirtiness, but things are complicated because when we run a computation it may fail (indicating we need to try again later), or may succeed.
+The best way to explain it is with the type of \hs{dynamic}:
+
+\begin{minted}{haskell}
+dynamic
+    :: (m ~ M (i, [k]) k v, Default i)
+    => (k -> Maybe [k] -> (k -> m (Maybe v)) -> m (Either k ([k], v)) -> m (Maybe k))
+    -> Build Monad (i, [k]) k v
+\end{minted}
+
+Dynamic requires a function that takes takes 4 arguments (we use \hs{m} to elide \hs{M (i, [k]) k v} from all the signatures):
+
+\begin{enumerate}
+\item \hs{k} - the key that is being built.
+\item \hs{Maybe [k]} - the list of dependencies, or \hs{Nothing} if the step makes use of monadic dependencies.
+Note that above if a step is truely monadic it is always rebuilt.
+In the case of spreadsheets such steps are very rare, so a very conservative approximation can be used.\footnote{In truth, many spreadsheets take the alternative approximation of always assuming monadic steps \textit{did not} change - trading correctness for performance.}
+\item \hs{k -> m (Maybe v)} - a function that looks up a key - it returns \hs{Nothing} if that key has not yet been built (and thus there is a dependency violation).
+\item \hs{m (Either k ([k], v))} - an action to produce either the \hs{k} which is required but not yet computed (the reason you fail your dependency order check), or a list of the keys used and value produced.
+\item It returns \hs{m (Maybe k)}, with a \hs{Nothing} to indicate the the build succeeded, or \hs{Just k} to indicate that the given key \hs{k} has not been built, and thus this key should be moved after \hs{k}.
+\end{enumerate}
+
+The implementation of \hs{dynamic} is 15 lines, but is substantially complicated by the requirement to ensure newly discovered keys are added to the list - something an actual spreadsheet with a fixed universe of cells need not do.
+
+\subsection{Shake}
+
+The Shake approach for dependency tracking is different yet again, with a recursive approach. We can implement Shake with:
+
+\begin{minted}{haskell}
+type Shake k v = Map.Map k (Hash v, [(k, Hash v)])
+
+-- | The simplified Shake approach of recording previous traces
+shake :: Hashable v => Build Monad (Shake k v) k v
+shake = recursive $ \k _ ask act -> do
+    info <- getInfo
+    valid <- case Map.lookup k info of
+        Nothing -> return False
+        Just (me, deps) ->
+            (maybe False (== me) <$> getStoreHashMaybe k) &&^
+            allM (\(d,h) -> (== h) . getHash <$> ask d) deps
+    unless valid $ do
+        (ds, v) <- act
+        putStore k v
+        dsh <- mapM getStoreHash ds
+        modifyInfo $ Map.insert k (getHash v, zip ds dsh)
+\end{minted}
+
+The basic implementation of Shake stores a trace - each target key is associated with a \hs{Hash v} (the value that resulted last time it was built), along with a list of \hs{(k, Hash v)} pairs, being the dependencies and the values it used at that point.
+The essence of Shake is that it considers a file dirty if the trace is not consistent with what currently exists on disk, and after rebuilding, captures a trace.
+The \hs{recursive} combinator is defined as:
+
+\begin{minted}{haskell}
+-- | Build a rule at most once in a single execution
+recursive :: Default i => (k -> Maybe [k] -> (k -> M i k v v) -> M i k v ([k], v) -> M i k v ()) -> Build Monad i k v
+recursive step compute = runM . ensure
+    where
+        ensure k = do
+            let ask x = ensure x >> getStore x
+            done <- getTemp
+            when (k `Set.notMember` done) $ do
+                modifyTemp $ Set.insert k set
+                case trackDependencies compute ask k of
+                    Nothing -> return ()
+                    Just act -> step k (getDependenciesMaybe compute k) ask act
+\end{minted}
+
+The required \hs{step} function gets given the \hs{k} to build, the dependencies as \hs{Maybe [k]} (where they are \hs{Nothing} if the rule is truely monadic), a function to demand the result of a dependent key, and a function to compute the current value and return the actual dependencies.
+To ensure each file is only visited once in a single execution a piece of temporary state is used with \hs{getTemp} and \hs{modifyTemp} to record which files were visited - this state is not persisted to the \hs{i} info parameter.
+
+\subsection{Bazel}
+
+Now we have seen the three dependency schemes, we can directly use \hs{topological} to define Bazel.
+Furthermore, in many ways Bazel is a tracing build system, so it has a number of similarities to Shake.
+Concretely we have:
+
+\begin{minted}{haskell}
+data Bazel k v = Bazel
+    {bzKnown :: Map.Map (k, [Hash v]) (Hash v)
+    ,bzContent :: Map.Map (Hash v) v
+    } deriving Show
+
+bazel :: Hashable v => Build Applicative (Bazel k v) k v
+bazel = topological $ \k ds act -> do
+    ds <- mapM getStoreHash ds
+    res <- Map.lookup (k, ds) . bzKnown <$> getInfo
+    case res of
+        Nothing -> do
+            res <- act
+            modifyInfo $ \i -> i
+                {bzKnown = Map.insert (k, ds) (getHash res) $ bzKnown i
+                ,bzContent = Map.insert (getHash res) res $ bzContent i}
+            putStore k res
+        Just res -> do
+            now <- getStoreHashMaybe k
+            when (now /= Just res) $
+                putStore k . (Map.! res) . bzContent =<< getInfo
+\end{minted}
+
+The state matches that described in \S\ref{todo} - we have a map indexed by a target key, and the hashes of the dependencies of that key.
+Given that, we point at a resultant hash.
+Separately we have a map giving the contents for a given hash.
+For simplicity we assume that all hashes have corresponding entries in the content index.
+
+The implementation works by getting the hashes of all dependencies and looking them up in the trace table. There are three possibilities:
+
+\begin{enumerate}
+\item This situation has never been encountered before. Execute the action and record it in the info.
+\item This situation has been encountered before, and the result was the same as we have now, so nothing to do.
+\item This situation has been encountered before but we don't have that result locally (usually we have nothing), so just download the result.
+\end{enumerate}
+
+\subsection{Cloud Shake}
+
+Using the framework above we have implemented the full 3x3 matrix, of dependency scheme and change approach.
+To us, the interesting point on the design space would be monadic dependencies using the \hs{recursive} strategy combined with cloud builds as per \Bazel, so we focus on that.
+The implementation is:
+
+\begin{minted}{haskell}
+data ShazelResult k v = ShazelResult [(k, Hash v)] (Hash v) deriving Show
+
+data Shazel k v = Shazel
+    {szKnown :: Map.Map k [ShazelResult k v]
+    ,szContent :: Map.Map (Hash v) v
+    } deriving Show
+
+instance Default (Shazel k v) where def = Shazel def def
+
+shazel :: Hashable v => Build Monad (Shazel k v) k v
+shazel = recursive $ \k _ ask act -> do
+    poss <- Map.findWithDefault [] k . szKnown <$> getInfo
+    res <- flip filterM poss $ \(ShazelResult ds r) -> allM (\(k,h) -> (==) h . getHash <$> ask k) ds
+    case res of
+        [] -> do
+            (ds, v) <- act
+            dsv <- mapM getStoreHash ds
+            modifyInfo $ \i -> i
+                {szKnown = Map.insertWith (++) k [ShazelResult (zip ds dsv) (getHash v)] $ szKnown i
+                ,szContent = Map.insert (getHash v) v $ szContent i}
+            putStore k v
+        _ -> do
+            let poss = [v | ShazelResult _ v <- res]
+            now <- getStoreHashMaybe k
+            when (now `notElem` map Just poss) $
+                putStore k . (Map.! head poss) . szContent =<< getInfo
+\end{minted}
+
+The difference between the data type from Bazel is worth looking at. In both cases they essentially store traces - a target key, target hash, list of dependency keys and hashes. However:
 
 \begin{itemize}
-    \item Call the compute function on the store containing all up-to-date
-    dependencies and put the computed value to the store.
-    \item Collect hashes of all up-to-date dependencies \hs{hs} and query the
-    \hs{Known} map. If this computation has been performed before, the map will
-    contain the hash \hs{h} of one valid result (recall that the compute function
-    may be non-deterministic). It is now possible to lookup \hs{h} in the
-    \hs{Cache} and if it contains a value \hs{v} it can be downloaded to the
-    local store instead of running the compute.
+\item Bazel doesn't bother to store the dependency keys, as they can be calculated from the target using \hs{getDependencies} due to the \hs{Applicative} nature.
+It indexes on the target key and dependency hashes, immediately providing the target hash.
+\item Cloud Shake has to store the dependency keys, but can't index the Map by them - the checking must mirror the monadic nature. A more refined/efficient data type would be:
+\begin{minted}{haskell}
+data Choice = Choice k (Map (Hash v) Choice)
+            | Result (Hash v)
+\end{minted}
 \end{itemize}
 
-...
-
-\todo{AM}{Also describe the XYZ algorithm.}
+We also technically allow multiple traces to be in the store at the same time, although that is merely something that falls out for free - we would have to do work to avoid it.

### 2018-03-08 paper/intro.tex
Minor revision

@@
 
-Build systems (such as make) are big, complicated, but unloved part of
+Build systems (such as \Make) are big, complicated, but unloved part of
 the software ecosystem.  Every developer on the planet uses one, but
@@
 
-Seldom do people ask questions like "What does it mean for my build
-system to be correct?" or "What are the trade-offs between different
-approaches?".  Complex build systems use subtle algorithms, but they
+Seldom do people ask questions like ``What does it mean for my build
+system to be correct?'' or ``What are the trade-offs between different
+approaches?''.  Complex build systems use subtle algorithms, but they
 are often hidden away, and not the object of study.
@@
 early cut-off; self-tracking build systems; and persistent metadata.
-These properties (which we define in Section~\ref{sec-background}) are often
+These properties (which we define in~\S\ref{sec-background}) are often
 deeply-built-in assumptions of a particular build system.
@@
 executable models of all the build systems we describe.  An unusual
-feature is that we include Excel in our line-up because, looked at
+feature is that we include \Excel in our line-up because, looked at
 in the right way, it certainly is a build system.

### 2018-03-08 paper/background.tex
Work on Background

@@
 Build systems automate the execution of simple repeatable tasks for individual
-users, as well as for large organisations. There are software build systems such
-as \Make, \Ninja, \Shake, \Bazel, \Buck and many others, as well as various
-incremental calculation engines, such as \Excel and \Calc. Existing build
-systems vary in several different ways:
+users, as well as for large organisations. There are software build systems,
+such as \Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake} and
+\Bazel~\cite{bazel}, as well as various incremental calculation engines, such
+as \Excel~\cite{advanced_excel}. In this section we use these four examples to
+introduce main domain-specific notions and requirements. Other notable examples
+of build systems and their relation to these four will be discussed
+in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.
+
+\Make\footnote{There are numerous implementations of \Make and none comes with a
+formal specification. In this paper we therefore use a simple and sensible
+approximation to a real \Make that you might find on your machine.} was developed
+more than 40 years ago to automatically build software libraries and executable
+programs from source code. It uses \emph{makefiles} to describe build tasks,
+typically referred to as \emph{build rules}, and their dependencies in a simple
+text form. For example:
+
+\begin{minted}[frame=single]{makefile}
+util.o: util.h util.c
+    gcc -c util.c
+
+main.o: util.h main.c
+    gcc -c main.c
+
+main.exe: util.o main.o
+    gcc util.o main.o -o main.exe
+\end{minted}
+
+\noindent
+The above makefile lists three tasks: (i) compile a utility library comprising
+files \cmd{util.h} and \cmd{util.c} into \cmd{main.o} by executing the command
+\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
+\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
+executable \cmd{main.exe}. The makefile contains the complete information about
+the task \emph{dependency graph}, which is shown in Fig.~\ref{fig-make}(a).
+
+\begin{figure}[h]
+\begin{subfigure}[b]{0.32\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/make-example.pdf}}
+\caption{A dependency graph}
+\end{subfigure}
+\begin{subfigure}[b]{0.32\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/make-example-full.pdf}}
+\caption{A full rebuild}
+\end{subfigure}
+\begin{subfigure}[b]{0.32\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/make-example-partial.pdf}}
+\caption{A partial rebuild}
+\end{subfigure}
+\caption{A dependency graph and two build scenarios. Input files are shown as
+rectangles, intermediate and output files are shown as rounded rectangles. Dirty
+inputs and files that are rebuilt are highlighted.
+\label{fig-make}}
+\end{figure}
+
+If the user modifies the sources of the utility library and runs \Make, it will
+perform a full rebuild, because all three tasks transitively depend on the
+library, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
+user modifies \cmd{main.c} then a partial rebuild is sufficient: indeed, the
+file \cmd{util.o} does not need to be rebuilt, since its inputs have not
+changed, see Fig.~\ref{fig-make}(c). Files that have changed since the previous
+build are called \emph{dirty}.
+
+The requirement to execute tasks \emph{at most once} and only if they
+\emph{transitively depend on dirty inputs} is essential for build systems,
+their raison d'\^etre. We will call build systems that satisfy this requirement
+\emph{minimal}.
+
+To achieve minimality \Make relies on two key ideas. First, it uses \emph{file
+modification time} to detect the files that are dirty: a file is marked dirty if
+it was modified after the previous build. Second, it constructs a complete task
+dependency graph from the information contained in the makefile and executes
+tasks in the \emph{topological order}.
+
+...
+
+\Excel
+
+...
+
+\Shake
+
+\begin{figure}[h]
+\centerline{\includegraphics[scale=0.28]{fig/make-example-cutoff.pdf}}
+\caption{An early cut-off example.\label{fig-cutoff}}
+\end{figure}
+
+...
+
+\Bazel
+
+...
 

### 2018-03-08 paper/background.tex
Minor revision

@@
 in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.
+
+\subsection{The venerable \Make}
 
@@
 The above makefile lists three tasks: (i) compile a utility library comprising
-files \cmd{util.h} and \cmd{util.c} into \cmd{main.o} by executing the command
-\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
-\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
-executable \cmd{main.exe}. The makefile contains the complete information about
-the task \emph{dependency graph}, which is shown in Fig.~\ref{fig-make}(a).
+files \cmd{util.h} and \cmd{util.c} into \cmd{main.o} by
+executing\footnote{In this example we treat \cmd{gcc} as a pure function for the
+sake of simplicity. In reality there are multiple versions of \cmd{gcc} and the
+actual binary that is used to compile and link files is also listed as a task
+dependency.} the command \cmd{gcc -c util.c}, (ii) compile the main source file
+\cmd{main.c} into \cmd{main.o}, and (iii) link object files \cmd{util.o} and
+\cmd{main.o} into the executable \cmd{main.exe}. The makefile contains the
+complete information about the task \emph{dependency graph}, which is shown in
+Fig.~\ref{fig-make}(a).
 
@@
 
-\Excel
+\subsection{\Excel: a build system in disguise}
 

### 2018-03-08 paper/engineering.tex
Add iterative computations
Minor revision

@@
 
+Or ``Build systems in the real world'', or ``Crazy things section''...
+
 \begin{itemize}
     \item Sandboxing
-    \item Non-determinism, specifically, it's useless form. This doesn't
+    \item Non-determinism, specifically, its useless form. This doesn't
           include \Excel's \textsf{RANDBETWEEN} function, which is useful.
@@
           non-deterministic results (with a notable exception of \Buck).
-    \item Concurrency
-    \item Cloud
+    \item Concurrency: nothing fundamentally difficult and fits our theory, but
+          hard to get right in practice.
+    \item Cloud aspects: eviction/garbage collection, `frankenbuilds', etc.
+    \item Iterative computations, e.g. LaTeX.
+    \item Polymorphic keys and values: tasks with multiple outputs, e.g. GHC's
+          \cmd{hi} and \cmd{o} files, phony tasks, etc.
 \end{itemize}

### 2018-03-08 paper/background.tex
Add dynamic dependencies

@@
 \cmd{main.o} into the executable \cmd{main.exe}. The makefile contains the
-complete information about the task \emph{dependency graph}, which is shown in
+complete information about the \emph{task dependency graph}, which is shown in
 Fig.~\ref{fig-make}(a).
@@
 \centerline{\includegraphics[scale=0.28]{fig/make-example.pdf}}
-\caption{A dependency graph}
+\caption{Task dependency graph}
 \end{subfigure}
@@
 \centerline{\includegraphics[scale=0.28]{fig/make-example-full.pdf}}
-\caption{A full rebuild}
+\caption{Full rebuild}
 \end{subfigure}
@@
 \centerline{\includegraphics[scale=0.28]{fig/make-example-partial.pdf}}
-\caption{A partial rebuild}
+\caption{Partial rebuild}
 \end{subfigure}
-\caption{A dependency graph and two build scenarios. Input files are shown as
+\caption{A task dependency graph and two build scenarios. Input files are shown as
 rectangles, intermediate and output files are shown as rounded rectangles. Dirty
@@
 
-The requirement to execute tasks \emph{at most once} and only if they
-\emph{transitively depend on dirty inputs} is essential for build systems,
-their raison d'\^etre. We will call build systems that satisfy this requirement
-\emph{minimal}.
+The requirement to \emph{execute tasks at most once and only if they transitively
+depend on dirty inputs} is essential for build systems, it is their raison
+d'\^etre. We will call build systems that satisfy this requirement \emph{minimal}.
 
-To achieve minimality \Make relies on two key ideas. First, it uses \emph{file
+To achieve minimality \Make relies on two main ideas. First, it uses \emph{file
 modification time} to detect the files that are dirty: a file is marked dirty if
@@
 dependency graph from the information contained in the makefile and executes
-tasks in the \emph{topological order}.
+tasks in a \emph{topological order}.
 
-...
+Note that it is not always possible to know the dependencies of a source file
+upfront, or \emph{statically}. To demonstrate this, let us first add another
+task to the above makefile:
+
+\begin{minted}[frame=single]{makefile}
+release.tar: main.exe README
+    tar -cf release.tar main.exe README
+\end{minted}
+
+So far so good: if either \cmd{main.exe} or \cmd{README} is dirty, \Make will
+recreate the release archive. But now imagine that we need to specify the files
+that go into \cmd{release.tar} in a text file \cmd{release.txt}. The dependencies
+of \cmd{release.tar} are no longer known statically: they \emph{depend on the
+contents} of \cmd{release.txt}, which might not even exist before the build
+starts, e.g. if it is produced by concatenating input files \cmd{release-bins.txt}
+and \cmd{release-docs.txt}. Alas, makefiles cannot express such \emph{dynamic
+dependencies}, requiring ad hoc workarounds such as \emph{build phases}, which
+are problematic~\cite{hadrian}.
 

### 2018-03-08 paper/background.tex
Finish Excel subsection

@@
 
-\subsection{The venerable \Make}
+\subsection{The venerable \Make: static dependencies and file modification times}
 
@@
 more than 40 years ago to automatically build software libraries and executable
-programs from source code. It uses \emph{makefiles} to describe build tasks,
-typically referred to as \emph{build rules}, and their dependencies in a simple
-text form. For example:
+programs from source code. It uses \emph{makefiles} to describe build tasks and
+their dependencies in a simple text form. For example:
 
@@
 modification time} to detect the files that are dirty: a file is marked dirty if
-it was modified after the previous build. Second, it constructs a complete task
+it was modified since the previous build. Second, it constructs a complete task
 dependency graph from the information contained in the makefile and executes
@@
 
-Note that it is not always possible to know the dependencies of a source file
-upfront, or \emph{statically}. To demonstrate this, let us first add another
-task to the above makefile:
+Note that it is not always possible to know the dependencies of a task upfront,
+or \emph{statically}. To demonstrate this, let us add another task to the above
+makefile:
 
@@
 
-\subsection{\Excel: a build system in disguise}
+\subsection{\Excel: dynamic dependencies at the cost of minimality}
 
-...
+\Excel is a build system that can deal with dynamic dependencies albeit at the
+cost of sacrificing minimality. Consider the following spreadsheet comprising
+four input cells \cmd{A1-A3} and \cmd{B1}, and a single task computing the sum
+of values \cmd{A1} $+\cdots+$ \cmd{An}, where \cmd{n} is specified indirectly in
+\cmd{B1}:
 
-\Shake
+\begin{minted}[frame=single]{bash}
+A1 = 10
+A2 = 20
+A3 = 30
+B1 = 2
+C1 = SUM(A1:INDIRECT("A" & B1))
+\end{minted}
+
+The output \cmd{C1} here is analogous to the \cmd{release.tar} from the previous
+example: one cannot statically determine the dependencies of the associated task.
+To deal with this, \Excel takes a simple conservative approach: it marks all
+cells with dynamic dependencies as dirty and therefore always recomputes them
+during a rebuild. This guarantees that the build result is correct, but clearly
+violates the minimality requirement: if the user modifies \cmd{A3}, \Excel will
+recompute \cmd{C1} even though it does not transitively depend on \cmd{A3}.
+
+Note that \Excel's approach to marking cells as dirty is a relatively minor
+variation of the approach used by \Make: a cell is considered dirty if it was
+modified since the previous build \emph{or if it is volatile}, where the notion
+of volatility includes dynamic dependencies, as well as \emph{non-determinism}
+that will be discussed in~\S\ref{sec-engineering}. We therefore put both \Make
+and \Excel in the same category of build systems that associate a \emph{dirty bit}
+with every file or cell. (To avoid the ``file or cell'' nuisance, we introduce the
+abstract notion of \emph{keys} in~\S\ref{sec-vocabulary}.)
+
+\Excel's recalculation algorithm~\cite{excel_recalc} on the other hand is
+significantly different from \Make. Since it is impossible to construct the full
+dependency graph upfront in the presence of dynamic dependencies, \Excel
+maintains the \emph{calculation chain}, which is an approximation to the correct
+topological order. During recalculation, \Excel processes cells in this order,
+but can \emph{defer recalculation of a cell} by moving it down the chain when newly
+discovered dynamic dependencies dictate that. We discuss this algorithm in more
+detail in~\S\ref{sec-examples}.
+
+Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
+systems track changes of inputs and intermediate results, executing dependent
+tasks whenever they change, but \Excel can also track changes in the tasks
+themselves: if a formula was modified, \Excel will recompute it and propagate
+the changes. Self-tracking is uncommon in software build systems, where one
+often needs to manually initiate a full rebuild even if just a single task has
+changed.
+
+\subsection{\Shake: dynamic dependencies with no remorse}
 
@@
 \centerline{\includegraphics[scale=0.28]{fig/make-example-cutoff.pdf}}
-\caption{An early cut-off example.\label{fig-cutoff}}
+\caption{An early cutoff example.\label{fig-cutoff}}
 \end{figure}
@@
 
-\Bazel
+\subsection{\Bazel: a cloud build system}
+
+...
+
+\subsection{Abstract vocabulary for build systems}\label{sec-vocabulary}
 
@@
 \begin{itemize}
-    \item Many build systems, such as the venerable \Make, need to know the
-    complete \emph{dependency graph} between tasks at the start of the build
-    process. This makes it possible to analyse the graph ahead of time, which
-    helps with efficient task scheduling but is fundamentally limited: parts of
-    the dependency graph can often be discovered only during the build process,
-    i.e. the dependency graph is \emph{dynamic}, not \emph{static}.
 
@@
     cut-off: \Make and \Excel do not, whereas \Shake and \Buck do.
-
-    \item Most build systems track changes of inputs and intermediate results,
-    executing dependent tasks whenever they change, but some build systems can
-    also track changes in the tasks themselves: if a task has changed, the build
-    system will execute it. For example, when a cell's formula has changed,
-    \Excel will recompute its value and propagate the changes. We call this
-    \emph{self-tracking}. Self-tracking is uncommon in software build systems,
-    where one often needs to manually initiate a rebuild of all tasks even if
-    just a single task has changed.
 
@@
 
-\emph{Keys} are used to distinguish \emph{values}. In build systems keys are
-typically filenames, e.g. \textsf{src/file.c}, whereas values are file contents
-(a C program source code in this case). In spreadsheets keys are cell names,
-e.g. \textsf{A1}, and values are numbers, text, etc. that are typically displayed
-inside cells. We will use type variables \hs{k} and \hs{v} to denote keys and
-values, respectively.
+Keys are
+used to distinguish values. In build systems keys are typically filenames, e.g.
+\cmd{main.c}, whereas values are file contents (a C program source code in this
+case). In spreadsheets keys are cell names, e.g. \cmd{A1}, and values are
+numbers, text, etc. that are typically displayed inside cells. In Haskell code,
+we will use type variables \hs{k} and \hs{v} to denote keys and values,
+respectively.
 

### 2018-03-08 paper/background.tex
Work on S2.3

@@
 \subsection{The venerable \Make: static dependencies and file modification times}
+\label{sec-background-make}
 
@@
 
-\begin{minted}[frame=single]{makefile}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{makefile}
 util.o: util.h util.c
@@
 \end{minted}
+\vspace{1mm}
 
@@
 
-\begin{minted}[frame=single]{makefile}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{makefile}
 release.tar: main.exe README
@@
 \end{minted}
+\vspace{1mm}
 
+\noindent
 So far so good: if either \cmd{main.exe} or \cmd{README} is dirty, \Make will
@@
 \subsection{\Excel: dynamic dependencies at the cost of minimality}
+\label{sec-background-excel}
 
@@
 
-\begin{minted}[frame=single]{bash}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
 A1 = 10
@@
 \end{minted}
+\vspace{1mm}
 
+\noindent
 The output \cmd{C1} here is analogous to the \cmd{release.tar} from the previous
@@
 with every file or cell. (To avoid the ``file or cell'' nuisance, we introduce the
-abstract notion of \emph{keys} in~\S\ref{sec-vocabulary}.)
+abstract notion of \emph{keys} in~\S\ref{sec-background-vocabulary}.)
 
@@
 \subsection{\Shake: dynamic dependencies with no remorse}
+\label{sec-background-shake}
+
+\Shake was developed specifically to address the issue of dynamic
+dependencies~\cite{mitchell2012shake} and it excels at handling them without
+sacrificing the minimality requirement. Here is how one can express the task
+for producing \cmd{release.tar} from the example discussed
+in~\S\ref{sec-background-make}.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+"release.tar" %> \_ -> do
+    need ["release.txt"]
+    contents <- lines <$> readFile "release.txt"
+    need contents
+    system "tar" $ ["-cf", "result.tar"] ++ contents
+\end{minted}
+\vspace{1mm}
+
+\noindent
+We first declare the static dependency on \cmd{release.txt}, then read its
+content and depend on it, dynamically. Finally, we specify the command line to
+produce the resulting archive. Crucially, the archive will only be rebuilt if
+one of the dependencies (static or dynamic) has changed.
 
@@
 \centerline{\includegraphics[scale=0.28]{fig/make-example-cutoff.pdf}}
-\caption{An early cutoff example.\label{fig-cutoff}}
+\vspace{-2mm}
+\caption{An early cutoff example: modify the source \cmd{main.c} by adding a
+comment. The rebuild is stopped after detecting that \cmd{main.o} is unchanged,
+which indicates that \cmd{main.exe} does not need to be rebuilt.\label{fig-cutoff}}
 \end{figure}
@@
 \subsection{\Bazel: a cloud build system}
+\label{sec-background-shake}
 
@@
 
-\subsection{Abstract vocabulary for build systems}\label{sec-vocabulary}
+\subsection{Abstract vocabulary for build systems}
+\label{sec-background-vocabulary}
 

### 2018-03-08 paper/background.tex
Further work on S2

@@
 tasks whenever they change, but \Excel can also track changes in the tasks
-themselves: if a formula was modified, \Excel will recompute it and propagate
+themselves: if a formula is modified, \Excel will recompute it and propagate
 the changes. Self-tracking is uncommon in software build systems, where one
@@
 We first declare the static dependency on \cmd{release.txt}, then read its
-content and depend on it, dynamically. Finally, we specify the command line to
+content and depend on it, dynamically. Finally, we specify the command to
 produce the resulting archive. Crucially, the archive will only be rebuilt if
 one of the dependencies (static or dynamic) has changed.
+
+\todo{AM}{How is \Shake different.}
 
@@
 
+\Shake also supports the following \emph{early cutoff optimisation}. When it
+executes a task and the result is unchanged from the previous build, it is
+unnecessary to execute the dependent tasks, and hence \Shake can stop a build
+earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
+early cutoff: \Make and \Excel do not, whereas \Shake and \Bazel (introduced
+below) do.
+
 ...
@@
 
-...
-
-\subsection{Abstract vocabulary for build systems}
-\label{sec-background-vocabulary}
-
-...
-
-\begin{itemize}
-
-    \item When build systems are used by large teams, different team members
+When build systems are used by large teams, different team members
 often end up executing exactly the same tasks on their local machines.
@@
 
-    \item Some build systems, e.g. \Buck, require all tasks to be
-    \emph{deterministic}, i.e. produce exactly the same output when run on the
-    same inputs. However, not all tasks are deterministic, and there are build
-    systems that support \emph{non-determinism}. A simple example is \Excel's
-    \textsf{RANDBETWEEN(low, high)} that returns a random integer in the
-    interval \textsf{[low, high]}.
-
-    \item If a build system executes a task and the result is unchanged from the
-    previous build, it is unnecessary to execute the dependent tasks. We call
-    this optimisation \emph{early cut-off}. Not all build systems support early
-    cut-off: \Make and \Excel do not, whereas \Shake and \Buck do.
-
-    \item Most build systems persistently store auxiliary \emph{build
-    information} for profiling and optimisation purposes: \Make stores file
-    modification times, \Shake stores the discovered dynamic dependency graph,
-    \Bazel and other cloud build systems store information about inputs and
-    outputs of previously executed tasks, etc.
-\end{itemize}
-
-This paper presents a purely functional abstraction for build systems that
-allows us to express all the above intricacies of build systems and design
-complex build systems from simple primitives. The presented abstraction fits in
-just two lines of Haskell code, which are explained
-in~\S\ref{sec-abstractions}:
-
-\begin{minted}{haskell}
-type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
-\end{minted}
-
-Below we define basic notions used in build systems and other similar domains,
-for example, spreadsheets.
+\subsection{Abstract vocabulary for build systems}
+\label{sec-background-vocabulary}
 
@@
 
-\subsection{Dynamic dependencies}
+\begin{itemize}
 
-...
+    \item Some build systems, e.g. \Buck, require all tasks to be
+    \emph{deterministic}, i.e. produce exactly the same output when run on the
+    same inputs. However, not all tasks are deterministic, and there are build
+    systems that support \emph{non-determinism}. A simple example is \Excel's
+    \textsf{RANDBETWEEN(low, high)} that returns a random integer in the
+    interval \textsf{[low, high]}.
 
-\todo{AM}{Add two examples: include files and cyclic spreadsheet computations.}
+    \item Most build systems persistently store auxiliary \emph{build
+    information} for profiling and optimisation purposes: \Make stores file
+    modification times, \Shake stores the discovered dynamic dependency graph,
+    \Bazel and other cloud build systems store information about inputs and
+    outputs of previously executed tasks, etc.
+\end{itemize}
+
+This paper presents a purely functional abstraction for build systems that
+allows us to express all the above intricacies of build systems and design
+complex build systems from simple primitives. The presented abstraction fits in
+just two lines of Haskell code, which are explained
+in~\S\ref{sec-abstractions}:
+
+\begin{minted}{haskell}
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
+\end{minted}
+
+Below we define basic notions used in build systems and other similar domains,
+for example, spreadsheets.
 

### 2018-03-09 paper/background.tex
Finish Shake, Bazel, add overview draft

@@
 of \cmd{release.tar} are no longer known statically: they \emph{depend on the
-contents} of \cmd{release.txt}, which might not even exist before the build
+content} of \cmd{release.txt}, which might not even exist before the build
 starts, e.g. if it is produced by concatenating input files \cmd{release-bins.txt}
@@
     need ["release.txt"]
-    contents <- lines <$> readFile "release.txt"
-    need contents
-    system "tar" $ ["-cf", "result.tar"] ++ contents
+    files <- lines <$> readFile "release.txt"
+    need files
+    system "tar" $ ["-cf", "result.tar"] ++ files
 \end{minted}
@@
 We first declare the static dependency on \cmd{release.txt}, then read its
-content and depend on it, dynamically. Finally, we specify the command to
-produce the resulting archive. Crucially, the archive will only be rebuilt if
-one of the dependencies (static or dynamic) has changed.
+content (a list of files) and depend on it, dynamically. Finally, we specify the
+command to produce the resulting archive. Crucially, the archive will only be
+rebuilt if one of the dependencies (static or dynamic) has changed.
 
-\todo{AM}{How is \Shake different.}
+\Shake's implementation is different from both \Make and \Excel in two aspects.
+First, it uses the dependency graph produced in the previous build to detect
+which files are dirty -- a powerful idea with a long history of use in
+\emph{incremental}~\cite{demers1981incremental},
+\emph{adaptive}~\cite{acar2002adaptive}, and
+\emph{self-adjusting}~\cite{acar2007selfadjusting} computation (see a discussion
+in related work~\S\ref{sec-related}). Second, instead of abandoning and deferring
+the computation of tasks whose rediscovered dependencies are dirty (as \Excel
+does), \Shake \emph{pauses} their computation until the dependencies are
+brought up-to-date.
 
@@
 
-...
-
 \subsection{\Bazel: a cloud build system}
@@
 that require all transitive dependencies of an end build product to be
-locally available. Non-cloud build systems cannot support shallow builds.
+locally available. % Non-cloud build systems cannot support shallow builds.
+
+\Bazel is one of the first examples of openly available cloud build systems.
+Like \Make, it does not support dynamic dependencies and can therefore benefit
+from the simplicity of building tasks in the statically known topological order.
+It is minimal and supports the early cutoff optimisation.
+
+To support cloud builds, \Bazel maintains a conventional \emph{content-addressable
+cache}, which can be used to fetch a previously built file given the hash of its
+content, as well as \emph{dependency graphs from all previous builds}, annotated
+with observed file hashes\footnote{Here we ignore the issue of limited cloud
+storage resources for the sake of simplicity; in practice, old entries are
+regularly evicted from the storage, as further discussed in~\S\ref{sec-engineering}.}.
+The latter allows to bypass the execution of a task, by predicting the hash of
+its result from the hashes of the dependencies, and subsequently fetch the file
+from the cache. A concrete implementation is provided in~\S\ref{sec-examples}.
+
+\subsection{Overview}
+\label{sec-background-overview}
+
+\begin{table}[t]
+\small
+\centering
+\begin{tabular}{l||l|l|l|l||l|l}
+Build system & Dependencies & Minimal & Early cutoff & Cloud & Metadata & Algorithm \\
+\hline
+\Make      & Static  & Yes & No  & No  & Dirty bit        & Topological   \\
+\Excel     & Dynamic & No  & No  & No  & Dirty bit, chain & Chain + defer \\
+\Shake     & Dynamic & Yes & Yes & No  & Graph            & Recursive     \\
+\Bazel     & Static  & Yes & Yes & Yes & Graphs, cache    & Topological   \\
+\hline
+This paper & Dynamic & Yes & Yes & Yes & Graphs, cache    & Recursive     \\
+\end{tabular}
+\vspace{2mm}
+\caption{Overview of build systems.\label{tab-overview}}
+\label{tab:use-cases}
+\end{table}
 

### 2018-03-09 paper/related.tex
Finish Shake, Bazel, add overview draft

@@
     \item Lens-like formulation
+    \item Other build systems: \Ninja, \cmd{nix},
+          cloud (Blaze -- first? \Buck, Pants, CloudBuild, XXX),
+          Shake derivatives (Jane Street's \cmd{Jenga}, Apple's \cmd{swift-llbuild}).
 \end{itemize}
+
+John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
+\url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.

### 2018-03-09 paper/background.tex
Finish S2.5

@@
 The above makefile lists three tasks: (i) compile a utility library comprising
-files \cmd{util.h} and \cmd{util.c} into \cmd{main.o} by
+files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
 executing\footnote{In this example we treat \cmd{gcc} as a pure function for the
-sake of simplicity. In reality there are multiple versions of \cmd{gcc} and the
-actual binary that is used to compile and link files is also listed as a task
+sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and the
+actual binary that is used to compile and link files is often also listed as a task
 dependency.} the command \cmd{gcc -c util.c}, (ii) compile the main source file
@@
 maintains the \emph{calculation chain}, which is an approximation to the correct
-topological order. During recalculation, \Excel processes cells in this order,
-but can \emph{defer recalculation of a cell} by moving it down the chain when newly
-discovered dynamic dependencies dictate that. We discuss this algorithm in more
-detail in~\S\ref{sec-examples}.
+topological order produced by the previous build. During recalculation, \Excel
+processes cells in this order, but can \emph{defer recalculation of a cell} by
+moving it down the chain when one of its newly discovered dynamic dependencies
+is dirty. We discuss this algorithm in more detail in~\S\ref{sec-examples}.
 
@@
 \emph{adaptive}~\cite{acar2002adaptive}, and
-\emph{self-adjusting}~\cite{acar2007selfadjusting} computation (see a discussion
-in related work~\S\ref{sec-related}). Second, instead of abandoning and deferring
-the computation of tasks whose rediscovered dependencies are dirty (as \Excel
-does), \Shake \emph{pauses} their computation until the dependencies are
-brought up-to-date.
+\emph{self-adjusting}~\cite{acar2007selfadjusting} \emph{computations} (see a
+discussion in related work~\S\ref{sec-related}). Second, instead of abandoning
+and deferring the computation of tasks whose rediscovered dependencies are dirty
+(as \Excel does), \Shake \emph{pauses} their computation until the dependencies
+are brought up-to-date. We refer to this build algorithm as \emph{recursive}.
 
@@
 The latter allows to bypass the execution of a task, by predicting the hash of
-its result from the hashes of the dependencies, and subsequently fetch the file
+its result from the hashes of the dependencies, and subsequently fetch the result
 from the cache. A concrete implementation is provided in~\S\ref{sec-examples}.
 
-\subsection{Overview}
-\label{sec-background-overview}
+\subsection{Summary}
+\label{sec-background-summary}
 
-\begin{table}[t]
-\small
+We summarise differences between four discussed build systems in
+Table~\ref{tab-summary}. The `Metadata' column refers to the \emph{auxiliary
+build information} persistently stored between builds by all build systems:
+\begin{itemize}
+    \item \Make stores file modification times, or rather, it relies on the file
+    system to do that. In principle a single dirty bit is sufficient, as will be
+    demonstrated in~\S\ref{sec-examples}.
+    \item \Excel stores the dirty bit as well as the calculation chain.
+    \item \Shake stores the dependency graph discovered in the previous build,
+    annotated with file content hashes for efficient checking of file changes.
+    \item \Bazel stores all dependency graphs discovered in previous builds
+    annotated with file hashes, and the content-addressable cache.
+\end{itemize}
+
+The last row of the table describes the new build system proposed in this paper,
+which shares the advantages of \Shake and \Bazel.
+
+\begin{table}[h]
+\smaller
 \centering
-\begin{tabular}{l||l|l|l|l||l|l}
-Build system & Dependencies & Minimal & Early cutoff & Cloud & Metadata & Algorithm \\
+\begin{tabular}{l||l|c|c|c|c||l|l}
 \hline
-\Make      & Static  & Yes & No  & No  & Dirty bit        & Topological   \\
-\Excel     & Dynamic & No  & No  & No  & Dirty bit, chain & Chain + defer \\
-\Shake     & Dynamic & Yes & Yes & No  & Graph            & Recursive     \\
-\Bazel     & Static  & Yes & Yes & Yes & Graphs, cache    & Topological   \\
+$\!\!$Build system$\!$&$\!$Dependencies$\!$&$\!$Minimal$\!$&$\!$Self-tracking$\!$&$\!$Cutoff$\!$&$\!$Cloud$\!$&$\!$Metadata$\!$&$\!$Algorithm$\!\!$\\\hline
+$\!\!$\Make       $\!$&$\!$Static  & Yes & No  & No  & No  &$\!$Dirty bit       $\!$&$\!$Topological $\!\!$\\
+$\!\!$\Excel      $\!$&$\!$Dynamic & No  & Yes & No  & No  &$\!$Dirty bit, chain$\!$&$\!$Chain, defer$\!\!$\\
+$\!\!$\Shake      $\!$&$\!$Dynamic & Yes & No  & Yes & No  &$\!$Graph           $\!$&$\!$Recursive   $\!\!$\\
+$\!\!$\Bazel      $\!$&$\!$Static  & Yes & No  & Yes & Yes &$\!$Graphs, cache   $\!$&$\!$Topological $\!\!$\\\hline
+$\!\!$This paper  $\!$&$\!$Dynamic & Yes & No? & Yes & Yes &$\!$Graphs, cache   $\!$&$\!$Recursive   $\!\!$\\
 \hline
-This paper & Dynamic & Yes & Yes & Yes & Graphs, cache    & Recursive     \\
 \end{tabular}
 \vspace{2mm}
-\caption{Overview of build systems.\label{tab-overview}}
+\caption{Summary of build system differences.\label{tab-summary}}
 \label{tab:use-cases}
@@
 
-\subsection{Abstract vocabulary for build systems}
+\subsection{Common vocabulary for build systems}
 \label{sec-background-vocabulary}
@@
 
-    \item Most build systems persistently store auxiliary \emph{build
-    information} for profiling and optimisation purposes: \Make stores file
-    modification times, \Shake stores the discovered dynamic dependency graph,
-    \Bazel and other cloud build systems store information about inputs and
-    outputs of previously executed tasks, etc.
 \end{itemize}

### 2018-03-09 paper/abstractions.tex
Finish S2

@@
 \clearpage
+
+This paper presents a purely functional abstraction for build systems that
+allows us to express all the above intricacies of build systems and design
+complex build systems from simple primitives. The presented abstraction fits in
+just two lines of Haskell code, which are explained
+in~\S\ref{sec-abstractions}:
+
+\begin{minted}{haskell}
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
+\end{minted}
 

### 2018-03-09 paper/background.tex
Finish S2

@@
 \caption{Summary of build system differences.\label{tab-summary}}
-\label{tab:use-cases}
+\vspace{-7mm}
 \end{table}
@@
 
-\subsection{Keys, values, hashes and store}
+In this section we introduce a common vocabulary for build systems that allows
+us to abstract away from specific application domains, such as software
+development or spreadsheets. We will use this vocabulary throughout the rest of
+the paper.
 
-Keys are
-used to distinguish values. In build systems keys are typically filenames, e.g.
-\cmd{main.c}, whereas values are file contents (a C program source code in this
-case). In spreadsheets keys are cell names, e.g. \cmd{A1}, and values are
-numbers, text, etc. that are typically displayed inside cells. In Haskell code,
-we will use type variables \hs{k} and \hs{v} to denote keys and values,
-respectively.
+\emph{Keys} are used to distinguish \emph{values}. In software build systems
+keys are typically filenames, e.g. \cmd{main.c}, whereas values are file
+contents (a C program source code in this case). In spreadsheets keys are cell
+names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
+displayed inside cells. In Haskell code, we will use type variables \hs{k}
+and \hs{v} to denote keys and values, respectively.
+
+We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
+tracking and sharing of build results, where \hs{Hash} is an abstract data type
+equipped with the equality test.
 
@@
 assume that the type of values is capable of encoding values corresponding to
-non-existent keys (missing files or empty cells).
-
-We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for
-efficient tracking and sharing of build results.
-
-\subsection{Input, intermediate and output values}
+non-existent keys (missing files or empty cells). This, in particular, suggests
+that it is possible to \emph{depend on the absence of a value} which is a useful
+feature for build systems. In addition to usual \hs{get} and \hs{put} operations,
+some stores support the \hs{getHash} operation implemented more efficiently than
+by hashing the result of \hs{get}. For example, GVFS (Git Virtual File
+System)~\cite{gvfs} stores file hashes and downloads file contents only on demand.
 
 Some values must be provided by the user as \emph{input}. For example,
-\textsf{src/file.c} can be edited by the user who relies on the build system to
-compile it into \textsf{obj/file.o}. Similarly, the user can input \textsf{A1 = 5}
-and \textsf{B1 = 9} expecting the spreadsheet to compute their sum in \textsf{C1},
-i.e. \textsf{C1 = 14}.
-
-In the above examples, \textsf{obj/file.o} and \textsf{C1} are \emph{output} values.
-
-In some situations we might also need the notion of \emph{intermediate} values,
-which are not interesting for the user but are produced in the process of turning
-inputs into outputs. For example, the user might only be interested in the
-executable \textsf{bin/file.exe} obtained by linking \textsf{obj/file.o} with
-standard libraries, in which case \textsf{obj/file.o} can be considered an
-intermediate value.
-
-\subsection{Non-deterministic computations}
-
-Build systems and spreadsheets compute output values from input and intermediate
-values. In the most typical case, these \emph{computations} are \emph{functions},
-such as \textsf{C1 = A1 + B1}, i.e. their result is uniquely determined by the
-input values. However, in general they can be \emph{relations}, i.e. have
-multiple valid results. A spreadsheet example: \textsf{A2 = A1 + RANDOM(1,6)}.
-This computation has six valid results for each input value \textsf{A1}. In
-build systems, the object file \textsf{obj/file.o} is sometimes not uniquely
-determined by the source \textsf{src/file.c} -- different compiler runs may
-produce different valid results.
-
-\begin{itemize}
-
-    \item Some build systems, e.g. \Buck, require all tasks to be
-    \emph{deterministic}, i.e. produce exactly the same output when run on the
-    same inputs. However, not all tasks are deterministic, and there are build
-    systems that support \emph{non-determinism}. A simple example is \Excel's
-    \textsf{RANDBETWEEN(low, high)} that returns a random integer in the
-    interval \textsf{[low, high]}.
-
-\end{itemize}
-
-This paper presents a purely functional abstraction for build systems that
-allows us to express all the above intricacies of build systems and design
-complex build systems from simple primitives. The presented abstraction fits in
-just two lines of Haskell code, which are explained
-in~\S\ref{sec-abstractions}:
-
-\begin{minted}{haskell}
-type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
-\end{minted}
-
-Below we define basic notions used in build systems and other similar domains,
-for example, spreadsheets.
-
-\subsection{Requirements for build systems}
-
-\begin{itemize}
-    \item Correctness
-    \item Minimality
-    \item Support for sharing and skipping intermediate values
-\end{itemize}
-...
+\cmd{main.c} can be edited by the user who relies on the build system to
+compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
+such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
+\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
+but are produced in the process of turning inputs into outputs.

### 2018-03-09 paper/engineering.tex
Finish S2

@@
 \end{itemize}
+
+\subsection{Non-deterministic computations}
+
+Build systems and spreadsheets compute output values from input and intermediate
+values. In the most typical case, these \emph{computations} are \emph{functions},
+such as \textsf{C1 = A1 + B1}, i.e. their result is uniquely determined by the
+input values. However, in general they can be \emph{relations}, i.e. have
+multiple valid results. A spreadsheet example: \textsf{A2 = A1 + RANDOM(1,6)}.
+This computation has six valid results for each input value \textsf{A1}. In
+build systems, the object file \textsf{obj/file.o} is sometimes not uniquely
+determined by the source \textsf{src/file.c} -- different compiler runs may
+produce different valid results.
+
+Some build systems, e.g. \Buck, require all tasks to be
+\emph{deterministic}, i.e. produce exactly the same output when run on the
+same inputs. However, not all tasks are deterministic, and there are build
+systems that support \emph{non-determinism}. A simple example is \Excel's
+\textsf{RANDBETWEEN(low, high)} that returns a random integer in the
+interval \textsf{[low, high]}.

### 2018-03-09 paper/abstractions.tex
Add draft of S3

--- 2018-03-07 paper/abstractions.tex
@@
-\clearpage
-
 \section{Build systems, abstractly}\label{sec-abstractions}
 
-In this section we present the \emph{compute} and \emph{build} abstractions:
-\begin{itemize}
-    \item Compute represents build rules (in build systems) or formulas (in
-    spreadsheets), and is completely isolated from the world of compilers, file
-    systems, memories, caches, and all other complexities of real build systems.
-    \item Build corresponds to the algorithm used for bringing a key-value store
-    to a coherent state by running compute.
-\end{itemize}
+This section presents purely functional abstractions for build systems that
+allow us to express all the intricacies of build systems discussed in the
+previous section~\S\ref{sec-background} and design complex build systems from
+simple primitives. Specifically, we present the \emph{compute} and \emph{build}
+abstractions:
 
-\begin{figure}
-\begin{minted}{haskell}
+\begin{minted}[xleftmargin=10pt]{haskell}
 type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
 \end{minted}
-\caption{Compute abstractions}\label{fig-compute}
-\end{figure}
 
-Consider abstractions in Fig.~\ref{fig-compute}.
+\noindent
+Compute represents build rules (in build systems) or formulas (in spreadsheets),
+and is completely isolated from the world of compilers, file systems, memories,
+caches, and all other complexities of real build systems. We explain the compute
+type in~\S\ref{sec-compute}, and then discuss several noteworthy special cases:
+functorial (\S\ref{sec-compute-functor}), applicative
+(\S\ref{sec-compute-applicative}) and monadic (\S\ref{sec-compute-monad}) types
+of compute.
+
+Build corresponds to the algorithm used for bringing a key-value store to a
+consistent state by recomputing dirty values in an appropriate order. We explain
+the build type in~\S\ref{sec-build}, and then provide implementations for several
+build systems in~\S\ref{sec-examples}.
+
+\subsection{Compute}\label{sec-compute}
+
+% \begin{figure}
+% \begin{minted}{haskell}
+% type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+% \end{minted}
+% \caption{Compute abstractions}\label{fig-compute}
+% \end{figure}
+% Consider abstractions in Fig.~\ref{fig-compute}.
+
+\subsection{Functorial compute}\label{sec-compute-functor}
+\subsection{Applicative compute}\label{sec-compute-applicative}
+\subsection{Monadic compute}\label{sec-compute-monad}
+
+\subsection{Build}\label{sec-build}
 
 ...
-
-We now describe several types of build systems according to their constraint
-on the functor \hs{f}.
-
-...

### 2018-03-09 paper/abstractions.tex
Work on S3

@@
 
+...
 % \begin{figure}
@@
 \subsection{Functorial compute}\label{sec-compute-functor}
+
+The \emph{Collatz sequence} $C_i$ is defined as follows:
+
+\[
+C_{i} = {\begin{cases}~n&{\text{for }}i=0\\~f(C_{i-1})&{\text{otherwise}},\end{cases}}\hspace{12pt}\text{where}\hspace{12pt}f(k)={\begin{cases}~k/2&{\text{if }}k\text{ is even}\\~3k+1&{\text{otherwise}}\end{cases}}
+\vspace{2mm}
+\]
+\noindent
+and $n$ is a positive integer parameter. The famous \emph{Collatz conjecture}
+states that the Collatz sequence eventually reaches 1 for all possible values of
+$n$. For example, if $n=6$, we reach 1 in eight steps:
+$(6, 3, 10, 5, 16, 8, 4, 2, 1, \dots)$, after which the sequence loops forever:
+$(4, 2, 1, 4, 2, 1, \dots)$.
+
+We can express the computation of values in the Collatz sequence as a functorial
+compute:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Collatz = Collatz Int
+
+collatz :: Compute Functor Collatz Int
+collatz get (Collatz k) | k <= 0    = Nothing
+                        | otherwise = Just $ f <$> get (Collatz (k - 1))
+  where
+    f n | even n    = n `div` 2
+        | otherwise = 3 * n + 1
+\end{minted}
+
+...
+
 \subsection{Applicative compute}\label{sec-compute-applicative}
+The \emph{generalised Fibonacci sequence} $F_i$ is defined as follows:
+
+\[
+F_{i} = {\begin{cases}~n&{\text{for }}i=0\\~m&{\text{for }}i=1\\~F_{i-1}+F_{i-2}&{\text{otherwise}}\end{cases}}
+\vspace{2mm}
+\]
+\noindent
+where $n$ and $m$ are integer parameters. By setting $n=0$ and $m=1$ we obtain
+the famous \emph{Fibonacci sequence}: $(0, 1, 1, 2, 3, 5, 8, 13, \dots$), and if
+$n=2$ and $m=1$, the result is the \emph{Lucas sequence}:
+$(2, 1, 3, 4, 7, 11, 18, 29, \dots)$.
+
+We can express the computation of values in the generalised Fibonacci sequence
+as an applicative compute:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Fibonacci = Fibonacci Int
+
+fibonacci :: Compute Applicative Fibonacci Int
+fibonacci get (Fibonacci k) | k <= 1    = Nothing
+                            | otherwise = Just $ (+) <$> get (Fibonacci (k - 1))
+                                                     <*> get (Fibonacci (k - 2))
+\end{minted}
+
+...
+
 \subsection{Monadic compute}\label{sec-compute-monad}
+
+The \emph{Ackermann function} $A(m, n)$ is defined as follows:
+
+\[
+A(m, n) = {\begin{cases}~n+1&{\text{for }}m=0\\~A(m-1, 1)&{\text{for }}n=0\\~A(m-1,A(m,n-1))&{\text{otherwise}}\end{cases}}
+\vspace{2mm}
+\]
+\noindent
+We can express the computation of the Ackermann function as a monadic compute:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Ackermann = Ackermann Int Int
+
+ackermann :: Compute Monad Ackermann Int
+ackermann get (Ackermann m n)
+    | m < 0 || n < 0 = Nothing
+    | m == 0    = Just $ return (n + 1)
+    | n == 0    = Just $ get (Ackermann (m - 1) 1)
+    | otherwise = Just $ do
+        index <- get (Ackermann m (n - 1))
+        get (Ackermann (m - 1) index)
+\end{minted}
+
+...
 

### 2018-03-09 paper/background.tex
Add draft of S3
Expand engineering aspects section

@@
 more than 40 years ago to automatically build software libraries and executable
-programs from source code. It uses \emph{makefiles} to describe build tasks and
-their dependencies in a simple text form. For example:
+programs from source code. It uses \emph{makefiles} to describe tasks (often
+referred to as \emph{build rules}) and their dependencies in a simple text form.
+For example:
 
@@
 of volatility includes dynamic dependencies, as well as \emph{non-determinism}
-that will be discussed in~\S\ref{sec-engineering}. We therefore put both \Make
-and \Excel in the same category of build systems that associate a \emph{dirty bit}
-with every file or cell. (To avoid the ``file or cell'' nuisance, we introduce the
-abstract notion of \emph{keys} in~\S\ref{sec-background-vocabulary}.)
+that will be discussed in~\S\ref{sec-non-determinism}. We therefore put both
+\Make and \Excel in the same category of build systems that associate a
+\emph{dirty bit} with every file or cell. (To avoid the ``file or cell''
+nuisance, we introduce the abstract notion of \emph{keys}
+in~\S\ref{sec-background-vocabulary}.)
 
@@
 storage resources for the sake of simplicity; in practice, old entries are
-regularly evicted from the storage, as further discussed in~\S\ref{sec-engineering}.}.
-The latter allows to bypass the execution of a task, by predicting the hash of
-its result from the hashes of the dependencies, and subsequently fetch the result
-from the cache. A concrete implementation is provided in~\S\ref{sec-examples}.
+regularly evicted from the storage, as further discussed
+in~\S\ref{sec-sec-cloud-aspects}.}. The latter allows to bypass the execution of
+a task, by predicting the hash of its result from the hashes of the dependencies,
+and subsequently fetch the result from the cache. A concrete implementation is
+provided in~\S\ref{sec-examples}.
 

### 2018-03-09 paper/engineering.tex
Expand engineering aspects section

@@
 
-\begin{itemize}
-    \item Sandboxing
-    \item Non-determinism, specifically, its useless form. This doesn't
+\subsection{Non-deterministic computations}\label{sec-non-determinism}
+
+Sandboxing: guarding against non-determinism due to missing dependencies.
+
+\todo{AM}{Rewrite everything below.}
+
+Non-determinism, specifically, its \emph{useless} form. This doesn't
 include \Excel's \textsf{RANDBETWEEN} function, which is useful.
@@
 non-deterministic results (with a notable exception of \Buck).
-    \item Concurrency: nothing fundamentally difficult and fits our theory, but
-          hard to get right in practice.
-    \item Cloud aspects: eviction/garbage collection, `frankenbuilds', etc.
-    \item Iterative computations, e.g. LaTeX.
-    \item Polymorphic keys and values: tasks with multiple outputs, e.g. GHC's
-          \cmd{hi} and \cmd{o} files, phony tasks, etc.
-\end{itemize}
 
-\subsection{Non-deterministic computations}
+... Some old text:
 
@@
 interval \textsf{[low, high]}.
+
+\subsection{Concurrency}\label{sec-concurrency}
+
+Nothing fundamentally difficult and fits our theory, but hard to get right in
+practice.
+
+\todo{NM}{Add a paragraph on how to introduce concurrency to implementations
+in~\S\ref{sec-examples}}.
+
+\subsection{Cloud aspects}\label{sec-cloud-aspects}
+
+Cloud aspects: eviction/garbage collection, `frankenbuilds', etc.
+
+\todo{AM}{Descibe frankenbuilds linking to CloudBuild paper.}
+
+\subsection{Tracking and self-tracking}\label{sec-tracking-aspects}
+
+Annotating dependency graphs with hashes, times, etc. is in many cases just an
+engineering trade-off.
+
+\todo{NM}{Clarify, e.g. Shake's \cmd{ChangeModtime}.}
+
+\todo{NM}{Explain how self-tracking can be implemented.}
+
+\subsection{Iterative computations}\label{sec-iterative-compute}
+
+Iterative computations, e.g. LaTeX. \Excel has built-in support for this.
+
+\todo{NM}{Clarify how this fits with the example implementations.}
+
+\subsection{Polymorphism}\label{sec-polymorphism}
+
+Polymorphic keys and values: tasks with multiple outputs, e.g. GHC's \cmd{hi}
+and \cmd{o} files, phony tasks, etc.
+
+\todo{AM}{Give an example.}
+
+\todo{NM}{Describe how to add this feature to \Shake.}

### 2018-03-09 paper/background.tex
Address some of Neil's comments

@@
 often needs to manually initiate a full rebuild even if just a single task has
-changed.
+changed. We further discuss self-tracking in~\S\ref{sec-tracking-aspects}.
 
@@
 
-The last row of the table describes the new build system proposed in this paper,
-which shares the advantages of \Shake and \Bazel.
+In this paper we show which build system properties are consequences of specific
+implementation choices (metadata and algorithm), and how one can obtain new
+build systems with desired properties by recombining parts of existing
+implementations. As a compelling example, we demonstrate how to combine
+advantages of \Shake and \Bazel in a cloud build system with dynamic dependencies.
 
@@
 \centering
-\begin{tabular}{l||l|c|c|c|c||l|l}
+\begin{tabular}{l||l|c|c|c||l|l}
 \hline
-$\!\!$Build system$\!$&$\!$Dependencies$\!$&$\!$Minimal$\!$&$\!$Self-tracking$\!$&$\!$Cutoff$\!$&$\!$Cloud$\!$&$\!$Metadata$\!$&$\!$Algorithm$\!\!$\\\hline
-$\!\!$\Make       $\!$&$\!$Static  & Yes & No  & No  & No  &$\!$Dirty bit       $\!$&$\!$Topological $\!\!$\\
-$\!\!$\Excel      $\!$&$\!$Dynamic & No  & Yes & No  & No  &$\!$Dirty bit, chain$\!$&$\!$Chain, defer$\!\!$\\
-$\!\!$\Shake      $\!$&$\!$Dynamic & Yes & No  & Yes & No  &$\!$Graph           $\!$&$\!$Recursive   $\!\!$\\
-$\!\!$\Bazel      $\!$&$\!$Static  & Yes & No  & Yes & Yes &$\!$Graphs, cache   $\!$&$\!$Topological $\!\!$\\\hline
-$\!\!$This paper  $\!$&$\!$Dynamic & Yes & No? & Yes & Yes &$\!$Graphs, cache   $\!$&$\!$Recursive   $\!\!$\\
+$\!\!$Build system & Dependencies & Minimal & Cutoff & Cloud & Metadata                       & Algorithm   $\!\!$\\\hline
+$\!\!$\Make        & Static       & Yes     & No     & No    & File modification times        & Topological $\!\!$\\
+$\!\!$\Excel       & Dynamic      & No      & No     & No    & Dirty cells, calculation chain & Chain, defer$\!\!$\\
+$\!\!$\Shake       & Dynamic      & Yes     & Yes    & No    & Last dependency graph          & Recursive   $\!\!$\\
+$\!\!$\Bazel       & Static       & Yes     & Yes    & Yes   & All dependency graphs, cache   & Topological $\!\!$\\\hline
 \hline

### 2018-03-09 paper/background.tex
Restructure the paper

@@
 moving it down the chain when one of its newly discovered dynamic dependencies
-is dirty. We discuss this algorithm in more detail in~\S\ref{sec-examples}.
+is dirty. We discuss this algorithm in more detail in~\S\ref{sec-implementations}.
 
@@
 \subsection{\Bazel: a cloud build system}
-\label{sec-background-shake}
+\label{sec-background-bazel}
 
@@
 regularly evicted from the storage, as further discussed
-in~\S\ref{sec-sec-cloud-aspects}.}. The latter allows to bypass the execution of
+in~\S\ref{sec-cloud-aspects}.}. The latter allows to bypass the execution of
 a task, by predicting the hash of its result from the hashes of the dependencies,
 and subsequently fetch the result from the cache. A concrete implementation is
-provided in~\S\ref{sec-examples}.
+provided in~\S\ref{sec-implementations}.
 
@@
     system to do that. In principle a single dirty bit is sufficient, as will be
-    demonstrated in~\S\ref{sec-examples}.
+    demonstrated in~\S\ref{sec-implementations}.
     \item \Excel stores the dirty bit as well as the calculation chain.

### 2018-03-09 paper/build.tex
Restructure the paper

\section{Build systems, abstractly}\label{sec-build}

In this section we further analyse the \hs{Build i c k v} abstraction, in
particular, we analyse two main aspects of every build system: the metadata
\hs{i} it persistently stores between builds, and the algorithm it uses to
compute values in the dependency order.

\subsection{Metadata: dirty bit, dependency graph, cache}\label{sec-build-metadata}

...

\subsection{Algorithm: topological, reordering, recursive}\label{sec-build-algorithm}

...

### 2018-03-09 paper/compute.tex
Add draft of the background section
Restructure the paper

@@
-\section{Compute}\label{sec-compute}
+\section{Build tasks, abstractly}\label{sec-compute}
 
-\begin{figure}
-\begin{minted}{haskell}
-type IdentityCompute    k v = forall f.                  (k -> f v) -> k -> f v
-type FunctorialCompute  k v = forall f. Functor     f => (k -> f v) -> k -> f v
-type ApplicativeCompute k v = forall f. Applicative f => (k -> f v) -> k -> f v
-type AlternativeCompute k v = forall f. Alternative f => (k -> f v) -> k -> f v
-type MonadicCompute     k v = forall m. Monad       m => (k -> m v) -> k -> m v
-type MonadPlusedCompute k v = forall m. MonadPlus   m => (k -> m v) -> k -> m v
+This section presents purely functional abstractions for build systems that
+allow us to express all the intricacies of build systems discussed in the
+previous section~\S\ref{sec-background} and design complex build systems from
+simple primitives. Specifically, we present the \emph{compute} and \emph{build}
+abstractions:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
 \end{minted}
-\caption{Compute abstractions}\label{fig-compute}
-\end{figure}
 
-Consider abstractions in Fig.~\ref{fig-compute}.
+\noindent
+Compute represents build rules (in build systems) or formulas (in spreadsheets),
+and is completely isolated from the world of compilers, file systems, memories,
+caches, and all other complexities of real build systems. We explain the compute
+type in~\S\ref{sec-general-compute}, and then discuss several noteworthy special cases:
+functorial (\S\ref{sec-compute-functor}), applicative
+(\S\ref{sec-compute-applicative}) and monadic (\S\ref{sec-compute-monad}) types
+of compute.
+
+Build corresponds to the algorithm used for bringing a key-value store to a
+consistent state by recomputing dirty values in an appropriate order. We explain
+the build type in~\S\ref{sec-build}, and then provide implementations for several
+build systems in~\S\ref{sec-implementations}.
+
+\subsection{Compute}\label{sec-general-compute}
+
+...
+% \begin{figure}
+% \begin{minted}{haskell}
+% type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+% \end{minted}
+% \caption{Compute abstractions}\label{fig-compute}
+% \end{figure}
+% Consider abstractions in Fig.~\ref{fig-compute}.
+
+\subsection{Functorial compute}\label{sec-compute-functor}
+
+The \emph{Collatz sequence} $C_i$ is defined as follows:
+
+\[
+C_{i} = {\begin{cases}~n&{\text{for }}i=0\\~f(C_{i-1})&{\text{otherwise}},\end{cases}}\hspace{12pt}\text{where}\hspace{12pt}f(k)={\begin{cases}~k/2&{\text{if }}k\text{ is even}\\~3k+1&{\text{otherwise}}\end{cases}}
+\vspace{2mm}
+\]
+\noindent
+and $n$ is a positive integer parameter. The famous \emph{Collatz conjecture}
+states that the Collatz sequence eventually reaches 1 for all possible values of
+$n$. For example, if $n=6$, we reach 1 in eight steps:
+$(6, 3, 10, 5, 16, 8, 4, 2, 1, \dots)$, after which the sequence loops forever:
+$(4, 2, 1, 4, 2, 1, \dots)$.
+
+We can express the computation of values in the Collatz sequence as a functorial
+compute:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Collatz = Collatz Int
+
+collatz :: Compute Functor Collatz Int
+collatz get (Collatz k) | k <= 0    = Nothing
+                        | otherwise = Just $ f <$> get (Collatz (k - 1))
+  where
+    f n | even n    = n `div` 2
+        | otherwise = 3 * n + 1
+\end{minted}
+
+...
+
+\subsection{Applicative compute}\label{sec-compute-applicative}
+The \emph{generalised Fibonacci sequence} $F_i$ is defined as follows:
+
+\[
+F_{i} = {\begin{cases}~n&{\text{for }}i=0\\~m&{\text{for }}i=1\\~F_{i-1}+F_{i-2}&{\text{otherwise}}\end{cases}}
+\vspace{2mm}
+\]
+\noindent
+where $n$ and $m$ are integer parameters. By setting $n=0$ and $m=1$ we obtain
+the famous \emph{Fibonacci sequence}: $(0, 1, 1, 2, 3, 5, 8, 13, \dots$), and if
+$n=2$ and $m=1$, the result is the \emph{Lucas sequence}:
+$(2, 1, 3, 4, 7, 11, 18, 29, \dots)$.
+
+We can express the computation of values in the generalised Fibonacci sequence
+as an applicative compute:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Fibonacci = Fibonacci Int
+
+fibonacci :: Compute Applicative Fibonacci Int
+fibonacci get (Fibonacci k) | k <= 1    = Nothing
+                            | otherwise = Just $ (+) <$> get (Fibonacci (k - 1))
+                                                     <*> get (Fibonacci (k - 2))
+\end{minted}
+
+...
+
+\subsection{Monadic compute}\label{sec-compute-monad}
+
+The \emph{Ackermann function} $A(m, n)$ is defined as follows:
+
+\[
+A(m, n) = {\begin{cases}~n+1&{\text{for }}m=0\\~A(m-1, 1)&{\text{for }}n=0\\~A(m-1,A(m,n-1))&{\text{otherwise}}\end{cases}}
+\vspace{2mm}
+\]
+\noindent
+We can express the computation of the Ackermann function as a monadic compute:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Ackermann = Ackermann Int Int
+
+ackermann :: Compute Monad Ackermann Int
+ackermann get (Ackermann m n)
+    | m < 0 || n < 0 = Nothing
+    | m == 0    = Just $ return (n + 1)
+    | n == 0    = Just $ get (Ackermann (m - 1) 1)
+    | otherwise = Just $ do
+        index <- get (Ackermann m (n - 1))
+        get (Ackermann (m - 1) index)
+\end{minted}
+
+...
+
+\subsection{Build}\label{sec-general-build}
+
+...

### 2018-03-09 paper/implementations.tex
Restructure the paper

@@
 \clearpage
-\section{Build systems, concretely}\label{sec-examples}
+\section{Build systems, concretely}\label{sec-implementations}
 
 In the previous sections we've discussed the types of build system in text, but the way we came to our conclusions was by developing a framework to fit the builds in.
-In this section we explain the framework, some of the abstractions we introduced (which model the properties from \S\ref{todo}), and how we can make a meaningful implementation of the build systems discussed in \S\ref{sec-background}.
+In this section we explain the framework, some of the abstractions we introduced (which model the properties from \S\ref{sec-background}), and how we can make a meaningful implementation of the build systems discussed in \S\ref{sec-background}.
 
@@
 
-\subsection{\Make}
+\subsection{\Make}\label{sec-implementation-make}
 
@@
 
-\subsection{Spreadsheet}
+\subsection{\Excel}\label{sec-implementation-excel}
 
@@
 
-\subsection{Shake}
+\subsection{\Shake}\label{sec-implementation-shake}
 
@@
 
-\subsection{Bazel}
+\subsection{\Bazel}\label{sec-implementation-bazel}
 
@@
 
-The state matches that described in \S\ref{todo} - we have a map indexed by a target key, and the hashes of the dependencies of that key.
+The state matches that described in \S\ref{sec-build} - we have a map indexed by a target key, and the hashes of the dependencies of that key.
 Given that, we point at a resultant hash.
@@
 
-\subsection{Cloud Shake}
+\subsection{Cloud \Shake}\label{sec-implementation-cloud-shake}
 

### 2018-03-09 paper/background.tex
Add dynamic dependencies example

@@
 
+\begin{figure}[h]
+\begin{subfigure}[b]{0.90\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
+\vspace{-1mm}
+\caption{Task dependency graph produced after the previous build.}
+\vspace{3mm}
+\end{subfigure}
+\begin{subfigure}[b]{0.90\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
+\vspace{-1mm}
+\caption{Since \cmd{docs.txt} is dirty, we rebuild \cmd{release.txt} and
+\cmd{release.tar}, discovering a new dependency.}
+\end{subfigure}
+\caption{Dynamic dependencies example: create \cmd{README} and add it to the
+list of release documents \cmd{docs.txt}.\label{fig-shake}}
+\end{figure}
+
 \Shake's implementation is different from both \Make and \Excel in two aspects.

### 2018-03-10 paper/background.tex
Rewrite S2.1-S2.2

@@
 If the user modifies the sources of the utility library and runs \Make, it will
-perform a full rebuild, because all three tasks transitively depend on the
-library, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
-user modifies \cmd{main.c} then a partial rebuild is sufficient: indeed, the
-file \cmd{util.o} does not need to be rebuilt, since its inputs have not
-changed, see Fig.~\ref{fig-make}(c). Files that have changed since the previous
-build are called \emph{dirty}.
+perform a \emph{full rebuild}, because all three tasks transitively depend on
+the library, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
+user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient: indeed,
+the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
+changed, see Fig.~\ref{fig-make}(c). Note that if the dependency graph is
+\emph{acyclic} then each task needs to be executed at most once. Cyclic task
+dependencies are typically not allowed in build systems although there are rare
+exceptions, see~\S\ref{sec-iterative-compute}.
 
-The requirement to \emph{execute tasks at most once and only if they transitively
-depend on dirty inputs} is essential for build systems, it is their raison
-d'\^etre. We will call build systems that satisfy this requirement \emph{minimal}.
+The following property is essential for build systems, it is their raison d'\^etre.
 
-To achieve minimality \Make relies on two main ideas. First, it uses \emph{file
-modification time} to detect the files that are dirty: a file is marked dirty if
-it was modified since the previous build. Second, it constructs a complete task
-dependency graph from the information contained in the makefile and executes
-tasks in a \emph{topological order}.
+\definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
+most once per build and only if they transitively depend on files that changed
+since the previous build.}
+\vspace{2mm}
+
+To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
+modification time} to detect the files that changed\footnote{Technically, you
+can fool \Make by altering the modification time of a file without changing its
+content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
+under the assumption that you do not do that.}, and (ii) it constructs a
+complete task dependency graph from the information contained in the makefile
+and executes tasks in a \emph{topological order}. See further details
+in~\S\ref{sec-implementation-make}.
+
+\subsection{\Excel: dynamic dependencies at the cost of minimality}
+\label{sec-background-excel}
+
+\Excel is a build system in disguise. Consider the following simple spreadsheet.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
+A1 = 10
+A2 = 20
+B1 = A1 + A2
+\end{minted}
+\vspace{1mm}
+
+\noindent
+There are two input cells \cmd{A1} and \cmd{A2}, and a single task that computes
+the sum of their values, writing the result into the cell \cmd{B1}. If either of
+the inputs change, \Excel will recompute the result.
+
+Unlike \Make, \Excel does not need to know all task dependencies upfront. Some
+dependencies may change \emph{dynamically} according to computation results. For
+example:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
+A1 = 10
+A2 = 20
+B1 = IF(C1=1,A1,A2)
+C1 = 1
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
+to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
+$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and this is
+not known statically. One might say that the value of \cmd{C1} is statically
+known in this example, but imagine that it is a result of a long computation
+chain -- its value will then only become available during the build.
+
+\Excel handles this example correctly: if \cmd{C1=1} and the user changes
+\cmd{A2}, \Excel will not recompute the result \cmd{B1}. Alas, other forms of
+dynamic dependencies can force \Excel to perform unnecessary computation.
+Consider the following modification of the above example:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
+A1 = 10
+A2 = 20
+B1 = INDIRECT("A" & C1)
+C1 = 1
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The new version uses the \cmd{INDIRECT} function, which allows us to reference a
+cell indirectly by a text string that is not necessarily known in advance. The
+current implementation of \Excel recomputes indirect references in every
+build~\cite{excel_recalc}. This clearly violates the minimality property: if
+\cmd{C1=1} and the user modifies \cmd{A2}, \Excel will recompute \cmd{B1},
+potentially triggering further unnecessary recalculation even though \cmd{B1}
+does not transitively depend on \cmd{A2}.
+
+\Excel's build algorithm~\cite{excel_recalc} is significantly different from
+\Make. Since it is impossible to construct the full dependency graph upfront in
+the presence of dynamic dependencies, \Excel uses the \emph{calculation chain}
+produced by the previous build as an approximation to the correct topological
+order. During recalculation, \Excel processes cells in this order, but can
+\emph{defer recalculation of a cell} by moving it down the chain if a newly
+discovered dependency has not yet been rebuilt. We further refer to this
+algorithm as \emph{reordering}, and will discuss it in more detail
+in~\S\ref{sec-implementation-excel}.
+
+Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
+systems track changes of inputs and intermediate results, executing dependent
+tasks whenever they change, but \Excel can also track changes in the tasks
+themselves: if a formula is modified, \Excel will recompute it and propagate
+the changes. Self-tracking is uncommon in software build systems, where one
+often needs to manually initiate a full rebuild even if just a single task has
+changed. We further discuss self-tracking in~\S\ref{sec-tracking-aspects}.
+
+\subsection{\Shake: dynamic dependencies with no remorse}
+\label{sec-background-shake}
 
@@
 are problematic~\cite{hadrian}.
-
-\subsection{\Excel: dynamic dependencies at the cost of minimality}
-\label{sec-background-excel}
-
-\Excel is a build system that can deal with dynamic dependencies albeit at the
-cost of sacrificing minimality. Consider the following spreadsheet comprising
-four input cells \cmd{A1-A3} and \cmd{B1}, and a single task computing the sum
-of values \cmd{A1} $+\cdots+$ \cmd{An}, where \cmd{n} is specified indirectly in
-\cmd{B1}:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{bash}
-A1 = 10
-A2 = 20
-A3 = 30
-B1 = 2
-C1 = SUM(A1:INDIRECT("A" & B1))
-\end{minted}
-\vspace{1mm}
-
-\noindent
-The output \cmd{C1} here is analogous to the \cmd{release.tar} from the previous
-example: one cannot statically determine the dependencies of the associated task.
-To deal with this, \Excel takes a simple conservative approach: it marks all
-cells with dynamic dependencies as dirty and therefore always recomputes them
-during a rebuild. This guarantees that the build result is correct, but clearly
-violates the minimality requirement: if the user modifies \cmd{A3}, \Excel will
-recompute \cmd{C1} even though it does not transitively depend on \cmd{A3}.
-
-Note that \Excel's approach to marking cells as dirty is a relatively minor
-variation of the approach used by \Make: a cell is considered dirty if it was
-modified since the previous build \emph{or if it is volatile}, where the notion
-of volatility includes dynamic dependencies, as well as \emph{non-determinism}
-that will be discussed in~\S\ref{sec-non-determinism}. We therefore put both
-\Make and \Excel in the same category of build systems that associate a
-\emph{dirty bit} with every file or cell. (To avoid the ``file or cell''
-nuisance, we introduce the abstract notion of \emph{keys}
-in~\S\ref{sec-background-vocabulary}.)
-
-\Excel's recalculation algorithm~\cite{excel_recalc} on the other hand is
-significantly different from \Make. Since it is impossible to construct the full
-dependency graph upfront in the presence of dynamic dependencies, \Excel
-maintains the \emph{calculation chain}, which is an approximation to the correct
-topological order produced by the previous build. During recalculation, \Excel
-processes cells in this order, but can \emph{defer recalculation of a cell} by
-moving it down the chain when one of its newly discovered dynamic dependencies
-is dirty. We discuss this algorithm in more detail in~\S\ref{sec-implementations}.
-
-Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
-systems track changes of inputs and intermediate results, executing dependent
-tasks whenever they change, but \Excel can also track changes in the tasks
-themselves: if a formula is modified, \Excel will recompute it and propagate
-the changes. Self-tracking is uncommon in software build systems, where one
-often needs to manually initiate a full rebuild even if just a single task has
-changed. We further discuss self-tracking in~\S\ref{sec-tracking-aspects}.
-
-\subsection{\Shake: dynamic dependencies with no remorse}
-\label{sec-background-shake}
 
@@
 \vspace{-1mm}
-\caption{Task dependency graph produced after the previous build.}
+\caption{Dependency graph produced after the previous build.}
 \vspace{3mm}

### 2018-03-10 paper/build.tex

@@
 
-In this section we further analyse the \hs{Build i c k v} abstraction, in
-particular, we analyse two main aspects of every build system: the metadata
-\hs{i} it persistently stores between builds, and the algorithm it uses to
-compute values in the dependency order.
+In this section we further describe our \hs{Build} abstraction, defined as:
 
-\subsection{Metadata: dirty bit, dependency graph, cache}\label{sec-build-metadata}
+\begin{minted}{haskell}
+type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
+\end{minted}
 
-...
+The build function takes 4 type arguments: \hs{c} is the constraint used for \hs{Compute} (typically either \hs{Applicative} or \hs{Monad}), \hs{i} is the information stored by the build system and \hs{k}/\hs{v} are the key/value types as described in \S\ref{todo}.
+Given a build system \hs{build}:
 
-\subsection{Algorithm: topological, reordering, recursive}\label{sec-build-algorithm}
+\begin{minted}{haskell}
+build compute key info store = (info', store')
+    where ...
+\end{minted}
 
-...
+The function \hs{build} is responsible for ensuring that \hs{key} and its dependencies are up-to-date in \hs{store'} starting from \hs{store}.
+As part of that work, the \hs{build} function \textit{may} store other information in \hs{info'}, which will be provided as \hs{info} in future calls to \hs{build}.
+The \hs{info} parameter will be \hs{Nothing} on the very first call to \hs{build}.
+
+All the \hs{Build} types we define are polymorphic in both \hs{k} and \hs{v} (perhaps adding constraints to them), and have specific instantiations of \hs{c} and \hs{i}.
+How to implement \hs{Build} can be captured by the idea:
+
+\begin{quote}
+A build system must build \underline{out-of-date targets} while \underline{respecting dependency orderings}.
+\end{quote}
+
+We have underlined two different phrases in this sentence, and tackle each aspect separately.
+
+\subsection{Respecting dependency orderings}
+
+If the \hs{Compute} function for \hs{k} calls the functional argument of \hs{d} then the build system is responsible for ensuring that \hs{d} is built \textit{before} \hs{k}.
+Looking at the build systems in \S\ref{todo} we have observed 3 distinct approaches to respecting dependency orderings.
+In this section of the paper we'll explore what properties are a consequence and how they operate.
+
+\subsubsection{Topological sort}
+
+The topological approach creates a linear ordering, which when followed, ensures each build is correct.
+Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort.
+However, to go from a \hs{Compute} to the dependencies, it must be the case that it is Applicative.
+
+\subsubsection{Reordering}
+
+The topological sort approach has two downside - it is limited to \hs{Applicative} build systems and requires a fresh topological sort each time - while the actions themselves may be incremental the pre-processing is not.
+An alternative approach is to keep the topological order between runs and assume it to be correct, but if execution determines it is wrong, reorder.
+
+This requires a way to abort computations that have failed.
+It's also not minimal in the sense that computations start, may do some meaningful work, then abort.
+However, in the case of an \hs{Applicative} system, that work is zero.
+It does require the ability to terminate a running computation.
+
+\subsubsection{Recursive}
+
+An alternative way is to simply build dependencies when they are requested.
+By combining that with a transient store of which keys have been built in this execution, you can describe dependencies in a simple way.
+You need a way to suspend a running action, which can be done with cheap green-threads and blocking (the original approach of Shake) or with a continuation-passing style (what Shake does now).
+An alternative approach to suspending a computation is to abort it and restart it again later, at the cost of doing additional work.
+
+\subsection{Determining what is out-of-date}
+
+The second part, determining what to rebuild, can be addressed in one of three fundamental ways, with a number of tweaks and variations within them.
+
+\subsubsection{A dirty bit}
+
+The idea of a dirty bit is to have one piece of information, whether a key is dirty or clean.
+After a run, all bits are set to clean.
+When a run starts anything that changed between the two states is marked dirty - by marking additional things dirty/clean the build system can track what needs to rebuild.
+When reaching a key, if it and all its dependencies are clean, the key does not need recomputing.
+
+\Excel models the dirty bit approach most directly, having an actual dirty bit associated with each cell.
+Since Excel is the only manner by which a cell may be changed by the user, it can dirty the bits as necessary.
+When rebuilding if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and marked dirty itself.
+The only wrinkle in this scheme is that \Excel supports monadic build systems, and does not separately record what rebuilt, so it has to approximate in this case.
+
+\Make uses file modification times, and compares files to their dependencies, which can be thought of as a dirty bit which is set when a file is newer than its dependencies.
+The interesting property of this dirty bit is that it is not under control of \Make, but an existing bit that is repurposed.
+In particular, updating a file automatically clears its dirty bit, and automatically sets the dirty bit of the nodes depending on it.
+One thing \Make does require is that file timestamps only go forward in time - something that can be violated by some backup software.
+
+With a dirty bit it is possible to achieve minimality.
+To achieve cut-off it would be important to not set the dirty bit after a computation that did not change the value.
+\Excel could model this approach, but does not.
+In contrast, \Make cannot model cut-off - to do so it would have to mark the node clean (so it didn't rebuild in the next run) and at the same time not mark the things it depends on dirty - an impossible task with only the ability to update to the latest modification time.
+
+\subsubsection{Recording predicate trace}
+
+An alternative way to determine if a key is dirty is to record what state the values/hashes of dependencies were used at last time, and if something has changed, the key is dirty and must be recomputed.
+For a monadic build system the trace can include what dependencies were used as well, although for an applicative system that information can be inferred.
+For each key, a single trace is recorded, so it naturally fits that the data is stored in an associative map indexed by target key.
+
+The recording of a trace requires storing extra information with the execution, but otherwise provides a simple approach which is capable of achieving all the properties without imposing restrictions.
+
+\subsubsection{Recording generative traces}
+
+A predicate trace allows us to mark the key dirty and rebuild it, extending that we can store a \textit{generative} trace which is the trace plus the complete result.
+Once we are storing the complete result it makes sense to record many generative traces per key, and to share them with other users, providing cloud-build functionality.
+
+For \hs{Applicative} computations, the traces need not record the dependency keys (they are implicit given the target key), and can be indexed by the hashes of the keys.
+Concretely, we could imagine storing:
+
+\begin{minted}{haskell}
+data GenerativeApplicativeTrace k v = GAT
+    { traces :: Map (k, [Hash v]) (Hash v)
+    , contents :: Map (Hash v) v
+    }
+\end{minted}
+
+Given a target key, we can grab the dependencies, grab their hashes and look up the associated result hash.
+If it doesn't match what we have on disk, we download it from the \hs{contents}.
+If the entry is not in \hs{traces} we can run it and add it.
+
+For \hs{Monadic} computations the situation is slightly more tricky - instead of being able to determine the dependencies in advance, we effectively have a set of traces for any given key, any of which could apply.
+While no existing build system actually implements monadic cloud builds, we can describe a plausible structure as:
+
+\begin{minted}{haskell}
+data Trace k v = Trace [(k, Hash v)] (Hash v)
+type GenerativeMonadicTrace k v = GMT
+    { traces :: Map k [Trace k v]
+    , contents :: Map (Hash v) v
+    }
+\end{minted}
+
+Now we can grab the list of traces for a given key, and see which (if any) matches our current state.
+In practice, to avoid an $O(n)$ scan of the number of traces, we can common up prefixes and use the data structure:
+
+\begin{minted}{haskell}
+data Trace k v = Trace k (Map (Hash v) (Trace k v))
+               | Done (Hash v)
+\end{minted}
+
+Now the \hs{Trace} requests the value of one specific \hs{k}, and by requesting it we can determine the next dependency.
+The inherent advantage of an \hs{Applicative} compute is that a single small request can be made to the server, with all the hashes of the dependencies, to get a final answer back.
+In contrast, a \hs{Monadic} cloud system either has to make multiple requests to the server, or pull down a potentially large \hs{Trace} tree.

### 2018-03-10 paper/background.tex
Finish S2

@@
 most once per build and only if they transitively depend on files that changed
-since the previous build.}
+since the previous build.}\label{def-minimal}
 \vspace{2mm}
@@
 current implementation of \Excel recomputes indirect references in every
-build~\cite{excel_recalc}. This clearly violates the minimality property: if
-\cmd{C1=1} and the user modifies \cmd{A2}, \Excel will recompute \cmd{B1},
-potentially triggering further unnecessary recalculation even though \cmd{B1}
-does not transitively depend on \cmd{A2}.
+build~\cite{excel_recalc}. This clearly violates the minimality
+property~\ref{def-minimal}: if \cmd{C1=1} and the user modifies \cmd{A2}, \Excel
+will recompute \cmd{B1}, potentially triggering further unnecessary
+recalculation even though \cmd{B1} does not transitively depend on \cmd{A2}.
 
@@
 Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
-systems track changes of inputs and intermediate results, executing dependent
-tasks whenever they change, but \Excel can also track changes in the tasks
-themselves: if a formula is modified, \Excel will recompute it and propagate
-the changes. Self-tracking is uncommon in software build systems, where one
-often needs to manually initiate a full rebuild even if just a single task has
-changed. We further discuss self-tracking in~\S\ref{sec-tracking-aspects}.
+systems only track changes of inputs and intermediate results, but \Excel can
+also track changes in the tasks themselves: if a formula is modified, \Excel
+will recompute it and propagate the changes. Self-tracking is uncommon in
+software build systems, where one often needs to manually initiate a full
+rebuild even if just a single task has changed. We further discuss
+self-tracking in~\S\ref{sec-tracking-aspects}.
+
+\begin{figure}[h]
+\vspace{-1mm}
+\begin{subfigure}[b]{0.90\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
+\caption{Dependency graph produced after the previous build.}
+\end{subfigure}
+\begin{subfigure}[b]{0.90\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
+\caption{Since \cmd{docs.txt} has changed, we rebuild \cmd{release.txt} and
+\cmd{release.tar}, discovering a new dependency.}
+\end{subfigure}
+\vspace{-1mm}
+\caption{Dynamic dependencies example: create \cmd{README} and add it to the
+list of release documents \cmd{docs.txt}.\label{fig-shake}}
+\vspace{-2mm}
+\end{figure}
 
@@
 
-Note that it is not always possible to know the dependencies of a task upfront,
-or \emph{statically}. To demonstrate this, let us add another task to the above
-makefile:
+\Shake was developed to solve the issue of dynamic
+dependencies~\cite{mitchell2012shake} and it excels at them without sacrificing
+the minimality requirement. To demonstrate this, we elaborate the \Make example
+from~\S\ref{sec-background-make} by adding the following files whose
+dependencies are shown in Fig.~\ref{fig-shake}(a):
 
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{makefile}
-release.tar: main.exe README
-    tar -cf release.tar main.exe README
-\end{minted}
-\vspace{1mm}
+\begin{itemize}
+    \item \cmd{LICENSE} is an input text file containing the project license.
+    \item \cmd{release.txt} is a text file listing all release files. This file
+    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt}
+    that list all binary and documentation files of the project.
+    \item \cmd{release.tar} is the release archive built by executing the
+    command \cmd{tar} on the release files.
+\end{itemize}
 
-\noindent
-So far so good: if either \cmd{main.exe} or \cmd{README} is dirty, \Make will
-recreate the release archive. But now imagine that we need to specify the files
-that go into \cmd{release.tar} in a text file \cmd{release.txt}. The dependencies
-of \cmd{release.tar} are no longer known statically: they \emph{depend on the
-content} of \cmd{release.txt}, which might not even exist before the build
-starts, e.g. if it is produced by concatenating input files \cmd{release-bins.txt}
-and \cmd{release-docs.txt}. Alas, makefiles cannot express such \emph{dynamic
-dependencies}, requiring ad hoc workarounds such as \emph{build phases}, which
-are problematic~\cite{hadrian}.
+The dependencies of \cmd{release.tar} are not known statically: they are
+determined by the content of \cmd{release.txt}, which might not even exist
+before the build. Makefiles cannot express such dependencies, requiring ad hoc
+workarounds such as \emph{build phases}, which are problematic~\cite{hadrian}.
+Here is how one can express the task for producing \cmd{release.tar} using \Shake.
 
-\Shake was developed specifically to address the issue of dynamic
-dependencies~\cite{mitchell2012shake} and it excels at handling them without
-sacrificing the minimality requirement. Here is how one can express the task
-for producing \cmd{release.tar} from the example discussed
-in~\S\ref{sec-background-make}.
-
-\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-\vspace{1mm}
 
@@
 command to produce the resulting archive. Crucially, the archive will only be
-rebuilt if one of the dependencies (static or dynamic) has changed.
-
-\begin{figure}[h]
-\begin{subfigure}[b]{0.90\linewidth}
-\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
-\vspace{-1mm}
-\caption{Dependency graph produced after the previous build.}
-\vspace{3mm}
-\end{subfigure}
-\begin{subfigure}[b]{0.90\linewidth}
-\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
-\vspace{-1mm}
-\caption{Since \cmd{docs.txt} is dirty, we rebuild \cmd{release.txt} and
-\cmd{release.tar}, discovering a new dependency.}
-\end{subfigure}
-\caption{Dynamic dependencies example: create \cmd{README} and add it to the
-list of release documents \cmd{docs.txt}.\label{fig-shake}}
-\end{figure}
+rebuilt if one of the dependencies (static or dynamic) has changed. For example,
+if we create another documentation file \cmd{README} and add it to
+\cmd{docs.txt}, \Shake will appropriately rebuild \cmd{release.txt} and
+\cmd{release.tar}, discovering the new dependency, as illustrated in
+Fig.~\ref{fig-shake}(b).
 
 \Shake's implementation is different from both \Make and \Excel in two aspects.
-First, it uses the dependency graph produced in the previous build to detect
-which files are dirty -- a powerful idea with a long history of use in
-\emph{incremental}~\cite{demers1981incremental},
+First, it uses the dependency from the previous build to detect which files need
+to be rebuilt. This is a powerful idea with a long history of
+use in \emph{incremental}~\cite{demers1981incremental},
 \emph{adaptive}~\cite{acar2002adaptive}, and
@@
 discussion in related work~\S\ref{sec-related}). Second, instead of abandoning
-and deferring the computation of tasks whose rediscovered dependencies are dirty
-(as \Excel does), \Shake \emph{pauses} their computation until the dependencies
-are brought up-to-date. We refer to this build algorithm as \emph{recursive}.
+and deferring the execution of tasks whose newly discovered dependencies have
+not yet been built (as \Excel does), \Shake \emph{pauses} their execution until
+the dependencies are brought up-to-date. We refer to this build algorithm as
+\emph{recursive}.
 
@@
     \item \Make stores file modification times, or rather, it relies on the file
-    system to do that. In principle a single dirty bit is sufficient, as will be
-    demonstrated in~\S\ref{sec-implementations}.
-    \item \Excel stores the dirty bit as well as the calculation chain.
+    system to do that. In principle a single \emph{dirty bit} per file is
+    sufficient, as will be demonstrated in~\S\ref{sec-implementations}.
+    \item \Excel stores a dirty bit per cell as well as the calculation chain.
     \item \Shake stores the dependency graph discovered in the previous build,
@@
 
-In this paper we show which build system properties are consequences of specific
-implementation choices (metadata and algorithm), and how one can obtain new
-build systems with desired properties by recombining parts of existing
-implementations. As a compelling example, we demonstrate how to combine
-advantages of \Shake and \Bazel in a cloud build system with dynamic dependencies.
-
 \begin{table}[h]
@@
 \centering
-\begin{tabular}{l||l|c|c|c||l|l}
+\begin{tabular}{l||l|l||l|c|c|c}
 \hline
-$\!\!$Build system & Dependencies & Minimal & Cutoff & Cloud & Metadata                       & Algorithm   $\!\!$\\\hline
-$\!\!$\Make        & Static       & Yes     & No     & No    & File modification times        & Topological $\!\!$\\
-$\!\!$\Excel       & Dynamic      & No      & No     & No    & Dirty cells, calculation chain & Chain, defer$\!\!$\\
-$\!\!$\Shake       & Dynamic      & Yes     & Yes    & No    & Last dependency graph          & Recursive   $\!\!$\\
-$\!\!$\Bazel       & Static       & Yes     & Yes    & Yes   & All dependency graphs, cache   & Topological $\!\!$\\\hline
+$\!$Build system$\!$& Metadata                       & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
+$\!$\Make       $\!$& File modification times        & Topological  & Static       & Yes     & No     & No   $\!$\\
+$\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
+$\!$\Shake      $\!$& Last dependency graph          & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
+$\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
 \hline
@@
 
-\subsection{Common vocabulary for build systems}
-\label{sec-background-vocabulary}
-
-In this section we introduce a common vocabulary for build systems that allows
-us to abstract away from specific application domains, such as software
-development or spreadsheets. We will use this vocabulary throughout the rest of
-the paper.
-
-\emph{Keys} are used to distinguish \emph{values}. In software build systems
-keys are typically filenames, e.g. \cmd{main.c}, whereas values are file
-contents (a C program source code in this case). In spreadsheets keys are cell
-names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
-displayed inside cells. In Haskell code, we will use type variables \hs{k}
-and \hs{v} to denote keys and values, respectively.
-
-We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
-tracking and sharing of build results, where \hs{Hash} is an abstract data type
-equipped with the equality test.
-
-A \emph{store} associates keys to values. It is convenient to assume that a store
-is total, i.e. it contains a value for every possible key. We therefore also
-assume that the type of values is capable of encoding values corresponding to
-non-existent keys (missing files or empty cells). This, in particular, suggests
-that it is possible to \emph{depend on the absence of a value} which is a useful
-feature for build systems. In addition to usual \hs{get} and \hs{put} operations,
-some stores support the \hs{getHash} operation implemented more efficiently than
-by hashing the result of \hs{get}. For example, GVFS (Git Virtual File
-System)~\cite{gvfs} stores file hashes and downloads file contents only on demand.
-
-Some values must be provided by the user as \emph{input}. For example,
-\cmd{main.c} can be edited by the user who relies on the build system to
-compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
-such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
-\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
-but are produced in the process of turning inputs into outputs.
+In this paper we show which build system properties are consequences of specific
+implementation choices (metadata and algorithm), and how one can obtain new
+build systems with desired properties by recombining parts of existing
+implementations. As a compelling example, we demonstrate how to combine
+advantages of \Shake and \Bazel in a cloud build system with dynamic dependencies.

### 2018-03-10 paper/compute.tex
Finish S2

@@
 type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
+type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
 \end{minted}
@@
 build systems in~\S\ref{sec-implementations}.
+
+\subsection{Common vocabulary for build systems}
+\label{sec-background-vocabulary}
+
+In this section we introduce a common vocabulary for build systems that allows
+us to abstract away from specific application domains, such as software
+development or spreadsheets. We will use this vocabulary throughout the rest of
+the paper.
+
+\emph{Keys} are used to distinguish \emph{values}. In software build systems
+keys are typically filenames, e.g. \cmd{main.c}, whereas values are file
+contents (a C program source code in this case). In spreadsheets keys are cell
+names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
+displayed inside cells. In Haskell code, we will use type variables \hs{k}
+and \hs{v} to denote keys and values, respectively.
+
+We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
+tracking and sharing of build results, where \hs{Hash} is an abstract data type
+equipped with the equality test.
+
+A \emph{store} associates keys to values. It is convenient to assume that a store
+is total, i.e. it contains a value for every possible key. We therefore also
+assume that the type of values is capable of encoding values corresponding to
+non-existent keys (missing files or empty cells). This, in particular, suggests
+that it is possible to \emph{depend on the absence of a value} which is a useful
+feature for build systems. In addition to usual \hs{get} and \hs{put} operations,
+some stores support the \hs{getHash} operation implemented more efficiently than
+by hashing the result of \hs{get}. For example, GVFS (Git Virtual File
+System)~\cite{gvfs} stores file hashes and downloads file contents only on demand.
+
+Some values must be provided by the user as \emph{input}. For example,
+\cmd{main.c} can be edited by the user who relies on the build system to
+compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
+such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
+\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
+but are produced in the process of turning inputs into outputs.
 

### 2018-03-10 paper/build.tex
Add TODOs for Neil

@@
 
-The build function takes 4 type arguments: \hs{c} is the constraint used for \hs{Compute} (typically either \hs{Applicative} or \hs{Monad}), \hs{i} is the information stored by the build system and \hs{k}/\hs{v} are the key/value types as described in \S\ref{todo}.
+The build function takes 4 type arguments: \hs{c} is the constraint used for \hs{Compute} (typically either \hs{Applicative} or \hs{Monad}), \hs{i} is the information stored by the build system and \hs{k}/\hs{v} are the key/value types as described in \S\ref{sec-vocabulary}.
 Given a build system \hs{build}:
@@
 
+\todo{NM}{We haven't used the term `target' before, perhaps change to `keys'?}
+
 We have underlined two different phrases in this sentence, and tackle each aspect separately.
@@
 If the \hs{Compute} function for \hs{k} calls the functional argument of \hs{d} then the build system is responsible for ensuring that \hs{d} is built \textit{before} \hs{k}.
-Looking at the build systems in \S\ref{todo} we have observed 3 distinct approaches to respecting dependency orderings.
+Looking at the build systems in \S\ref{sec-background} we have observed 3 distinct approaches to respecting dependency orderings.
 In this section of the paper we'll explore what properties are a consequence and how they operate.
@@
 However, to go from a \hs{Compute} to the dependencies, it must be the case that it is Applicative.
+
+\todo{NM}{In this case we can give actual code, so let's do that?}
 
@@
 
+\todo{NM}{I found terms `predicate' and `generative' hard to grasp and a bit abstract. Especially `predicate'.}
+
+\todo{NM}{I would also like to see an actual \hs{upToDate} function for predicate, and a \hs{recover}/\hs{fetch} or something like that for generative.}
+
 For \hs{Applicative} computations, the traces need not record the dependency keys (they are implicit given the target key), and can be indexed by the hashes of the keys.
@@
 In contrast, a \hs{Monadic} cloud system either has to make multiple requests to the server, or pull down a potentially large \hs{Trace} tree.
+
+\todo{NM}{This sounds very interesting, but I didn't get it after a quick read, and it's late. I think we need examples for all these cases, ideally reusing the examples from the background section.}
+
+\todo{NM}{I'm becoming a bit more optimistic about the term `trace'.
+It sounds conceptually simpler than a `dependency graph' even though it's isomorphic.
+So, let's keep it, with a note somewhere, pointing out that traces actually contain enough information to recover dependency graphs.}

### 2018-03-10 paper/implementations.tex
Add a few more comments

@@
 We can then provide functions like \hs{getStore} to load from the store and \hs{putStore} to write to the store.
+
+\todo{NM}{To me \hs{M} is a bit mysterious. Maybe show its internals?}
+
+\todo{NM}{What about using shorter \hs{get} and \hs{put}?
+I can introduce it right in the common vocabulary, mentioning the usual get-put laws.
+I would also remove `get' prefixes from many functions, e.g.
+\hs{dependencies compute} reads fine.
+This is a minor issue, happy to do this at the end.}
+
+\todo{NM}{The API still uses \hs{Map k v} and it looks this needs updating throughout the section, so I'll stop here.}
 

### 2018-03-10 paper/build.tex
Expand some TODOs

@@
 \todo{NM}{We haven't used the term `target' before, perhaps change to `keys'?}
+
+\todo{AM}{Turn the above into a formal definition of `correctness' and add a section that provides a mathematical definition of correctness, which does not rely on materialising intermediate values.}
 
@@
 
-\todo{NM}{I found terms `predicate' and `generative' hard to grasp and a bit abstract. Especially `predicate'.}
+\todo{NM}{I found terms `predicate' and `generative' hard to grasp and a bit abstract.
+Especially `predicate'.
+Let's replace it with `verifying'?
+And maybe `generative' with `recreating' since it allows to recreate values which were previously created by compute?}
 
-\todo{NM}{I would also like to see an actual \hs{upToDate} function for predicate, and a \hs{recover}/\hs{fetch} or something like that for generative.}
+\todo{NM}{I would also like to see an actual \hs{upToDate} function for verifying, and a \hs{recreate} for `recreating'.}
 

### 2018-03-10 paper/implementations.tex
Expand some TODOs

@@
 
-\todo{NM}{What about using shorter \hs{get} and \hs{put}?
+\todo{AM}{What about using shorter \hs{get} and \hs{put}?
 I can introduce it right in the common vocabulary, mentioning the usual get-put laws.
@@
 \hs{dependencies compute} reads fine.
-This is a minor issue, happy to do this at the end.}
+I suggest not to rush with this TODO -- let me first finish the Compute section, where many of these helper functions will be defined.}
 

### 2018-03-10 paper/background.tex
Polish S2

@@
 software build systems, where one often needs to manually initiate a full
-rebuild even if just a single task has changed. We further discuss
+rebuild even if just a single build task has changed. We further discuss
 self-tracking in~\S\ref{sec-tracking-aspects}.
@@
 \begin{figure}[h]
-\vspace{-1mm}
+\vspace{1mm}
 \begin{subfigure}[b]{0.90\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
+\vspace{-1mm}
 \caption{Dependency graph produced after the previous build.}
+\vspace{1mm}
 \end{subfigure}
+\vspace{1mm}
 \begin{subfigure}[b]{0.90\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
+\vspace{-1mm}
 \caption{Since \cmd{docs.txt} has changed, we rebuild \cmd{release.txt} and
@@
 \end{subfigure}
-\vspace{-1mm}
+\vspace{-2mm}
 \caption{Dynamic dependencies example: create \cmd{README} and add it to the
@@
 \Shake's implementation is different from both \Make and \Excel in two aspects.
-First, it uses the dependency from the previous build to detect which files need
-to be rebuilt. This is a powerful idea with a long history of
-use in \emph{incremental}~\cite{demers1981incremental},
+First, it uses the dependency graph from the previous build to decide which
+files need to be rebuilt. This idea has a long history, going back to
+\emph{incremental}~\cite{demers1981incremental},
 \emph{adaptive}~\cite{acar2002adaptive}, and
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.28]{fig/make-example-cutoff.pdf}}
+\centerline{\includegraphics[scale=0.28]{fig/shake-example-cutoff.pdf}}
 \vspace{-2mm}
-\caption{An early cutoff example: modify the source \cmd{main.c} by adding a
-comment. The rebuild is stopped after detecting that \cmd{main.o} is unchanged,
-which indicates that \cmd{main.exe} does not need to be rebuilt.\label{fig-cutoff}}
+\caption{An early cutoff example: if a comment is added to \cmd{main.c}, the
+rebuild is stopped after detecting that \cmd{main.o} is unchanged, since this
+indicates that \cmd{main.exe} and its dependents do not need to be
+rebuilt.\label{fig-cutoff}}
 \end{figure}
@@
 Like \Make, it does not support dynamic dependencies and can therefore benefit
-from the simplicity of building tasks in the statically known topological order.
+from the simplicity of building tasks in a statically known topological order.
 It is minimal and supports the early cutoff optimisation.
@@
 in~\S\ref{sec-cloud-aspects}.}. The latter allows to bypass the execution of
-a task, by predicting the hash of its result from the hashes of the dependencies,
-and subsequently fetch the result from the cache. A concrete implementation is
+a task, by predicting the hash of the result from the hashes of its dependencies,
+and subsequently fetching the result from the cache. A concrete implementation is
 provided in~\S\ref{sec-implementations}.
@@
     sufficient, as will be demonstrated in~\S\ref{sec-implementations}.
-    \item \Excel stores a dirty bit per cell as well as the calculation chain.
+    \item \Excel stores one dirty bit per cell and the calculation chain from
+    the previous build.
     \item \Shake stores the dependency graph discovered in the previous build,
@@
 $\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
-$\!$\Shake      $\!$& Last dependency graph          & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
+$\!$\Shake      $\!$& Previous dependency graph      & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
 $\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
@@
 
-In this paper we show which build system properties are consequences of specific
-implementation choices (metadata and algorithm), and how one can obtain new
-build systems with desired properties by recombining parts of existing
+In this paper we elucidate which build system properties are consequences of
+specific implementation choices (metadata and algorithm), and how one can obtain
+new build systems with desired properties by recombining parts of existing
 implementations. As a compelling example, we demonstrate how to combine
-advantages of \Shake and \Bazel in a cloud build system with dynamic dependencies.
+advantages of \Shake and \Bazel in a cloud build system with dynamic
+dependencies, see~\S\ref{sec-implementation-cloud-shake}.

### 2018-03-10 paper/build.tex
More todos

@@
 \begin{quote}
-A build system must build \underline{out-of-date targets} while \underline{respecting dependency orderings}.
+A build system must build \underline{out-of-date keys} while \underline{respecting dependency orderings}.
 \end{quote}
-
-\todo{NM}{We haven't used the term `target' before, perhaps change to `keys'?}
 

### 2018-03-10 paper/1-intro.tex
Rename the chapters so they are in order

--- 2018-03-08 paper/intro.tex
@@
 reasoned about, and varied.
+\item We show that all build system we look at can be broken along two axes - the order in which they build dependencies, and whether they decide to build a particular dependency.
+Interestingly, these two aspects can be recombined without restriction to produce new build systems.
 \item Two particularly desirable properties are dynamic dependencies
@@
 
-Papers about "frameworks" are often fuzzy.  This one is not: all our
+Papers about ``frameworks'' are often fuzzy.  This one is not: all our
 abstractions are defined in Haskell, and we have (freely-available)
@@
 in the right way, it certainly is a build system.
-
-\todo{AM/NM}{Add a paragraph on the decomposition of build systems into
-\emph{compute} and \emph{build} components, with the latter split into
-\emph{when} and \emph{how} parts.}

### 2018-03-10 paper/2-background.tex
Rename the chapters so they are in order

--- 2018-03-10 paper/background.tex
@@
 in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.
+
+\begin{quote}
+A build system must build \underline{out-of-date keys} while \underline{respecting dependency orderings}.
+\end{quote}
+
+\todo{AM}{Turn the above into a formal definition of `correctness' and add a section that provides a mathematical definition of correctness, which does not rely on materialising intermediate values.}
 

### 2018-03-10 paper/6-engineering.tex
Rename the chapters so they are in order

--- 2018-03-09 paper/engineering.tex
@@
 Or ``Build systems in the real world'', or ``Crazy things section''...
+
+\subsection{Partial stores, exceptions}
+
+Given a \hs{Build c i k v} defined for a total store, one can use it
+with a partial store \hs{k -> Maybe v}, obtaining a special case
+\hs{Build c i k (Maybe v)}, or with a store \hs{k -> Either e v} that can
+throw exceptions of type \hs{e}, obtaining a special case
+\hs{Build c i k (Either e v)}.
 
@@
 \todo{NM}{Add a paragraph on how to introduce concurrency to implementations
-in~\S\ref{sec-examples}}.
+in~\S\ref{sec-implementations}}.
 

### 2018-03-10 paper/8-conclusions.tex
Rename the chapters so they are in order

\clearpage
\section{Conclusions}\label{sec-conclusions}

...

### 2018-03-10 paper/4-build.tex
Clean up build a bit

--- 2018-03-10 paper/build.tex
@@
 
-\todo{AM}{Turn the above into a formal definition of `correctness' and add a section that provides a mathematical definition of correctness, which does not rely on materialising intermediate values.}
-
 We have underlined two different phrases in this sentence, and tackle each aspect separately.
@@
 Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort.
-However, to go from a \hs{Compute} to the dependencies, it must be the case that it is Applicative.
-
-\todo{NM}{In this case we can give actual code, so let's do that?}
+However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{getdependencies}.
 
@@
 
-\subsubsection{Recording predicate trace}
+\subsubsection{Verifying traces}
 
-An alternative way to determine if a key is dirty is to record what state the values/hashes of dependencies were used at last time, and if something has changed, the key is dirty and must be recomputed.
+An alternative way to determine if a key is dirty is to record what state the values/hashes of dependencies were used at last time, and if something has changed, the key is dirty and must be recomputed - in essence a trace with which we can use to \textit{verify} existing values.
 For a monadic build system the trace can include what dependencies were used as well, although for an applicative system that information can be inferred.
@@
 
-\subsubsection{Recording generative traces}
+\subsubsection{Constructive traces}
 
-A predicate trace allows us to mark the key dirty and rebuild it, extending that we can store a \textit{generative} trace which is the trace plus the complete result.
-Once we are storing the complete result it makes sense to record many generative traces per key, and to share them with other users, providing cloud-build functionality.
+A verifying trace allows us to mark the key dirty and rebuild it, extending that we can store a \textit{constructive} trace which is the trace plus the complete result.
+Once we are storing the complete result it makes sense to record many constructive traces per key, and to share them with other users, providing cloud-build functionality.
 
-\todo{NM}{I found terms `predicate' and `generative' hard to grasp and a bit abstract.
-Especially `predicate'.
-Let's replace it with `verifying'?
-And maybe `generative' with `recreating' since it allows to recreate values which were previously created by compute?}
-
-\todo{NM}{I would also like to see an actual \hs{upToDate} function for verifying, and a \hs{recreate} for `recreating'.}
+\todo{AM}{AM: I would also like to see an actual \hs{upToDate} function for verifying, and a \hs{recreate} for `recreating'. NM: }
 

### 2018-03-10 src/Neil/Builder.hs
Use consistent variable names throughout

@@ -1,4 +1,4 @@
-{-# LANGUAGE Rank2Types, FlexibleContexts, GeneralizedNewtypeDeriving, ScopedTypeVariables, RecordWildCards, ConstraintKinds, TypeFamilies #-}
+{-# LANGUAGE Rank2Types, FlexibleContexts, GeneralizedNewtypeDeriving, ScopedTypeVariables, RecordWildCards, ConstraintKinds, TypeFamilies, LambdaCase #-}
 
 module Neil.Builder(
     dumb,
@@ -54,7 +54,7 @@ recursive step compute = runM . ensure
             let ask x = ensure x >> getStore x
             Recursive done <- getTemp
             when (k `Set.notMember` done) $ do
-                modifyTemp $ \(Recursive set) -> Recursive $ Set.insert k set
+                modifyTemp $ \(Recursive done) -> Recursive $ Set.insert k done
                 case trackDependencies compute ask k of
                     Nothing -> return ()
                     Just act -> step k (getDependenciesMaybe compute k) ask act
@@ -74,8 +74,7 @@ reordering step compute k = runM $ do
             case failDependencies compute f' k of
                 Nothing -> (k :) <$> f (Set.insert k done) ks
                 Just act -> do
-                    res <- step k (getDependenciesMaybe compute k) (fmap eitherToMaybe . f') act
-                    case res of
+                    step k (getDependenciesMaybe compute k) (fmap eitherToMaybe . f') act >>= \case
                         Nothing -> (k :) <$> f (Set.insert k done) ks
                         Just e -> f done $ [e | e `notElem` ks] ++ ks ++ [k]
 
@@ -88,8 +87,8 @@ data Trace k v = Trace k [(k, Hash v)] (Hash v) deriving Show
 -- | Given a way to ask for a dependency, and the key under construction, and the traces, pick the hashes that match
 traces :: (Monad m, Eq k) => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
 traces check k ts = mapMaybeM f ts
-    where f (Trace k2 ds v) = do
-                b <- return (k == k2) &&^ allM (uncurry check) ds
+    where f (Trace k2 dkv v) = do
+                b <- return (k == k2) &&^ allM (uncurry check) dkv
                 return $ if b then Just v else Nothing
 
 
[…]

### 2018-03-10 paper/4-build.tex
More changes

@@
 
-In this section we further describe our \hs{Build} abstraction, defined as:
+In this section we describe our \hs{Build} abstraction, defined as:
 
@@
 \subsection{Respecting dependency orderings}
+\label{sec-dependency-orderings}
 
@@
 An alternative way to determine if a key is dirty is to record what state the values/hashes of dependencies were used at last time, and if something has changed, the key is dirty and must be recomputed - in essence a trace with which we can use to \textit{verify} existing values.
-For a monadic build system the trace can include what dependencies were used as well, although for an applicative system that information can be inferred.
-For each key, a single trace is recorded, so it naturally fits that the data is stored in an associative map indexed by target key.
+We can describe a trace as:
 
-The recording of a trace requires storing extra information with the execution, but otherwise provides a simple approach which is capable of achieving all the properties without imposing restrictions.
+\begin{minted}{haskell}
+data Trace k v = Trace
+    {key :: k
+    ,result :: Hash v
+    ,dependencies :: [(k, Hash v)]
+    }
+\end{minted}
+
+We assume that \hs{Hash v} is a small constant size, constructed from hashing the underlying \hs{v} rather than storing it directly.
+Checking a trace requires ensuring all the dependencies are up to date (using whatever ordering strategy as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are same as the current value and that the result is the same.
+After computing a fresh value we add its \hs{Trace} to the list.
+Therefore the information stored by a build system that verifies traces is \hs{[Trace k v]}.
 
@@
 
-A verifying trace allows us to mark the key dirty and rebuild it, extending that we can store a \textit{constructive} trace which is the trace plus the complete result.
+A verifying trace allows us to mark a key dirty and rebuild it.
+Extending that information we can store a \textit{constructive} trace which is the trace plus the actual result.
 Once we are storing the complete result it makes sense to record many constructive traces per key, and to share them with other users, providing cloud-build functionality.
-
-\todo{AM}{AM: I would also like to see an actual \hs{upToDate} function for verifying, and a \hs{recreate} for `recreating'. NM: }
-
-For \hs{Applicative} computations, the traces need not record the dependency keys (they are implicit given the target key), and can be indexed by the hashes of the keys.
-Concretely, we could imagine storing:
+We can represent that as:
 
 \begin{minted}{haskell}
-data GenerativeApplicativeTrace k v = GAT
-    { traces :: Map (k, [Hash v]) (Hash v)
-    , contents :: Map (Hash v) v
+data Traces k v = Traces
+    {traces :: [Trace k v]
+    ,contents :: Map (Hash v) v
     }
@@
 
-Given a target key, we can grab the dependencies, grab their hashes and look up the associated result hash.
-If it doesn't match what we have on disk, we download it from the \hs{contents}.
-If the entry is not in \hs{traces} we can run it and add it.
+We have a list of traces, plus a \hs{Map} from the hash to the actual contents.
+Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{contents} \textit{without} recomputing it.
 
-For \hs{Monadic} computations the situation is slightly more tricky - instead of being able to determine the dependencies in advance, we effectively have a set of traces for any given key, any of which could apply.
-While no existing build system actually implements monadic cloud builds, we can describe a plausible structure as:
+For both verifying and constructive choices it is possible to choose smarter data structures than simply a list of \hs{Trace}, which we discuss in \S\ref{sec-smart-traces}.
 
-\begin{minted}{haskell}
-data Trace k v = Trace [(k, Hash v)] (Hash v)
-type GenerativeMonadicTrace k v = GMT
-    { traces :: Map k [Trace k v]
-    , contents :: Map (Hash v) v
-    }
-\end{minted}
-
-Now we can grab the list of traces for a given key, and see which (if any) matches our current state.
-In practice, to avoid an $O(n)$ scan of the number of traces, we can common up prefixes and use the data structure:
-
-\begin{minted}{haskell}
-data Trace k v = Trace k (Map (Hash v) (Trace k v))
-               | Done (Hash v)
-\end{minted}
-
-Now the \hs{Trace} requests the value of one specific \hs{k}, and by requesting it we can determine the next dependency.
-The inherent advantage of an \hs{Applicative} compute is that a single small request can be made to the server, with all the hashes of the dependencies, to get a final answer back.
-In contrast, a \hs{Monadic} cloud system either has to make multiple requests to the server, or pull down a potentially large \hs{Trace} tree.
-
-\todo{NM}{This sounds very interesting, but I didn't get it after a quick read, and it's late. I think we need examples for all these cases, ideally reusing the examples from the background section.}
-
-\todo{NM}{I'm becoming a bit more optimistic about the term `trace'.
-It sounds conceptually simpler than a `dependency graph' even though it's isomorphic.
-So, let's keep it, with a note somewhere, pointing out that traces actually contain enough information to recover dependency graphs.}
+\todo{AM}{AM: I would also like to see an actual \hs{upToDate} function for verifying, and a \hs{recreate} for `recreating'.
+NM: It actually gets a bit more confusing than that - since you end up needing to pass functions to get yourself, your dependencies etc.
+Unclear there is a win.
+The functions would occur in the next section anyway.}

### 2018-03-10 paper/5-implementations.tex
More changes

--- 2018-03-10 paper/implementations.tex
@@
 We also technically allow multiple traces to be in the store at the same time, although that is merely something that falls out for free - we would have to do work to avoid it.
+
+\subsection{Smart Trace structures}
+\label{sec-smart-traces}
+
+For \hs{Applicative} computations, the traces need not record the dependency keys (they are implicit given the target key), and can be indexed by the hashes of the keys.
+Concretely, we could imagine storing:
+
+\begin{minted}{haskell}
+data GenerativeApplicativeTrace k v = GAT
+    { traces :: Map (k, [Hash v]) (Hash v)
+    , contents :: Map (Hash v) v
+    }
+\end{minted}
+
+Given a target key, we can grab the dependencies, grab their hashes and look up the associated result hash.
+If it doesn't match what we have on disk, we download it from the \hs{contents}.
+If the entry is not in \hs{traces} we can run it and add it.
+
+For \hs{Monadic} computations the situation is slightly more tricky - instead of being able to determine the dependencies in advance, we effectively have a set of traces for any given key, any of which could apply.
+While no existing build system actually implements monadic cloud builds, we can describe a plausible structure as:
+
+\begin{minted}{haskell}
+data Trace k v = Trace [(k, Hash v)] (Hash v)
+type GenerativeMonadicTrace k v = GMT
+    { traces :: Map k [Trace k v]
+    , contents :: Map (Hash v) v
+    }
+\end{minted}
+
+Now we can grab the list of traces for a given key, and see which (if any) matches our current state.
+In practice, to avoid an $O(n)$ scan of the number of traces, we can common up prefixes and use the data structure:
+
+\begin{minted}{haskell}
+data Trace k v = Trace k (Map (Hash v) (Trace k v))
+               | Done (Hash v)
+\end{minted}
+
+Now the \hs{Trace} requests the value of one specific \hs{k}, and by requesting it we can determine the next dependency.
+The inherent advantage of an \hs{Applicative} compute is that a single small request can be made to the server, with all the hashes of the dependencies, to get a final answer back.
+In contrast, a \hs{Monadic} cloud system either has to make multiple requests to the server, or pull down a potentially large \hs{Trace} tree.
+
+\todo{NM}{This sounds very interesting, but I didn't get it after a quick read, and it's late. I think we need examples for all these cases, ideally reusing the examples from the background section.}
+
+ For a monadic build system the trace can include what dependencies were used as well, although for an applicative system that information can be inferred. In generality, a trace can be described as:
+
+For each key, a single trace is recorded, so it naturally fits that the data is stored in an associative map indexed by target key.
+
+The recording of a trace requires storing extra information with the execution, but otherwise provides a simple approach which is capable of achieving all the properties without imposing restrictions.

### 2018-03-10 src/Neil/Builder.hs

@@ -196,10 +196,6 @@ data TraceContent k v = TraceContent
     ,tcContent :: Map.Map (Hash v) v
     } deriving Show
 
-{-
-addTrace :: k -> [k] -> M i k v -> M i k v (TracedContent k v -> TracedContent k v)
-addTrace k dk 
--}
 
 instance Default (TraceContent k v) where def = TraceContent def def
 

### 2018-03-11 paper/2-background.tex
Add an intuitive definition of a build system

--- 2018-03-10 paper/background.tex
@@
 in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.
+
+We start by giving an intuitive definition of a build system. After providing
+more context and examples in the rest of this section, we will be able to
+develop a formal definition in~\S\ref{sec-general-build}.
+
+\definition[Build system]{Given a description of tasks, a \emph{build system}
+executes some of them in an order that respects their dependencies to produce a
+specified output from supplied inputs.}
+\label{def-build}
 
@@
 
-If the user modifies the sources of the utility library and runs \Make, it will
+If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
+will first build \cmd{util.o} and \cmd{main.o}, in any order since the
+corresponding tasks are independent, and will then build \cmd{main.exe}. If the
+user modifies the sources of the utility library and runs \Make again, it will
 perform a \emph{full rebuild}, because all three tasks transitively depend on
@@
 \begin{figure}[h]
-\vspace{1mm}
 \begin{subfigure}[b]{0.90\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
-\vspace{-1mm}
 \caption{Dependency graph produced after the previous build.}
-\vspace{1mm}
 \end{subfigure}
-\vspace{1mm}
 \begin{subfigure}[b]{0.90\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
-\vspace{-1mm}
 \caption{Since \cmd{docs.txt} has changed, we rebuild \cmd{release.txt} and
@@
 \end{subfigure}
-\vspace{-2mm}
 \caption{Dynamic dependencies example: create \cmd{README} and add it to the
 list of release documents \cmd{docs.txt}.\label{fig-shake}}
-\vspace{-2mm}
 \end{figure}
@@
 
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}

### 2018-03-11 paper/3-compute.tex
Clean up build a bit
Add an intuitive definition of a build system

--- 2018-03-10 paper/compute.tex
@@
 \subsection{Common vocabulary for build systems}
-\label{sec-background-vocabulary}
+\label{sec-vocabulary}
 
@@
 
+\label{getdependencies} We can define the dependencies for a \hs{Compute Applicative} using the \hs{Const} functor:
+
+\begin{minted}{haskell}
+getDependencies :: Compute Applicative k v -> k -> [k]
+getDependencies comp = maybe [] getConst . comp (\k -> Const [k])
+\end{minted}
+
 \subsection{Monadic compute}\label{sec-compute-monad}
@@
 ...
+
+\todo{AM}{Turn the above into a formal definition of `correctness' and add a section that provides a mathematical definition of correctness, which does not rely on materialising intermediate values.}

### 2018-03-11 src/Neil/Build.hs
Remove unused functions on M

@@ -6,8 +6,8 @@ module Neil.Build(
     getStore, putStore,
     Changed, getChanged, withChangedApplicative, withChangedMonad,
     Time, getStoreTime,
-    getInfo, putInfo, modifyInfo,
-    getTemp, putTemp, modifyTemp,
+    getInfo, modifyInfo,
+    getTemp, modifyTemp,
     Hash, getHash, Hashable, getStoreHash,
     ) where
 

### 2018-03-11 src/Neil/Builder.hs
Changes to implementations to more closely follow the paper

@@ -28,13 +28,18 @@ import Data.Either.Extra
 import Debug.Trace
 import Data.List
 import Data.Typeable
-import qualified Data.Set as Set
-import qualified Data.Map as Map
+import Data.Map(Map)
+import Data.Set(Set)
+import qualified Data.Set as Set hiding (Set)
+import qualified Data.Map as Map hiding (Map)
 
 ---------------------------------------------------------------------
 -- DEPENDENCY ORDER SCHEMES
 
-topological :: Default i => (k -> [k] -> M i k v v -> M i k v ()) -> Build Applicative i k v
+topological
+    :: (m ~ M i k v, Default i)
+    => (k -> [k] -> m v -> m ())
+    -> Build Applicative i k v
 topological step compute k = runM $ do
     let depends = getDependencies compute
     forM_ (topSort depends $ transitiveClosure depends k) $ \k ->
@@ -43,11 +48,14 @@ topological step compute k = runM $ do
             Just act -> step k (depends k) act
 
 
-newtype Recursive k = Recursive (Set.Set k)
+newtype Recursive k = Recursive (Set k)
     deriving Default
 
 -- | Build a rule at most once in a single execution
-recursive :: Default i => (k -> Maybe [k] -> (k -> M i k v v) -> M i k v ([k], v) -> M i k v ()) -> Build Monad i k v
+recursive
+    :: (m ~ M i k v, Default i)
+    => (k -> Maybe [k] -> (k -> m v) -> m ([k], v) -> m ())
+    -> Build Monad i k v
 recursive step compute = runM . ensure
     where
         ensure k = do
@@ -62,7 +70,8 @@ recursive step compute = runM . ensure
 
 reordering
     :: (m ~ M (i, [k]) k v, Default i)
-    => (k -> Maybe [k] -> (k -> m (Maybe v)) -> m (Either k ([k], v)) -> m (Maybe k)) -> Build Monad (i, [k]) k v
+    => (k -> Maybe [k] -> (k -> m (Maybe v)) -> m (Either k ([k], v)) -> m (Maybe k))
+    -> Build Monad (i, [k]) k v
 reordering step compute k = runM $ do
     order <- snd <$> getInfo
     order <- f Set.empty $ order ++ [k | k `notElem` order]
@@ -82,11 +91,22 @@ reordering step compute k = runM $ do
 ---------------------------------------------------------------------
 -- DIRTY SCHEMES
 
-data Trace k v = Trace k [(k, Hash v)] (Hash v) deriving Show
[…]

### 2018-03-11 paper/3-compute.tex
Add some pieces for Andrey to include
Work on S3

@@
 \end{minted}
+\vspace{1mm}
 
 \noindent
-Compute represents build rules (in build systems) or formulas (in spreadsheets),
-and is completely isolated from the world of compilers, file systems, memories,
+\hs{Compute} represents an abstract set of \emph{build tasks}, for example, build
+rules in software build systems or formulas in spreadsheets. It is completely
+isolated from the world of compilers, file systems, calculation chains, memories,
 caches, and all other complexities of real build systems. We explain the compute
-type in~\S\ref{sec-general-compute}, and then discuss several noteworthy special cases:
-functorial (\S\ref{sec-compute-functor}), applicative
+type in~\S\ref{sec-general-compute}, and then discuss several noteworthy special
+cases: functorial (\S\ref{sec-compute-functor}), applicative
 (\S\ref{sec-compute-applicative}) and monadic (\S\ref{sec-compute-monad}) types
@@
 
-Build corresponds to the algorithm used for bringing a key-value store to a
-consistent state by recomputing dirty values in an appropriate order. We explain
-the build type in~\S\ref{sec-build}, and then provide implementations for several
-build systems in~\S\ref{sec-implementations}.
+\hs{Build} corresponds to the algorithm used for bringing a key-value store
+\store to a consistent state by recomputing out-of-date values in an
+appropriate order. We explain the build type and use it to develop a formal
+definition of build systems in~\S\ref{sec-general-build}.
+Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} further deconstruct
+build algorithms and provide concrete implementations for several build systems.
 
@@
 
-In this section we introduce a common vocabulary for build systems that allows
+We start by introducing a common vocabulary for build systems that allows
 us to abstract away from specific application domains, such as software
@@
 
-\emph{Keys} are used to distinguish \emph{values}. In software build systems
-keys are typically filenames, e.g. \cmd{main.c}, whereas values are file
-contents (a C program source code in this case). In spreadsheets keys are cell
+\emph{Keys} are used to distinguish \emph{values}. In software build systems,
+keys are typically filenames, e.g.~\cmd{main.c}, whereas values are file
+contents (a C program source code in this case). In spreadsheets, keys are cell
 names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
@@
 We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
-tracking and sharing of build results, where \hs{Hash} is an abstract data type
+tracking and sharing of~build results, where \hs{Hash} is an abstract data type
 equipped with the equality test.
 
-A \emph{store} associates keys to values. It is convenient to assume that a store
-is total, i.e. it contains a value for every possible key. We therefore also
-assume that the type of values is capable of encoding values corresponding to
-non-existent keys (missing files or empty cells). This, in particular, suggests
-that it is possible to \emph{depend on the absence of a value} which is a useful
-feature for build systems. In addition to usual \hs{get} and \hs{put} operations,
-some stores support the \hs{getHash} operation implemented more efficiently than
-by hashing the result of \hs{get}. For example, GVFS (Git Virtual File
-System)~\cite{gvfs} stores file hashes and downloads file contents only on demand.
+A \emph{store} associates keys to values. It is convenient to assume that a
+store is total function \store, i.e. it contains a value for every possible key.
+We therefore also assume that the type of values is capable of encoding values
+corresponding to non-existent keys (missing files or empty cells). This, in
+particular, suggests that it is possible to \emph{depend on the absence of a
+value}, which is a useful feature for build systems. In addition to usual
+\hs{get} and \hs{put} operations, some stores support the \hs{getHash} operation
+implemented more efficiently than by hashing the result of \hs{get}. For example,
+GVFS (Git Virtual File System)~\cite{gvfs} stores file hashes and downloads file
+contents only on demand -- an important feature for cloud build systems.
 

### 2018-03-11 paper/4-build.tex

@@
 The \hs{info} parameter will be \hs{Nothing} on the very first call to \hs{build}.
+We assume that \hs{store} is total, where "missing files" can be represented by some sentinel value of \hs{v} if necessary.
 
@@
 \subsubsection{A dirty bit}
+\label{sec-dirty-bit}
 
@@
 
-\subsubsection{Verifying traces}
+\subsubsection{Verifying traces}\label{sec-verifying-traces}
 
@@
     {key :: k
+    ,dependencies :: [(k, Hash v)]
     ,result :: Hash v
-    ,dependencies :: [(k, Hash v)]
     }
@@
 
-\subsubsection{Constructive traces}
+\subsubsection{Constructive traces}\label{sec-constructive-traces}
 

### 2018-03-11 paper/5-implementations.tex

--- 2018-03-10 paper/implementations.tex
@@
 
-In the previous sections we've discussed the types of build system in text, but the way we came to our conclusions was by developing a framework to fit the builds in.
-In this section we explain the framework, some of the abstractions we introduced (which model the properties from \S\ref{sec-background}), and how we can make a meaningful implementation of the build systems discussed in \S\ref{sec-background}.
+In the previous sections we discussed the types of build system, and how they broken down.
+But these divisions were not obvious to us, and only by concretely implementing and refactoring each build system did we determine the underlying commonalities.
+In this section we share some of the code that got us there, but full code is available in the supporting material and at \url{url-hidden-for-anonymous-submission}.
+Our code comprises a framework, utility functions implementing the various dependency orderings as per \S\ref{sec-dependency-orderings}, and actual implementations.
 
@@
 
-As we saw in the introduction, a build system can be defined as:
+As we saw in \S\ref{sec-build}, a build system can be defined as:
 
 \begin{minted}{haskell}
-type Build c i k v = Compute c k v -> k -> Maybe i -> Map.Map k v -> (i, Map.Map k v)
+type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
 \end{minted}
 
-Using the framework we have built we can express the simplest and stupidest build system as:
+Using the framework we build in this section, we can express a dumb build system as:
 
@@
 dumb compute k = runM (f k)
-    where f k = maybe (getStore k) (\act -> do v <- act; putStore k v; return v) $ compute f k
+    where
+        f k = case compute f k of
+            Nothing -> getStore k
+            Just act -> do
+                v <- act
+                putStore k v
+                return v
 \end{minted}
+
+The \hs{dumb} build system can build with monadic computes, using a function \hs{f} which given a key returns the associated value.
+To do so, \hs{f} calls \hs{compute} passing \hs{f} itself as an argument.
+If the result is \hs{Nothing} (this key is an input) it simply reads from the store.
+Otherwise it runs the action to produce a value, stores it and returns it.
 
@@
 \begin{minted}{haskell}
-data M i k v r
-runM :: Default i => M i k v a -> Maybe i -> Map.Map k v -> (i, Map.Map k v)
+type M i k v r = State (i, k -> v, ...) r
+runM :: Default i => M i k v a -> Maybe i -> (k -> v) -> (i, k -> v)
+
 getStore :: k -> M i k v v
 putStore :: k -> v -> M i k v ()
+getInfo :: M i k v i
+modifyInfo :: (i -> i) -> M i k v ()
 \end{minted}
-Note: we have omitted the context on \hs{k}, things like \hs{Ord} (so the implementation can use a finite map) or \hs{Show} (so the implementation can produce nice error messages).
+Note: we have omitted the context on \hs{k}, things like \hs{Eq} (to implement the store).
 \caption{API with which to implement build systems}
@@
 
-The types of important definitions are given in Figure \ref{fig-M-api}.
-In essence we take the pure \hs{Map} to \hs{Map} interface of \hs{Build} and instead provide a Monad \hs{M} which keeps track of the \hs{i} info and \hs{Map k v} as a store.
-We can then provide functions like \hs{getStore} to load from the store and \hs{putStore} to write to the store.
+We use \hs{runM} to switch between the pure world of \hs{Build} and a state monad \hs{M} which lets us more precisely express the ideas behind the build system without threading the state manually - the types of important operations are given in Figure~\ref{fig-M-api}.
+We provide functions like \hs{getStore} to load from the store and \hs{putStore} to write to the store.
+We'll introduce the other functions in in the monad \hs{M} as we need them while exploring the build systems.
 
-\todo{NM}{To me \hs{M} is a bit mysterious. Maybe show its internals?}
-
-\todo{NM}{What about using shorter \hs{get} and \hs{put}?
+\todo{AM}{AM: What about using shorter \hs{get} and \hs{put}?
 I can introduce it right in the common vocabulary, mentioning the usual get-put laws.
@@
 \hs{dependencies compute} reads fine.
-This is a minor issue, happy to do this at the end.}
-
-\todo{NM}{The API still uses \hs{Map k v} and it looks this needs updating throughout the section, so I'll stop here.}
-
-With that the \hs{dumb} build from above should start to make sense.
-The function \hs{dumb} takes \hs{compute} and \hs{k}, then calls \hs{runM} on \hs{f x}, where \hs{f x} has type \hs{M () k v v}.
-The \hs{runM} function takes care of putting the information into a state monad, running the computation, then extracting it afterwards - to match the signature of \hs{build}.
-
-We'll introduce the other functions in in the monad \hs{M} as we need them while exploring the build systems.
+I suggest not to rush with this TODO -- let me first finish the Compute section, where many of these helper functions will be defined.
+NM: I think this is a bad idea - conflicts with a state Monad which is what we're using underneath so will confuse some readers.}
 
@@
 The \Make build system builds files in a linear order based on a topological sort.
-For each file, it builds it if the output file does not exist or is older than any of the inputs.
+For each file, it builds it if the output file is older than any of the inputs.
 We can express this in our framework with:
@@
 make :: Eq v => Build Applicative (Changed k v, ()) k v
-make = withChangedApplicative $ topological $ \k ds act -> do
-    kt <- getStoreTimeMaybe k
-    ds <- mapM getStoreTime ds
-    case kt of
-        Just xt | all (<= xt) ds -> return ()
-        _ -> putStore k =<< act
+make = withChanged $ topological $ \k dk act -> do
+    t <- getStoreTime k
+    dt <- mapM getStoreTime dk
+    let dirty = any (>= t) dt
+    when dirty $
+        putStore k =<< act
 \end{minted}
 
-We'll come back to \hs{withChangedApplicative} afterwards, but ignoring that, reading the rest of the implementation you can see the underying essence of \Make.
-We use the helper function \hs{topological} which gets called with each file in a dependency-respecting order given the file (\hs{k}), the dependencies (\hs{ds}) and an action to generate the result (\hs{act}).
-With those three things in hand, the implementation first tries to get the timestamp of the target, then all the dependencies (which must have been rebuilt if necessary, thanks to the guarantees on \hs{topological}), and if everything is fine does nothing, otherwise rebuilds and stores the result.
+We'll come back to \hs{withChanged} afterwards, but ignoring that, reading the rest of the implementation you can see the underying essence of \Make.
+We use the helper function \hs{topological} which gets called with each file in a dependency-respecting order given the file (\hs{k}), the dependencies (\hs{dk}) and an action to generate the result (\hs{act}).
+With those three things in hand, the implementation first gets the timestamp of the key being built, then the timestamps of the dependencies (which must have been rebuilt if necessary, thanks to the guarantees on \hs{topological}), and considers it dirty if any dependency was built after \hs{k}.
+If it is dirty, it rebuilds and stores the result.
 
@@
 \begin{minted}{haskell}
-topological :: Default i => (k -> [k] -> M i k v v -> M i k v ()) -> Build Applicative i k v
+topological
+    :: (m ~ M i k v, Default i)
+    => (k -> [k] -> m v -> m ())
+    -> Build Applicative i k v
 topological step compute k = runM $ do
@@
 
-The ``nasty'' element of this implementation is \hs{getStoreTime}, which is supported with a few somewhat abstract functions on \hs{M}:
+The greatest impedence mismatch between our framework and this implementation is the notion of \hs{Time} in \hs{getStoreTime}, which is missing from \hs{Build}.
+However, as we discussed in \S\ref{sec-dirty-bit}, \Make is really using modification-time to encode a dirty bit with specific properties.
+We model the dirty bit and time with the following API on \hs{M}:
 
@@
 getChanged :: Eq v => k -> M (Changed k v, i) k v Bool
-withChangedApplicative :: Build Applicative (Changed k v, i) k v -> Build Applicative (Changed k v, i) k v
+withChanged :: Build c (Changed k v, i) k v -> Build c (Changed k v, i) k v
 
 data Time = LastBuild | AfterLastBuild
-getStoreTimeMaybe :: Eq v => k -> M (Changed k v, i) k v (Maybe Time)
 getStoreTime :: Eq v => k -> M (Changed k v, i) k v Time
@@
 
-Recall that our \hs{Build} operation only gets a \hs{Map} - not a set of timestamps at which the build was updated.
-To model a timed store on top the build system implementation needs to store as info (the \hs{i}) field the old file store at the end, and then it can consider anything that has changed to be dirty.
-With this implementation \hs{getChanged k} will return \hs{True} if either the value associated with \hs{k} changed between the last execution and the start of this execution (tested for using the \hs{Eq} class), \textit{or} if \hs{putStore} has been called on it in this run.
-That essentially encodes the dirty bit required for Excel.
-We use \hs{withChangedApplicative} to do the tedious step of capturing the state at the end of the build and storing it as the info.
+To model a dirty bit we extend the \hs{i} information with \hs{Changed} to capture the \hs{k -> v} store at the end of the previous run, and extend \hs{M} to record which files have been written to during this \hs{runM}.
+A file is considered dirty if either the current value as per \hs{getStore} has changed since the last run, or if it has had \hs{putStore} called on it\footnote{It would be reasonable to declare files unchanged if they have \hs{putStore} called at their previous value, but \Excel does not and \Make cannot using modification times, so we ignore this option.}.
+To track the values at the last run we use \hs{withChanged} to copy the resultant \hs{k -> v} store into the info after the build completes\footnote{For tedious reasons our implementation requires \hs{withChangedApplicative} and \hs{withChangedMonad} which are variants of \hs{withChanged} for concrete constraints \hs{c} - but with identical implementations.}
 
@@
 That domain is sufficient for the \Make system.
-At first the fact that was sufficient surprised us, but it eventually became clear that \Make is using modification times as a dirty bit!
-Moreover, that explains why that a \hs{putStore} of an equal value doesn't avoid setting the changed bit - it can't because modification times can't encode that.
-Now we can give an equivalent but simpler formulation:
+At first the fact that this domain is sufficient surprised us, until we realised \Make is only interested in the dirty bit anyway!
+If we strip back the modification time and directly implement \Make in terms of \hs{getChanged} we obtain:
 
@@
 makeDirtyBit :: Eq v => Build Applicative (Changed k v, ()) k v
-makeDirtyBit = withChangedApplicative $ topological $ \k ds act -> do
-    dirty <- fmap isNothing (getStoreMaybe k) ||^ anyM getChanged ds
+makeDirtyBit = withChanged $ topological $ \k dk act -> do
+    dirty <- getChanged k ||^ anyM getChanged dk
     when dirty $
@@
 
-Here we directly test for changes, which is simpler.
-
 \subsection{\Excel}\label{sec-implementation-excel}
 
-The spreadsheet is actually one of the most challenging to implement, and one of the most challenging to into pieces. The code is:
+As a programming exercise, \Excel is the most challenging, due to the complicated reordering dependency scheme, which requires retrying out-of-order operations. The complete implementation is:
 
 \begin{minted}{haskell}
-spreadsheet :: Eq v => Build Monad (Changed k v, [k]) k v
-spreadsheet = withChangedMonad $ dynamic $ \k ds _ act -> do
-    dirty <- fmap isNothing (getStoreMaybe k) ||^ getChanged k ||^ maybe (return True) (anyM getChanged) ds
+excel :: Eq v => Build Monad (Changed k v, [k]) k v
+excel = withChanged $ reordering $ \k dk _ act -> do
+    dirty <- getChanged k ||^ maybe (return True) (anyM getChanged) dk
     if not dirty then
         return Nothing
-    else do
-        res <- act
-        case res of
+    else
+        act >>= \case
             Left e -> return $ Just e
@@
 
-We use basically the same change computation to test for dirtiness, but things are complicated because when we run a computation it may fail (indicating we need to try again later), or may succeed.
-The best way to explain it is with the type of \hs{dynamic}:
+We use the helper function \hs{reordering} which provides the key to build (\hs{k}), the dependencies of \hs{k} (\hs{dk}) and an action to build \hs{k} (\hs{act}).
+The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Compute} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway.
+If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possiblities:
+
+\begin{enumerate}
+\item The key cannot be built at this time because it depends on a key which has not yet been computed - and thus \hs{act} returns \hs{Left e}, where \hs{e} is the key that we are waiting for.
+In response we return \hs{e} as the result of step - asking \hs{reordering} to requeue this key after \hs{e}.
+\item The key builds successfully, so we store it and return \hs{Nothing} to indicate \hs{reordering} can move on to the next key.
+\end{enumerate}
+
+The astute reader may notice that \hs{reordering} supplies extra information that is ignored with \hs{_} patterns.
+While not required for \Excel, we have implemented build systems using \hs{reordering} which use verifying and constructive traces, effectively turning \Excel into a cloud build system and ensuring \hs{reordering} is not overly fitted to \Excel alone (additional implementations available in the suplementary material).
+The definition of \hs{reordering} is:
 
 \begin{minted}{haskell}
-dynamic
+reordering
     :: (m ~ M (i, [k]) k v, Default i)
@@
 
-The Shake approach for dependency tracking is different yet again, with a recursive approach. We can implement Shake with:
+The Shake approach for dependency tracking involves recording traces and verifying them. Using the \hs{Trace} type defined in \ref{sec-verifying-traces} we can implement:
 
 \begin{minted}{haskell}
-type Shake k v = Map.Map k (Hash v, [(k, Hash v)])
-
--- | The simplified Shake approach of recording previous traces
-shake :: Hashable v => Build Monad (Shake k v) k v
+shake :: Hashable v => Build Monad [Trace k v] k v
 shake = recursive $ \k _ ask act -> do
-    info <- getInfo
-    valid <- case Map.lookup k info of
-        Nothing -> return False
-        Just (me, deps) ->
-            (maybe False (== me) <$> getStoreHashMaybe k) &&^
-            allM (\(d,h) -> (== h) . getHash <$> ask d) deps
-    unless valid $ do
-        (ds, v) <- act
+    poss <- traceMatch (\k v -> (==) v . getHash <$> ask k) k =<< getInfo
+    h <- getStoreHash k
+    when (h `notElem` poss) $ do
+        (dk, v) <- act
         putStore k v
-        dsh <- mapM getStoreHash ds
-        modifyInfo $ Map.insert k (getHash v, zip ds dsh)
+        dh <- mapM getStoreHash dk
+        modifyInfo (Trace k (zip dk dh) (getHash v) :)
 \end{minted}
 
-The basic implementation of Shake stores a trace - each target key is associated with a \hs{Hash v} (the value that resulted last time it was built), along with a list of \hs{(k, Hash v)} pairs, being the dependencies and the values it used at that point.
-The essence of Shake is that it considers a file dirty if the trace is not consistent with what currently exists on disk, and after rebuilding, captures a trace.
-The \hs{recursive} combinator is defined as:
+This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to ask for the result of a dependecy (\hs{ask}), and a way to build \hs{k} (\hs{act}).
+The code first looks up the traces that match the current situation, creating a list of possible results as \hs{poss}, using the helper \hs{traceMatch}.
+We check if the current value for the key is one of the possible results, and it not, build the key and store it in the trace.
+We can implement \hs{traceMatch}, which is used in all build systems working with \hs{Trace}, as:
 
 \begin{minted}{haskell}
--- | Build a rule at most once in a single execution
-recursive :: Default i => (k -> Maybe [k] -> (k -> M i k v v) -> M i k v ([k], v) -> M i k v ()) -> Build Monad i k v
+traceMatch :: (Monad m, Eq k) => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
+traceMatch check k ts = mapMaybeM f ts
+    where f (Trace k2 dkv v) = do
+                b <- return (k == k2) &&^ allM (uncurry check) dkv
+                return $ if b then Just v else Nothing
+\end{minted}
+
+The function \hs{traceMatch} requires a way to check a dependency, and the key, and the traces.
+We return the \hs{result} field from all traces where the key and dependencies match.
+We define the recursive combinator as:
+
+\begin{minted}{haskell}
+recursive
+    :: (m ~ M i k v, Default i)
+    => (k -> Maybe [k] -> (k -> m v) -> m ([k], v) -> m ())
+    -> Build Monad i k v
 recursive step compute = runM . ensure
@@
 The required \hs{step} function gets given the \hs{k} to build, the dependencies as \hs{Maybe [k]} (where they are \hs{Nothing} if the rule is truely monadic), a function to demand the result of a dependent key, and a function to compute the current value and return the actual dependencies.
-To ensure each file is only visited once in a single execution a piece of temporary state is used with \hs{getTemp} and \hs{modifyTemp} to record which files were visited - this state is not persisted to the \hs{i} info parameter.
+To ensure each file is only visited once in a single execution a piece of temporary state is used with \hs{getTemp} and \hs{modifyTemp} to record which files were visited - this state is not persisted to the \hs{i} info parameter and is reset on each fresh execution of \hs{M}.
 
@@
 
-Now we have seen the three dependency schemes, we can directly use \hs{topological} to define Bazel.
-Furthermore, in many ways Bazel is a tracing build system, so it has a number of similarities to Shake.
-Concretely we have:
+Now we have seen the three dependency schemes, we can directly reuse \hs{topological} to define Bazel.
+Furthermore, as Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type.
+We define \Bazel as:
 
 \begin{minted}{haskell}
-data Bazel k v = Bazel
-    {bzKnown :: Map.Map (k, [Hash v]) (Hash v)
-    ,bzContent :: Map.Map (Hash v) v
-    } deriving Show
-
-bazel :: Hashable v => Build Applicative (Bazel k v) k v
-bazel = topological $ \k ds act -> do
-    ds <- mapM getStoreHash ds
-    res <- Map.lookup (k, ds) . bzKnown <$> getInfo
-    case res of
-        Nothing -> do
-            res <- act
+bazel :: Hashable v => Build Applicative (Traces k v) k v
+bazel = topological $ \k dk act -> do
+    poss <- traceMatch (\k v -> (== v) <$> getStoreHash k) k . traces =<< getInfo
+    if null poss then do
+        v <- act
+        dh <- mapM getStoreHash dk
         modifyInfo $ \i -> i
-                {bzKnown = Map.insert (k, ds) (getHash res) $ bzKnown i
-                ,bzContent = Map.insert (getHash res) res $ bzContent i}
-            putStore k res
-        Just res -> do
-            now <- getStoreHashMaybe k
-            when (now /= Just res) $
-                putStore k . (Map.! res) . bzContent =<< getInfo
+            {traces = Trace k (zip dk dh) (getHash v) : traces i
+            ,contents = Map.insert (getHash v) v $ contents i}
+        putStore k v
+    else do
+        h <- getStoreHash k
+        when (h `notElem` poss) $
+            putStore k . (Map.! head poss) . contents =<< getInfo
 \end{minted}
 
-The state matches that described in \S\ref{sec-build} - we have a map indexed by a target key, and the hashes of the dependencies of that key.
-Given that, we point at a resultant hash.
-Separately we have a map giving the contents for a given hash.
-For simplicity we assume that all hashes have corresponding entries in the content index.
-
-The implementation works by getting the hashes of all dependencies and looking them up in the trace table. There are three possibilities:
-
-\begin{enumerate}
-\item This situation has never been encountered before. Execute the action and record it in the info.
-\item This situation has been encountered before, and the result was the same as we have now, so nothing to do.
-\item This situation has been encountered before but we don't have that result locally (usually we have nothing), so just download the result.
-\end{enumerate}
+We execute keys in a topological order.
+For each key we first get a list of the traces that match.
+If there are none we rebuild and record the trace, much like \Shake, but also recording the value of the result.
+If there are some matching traces, and the current value of \hs{k} is not compatible, we grab the first possiblity and grab its value from \hs{contents}.
 
@@
 
-Using the framework above we have implemented the full 3x3 matrix, of dependency scheme and change approach.
-To us, the interesting point on the design space would be monadic dependencies using the \hs{recursive} strategy combined with cloud builds as per \Bazel, so we focus on that.
-The implementation is:
+Using the abstractions and approaches built thus far, we have shown how to combine dependency scheme and change approach to reproduce existing build systems.
+In the attached materials we have implemented 9 build systems corresponding to all three dependency schemes, matched with all three change approaches.
+To us, the most interesting build system as yet unavailable would matching recursive ordering with constructive traces - providing a cloud-capable build system with minimality, cut-off and monadic dependencies.
+Using our framework it is possible to define and test such a system:
 
 \begin{minted}{haskell}
-data ShazelResult k v = ShazelResult [(k, Hash v)] (Hash v) deriving Show
-
-data Shazel k v = Shazel
-    {szKnown :: Map.Map k [ShazelResult k v]
-    ,szContent :: Map.Map (Hash v) v
-    } deriving Show
-
-instance Default (Shazel k v) where def = Shazel def def
-
-shazel :: Hashable v => Build Monad (Shazel k v) k v
-shazel = recursive $ \k _ ask act -> do
-    poss <- Map.findWithDefault [] k . szKnown <$> getInfo
-    res <- flip filterM poss $ \(ShazelResult ds r) -> allM (\(k,h) -> (==) h . getHash <$> ask k) ds
-    case res of
-        [] -> do
-            (ds, v) <- act
-            dsv <- mapM getStoreHash ds
+cloudShake :: Hashable v => Build Monad (Traces k v) k v
+cloudShake = recursive $ \k _ ask act -> do
+    poss <- traceMatch (\k v -> (== v) . getHash <$> ask k) k . traces =<< getInfo
+    if null poss then do
+        (dk, v) <- act
+        dh <- mapM getStoreHash dk
         modifyInfo $ \i -> i
-                {szKnown = Map.insertWith (++) k [ShazelResult (zip ds dsv) (getHash v)] $ szKnown i
-                ,szContent = Map.insert (getHash v) v $ szContent i}
+            {traces = Trace k (zip dk dh) (getHash v) : traces i
+            ,contents = Map.insert (getHash v) v $ contents i}
         putStore k v
-        _ -> do
-            let poss = [v | ShazelResult _ v <- res]
-            now <- getStoreHashMaybe k
-            when (now `notElem` map Just poss) $
-                putStore k . (Map.! head poss) . szContent =<< getInfo
+    else do
+        now <- getStoreHash k
+        when (now `notElem` poss) $
+            putStore k . (Map.! head poss) . contents =<< getInfo
 \end{minted}
 
-The difference between the data type from Bazel is worth looking at. In both cases they essentially store traces - a target key, target hash, list of dependency keys and hashes. However:
+The differences from \hs{bazel} are minor - the dependency scheme has changed from \hs{topological} to \hs{recursive}, and thus the dependency keys \hs{dk} are captured from the action rather than in advance, and the checking calls \hs{ask} to get the result instead of accessing the store directly.
+We believe this accurately captures and can be used to reason about how to make \Shake work with a cloud-based cache -- a challenge we intend to attempt in the future.
 
-\begin{itemize}
-\item Bazel doesn't bother to store the dependency keys, as they can be calculated from the target using \hs{getDependencies} due to the \hs{Applicative} nature.
-It indexes on the target key and dependency hashes, immediately providing the target hash.
-\item Cloud Shake has to store the dependency keys, but can't index the Map by them - the checking must mirror the monadic nature. A more refined/efficient data type would be:
+\subsection{Smarter \hs{[Trace]} Structures}\label{sec-smart-traces}
+
+In the examples above, we have used \hs{[Trace]} to capture a list of traces stored in the info field, which accurately models the data and operations.
+However, using a list necessarily means that finding the right trace takes $O(n)$.
+For each of the \hs{Trace} based systems it is possible to devise a smarter representation, which we sketch below.
+Note that these implementations do not avoid calls to \hs{compute}, merely overheads in the build system itself.
+
+\begin{enumerate}
+\item Any system using verifying traces, e.g.
+\Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}.
+Therefore, such systems can store \hs{Map k (Trace k v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
+\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
+\item Any \hs{Applicative} system storing constructive traces, e.g.
+\Bazel, can index directly from the key and results to the output result - i.e.
+\hs{Map (k, [Hash v]) (Hash v)}.
+Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash.
+In this formulation we have removed the possibilty for a single key/dependency state to map to multiple different hashes, e.g.
+on a non-determinstic build - something \Bazel already prohibits which is discussed more in \S\ref{sec-non-determinism}.
+\item Finally, a \hs{Monad} system with constructive traces can be efficiently encoded as \hs{Map k (Choice k v)}, assuming a definition of \hs{Choice} as:
 \begin{minted}{haskell}
-data Choice = Choice k (Map (Hash v) Choice)
+data Choice k v = Choice k (Map (Hash v) Choice)
                 | Result (Hash v)
 \end{minted}
-\end{itemize}
+Here the \hs{Choice} effectively encodes a tree, asking successive questions about keys, and taking different branches based on the resulting value, until it reaches a final result.
+Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
+\end{enumerate}
 
-We also technically allow multiple traces to be in the store at the same time, although that is merely something that falls out for free - we would have to do work to avoid it.
+As an aside we note that the optimal cloud structure for applicative and monadic constructive systems has a resemblance to the free dependency structures described in \S\ref{sec-free-build}.

### 2018-03-11 paper/3-compute.tex
Minor revision

@@
 
-This section presents purely functional abstractions for build systems that
-allow us to express all the intricacies of build systems discussed in the
-previous section~\S\ref{sec-background} and design complex build systems from
-simple primitives. Specifically, we present the \emph{compute} and \emph{build}
+This section presents purely functional abstractions that allow us to express
+all the intricacies of build systems discussed in the previous
+section~\S\ref{sec-background} and design complex build systems from simple
+primitives. Specifically, we present the \emph{compute} and \emph{build}
 abstractions:
@@
 rules in software build systems or formulas in spreadsheets. It is completely
-isolated from the world of compilers, file systems, calculation chains, memories,
-caches, and all other complexities of real build systems. We explain the compute
-type in~\S\ref{sec-general-compute}, and then discuss several noteworthy special
+isolated from the world of compilers, file systems, calculation chains, caches,
+and all other complexities of real build systems. We explain the compute type
+in~\S\ref{sec-general-compute}, and then discuss several noteworthy special
 cases: functorial (\S\ref{sec-compute-functor}), applicative
@@
 definition of build systems in~\S\ref{sec-general-build}.
-Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} further deconstruct
-build algorithms and provide concrete implementations for several build systems.
+Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the build
+abstraction further and provide concrete implementations for several build
+systems.
 
@@
 \todo{AM}{Turn the above into a formal definition of `correctness' and add a section that provides a mathematical definition of correctness, which does not rely on materialising intermediate values.}
+
+\subsection{Free build structures}\label{sec-free-build}
+
+\begin{minted}{haskell}
+data Depend k v r = Depend [k] ([v] -> r)
+
+data Depends k v r = Depends [k] ([v] -> Depends k v r)
+                   | Done r
+\end{minted}

### 2018-03-11 paper/3-compute.tex
Avoid specific function names in the vocabulary

@@
 value}, which is a useful feature for build systems. In addition to usual
-\hs{get} and \hs{put} operations, some stores support the \hs{getHash} operation
-implemented more efficiently than by hashing the result of \hs{get}. For example,
-GVFS (Git Virtual File System)~\cite{gvfs} stores file hashes and downloads file
-contents only on demand -- an important feature for cloud build systems.
+read and write operations, some stores also support a dedicated \emph{read hash}
+operation, which is implemented more efficiently than by reading and hashing the
+result. For example, GVFS (Git Virtual File System)~\cite{gvfs} stores file
+hashes and downloads file contents only on demand -- an important feature for
+cloud build systems.
 

### 2018-03-11 paper/6-engineering.tex
Lots of stuff on engineering

--- 2018-03-09 paper/engineering.tex
@@
 
-Or ``Build systems in the real world'', or ``Crazy things section''...
+In the previous sections we have modelled the most critical subset of various build systems.
+However, like all real-world systems, there are many corners that obscure the essence.
+In this section we discuss some of those details, what would need to be done to capture them in our model, and what the impact would be.
 
-\subsection{Non-deterministic computations}\label{sec-non-determinism}
+\subsection{Partial stores, exceptions}
 
-Sandboxing: guarding against non-determinism due to missing dependencies.
-
-\todo{AM}{Rewrite everything below.}
-
-Non-determinism, specifically, its \emph{useless} form. This doesn't
-include \Excel's \textsf{RANDBETWEEN} function, which is useful.
-To define correctness of non-deterministic build systems, one needs to
-switch from \hs{Applicative} and \hs{Monad} abstractions to
-\hs{Alternative} and \hs{MonadPlus}, respectively, but this is not
-required for actual implementations, most of which can happily accept
-non-deterministic results (with a notable exception of \Buck).
-
-... Some old text:
-
-Build systems and spreadsheets compute output values from input and intermediate
-values. In the most typical case, these \emph{computations} are \emph{functions},
-such as \textsf{C1 = A1 + B1}, i.e. their result is uniquely determined by the
-input values. However, in general they can be \emph{relations}, i.e. have
-multiple valid results. A spreadsheet example: \textsf{A2 = A1 + RANDOM(1,6)}.
-This computation has six valid results for each input value \textsf{A1}. In
-build systems, the object file \textsf{obj/file.o} is sometimes not uniquely
-determined by the source \textsf{src/file.c} -- different compiler runs may
-produce different valid results.
-
-Some build systems, e.g. \Buck, require all tasks to be
-\emph{deterministic}, i.e. produce exactly the same output when run on the
-same inputs. However, not all tasks are deterministic, and there are build
-systems that support \emph{non-determinism}. A simple example is \Excel's
-\textsf{RANDBETWEEN(low, high)} that returns a random integer in the
-interval \textsf{[low, high]}.
+Our model assumes a world where the store is fully-defined, every \hs{k} is associated with a \hs{v}, and every compute successfully completes returning a valid value.
+In the real world, build systems frequently deal with errors - e.g.
+``file not found'', or ``compilation failed''.
+We can model such failure conditions by instantiating \hs{v} to either \hs{Maybe v} (for missing values) or \hs{Either e v} (for exceptions of type \hs{e}).
+That can model the the values inside the store and the compute function, but because \hs{v} is polymorphic for builds, it does not let the build system apply special behaviour for errors - e.g.
+early aborting.
+We consider such details an engineering concern, which has no real impact on the power or high-level implementation of the build system.
 
@@
 
-Nothing fundamentally difficult and fits our theory, but hard to get right in
-practice.
+While we have given simple implementations assuming a single thread of execution, all the build systems we address are actually concurrent.
+While it complicates the model to add concurrency, the complications can be restricted exclusively to the dependency strategy:
 
-\todo{NM}{Add a paragraph on how to introduce concurrency to implementations
-in~\S\ref{sec-examples}}.
+\begin{enumerate}
+\item The \hs{topological} function can instead retain the dependency graph, and whenever all the dependencies of a node are complete, the node itself can be started.
+\item The \hs{reordering} function can be made parallel in a few ways, but the most direct is to have $n$ threads reading entries from the list of keys.
+As before, if a key requires something not yet built, it is added to the end -- the difference is that sometimes things will be moved back not because they are out of order but because of races with earlier nodes that had not yet finished.
+As a consequence, over successive runs, potentially racey dependencies will be separated, giving better parallelism over time.
+\item The \hs{recursive} function can be made parallel by working more directly on the \hs{Depeneds} structure, which gives multiple dependency keys at a time (all those arising from \hs{Applicative} dependencies).
+All such dependencies can be executed simultaneously, with a slightly smarter check for whether something is already executing that also blocks until an existing execution has finished.
+\end{enumerate}
+
+The actual implementation of the parallel strategies is not overly onerous, but neither is it beautiful or informative.
+
+\subsection{Impure computations}\label{sec-non-determinism}
+
+In our model we define \hs{Compute} as a function - when give the same inputs it will always produce the same outputs.
+Alas, the real-world is not as obliging.
+Some examples of impure \hs{Compute} functions include:
+
+\begin{description}
+\item[Untracked state] Some actions depend on untracked state - for example your C compilation may explicitly list the \texttt{source.c} file as a dependency, but it may not record that the version of \texttt{gcc} is also a dependency.
+\item[Volatility] Some actions are defined to change in every execution, for example \Excel has a function \textsf{RANDBETWEEN} which produces a random number - on each recalculation this function is defined to change.
+Similarly, build systems like \Make and \Shake provide \textsf{phony} rules which are also volatile.
+\item[Non-determinism] Some actions are non-determinstic, producing one result from a possble set.
+As an example, GHC when compiled using parallelism can change the order in which unique variables are obtained from the supply, producing semantically identical results but with different hidden variables.
+\end{description}
+
+Interestingly, there is signficant interplay between all three sources of impurity.
+Systems like \Bazel use various sandboxing techniques to guard against missing dependencies, but none are likely to capture all dependencies right down to CPU model and microcode version.
+Rules that do have untracked state can be marked as volatile, a technique \Excel takes with the \textsf{INDIRECT} function, removing the untracked-state at the cost of more recomputation.
+
+We note that while all the simple implementations presented in \S\ref{sec-implementations} can deal with non-determinism, some real build systems cannot, e.g. \Buck.
+
+One tempting way of modelling non-determinism is to enrich \hs{Compute} from \hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus} respectively - both provide the underlying class with zero and choice.
+However, that seems to encode a different type of non-determinism - a kind of dependency-level non-determinism when most rules really experience a value-level non-determinism.
+
+Volatility can be modelled by introducing a special \hs{k} whose value is changed between every run.
 
@@
 
-Cloud aspects: eviction/garbage collection, `frankenbuilds', etc.
+Our model of cloud builds provides a basic framework to discuss and reason about them, but lacks a number of important engineering corners:
 
-\todo{AM}{Descibe frankenbuilds linking to CloudBuild paper.}
+\begin{description}
+\item[Materialisation] Sometimes input files will go through many transformations before producing the end result, e.g.
+an installer package.
+These intermediate steps may be large, so some cloud builds can execute the graph \textit{without} downloading or \textit{materialising} the intermediate nodes - only the final one.
+Some build systems can go even further, integrating with the file system to only materialise the file when the user accesses it \cite{google_internal}.
+\item[Eviction] The store of traces as shown grows indefinitely.
+In many cases resource constraints mean that eventually items must be evicted from the store that are predicted not to be used again.
+One option is to evict the contents and any trace that mentions the now-defunct \hs{Hash}.
+However, if the build system can defer materialisation, it may be possible to only evict the contents, allowing builds to pass-through hashes of values where the underlying value is not known.
+If so, the build must be able to recreate the value if required, potentially dealing with a different result in a future run.
+\item[Frankenbuilds] A build is considered a franken-build \cite{cloudbuild} if a value is calculated locally, but something that depends on that key is pulled from the cache, and the value calculated locally does not match what was previously calculated by the cloud.
+Our implementations avoid this issue by storing complete traces, but if a cloud build system was to only reference input nodes it could be a problem.
+\item[Communication] With a cloud build communication with a central server can be a bottleneck, so it is important to send only the minimum amount of information, optimising with respect to build-system specific invariants - see \S\ref{sec-smart-traces} for some possible optimisations.
+\item[Offloading] Once the cloud is storing build products and traces, it is not infeasible for the cloud to also contain dedicated workers that can help with jobs run locally - offloading some of the computation and potentially running vastly more commands in parallel.
+\end{description}
 
@@
 
-Annotating dependency graphs with hashes, times, etc. is in many cases just an
-engineering trade-off.
+Some build systems, for example \Excel and \Ninja are capable of recomputing a node if either its dependencies change, \textit{or} the rule itself changes. For example:
 
-\todo{NM}{Clarify, e.g. Shake's \cmd{ChangeModtime}.}
+\begin{minted}[xleftmargin=10pt]{bash}
+A1 = 20
+A2 = 10
+B1 = A1+A2
+\end{minted}
 
-\todo{NM}{Explain how self-tracking can be implemented.}
+In \Excel the user can alter the value produced by \textsf{B1} by either editing the inputs of \textsf{A1} or \textsf{A2}, \textit{or} editing the formula in \textsf{B1} - e.g.
+to \textsf{A1-A2}.
+This pattern can be captured by describing the rule producing \textsf{B1} as also depending on the value \textsf{B1.formula}.
+In a \hs{Monad} build system the implementation can be given very directly as:
+
+\begin{minted}{haskell}
+compute ask "B1" = do
+    formula <- ask "B1-formula"
+    evalFormula ask formula
+\end{minted}
+
+Namely, first look up the formula, then interpret it.
+In an \hs{Applicative} system it is not possible to change dependencies based on the formula, as would be required, so instead the formula can be captured as a dependency (but its value not used) and \textit{also} baked directly into the \hs{compute} function.
+
+The build systems that have precise tracking of self-computation are all onces which use a non-Embedded Domain Specific Language to describe computations.
+Those which make use of a full programming language, e.g.
+\Shake, are much more difficult to implement equality on rules as they are arbitrary functions making use of opqaue abstractions.
+For such build systems the incredibly pessimistic assuming of saying that any change to the build system potentially changes any rule can often be used -- the classic example being a \texttt{Makefile} depending on itself.
 
 \subsection{Iterative computations}\label{sec-iterative-compute}
+
+Some computations are best described not by a chain of acyclic dependencies, but by a loop - for example \LaTeX requires repeated rebuilding until it reaches a fixed point - something that can be directly expressed in build systems such as Pluto \cite{pluto}.
+Another example of cyclic computations is \Excel, where a cell can depend on itself, for example:
+
+\begin{minted}{bash}
+A1 = A1 + 1
+\end{minted}
+
+In such cases \Excel guarantees to execute the computation exactly once per refresh cycle.
 

### 2018-03-11 paper/5-implementations.tex
Rename ask to fetch, in agreement with Andrey

@@
 shake :: Hashable v => Build Monad [Trace k v] k v
-shake = recursive $ \k _ ask act -> do
-    poss <- traceMatch (\k v -> (==) v . getHash <$> ask k) k =<< getInfo
+shake = recursive $ \k _ fetch act -> do
+    poss <- traceMatch (\k v -> (==) v . getHash <$> fetch k) k =<< getInfo
     h <- getStoreHash k
@@
 
-This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to ask for the result of a dependecy (\hs{ask}), and a way to build \hs{k} (\hs{act}).
+This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to fetch the result of a dependecy (\hs{fetch}), and a way to build \hs{k} (\hs{act}).
 The code first looks up the traces that match the current situation, creating a list of possible results as \hs{poss}, using the helper \hs{traceMatch}.
@@
         ensure k = do
-            let ask x = ensure x >> getStore x
+            let fetch x = ensure x >> getStore x
             done <- getTemp
@@
                 modifyTemp $ Set.insert k set
-                case trackDependencies compute ask k of
+                case trackDependencies compute fetch k of
                     Nothing -> return ()
-                    Just act -> step k (getDependenciesMaybe compute k) ask act
+                    Just act -> step k (getDependenciesMaybe compute k) fetch act
 \end{minted}
@@
 cloudShake :: Hashable v => Build Monad (Traces k v) k v
-cloudShake = recursive $ \k _ ask act -> do
-    poss <- traceMatch (\k v -> (== v) . getHash <$> ask k) k . traces =<< getInfo
+cloudShake = recursive $ \k _ fetch act -> do
+    poss <- traceMatch (\k v -> (== v) . getHash <$> fetch k) k . traces =<< getInfo
     if null poss then do
@@
 
-The differences from \hs{bazel} are minor - the dependency scheme has changed from \hs{topological} to \hs{recursive}, and thus the dependency keys \hs{dk} are captured from the action rather than in advance, and the checking calls \hs{ask} to get the result instead of accessing the store directly.
+The differences from \hs{bazel} are minor - the dependency scheme has changed from \hs{topological} to \hs{recursive}, and thus the dependency keys \hs{dk} are captured from the action rather than in advance, and the checking calls \hs{fetch} to get the result instead of accessing the store directly.
 We believe this accurately captures and can be used to reason about how to make \Shake work with a cloud-based cache -- a challenge we intend to attempt in the future.

### 2018-03-11 src/Neil/Build.hs
Rename ask to fetch

@@ -12,7 +12,6 @@ module Neil.Build(
     ) where
 
 import Neil.Compute
-import Control.Monad.Extra
 import qualified Neil.DynamicMap as DM
 import Data.Typeable
 import Data.Hashable
@@ -20,7 +19,6 @@ import Control.Monad.Trans.State
 import Data.Maybe
 import Data.Default
 import qualified Data.Set as Set
-import qualified Data.Map as Map
 
 
 type Build c i k v = (Ord k, Typeable k) => Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
@@ -101,4 +99,3 @@ getHash = Hash . hash
 
 getStoreHash :: (Hashable v) => k -> M i k v (Hash v)
 getStoreHash = fmap getHash . getStore
-

### 2018-03-11 src/Neil/Builder.hs

@@ -23,11 +23,7 @@ import Neil.Compute
 import Control.Monad.Extra
 import Data.Tuple.Extra
 import Data.Default
-import Data.Maybe
 import Data.Either.Extra
-import Debug.Trace
-import Data.List
-import Data.Typeable
 import Data.Map(Map)
 import Data.Set(Set)
 import qualified Data.Set as Set hiding (Set)
@@ -59,13 +55,13 @@ recursive
 recursive step compute = runM . ensure
     where
         ensure k = do
-            let ask x = ensure x >> getStore x
+            let fetch x = ensure x >> getStore x
             Recursive done <- getTemp
             when (k `Set.notMember` done) $ do
                 modifyTemp $ \(Recursive done) -> Recursive $ Set.insert k done
-                case trackDependencies compute ask k of
+                case trackDependencies compute fetch k of
                     Nothing -> return ()
-                    Just act -> step k (getDependenciesMaybe compute k) ask act
+                    Just act -> step k (getDependenciesMaybe compute k) fetch act
 
 
 reordering
@@ -104,7 +100,7 @@ data Traces k v = Traces
 
 instance Default (Traces k v) where def = Traces def def
 
--- | Given a way to ask for a dependency, and the key under construction, and the traces, pick the hashes that match
+-- | Given a way to fetch for a dependency, and the key under construction, and the traces, pick the hashes that match
 traceMatch :: (Monad m, Eq k) => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
 traceMatch check k ts = mapMaybeM f ts
     where f (Trace k2 dkv v) = do
@@ -171,16 +167,16 @@ makeTrace = topological $ \k dk act -> do
 
 
 shakeDirtyBit :: Eq v => Build Monad (Changed k v, ()) k v
-shakeDirtyBit = withChangedMonad $ recursive $ \k dk ask act -> do
-    dirty <- getChanged k ||^ maybe (return True) (anyM (\k -> ask k >> getChanged k)) dk
+shakeDirtyBit = withChangedMonad $ recursive $ \k dk fetch act -> do
+    dirty <- getChanged k ||^ maybe (return True) (anyM (\k -> fetch k >> getChanged k)) dk
     when dirty $
         putStore k . snd =<< act
 
 
[…]

### 2018-03-11 paper/2-background.tex
Tweaks to background

--- 2018-03-10 paper/background.tex
@@
 introduce main domain-specific notions and requirements. Other notable examples
-of build systems and their relation to these four will be discussed
+of build systems and their relation to these four are discussed
 in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.
+
+We start by giving an intuitive definition of a build system. After providing
+more context and examples in the rest of this section, we
+develop a formal definition in~\S\ref{sec-general-build}.
+
+\definition[Build system]{Given a description of tasks, a \emph{build system}
+executes them in an order that respects their dependencies to produce a
+specified output from supplied inputs.}
+\label{def-build}
 
@@
 programs from source code. It uses \emph{makefiles} to describe tasks (often
-referred to as \emph{build rules}) and their dependencies in a simple text form.
+referred to as \emph{build rules}) and their dependencies in a simple textual form.
 For example:
@@
 files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
-executing\footnote{In this example we treat \cmd{gcc} as a pure function for the
+executing\footnote{In this example we pretend \cmd{gcc} is a pure function for the
 sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and the
@@
 
-If the user modifies the sources of the utility library and runs \Make, it will
+If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
+will first build \cmd{util.o} and \cmd{main.o}, in any order since the
+corresponding tasks are independent, and then build \cmd{main.exe}. If the
+user modifies the sources of \cmd{util.h} and runs \Make again, it will
 perform a \emph{full rebuild}, because all three tasks transitively depend on
-the library, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
-user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient: indeed,
+the \cmd{util.h}, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
+user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient:
 the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
@@
 
-The following property is essential for build systems, it is their raison d'\^etre.
+The following property is essential for build systems, it is their raison d'\^etre:
 
 \definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
-most once per build and only if they transitively depend on files that changed
+most once per build and only if they transitively depend on inputs that changed
 since the previous build.}\label{def-minimal}
@@
 To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
-modification time} to detect the files that changed\footnote{Technically, you
+modification time} to detect which files changed\footnote{Technically, you
 can fool \Make by altering the modification time of a file without changing its
@@
 complete task dependency graph from the information contained in the makefile
-and executes tasks in a \emph{topological order}. See further details
-in~\S\ref{sec-implementation-make}.
+and executes tasks in a \emph{topological order}. For a more concrete description
+see~\S\ref{sec-implementation-make}.
 
@@
 to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
-$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and this is
+$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and which is
 not known statically. One might say that the value of \cmd{C1} is statically
-known in this example, but imagine that it is a result of a long computation
-chain -- its value will then only become available during the build.
+known in this particular example, but imagine that it is the result of a long computation
+chain -- its value will only become available during the build.
 

### 2018-03-11 paper/6-engineering.tex

@@
 
-In such cases \Excel guarantees to execute the computation exactly once per refresh cycle.
+In such cases \Excel will normally not execute anything, but if the ``Iterative Calculations'' feature is enabled will execute the formula a maximum of $n$ times per calculation (where $n$ is a setting that defaults to 100).
 
-Iterative computations, e.g. LaTeX. \Excel has built-in support for this.
+For examples like \LaTeX we consider the proper encoding to not be with circular rules, but with a series of iterative steps, as described in \S\ref{neil-iterative-on-stack-overflow}.
+It is important that the number of executions is bounded, otherwise the build system may not terminate (a legitimate concern with \LaTeX which can be put into a situation where it is bistable or divering over multiple executions).
 
-\todo{NM}{Clarify how this fits with the example implementations.}
+The examples in \Excel tend to encode either mutable state, or recurrence relations.
+The former is only required because \Excel inherently lacks the ability to write mutable state, and the latter is probably better solved using explicit recurrence formulae.
+
+Overall we choose not to deal with cyclic rules, a choice that most build systems also follow.
 
@@
 
-Polymorphic keys and values: tasks with multiple outputs, e.g. GHC's \cmd{hi}
-and \cmd{o} files, phony tasks, etc.
+Our build system abstraction assumes a \hs{k -> v} store, along with a build system that works directly on \hs{k} and \hs{v} values.
+However, certain build systems provide greater flexiblity, e.g.
+\Shake permits polymorphic keys and values, allowing types that are only stored in the Shake information, and never persisted to the Store.
 
-\todo{AM}{Give an example.}
+As one example of richer keys/values, consider the version of \textsf{gcc} - for many builds it should be a dependency.
+In \Shake we can define an \textit{oracle} rule as per \cite{mitchell2012shake} whose value is the result of running \hs{gcc --version} and which is volatile, making the \textsf{gcc} version something that can be depended upon.
+Of course, provided the build can express volatile dependencies and supports cutoff, the version number could equally be written to a file and used in a similar way.
 
-\todo{NM}{Describe how to add this feature to \Shake.}
+Another more compelling example is build rules that produce multiple keys -- for example \textsf{ghc Foo.hs} produces both \textsf{Foo.hi} and \textsf{Foo.o}.
+That can be represented by having a key whose value is a pair of file names, and whose result is a pair of file contents.
+From that, the rule for \textsf{Foo.hi} can be the first-component of the result of the pair.
+Again, such an operation can be encoded without polymorphic keys provided the pair of files (or a dummy file representing the pair) is marked as changed if either of the contained files change.
+Once again, polymorphic dependencies provide convenience rather than power.
+
+Authors have remarked that polymorphism provides a much easier expression of concepts, e.g. \citet{hadrian}, but it is not essential and thus not necessary to model.

### 2018-03-11 paper/2-background.tex
Minor tweaks

@@
 perform a \emph{full rebuild}, because all three tasks transitively depend on
-the \cmd{util.h}, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
-user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient:
+\cmd{util.h}, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if
+the user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient:
 the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
@@
 content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
-under the assumption that you do not do that.}, and (ii) it constructs a
-complete task dependency graph from the information contained in the makefile
-and executes tasks in a \emph{topological order}. For a more concrete description
+under the assumption that you do not do that.}, and (ii) it constructs a task
+dependency graph from the information contained in the makefile and executes
+tasks in a \emph{topological order}. For a more concrete description
 see~\S\ref{sec-implementation-make}.

### 2018-03-11 paper/7-related.tex
Rename the chapters so they are in order
Write the related work

--- 2018-03-07 paper/related.tex
@@
 
+In section \S\ref{sec-background} we cover many build systems, but there is frustrating little work comparatively analysing build systems. In this section we cover a few build systems that may not
+
+\subsection{Other Build Systems}
+
+Most build systems, when viewed at the level we talk, can be captured with minor variations on the code presented in \S\ref{sec-implementations}. As some examples:
+
 \begin{itemize}
-    \item Memoization
-    \item Self-adjusting computation
-    \item Lens-like formulation
+\item \Ninja \cite{ninja} combines the dependency strategy of \Make with the validating traces of \Shake - our associated implementation provides such a combination.
+Ninja is also capable of modelling rules that produce multiple results, a limited form of polymorphism \S\ref{sec-polymorphism}.
+\item \Tup \cite{tup} functions much like \Make, but with a refined dirty-bit implementation that watches the file system for changes and can thus avoid rechecking the entire graph, and with automatic deleting of stale results.
+\item \Redo \cite{redo} almost exactly matches \Shake at the level of detail given here, differing only on aspects like polymorphic dependencies \S\ref{sec-polymorphism}.
+\item \Buck \cite{buck} and \CloudBuild \cite{cloudbuild} are both fundamentally similar to \Bazel, differing around areas such as non-determinism \S\ref{sec-sec-non-determinism}.
+\item \Nix \cite{nix} has coarse-grained monadic dependencies, with precise hashing of dependencies and downloading of precomputed build products, so may be, at some level of abstraction, the same as the hypothetical Cloud Shake from \S\ref{sec-implementation-cloud-shake}.
+However, \Nix is not intended as a build system, and the coarse grained nature (packages, not individual files) makes it targetted to a different purpose.
+\todo{NM}{Read comment below when not on an airplane.}
+% John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
+% \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.
+\item \Pluto \cite{pluto} has power to express cyclic dependenies, but that is implemented \todo{NM}{Read the paper so I can say something precise}
 \end{itemize}
+
+The one build system we are aware of that cannot be modelled in our framework is \Fabricate \cite{fabricate}.
+In \Fabricate a build system is a script which is run in-order, in the spirit of\footnote{\Fabricate requires scripts to be written in Python, but those details are not fundamental to what makes \Fabricate special.}:
+
+\begin{minted}{bash}
+gcc -c util.c
+gcc -c main.c
+gcc util.o main.o -o main.exe
+\end{minted}
+
+To achieve minimality, each separate command is traced at the OS-level, allowing \Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from \cmd{util.c}.
+In future runs \Fabricate runs the script from start to finish, skipping any commands where no inputs have changed.
+
+Taking our abstraction, it is possible to encode \Fabricate assuming that nodes like \cmd{gcc -c util.c} are keys, there is a linear dependency between each successive node, and that the OS-level tracing can be lifted back as a monadic \hs{Compute} function\footnote{In fact, \Shake has an execution mode that can model \Fabricate{}-like build systems -- see \hs{Development.Shake.Forward} in the \Shake library.}.
+However, in our pure model the mapping is not perfect as \hs{gcc} writes to arbitrary files whose locations are not known in advance.
+
+\subsection{Memoization}
+
+\todo{AM}{Not convined this is really related, but if you think it is, please write something.}
+
+\subsection{Self-adjusting computation}
+
+While not typically considered build systems, self-adjusting computation is a well studied area, and in particular the constrast between different formulations has been well studied.
+\todo{AM}{Say something.}
+
+\subsection{Profunctor Optics}
+
+The \hs{Compute} definition of:
+
+\begin{minted}{haskell}
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+\end{minted}
+
+Looks tanilisingly close to the profunctor optics definition in \cite{gibbons-profunctor-optics}:
+
+\begin{minted}{haskell}
+type Optic c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> f v
+\end{minted}
+
+The properties of such optics are well studied, and the functions like \hs{getDependencies} are very much based on observations from that field of work.
+Alas, we have been unable to remove the \hs{Maybe} used to encode whether a file is an input, without complicating other aspects of our definition.
+Furthermore, the \hs{Build} operation lacks any further such symmetry.

### 2018-03-11 src/Development/Build/Compute.hs
Add a spreadsheet example

@@ -1,5 +1,6 @@
 {-# LANGUAGE ConstraintKinds, RankNTypes #-}
-module Development.Build.Compute where
+{-# OPTIONS_GHC -Wno-unused-top-binds #-}
+module Development.Build.Compute (Compute, inputCompute, isInput) where
 
 import Control.Applicative
 import Control.Monad
@@ -77,6 +78,14 @@ ackermann get (Ackermann m n)
 -- Unlike Collatz and Fibonacci computations, the Ackermann computation cannot
 -- be statically analysed for dependencies. We can only find the first dependency
 -- statically (Ackermann m (n - 1)), but not the second one.
+
+----------------------------- Spreadsheet examples -----------------------------
+data Cell = A1 | A2 | B1 deriving Eq
+
+add :: Compute Applicative Cell Integer
+add fetch key | key /= B1 = Nothing
+              | otherwise = Just $ (+) <$> fetch A1 <*> fetch A2
+
 
 -- These type synonyms are not very useful, but enumerate all interesting cases.
 type FunctorialCompute  k v = forall f. Functor     f => (k -> f v) -> k -> Maybe (f v)

### 2018-03-11 paper/3-compute.tex

@@
 
+We describe build tasks and their dependencies using functions of type
+\hs{Compute}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+\end{minted}
+\vspace{1mm}
+
+\noindent
+A compute takes two arguments: \hs{fetch :: k -> f v} and \hs{key :: k}, and
+returns either \hs{Nothing} to indicate that the \hs{key} is an input and hence
+cannot be computed, or \hs{Just value} obtained by \hs{fetch}ing the
+dependencies of the \hs{key} and performing the required computation. A compute
+must be polymorphic in \hs{f} but may request certain computational
+power\footnote{Readers familiar with \emph{lenses} might recognise a familiar
+pattern. We discuss this in~\S\ref{sec-related}.} by specifying the
+constraint~\hs{c}. Let us look at an example by recalling the following
+spreadsheet from~\S\ref{sec-background-excel}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
+A1 = 10
+A2 = 20
+B1 = A1 + A2
+\end{minted}
+\vspace{1mm}
+
+\noindent
+We can express it as an \emph{applicative compute} as follows:
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Cell = A1 | A2 | B1 deriving Eq
+
+add :: Compute Applicative Cell Integer
+add fetch key | key /= B1 = Nothing
+              | otherwise = Just $ (+) <$> fetch A1 <*> fetch A2
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Cells \cmd{A1} and \cmd{A2} are inputs and \hs{add} returns \hs{Nothing} to
+indicate this. The only task is associated with the cell \cmd{B1}: fetch the
+values of \cmd{A1} and \cmd{A2} and apply the pure function~\hs{(+)} to them
+using the methods of \hs{Applicative} type class.
+
+%\hs{($) :: (a -> b) -> f a -> f b} and \hs{(<*>) :: f (a -> b) -> f a -> f b}
+
 ...
+
 % \begin{figure}
+% \begin{subfigure}[b]{\linewidth}
 % \begin{minted}{haskell}
@@
 % \end{minted}
+% \caption{Compute}
+% \end{subfigure}
+% ~\\
+% \begin{subfigure}[b]{\linewidth}
+% \begin{minted}{haskell}
+% dependency :: Compute Functor k v -> k -> Maybe k
+% dependency compute = fmap getConst . compute Const
+% \end{minted}
+% \caption{Functorial compute}
+% \end{subfigure}
 % \caption{Compute abstractions}\label{fig-compute}
 % \end{figure}
+
 % Consider abstractions in Fig.~\ref{fig-compute}.

### 2018-03-11 src/Development/Build/Compute.hs
Switch to String keys

@@ -80,12 +80,9 @@ ackermann get (Ackermann m n)
 -- statically (Ackermann m (n - 1)), but not the second one.
 
 ----------------------------- Spreadsheet examples -----------------------------
-data Cell = A1 | A2 | B1 deriving Eq
-
-add :: Compute Applicative Cell Integer
-add fetch key | key /= B1 = Nothing
-              | otherwise = Just $ (+) <$> fetch A1 <*> fetch A2
-
+add :: Compute Applicative String Integer
+add fetch key | key /= "B1" = Nothing
+              | otherwise = Just $ (+) <$> fetch "A1" <*> fetch "A2"
 
 -- These type synonyms are not very useful, but enumerate all interesting cases.
 type FunctorialCompute  k v = forall f. Functor     f => (k -> f v) -> k -> Maybe (f v)

### 2018-03-12 paper/3-compute.tex
Rewrite the example with String keys

--- 2018-03-11 paper/3-compute.tex
@@
 
-...
-% \begin{figure}
-% \begin{minted}{haskell}
-% type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-% \end{minted}
-% \caption{Compute abstractions}\label{fig-compute}
-% \end{figure}
-% Consider abstractions in Fig.~\ref{fig-compute}.
+We describe build tasks and their dependencies using functions of type
+\hs{Compute}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+\end{minted}
+\vspace{1mm}
+
+\noindent
+A compute takes two arguments: \hs{fetch :: k -> f v} and \hs{key :: k}, and
+returns either \hs{Nothing} to indicate that the \hs{key} is an input and hence
+cannot be computed, or \hs{Just}~\hs{value} obtained by \hs{fetch}ing the
+dependencies of the \hs{key} and performing the required computation. A compute
+must be polymorphic in \hs{f} but may request certain computational
+power\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
+might recognise a familiar pattern. We discuss this in~\S\ref{sec-related}.} by
+specifying the constraint~\hs{c}. Let us look at an example by recalling the
+following spreadsheet from~\S\ref{sec-background-excel}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
+A1 = 10
+A2 = 20
+B1 = A1 + A2
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The only build task in this spreadsheet can be expressed as an \emph{applicative
+compute}, operating on keys (cell names) of type \hs{String} and values of
+type \hs{Integer}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+add :: Compute Applicative String Integer
+add fetch key | key /= "B1" = Nothing
+              | otherwise   = Just $ (+) <$> fetch "A1" <*> fetch "A2"
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The cells \cmd{A1} and \cmd{A2} are inputs, which is indicated by returning
+\hs{Nothing}. The cell \cmd{B1} is an output, whose value is computed by
+fetching the values of \cmd{A1} and \cmd{A2} and applying the pure
+function~\hs{(+)} using the methods of the \hs{Applicative} type
+class~\cite{mcbride2008applicative}.
+
+The \hs{Compute} type and the example above might seem mysterious. Why insist on
+\hs{Compute} to be polymorphic in \hs{f}? Why choose the \hs{Applicative}
+constraint?
 

### 2018-03-12 paper/7-related.tex
Add labels and a TODO for free constructions

@@
 
-\subsection{Other Build Systems}
+\subsection{Other Build Systems}\label{sec-related-build}
 
@@
 
-\subsection{Memoization}
+\subsection{Memoization}\label{sec-related-memo}
 
@@
 
-\subsection{Profunctor Optics}
+\subsection{Profunctor Optics}\label{sec-related-optics}
 
@@
 \begin{minted}{haskell}
-type Optic c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> f v
+type Optic p a b s t = p a b -> p s t
 \end{minted}
 
+Assuming we instantiate \hs{p} to something that provides \hs{k -> f v} -- which many of the actual instances in that paper do.
 The properties of such optics are well studied, and the functions like \hs{getDependencies} are very much based on observations from that field of work.
@@
 Furthermore, the \hs{Build} operation lacks any further such symmetry.
+
+\subsection{Free applicative functors and monads}\label{sec-related-free}
+
+\todo{AM}{Mention that one could have chosen a free construction to create
+a small DSL for build tasks. Although this simplifies some of the presented
+combinators, in general this seems unnecessary.}

### 2018-03-12 paper/3-compute.tex
Work on S3.2

@@
 and all other complexities of real build systems. We explain the compute type
-in~\S\ref{sec-general-compute}, and then discuss several noteworthy special
-cases: functorial (\S\ref{sec-compute-functor}), applicative
-(\S\ref{sec-compute-applicative}) and monadic (\S\ref{sec-compute-monad}) types
-of compute.
+in~\S\ref{sec-general-compute}, discussing several noteworthy special cases:
+functorial, applicative and monadic types of compute.
 
@@
 us to abstract away from specific application domains, such as software
-development or spreadsheets. We will use this vocabulary throughout the rest of
-the paper.
+development or spreadsheets.
+% We will use this vocabulary throughout the rest of the paper.
 
@@
 
-\subsection{Compute}\label{sec-general-compute}
+\subsection{Compute}\label{sec-compute-applicative}
 
@@
 power\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
-might recognise a familiar pattern. We discuss this in~\S\ref{sec-related}.} by
-specifying the constraint~\hs{c}. Let us look at an example by recalling the
+might recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}
+by specifying the constraint~\hs{c}. Let us look at an example by recalling the
 following spreadsheet from~\S\ref{sec-background-excel}:
@@
 
-\subsection{Functorial compute}\label{sec-compute-functor}
+The answer to both questions is that we can statically analyse an applicative
+compute and extract all dependencies of a key using the following function:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+dependencies :: Compute Applicative k v -> k -> [k]
+dependencies compute = maybe [@@] getConst . compute (\k -> Const [k])
+\end{minted}
+\vspace{1mm}
+
+\noindent
+We can use this function to extract dependencies of \hs{add}, as shown in the
+following GHCi session:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ dependencies add "A1"
+[@@]
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ dependencies add "B1"
+["A1", "A2"]
+\end{minted}
+\vspace{1mm}
+
+\noindent
+It is important to note that we extract the dependencies without providing any
+values to the compute. This guarantees that no actual computation is performed
+during the analysis -- we cannot execute the function \hs{(+)} if we have no
+values for it! Static analysis of applicative computations is not new and has
+been used to similar effect before, e.g. see~\cite{free-applicatives}.
+
+To understand how \hs{dependencies} works, let us examine its type more closely,
+expanding the type synonym \hs{Compute}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+(@\std{forall}@ f. Applicative f => (k -> f v) -> k -> Maybe (f v)) -> k -> [k]
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Since the first argument is polymorphic in \hs{f}, we can choose any \hs{f} that
+satisfies the \hs{Applicative} constraint. A good choice in this case is the
+applicative functor \hs{Const} defined in the standard module
+\cmd{Data.Functor.Const}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+newtype Const m a = Const { getConst :: m }
+
+instance Functor (Const m) where
+    fmap _ (Const m) = Const m
+
+instance Monoid m => Applicative (Const m) where
+    pure _              = Const mempty
+    Const x <*> Const y = Const (x <> y)
+\end{minted}
+\vspace{1mm}
+
+\noindent
+A value of type \hs{Const}~\hs{[}\hs{k]}~\hs{v} contains no values \hs{v} but
+allows us to record all invocations of the custom \hs{fetch} function that we
+feed to the given compute: \hs{fetch k = Const [k]}. The compute returns a
+value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)}, which we unwrap
+using \hs{maybe [@@] getConst} since inputs have no dependencies.
+
+We can also execute a compute on a given store to obtain the resulting value.
+This uses a very similar transformation based on the \hs{Identity} monad
+defined in \cmd{Data.Functor.Identity}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+execute :: Compute Monad k v -> (k -> v) -> k -> Maybe v
+execute compute store = fmap runIdentity . compute (Identity . store)
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Note that \hs{execute} can accept both an applicative as well as a \emph{monadic
+compute} introduced in~\S\ref{sec-compute-monad}. Below we run \hs{execute} on
+the above \hs{add} example. Note that \hs{execute} returns \hs{Nothing} for
+inputs.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store key = if key == "A1" then 10 else 20
+@\ghci@ execute add store "A1"
+Nothing
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ execute add store "B1"
+Just 30
+\end{minted}
+\vspace{1mm}
+
+\subsection{Monadic compute}\label{sec-compute-monad}
+
+\subsection{Compute examples (likely to be removed)}
 
@@
 
-\subsection{Applicative compute}\label{sec-compute-applicative}
 The \emph{generalised Fibonacci sequence} $F_i$ is defined as follows:
@@
 ...
-
-\label{getdependencies} We can define the dependencies for a \hs{Compute Applicative} using the \hs{Const} functor:
-
-\begin{minted}{haskell}
-getDependencies :: Compute Applicative k v -> k -> [k]
-getDependencies comp = maybe [] getConst . comp (\k -> Const [k])
-\end{minted}
-
-\subsection{Monadic compute}\label{sec-compute-monad}
 
@@
 
-\todo{AM}{Turn the above into a formal definition of `correctness' and add a section that provides a mathematical definition of correctness, which does not rely on materialising intermediate values.}
+\todo{AM}{Provide a formal definition of `correctness', which does not rely on materialising intermediate values.}
 

### 2018-03-12 paper/4-build.tex
Work on S3.2

@@
 Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort.
-However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{getdependencies}.
+However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{sec-compute-applicative}.
 

### 2018-03-12 src/Development/Build/Compute.hs
Follow changes in the paper

@@ -82,7 +82,7 @@ ackermann get (Ackermann m n)
 ----------------------------- Spreadsheet examples -----------------------------
 add :: Compute Applicative String Integer
 add fetch key | key /= "B1" = Nothing
-              | otherwise = Just $ (+) <$> fetch "A1" <*> fetch "A2"
+              | otherwise   = Just $ (+) <$> fetch "A1" <*> fetch "A2"
 
 -- These type synonyms are not very useful, but enumerate all interesting cases.
 type FunctorialCompute  k v = forall f. Functor     f => (k -> f v) -> k -> Maybe (f v)

### 2018-03-12 src/Development/Build/Compute/Applicative.hs

@@ -17,7 +17,7 @@ pureCompute f _ = Just . pure . f
 
 -- TODO: Does this always terminate? It's not obvious!
 dependencies :: Compute Applicative k v -> k -> [k]
-dependencies compute = getConst . sequenceA . compute (Const . return)
+dependencies compute = maybe [] getConst . compute (\k -> Const [k])
 
 transitiveDependencies :: Eq k => Compute Applicative k v -> k -> Maybe [k]
 transitiveDependencies compute = reach (dependencies compute)

### 2018-03-12 src/Development/Build/Compute/Monad.hs

@@ -1,7 +1,7 @@
 {-# LANGUAGE FlexibleInstances, GADTs, MultiParamTypeClasses, RankNTypes #-}
 module Development.Build.Compute.Monad (
     dependencies, transitiveDependencies, inputs, acyclic,
-    consistent, correctBuild, runPure, debugPartial, partial, exceptional,
+    consistent, correctBuild, execute, debugPartial, partial, exceptional,
     staticDependencies, Script (..), getScript, runScript, isStatic
     ) where
 
@@ -39,7 +39,7 @@ pureInputs compute f = runIdentity . inputs compute (Identity . f)
 -- | Check that a compute is /consistent/ with a pure lookup function @f@, i.e.
 -- if it returns @Just v@ for some key @k@ then @f k == v@.
 consistent :: Eq v => Compute Monad k v -> (k -> v) -> Bool
-consistent compute f = forall $ \k -> maybe True (f k ==) $ runPure compute f k
+consistent compute f = forall $ \k -> maybe True (f k ==) $ execute compute f k
 
 -- | Given a @compute@, a pair of key-value maps describing the contents of a
 -- store @before@ and @after@ a build system was executed to build a given @key@,
@@ -59,8 +59,8 @@ correctBuild compute before after key =
 
 -- | Run a compute with a pure lookup function. Returns @Nothing@ to indicate
 -- that a given key is an input.
-runPure :: Compute Monad k v -> (k -> v) -> k -> Maybe v
-runPure compute f = fmap runIdentity . compute (Identity . f)
+execute :: Compute Monad k v -> (k -> v) -> k -> Maybe v
+execute compute f = fmap runIdentity . compute (Identity . f)
 
 -- | Run a compute with a partial lookup function. The result @Left k@ indicates
 -- that the compute failed due to a missing dependency @k@. Otherwise, the

### 2018-03-12 paper/3-compute.tex
Move math compute examples to Appendix

@@
 
-\subsection{Compute examples (likely to be removed)}
-
-The \emph{Collatz sequence} $C_i$ is defined as follows:
-
-\[
-C_{i} = {\begin{cases}~n&{\text{for }}i=0\\~f(C_{i-1})&{\text{otherwise}},\end{cases}}\hspace{12pt}\text{where}\hspace{12pt}f(k)={\begin{cases}~k/2&{\text{if }}k\text{ is even}\\~3k+1&{\text{otherwise}}\end{cases}}
-\vspace{2mm}
-\]
-\noindent
-and $n$ is a positive integer parameter. The famous \emph{Collatz conjecture}
-states that the Collatz sequence eventually reaches 1 for all possible values of
-$n$. For example, if $n=6$, we reach 1 in eight steps:
-$(6, 3, 10, 5, 16, 8, 4, 2, 1, \dots)$, after which the sequence loops forever:
-$(4, 2, 1, 4, 2, 1, \dots)$.
-
-We can express the computation of values in the Collatz sequence as a functorial
-compute:
-
-\begin{minted}[xleftmargin=10pt]{haskell}
-data Collatz = Collatz Int
-
-collatz :: Compute Functor Collatz Int
-collatz get (Collatz k) | k <= 0    = Nothing
-                        | otherwise = Just $ f <$> get (Collatz (k - 1))
-  where
-    f n | even n    = n `div` 2
-        | otherwise = 3 * n + 1
-\end{minted}
-
-...
-
-The \emph{generalised Fibonacci sequence} $F_i$ is defined as follows:
-
-\[
-F_{i} = {\begin{cases}~n&{\text{for }}i=0\\~m&{\text{for }}i=1\\~F_{i-1}+F_{i-2}&{\text{otherwise}}\end{cases}}
-\vspace{2mm}
-\]
-\noindent
-where $n$ and $m$ are integer parameters. By setting $n=0$ and $m=1$ we obtain
-the famous \emph{Fibonacci sequence}: $(0, 1, 1, 2, 3, 5, 8, 13, \dots$), and if
-$n=2$ and $m=1$, the result is the \emph{Lucas sequence}:
-$(2, 1, 3, 4, 7, 11, 18, 29, \dots)$.
-
-We can express the computation of values in the generalised Fibonacci sequence
-as an applicative compute:
-
-\begin{minted}[xleftmargin=10pt]{haskell}
-data Fibonacci = Fibonacci Int
-
-fibonacci :: Compute Applicative Fibonacci Int
-fibonacci get (Fibonacci k) | k <= 1    = Nothing
-                            | otherwise = Just $ (+) <$> get (Fibonacci (k - 1))
-                                                     <*> get (Fibonacci (k - 2))
-\end{minted}
-
-...
-
-The \emph{Ackermann function} $A(m, n)$ is defined as follows:
-
-\[
-A(m, n) = {\begin{cases}~n+1&{\text{for }}m=0\\~A(m-1, 1)&{\text{for }}n=0\\~A(m-1,A(m,n-1))&{\text{otherwise}}\end{cases}}
-\vspace{2mm}
-\]
-\noindent
-We can express the computation of the Ackermann function as a monadic compute:
-
-\begin{minted}[xleftmargin=10pt]{haskell}
-data Ackermann = Ackermann Int Int
-
-ackermann :: Compute Monad Ackermann Int
-ackermann get (Ackermann m n)
-    | m < 0 || n < 0 = Nothing
-    | m == 0    = Just $ return (n + 1)
-    | n == 0    = Just $ get (Ackermann (m - 1) 1)
-    | otherwise = Just $ do
-        index <- get (Ackermann m (n - 1))
-        get (Ackermann (m - 1) index)
-\end{minted}
-
 ...

### 2018-03-12 paper/9-appendix.tex
Move math compute examples to Appendix

--- 2018-03-09 paper/abstractions.tex
@@
-\section{Build systems, abstractly}\label{sec-abstractions}
+\clearpage
+\section{Appendix}\label{sec-appendix}
 
-This section presents purely functional abstractions for build systems that
-allow us to express all the intricacies of build systems discussed in the
-previous section~\S\ref{sec-background} and design complex build systems from
-simple primitives. Specifically, we present the \emph{compute} and \emph{build}
-abstractions:
-
-\begin{minted}[xleftmargin=10pt]{haskell}
-type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
-\end{minted}
-
-\noindent
-Compute represents build rules (in build systems) or formulas (in spreadsheets),
-and is completely isolated from the world of compilers, file systems, memories,
-caches, and all other complexities of real build systems. We explain the compute
-type in~\S\ref{sec-compute}, and then discuss several noteworthy special cases:
-functorial (\S\ref{sec-compute-functor}), applicative
-(\S\ref{sec-compute-applicative}) and monadic (\S\ref{sec-compute-monad}) types
-of compute.
-
-Build corresponds to the algorithm used for bringing a key-value store to a
-consistent state by recomputing dirty values in an appropriate order. We explain
-the build type in~\S\ref{sec-build}, and then provide implementations for several
-build systems in~\S\ref{sec-examples}.
-
-\subsection{Compute}\label{sec-compute}
-
-...
-% \begin{figure}
-% \begin{minted}{haskell}
-% type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-% \end{minted}
-% \caption{Compute abstractions}\label{fig-compute}
-% \end{figure}
-% Consider abstractions in Fig.~\ref{fig-compute}.
-
-\subsection{Functorial compute}\label{sec-compute-functor}
+\subsection{Compute examples}
 
@@
 
-\subsection{Monadic compute}\label{sec-compute-monad}
-
 The \emph{Ackermann function} $A(m, n)$ is defined as follows:
@@
 \end{minted}
-
-...
-
-\subsection{Build}\label{sec-build}
-
-...

### 2018-03-12 paper/3-compute.tex
Minor revision (typos, TODOs)

@@
 and all other complexities of real build systems. We explain the compute type
-in~\S\ref{sec-general-compute}, discussing several noteworthy special cases:
+in~\S\ref{sec-compute-applicative}, discussing several noteworthy special cases:
 functorial, applicative and monadic types of compute.
@@
 We can also execute a compute on a given store to obtain the resulting value.
-This uses a very similar transformation based on the \hs{Identity} monad
+This uses a similar transformation based on the \hs{Identity} monad
 defined in \cmd{Data.Functor.Identity}:

### 2018-03-12 paper/5-implementations.tex
Minor revision (typos, TODOs)

@@
 
-\todo{AM}{AM: What about using shorter \hs{get} and \hs{put}?
-I can introduce it right in the common vocabulary, mentioning the usual get-put laws.
-I would also remove `get' prefixes from many functions, e.g.
-\hs{dependencies compute} reads fine.
-I suggest not to rush with this TODO -- let me first finish the Compute section, where many of these helper functions will be defined.
-NM: I think this is a bad idea - conflicts with a state Monad which is what we're using underneath so will confuse some readers.}
-
 \subsection{\Make}\label{sec-implementation-make}
@@
 
-We'll come back to \hs{withChanged} afterwards, but ignoring that, reading the rest of the implementation you can see the underying essence of \Make.
+We'll come back to \hs{withChanged} afterwards, but ignoring that, reading the rest of the implementation you can see the underlying essence of \Make.
 We use the helper function \hs{topological} which gets called with each file in a dependency-respecting order given the file (\hs{k}), the dependencies (\hs{dk}) and an action to generate the result (\hs{act}).
@@
 topological step compute k = runM $ do
-    let depends = getDependencies compute
+    let depends = dependencies compute
     forM_ (topSort depends $ transitiveClosure depends k) $ \k ->
@@
 
-It uses a \hs{getDependencies} on the computation function, takes the transitive-closure and topological-sort to provide an order in which to build the keys.
+It uses a \hs{dependencies} on the computation function, takes the transitive-closure and topological-sort to provide an order in which to build the keys.
 For each key, if it is an input it does nothing, otherwise it calls the supplied \hs{step} function.
-Note that \hs{getDependencies} is only defined for \hs{Applicative} computes, which is what restricts \Make to Applicative build systems.
+Note that \hs{dependencies} is only defined for \hs{Applicative} computes, which is what restricts \Make to Applicative build systems.
 Moreover, anyone following the same topological-sort approach will also have the \hs{Applicative} restriction.
 
-The greatest impedence mismatch between our framework and this implementation is the notion of \hs{Time} in \hs{getStoreTime}, which is missing from \hs{Build}.
+The greatest impedance mismatch between our framework and this implementation is the notion of \hs{Time} in \hs{getStoreTime}, which is missing from \hs{Build}.
 However, as we discussed in \S\ref{sec-dirty-bit}, \Make is really using modification-time to encode a dirty bit with specific properties.
@@
 The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Compute} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway.
-If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possiblities:
+If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possibilities:
 
@@
 The astute reader may notice that \hs{reordering} supplies extra information that is ignored with \hs{_} patterns.
-While not required for \Excel, we have implemented build systems using \hs{reordering} which use verifying and constructive traces, effectively turning \Excel into a cloud build system and ensuring \hs{reordering} is not overly fitted to \Excel alone (additional implementations available in the suplementary material).
+While not required for \Excel, we have implemented build systems using \hs{reordering} which use verifying and constructive traces, effectively turning \Excel into a cloud build system and ensuring \hs{reordering} is not overly fitted to \Excel alone (additional implementations available in the supplementary material).
 The definition of \hs{reordering} is:
@@
 \item \hs{Maybe [k]} - the list of dependencies, or \hs{Nothing} if the step makes use of monadic dependencies.
-Note that above if a step is truely monadic it is always rebuilt.
+Note that above if a step is truly monadic it is always rebuilt.
 In the case of spreadsheets such steps are very rare, so a very conservative approximation can be used.\footnote{In truth, many spreadsheets take the alternative approximation of always assuming monadic steps \textit{did not} change - trading correctness for performance.}
@@
 
-This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to fetch the result of a dependecy (\hs{fetch}), and a way to build \hs{k} (\hs{act}).
+This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to fetch the result of a dependency (\hs{fetch}), and a way to build \hs{k} (\hs{act}).
 The code first looks up the traces that match the current situation, creating a list of possible results as \hs{poss}, using the helper \hs{traceMatch}.
@@
 
-The required \hs{step} function gets given the \hs{k} to build, the dependencies as \hs{Maybe [k]} (where they are \hs{Nothing} if the rule is truely monadic), a function to demand the result of a dependent key, and a function to compute the current value and return the actual dependencies.
+The required \hs{step} function gets given the \hs{k} to build, the dependencies as \hs{Maybe [k]} (where they are \hs{Nothing} if the rule is truly monadic), a function to demand the result of a dependent key, and a function to compute the current value and return the actual dependencies.
 To ensure each file is only visited once in a single execution a piece of temporary state is used with \hs{getTemp} and \hs{modifyTemp} to record which files were visited - this state is not persisted to the \hs{i} info parameter and is reset on each fresh execution of \hs{M}.
@@
 
-Now we have seen the three dependency schemes, we can directly reuse \hs{topological} to define Bazel.
-Furthermore, as Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type.
+Now we have seen the three dependency schemes, we can directly reuse \hs{topological} to define \Bazel.
+Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type.
 We define \Bazel as:
@@
 If there are none we rebuild and record the trace, much like \Shake, but also recording the value of the result.
-If there are some matching traces, and the current value of \hs{k} is not compatible, we grab the first possiblity and grab its value from \hs{contents}.
+If there are some matching traces, and the current value of \hs{k} is not compatible, we grab the first possibility and grab its value from \hs{contents}.
 
@@
 Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash.
-In this formulation we have removed the possibilty for a single key/dependency state to map to multiple different hashes, e.g.
-on a non-determinstic build - something \Bazel already prohibits which is discussed more in \S\ref{sec-non-determinism}.
+In this formulation we have removed the possibility for a single key/dependency state to map to multiple different hashes, e.g.
+on a non-deterministic build - something \Bazel already prohibits which is discussed more in \S\ref{sec-non-determinism}.
 \item Finally, a \hs{Monad} system with constructive traces can be efficiently encoded as \hs{Map k (Choice k v)}, assuming a definition of \hs{Choice} as:

### 2018-03-12 paper/6-engineering.tex
Minor revision (typos, TODOs)

@@
 
-Interestingly, there is signficant interplay between all three sources of impurity.
+Interestingly, there is significant interplay between all three sources of impurity.
 Systems like \Bazel use various sandboxing techniques to guard against missing dependencies, but none are likely to capture all dependencies right down to CPU model and microcode version.
@@
 
-The build systems that have precise tracking of self-computation are all onces which use a non-Embedded Domain Specific Language to describe computations.
+The build systems that have precise tracking of self-computation are all ones which use a non-Embedded Domain Specific Language to describe computations.
 Those which make use of a full programming language, e.g.
-\Shake, are much more difficult to implement equality on rules as they are arbitrary functions making use of opqaue abstractions.
+\Shake, are much more difficult to implement equality on rules as they are arbitrary functions making use of opaque abstractions.
 For such build systems the incredibly pessimistic assuming of saying that any change to the build system potentially changes any rule can often be used -- the classic example being a \texttt{Makefile} depending on itself.
@@
 For examples like \LaTeX we consider the proper encoding to not be with circular rules, but with a series of iterative steps, as described in \S\ref{neil-iterative-on-stack-overflow}.
-It is important that the number of executions is bounded, otherwise the build system may not terminate (a legitimate concern with \LaTeX which can be put into a situation where it is bistable or divering over multiple executions).
+It is important that the number of executions is bounded, otherwise the build system may not terminate (a legitimate concern with \LaTeX which can be put into a situation where it is bistable or diverging over multiple executions).
 
@@
 Our build system abstraction assumes a \hs{k -> v} store, along with a build system that works directly on \hs{k} and \hs{v} values.
-However, certain build systems provide greater flexiblity, e.g.
+However, certain build systems provide greater flexibility, e.g.
 \Shake permits polymorphic keys and values, allowing types that are only stored in the Shake information, and never persisted to the Store.

### 2018-03-12 paper/7-related.tex
Minor revision (typos, TODOs)

@@
 \item \Redo \cite{redo} almost exactly matches \Shake at the level of detail given here, differing only on aspects like polymorphic dependencies \S\ref{sec-polymorphism}.
-\item \Buck \cite{buck} and \CloudBuild \cite{cloudbuild} are both fundamentally similar to \Bazel, differing around areas such as non-determinism \S\ref{sec-sec-non-determinism}.
+\item \Buck \cite{buck} and \CloudBuild \cite{cloudbuild} are both fundamentally similar to \Bazel, differing around areas such as non-determinism \S\ref{sec-non-determinism}.
 \item \Nix \cite{nix} has coarse-grained monadic dependencies, with precise hashing of dependencies and downloading of precomputed build products, so may be, at some level of abstraction, the same as the hypothetical Cloud Shake from \S\ref{sec-implementation-cloud-shake}.
-However, \Nix is not intended as a build system, and the coarse grained nature (packages, not individual files) makes it targetted to a different purpose.
+However, \Nix is not intended as a build system, and the coarse grained nature (packages, not individual files) makes it targeted to a different purpose.
 \todo{NM}{Read comment below when not on an airplane.}
@@
 % \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.
-\item \Pluto \cite{pluto} has power to express cyclic dependenies, but that is implemented \todo{NM}{Read the paper so I can say something precise}
+\item \Pluto \cite{pluto} has power to express cyclic dependencies, but that is implemented \todo{NM}{Read the paper so I can say something precise}
 \end{itemize}
@@
 
-While not typically considered build systems, self-adjusting computation is a well studied area, and in particular the constrast between different formulations has been well studied.
+While not typically considered build systems, self-adjusting computation is a well studied area, and in particular the contrast between different formulations has been well studied.
 \todo{AM}{Say something.}
@@
 
-Looks tanilisingly close to the profunctor optics definition in \cite{gibbons-profunctor-optics}:
+Looks tantalisingly close to the profunctor optics definition in \cite{gibbons-profunctor-optics}:
 
@@
 Assuming we instantiate \hs{p} to something that provides \hs{k -> f v} -- which many of the actual instances in that paper do.
-The properties of such optics are well studied, and the functions like \hs{getDependencies} are very much based on observations from that field of work.
+The properties of such optics are well studied, and the functions like \hs{dependencies} are very much based on observations from that field of work.
 Alas, we have been unable to remove the \hs{Maybe} used to encode whether a file is an input, without complicating other aspects of our definition.

### 2018-03-12 src/Development/Build/Compute/Monad.hs
Add trackDependencies

@@ -1,6 +1,6 @@
 {-# LANGUAGE FlexibleInstances, GADTs, MultiParamTypeClasses, RankNTypes #-}
 module Development.Build.Compute.Monad (
-    dependencies, transitiveDependencies, inputs, acyclic,
+    dependencies, trackDependencies, transitiveDependencies, inputs, acyclic,
     consistent, correctBuild, execute, debugPartial, partial, exceptional,
     staticDependencies, Script (..), getScript, runScript, isStatic
     ) where
@@ -20,6 +20,11 @@ dependencies :: Monad m => Compute Monad k v -> (k -> m v) -> k -> m [k]
 dependencies compute get = execWriterT . sequenceA . compute tracingGet
   where
     tracingGet k = tell [k] >> lift (get k)
+
+trackDependencies :: Monad m => Compute Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
+trackDependencies compute fetch = fmap runWriterT . compute tracingFetch
+  where
+    tracingFetch k = tell [k] >> lift (fetch k)
 
 transitiveDependencies :: (Eq k, Monad m) => Compute Monad k v
                                           -> (k -> m v) -> k -> m (Maybe [k])

### 2018-03-12 paper/3-compute.tex

@@
 \noindent
-\hs{Compute} represents an abstract set of \emph{build tasks}, for example, build
-rules in software build systems or formulas in spreadsheets. It is completely
-isolated from the world of compilers, file systems, calculation chains, caches,
-and all other complexities of real build systems. We explain the compute type
-in~\S\ref{sec-compute-applicative}, discussing several noteworthy special cases:
-functorial, applicative and monadic types of compute.
+\hs{Compute} represents an abstract collection of \emph{build tasks}, for
+example, build rules in software build systems or formulas in spreadsheets. It
+is completely isolated from the world of compilers, calculation chains, file
+systems, caches, and all other complexities of real build systems. We explain
+the compute type in~\S\ref{sec-compute-applicative}, discussing several
+noteworthy special cases: functorial, applicative and monadic types of compute.
 
@@
 satisfies the \hs{Applicative} constraint. A good choice in this case is the
-applicative functor \hs{Const} defined in the standard module
+applicative functor \hs{Const} defined in GHC's standard library module
 \cmd{Data.Functor.Const}:
@@
 feed to the given compute: \hs{fetch k = Const [k]}. The compute returns a
-value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)}, which we unwrap
-using \hs{maybe [@@] getConst} since inputs have no dependencies.
+value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)}, which we can unwrap
+using \hs{maybe [@@] getConst}, i.e. returning the empty list of dependencies
+in case of an input.
 
-We can also execute a compute on a given store to obtain the resulting value.
-This uses a similar transformation based on the \hs{Identity} monad
-defined in \cmd{Data.Functor.Identity}:
+We can also execute a compute on a given store. This relies on a similar
+transformation, now based on the \hs{Identity} monad defined in
+\cmd{Data.Functor.Identity}:
 
@@
 
-...
+As discussed in~\S\ref{sec-background}, some build tasks have dynamic
+dependencies, which are determined by values of intermediate computations. In
+our framework, such build tasks correspond to \emph{monadic compute}, i.e.
+the type \hs{Compute}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
+example:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
+A1 = 10
+A2 = 20
+B1 = IF(C1=1,A1,A2)
+C1 = 1
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
+dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
+our compute abstraction as follows\footnote{The spreadsheet example that uses
+the \hs{INDIRECT} function can be expressed as well: simply replace the last
+line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+select :: Compute Monad String Integer
+select fetch key | key /= "B1" = Nothing
+                 | otherwise   = Just $ do
+                     c1 <- fetch "C1"
+                     if c1 == 1 then fetch "A1" else fetch "A2"
+\end{minted}
+\vspace{1mm}
+
+We cannot statically analyse \hs{select} using the function \hs{dependencies}
+from the previous section, but we can still \hs{execute} it on a given store:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store key = if key == "A1" then 10 else 20
+@\ghci@ execute select store "B1"
+Just 20
+\end{minted}
+\vspace{1mm}
+
+% Note that this presentation of build tasks makes dynamic dependencies
+% immediately recognisable: we need to examine the actual value \hs{c1} by
+% extracting it from the computational context using \hs{Monad}'s bind statement
+% \hs{c1 <- fetch}
 
 \subsection{Build}\label{sec-general-build}
-
-...
 

### 2018-03-12 paper/3-compute.tex
Expand the Indentity monad construction

@@
 satisfies the \hs{Applicative} constraint. A good choice in this case is the
-applicative functor \hs{Const} defined in GHC's standard library module
-\cmd{Data.Functor.Const}:
+applicative functor \hs{Const}:
 
@@
 newtype Const m a = Const { getConst :: m }
-
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 instance Functor (Const m) where
     fmap _ (Const m) = Const m
-
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 instance Monoid m => Applicative (Const m) where
@@
 
-We can also execute a compute on a given store. This relies on a similar
-transformation, now based on the \hs{Identity} monad defined in
-\cmd{Data.Functor.Identity}:
+We can also \hs{execute} a compute on a given store. This relies on a similar
+transformation, now based on the \hs{Identity} monad and feeding
+\hs{fetch k = pure (store k)} to the compute:
 
@@
 execute :: Compute Monad k v -> (k -> v) -> k -> Maybe v
-execute compute store = fmap runIdentity . compute (Identity . store)
+execute compute store = fmap runIdentity . compute (pure . store)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+newtype Identity a = Identity { runIdentity :: a }
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+instance Functor Identity where
+    fmap f (Identity a) = Identity (f a)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+instance Applicative Identity where
+    pure a = Identity a
+    Identity f <*> Identity a = Identity (f a)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+instance Monad Identity where
+    Identity a >>= f = f a
 \end{minted}
@@
 \end{minted}
-\vspace{1mm}
+% \vspace{1mm}
 

### 2018-03-12 paper/3-compute.tex
Move Identity monad to Appendix

@@
 \begin{minted}[xleftmargin=10pt]{haskell}
+import Data.Functor.Const
+\end{minted}
+\vspace{0.5mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 dependencies :: Compute Applicative k v -> k -> [k]
@@
 \end{minted}
-\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-\vspace{1mm}
+\vspace{0.5mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 feed to the given compute: \hs{fetch k = Const [k]}. The compute returns a
-value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)}, which we can unwrap
-using \hs{maybe [@@] getConst}, i.e. returning the empty list of dependencies
-in case of an input.
+value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)} that can be
+unwrapped using \hs{maybe [@@] getConst}, i.e. returning the empty list of
+dependencies in case of an input.
 
 We can also \hs{execute} a compute on a given store. This relies on a similar
-transformation, now based on the \hs{Identity} monad and feeding
-\hs{fetch k = pure (store k)} to the compute:
+transformation, now based on the \hs{Identity} monad (see
+Appendix~\S\ref{sec-appendix-transformers} for the implementation of
+\hs{Identity}):
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
+import Data.Functor.Identity
+\end{minted}
+\vspace{0.5mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 execute :: Compute Monad k v -> (k -> v) -> k -> Maybe v
 execute compute store = fmap runIdentity . compute (pure . store)
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-newtype Identity a = Identity { runIdentity :: a }
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-instance Functor Identity where
-    fmap f (Identity a) = Identity (f a)
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-instance Applicative Identity where
-    pure a = Identity a
-    Identity f <*> Identity a = Identity (f a)
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-instance Monad Identity where
-    Identity a >>= f = f a
 \end{minted}
@@
 \end{minted}
-% \vspace{1mm}
 
@@
 
+Now we need a combination of \hs{Const} and \hs{Identity}, which is the
+\hs{Writer} monad!
+
 \subsection{Build}\label{sec-general-build}

### 2018-03-12 paper/9-appendix.tex
Move Identity monad to Appendix

@@
 
-\subsection{Compute examples}
+\subsection{Compute transformers}\label{sec-appendix-transformers}
+
+In this section we clarify some of the compute transformers used in this paper.
+
+\hs{execute} uses the transformation based on the \hs{Identity} monad, feeding
+\hs{fetch k = pure (store k)} to the compute:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+execute :: Compute Monad k v -> (k -> v) -> k -> Maybe v
+execute compute store = fmap runIdentity . compute (pure . store)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+newtype Identity a = Identity { runIdentity :: a }
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+instance Functor Identity where
+    fmap f (Identity a) = Identity (f a)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+instance Applicative Identity where
+    pure a = Identity a
+    Identity f <*> Identity a = Identity (f a)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+instance Monad Identity where
+    Identity a >>= f = f a
+\end{minted}
+\vspace{1mm}
+
+\todo{AM}{Explain \hs{trackDependencies}.}
+
+\subsection{Compute examples}\label{sec-appendix-compute-examples}
+
+\todo{AM}{Add some explanatory text.}
 

### 2018-03-12 src/Development/Build/Compute/Monad.hs
Rename trackDependencies to track

@@ -1,8 +1,8 @@
 {-# LANGUAGE FlexibleInstances, GADTs, MultiParamTypeClasses, RankNTypes #-}
 module Development.Build.Compute.Monad (
-    dependencies, trackDependencies, trackDependenciesM, transitiveDependencies,
-    inputs, acyclic, consistent, correctBuild, execute, debugPartial, partial,
-    exceptional, staticDependencies, Script (..), getScript, runScript, isStatic
+    dependencies, track, trackM, transitiveDependencies, inputs, acyclic,
+    consistent, correctBuild, execute, debugPartial, partial, exceptional,
+    staticDependencies, Script (..), getScript, runScript, isStatic
     ) where
 
 import Control.Monad.Trans
@@ -21,13 +21,11 @@ dependencies compute store = execWriterT . sequenceA . compute fetch
   where
     fetch k = tell [k] >> lift (store k)
 
-trackDependencies :: Compute Monad k v -> (k -> v) -> k -> Maybe (v, [k])
-trackDependencies compute store = fmap runWriter . compute fetch
-  where
-    fetch k = tell [k] >> lift (pure (store k))
+track :: Compute Monad k v -> (k -> v) -> k -> Maybe (v, [k])
+track compute store = fmap runWriter . compute (\k -> writer (store k, [k]))
 
-trackDependenciesM :: Monad m => Compute Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
-trackDependenciesM compute store = fmap runWriterT . compute fetch
+trackM :: Monad m => Compute Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
+trackM compute store = fmap runWriterT . compute fetch
   where
     fetch k = tell [k] >> lift (store k)
 

### 2018-03-12 paper/3-compute.tex
Add description of track

@@
 our compute abstraction as follows\footnote{The spreadsheet example that uses
-the \hs{INDIRECT} function can be expressed as well: simply replace the last
-line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:
+the \hs{INDIRECT} function can be expressed very similarly: simply replace the
+last line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:
 
@@
 
-% Note that this presentation of build tasks makes dynamic dependencies
-% immediately recognisable: we need to examine the actual value \hs{c1} by
-% extracting it from the computational context using \hs{Monad}'s bind statement
-% \hs{c1 <- fetch}
+\noindent
+In the above execution $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
+To extract the dependencies of \hs{select} we introduce the function \hs{track}:
+a combination of \hs{execute} and \hs{dependencies} that takes the store~\store
+as input and computes both the resulting value and the list of its dependencies:
 
-Now we need a combination of \hs{Const} and \hs{Identity}, which is the
-\hs{Writer} monad!
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+import Control.Monad.Writer
+\end{minted}
+\vspace{0.5mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+track :: Compute Monad k v -> (k -> v) -> k -> Maybe (v, [k])
+track compute store = fmap runWriter . compute (\k -> writer (store k, [k]))
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
+we feed \hs{fetch k = writer (store k, [k])} to the compute, which in addition
+to fetching a value from the store, tracks the corresponding key. We can use
+\hs{track} to track both applicative and monadic compute:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store key = if key == "A1" then 10 else 20
+@\ghci@ track add store "B1"
+Just (30,["A1","A2"])
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ track select store "B1"
+Just (20,["C1","A2"])
+\end{minted}
+\vspace{1mm}
 

### 2018-03-13 paper/5-implementations.tex
Rename trackDependencies to track

@@
                 modifyTemp $ Set.insert k set
-                case trackDependencies compute fetch k of
+                case track compute fetch k of
                     Nothing -> return ()

### 2018-03-13 paper/3-compute.tex
Add correctness definition

@@
 \noindent
-In the above execution $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
+In the example above $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
 To extract the dependencies of \hs{select} we introduce the function \hs{track}:
@@
 we feed \hs{fetch k = writer (store k, [k])} to the compute, which in addition
-to fetching a value from the store, tracks the corresponding key. We can use
-\hs{track} to track both applicative and monadic compute:
+to fetching a value from the store, tracks the corresponding key. We can
+\hs{track} both applicative and monadic compute:
 
@@
 \vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store' key = if key == "A1" then 10 else 1
+@\ghci@ track select store' "B1"
+Just (10,["C1","A1"])
+\end{minted}
+\vspace{1mm}
+
+\noindent
+As expected, the dependencies of \hs{select} are determined by the contents of
+the store.
+
+% \subsection{Recognising an applicative compute wearing a monadic hat}
+
+% Although dynamic dependencies are very useful, they typically represent only a
+% small fraction of all dependencies. One can exploit this in a build system by
+% recognising applicative computations even though they have the type
+% \hs{Compute}~\hs{Monad} and treating them specially.
+
+% Consider the following data type.
+
+% \subsection{Free build structures}\label{sec-free-build}
+
+% \begin{minted}{haskell}
+% data Depend k v r = Depend [k] ([v] -> r)
+
+% data Depends k v r = Depends [k] ([v] -> Depends k v r)
+%                    | Done r
+% \end{minted}
 
@@
 
-\todo{AM}{Provide a formal definition of `correctness', which does not rely on materialising intermediate values.}
+Equipped with the \hs{Compute} abstraction, we can now introduce \hs{Build} in
+more detail. Recall the informal definition of a build system
+from~\S\ref{sec-background}:
 
-\subsection{Free build structures}\label{sec-free-build}
+\vspace{2mm}
+\emph{Definition (Build system).}\hspace{1.5mm}Given a description of tasks, a
+\emph{build system} executes them in an order that respects their dependencies
+to produce a specified output from supplied inputs.
+\vspace{2mm}
 
-\begin{minted}{haskell}
-data Depend k v r = Depend [k] ([v] -> r)
+\noindent
+To make the definition more precise, we express it as the following function:
 
-data Depends k v r = Depends [k] ([v] -> Depends k v r)
-                   | Done r
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
 \end{minted}
+\vspace{1mm}
+
+\noindent
+The type closely matches the above definition: \hs{Compute c k v} is `a
+description of tasks', a key~\hs{k} is~`a~specified output', and the store
+\store holds `supplied inputs'. The additional input parameter \hs{Maybe i}
+may provide auxiliary information from the previous build, as discussed
+in~\S\ref{sec-background-summary}. The build system produces auxiliary
+information \hs{i} for the next build as well as an updated store, where the
+output \hs{k} must be up-to-date. We formally capture the correctness of the
+resulting store in the following definition.
+
+\definition[Correctness]{Let \hs{build :: Build c i k v} be a build system,
+\hs{compute} describe an acyclic collection of build tasks, \hs{key} be an output
+key, \hs{mi :: Maybe i} be auxiliary information, and \hs{store} be a store.
+Then the \hs{result}ing store \hs{(_, result) = build compute k mi store}
+is~\emph{correct} if there exists an \hs{ideal} store, such that
+
+\begin{itemize}
+    \item The \hs{ideal} store is \emph{consistent} with the \hs{compute}, i.e.
+    for all possible non-input keys \hs{k}, the result of executing the
+    \hs{compute} matches the value in the \hs{ideal} store:
+    \[
+    \hs{execute}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{ideal}~\hs{k)}.
+    \]
+    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.
+    \hs{result}~\hs{key}~\hs{==}~\hs{ideal}~\hs{key}. In other words, we require
+    that the output \hs{key} has the correct value, but do not impose any
+    restrictions on intermediate keys, therefore permitting shallow cloud builds.
+    \item The \hs{store}, \hs{ideal} and \hs{result} all agree on the inputs
+    that belong to the transitive closure of \hs{key}'s dependencies. This means
+    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
+    store has constraints not only on the output, but also on the inputs.
+\end{itemize}
+}
+\label{def-correct}

### 2018-03-13 paper/3-compute.tex
Stuff from Simon

@@
-\section{Build tasks, abstractly}\label{sec-compute}
+\section{Build systems, abstractly}\label{sec-compute}
 
@@
 primitives. Specifically, we present the \emph{compute} and \emph{build}
-abstractions:
+abstractions, which are described in
+Sections~\ref{sec-task} and \ref{sec-general-build} respectively.
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
+type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
 \end{minted}
 \vspace{1mm}
-
-\noindent
-\hs{Compute} represents an abstract collection of \emph{build tasks}, for
-example, build rules in software build systems or formulas in spreadsheets. It
-is completely isolated from the world of compilers, calculation chains, file
-systems, caches, and all other complexities of real build systems. We explain
-the compute type in~\S\ref{sec-compute-applicative}, discussing several
-noteworthy special cases: functorial, applicative and monadic types of compute.
 
@@
 
-We start by introducing a common vocabulary for build systems that allows
+\begin{figure}
+\begin{minted}{haskell}
+-- Stores
+data Store i k v   -- Abstract
+getInfo  :: Store i k v -> i
+putInfo  :: Store i k v -> i -> Store i k v
+getValue :: Store i k v -> k -> v
+putValue :: Hashable v => Store i k v -> k -> v -> Store i k v
+getHash  :: Hashable v => Store i k v -> k -> Hash
+
+-- Hashing
+hash :: Hashable a => a -> Hash
+
+-- Applicative
+(<*>) :: Applicative f => f (a->b) -> f a -> f b  -- Left-associative
+(<$>) :: Applicative f => (a->b) -> f a -> f b    -- Left-associative
+pure  :: Applicative f => a -> f a
+
+-- State monad
+data State s
+instance Applicative State
+instance Monad       State
+evalState :: State s a -> s -> s
+getState  :: State s s
+\end{minted}
+\caption{Signatures of library functions}
+\label{fig-lib-funs}
+\end{figure}
+
+We start by establishing a common vocabulary for build systems that allows
 us to abstract away from specific application domains, such as software
@@
 
-\emph{Keys} are used to distinguish \emph{values}. In software build systems,
-keys are typically filenames, e.g.~\cmd{main.c}, whereas values are file
-contents (a C program source code in this case). In spreadsheets, keys are cell
-names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
-displayed inside cells. In Haskell code, we will use type variables \hs{k}
-and \hs{v} to denote keys and values, respectively.
+A \emph{build system} takes a \emph{task description}, a target \emph{key},
+and a \emph{store}, and returns a new store in which the target key is
+up to date.  The store is a key/value mapping, togther with
+\emph{persistent build state} that is maintained by the build system itself.
 
-We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
-tracking and sharing of~build results, where \hs{Hash} is an abstract data type
-equipped with the equality test.
+In software build systems, keys are typically filenames,
+e.g.~\cmd{main.c}, and values are file contents (a C program source
+code in this case). In spreadsheets, keys are cell names,
+e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
+displayed inside cells. In our Haskell code, we will use type
+variables \hs{k} and \hs{v} to denote keys and values, respectively.
 
-A \emph{store} associates keys to values. It is convenient to assume that a
+Some values must be provided by the user as \emph{input}. For example,
+\cmd{main.c} can be edited by the user who relies on the build system to
+compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
+such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
+\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
+but are produced in the process of turning inputs into outputs.
+
+The types of the library functions we use in this paper is
+given in Figure~\ref{fig-lib-funs}.
+We use a cryptographic \emph{hash function} \hs{hash :: Hashable v =>
+v -> Hash} for efficient tracking and sharing of~build results, where
+\hs{Hash} is an abstract data type equipped with the equality test.
+
+A \emph{store}, of type \hs{Store i k v}, associates keys to values.
+It is convenient to assume that a
 store is total function \store, i.e. it contains a value for every possible key.
@@
 
-Some values must be provided by the user as \emph{input}. For example,
-\cmd{main.c} can be edited by the user who relies on the build system to
-compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
-such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
-\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
-but are produced in the process of turning inputs into outputs.
+The store also contains a single persistent value, of type \hs{i}, representing
+the persistent state maintained by the build system itself -- its ``memory''.
 
-\subsection{Compute}\label{sec-compute-applicative}
+\subsection{The Task abstraction}\label{sec-task}
 
-We describe build tasks and their dependencies using functions of type
-\hs{Compute}:
-
+Our first key idea is the task-description abstraction:
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+\end{minted}
+This highly-abstracted type is best introduced through an example.
+Consider this Excel spreadsheet (c.f. \S\ref{sec-background-excel}):
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
+  A1: 10     B1: =A1 + A2
+  A2: 20     B2: =B1 * 2
+\end{minted}
+\vspace{1mm}
+Here cell \hs{A1} contains the value \hs{10} while cell \hs{B1} contains
+the formula \hs{=A1+A2}.
+We can represent the formulae of of this spreadsheet
+with the following task description
 \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+sprsh1 :: Compute Applicative String Integer
+sprsh1 get "B1" = Just ((+) <$> get "A1" <*> "A2")
+sprsh1 get "B2" = Just ((* 2) <$> get "B1")
+sprsh1 _   _    = Nothing
 \end{minted}
 \vspace{1mm}
+Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
+(Real spreadsheet cells would contain a wider range of values, of course.)
+The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
+but not the constant values.  Like every \hs{Task}, \hs{sprsh1} is given
+a \hs{get} function and a key. It pattern-matches on the key to see if it has
+a task description (in the Excel case, a formula)
+for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
+that cell.  If there is a formula in the cell, it computes
+the value of the formula, using \hs{get} to find the value of any keys on
+which it depends.
 
-\noindent
-A compute takes two arguments: \hs{fetch :: k -> f v} and \hs{key :: k}, and
-returns either \hs{Nothing} to indicate that the \hs{key} is an input and hence
-cannot be computed, or \hs{Just}~\hs{value} obtained by \hs{fetch}ing the
-dependencies of the \hs{key} and performing the required computation. A compute
-must be polymorphic in \hs{f} but may request certain computational
-power\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
-might recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}
-by specifying the constraint~\hs{c}. Let us look at an example by recalling the
-following spreadsheet from~\S\ref{sec-background-excel}:
+The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
+because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
+-- the relevant type signatures
+are given in Figure~\ref{fig-lib-funs}.  We will return to the need for this
+generality in
+Section~\ref{sec-deps}\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
+might recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}.
 
+For now, we content ourselves with observing that a task description,
+of type \hs{Task c k v}, is completely isolated from the world of
+compilers, calculation chains, file systems, caches, and all other
+complexities of real build systems.  It just computes a single output, in
+a side-effect-free way, using a call-back (\hs{get}) to get the values
+of its dependencies.
+
+\subsection{The Build abstraction}\label{sec-general-build}
+
+Next comes our second key abstraction: a build system:
 \vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{bash}
-A1 = 10
-A2 = 20
-B1 = A1 + A2
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
 \end{minted}
 \vspace{1mm}
+The signature is very straightforward.  Given a task description, a target key,
+and a store, the build system returns a new store in which the value of the
+target key is up-to-date.  (What exactly does ``up to date'' mean?  We answer
+that precisely in \S\ref{sec-build-correctness}.
 
-\noindent
-The only build task in this spreadsheet can be expressed as an \emph{applicative
-compute}, operating on keys (cell names) of type \hs{String} and values of
-type \hs{Integer}:
-
+\emph{SLPJ: is this right?}
 \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-add :: Compute Applicative String Integer
-add fetch key | key /= "B1" = Nothing
-              | otherwise   = Just $ (+) <$> fetch "A1" <*> fetch "A2"
+dumbBuild :: Build Monad () k v
+dumbBuild task k store = execState (go k) store
+  where
+    go :: k -> State s v
+    go k = case (task go k) of
+             Nothing  -> do { s <- getState; return (getValue s k) }
+             Just act -> do { v <- act; updState (\s -> putStore s k v); return v }
 \end{minted}
 \vspace{1mm}
+\emph{SLPJ: explain}
 
-\noindent
-The cells \cmd{A1} and \cmd{A2} are inputs, which is indicated by returning
-\hs{Nothing}. The cell \cmd{B1} is an output, whose value is computed by
-fetching the values of \cmd{A1} and \cmd{A2} and applying the pure
-function~\hs{(+)} using the methods of the \hs{Applicative} type
-class~\cite{mcbride2008applicative}.
+\subsection{Correctness of a build system} \label{sec-build-correctness}
+
+We can now say what it means for a build system to be \emph{correct}, something
+that is seldom stated formally:
+
+\definition[Correctness]{Suppose \hs{build :: Build c i k v} is a build system,
+\hs{task :: Task c k v} describes an acyclic collection of build tasks, \hs{key} is an output
+key, \hs{store} be a store.
+Then the \hs{result}ing store \hs{result = build compute key store}
+is~\emph{correct} if there exists a store \hs{ideal}, such that
+
+\begin{itemize}
+    \item The \hs{ideal} store is \emph{consistent} with \hs{task}, i.e.
+    for all possible non-input keys \hs{k}, the result of executing the
+    \hs{compute} matches the value in the \hs{ideal} store:
+    \[
+    \hs{execute}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{ideal}~\hs{k)}.
+    \]
+    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.
+    \hs{result}~\hs{key}~\hs{==}~\hs{ideal}~\hs{key}. In other words, we require
+    that the output \hs{key} has the correct value, but do not impose any
+    restrictions on intermediate keys, therefore permitting shallow cloud builds.
+    \item The \hs{store}, \hs{ideal} and \hs{result} all agree on the inputs
+    that belong to the transitive closure of \hs{key}'s dependencies. This means
+    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
+    store has constraints not only on the output, but also on the inputs.
+\end{itemize}
+}
+\label{def-correct}
+
+\subsection{Computing dependencies}
 
@@
 % \end{minted}
-
-\subsection{Build}\label{sec-general-build}
-
-Equipped with the \hs{Compute} abstraction, we can now introduce \hs{Build} in
-more detail. Recall the informal definition of a build system
-from~\S\ref{sec-background}:
-
-\vspace{2mm}
-\emph{Definition (Build system).}\hspace{1.5mm}Given a description of tasks, a
-\emph{build system} executes them in an order that respects their dependencies
-to produce a specified output from supplied inputs.
-\vspace{2mm}
-
-\noindent
-To make the definition more precise, we express it as the following function:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
-\end{minted}
-\vspace{1mm}
-
-\noindent
-The type closely matches the above definition: \hs{Compute c k v} is `a
-description of tasks', a key~\hs{k} is~`a~specified output', and the store
-\store holds `supplied inputs'. The additional input parameter \hs{Maybe i}
-may provide auxiliary information from the previous build, as discussed
-in~\S\ref{sec-background-summary}. The build system produces auxiliary
-information \hs{i} for the next build as well as an updated store, where the
-output \hs{k} must be up-to-date. We formally capture the correctness of the
-resulting store in the following definition.
-
-\definition[Correctness]{Let \hs{build :: Build c i k v} be a build system,
-\hs{compute} describe an acyclic collection of build tasks, \hs{key} be an output
-key, \hs{mi :: Maybe i} be auxiliary information, and \hs{store} be a store.
-Then the \hs{result}ing store \hs{(_, result) = build compute k mi store}
-is~\emph{correct} if there exists an \hs{ideal} store, such that
-
-\begin{itemize}
-    \item The \hs{ideal} store is \emph{consistent} with the \hs{compute}, i.e.
-    for all possible non-input keys \hs{k}, the result of executing the
-    \hs{compute} matches the value in the \hs{ideal} store:
-    \[
-    \hs{execute}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{ideal}~\hs{k)}.
-    \]
-    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.
-    \hs{result}~\hs{key}~\hs{==}~\hs{ideal}~\hs{key}. In other words, we require
-    that the output \hs{key} has the correct value, but do not impose any
-    restrictions on intermediate keys, therefore permitting shallow cloud builds.
-    \item The \hs{store}, \hs{ideal} and \hs{result} all agree on the inputs
-    that belong to the transitive closure of \hs{key}'s dependencies. This means
-    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
-    store has constraints not only on the output, but also on the inputs.
-\end{itemize}
-}
-\label{def-correct}

### 2018-03-13 src/Development/Build.hs
Switch to new Build type

@@ -1,16 +1,14 @@
-{-# LANGUAGE ConstraintKinds, RankNTypes #-}
+{-# LANGUAGE ConstraintKinds, FlexibleContexts, RankNTypes #-}
 module Development.Build (
     -- * Build
-    Build, MultiBuild, sequentialMultiBuild, sequentialMultiStoreBuild,
-    dumb, purify, slow, dumbTracing,
+    Build, MultiBuild, sequentialMultiBuild, --, sequentialMultiStoreBuild,
+    dumb, slow, -- purify, slow, dumbTracing,
 
     -- * Properties
     correct, idempotent
     ) where
 
-import Control.Monad.IO.Class
 import Control.Monad.State
-import Data.List.NonEmpty (NonEmpty (..))
 
 import Development.Build.Task
 import Development.Build.Task.Monad
@@ -22,53 +20,48 @@ import Development.Build.Utilities
 -- and a key-value map @k -> v@, and computes information for the next build and
 -- an updated key-value map. Note that we require @Eq k@ since without it one
 -- has no way of updating the map.
-type Build c i k v = Eq k => Task c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
+type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
 
-type MultiBuild c i k v = Eq k => Task c k v -> NonEmpty k -> Maybe i -> (k -> v) -> (i, k -> v)
+type MultiBuild c i k v = Task c k v -> [k] -> Store i k v -> Store i k v
 
-type StoreBuild c m k v = Store m k v => Task c k v -> k -> m ()
+-- type StoreBuild c m k v = Store m k v => Task c k v -> k -> m ()
 
-type MultiStoreBuild c m k v = Store m k v => Task c k v -> [k] -> m ()
+-- type MultiStoreBuild c m k v = Store m k v => Task c k v -> [k] -> m ()
 
 -- How do I express that we can do it for any c, not just c = Monad?
-purify :: (forall m. StoreBuild Monad m k v) -> Build Monad () k v
-purify build task key _ = runState (build task key)
+-- purify :: (forall m. StoreBuild Monad m k v) -> Build Monad () k v
+-- purify build task key _ = runState (build task key)
 
 sequentialMultiBuild :: Build Monad i k v -> MultiBuild Monad i k v
-sequentialMultiBuild build task (x :| xs) i store =
-    let (i', store') = build task x i store
-    in case xs of
[…]

### 2018-03-13 src/Development/Build/Plan.hs

@@ -20,7 +20,7 @@ import Development.Build.Utilities
 -- about the key and if it is required for building any of the /outputs/ the
 -- build system will need to discover its dependencies. We require that every
 -- plan is 'acyclic'.
-type Plan k v = k -> Maybe (Hash v, [(k, Hash v)])
+type Plan k v = k -> Maybe (Hash, [(k, Hash)])
 
 -- | Sometimes you have no plan at all, i.e. @noPlan = const Nothing@.
 noPlan :: Plan k v
@@ -47,10 +47,10 @@ acyclic plan = forall $ \key -> isJust (reach knownDependencies key)
 --     1. The value has expected hash, i.e. @getHash key@ returns @keyHash@.
 --     2. All dependencies have expected hashes: i.e. @getHash k@ returns @h@
 --        for each @(k, h)@ in @deps@.
-upToDate :: (Eq v, Store m k v) => Plan k v -> k -> m Bool
-upToDate plan key = case plan key of
-    Nothing -> return False -- We don't know and conservatively return False
-    Just (keyHash, deps) -> checkHashes ((key, keyHash) : deps)
+upToDate :: Hashable v => Store i k v -> Plan k v -> k -> Bool
+upToDate store plan key = case plan key of
+    Nothing -> False -- We don't know and conservatively return False
+    Just (keyHash, deps) -> checkHashes store ((key, keyHash) : deps)
 
 -- | Check that a 'Plan' is consistent with respect to a given 'Store'.
 -- * An empty plan (@plan key == Nothing@) is consistent.
@@ -59,7 +59,7 @@ upToDate plan key = case plan key of
 --     1. The value has expected hash, i.e. @getHash key@ returns @keyHash@.
 --     2. All dependencies have expected hashes: i.e. @getHash k@ returns @h@
 --        for each @(k, h)@ in @deps@.
-consistent :: (Eq v, Store m k v) => Plan k v -> m Bool
-consistent plan = forallM $ \key -> case plan key of
-    Nothing -> return True -- Incomplete plan is consistent
-    Just (keyHash, deps) -> checkHashes ((key, keyHash) : deps)
+consistent :: Hashable v => Store i k v -> Plan k v -> Bool
+consistent store plan = forall $ \key -> case plan key of
+    Nothing -> True -- Incomplete plan is consistent
+    Just (keyHash, deps) -> checkHashes store ((key, keyHash) : deps)

### 2018-03-13 src/Development/Build/Store.hs

@@ -1,26 +1,41 @@
 {-# LANGUAGE FlexibleInstances, DefaultSignatures, FunctionalDependencies #-}
 module Development.Build.Store (
     -- * Hashing
-    Hash, hash,
+    Hash, Hashable,
+
+    -- * Store
+    Store, getValue, putValue, getHash, getInfo, putInfo, initialise, checkHashes
 
     -- * The store monad
-    Store (..), Snapshot (..), checkHashes, PureStore, runPureStore,
-    UnsafeMapStore, runUnsafeMapStore, MapStore, runMapStoreT, runMapStore
+    -- Store (..), Snapshot (..), checkHashes, PureStore, runPureStore,
+    -- UnsafeMapStore, runUnsafeMapStore, MapStore, runMapStoreT, runMapStore
     ) where
 
-import Data.Map
-import Data.Functor.Identity
-import Control.Monad.State
-import Control.Monad.Trans.Reader
-
 -- | A 'Hash' is used for efficient tracking and sharing of build results. We
--- use @newtype Hash v = Hash v@ for prototyping.
-newtype Hash v = Hash v deriving (Eq, Show)
+-- use @type Hash = Int@ for prototyping.
+type Hash = Int
 
--- | Compute the hash of a given value. We typically assume cryptographic
--- hashing, e.g. SHA256. Here we use @hash v = Hash v@ for prototyping.
-hash :: v -> Hash v
-hash = Hash
+class Hashable a where
+    -- | Compute the hash of a given value. We typically assume cryptographic
+    -- hashing, e.g. SHA256.
+    hash :: a -> Hash
+
+data Store i k v = Store { getInfo :: i, getValue :: k -> v }
+
+getHash :: Hashable v => Store i k v -> k -> Hash
+getHash s = hash . getValue s
+
+putInfo :: Store i k v -> i -> Store i k v
+putInfo s i = s { getInfo = i }
+
+putValue :: Eq k => Store i k v -> k -> v -> Store i k v
+putValue s k v = s { getValue = \key -> if key == k then v else getValue s key }
+
+checkHashes :: Hashable v => Store m k v -> [(k, Hash)] -> Bool
+checkHashes store = all (\(k, h) -> getHash store k == h)
+
+initialise :: i -> (k -> v) -> Store i k v
+initialise = Store
 
 -- | A key-value store monad that in addition to usual 'getValue' and 'putValue'
 -- queries supports 'getHash', which can in some cases be implemented more
@@ -28,64 +43,58 @@ hash = Hash
[…]

### 2018-03-13 src/Development/Build/Task/Monad.hs

@@ -69,7 +69,7 @@ correctBuild task before after key =
 -- | Run a task with a pure lookup function. Returns @Nothing@ to indicate
 -- that a given key is an input.
 execute :: Task Monad k v -> (k -> v) -> k -> Maybe v
-execute task store = fmap runIdentity . task (pure . store)
+execute task fetch = fmap runIdentity . task (pure . fetch)
 
 -- | Run a task with a partial lookup function. The result @Left k@ indicates
 -- that the task failed due to a missing dependency @k@. Otherwise, the

### 2018-03-13 test/Main.hs

@@ -1,6 +1,5 @@
 {-# LANGUAGE FlexibleContexts, OverloadedStrings, ScopedTypeVariables #-}
 import Control.Monad
-import Data.List.NonEmpty (NonEmpty (..), toList)
 import Data.Maybe
 
 import Development.Build
@@ -9,8 +8,8 @@ import Development.Build.Store
 
 import Development.Build.Example.Spreadsheet
 
-inputs :: Cell -> Int
-inputs cell = fromMaybe (cellNotFoundValue cell) $ lookup cell
+inputs :: Store () Cell Int
+inputs = initialise () $ \cell -> fromMaybe (cellNotFoundValue cell) $ lookup cell
     [ ("A1", 1)
     , ("A2", 2)
     , ("A3", 3) ]
@@ -28,8 +27,8 @@ spreadsheet cell = case name cell of
     'F':_ -> Just $ rel (-1) 0 + rel (-2) 0 --          Fn = F(n - 1) + F(n - 2)
     _     -> Nothing
 
-outputs :: NonEmpty Cell
-outputs = "B1" :| ["B2", "B3", "C1", "C2", "F30" ]
+outputs :: [Cell]
+outputs = [ "B1", "B2", "B3", "C1", "C2", "F30" ]
 
 -- TODO: Handle lookup errors nicer
 -- cellNotFoundError :: Cell -> Int
@@ -42,26 +41,26 @@ cellNotFoundValue _ = 0
 task :: Task Monad Cell Int
 task = spreadsheetTask spreadsheet
 
-dumbResult :: Cell -> Int
-dumbResult = snd $ sequentialMultiBuild dumb task outputs Nothing inputs
+dumbResult :: Store () Cell Int
+dumbResult = sequentialMultiBuild dumb task outputs inputs
 
-slowResult :: Cell -> Int
-slowResult = snd $ sequentialMultiBuild slow task outputs Nothing inputs
+slowResult :: Store () Cell Int
+slowResult = sequentialMultiBuild slow task outputs inputs
 
-tracingDumbResult :: IO (Cell -> Int)
-tracingDumbResult = snd <$> runPureStore build inputs
-  where
-    build = sequentialMultiStoreBuild dumbTracing task (toList outputs)
+-- tracingDumbResult :: IO (Cell -> Int)
+-- tracingDumbResult = snd <$> runPureStore build inputs
+--   where
+--     build = sequentialMultiStoreBuild dumbTracing task (toList outputs)
 
-printOutputs :: (Cell -> Int) -> IO ()
[…]

### 2018-03-13 paper/1-intro.tex

@@
 
-Build systems (such as \Make) are big, complicated, but unloved part of
-the software ecosystem.  Every developer on the planet uses one, but
-they are very much a means to an end, and seldom the focus of
-attention.  But the challenges of scale have driven large software firms
-like Microsoft, Facebook, and Google to develop their own build
-systems, each with its own choices and idiosyncrasies.
+Build systems (such as \Make) are big, complicated, and used by every
+software developer on the planet.  But they are a sadly unloved part
+of the software ecosystem, very much a means to an end, and seldom the
+focus of attention.  Despite this, the challenges of scale have driven
+large software firms like Microsoft, Facebook, and Google to develop
+their own build systems, each with its own choices and idiosyncrasies.
 
@@
 brought out clearly.
-\item Build systems vary on many axes;
-for example: static vs dynamic dependencies; cloud-build, including
+\item Build systems vary on many axes,
+including: static vs dynamic dependencies; cloud-build, including
 shallow vs deep; deterministic vs non-deterministic build rules;
@@
 These properties (which we define in~\S\ref{sec-background}) are often
-deeply-built-in assumptions of a particular build system.
+deeply built into any particular build system.
 In contrast, our framework allows them to be distinguished,
@@
 \end{itemize}
-Thus equipped, instead of seeing build systems as unrelated
+In short, instead of seeing build systems as unrelated
 points in space, we can re-envisage them as locations in a landscape,

### 2018-03-13 paper/2-background.tex

@@
 Build systems automate the execution of simple repeatable tasks for individual
-users, as well as for large organisations. There are software build systems,
-such as \Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake} and
-\Bazel~\cite{bazel}, as well as various incremental calculation engines, such
-as \Excel~\cite{advanced_excel}. In this section we use these four examples to
-introduce main domain-specific notions and requirements. Other notable examples
-of build systems and their relation to these four are discussed
+users, as well as for large organisations. In this section we explore the design
+space of build systems, using four concrete examples:
+\Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake},
+\Bazel~\cite{bazel}, and \Excel~\cite{advanced_excel}\footnote{Excel appears
+very different to the others but, seen through the lens of this paper, it is
+very close indeed.}.
+We have carefully chosen these four to illustrate the various axes on which build systems
+differ; we discuss many other notable examples
+of build systems, and their relationships,
 in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.
-
-We start by giving an intuitive definition of a build system. After providing
-more context and examples in the rest of this section, we
-develop a formal definition in~\S\ref{sec-general-build}.
-
-\definition[Build system]{Given a description of tasks, a \emph{build system}
-executes them in an order that respects their dependencies to produce a
-specified output from supplied inputs.}
-\label{def-build}
 
@@
 $\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and which is
-not known statically. One might say that the value of \cmd{C1} is statically
+not known statically\footnote{One might say that the value of \cmd{C1} is statically
 known in this particular example, but imagine that it is the result of a long computation
-chain -- its value will only become available during the build.
+chain -- its value will only become available during the build.}.
 
@@
 "release.tar" %> \_ -> do
-    need ["release.txt"]
-    files <- lines <$> readFile "release.txt"
-    need files
-    system "tar" $ ["-cf", "result.tar"] ++ files
+    { need ["release.txt"]
+    ; files <- lines <$> readFile "release.txt"
+    ; need files
+    ; system "tar" $ ["-cf", "result.tar"] ++ files }
 \end{minted}
@@
 \emph{adaptive}~\cite{acar2002adaptive}, and
-\emph{self-adjusting}~\cite{acar2007selfadjusting} \emph{computations} (see a
-discussion in related work~\S\ref{sec-related}). Second, instead of abandoning
+\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and \S\ref{sec-related}).
+Second, instead of abandoning
 and deferring the execution of tasks whose newly discovered dependencies have

### 2018-03-13 paper/3-compute.tex
Bring fetch back

@@
 sprsh1 :: Compute Applicative String Integer
-sprsh1 get "B1" = Just ((+) <$> get "A1" <*> "A2")
-sprsh1 get "B2" = Just ((* 2) <$> get "B1")
+sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
+sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
 sprsh1 _   _      = Nothing
@@
 but not the constant values.  Like every \hs{Task}, \hs{sprsh1} is given
-a \hs{get} function and a key. It pattern-matches on the key to see if it has
+a \hs{fetch} function and a key. It pattern-matches on the key to see if it has
 a task description (in the Excel case, a formula)
@@
 that cell.  If there is a formula in the cell, it computes
-the value of the formula, using \hs{get} to find the value of any keys on
+the value of the formula, using \hs{fetch} to find the value of any keys on
 which it depends.
@@
 complexities of real build systems.  It just computes a single output, in
-a side-effect-free way, using a call-back (\hs{get}) to get the values
+a side-effect-free way, using a call-back (\hs{fetch}) to obtain the values
 of its dependencies.

### 2018-03-14 paper/2-background.tex
Add cloud build example

@@
 \Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake},
-\Bazel~\cite{bazel}, and \Excel~\cite{advanced_excel}\footnote{Excel appears
+\Bazel~\cite{bazel}, and \Excel~\cite{advanced_excel}\footnote{\Excel appears
 very different to the others but, seen through the lens of this paper, it is
@@
 of build systems, and their relationships,
-in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.
+in~\S\ref{sec-related}-\S\ref{sec-conclusions}.
 
@@
     gcc -c util.c
-
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{makefile}
 main.o: util.h main.c
     gcc -c main.c
-
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{makefile}
 main.exe: util.o main.o
@@
 \end{subfigure}
-\caption{A task dependency graph and two build scenarios. Input files are shown as
-rectangles, intermediate and output files are shown as rounded rectangles. Dirty
-inputs and files that are rebuilt are highlighted.
+\vspace{-2mm}
+\caption{A task dependency graph and two build scenarios. Input files are shown
+as rectangles, intermediate and output files are shown as rounded rectangles.
+Modified inputs and files that are rebuilt are highlighted.
 \label{fig-make}}
+\vspace{-2mm}
 \end{figure}
@@
 \vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{bash}
-A1 = 10
-A2 = 20
-B1 = A1 + A2
+\begin{minted}[xleftmargin=10pt]{text}
+A1: 10      B1: A1 + A2
+A2: 20
 \end{minted}
@@
 \vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{bash}
-A1 = 10
-A2 = 20
-B1 = IF(C1=1,A1,A2)
-C1 = 1
+\begin{minted}[xleftmargin=10pt]{text}
+A1: 10      B1: IF(C1=1,A1,A2)      C1: 1
+A2: 20
 \end{minted}
@@
 \vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{bash}
-A1 = 10
-A2 = 20
-B1 = INDIRECT("A" & C1)
-C1 = 1
+\begin{minted}[xleftmargin=10pt]{text}
+A1: 10      B1: INDIRECT("A" & C1)      C1: 1
+A2: 20
 \end{minted}
@@
 \centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
-\caption{Since \cmd{docs.txt} has changed, we rebuild \cmd{release.txt} and
+\caption{Since \cmd{docs.txt} was modified, we rebuild \cmd{release.txt} and
 \cmd{release.tar}, discovering a new dependency.}
 \end{subfigure}
+\vspace{-2mm}
 \caption{Dynamic dependencies example: create \cmd{README} and add it to the
 list of release documents \cmd{docs.txt}.\label{fig-shake}}
+\vspace{-2mm}
 \end{figure}
@@
 \begin{figure}[h]
+\vspace{-5mm}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example-cutoff.pdf}}
-\vspace{-2mm}
+\vspace{-3mm}
 \caption{An early cutoff example: if a comment is added to \cmd{main.c}, the
@@
 rebuilt.\label{fig-cutoff}}
+\vspace{-2mm}
 \end{figure}
@@
 
-When build systems are used by large teams, different team members
-often end up executing exactly the same tasks on their local machines.
-A \emph{cloud build system} can speed up builds dramatically by
-transparently sharing build results among team members. Furthermore, cloud
-build systems allow one to perform \emph{shallow builds} that materialise
-only end build products on a local machine, leaving all intermediates in the
-cloud. This is a significant optimisation compared to \emph{deep builds}
-that require all transitive dependencies of an end build product to be
-locally available. % Non-cloud build systems cannot support shallow builds.
+When build systems are used by large teams, different team members often end up
+executing exactly the same tasks on their local machines. A \emph{cloud build
+system} can speed up builds dramatically by transparently sharing build results
+among team members. Furthermore, cloud build systems allow one to perform
+\emph{shallow builds} that materialise only end build products locally, leaving
+all intermediates in the cloud.
 
@@
 
-To support cloud builds, \Bazel maintains a conventional \emph{content-addressable
-cache}, which can be used to fetch a previously built file given the hash of its
+To support cloud builds, \Bazel maintains a \emph{content-addressable cache},
+which can be used to fetch a previously built file given the hash of its
 content, as well as \emph{dependency graphs from all previous builds}, annotated
@@
 provided in~\S\ref{sec-implementations}.
+
+Consider a cloud build example in Fig.~\ref{fig-bazel}. The user starts by
+checking out the project sources, whose hashes happen to be 1, 2 and 3, and
+requests to build \cmd{main.exe}. By looking up the history of all previous
+builds, the build system finds that someone has already compiled these sources
+before and obtains the hashes (4~and~5) of the resulting \cmd{util.o} and
+\cmd{main.o} files (the files themselves do not need to be downloaded).
+Similarly, the build system finds that the hash of the resulting \cmd{main.exe}
+should be 6 and downloads the actual binary from the cache, since it is the end
+build product and must be materialised. In the second scenario, shown in
+Fig.~\ref{fig-bazel}(c), the user modifies \cmd{util.c}, thereby changing its
+hash from 1 to 7. The build system finds that nobody has ever compiled the new
+$\{\cmd{util.c}, \cmd{util.h}\}$ combination and must therefore build
+\cmd{util.o}, which results in changing its hash from~4~to~8. The obtained
+combination of hashes of \cmd{util.o} and \cmd{main.o} has not been encountered
+before, therefore the build system first downloads \cmd{main.o} from the cache
+and then builds \cmd{main.exe} by linking the two object files. When the build
+is complete, the obtained files as well as the new dependency graph annotated
+with hashes are stored in the cloud for future reuse.
+
+\begin{figure}[t]
+\begin{subfigure}[b]{0.25\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
+\caption{Checkout source files}
+\end{subfigure}
+\begin{subfigure}[b]{0.40\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
+\caption{Build \cmd{main.exe}}
+\end{subfigure}
+\begin{subfigure}[b]{0.31\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
+\caption{Modify \cmd{util.c} and rebuild}
+\end{subfigure}
+\vspace{-2mm}
+\caption{A cloud build example: (a)~checkout sources, (b)~download \cmd{main.exe}
+from the cloud and skip intermediate files (only their hashes are needed),
+(c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which requires building
+\cmd{util.o} (since nobody has compiled \cmd{util.c} before) and downloading
+\cmd{main.o}. File hashes are shown inside circles.
+\label{fig-bazel}}
+\vspace{-2mm}
+\end{figure}
 

### 2018-03-14 paper/3-compute.tex
Work on S3

@@
 section~\S\ref{sec-background} and design complex build systems from simple
-primitives. Specifically, we present the \emph{compute} and \emph{build}
-abstractions, which are described in
-Sections~\ref{sec-task} and \ref{sec-general-build} respectively.
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
-\end{minted}
-\vspace{1mm}
-
-\hs{Build} corresponds to the algorithm used for bringing a key-value store
-\store to a consistent state by recomputing out-of-date values in an
-appropriate order. We explain the build type and use it to develop a formal
-definition of build systems in~\S\ref{sec-general-build}.
-Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the build
-abstraction further and provide concrete implementations for several build
+primitives. Specifically, we present the \emph{task} and \emph{build}
+abstractions in~\S\ref{sec-task} and~\S\ref{sec-general-build}, respectively.
+Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the
+abstractions further and provide concrete implementations for several build
 systems.
-
-\subsection{Common vocabulary for build systems}
-\label{sec-vocabulary}
-
-\begin{figure}
-\begin{minted}{haskell}
--- Stores
-data Store i k v   -- Abstract
-getInfo  :: Store i k v -> i
-putInfo  :: Store i k v -> i -> Store i k v
-getValue :: Store i k v -> k -> v
-putValue :: Hashable v => Store i k v -> k -> v -> Store i k v
-getHash  :: Hashable v => Store i k v -> k -> Hash
-
--- Hashing
-hash :: Hashable a => a -> Hash
-
--- Applicative
-(<*>) :: Applicative f => f (a->b) -> f a -> f b  -- Left-associative
-(<$>) :: Applicative f => (a->b) -> f a -> f b    -- Left-associative
-pure  :: Applicative f => a -> f a
-
--- State monad
-data State s
-instance Applicative State
-instance Monad       State
-evalState :: State s a -> s -> s
-getState  :: State s s
-\end{minted}
-\caption{Signatures of library functions}
-\label{fig-lib-funs}
-\end{figure}
 
@@
 development or spreadsheets.
-% We will use this vocabulary throughout the rest of the paper.
+
+\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
+
+\begin{figure}
+\begin{minted}{haskell}
+-- An abstract store
+data Store i k v
+getInfo    :: Store i k v -> i
+putInfo    :: Store i k v -> i -> Store i k v
+getValue   :: Store i k v -> k -> v
+putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
+getHash    :: Hashable v => Store i k v -> k -> Hash
+initialise :: Hashable v => i -> (k -> v) -> Store i k v
+\end{minted}
+\vspace{1mm}
+\begin{minted}{haskell}
+-- Hashing
+hash :: Hashable a => a -> Hash
+\end{minted}
+\vspace{1mm}
+\begin{minted}{haskell}
+-- Applicative functors
+pure  :: Applicative f => a -> f a
+(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
+(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
+\end{minted}
+\vspace{1mm}
+\begin{minted}{haskell}
+-- State monad
+data State s
+instance Applicative (State s)
+instance Monad       (State s)
+get       :: State s s
+put       :: s -> State s ()
+modify    :: (s -> s) -> State s ()
+execState :: State s a -> s -> s
+\end{minted}
+\vspace{-2mm}
+\caption{Signatures of library functions}\label{fig-lib-funs}
+\vspace{-4mm}
+\end{figure}
 
@@
 and a \emph{store}, and returns a new store in which the target key is
-up to date.  The store is a key/value mapping, togther with
-\emph{persistent build state} that is maintained by the build system itself.
+up to date. The store is a key-value map, together with \emph{persistent build
+information} that is maintained by the build system itself.
 
@@
 variables \hs{k} and \hs{v} to denote keys and values, respectively.
+Fig.~\ref{fig-lib-funs} provides a Haskell encoding of our main definitions,
+and gives types of the library functions we use in this paper.
 
@@
 
-The types of the library functions we use in this paper is
-given in Figure~\ref{fig-lib-funs}.
-We use a cryptographic \emph{hash function} \hs{hash :: Hashable v =>
-v -> Hash} for efficient tracking and sharing of~build results, where
-\hs{Hash} is an abstract data type equipped with the equality test.
+We use a cryptographic \emph{hash function}
+\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash} for
+efficient tracking and sharing of build results, where \hs{Hash} is an abstract
+data type equipped with the equality test.
 
-A \emph{store}, of type \hs{Store i k v}, associates keys to values.
-It is convenient to assume that a
-store is total function \store, i.e. it contains a value for every possible key.
-We therefore also assume that the type of values is capable of encoding values
-corresponding to non-existent keys (missing files or empty cells). This, in
-particular, suggests that it is possible to \emph{depend on the absence of a
-value}, which is a useful feature for build systems. In addition to usual
-read and write operations, some stores also support a dedicated \emph{read hash}
-operation, which is implemented more efficiently than by reading and hashing the
-result. For example, GVFS (Git Virtual File System)~\cite{gvfs} stores file
-hashes and downloads file contents only on demand -- an important feature for
-cloud build systems.
-
-The store also contains a single persistent value, of type \hs{i}, representing
-the persistent state maintained by the build system itself -- its ``memory''.
+A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
+values. It is convenient to assume that a store is total, i.e. it contains a
+value for every possible key. We therefore also assume that the type of values
+is capable of encoding values corresponding to non-existent keys (missing files
+or empty cells). This, in particular, suggests that it is possible to
+\emph{depend on the absence of a value}, which is a useful feature for build
+systems. In addition to usual \hs{getValue} and \hs{putValue} operations, some
+stores also support a dedicated \hs{getHash} operation, which is implemented
+more efficiently than by hashing the result of \hs{getValue}. For example, GVFS
+(Git Virtual File System)~\cite{gvfs} stores file hashes and downloads file
+contents only on demand. The store also contains a single persistent value, of
+type \hs{i}, representing the persistent state maintained by the build system
+itself~--~its~``memory''. We can \hs{initialise} a store by supplying an initial
+value \hs{i} and a key-value map \store.
 
@@
 
-Our first key idea is the task-description abstraction:
+Our first main idea is the task-description abstraction:
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-This highly-abstracted type is best introduced through an example.
-Consider this Excel spreadsheet (c.f. \S\ref{sec-background-excel}):
+This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
+\emph{profunctor optics} might recognise a familiar pattern. We discuss this
+in~\S\ref{sec-related-optics}.} is best introduced by an example.
+Consider this \Excel spreadsheet:
 \vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{bash}
-  A1: 10     B1: =A1 + A2
-  A2: 20     B2: =B1 * 2
+\begin{minted}[xleftmargin=10pt]{text}
+  A1: 10     B1: A1 + A2
+  A2: 20     B2: B1 * 2
 \end{minted}
 \vspace{1mm}
-Here cell \hs{A1} contains the value \hs{10} while cell \hs{B1} contains
-the formula \hs{=A1+A2}.
-We can represent the formulae of of this spreadsheet
-with the following task description
+Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
+the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
+with the following task description:
 \vspace{1mm}
@@
 The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
-but not the constant values.  Like every \hs{Task}, \hs{sprsh1} is given
+but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given
 a \hs{fetch} function and a key. It pattern-matches on the key to see if it has
-a task description (in the Excel case, a formula)
+a task description (in the \Excel case, a formula)
 for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
@@
 -- the relevant type signatures
-are given in Figure~\ref{fig-lib-funs}.  We will return to the need for this
-generality in
-Section~\ref{sec-deps}\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
-might recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}.
+are given in Fig.~\ref{fig-lib-funs}. We will return to the need for this
+generality in~\S\ref{sec-deps}.
 
@@
 complexities of real build systems.  It just computes a single output, in
-a side-effect-free way, using a call-back (\hs{fetch}) to obtain the values
+a side-effect-free way, using a callback (\hs{fetch}) to find the values
 of its dependencies.
@@
 
-Next comes our second key abstraction: a build system:
+Next comes our second main abstraction -- a build system:
 \vspace{1mm}
@@
 and a store, the build system returns a new store in which the value of the
-target key is up-to-date.  (What exactly does ``up to date'' mean?  We answer
+target key is up-to-date. What exactly does ``up to date'' mean?  We answer
 that precisely in \S\ref{sec-build-correctness}.
 
-\emph{SLPJ: is this right?}
 \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-dumbBuild :: Build Monad () k v
-dumbBuild task k store = execState (go k) store
+busy :: (Eq k, Hashable v) => Build Monad () k v
+busy task key store = execState (go key) store
   where
-    go :: k -> State s v
-    go k = case (task go k) of
-             Nothing  -> do { s <- getState; return (getValue s k) }
-             Just act -> do { v <- act; updState (\s -> putStore s k v); return v }
+    go k = case task go k of
+        Nothing  -> do { s <- get; return (getValue s k) }
+        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
 \end{minted}
 \vspace{1mm}
+
 \emph{SLPJ: explain}
@@
 
-\subsection{Computing dependencies}
+\subsection{Computing dependencies}\label{sec-deps}
 

### 2018-03-14 paper/2-background.tex
Work on S3
Polish S2

@@
 \emph{shallow builds} that materialise only end build products locally, leaving
-all intermediates in the cloud.
+all intermediates in the cloud. We illustrate shallow cloud builds by an example
+(see~Fig.~\ref{fig-bazel}).
+
+The user starts by checking out the project sources, whose hashes are (for
+simplicity)~1,~2~and~3, and requests to build \cmd{main.exe}, see
+Fig.~\ref{fig-bazel}(a,b). By looking up the history of all previous builds of
+the project\footnote{Here we ignore the issue of limited cloud storage resources
+for the sake of simplicity; in practice, old entries are regularly evicted from
+the storage, as further discussed in~\S\ref{sec-cloud-aspects}.}, the build
+system finds that someone has already compiled these exact sources before and
+their resulting files \cmd{util.o} and \cmd{main.o} had hashes 4~and~5.
+Similarly, the build system finds that the hash of the resulting \cmd{main.exe}
+should be 6 and downloads the actual binary from the shared cloud storage, since
+it is the end build product and must therefore be materialised.
+
+In the second scenario, shown in Fig.~\ref{fig-bazel}(c), the user modifies the
+source \cmd{util.c}, thereby changing its hash from~1~to~7. The build system
+finds that nobody has ever compiled the new $\{\cmd{util.c}, \cmd{util.h}\}$
+combination and must therefore build \cmd{util.o}, which results in changing its
+hash from~4~to~8. The obtained combination of hashes of \cmd{util.o} and
+\cmd{main.o} has not been encountered before either, therefore the build system
+first downloads \cmd{main.o} from the cloud and then builds \cmd{main.exe} by
+linking the two object files. When the build is complete, the obtained results
+may be uploaded to the cloud for future reuse by other team members.
 
@@
 from the simplicity of building tasks in a statically known topological order.
-It is minimal and supports the early cutoff optimisation.
-
-To support cloud builds, \Bazel maintains a \emph{content-addressable cache},
-which can be used to fetch a previously built file given the hash of its
-content, as well as \emph{dependency graphs from all previous builds}, annotated
-with observed file hashes\footnote{Here we ignore the issue of limited cloud
-storage resources for the sake of simplicity; in practice, old entries are
-regularly evicted from the storage, as further discussed
-in~\S\ref{sec-cloud-aspects}.}. The latter allows to bypass the execution of
-a task, by predicting the hash of the result from the hashes of its dependencies,
-and subsequently fetching the result from the cache. A concrete implementation is
-provided in~\S\ref{sec-implementations}.
-
-Consider a cloud build example in Fig.~\ref{fig-bazel}. The user starts by
-checking out the project sources, whose hashes happen to be 1, 2 and 3, and
-requests to build \cmd{main.exe}. By looking up the history of all previous
-builds, the build system finds that someone has already compiled these sources
-before and obtains the hashes (4~and~5) of the resulting \cmd{util.o} and
-\cmd{main.o} files (the files themselves do not need to be downloaded).
-Similarly, the build system finds that the hash of the resulting \cmd{main.exe}
-should be 6 and downloads the actual binary from the cache, since it is the end
-build product and must be materialised. In the second scenario, shown in
-Fig.~\ref{fig-bazel}(c), the user modifies \cmd{util.c}, thereby changing its
-hash from 1 to 7. The build system finds that nobody has ever compiled the new
-$\{\cmd{util.c}, \cmd{util.h}\}$ combination and must therefore build
-\cmd{util.o}, which results in changing its hash from~4~to~8. The obtained
-combination of hashes of \cmd{util.o} and \cmd{main.o} has not been encountered
-before, therefore the build system first downloads \cmd{main.o} from the cache
-and then builds \cmd{main.exe} by linking the two object files. When the build
-is complete, the obtained files as well as the new dependency graph annotated
-with hashes are stored in the cloud for future reuse.
+It is minimal and supports the early cutoff optimisation. To support cloud
+builds, \Bazel maintains a \emph{content-addressable cache} that can be used to
+fetch a previously built file given the hash of its content, and
+\emph{dependency graphs from all previous builds}, annotated with observed file
+hashes. The latter allows to bypass the execution of a task, by predicting the
+hash of the result from the hashes of its dependencies, and subsequently
+fetch the result from the cache. A concrete implementation is provided
+in~\S\ref{sec-implementations}.
 
@@
 \begin{table}[h]
+\vspace{-1mm}
 \smaller
@@
 \end{tabular}
-\vspace{2mm}
+\vspace{0.5mm}
 \caption{Summary of build system differences.\label{tab-summary}}
-\vspace{-7mm}
+\vspace{-8mm}
 \end{table}

### 2018-03-14 paper/2-background.tex
Rearrange the order of subsections in S3

@@
 simplicity)~1,~2~and~3, and requests to build \cmd{main.exe}, see
-Fig.~\ref{fig-bazel}(a,b). By looking up the history of all previous builds of
-the project\footnote{Here we ignore the issue of limited cloud storage resources
-for the sake of simplicity; in practice, old entries are regularly evicted from
-the storage, as further discussed in~\S\ref{sec-cloud-aspects}.}, the build
-system finds that someone has already compiled these exact sources before and
-their resulting files \cmd{util.o} and \cmd{main.o} had hashes 4~and~5.
+Fig.~\ref{fig-bazel}(a,b). By looking up the global history of all previous
+builds of the project\footnote{Here we ignore the issue of limited cloud storage
+resources for the sake of simplicity; in practice, old entries are regularly
+evicted from the storage, as further discussed in~\S\ref{sec-cloud-aspects}.},
+the build system finds that someone has already compiled these exact sources
+before and their resulting files \cmd{util.o} and \cmd{main.o} had hashes 4~and~5.
 Similarly, the build system finds that the hash of the resulting \cmd{main.exe}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
+\vspace{-1mm}
 \caption{Checkout source files}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
+\vspace{-1mm}
 \caption{Build \cmd{main.exe}}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
+\vspace{-1mm}
 \caption{Modify \cmd{util.c} and rebuild}
@@
 \label{fig-bazel}}
-\vspace{-2mm}
+\vspace{-1mm}
 \end{figure}
@@
 \begin{table}[h]
-\vspace{-1mm}
+\vspace{-2mm}
 \smaller

### 2018-03-14 paper/3-compute.tex
Rearrange the order of subsections in S3

@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh1 :: Compute Applicative String Integer
+sprsh1 :: Task Applicative String Integer
 sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
@@
 
+\subsection{Monadic tasks}\label{sec-task-monad}
+
+As discussed in~\S\ref{sec-background}, some build tasks have dynamic
+dependencies, which are determined by values of intermediate computations. In
+our framework, such build tasks correspond to \emph{monadic task}, i.e.
+the type \hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
+example:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{bash}
+A1 = 10
+A2 = 20
+B1 = IF(C1=1,A1,A2)
+C1 = 1
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
+dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
+our task abstraction as follows\footnote{The spreadsheet example that uses
+the \hs{INDIRECT} function can be expressed very similarly: simply replace the
+last line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+select :: Task Monad String Integer
+select fetch key | key /= "B1" = Nothing
+                 | otherwise   = Just $ do
+                     c1 <- fetch "C1"
+                     if c1 == 1 then fetch "A1" else fetch "A2"
+\end{minted}
+\vspace{1mm}
+
+We cannot statically analyse \hs{select} using the function \hs{dependencies}
+from the previous section, but we can still \hs{execute} it on a given store:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store key = if key == "A1" then 10 else 20
+@\ghci@ execute select store "B1"
+Just 20
+\end{minted}
+\vspace{1mm}
+
+\noindent
+In the example above $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
+To extract the dependencies of \hs{select} we introduce the function \hs{track}:
+a combination of \hs{execute} and \hs{dependencies} that takes the store~\store
+as input and computes both the resulting value and the list of its dependencies:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+import Control.Monad.Writer
+\end{minted}
+\vspace{0.5mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
+track task store = fmap runWriter . task (\k -> writer (store k, [k]))
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
+we feed \hs{fetch k = writer (store k, [k])} to the compute, which in addition
+to fetching a value from the store, tracks the corresponding key. We can
+\hs{track} both applicative and monadic compute:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store key = if key == "A1" then 10 else 20
+@\ghci@ track add store "B1"
+Just (30,["A1","A2"])
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ track select store "B1"
+Just (20,["C1","A2"])
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store' key = if key == "A1" then 10 else 1
+@\ghci@ track select store' "B1"
+Just (10,["C1","A1"])
+\end{minted}
+\vspace{1mm}
+
+\noindent
+As expected, the dependencies of \hs{select} are determined by the contents of
+the store.
+
 \subsection{Computing dependencies}\label{sec-deps}
 
-The \hs{Compute} type and the example above might seem mysterious. Why insist on
-\hs{Compute} to be polymorphic in \hs{f}? Why choose the \hs{Applicative}
+The \hs{Task} type and the example above might seem mysterious. Why insist on
+\hs{Task} to be polymorphic in \hs{f}? Why choose the \hs{Applicative}
 constraint?
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-dependencies :: Compute Applicative k v -> k -> [k]
-dependencies compute = maybe [@@] getConst . compute (\k -> Const [k])
+dependencies :: Task Applicative k v -> k -> [k]
+dependencies task = maybe [@@] getConst . task (\k -> Const [k])
 \end{minted}
@@
 To understand how \hs{dependencies} works, let us examine its type more closely,
-expanding the type synonym \hs{Compute}:
+expanding the type synonym \hs{Task}:
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-execute :: Compute Monad k v -> (k -> v) -> k -> Maybe v
+execute :: Task Monad k v -> (k -> v) -> k -> Maybe v
 execute compute store = fmap runIdentity . compute (pure . store)
@@
 Note that \hs{execute} can accept both an applicative as well as a \emph{monadic
-compute} introduced in~\S\ref{sec-compute-monad}. Below we run \hs{execute} on
+task} introduced in~\S\ref{sec-task-monad}. Below we run \hs{execute} on
 the above \hs{add} example. Note that \hs{execute} returns \hs{Nothing} for
@@
 \end{minted}
-
-\subsection{Monadic compute}\label{sec-compute-monad}
-
-As discussed in~\S\ref{sec-background}, some build tasks have dynamic
-dependencies, which are determined by values of intermediate computations. In
-our framework, such build tasks correspond to \emph{monadic compute}, i.e.
-the type \hs{Compute}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
-example:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{bash}
-A1 = 10
-A2 = 20
-B1 = IF(C1=1,A1,A2)
-C1 = 1
-\end{minted}
-\vspace{1mm}
-
-\noindent
-The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
-dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
-our compute abstraction as follows\footnote{The spreadsheet example that uses
-the \hs{INDIRECT} function can be expressed very similarly: simply replace the
-last line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-select :: Compute Monad String Integer
-select fetch key | key /= "B1" = Nothing
-                 | otherwise   = Just $ do
-                     c1 <- fetch "C1"
-                     if c1 == 1 then fetch "A1" else fetch "A2"
-\end{minted}
-\vspace{1mm}
-
-We cannot statically analyse \hs{select} using the function \hs{dependencies}
-from the previous section, but we can still \hs{execute} it on a given store:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store key = if key == "A1" then 10 else 20
-@\ghci@ execute select store "B1"
-Just 20
-\end{minted}
-\vspace{1mm}
-
-\noindent
-In the example above $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
-To extract the dependencies of \hs{select} we introduce the function \hs{track}:
-a combination of \hs{execute} and \hs{dependencies} that takes the store~\store
-as input and computes both the resulting value and the list of its dependencies:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-import Control.Monad.Writer
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-track :: Compute Monad k v -> (k -> v) -> k -> Maybe (v, [k])
-track compute store = fmap runWriter . compute (\k -> writer (store k, [k]))
-\end{minted}
-\vspace{1mm}
-
-\noindent
-The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
-we feed \hs{fetch k = writer (store k, [k])} to the compute, which in addition
-to fetching a value from the store, tracks the corresponding key. We can
-\hs{track} both applicative and monadic compute:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store key = if key == "A1" then 10 else 20
-@\ghci@ track add store "B1"
-Just (30,["A1","A2"])
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ track select store "B1"
-Just (20,["C1","A2"])
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store' key = if key == "A1" then 10 else 1
-@\ghci@ track select store' "B1"
-Just (10,["C1","A1"])
-\end{minted}
-\vspace{1mm}
-
-\noindent
-As expected, the dependencies of \hs{select} are determined by the contents of
-the store.
 

### 2018-03-14 paper/3-abstractions.tex
Rename section file
Work on S3

--- 2018-03-14 paper/3-compute.tex
@@
-\section{Build systems, abstractly}\label{sec-compute}
+\section{Build systems, abstractly}\label{sec-abstractions}
 
@@
 
-\definition[Correctness]{Suppose \hs{build :: Build c i k v} is a build system,
-\hs{task :: Task c k v} describes an acyclic collection of build tasks, \hs{key} is an output
-key, \hs{store} be a store.
-Then the \hs{result}ing store \hs{result = build compute key store}
-is~\emph{correct} if there exists a store \hs{ideal}, such that
+\definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
+build task description, \hs{key} is an output key, \hs{store} is an initial
+store, and \hs{result} is the store produced by running the build system with
+parameters \hs{task}, \hs{key} and \hs{store}. Or, using the precise language of
+our abstractions:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+build  :: Build c i k v
+task   :: Task c k v
+key    :: k
+store  :: Store i k v
+result :: Store i k v
+result = build task key store
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Then the \hs{result} is~\emph{correct} if there exists an \hs{ideal} store, such
+that:
 
 \begin{itemize}
-    \item The \hs{ideal} store is \emph{consistent} with \hs{task}, i.e.
+    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
     for all possible non-input keys \hs{k}, the result of executing the
-    \hs{compute} matches the value in the \hs{ideal} store:
+    \hs{task} matches the value in the \hs{ideal} store:
     \[
-    \hs{execute}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{ideal}~\hs{k)}.
+    \hs{execute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{ideal}~\hs{k)}.
     \]

### 2018-03-14 paper/3-abstractions.tex
Work on S3.1-S3.3

@@
 \begin{figure}
-\begin{minted}{haskell}
+\begin{minted}[fontsize=\small]{haskell}
 -- An abstract store
@@
 \vspace{1mm}
-\begin{minted}{haskell}
+\begin{minted}[fontsize=\small]{haskell}
 -- Hashing
@@
 \vspace{1mm}
-\begin{minted}{haskell}
+\begin{minted}[fontsize=\small]{haskell}
 -- Applicative functors
@@
 \vspace{1mm}
-\begin{minted}{haskell}
+\begin{minted}[fontsize=\small]{haskell}
 -- State monad
-data State s
+data State s a
 instance Applicative (State s)
@@
 \vspace{-2mm}
-\caption{Signatures of library functions}\label{fig-lib-funs}
+\caption{Signatures of main data types and library functions.}\label{fig-types}
 \vspace{-4mm}
@@
 and a \emph{store}, and returns a new store in which the target key is
-up to date. The store is a key-value map, together with \emph{persistent build
+up-to-date. The store is a key-value map, together with \emph{persistent build
 information} that is maintained by the build system itself.
@@
 variables \hs{k} and \hs{v} to denote keys and values, respectively.
-Fig.~\ref{fig-lib-funs} provides a Haskell encoding of our main definitions,
+Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
 and gives types of the library functions we use in this paper.
@@
 is capable of encoding values corresponding to non-existent keys (missing files
-or empty cells). This, in particular, suggests that it is possible to
-\emph{depend on the absence of a value}, which is a useful feature for build
-systems. In addition to usual \hs{getValue} and \hs{putValue} operations, some
-stores also support a dedicated \hs{getHash} operation, which is implemented
-more efficiently than by hashing the result of \hs{getValue}. For example, GVFS
-(Git Virtual File System)~\cite{gvfs} stores file hashes and downloads file
-contents only on demand. The store also contains a single persistent value, of
-type \hs{i}, representing the persistent state maintained by the build system
-itself~--~its~``memory''. We can \hs{initialise} a store by supplying an initial
-value \hs{i} and a key-value map \store.
+or empty cells). In addition to usual \hs{getValue} and \hs{putValue} operations,
+some stores support a dedicated \hs{getHash} operation, which is implemented
+more efficiently than by hashing the result of \hs{getValue}\footnote{For
+example, Git Virtual File System~\cite{gvfs} stores file hashes and downloads
+file contents only on demand.}. The store also holds a value of type \hs{i},
+representing the persistent build information maintained by the build
+system~--~its ``memory''. Given an \hs{i} and a key-value map \store, we can
+\hs{initialise} a store.
+
+% This, in particular, suggests that it is possible to \emph{depend on the
+% absence of a value}, which is a useful feature for build systems.
 
@@
 The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
-but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given
-a \hs{fetch} function and a key. It pattern-matches on the key to see if it has
-a task description (in the \Excel case, a formula)
-for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
-that cell.  If there is a formula in the cell, it computes
-the value of the formula, using \hs{fetch} to find the value of any keys on
-which it depends.
+but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
+\hs{fetch} function and a key. It pattern-matches on the key to see if it has a
+task description (in the \Excel case, a formula) for it. If not, it returns
+\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
+it computes the value of the formula, using \hs{fetch} to find the value of any
+keys on which it depends.
 
@@
 because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
--- the relevant type signatures
-are given in Fig.~\ref{fig-lib-funs}. We will return to the need for this
-generality in~\S\ref{sec-deps}.
+-- the relevant type signatures are given in Fig.~\ref{fig-types}. We will
+return to the need for this generality in~\S\ref{sec-deps}.
 
 For now, we content ourselves with observing that a task description,
-of type \hs{Task c k v}, is completely isolated from the world of
+of type \hs{Task}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
 compilers, calculation chains, file systems, caches, and all other
@@
 and a store, the build system returns a new store in which the value of the
-target key is up-to-date. What exactly does ``up to date'' mean?  We answer
-that precisely in \S\ref{sec-build-correctness}.
+target key is up-to-date. What exactly does ``up-to-date'' mean?  We answer
+that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:
 
@@
 busy :: (Eq k, Hashable v) => Build Monad () k v
-busy task key store = execState (go key) store
+busy task key store = execState (compute key) store
   where
-    go k = case task go k of
+    compute k = case task compute k of
         Nothing  -> do { s <- get; return (getValue s k) }
@@
 
-\emph{SLPJ: explain}
+\noindent
+The \hs{busy} build system uses a function
+\hs{compute}~\hs{::}~\hs{k}~\hs{->}~\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)}~\hs{v},
+which given a key brings the associated value up-to-date and returns it. To do
+so, \hs{compute} calls \hs{task} passing itself as an argument, so that \hs{task}
+can use \hs{compute} to recursively find the values of its dependencies. If
+\hs{task} returns \hs{Nothing} (this key is an input), \hs{compute} simply reads
+the result from the store~\hs{s}, which is available in the \hs{State} monad (in
+which \hs{compute} runs) via the \hs{get} method~--~see the \hs{State} monad
+interface in Fig.~\ref{fig-types}. Otherwise it runs the action \hs{act}
+returned by the \hs{task} to produce a resulting value~\hs{v}, stores it
+(modifying the store \hs{s}), and returns it.
+
+Given an acyclic task description, the \hs{busy} build system terminates with a
+correct result, but it is not a minimal build system (c.f. the
+Definition~\ref{def-minimal}); \hs{busy} keeps no track of keys it has already
+brought up-to-date, and will therefore busily recompute the same keys again and
+again if they have multiple dependents. We will develop much more efficient
+build systems in~\S\ref{sec-implementations}.
 

### 2018-03-14 paper/4-build.tex
Rename section file
Work on S3.1-S3.3

--- 2018-03-11 paper/4-build.tex
@@
 Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort.
-However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{getdependencies}.
+However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{sec-deps}.
 
@@
 We assume that \hs{Hash v} is a small constant size, constructed from hashing the underlying \hs{v} rather than storing it directly.
-Checking a trace requires ensuring all the dependencies are up to date (using whatever ordering strategy as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are same as the current value and that the result is the same.
+Checking a trace requires ensuring all the dependencies are up-to-date (using whatever ordering strategy as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are same as the current value and that the result is the same.
 After computing a fresh value we add its \hs{Trace} to the list.

### 2018-03-14 paper/3-abstractions.tex
Add busy build example

@@
 sprsh1 :: Task Applicative String Integer
-sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
+sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
 sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
@@
 correct result, but it is not a minimal build system (c.f. the
-Definition~\ref{def-minimal}); \hs{busy} keeps no track of keys it has already
-brought up-to-date, and will therefore busily recompute the same keys again and
-again if they have multiple dependents. We will develop much more efficient
-build systems in~\S\ref{sec-implementations}.
+Definition~\ref{def-minimal}). Since \hs{busy} has no memory (\hs{i = ()}), it
+cannot keep track of keys it has already built, and will therefore busily
+recompute the same keys again and again if they have multiple dependents. We
+will develop much more efficient build systems in~\S\ref{sec-implementations}.
+
+Despite its limitations, \hs{busy} can easily handle our \hs{sprsh1} example
+from the previous section~\S\ref{sec-task}. Below we initialise the store with
+\cmd{A1} set to 10 and all other cells set to 20.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store  = initialise () $ \key -> if key == "A1" then 10 else 20
+@\ghci@ result = busy sprsh1 "B2" store
+@\ghci@ getValue result "B1"
+30
+@\ghci@ getValue result "B2"
+60
+\end{minted}
+\vspace{1mm}
+
+\noindent
+As one can see, \hs{busy} correctly built both \cmd{B1} and \cmd{B2}, and in the
+right order (if it had built \cmd{B2} before building \cmd{B1}, the result would
+have been $20 * 2 = 40$ instead of $(10 + 20) * 2 = 60$). As an example showing
+that \hs{busy} is not minimal, imagine that the formula in cell \cmd{B2} was
+\cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
+\hs{compute}~\hs{"B1"} twice -- once per each occurrence of \cmd{B1} in the
+formula.
 

### 2018-03-14 src/Development/Build/Task.hs
Add sprsh2 example

@@ -99,7 +99,12 @@ indirect fetch key | key /= "B1" = Nothing
 sprsh1 :: Task Applicative String Integer
 sprsh1 fetch "B1" = Just ((+) <$> fetch "A1" <*> fetch "A2")
 sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
-sprsh1 _ _ = Nothing
+sprsh1 _     _    = Nothing
+
+sprsh2 :: Task Monad String Integer
+sprsh2 fetch "B1" = Just $ do c1 <- fetch "C1"
+                              if c1 == 1 then fetch "A1" else fetch "A2"
+sprsh2 _     _    = Nothing
 
 -- These type synonyms are not very useful, but enumerate all interesting cases.
 type FunctorialTask  k v = forall f. Functor     f => (k -> f v) -> k -> Maybe (f v)

### 2018-03-14 paper/3-abstractions.tex
Work on S3.4

@@
 
-Despite its limitations, \hs{busy} can easily handle our \hs{sprsh1} example
-from the previous section~\S\ref{sec-task}. Below we initialise the store with
+Despite limitations, \hs{busy} can easily handle the example \hs{sprsh1}
+from the previous subsection~\S\ref{sec-task}. Below we initialise the store with
 \cmd{A1} set to 10 and all other cells set to 20.
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store  = initialise () $ \key -> if key == "A1" then 10 else 20
+@\ghci@ store  = initialise () (\key -> if key == "A1" then 10 else 20)
 @\ghci@ result = busy sprsh1 "B2" store
@@
 formula.
+
+An alert reader might notice that \hs{busy} accepts \emph{monadic tasks} of
+type \hs{Task}~\hs{Monad}, whereas \hs{sprsh1} is a \hs{Task}~\hs{Applicative}.
+This indicates that \hs{busy} can cope with tasks containing dynamic
+dependencies, which we address in the next subsection~\S\ref{sec-task-monad}.
+
+\subsection{Monadic tasks}\label{sec-task-monad}
+
+As discussed in~\S\ref{sec-background-excel}, some build tasks have dynamic
+dependencies, which are determined by values of intermediate computations. In
+our framework, such build tasks correspond to the type
+\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
+example:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{text}
+A1: 10      B1: IF(C1=1,A1,A2)      C1: 1
+A2: 20
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
+dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
+our task abstraction as follows\footnote{The spreadsheet example that uses
+the \hs{INDIRECT} function can be expressed very similarly: simply replace the
+line containing the \cmd{if} statement with \hs{fetch ("A" ++ show c1)}.}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+sprsh2 :: Task Monad String Integer
+sprsh2 fetch "B1" = Just $ do c1 <- fetch "C1"
+                              if c1 == 1 then fetch "A1" else fetch "A2"
+sprsh2 _     _    = Nothing
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
+rebuilds every dependency it encounters, it is easy for it to handle dynamic
+dependencies.
+
+\todo{AM}{Describe another simple build system that does topological sort and
+builds everything in order. It will not be able to handle dynamic dependencies.}
 
@@
 \label{def-correct}
-
-\subsection{Monadic tasks}\label{sec-task-monad}
-
-As discussed in~\S\ref{sec-background}, some build tasks have dynamic
-dependencies, which are determined by values of intermediate computations. In
-our framework, such build tasks correspond to \emph{monadic task}, i.e.
-the type \hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
-example:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{bash}
-A1 = 10
-A2 = 20
-B1 = IF(C1=1,A1,A2)
-C1 = 1
-\end{minted}
-\vspace{1mm}
-
-\noindent
-The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
-dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
-our task abstraction as follows\footnote{The spreadsheet example that uses
-the \hs{INDIRECT} function can be expressed very similarly: simply replace the
-last line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-select :: Task Monad String Integer
-select fetch key | key /= "B1" = Nothing
-                 | otherwise   = Just $ do
-                     c1 <- fetch "C1"
-                     if c1 == 1 then fetch "A1" else fetch "A2"
-\end{minted}
-\vspace{1mm}
-
-We cannot statically analyse \hs{select} using the function \hs{dependencies}
-from the previous section, but we can still \hs{execute} it on a given store:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store key = if key == "A1" then 10 else 20
-@\ghci@ execute select store "B1"
-Just 20
-\end{minted}
-\vspace{1mm}
-
-\noindent
-In the example above $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
-To extract the dependencies of \hs{select} we introduce the function \hs{track}:
-a combination of \hs{execute} and \hs{dependencies} that takes the store~\store
-as input and computes both the resulting value and the list of its dependencies:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-import Control.Monad.Writer
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
-track task store = fmap runWriter . task (\k -> writer (store k, [k]))
-\end{minted}
-\vspace{1mm}
-
-\noindent
-The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
-we feed \hs{fetch k = writer (store k, [k])} to the compute, which in addition
-to fetching a value from the store, tracks the corresponding key. We can
-\hs{track} both applicative and monadic compute:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store key = if key == "A1" then 10 else 20
-@\ghci@ track add store "B1"
-Just (30,["A1","A2"])
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ track select store "B1"
-Just (20,["C1","A2"])
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store' key = if key == "A1" then 10 else 1
-@\ghci@ track select store' "B1"
-Just (10,["C1","A1"])
-\end{minted}
-\vspace{1mm}
-
-\noindent
-As expected, the dependencies of \hs{select} are determined by the contents of
-the store.
 
@@
 
+\subsection{Left-over stuff, need to relocate}
+
+We cannot statically analyse \hs{select} using the function \hs{dependencies}
+from the previous section, but we can still \hs{execute} it on a given store:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store key = if key == "A1" then 10 else 20
+@\ghci@ execute select store "B1"
+Just 20
+\end{minted}
+\vspace{1mm}
+
+\noindent
+In the example above $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
+To extract the dependencies of \hs{select} we introduce the function \hs{track}:
+a combination of \hs{execute} and \hs{dependencies} that takes the store~\store
+as input and computes both the resulting value and the list of its dependencies:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+import Control.Monad.Writer
+\end{minted}
+\vspace{0.5mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
+track task store = fmap runWriter . task (\k -> writer (store k, [k]))
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
+we feed \hs{fetch k = writer (store k, [k])} to the compute, which in addition
+to fetching a value from the store, tracks the corresponding key. We can
+\hs{track} both applicative and monadic compute:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store key = if key == "A1" then 10 else 20
+@\ghci@ track add store "B1"
+Just (30,["A1","A2"])
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ track select store "B1"
+Just (20,["C1","A2"])
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store' key = if key == "A1" then 10 else 1
+@\ghci@ track select store' "B1"
+Just (10,["C1","A1"])
+\end{minted}
+\vspace{1mm}
+
+\noindent
+As expected, the dependencies of \hs{select} are determined by the contents of
+the store.
+
 % \subsection{Recognising an applicative compute wearing a monadic hat}

### 2018-03-14 paper/3-abstractions.tex
Finish S3.3 and S3.4

@@
 
-\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-\vspace{1mm}
 
 \noindent
-As one can see, \hs{busy} correctly built both \cmd{B1} and \cmd{B2}, and in the
+As one can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
 right order (if it had built \cmd{B2} before building \cmd{B1}, the result would
@@
 This indicates that \hs{busy} can cope with tasks containing dynamic
-dependencies, which we address in the next subsection~\S\ref{sec-task-monad}.
+dependencies, which we discuss in the next subsection~\S\ref{sec-task-monad}.
+
+It is interesting to note that if \hs{busy} is given a \emph{functorial task}
+of type \hs{Task}~\hs{Functor}, which cannot use the application operator
+\hs{<*>}, then \hs{busy} can rejoice in knowing that it will never have to
+rebuild a key twice. Indeed, functorial tasks are very much like tail recursive
+functions -- they cannot branch into multiple subproblems, always leading to a
+straight-line build.
 
@@
 
-As discussed in~\S\ref{sec-background-excel}, some build tasks have dynamic
+As explained in~\S\ref{sec-background-excel}, some build tasks have dynamic
 dependencies, which are determined by values of intermediate computations. In
@@
 \noindent
+The main difference compared to \hs{sprsh1} is that the computation now takes
+place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
+\hs{fetch} different keys depending on whether \hs{c1}~\hs{==}~\hs{1} or not.
+
 Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
 rebuilds every dependency it encounters, it is easy for it to handle dynamic
-dependencies.
-
-\todo{AM}{Describe another simple build system that does topological sort and
-builds everything in order. It will not be able to handle dynamic dependencies.}
+dependencies. For minimal build systems, however, dynamic dependencies and hence
+monadic tasks are much more challenging, as we will see
+in~\S\ref{sec-implementations}.
 

### 2018-03-14 src/Development/Build/Task/Monad.hs
Add execute' as defined in the paper

@@ -2,7 +2,7 @@
 module Development.Build.Task.Monad (
     dependencies, track, trackM, transitiveDependencies, inputs, acyclic,
     consistent, correctBuild, execute, debugPartial, partial, exceptional,
-    staticDependencies, Script (..), getScript, runScript, isStatic
+    staticDependencies, Script (..), getScript, runScript, isStatic, execute'
     ) where
 
 import Control.Monad.Trans
@@ -12,6 +12,7 @@ import Control.Monad.Writer
 import Data.Functor.Identity
 import Data.Maybe
 
+import Development.Build.Store
 import Development.Build.Task
 import Development.Build.Utilities
 
@@ -70,6 +71,10 @@ correctBuild task before after key =
 -- that a given key is an input.
 execute :: Task Monad k v -> (k -> v) -> k -> Maybe v
 execute task fetch = fmap runIdentity . task (pure . fetch)
+
+-- The version used in the paper
+execute' :: Task Monad k v -> Store i k v -> k -> Maybe v
+execute' task store key = runIdentity <$> task (Identity . getValue store) key
 
 -- | Run a task with a partial lookup function. The result @Left k@ indicates
 -- that the task failed due to a missing dependency @k@. Otherwise, the

### 2018-03-14 paper/3-abstractions.tex
Finish S3.5

@@
 We can now say what it means for a build system to be \emph{correct}, something
-that is seldom stated formally:
+that is seldom stated formally. Before proceeding, we need to introduce a
+useful utility function called \hs{execute} that executes a \hs{task} on a
+\hs{key} by reading its dependencies from a supplied \hs{store}:
 
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+newtype Identity a = Identity { runIdentity :: a }
+\end{minted}
+\vspace{0mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+execute :: Task Monad k v -> Store i k v -> k -> Maybe v
+execute task store key = runIdentity <$> task (Identity . getValue store) key
+\end{minted}
+\vspace{1mm}
+
+\noindent
+This function is simpler than it might look: it fixes the `impedance mismatch'
+between the pure function \hs{getValue}~\hs{store}, whose type is \store, and
+the \hs{fetch} argument of the \hs{task}, whose type must be
+\hs{k}~\hs{->}~\hs{f}~\hs{v} for some \hs{f}. To fix the mismatch, we wrap the
+result of the pure function in the \hs{Identity} monad: the function
+\hs{Identity}~\hs{.}~\hs{getValue}~\hs{store} has the type
+\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
+result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now need to get rid
+of the \hs{Identity} wrapper by applying \hs{runIdentity} to the contents of
+\hs{Maybe}.
+
+\vspace{-1mm}
+\noindent\rule{\textwidth}{0.4pt}
+\vspace{-7mm}
 \definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
@@
     \hs{task} matches the value in the \hs{ideal} store:
+    \vspace{-1mm}
     \[
-    \hs{execute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{ideal}~\hs{k)}.
+    \hs{execute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
     \]
-    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.
-    \hs{result}~\hs{key}~\hs{==}~\hs{ideal}~\hs{key}. In other words, we require
-    that the output \hs{key} has the correct value, but do not impose any
-    restrictions on intermediate keys, therefore permitting shallow cloud builds.
-    \item The \hs{store}, \hs{ideal} and \hs{result} all agree on the inputs
-    that belong to the transitive closure of \hs{key}'s dependencies. This means
-    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
-    store has constraints not only on the output, but also on the inputs.
+    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
+    \vspace{-1mm}
+    \[
+        \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
+    \]
+    In other words, we require that the output \hs{key} has the correct value,
+    but do not impose any restrictions on intermediate keys, therefore
+    permitting shallow cloud builds.
+    \item The \hs{store}, \hs{result} and \hs{ideal} all agree on the inputs
+    that belong to the transitive closure of \hs{key}'s dependencies:
+    \[
+        \hs{and}~\hs{[}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{key}~\hs{]}
+    \]
+    \vspace{-6mm}
+    \[
+        \hs{and}~\hs{[}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{key}~\hs{]}
+    \]
+    where the implementation of \hs{inputs} will be provided in the next
+    subsection~\S\ref{sec-deps}, which is dedicated to the task dependency
+    analysis. This requirement guarantees that (i) no inputs were corrupted
+    during the build, and (ii)~the \hs{ideal} store has constraints not only
+    on the output, but also on the inputs.
 \end{itemize}
+A build system is \emph{correct} if it produces a correct \hs{result} for any
+given \hs{task}, \hs{key} and \hs{store}.
 }
 \label{def-correct}
+\vspace{-2mm}
+\rule{\textwidth}{0.4pt}
+
+It is hard to satisfy the above definition of correctness given a task
+description with cycles. All build systems discussed in this paper are correct
+only under the assumption that the given task description is acyclic. This
+includes the \hs{busy} build system introduced earlier: it will loop
+indefinitely given a cyclic \hs{task}. Some build systems provide a limited
+support for cyclic tasks, see~\S\ref{sec-iterative-compute}.
 
@@
 dependencies in case of an input.
-
-We can also \hs{execute} a compute on a given store. This relies on a similar
-transformation, now based on the \hs{Identity} monad (see
-Appendix~\S\ref{sec-appendix-transformers} for the implementation of
-\hs{Identity}):
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-import Data.Functor.Identity
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-execute :: Task Monad k v -> (k -> v) -> k -> Maybe v
-execute compute store = fmap runIdentity . compute (pure . store)
-\end{minted}
-\vspace{1mm}
-
-\noindent
-Note that \hs{execute} can accept both an applicative as well as a \emph{monadic
-task} introduced in~\S\ref{sec-task-monad}. Below we run \hs{execute} on
-the above \hs{add} example. Note that \hs{execute} returns \hs{Nothing} for
-inputs.
 

### 2018-03-15 paper/3-abstractions.tex
Add some text and TODOs for S3.6-S3.7

@@
 Despite limitations, \hs{busy} can easily handle the example \hs{sprsh1}
-from the previous subsection~\S\ref{sec-task}. Below we initialise the store with
-\cmd{A1} set to 10 and all other cells set to 20.
+from the previous subsection~\S\ref{sec-task}. In the GHCi session below we
+initialise the store with \cmd{A1} set to 10 and all other cells set to 20.
 
@@
     permitting shallow cloud builds.
-    \item The \hs{store}, \hs{result} and \hs{ideal} all agree on the inputs
-    that belong to the transitive closure of \hs{key}'s dependencies:
+    \item The \hs{result}, \hs{store} and \hs{ideal} all agree on the inputs
+    that belong to the transitive closure of \hs{key}'s dependencies with
+    respect to the \hs{result}:
     \[
-        \hs{and}~\hs{[}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{key}~\hs{]}
+        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{store}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
     \]
@@
     \[
-        \hs{and}~\hs{[}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{key}~\hs{]}
+        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
     \]
@@
 
-The \hs{Task} type and the example above might seem mysterious. Why insist on
-\hs{Task} to be polymorphic in \hs{f}? Why choose the \hs{Applicative}
-constraint?
+\todo{SLPJ}{Please edit as you see fit. We need \hs{dependencies} and
+\hs{track} to be defined for future sections. We also promised to define
+\hs{inputs} in the previous section, but I think it's not too critical and can
+be skipped/moved to Appendix if need be.}
 
-The answer to both questions is that we can statically analyse an applicative
-compute and extract all dependencies of a key using the following function:
+...
+
+We can statically analyse an applicative task and extract all dependencies of a
+key using the following function:
 
@@
 \noindent
-We can use this function to extract dependencies of \hs{add}, as shown in the
-following GHCi session:
+We can use this function to extract dependencies of \hs{sprsh1}:
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ dependencies add "A1"
+@\ghci@ dependencies sprsh1 "A1"
 [@@]
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ dependencies add "B1"
+@\ghci@ dependencies sprsh1 "B1"
 ["A1", "A2"]
@@
 \noindent
-It is important to note that we extract the dependencies without providing any
-values to the compute. This guarantees that no actual computation is performed
+It is important to note that we extract the dependencies without supplying any
+values to the build task. This guarantees that no actual computation is performed
 during the analysis -- we cannot execute the function \hs{(+)} if we have no
@@
 
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store key = if key == "A1" then 10 else 20
-@\ghci@ execute add store "A1"
-Nothing
-\end{minted}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ execute add store "B1"
-Just 30
-\end{minted}
-
-\subsection{Left-over stuff, need to relocate}
-
-We cannot statically analyse \hs{select} using the function \hs{dependencies}
-from the previous section, but we can still \hs{execute} it on a given store:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store key = if key == "A1" then 10 else 20
-@\ghci@ execute select store "B1"
-Just 20
-\end{minted}
-\vspace{1mm}
-
-\noindent
-In the example above $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
-To extract the dependencies of \hs{select} we introduce the function \hs{track}:
-a combination of \hs{execute} and \hs{dependencies} that takes the store~\store
-as input and computes both the resulting value and the list of its dependencies:
+To extract the dependencies of monadic tasks we introduce the function \hs{track}:
+a combination of \hs{execute} and \hs{dependencies} that takes a store as input
+and computes both the resulting value and the list of its dependencies:
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
-track task store = fmap runWriter . task (\k -> writer (store k, [k]))
+track :: Task Monad k v -> Store i k v -> k -> Maybe (v, [k])
+track task store = fmap runWriter . task (\k -> writer (getValue store k, [k]))
 \end{minted}
@@
 
-\noindent
 The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
-we feed \hs{fetch k = writer (store k, [k])} to the compute, which in addition
-to fetching a value from the store, tracks the corresponding key. We can
-\hs{track} both applicative and monadic compute:
+we feed \hs{fetch k = writer (getValue store k, [k])} to the task, which in
+addition to fetching a value from the store, tracks the corresponding key. We
+can \hs{track} both applicative and monadic tasks:
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store key = if key == "A1" then 10 else 20
-@\ghci@ track add store "B1"
+@\ghci@ store = initialise () (\key -> if key == "A1" then 10 else 20)
+@\ghci@ track sprsh1 store "B1"
 Just (30,["A1","A2"])
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ track select store "B1"
+@\ghci@ track sprsh2 store "B1"
 Just (20,["C1","A2"])
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store' key = if key == "A1" then 10 else 1
-@\ghci@ track select store' "B1"
+@\ghci@ store' = initialise () (\key -> if key == "A1" then 10 else 1)
+@\ghci@ track sprsh2 store' "B1"
 Just (10,["C1","A1"])
@@
 \noindent
-As expected, the dependencies of \hs{select} are determined by the contents of
+As expected, the dependencies of \hs{sprsh2} are determined by the contents of
 the store.
+
+...
+
+Here is a draft implementation of \hs{inputs} used in the definition of
+build system correctness in \S\ref{sec-build-correctness}:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
+inputs task store key = filter (isInput task) (closure deps key)
+  where
+    deps k = maybe [] snd (track task store k)
+
+closure :: Eq a => (a -> [a]) -> a -> [a] -- Standard graph transitive closure
+
+data Proxy a = Proxy
+
+isInput :: Task Monad k v -> k -> Bool
+isInput task = isNothing . task (const Proxy)
+\end{minted}
+
+\subsection{Free tasks (?)}
+
+\todo{AM}{Introduce free constructions \hs{TaskA} and \hs{TaskM} giving
+intuitions on what tasks really are. Requires some further discussion with Neil.
+This will take at least one page.}
 

### 2018-03-15 paper/3-abstractions.tex
Work on S4

@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-busy :: (Eq k, Hashable v) => Build Monad () k v
+busy :: (Eq k, Hashable v) => Build Monad () k v -- Drop Hashable? It's confusing here
 busy task key store = execState (compute key) store

### 2018-03-15 paper/4-build.tex
Work on S4

@@
-\section{Build systems, abstractly}\label{sec-build}
+\section{Build Systems \`a la Carte}\label{sec-build}
 
-In this section we describe our \hs{Build} abstraction, defined as:
+The focus of this paper is on a variety of implementations of \hs{Build}, given
+a client-supplied implementation of \hs{Task}. That is, we are going to take
+\hs{Task} as given from now on, and explore \hs{Build} variants: first
+abstractly (in this section) and then concretely in~\S\ref{sec-implementations}.
 
-\begin{minted}{haskell}
-type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
-\end{minted}
+Let \hs{result}~\hs{=}~\hs{build}~\hs{task}~\hs{key}~\hs{store}. The build
+system \hs{build} is responsible for ensuring that \hs{key} and its dependencies
+are up-to-date in \hs{result} starting from \hs{store} (see the formal definition
+of correctness in~\S\ref{sec-build-correctness}). As part of that work,
+\hs{build} \emph{may} store auxiliary information \hs{info} of type \hs{i} in
+the \hs{store}, which will be provided to future calls of \hs{build}.
 
-The build function takes 4 type arguments: \hs{c} is the constraint used for \hs{Compute} (typically either \hs{Applicative} or \hs{Monad}), \hs{i} is the information stored by the build system and \hs{k}/\hs{v} are the key/value types as described in \S\ref{sec-vocabulary}.
-Given a build system \hs{build}:
+All the build systems we define are polymorphic in both \hs{k} and \hs{v}
+(perhaps adding constraints to them), and have specific instantiations of \hs{c}
+and \hs{i}. As an example, recall that the \hs{busy} build system defined
+in~\S\ref{sec-general-build}, has \hs{c}~\hs{=}~\hs{Monad}, i.e. it supports
+monadic tasks, and \hs{i = ()}, i.e. it does not store any build information.
 
-\begin{minted}{haskell}
-build compute key info store = (info', store')
-    where ...
-\end{minted}
+According to the definition of minimality~\ref{def-minimal}, a minimal build
+system must \textbf{rebuild only out-of-date keys} and at most once. The only
+way to achieve the ``at most once'' requirement is to \textbf{build all keys in
+the order that respects their dependencies}.
 
-The function \hs{build} is responsible for ensuring that \hs{key} and its dependencies are up-to-date in \hs{store'} starting from \hs{store}.
-As part of that work, the \hs{build} function \textit{may} store other information in \hs{info'}, which will be provided as \hs{info} in future calls to \hs{build}.
-The \hs{info} parameter will be \hs{Nothing} on the very first call to \hs{build}.
-We assume that \hs{store} is total, where "missing files" can be represented by some sentinel value of \hs{v} if necessary.
+We have emphasised two different phrases above, and tackle each aspect separately.
 
-All the \hs{Build} types we define are polymorphic in both \hs{k} and \hs{v} (perhaps adding constraints to them), and have specific instantiations of \hs{c} and \hs{i}.
-How to implement \hs{Build} can be captured by the idea:
-
-\begin{quote}
-A build system must build \underline{out-of-date keys} while \underline{respecting dependency orderings}.
-\end{quote}
-
-We have underlined two different phrases in this sentence, and tackle each aspect separately.
-
-\subsection{Respecting dependency orderings}
+\subsection{Respecting the dependency order}
 \label{sec-dependency-orderings}
 
-If the \hs{Compute} function for \hs{k} calls the functional argument of \hs{d} then the build system is responsible for ensuring that \hs{d} is built \textit{before} \hs{k}.
-Looking at the build systems in \S\ref{sec-background} we have observed 3 distinct approaches to respecting dependency orderings.
-In this section of the paper we'll explore what properties are a consequence and how they operate.
+The build systems overview (\S\ref{sec-background-summary}) highlighted three
+distinct approaches to respecting the dependency order. This subsection explores
+their properties and possible implementations.
 
-\subsubsection{Topological sort}
+\subsubsection{Topological}
 
-The topological approach creates a linear ordering, which when followed, ensures each build is correct.
-Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort.
-However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{sec-deps}.
+The topological approach computes a linear order, which when followed, ensures
+the \hs{build} is correct regardless of the initial \hs{store}. Given a function
+from a key to its dependencies, and the output \hs{key}, you can compute the
+linear order by first taking the transitive closure of \hs{key}'s dependencies,
+and then computing its topological sort. However, as we have seen
+in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task,
+which requires the build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling
+out dynamic dependencies.
 
@@
 
-The topological sort approach has two downside - it is limited to \hs{Applicative} build systems and requires a fresh topological sort each time - while the actions themselves may be incremental the pre-processing is not.
-An alternative approach is to keep the topological order between runs and assume it to be correct, but if execution determines it is wrong, reorder.
+The topological approach has two downsides: it is limited to \hs{Applicative}
+build systems and requires a fresh topological sort each time -- while the
+actions themselves may be incremental the pre-processing is not. An alternative
+approach is to store the topological order between build runs and assume it to
+be correct, but if the build determines it is wrong, reorder.
 
-This requires a way to abort computations that have failed.
-It's also not minimal in the sense that computations start, may do some meaningful work, then abort.
-However, in the case of an \hs{Applicative} system, that work is zero.
-It does require the ability to terminate a running computation.
+This requires a way to abort tasks that have failed due to out-of-date
+dependencies. It is also not minimal in the sense that tasks start, may do some
+meaningful work, then abort. However, in the case of an \hs{Applicative} system,
+that work is zero.
 
@@
 
-An alternative way is to simply build dependencies when they are requested.
-By combining that with a transient store of which keys have been built in this execution, you can describe dependencies in a simple way.
-You need a way to suspend a running action, which can be done with cheap green-threads and blocking (the original approach of Shake) or with a continuation-passing style (what Shake does now).
-An alternative approach to suspending a computation is to abort it and restart it again later, at the cost of doing additional work.
+An alternative approach, utilised by the \hs{busy} build system
+(\S\ref{sec-general-build}) is to simply build dependencies when they are
+requested. By combining that with a transient store of which keys have already
+been built, you can obtain a minimal build system.
 
-\subsection{Determining what is out-of-date}
+This requires a way to suspend a running task, which can be done with cheap
+green threads and blocking (the original approach of \Shake) or with a
+continuation-passing style (what \Shake does now). An alternative approach to
+suspending a task is to abort it and restart it again later, at the cost
+of doing additional work.
 
-The second part, determining what to rebuild, can be addressed in one of three fundamental ways, with a number of tweaks and variations within them.
+\subsection{Determining out-of-date keys}
 
-\subsubsection{A dirty bit}
-\label{sec-dirty-bit}
+The second aspect, determining what to rebuild, can be addressed in one of three
+fundamental ways, with a number of tweaks and variations within them.
 
-The idea of a dirty bit is to have one piece of information, whether a key is dirty or clean.
-After a run, all bits are set to clean.
-When a run starts anything that changed between the two states is marked dirty - by marking additional things dirty/clean the build system can track what needs to rebuild.
-When reaching a key, if it and all its dependencies are clean, the key does not need recomputing.
+\subsubsection{A dirty bit}\label{sec-dirty-bit}
 
-\Excel models the dirty bit approach most directly, having an actual dirty bit associated with each cell.
-Since Excel is the only manner by which a cell may be changed by the user, it can dirty the bits as necessary.
-When rebuilding if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and marked dirty itself.
-The only wrinkle in this scheme is that \Excel supports monadic build systems, and does not separately record what rebuilt, so it has to approximate in this case.
+The idea of a dirty bit is to have one piece of information, whether a key is
+\emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
+next build starts anything that changed between the two states is marked
+dirty~--~by marking additional things dirty/clean the build system can track
+what needs to rebuild. When reaching a key, if it and all its dependencies are
+clean, the key does not need recomputing.
 
-\Make uses file modification times, and compares files to their dependencies, which can be thought of as a dirty bit which is set when a file is newer than its dependencies.
-The interesting property of this dirty bit is that it is not under control of \Make, but an existing bit that is repurposed.
-In particular, updating a file automatically clears its dirty bit, and automatically sets the dirty bit of the nodes depending on it.
-One thing \Make does require is that file timestamps only go forward in time - something that can be violated by some backup software.
+\Excel models the dirty bit approach most directly, having an actual dirty bit
+associated with each cell. Since \Excel is the only manner by which a cell may
+be changed by the user, it can mark cells dirty as necessary. When rebuilding,
+if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and
+marked dirty itself.
 
-With a dirty bit it is possible to achieve minimality.
-To achieve cut-off it would be important to not set the dirty bit after a computation that did not change the value.
-\Excel could model this approach, but does not.
-In contrast, \Make cannot model cut-off - to do so it would have to mark the node clean (so it didn't rebuild in the next run) and at the same time not mark the things it depends on dirty - an impossible task with only the ability to update to the latest modification time.
+% AM: I didn't understand this bit, so commented it out.
+% The only wrinkle in this scheme is that \Excel supports
+% monadic tasks, and does not separately record the rebuilt, so it has
+% to approximate in this case.
+
+\Make uses file modification times, and compares files to their dependencies,
+which can be thought of as a dirty bit which is set when a file is newer than
+its dependencies. The interesting property of this dirty bit is that it is not
+under control of \Make, but an existing bit that is repurposed. In particular,
+rebuilding a file automatically clears its dirty bit, and automatically sets
+the dirty bit of the nodes depending on it. One thing \Make does require is that
+file timestamps only go forward in time -- something that can be violated by
+backup software.
+
+With a dirty bit it is possible to achieve minimality. To achieve early cutoff
+(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
+after a computation that did not change the value. \Excel could model this
+approach, but does not. In contrast, \Make cannot model early cutoff -- to do so
+it would have to mark the node clean (so it didn't rebuild in the next run) and
+at the same time not mark the things it depends on dirty -- an impossible task
+with only the ability to update to the latest modification time.
 

### 2018-03-15 paper/4-build.tex
Add TODOs for S4

@@
 
-An alternative way to determine if a key is dirty is to record what state the values/hashes of dependencies were used at last time, and if something has changed, the key is dirty and must be recomputed - in essence a trace with which we can use to \textit{verify} existing values.
-We can describe a trace as:
+An alternative way to determine if a key is dirty is to record what state the
+values/hashes of dependencies were used at last time, and if something has
+changed, the key is dirty and must be recomputed -- in essence a \emph{trace}
+which we can use to \textit{verify} existing values. We can describe a trace as:
+
+\todo{AM/NM}{Change from \hs{[Trace k v]} to \hs{Map k (Trave k v)} or something
+similar, so that it's obvious we have one trace per key .}
 
@@
 data Trace k v = Trace
-    {key :: k
-    ,dependencies :: [(k, Hash v)]
-    ,result :: Hash v
-    }
+    { key          :: k
+    , dependencies :: [(k, Hash)]
+    , result       :: Hash }
 \end{minted}
 
-We assume that \hs{Hash v} is a small constant size, constructed from hashing the underlying \hs{v} rather than storing it directly.
-Checking a trace requires ensuring all the dependencies are up-to-date (using whatever ordering strategy as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are same as the current value and that the result is the same.
-After computing a fresh value we add its \hs{Trace} to the list.
-Therefore the information stored by a build system that verifies traces is \hs{[Trace k v]}.
+We assume that \hs{Hash} is a small constant size, constructed from hashing the
+underlying \hs{v} rather than storing it directly. Checking a trace requires
+ensuring all the dependencies are up-to-date (using whatever ordering strategy
+as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are
+same as the current value and that the result is the same. After computing a
+fresh value we add its \hs{Trace} to the list. Therefore the information stored
+by a build system that verifies traces is \hs{[Trace k v]}.
 
@@
 
+\todo{AM/NM}{Traces -- why the indirection (first get the hash, then look it up)?
+Can we drop it? Can we switch to a more efficient/explicit data structure than a list?}
+
 \begin{minted}{haskell}
 data Traces k v = Traces
-    {traces :: [Trace k v]
-    ,contents :: Map (Hash v) v
-    }
+    { traces :: [Trace k v]
+    , cache  :: Map Hash v }
 \end{minted}
@@
 We have a list of traces, plus a \hs{Map} from the hash to the actual contents.
-Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{contents} \textit{without} recomputing it.
+Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \textit{without} recomputing it.
 
@@
 
-\todo{AM}{AM: I would also like to see an actual \hs{upToDate} function for verifying, and a \hs{recreate} for `recreating'.
+\todo{AM}{AM: I would also like to see actual \hs{verify} and \hs{construct} functions.
+
 NM: It actually gets a bit more confusing than that - since you end up needing to pass functions to get yourself, your dependencies etc.
 Unclear there is a win.
-The functions would occur in the next section anyway.}
+The functions would occur in the next section anyway.
+
+AM: Now, after we've switched to \hs{Store i k v} it might work better?}

### 2018-03-15 paper/4-build.tex
Add 3x3 table

@@
 AM: Now, after we've switched to \hs{Store i k v} it might work better?}
+
+\subsection{Summary}
+
+\todo{AM/NM/SLPJ}{Add a decription for the table.}
+
+\begin{table}[h]
+\smaller
+\centering
+\begin{tabular}{l||c|c|c}
+\hline
+Property           & Topological & Reordering & Recursive    \\\hline
+\hline
+Dirty bit          & \Make                   & \Excel                   & \Shake without early cutoff \\\hline
+Verifying trace    & \Make with early cutoff & \Excel with early cutoff & \Shake                      \\\hline
+Constructive trace & \Bazel                  & Cloud \Excel             & Cloud \Shake                \\\hline
+\end{tabular}
+\vspace{0.5mm}
+\caption{Build systems \`a la carte.\label{tab-build-systems}}
+\end{table}

### 2018-03-15 paper/1-intro.tex
Lots more from Simon

@@
 of the software ecosystem, very much a means to an end, and seldom the
-focus of attention.  Despite this, the challenges of scale have driven
-large software firms like Microsoft, Facebook, and Google to develop
-their own build systems, each with its own choices and idiosyncrasies.
-
+focus of attention.
 Seldom do people ask questions like ``What does it mean for my build
 system to be correct?'' or ``What are the trade-offs between different
-approaches?''.  Complex build systems use subtle algorithms, but they
+approaches?''.  Moreover, complex build systems use subtle algorithms, but they
 are often hidden away, and not the object of study.
 
-In this paper we offer a general framework in which to discuss these
-questions in a way that is both abstract (omitting incidental detail)
+Recently, the challenges of scale have driven large software firms
+like Microsoft, Facebook, and Google to develop their own build
+systems, each with its own choices and idiosyncrasies.
+In this paper we offer a general framework in which to understand
+and compare build systems,
+in a way that is both abstract (omitting incidental detail)
 and yet precise (implemented as Haskell code).  Specifically we make
@@
 \begin{itemize}
-\item We describe some simple but novel abstractions that
-crisply encapsulate what a build system is.
-\item We show that we can instantiate
-these abstractions to describe the essence of a variety of different
-build systems, including \Make, \Shake, \Bazel, and \Excel, each in
-a dozen lines of code or so.
-Doing this in a single setting allows
-the differences and similarities between these huge systems to be
-brought out clearly.
 \item Build systems vary on many axes,
@@
 early cut-off; self-tracking build systems; and persistent metadata.
-These properties (which we define in~\S\ref{sec-background}) are often
-deeply built into any particular build system.
-In contrast, our framework allows them to be distinguished,
-reasoned about, and varied.
-\item We show that all build system we look at can be broken along two axes - the order in which they build dependencies, and whether they decide to build a particular dependency.
-Interestingly, these two aspects can be recombined without restriction to produce new build systems.
-\item Two particularly desirable properties are dynamic dependencies
-and cloud-build; yet no currently-available build system supports
-both.  Our framework makes it rather easy to do so, for the first
-time.
+In \S\ref{sec-background} we identify some key properties, in the context of
+four carefully-chosen build systems.
+\item We describe some simple but novel abstractions that
+  crisply encapsulate what a build system is (\S\ref{sec-abstractions}),
+  allowing us, for example, to speak about what it means for a build system to be correct.
+
+\item We identify two key design choices
+  that are typically deeply built into any build system:
+  \emph{the order in which dependencies are built} (\S\ref{sec-dependency-orderings})
+  and \emph{whether or not a dependency is (re-)built} (\S\ref{sec-out-of-date}).
+  These choices turn out to be orthogonal, which leads us to a new
+  classification of the design space (\S\ref{sec-design-space}).
+
+\item We show that we can instantiate
+our abstractions to describe the essence of a variety of different
+real-life build systems, including \Make, \Shake, \Bazel, and \Excel, each in
+a dozen lines of code or so (\S\ref{sec-implementations}).
+Doing this in a single setting allows
+the differences and similarities between these huge systems to be
+brought out clearly.
+
+\item Moreover, we can readily mix the ingredients differently to produce the first
+  build system that supports both \emph{dynamic dependencies}
+  and \emph{cloud build} (\S\ref{sec-implementation-cloud-shake}).
+
 \end{itemize}
 In short, instead of seeing build systems as unrelated
-points in space, we can re-envisage them as locations in a landscape,
+points in space, we now see them as locations in a landscape,
 leading to better understanding of what they do and how they compare,
@@
 landscape.
+We discuss engineering aspects in \S\ref{sec-engineering}, and related
+work in \S\ref{sec-related}.
 

### 2018-03-15 paper/3-abstractions.tex
Lots more from Simon

@@
 
-We start by establishing a common vocabulary for build systems that allows
-us to abstract away from specific application domains, such as software
-development or spreadsheets.
-
 \subsection{Common vocabulary for build systems}\label{sec-vocabulary}
-
 \begin{figure}
@@
 \end{figure}
+We begin by establishing a common vocabulary for build systems.
+% that allows
+% us to abstract away from specific application domains, such as software
+% development or spreadsheets.
+Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
+and gives types of the library functions we use in this paper.
 
@@
 
-In software build systems, keys are typically filenames,
+\paragraph{Keys and values.} In software build systems, keys are typically filenames,
 e.g.~\cmd{main.c}, and values are file contents (a C program source
@@
 variables \hs{k} and \hs{v} to denote keys and values, respectively.
-Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
-and gives types of the library functions we use in this paper.
 
+\paragraph{Input, output, and intermediate values}
 Some values must be provided by the user as \emph{input}. For example,
@@
 
+\paragraph{The store.}
+A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
+values. It is convenient to assume that a store is total, i.e. it contains a
+value for every possible key. We therefore also assume that the type of values
+is capable of encoding values corresponding to non-existent keys (missing files
+or empty cells).
+
+\paragraph{Hashes.}
 We use a cryptographic \emph{hash function}
@@
 data type equipped with the equality test.
-
-A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
-values. It is convenient to assume that a store is total, i.e. it contains a
-value for every possible key. We therefore also assume that the type of values
-is capable of encoding values corresponding to non-existent keys (missing files
-or empty cells). In addition to usual \hs{getValue} and \hs{putValue} operations,
+In addition to usual \hs{getValue} and \hs{putValue} operations,
 some stores support a dedicated \hs{getHash} operation, which is implemented
@@
 example, Git Virtual File System~\cite{gvfs} stores file hashes and downloads
-file contents only on demand.}. The store also holds a value of type \hs{i},
+file contents only on demand.}.
+
+\paragraph{Persistent build information.}
+The store also holds a value of type \hs{i},
 representing the persistent build information maintained by the build
-system~--~its ``memory''. Given an \hs{i} and a key-value map \store, we can
-\hs{initialise} a store.
+system~--~its ``memory''.
+
+% Given an \hs{i} and a key-value map \store, we can
+% \hs{initialise} a store.
 
@@
 
-Our first main idea is the task-description abstraction:
+Our first main abstraction is for \emph{task descriptions}:
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 -- the relevant type signatures are given in Fig.~\ref{fig-types}.   We will
-return to the need for this generality in~\S\ref{sec-deps}.
+explain why in the next subsection (\S\ref{sec-general-build}).
 
@@
   where
+    compute :: k -> State (Store i k v) v
     compute k = case task compute k of
@@
 \noindent
-The \hs{busy} build system uses a function
-\hs{compute}~\hs{::}~\hs{k}~\hs{->}~\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)}~\hs{v},
-which given a key brings the associated value up-to-date and returns it. To do
-so, \hs{compute} calls \hs{task} passing itself as an argument, so that \hs{task}
-can use \hs{compute} to recursively find the values of its dependencies. If
-\hs{task} returns \hs{Nothing} (this key is an input), \hs{compute} simply reads
-the result from the store~\hs{s}, which is available in the \hs{State} monad (in
-which \hs{compute} runs) via the \hs{get} method~--~see the \hs{State} monad
-interface in Fig.~\ref{fig-types}. Otherwise it runs the action \hs{act}
-returned by the \hs{task} to produce a resulting value~\hs{v}, stores it
-(modifying the store \hs{s}), and returns it.
+The \hs{busy} build system uses a helper function
+\hs{compute} that, given a key, brings the key up to date in the store, and returns its value.
+The function \hs{compute} runs in the standard Haskell \hs{State} monad -- see
+Figure~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
+To bring a key up to date, \hs{compute} asks the task description \hs{task} how
+to compute \hs{k}.
+If \hs{task} returns \hs{Nothing} the key is an input, so \hs{compute} simply reads
+the result from the store.
+Otherwise \hs{compute} runs the action \hs{act}
+returned by the \hs{task} to produce a resulting value~\hs{v}, records
+the new key/value mapping in the store, and returns \hs{v}.
+Notice that \hs{compute} passes itself to \hs{task} as an argument,
+so that the latter can use \hs{compute} to recursively find the
+values of \hs{k}'s dependencies.
+
+Now we can see why the \hs{Task} abstraction needed to be abstracted over \hs{f}:
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+\end{minted}
+The \hs{busy} build system instantiates \hs{c} to \hs{Monad} and \hs{f} to \hs{State (Store i k v)},
+so that the call-back to \hs{fetch :: k -> f v} can side-effect the \hs{Store},
+thereby allowing successive calls to \hs{fetch} to communicate with one another.
+We will see why we want \hs{Task} to be \emph{polymorphic} in \hs{f} in \S\ref{sec-deps}.
 
 Given an acyclic task description, the \hs{busy} build system terminates with a
-correct result, but it is not a minimal build system (c.f. the
-Definition~\ref{def-minimal}). Since \hs{busy} has no memory (\hs{i = ()}), it
+correct result, but it is not a \emph{minimal} build system
+(Definition~\ref{def-minimal}). Since \hs{busy} has no memory (\hs{i=()}), it
 cannot keep track of keys it has already built, and will therefore busily
@@
 
-Despite limitations, \hs{busy} can easily handle the example \hs{sprsh1}
+Nevertheless, \hs{busy} can easily handle the example \hs{sprsh1}
 from the previous subsection~\S\ref{sec-task}. In the GHCi session below we
@@
 This indicates that \hs{busy} can cope with tasks containing dynamic
-dependencies, which we discuss in the next subsection~\S\ref{sec-task-monad}.
+dependencies, which we discuss next (\S\ref{sec-task-monad}).
 
@@
 
-As explained in~\S\ref{sec-background-excel}, some build tasks have dynamic
+As explained in~\S\ref{sec-background-excel}, some task descriptions have dynamic
 dependencies, which are determined by values of intermediate computations. In
-our framework, such build tasks correspond to the type
+our framework, such task descriptions correspond to the type
 \hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
@@
 \begin{minted}[xleftmargin=10pt]{text}
-A1: 10      B1: IF(C1=1,A1,A2)      C1: 1
-A2: 20
+A1: 10      B1: IF(C1=1,B2,A2)      C1: 1
+A2: 20      B2: IF(C1=1,A1,B1)
 \end{minted}
@@
 \noindent
-The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
-dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
+Here, \cmd{B1} and \cmd{B2} statically depend on each other, but Excel (which
+uses dynamic dependencies) is perfectly happy. We can express this using
 our task abstraction as follows\footnote{The spreadsheet example that uses
@@
 sprsh2 fetch "B1" = Just $ do c1 <- fetch "C1"
-                              if c1 == 1 then fetch "A1" else fetch "A2"
+                              if c1 == 1 then fetch "B2" else fetch "A2"
+sprsh2 fetch "B2" = Just $ do c1 <- fetch "C1"
+                              if c1 == 1 then fetch "A1" else fetch "B1"
 sprsh2 _     _    = Nothing
@@
 The main difference compared to \hs{sprsh1} is that the computation now takes
-place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
-\hs{fetch} different keys depending on whether \hs{c1}~\hs{==}~\hs{1} or not.
+place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
+then \hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.
 
@@
 rebuilds every dependency it encounters, it is easy for it to handle dynamic
-dependencies. For minimal build systems, however, dynamic dependencies and hence
-monadic tasks are much more challenging, as we will see
+dependencies. For minimal build systems, however, dynamic dependencies, and hence
+monadic tasks, are much more challenging, as we will see
 in~\S\ref{sec-implementations}.
@@
 We can now say what it means for a build system to be \emph{correct}, something
-that is seldom stated formally. Before proceeding, we need to introduce a
-useful utility function called \hs{execute} that executes a \hs{task} on a
-\hs{key} by reading its dependencies from a supplied \hs{store}:
+that is seldom stated formally. Our intuition is this: \emph{when the build system
+  returns, the specified key, and all its dependencies, should be up to date}.
+What does ``up to date'' mean?  It means that if we recompute the value of the
+key (using the task description, and the final store), we should get exactly
+the same value as we see in the final store.
 
+To express this formally we need an auxiliary
+function \hs{recompute}, that computes the value of a key in a given store
+\emph{without attempting to update any dependencies}:
 \vspace{1mm}
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-execute :: Task Monad k v -> Store i k v -> k -> Maybe v
-execute task store key = runIdentity <$> task (Identity . getValue store) key
+recompute :: Task Monad k v -> Store i k v -> k -> Maybe v
+recompute task store key = runIdentity <$> task (Identity . getValue store) key
 \end{minted}
 \vspace{1mm}
-
 \noindent
-This function is simpler than it might look: it fixes the `impedance mismatch'
+\todo{SLPJ}{Why monad?}
+Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
+we can use the standard Haskell \hs{Identity} monad.
+The use of \hs{Identity} just fixes the `impedance mismatch'
 between the pure function \hs{getValue}~\hs{store}, whose type is \store, and
@@
 key           :: k
-store  :: Store i k v
-result :: Store i k v
+store, result :: Store i k v
 result = build task key store
@@
     \[
-    \hs{execute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
+    \hs{recompute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
     \]
@@
 To extract the dependencies of monadic tasks we introduce the function \hs{track}:
-a combination of \hs{execute} and \hs{dependencies} that takes a store as input
+a combination of \hs{recompute} and \hs{dependencies} that takes a store as input
 and computes both the resulting value and the list of its dependencies:

### 2018-03-15 paper/4-build.tex
Lots more from Simon

@@
 The focus of this paper is on a variety of implementations of \hs{Build}, given
-a client-supplied implementation of \hs{Task}. That is, we are going to take
-\hs{Task} as given from now on, and explore \hs{Build} variants: first
+a \emph{client-supplied} implementation of \hs{Task}. That is, we are going to take
+\hs{Task} as given from now on, and explore variants of \hs{Build}: first
 abstractly (in this section) and then concretely in~\S\ref{sec-implementations}.
 
+\todo{SLPJ}{What does this para add?}
 Let \hs{result}~\hs{=}~\hs{build}~\hs{task}~\hs{key}~\hs{store}. The build
 system \hs{build} is responsible for ensuring that \hs{key} and its dependencies
-are up-to-date in \hs{result} starting from \hs{store} (see the formal definition
-of correctness in~\S\ref{sec-build-correctness}). As part of that work,
+are up-to-date in \hs{result} starting from \hs{store} (\S\ref{sec-build-correctness}). As part of that work,
 \hs{build} \emph{may} store auxiliary information \hs{info} of type \hs{i} in
@@
 
+\todo{SLPJ}{What does this para add?}
 All the build systems we define are polymorphic in both \hs{k} and \hs{v}
@@
 
-The topological approach computes a linear order, which when followed, ensures
+The topological approach pre-computes a linear order, which when followed, ensures
 the \hs{build} is correct regardless of the initial \hs{store}. Given a function
@@
 The topological approach has two downsides: it is limited to \hs{Applicative}
-build systems and requires a fresh topological sort each time -- while the
-actions themselves may be incremental the pre-processing is not. An alternative
-approach is to store the topological order between build runs and assume it to
-be correct, but if the build determines it is wrong, reorder.
+build systems and requires a fresh topological sort each time.  So, while the
+actions themselves may be incremental (i.e. unnecessary tasks will not be performed),
+the pre-processing is not. We may try to incrementalise the topological sort by
+storing the topological order between build runs and assume it to
+be correct, but if the build determines it is wrong, fix it up.
 
 This requires a way to abort tasks that have failed due to out-of-date
-dependencies. It is also not minimal in the sense that tasks start, may do some
+dependencies. It is also not minimal in the sense that a task may start, do some
 meaningful work, then abort. However, in the case of an \hs{Applicative} system,
@@
 (\S\ref{sec-general-build}) is to simply build dependencies when they are
-requested. By combining that with a transient store of which keys have already
+requested. By combining that with a transient log of which keys have already
 been built, you can obtain a minimal build system.
@@
 
-\subsection{Determining out-of-date keys}
+\subsection{Determining out-of-date keys} \label{sec-out-of-date}
 
@@
 
-The idea of a dirty bit is to have one piece of information, whether a key is
+The idea of a dirty bit is to have one piece of persistent
+information per key, saying whether the key is
 \emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
 next build starts anything that changed between the two states is marked
-dirty~--~by marking additional things dirty/clean the build system can track
-what needs to rebuild. When reaching a key, if it and all its dependencies are
+dirty.
+% ; and by marking additional things dirty/clean the build system can track
+% what needs to rebuild.
+When reaching a key, if it and all its dependencies are
 clean, the key does not need recomputing.
@@
 \Excel models the dirty bit approach most directly, having an actual dirty bit
-associated with each cell. Since \Excel is the only manner by which a cell may
-be changed by the user, it can mark cells dirty as necessary. When rebuilding,
+associated with each cell, marking the cell it (and its dependencies) dirty
+if the user modifies it.
+When rebuilding,
 if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and
 marked dirty itself.
+\todo{SLPJ}{What does ``marked dirty itself'' mean?}
 
@@
 
-\Make uses file modification times, and compares files to their dependencies,
-which can be thought of as a dirty bit which is set when a file is newer than
-its dependencies. The interesting property of this dirty bit is that it is not
-under control of \Make, but an existing bit that is repurposed. In particular,
-rebuilding a file automatically clears its dirty bit, and automatically sets
-the dirty bit of the nodes depending on it. One thing \Make does require is that
-file timestamps only go forward in time -- something that can be violated by
-backup software.
+\Make uses file modification times, and compares files to their
+dependencies, which can be thought of as a dirty bit which is set when
+a file is newer than its dependencies. The interesting property of
+this dirty bit is that it is not under control of \Make; rather it is
+existing file-system information that is repurposed. In particular,
+rebuilding a file automatically clears its dirty bit, and
+automatically sets the dirty bit of the nodes depending on it. One
+thing \Make does require is that file timestamps only go forward in
+time -- something that can be violated by backup software.
 
@@
 
-\subsection{Summary}
+\subsection{Summary} \label{sec-design-space}
 

### 2018-03-15 paper/5-implementations.tex
Lots more from Simon

@@
 \subsection{Framework}
+
+todo{SLPJ}{I think this entire section, including the mysterious M, can be deleted.  Or is
+there anything that remains?}
 
@@
 The \Make build system builds files in a linear order based on a topological sort.
-For each file, it builds it if the output file is older than any of the inputs.
-We can express this in our framework with:
+(See \S\ref{sec-concurrency} for parallel \Make.)
+For each file, it builds it if the output file is older than any of the inputs. We can express this in our framework with:
 
@@
 We use the helper function \hs{reordering} which provides the key to build (\hs{k}), the dependencies of \hs{k} (\hs{dk}) and an action to build \hs{k} (\hs{act}).
-The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Compute} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway.
+The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Task} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway.
 If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possibilities:

### 2018-03-15 paper/6-engineering.tex
Lots more from Simon

@@
 
+\todo{SLPJ}{Can we replace ``node'' with ``task''? Everywhere.}
 \begin{enumerate}
@@
 
-In our model we define \hs{Compute} as a function - when give the same inputs it will always produce the same outputs.
+In our model we define \hs{Task} as a function - when give the same inputs it will always produce the same outputs.
 Alas, the real-world is not as obliging.
-Some examples of impure \hs{Compute} functions include:
+Some examples of impure \hs{Task} functions include:
 
@@
 
-One tempting way of modelling non-determinism is to enrich \hs{Compute} from \hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus} respectively - both provide the underlying class with zero and choice.
+One tempting way of modelling non-determinism is to enrich \hs{Task} from \hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus} respectively - both provide the underlying class with zero and choice.
 However, that seems to encode a different type of non-determinism - a kind of dependency-level non-determinism when most rules really experience a value-level non-determinism.

### 2018-03-15 paper/7-related.tex
Lots more from Simon

@@
 
-Taking our abstraction, it is possible to encode \Fabricate assuming that nodes like \cmd{gcc -c util.c} are keys, there is a linear dependency between each successive node, and that the OS-level tracing can be lifted back as a monadic \hs{Compute} function\footnote{In fact, \Shake has an execution mode that can model \Fabricate{}-like build systems -- see \hs{Development.Shake.Forward} in the \Shake library.}.
+Taking our abstraction, it is possible to encode \Fabricate assuming that nodes like \cmd{gcc -c util.c} are keys, there is a linear dependency between each successive node, and that the OS-level tracing can be lifted back as a monadic \hs{Task} function\footnote{In fact, \Shake has an execution mode that can model \Fabricate{}-like build systems -- see \hs{Development.Shake.Forward} in the \Shake library.}.
 However, in our pure model the mapping is not perfect as \hs{gcc} writes to arbitrary files whose locations are not known in advance.
@@
 
-The \hs{Compute} definition of:
+The \hs{Task} definition of:
 
 \begin{minted}{haskell}
-type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
 \end{minted}

### 2018-03-15 src/Development/Build/Plan.hs
Add verifying trace

@@ -1,4 +1,7 @@
 module Development.Build.Plan (
+    -- Traces
+    VerifyingTrace, verify,
+
     -- * Plan
     Plan, noPlan,
 
@@ -7,8 +10,24 @@ module Development.Build.Plan (
     ) where
 
 import Data.Maybe
+import Data.Map (Map)
 import Development.Build.Store
 import Development.Build.Utilities
+
+import qualified Data.Map as Map
+
+------------------------------- New traces stuff -------------------------------
+
+type VerifyingTrace k v = Map k (Hash, [(k, Hash)])
+
+verify :: (Ord k, Hashable v) => VerifyingTrace k v -> Store i k v -> k -> Bool
+verify trace store key = case Map.lookup key trace of
+    Nothing -> False
+    Just (result, deps) -> getHash store key == result
+                        && and [ getHash store k == h | (k, h) <- deps ]
+
+
+------------------------ Old plan stuff (to be deleted) ------------------------
 
 -- | A /build plan/ is a partial map from a key to the hash of its value, plus a
 -- list of its @(key, hash)@ dependencies. Build plans are typically reused from

### 2018-03-15 paper/3-abstractions.tex

@@
 \begin{minted}[fontsize=\small]{haskell}
--- State monad
+-- Standard State monad (from Control.Monad.State)
 data State s a
@@
 execState :: State s a -> s -> s
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Standard types from Data.Functor.Identity and Data.Functor.Const
+newtype Identity a = Identity { runIdentity :: a }
+\end{minted}
+\begin{minted}[fontsize=\small]{haskell}
+newtype Const ks a = Const { getConst :: ks }
+\end{minted}
+\vspace{0.5mm}
+\begin{minted}[fontsize=\small]{haskell}
+instance Functor (Const ks) where
+    fmap _ (Const ks) = Const ks
+instance Monoid ks => Applicative (Const ks) where
+    pure _              = Const mempty
+    Const x <*> Const y = Const (x <> y)
 \end{minted}
@@
 
-Now we can see why the \hs{Task} abstraction needed to be abstracted over \hs{f}:
-\begin{minted}[xleftmargin=10pt]{haskell}
-type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
-\end{minted}
-The \hs{busy} build system instantiates \hs{c} to \hs{Monad} and \hs{f} to \hs{State (Store i k v)},
-so that the call-back to \hs{fetch :: k -> f v} can side-effect the \hs{Store},
-thereby allowing successive calls to \hs{fetch} to communicate with one another.
-We will see why we want \hs{Task} to be \emph{polymorphic} in \hs{f} in \S\ref{sec-deps}.
-
 Given an acyclic task description, the \hs{busy} build system terminates with a
@@
 
-An alert reader might notice that \hs{busy} accepts \emph{monadic tasks} of
-type \hs{Task}~\hs{Monad}, whereas \hs{sprsh1} is a \hs{Task}~\hs{Applicative}.
-This indicates that \hs{busy} can cope with tasks containing dynamic
-dependencies, which we discuss next (\S\ref{sec-task-monad}).
+\subsection{The need for polymorphism in \hs{Task}}
 
-It is interesting to note that if \hs{busy} is given a \emph{functorial task}
-of type \hs{Task}~\hs{Functor}, which cannot use the application operator
-\hs{<*>}, then \hs{busy} can rejoice in knowing that it will never have to
-rebuild a key twice. Indeed, functorial tasks are very much like tail recursive
-functions -- they cannot branch into multiple subproblems, always leading to a
-straight-line build.
+Now we can see why the \hs{Task} abstraction is polymorphic in \hs{f}:
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+\end{minted}
+The \hs{busy} build system instantiates \hs{f} to \hs{State (Store i k v)},
+so that the call-back to \hs{fetch :: k -> f v} can side-effect the \hs{Store},
+thereby allowing successive calls to \hs{fetch} to communicate with one another.
+
+We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
+Given \emph{one} task description T, we want to explore \emph{many} build systems that
+can build T -- and we will do so in Sections~\ref{sec-build} and \ref{sec-implementations}.
+As we shall see, each build system will use a different \hs{f}, so the task description
+must not fix \hs{f}.
+
+But nor can the task description work for \emph{any} \hs{f}; most
+task descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that
+\hs{f} satisfies certain properties, such as \hs{Monad} or
+\hs{Applicatove}.  That is why \hs{task} has the ``\hs{c f =>}''
+constraint in its type, expressing that \hs{f} can only be instantiated by types
+that satisfy the constraints \hs{c}.
+
+So the type \hs{Task} emerges naturally, almost inevitably.  But now that
+it \emph{has} emgerged, we find the that constraints \hs{c} classify task descriptions
+in a very interesting way:
+\begin{itemize}
+\item \hs{Task Applicative}. In \hs{sprsh1} we needed only \hs{Applicative} operations,
+  expressing the fact that the dependencies between cells can be determined \emph{statically};
+  that is, by looking at the formulae, without ``executing'' them (see \S\ref{sec-deps}).
+\item \hs{Task Monad}.  As we shall see in \S\ref{sec-task-monad}, a monadic task
+  description allows \emph{dynamic} dependencies, in which a formula
+  may depend on the value of cell C, but \emph{which} cell C depends on
+  the value of another cell D.
+\item \hs{Task Functor} is somehat degenerate: the task description cannot even use
+  the application operator \hs{<*>}, which limits dependencies to a single linear chain.
+\end{itemize}
+
+Notice also that, even though \hs{busy} takes a \hs{Task Monad} as its argument, an application of
+\hs{busy} to a \hs{Task Applicative} will typecheck just fine.  It feels a bit
+like sub-typing, but it is actually just ordinary higher-rank polymorphism at work.
 
@@
 our framework, such task descriptions correspond to the type
-\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
+\hs{Task Monad k v}. Consider this spreadsheet
 example:
@@
 \noindent
-The main difference compared to \hs{sprsh1} is that the computation now takes
+The big difference compared to \hs{sprsh1} is that the computation now takes
 place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
@@
 \emph{without attempting to update any dependencies}:
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-newtype Identity a = Identity { runIdentity :: a }
-\end{minted}
-\vspace{0mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-\vspace{1mm}
 \noindent
@@
 
-...
-
-We can statically analyse an applicative task and extract all dependencies of a
-key using the following function:
-
+Earlier we remarked that a \hs{Task Applicative} could only have static dependencies.
+Usually we would extract such static dependencies by (in the case of Excel)
+looking at the syntax tree of the formula.  But a task description has no
+such syntax tree.  Yet, remarkably, we can use the polymorphism of a
+\hs{Task Applicative} to find its dependencies \emph{without dong any of the actual work}.
+Here is the code:
 \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-import Data.Functor.Const
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
 dependencies :: Task Applicative k v -> k -> [k]
-dependencies task = maybe [@@] getConst . task (\k -> Const [k])
+dependencies task k = case task (\k -> Const [k]) k of
+                        Nothing         -> []
+                        Just (Const ks) -> ks
 \end{minted}
 \vspace{1mm}
-
-\noindent
-We can use this function to extract dependencies of \hs{sprsh1}:
-
+Here \hs{Const} is a standard Haskell type defined in Figure~\ref{fig-types}.
+We instantiate \hs{f} to \hs{Const [k]}.  So a value of type \hs{f v},
+or in this case \hs{Const [k] v}, contains no value \hs{v}, but does contain
+a list of keys of type \hs{[k]} which we use to record dependencies.
+The \hs{fetch} callback that we pass to \hs{task}
+records a single dependency; and the standard definitions of \hs{Applicative}
+for \hs{Const} (which we give in Figure~\ref{fig-types}) combines the
+dependencies from different parts of the task.
+As a result, running the task with \hs{f} = \hs{Const [k]} will accumulate a list of
+the task's dependencies -- and that is just what \hs{dependencies} does:
 \vspace{1mm}
@@
 \noindent
-It is important to note that we extract the dependencies without supplying any
-values to the build task. This guarantees that no actual computation is performed
-during the analysis -- we cannot execute the function \hs{(+)} if we have no
-values for it! Static analysis of applicative computations is not new and has
-been used to similar effect before, e.g. see~\cite{free-applicatives}.
+Notice that these calls to \hs{dependencies} do no actual computation
+(in this case, spreadsheet arithmetic). They can't:
+we are not supplying a store or any input numbers.
+So, through the wonders of polymorphism, we are able to extract the
+dependencies of the spreadsheet formula, and to do so efficiently,
+simply by running its code in a different \hs{Applicative}!
+This is not new -- e.g. see~\cite{free-applicatives} -- but it is cool.
 
-To understand how \hs{dependencies} works, let us examine its type more closely,
-expanding the type synonym \hs{Task}:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-(@\std{forall}@ f. Applicative f => (k -> f v) -> k -> Maybe (f v)) -> k -> [k]
-\end{minted}
-\vspace{1mm}
-
-\noindent
-Since the first argument is polymorphic in \hs{f}, we can choose any \hs{f} that
-satisfies the \hs{Applicative} constraint. A good choice in this case is the
-applicative functor \hs{Const}:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-newtype Const m a = Const { getConst :: m }
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-instance Functor (Const m) where
-    fmap _ (Const m) = Const m
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-instance Monoid m => Applicative (Const m) where
-    pure _              = Const mempty
-    Const x <*> Const y = Const (x <> y)
-\end{minted}
-\vspace{1mm}
-
-\noindent
-A value of type \hs{Const}~\hs{[}\hs{k]}~\hs{v} contains no values \hs{v} but
-allows us to record all invocations of the custom \hs{fetch} function that we
-feed to the given compute: \hs{fetch k = Const [k]}. The compute returns a
-value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)} that can be
-unwrapped using \hs{maybe [@@] getConst}, i.e. returning the empty list of
-dependencies in case of an input.
-
-To extract the dependencies of monadic tasks we introduce the function \hs{track}:
+So much for applicatives.  What about monadic tasks with dynamic dependencies?
+We introduce the function \hs{track}:
 a combination of \hs{recompute} and \hs{dependencies} that takes a store as input
@@
 the store.
+\todo{slpj}{This track stuff is not well motivated. When do we want it?  How do we use it?}
 
-...
+\todo{slpj}{I think we should push 'inputs' into the appendix.}
 

### 2018-03-15 paper/5-implementations.tex
Wibbles

@@
     let dirty = any (>= t) dt
-    when dirty $
-        putStore k =<< act
+    when dirty (putStore k =<< act)
 \end{minted}
@@
     dirty <- getChanged k ||^ anyM getChanged dk
-    when dirty $
-        putStore k =<< act
+    when dirty (putStore k =<< act)
 \end{minted}
@@
     dirty <- getChanged k ||^ maybe (return True) (anyM getChanged) dk
-    if not dirty then
-        return Nothing
-    else
-        act >>= \case
+    if not dirty
+       then return Nothing
+       else act >>= \case
               Left e -> return $ Just e
-            Right (_, v) -> do
-                putStore k v
+              Right (_, v) -> do putStore k v
                                  return Nothing

### 2018-03-15 src/Development/Build/Plan.hs
Add constructive traces

@@ -1,6 +1,6 @@
 module Development.Build.Plan (
     -- Traces
-    VerifyingTrace, verify,
+    Result, checkResult, VerifyingTrace, verify, ConstructiveTrace, construct,
 
     -- * Plan
     Plan, noPlan,
@@ -18,14 +18,25 @@ import qualified Data.Map as Map
 
 ------------------------------- New traces stuff -------------------------------
 
-type VerifyingTrace k v = Map k (Hash, [(k, Hash)])
+type Result k = (Hash, [(k, Hash)])
 
-verify :: (Ord k, Hashable v) => VerifyingTrace k v -> Store i k v -> k -> Bool
+checkResult :: Hashable v => Store i k v -> k -> Result k -> Bool
+checkResult store key (keyHash, deps) =
+    and [ getHash store k == h | (k, h) <- (key, keyHash) : deps ]
+
+type VerifyingTrace k = Map k [Result k]
+
+verify :: (Ord k, Hashable v) => VerifyingTrace k -> Store i k v -> k -> Bool
 verify trace store key = case Map.lookup key trace of
-    Nothing -> False
-    Just (result, deps) -> getHash store key == result
-                        && and [ getHash store k == h | (k, h) <- deps ]
+    Nothing      -> False
+    Just results -> any (checkResult store key) results
 
+data ConstructiveTrace k v = ConstructiveTrace (VerifyingTrace k) (Map Hash v)
+
+construct :: (Ord k, Hashable v) => ConstructiveTrace k v -> Store i k v -> k -> Maybe v
+construct (ConstructiveTrace trace cache) store key
+    | verify trace store key == False = Nothing
+    | otherwise                       = Map.lookup (getHash store key) cache
 
 ------------------------ Old plan stuff (to be deleted) ------------------------
 

### 2018-03-15 paper/3-abstractions.tex
Revise as discussed

@@
 getValue   :: Store i k v -> k -> v
-putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
-getHash    :: Hashable v => Store i k v -> k -> Hash
-initialise :: Hashable v => i -> (k -> v) -> Store i k v
+putValue   :: Eq k => Store i k v -> k -> v -> Store i k v
+getHash    :: Hashable v => Store i k v -> k -> Hash v
+initialise :: i -> (k -> v) -> Store i k v
 \end{minted}
@@
 -- Hashing
-hash :: Hashable a => a -> Hash
+hash :: Hashable a => a -> Hash a
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Standard State monad (from Control.Monad.State)
+-- Standard State monad from Control.Monad.State
 data State s a
@@
 newtype Identity a = Identity { runIdentity :: a }
-\end{minted}
-\begin{minted}[fontsize=\small]{haskell}
-newtype Const ks a = Const { getConst :: ks }
+newtype Const m a  = Const    { getConst    :: m }
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
-instance Functor (Const ks) where
-    fmap _ (Const ks) = Const ks
-instance Monoid ks => Applicative (Const ks) where
+instance Functor (Const m) where
+    fmap _ (Const m) = Const m
+\end{minted}
+\vspace{0.5mm}
+\begin{minted}[fontsize=\small]{haskell}
+instance Monoid m => Applicative (Const m) where
     pure _              = Const mempty
@@
 \end{minted}
-\vspace{-2mm}
+\vspace{-3mm}
 \caption{Signatures of main data types and library functions.}\label{fig-types}
@@
 We begin by establishing a common vocabulary for build systems.
-% that allows
-% us to abstract away from specific application domains, such as software
-% development or spreadsheets.
+% that allows us to abstract away from specific application domains, such as
+% software development or spreadsheets.
 Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
@@
 
+\vspace{1mm}
 A \emph{build system} takes a \emph{task description}, a target \emph{key},
 and a \emph{store}, and returns a new store in which the target key is
-up-to-date. The store is a key-value map, together with \emph{persistent build
+up to date. The store is a key/value map, together with \emph{persistent build
 information} that is maintained by the build system itself.
 
+\vspace{-2mm}
 \paragraph{Keys and values.} In software build systems, keys are typically filenames,
@@
 
+\vspace{-2mm}
 \paragraph{Input, output, and intermediate values}
@@
 
-\paragraph{The store.}
+% \vspace{-2mm}
+% \paragraph{The store.}
 A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
@@
 is capable of encoding values corresponding to non-existent keys (missing files
-or empty cells).
+or empty cells). The store also holds a value of type \hs{i},
+representing the persistent build information maintained by the build
+system~--~its ``memory''.
+% Given an \hs{i} and a key/value map \store, we can
+% \hs{initialise} a store.
 
+\vspace{-2mm}
 \paragraph{Hashes.}
 We use a cryptographic \emph{hash function}
-\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash} for
-efficient tracking and sharing of build results, where \hs{Hash} is an abstract
-data type equipped with the equality test.
+\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash}~\hs{v}
+for efficient tracking and sharing of build results, where \hs{Hash}~\hs{v} is
+an abstract data type equipped with~the~equality test.
 In addition to usual \hs{getValue} and \hs{putValue} operations,
@@
 file contents only on demand.}.
-
-\paragraph{Persistent build information.}
-The store also holds a value of type \hs{i},
-representing the persistent build information maintained by the build
-system~--~its ``memory''.
-
-% Given an \hs{i} and a key-value map \store, we can
-% \hs{initialise} a store.
 
@@
 but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
-\hs{fetch} function and a key. It pattern-matches on the key to see if it has a
-task description (in the \Excel case, a formula) for it. If not, it returns
+\emph{callback} \hs{fetch} and a key. It pattern-matches on the key to see if it
+has a task description (in the \Excel case, a formula) for it. If not, it returns
 \hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
@@
 -- the relevant type signatures are given in Fig.~\ref{fig-types}. We will
-explain why in the next subsection (\S\ref{sec-general-build}).
+explain why in subsection~\S\ref{sec-general-build}.
 
@@
 and a store, the build system returns a new store in which the value of the
-target key is up-to-date. What exactly does ``up-to-date'' mean?  We answer
-that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:
+target key is up to date. What exactly does ``up to date'' mean?  We answer
+that precisely in \S\ref{sec-build-correctness}.
 
+\noindent
+Here is a simple build system:
 \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-busy :: (Eq k, Hashable v) => Build Monad () k v -- Drop Hashable? It's confusing here
-busy task key store = execState (compute key) store
+busy :: Eq k => Build Monad () k v
+busy task key store = execState (fetch key) store
   where
-    compute :: k -> State (Store i k v) v
-    compute k = case task compute k of
-        Nothing  -> do { s <- get; return (getValue s k) }
-        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
+    fetch :: k -> State (Store i k v) v
+    fetch k = case task fetch k of
+        Nothing  -> do s <- get; return (getValue s k)
+        Just act -> do v <- act; modify (\s -> putValue s k v); return v
 \end{minted}
@@
 \noindent
-The \hs{busy} build system uses a helper function
-\hs{compute} that, given a key, brings the key up to date in the store, and returns its value.
-The function \hs{compute} runs in the standard Haskell \hs{State} monad -- see
-Figure~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
-To bring a key up to date, \hs{compute} asks the task description \hs{task} how
-to compute \hs{k}.
-If \hs{task} returns \hs{Nothing} the key is an input, so \hs{compute} simply reads
-the result from the store.
-Otherwise \hs{compute} runs the action \hs{act}
-returned by the \hs{task} to produce a resulting value~\hs{v}, records
-the new key/value mapping in the store, and returns \hs{v}.
-Notice that \hs{compute} passes itself to \hs{task} as an argument,
-so that the latter can use \hs{compute} to recursively find the
-values of \hs{k}'s dependencies.
+The \hs{busy} build system defines the callback \hs{fetch} so that, when given a
+key, it brings the key up to date in the store, and returns its value.
+The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
+Fig.~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
+To bring a key up to date, \hs{fetch} asks the task description \hs{task} how
+to compute \hs{k}. If \hs{task} returns \hs{Nothing} the key is an input, so
+\hs{fetch} simply reads the result from the store. Otherwise \hs{fetch} runs
+the action \hs{act} returned by the \hs{task} to produce a resulting
+value~\hs{v}, records the new key/value mapping in the store, and returns \hs{v}.
+Notice that \hs{fetch} passes itself to \hs{task} as an argument, so that the
+latter can use \hs{fetch} to recursively find the values of \hs{k}'s dependencies.
 
@@
 correct result, but it is not a \emph{minimal} build system
-(Definition~\ref{def-minimal}). Since \hs{busy} has no memory (\hs{i=()}), it
-cannot keep track of keys it has already built, and will therefore busily
-recompute the same keys again and again if they have multiple dependents. We
-will develop much more efficient build systems in~\S\ref{sec-implementations}.
+(Definition~\ref{def-minimal}). Since \hs{busy} has no memory
+(\hs{i}~\hs{=}~\hs{()}), it cannot keep track of keys it has already built, and
+will therefore busily recompute the same keys again and again if they have
+multiple dependents. We will develop much more efficient build systems
+in~\S\ref{sec-implementations}.
 
@@
 \cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
-\hs{compute}~\hs{"B1"} twice -- once per each occurrence of \cmd{B1} in the
+\hs{fetch}~\hs{"B1"} twice -- once per each occurrence of \cmd{B1} in the
 formula.
 
-\subsection{The need for polymorphism in \hs{Task}}
+\subsection{The need for polymorphism in \hs{Task}}\label{sec-why-polymorphism}
 
@@
 \end{minted}
-The \hs{busy} build system instantiates \hs{f} to \hs{State (Store i k v)},
-so that the call-back to \hs{fetch :: k -> f v} can side-effect the \hs{Store},
-thereby allowing successive calls to \hs{fetch} to communicate with one another.
+The \hs{busy} build system instantiates \hs{f} to
+\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)},
+so that \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} can side-effect the
+\hs{Store}, thereby allowing successive calls to \hs{fetch} to communicate with
+one another.
 
 We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
-Given \emph{one} task description T, we want to explore \emph{many} build systems that
-can build T -- and we will do so in Sections~\ref{sec-build} and \ref{sec-implementations}.
-As we shall see, each build system will use a different \hs{f}, so the task description
-must not fix \hs{f}.
+Given \emph{one} task description \cmd{T}, we want to explore \emph{many} build
+systems that can build \cmd{T} -- and we will do so in sections~\S\ref{sec-build}
+and~\S\ref{sec-implementations}. As we shall see, each build system will use a
+different \hs{f}, so the task description must not fix \hs{f}.
 
-But nor can the task description work for \emph{any} \hs{f}; most
-task descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that
-\hs{f} satisfies certain properties, such as \hs{Monad} or
-\hs{Applicatove}.  That is why \hs{task} has the ``\hs{c f =>}''
-constraint in its type, expressing that \hs{f} can only be instantiated by types
-that satisfy the constraints \hs{c}.
+But nor can the task description work for \emph{any} \hs{f}; most task
+descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
+satisfies certain properties, such as \hs{Applicatove} or \hs{Monad}. That is
+why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
+expressing that \hs{f} can only be instantiated by types that satisfy the
+constraint \hs{c}.
 
@@
 \begin{itemize}
-\item \hs{Task Applicative}. In \hs{sprsh1} we needed only \hs{Applicative} operations,
-  expressing the fact that the dependencies between cells can be determined \emph{statically};
-  that is, by looking at the formulae, without ``executing'' them (see \S\ref{sec-deps}).
-\item \hs{Task Monad}.  As we shall see in \S\ref{sec-task-monad}, a monadic task
-  description allows \emph{dynamic} dependencies, in which a formula
-  may depend on the value of cell C, but \emph{which} cell C depends on
-  the value of another cell D.
-\item \hs{Task Functor} is somehat degenerate: the task description cannot even use
-  the application operator \hs{<*>}, which limits dependencies to a single linear chain.
+\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
+  operations, expressing the fact that the dependencies between cells can be
+  determined \emph{statically}; that is, by looking at the formulae, without
+  ``computing'' them (see \S\ref{sec-deps}).
+\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
+  description allows \emph{dynamic} dependencies, in which a formula may depend
+  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
+  value of another cell \cmd{D}.
+\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot even
+  use the application operator \hs{<*>}, which limits dependencies to a single
+  linear chain.
+  % It is interesting to note that if \hs{busy} is given a \emph{functorial task}
+  % of type \hs{Task}~\hs{Functor}, which cannot use the application operator
+  % \hs{<*>}, then \hs{busy} can rejoice in knowing that it will never have to
+  % rebuild a key twice. Indeed, functorial tasks are very much like tail recursive
+  % functions, which cannot branch into multiple subproblems.
+\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
+  variants can be used for describing tasks with a certain type of
+  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
 \end{itemize}
 
-Notice also that, even though \hs{busy} takes a \hs{Task Monad} as its argument, an application of
-\hs{busy} to a \hs{Task Applicative} will typecheck just fine.  It feels a bit
-like sub-typing, but it is actually just ordinary higher-rank polymorphism at work.
+Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
+argument, an application of \hs{busy} to a \hs{Task}~\hs{Applicative} will
+typecheck just fine.  It feels a bit like sub-typing, but it is actually just
+ordinary higher-rank polymorphism at work.
 
@@
 our framework, such task descriptions correspond to the type
-\hs{Task Monad k v}. Consider this spreadsheet
-example:
+\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Consider this spreadsheet example:
 
@@
 \noindent
-Here, \cmd{B1} and \cmd{B2} statically depend on each other, but Excel (which
+Here, \cmd{B1} and \cmd{B2} statically depend on each other, but \Excel (which
 uses dynamic dependencies) is perfectly happy. We can express this using
-our task abstraction as follows\footnote{The spreadsheet example that uses
-the \hs{INDIRECT} function can be expressed very similarly: simply replace the
-line containing the \cmd{if} statement with \hs{fetch ("A" ++ show c1)}.}:
+our task abstraction as follows:
+
+% The spreadsheet example that uses
+% the \hs{INDIRECT} function can be expressed very similarly: simply replace the
+% line containing the \cmd{if} statement with \hs{fetch ("A" ++ show c1)}.
 
@@
 place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
-then \hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.
+\hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.
 
@@
 We can now say what it means for a build system to be \emph{correct}, something
-that is seldom stated formally. Our intuition is this: \emph{when the build system
-  returns, the specified key, and all its dependencies, should be up to date}.
-What does ``up to date'' mean?  It means that if we recompute the value of the
-key (using the task description, and the final store), we should get exactly
+that is seldom stated formally. Our intuition is this: \emph{when the build
+system completes, the specified key, and all its dependencies, should be up to
+date}. What does ``up to date'' mean? It means that if we recompute the value of
+the key (using the task description, and the final store), we should get exactly
 the same value as we see in the final store.
 
-To express this formally we need an auxiliary
-function \hs{recompute}, that computes the value of a key in a given store
-\emph{without attempting to update any dependencies}:
+To express this formally we need an auxiliary function \hs{compute}, that
+computes the value of a key in a given store \emph{without attempting to update
+any dependencies}:
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-\noindent
-\todo{SLPJ}{Why monad?}
+
 Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
-we can use the standard Haskell \hs{Identity} monad.
-The use of \hs{Identity} just fixes the `impedance mismatch'
-between the pure function \hs{getValue}~\hs{store}, whose type is \store, and
-the \hs{fetch} argument of the \hs{task}, whose type must be
-\hs{k}~\hs{->}~\hs{f}~\hs{v} for some \hs{f}. To fix the mismatch, we wrap the
-result of the pure function in the \hs{Identity} monad: the function
-\hs{Identity}~\hs{.}~\hs{getValue}~\hs{store} has the type
-\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
-result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now need to get rid
-of the \hs{Identity} wrapper by applying \hs{runIdentity} to the contents of
-\hs{Maybe}.
+we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
+The use of \hs{Identity} just fixes the `impedance mismatch' between the pure
+function \hs{getValue}~\hs{store}, whose type is \store, and the \hs{fetch}
+argument of the \hs{task}, whose type must be \hs{k}~\hs{->}~\hs{f}~\hs{v} for
+some \hs{f}. To fix the mismatch, we wrap the result of the pure function in the
+\hs{Identity} monad: the function \hs{Identity}~\hs{.}~\hs{getValue}~\hs{store}
+has the type \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a
+\hs{task}. The result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now
+need to get rid of the \hs{Identity} wrapper by applying \hs{runIdentity} to the
+contents of \hs{Maybe}.
 
@@
 key           :: k
-store, result :: Store i k v
+store, @@result :: Store i k v
 result = build task key store
@@
     \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
-    for all possible non-input keys \hs{k}, the result of executing the
+    for all possible non-input keys \hs{k}, the result of recomputing the
     \hs{task} matches the value in the \hs{ideal} store:
@@
     \[
-    \hs{recompute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
+    \hs{compute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
     \]
@@
     \]
-    where the implementation of \hs{inputs} will be provided in the next
-    subsection~\S\ref{sec-deps}, which is dedicated to the task dependency
-    analysis. This requirement guarantees that (i) no inputs were corrupted
-    during the build, and (ii)~the \hs{ideal} store has constraints not only
-    on the output, but also on the inputs.
+    where the implementation of \hs{inputs} is provided in the Appendix, which
+    is dedicated to the task dependency analysis. This requirement guarantees
+    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
+    store has constraints not only on the output, but also on the inputs.
 \end{itemize}
@@
 
-\todo{SLPJ}{Please edit as you see fit. We need \hs{dependencies} and
-\hs{track} to be defined for future sections. We also promised to define
-\hs{inputs} in the previous section, but I think it's not too critical and can
-be skipped/moved to Appendix if need be.}
-
-Earlier we remarked that a \hs{Task Applicative} could only have static dependencies.
-Usually we would extract such static dependencies by (in the case of Excel)
-looking at the syntax tree of the formula.  But a task description has no
-such syntax tree.  Yet, remarkably, we can use the polymorphism of a
-\hs{Task Applicative} to find its dependencies \emph{without dong any of the actual work}.
-Here is the code:
+Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
+dependencies. Usually we would extract such static dependencies by (in the case
+of \Excel) looking at the syntax tree of the formula.  But a task description
+has no such syntax tree. Yet, remarkably, we can use the polymorphism of a
+\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
+the actual work}. Here is the code:
 \vspace{1mm}

### 2018-03-15 paper/4-build.tex
Revise as discussed

@@
 system \hs{build} is responsible for ensuring that \hs{key} and its dependencies
-are up-to-date in \hs{result} starting from \hs{store} (\S\ref{sec-build-correctness}). As part of that work,
+are up to date in \hs{result} starting from \hs{store} (\S\ref{sec-build-correctness}). As part of that work,
 \hs{build} \emph{may} store auxiliary information \hs{info} of type \hs{i} in
@@
 underlying \hs{v} rather than storing it directly. Checking a trace requires
-ensuring all the dependencies are up-to-date (using whatever ordering strategy
+ensuring all the dependencies are up to date (using whatever ordering strategy
 as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are

### 2018-03-15 paper/7-related.tex
A few more sections

@@
-\clearpage
 \section{Related work}\label{sec-related}
 
-In section \S\ref{sec-background} we cover many build systems, but there is frustrating little work comparatively analysing build systems. In this section we cover a few build systems that may not
+While there is research on individual build systems, there has been little research to date comparing different build systems.
+In \S\ref{sec-background} we cover several important build systems -- in this section we relate a few additional build systems to our abstractions, and discuss other places where similar abstractions arise.
 
@@
 \item \Buck \cite{buck} and \CloudBuild \cite{cloudbuild} are both fundamentally similar to \Bazel, differing around areas such as non-determinism \S\ref{sec-non-determinism}.
-\item \Nix \cite{nix} has coarse-grained monadic dependencies, with precise hashing of dependencies and downloading of precomputed build products, so may be, at some level of abstraction, the same as the hypothetical Cloud Shake from \S\ref{sec-implementation-cloud-shake}.
+\item \Nix \cite{nix} has coarse-grained dependencies, with precise hashing of dependencies and downloading of precomputed build products.
+When combined with \hs{import-from-derivation} \Nix can also be considered monadic, making it similar to Cloud Shake from \S\ref{sec-implementation-cloud-shake}.
 However, \Nix is not intended as a build system, and the coarse grained nature (packages, not individual files) makes it targeted to a different purpose.
-\todo{NM}{Read comment below when not on an airplane.}
 % John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
 % \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.
-\item \Pluto \cite{pluto} has power to express cyclic dependencies, but that is implemented \todo{NM}{Read the paper so I can say something precise}
+\item \Pluto \cite{pluto} is based on a similar model to \Shake, but additionally allows cyclic build rules combined with a user-specific resolution strategy.
+Often such a strategy can be unfolded into the user rules without loss of precision, but a fully general resolution handler extends the \hs{Task} abstraction with additional features.
 \end{itemize}
@@
 
-\subsection{Memoization}\label{sec-related-memo}
-
-\todo{AM}{Not convined this is really related, but if you think it is, please write something.}
+% \todo{AM}{Not convined this is really related, but if you think it is, please write something.}
+% \subsection{Memoization}\label{sec-related-memo}
 
@@
 While not typically considered build systems, self-adjusting computation is a well studied area, and in particular the contrast between different formulations has been well studied.
-\todo{AM}{Say something.}
+\todo{AM}{Write this section.}
 
@@
 
-The \hs{Task} definition of:
+The definition of \hs{Task} is:
 
@@
 
-Looks tantalisingly close to the profunctor optics definition in \cite{gibbons-profunctor-optics}:
+Which looks tantalisingly close to the profunctor optics definition by \citet{gibbons-profunctor-optics}:
 
@@
 
-Assuming we instantiate \hs{p} to something that provides \hs{k -> f v} -- which many of the actual instances in that paper do.
+Provided we instantiate \hs{p} to something like \hs{k -> f v} -- which many of the actual instances in that paper do.
 The properties of such optics are well studied, and the functions like \hs{dependencies} are very much based on observations from that field of work.
@@
 
-\subsection{Free applicative functors and monads}\label{sec-related-free}
-
-\todo{AM}{Mention that one could have chosen a free construction to create
-a small DSL for build tasks. Although this simplifies some of the presented
-combinators, in general this seems unnecessary.}
+% \subsection{Free applicative functors and monads}\label{sec-related-free}
+%
+% NM: Unconvinced it is worth it.
+% \todo{AM}{Mention that one could have chosen a free construction to create
+% a small DSL for build tasks. Although this simplifies some of the presented
+% combinators, in general this seems unnecessary.}

### 2018-03-15 paper/2-background.tex

@@
 "release.tar" %> \_ -> do
-    { need ["release.txt"]
-    ; files <- lines <$> readFile "release.txt"
-    ; need files
-    ; system "tar" $ ["-cf", "result.tar"] ++ files }
+    need ["release.txt"]
+    files <- lines <$> readFile "release.txt"
+    need files
+    system "tar" $ ["-cf", "result.tar"] ++ files
 \end{minted}
@@
 not yet been built (as \Excel does), \Shake \emph{pauses} their execution until
-the dependencies are brought up-to-date. We refer to this build algorithm as
+the dependencies are brought up to date. We refer to this build algorithm as
 \emph{recursive}.

### 2018-03-15 src/Development/Build.hs
Translate changes from the paper

@@ -38,8 +38,8 @@ busy task key store = execState (fetch key) store
   where
     fetch :: k -> State (Store () k v) v
     fetch k = case task fetch k of
-        Nothing  -> do { s <- get; return (getValue s k) }
-        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
+        Nothing  -> do s <- get; return (getValue s k)
+        Just act -> do v <- act; modify (\s -> putValue s k v); return v
 
 memo :: forall k v. Eq k => Build Monad () k v
 memo task key store = fst $ execState (fetch key) (store, [])

### 2018-03-15 src/Development/Build/Task.hs

@@ -1,6 +1,6 @@
 {-# LANGUAGE ConstraintKinds, RankNTypes #-}
 {-# OPTIONS_GHC -Wno-unused-top-binds #-}
-module Development.Build.Task (Task, inputTask, isInput, sprsh1) where
+module Development.Build.Task (Task, inputTask, isInput, sprsh1, sprsh2) where
 
 import Control.Applicative
 import Control.Monad
@@ -103,7 +103,9 @@ sprsh1 _     _    = Nothing
 
 sprsh2 :: Task Monad String Integer
 sprsh2 fetch "B1" = Just $ do c1 <- fetch "C1"
-                              if c1 == 1 then fetch "A1" else fetch "A2"
+                              if c1 == 1 then fetch "B2" else fetch "A2"
+sprsh2 fetch "B2" = Just $ do c1 <- fetch "C1"
+                              if c1 == 1 then fetch "A1" else fetch "B1"
 sprsh2 _     _    = Nothing
 
 -- These type synonyms are not very useful, but enumerate all interesting cases.

### 2018-03-15 paper/3-abstractions.tex
Finalise S3

@@
   value of another cell \cmd{D}.
-\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot even
-  use the application operator \hs{<*>}, which limits dependencies to a single
-  linear chain.
-  % It is interesting to note that if \hs{busy} is given a \emph{functorial task}
-  % of type \hs{Task}~\hs{Functor}, which cannot use the application operator
-  % \hs{<*>}, then \hs{busy} can rejoice in knowing that it will never have to
-  % rebuild a key twice. Indeed, functorial tasks are very much like tail recursive
-  % functions, which cannot branch into multiple subproblems.
+\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
+  even use the application operator \hs{<*>}, which limits dependencies to a
+  single linear chain. It is interesting to note that, when run on a
+  \hs{Task}~\hs{Functor}, the \hs{busy} build system will build each key at most
+  once, thus partially fulfilling the minimality requirement~\ref{def-minimal}.
+  Alas, it still has no mechanism to decide which input keys changed since the
+  previous build.
 \item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
@@
 Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
-argument, an application of \hs{busy} to a \hs{Task}~\hs{Applicative} will
-typecheck just fine.  It feels a bit like sub-typing, but it is actually just
-ordinary higher-rank polymorphism at work.
+argument, an application of \hs{busy} to a \hs{Task}~\hs{Functor} or
+a \hs{Task}~\hs{Applicative} will typecheck just fine. It feels a bit like
+sub-typing, but it is actually just ordinary higher-rank polymorphism at
+work~\cite{jones2007practical}.
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-recompute :: Task Monad k v -> Store i k v -> k -> Maybe v
-recompute task store key = runIdentity <$> task (Identity . getValue store) key
+compute :: Task Monad k v -> Store i k v -> k -> Maybe v
+compute task store key = runIdentity <$> task (Identity . getValue store) key
 \end{minted}
 
+\noindent
 Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
@@
 store, @@result :: Store i k v
-result = build task key store
+result = @@build @@task @@key @@store
 \end{minted}
@@
     permitting shallow cloud builds.
-    \item The \hs{result}, \hs{store} and \hs{ideal} all agree on the inputs
-    that belong to the transitive closure of \hs{key}'s dependencies with
-    respect to the \hs{result}:
+    \item The \hs{store}, \hs{result} and \hs{ideal} agree on all input keys
+    \hs{k} that belong to the transitive closure of \hs{key}'s dependencies:
     \[
-        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{store}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
+    \,\,\,\,\,\,\,\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{&&}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}
     \]
-    \vspace{-6mm}
-    \[
-        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
-    \]
-    where the implementation of \hs{inputs} is provided in the Appendix, which
-    is dedicated to the task dependency analysis. This requirement guarantees
-    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
-    store has constraints not only on the output, but also on the inputs.
+    % where the implementation of \hs{inputs} is provided in the Appendix, which
+    % is dedicated to the task dependency analysis.
+    This requirement asserts that (i) no inputs were corrupted during the
+    build, and (ii)~the \hs{ideal} store has constraints not only on the output,
+    but also on the inputs.
 \end{itemize}
@@
 dependencies :: Task Applicative k v -> k -> [k]
-dependencies task k = case task (\k -> Const [k]) k of
+dependencies task key = case task (\k -> Const [k]) key of
                             Nothing         -> []
@@
 \vspace{1mm}
-Here \hs{Const} is a standard Haskell type defined in Figure~\ref{fig-types}.
-We instantiate \hs{f} to \hs{Const [k]}.  So a value of type \hs{f v},
-or in this case \hs{Const [k] v}, contains no value \hs{v}, but does contain
-a list of keys of type \hs{[k]} which we use to record dependencies.
-The \hs{fetch} callback that we pass to \hs{task}
-records a single dependency; and the standard definitions of \hs{Applicative}
-for \hs{Const} (which we give in Figure~\ref{fig-types}) combines the
-dependencies from different parts of the task.
-As a result, running the task with \hs{f} = \hs{Const [k]} will accumulate a list of
-the task's dependencies -- and that is just what \hs{dependencies} does:
+Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
+instantiate \hs{f} to \hs{Const}~\hs{[@@k]}.  So a value of type \hs{f}~\hs{v},
+or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
+contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
+The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
+and the standard definition of \hs{Applicative} for \hs{Const} (which we give
+in Fig.~\ref{fig-types}) combines the dependencies from different parts of the
+task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
+accumulate a list of the task's dependencies -- and that is just what
+\hs{dependencies} does:
 \vspace{1mm}
@@
 \noindent
-Notice that these calls to \hs{dependencies} do no actual computation
-(in this case, spreadsheet arithmetic). They can't:
-we are not supplying a store or any input numbers.
-So, through the wonders of polymorphism, we are able to extract the
-dependencies of the spreadsheet formula, and to do so efficiently,
-simply by running its code in a different \hs{Applicative}!
-This is not new -- e.g. see~\cite{free-applicatives} -- but it is cool.
+Notice that these calls to \hs{dependencies} do no actual computation (in this
+case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
+input numbers. So, through the wonders of polymorphism, we are able to extract
+the dependencies of the spreadsheet formula, and to do so efficiently, simply by
+running its code in a different \hs{Applicative}! This is not new, for example
+see~\cite{free-applicatives}, but it is cool.
 
-So much for applicatives.  What about monadic tasks with dynamic dependencies?
-We introduce the function \hs{track}:
-a combination of \hs{recompute} and \hs{dependencies} that takes a store as input
-and computes both the resulting value and the list of its dependencies:
+So much for applicative tasks. What about monadic tasks with dynamic
+dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
+dependencies need to be tracked too. This cannot be done statically; note that
+the application of the function \hs{dependencies} to a \hs{Task}~\hs{Monad} will
+not typecheck. We need to run a monadic task on a store with concrete values,
+which will determine the discovered dependencies. Accordingly, we introduce
+the function \hs{track}: a combination of \hs{compute} and \hs{dependencies}
+that computes both the resulting value and the list of its dependencies:
 
@@
 track :: Task Monad k v -> Store i k v -> k -> Maybe (v, [k])
-track task store = fmap runWriter . task (\k -> writer (getValue store k, [k]))
-\end{minted}
-\vspace{1mm}
-
-The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
-we feed \hs{fetch k = writer (getValue store k, [k])} to the task, which in
-addition to fetching a value from the store, tracks the corresponding key. We
-can \hs{track} both applicative and monadic tasks:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store = initialise () (\key -> if key == "A1" then 10 else 20)
-@\ghci@ track sprsh1 store "B1"
-Just (30,["A1","A2"])
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ track sprsh2 store "B1"
-Just (20,["C1","A2"])
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store' = initialise () (\key -> if key == "A1" then 10 else 1)
-@\ghci@ track sprsh2 store' "B1"
-Just (10,["C1","A1"])
+track task store = runWriter . task (\k -> writer (getValue store k, [k]))
 \end{minted}
@@
 \noindent
-As expected, the dependencies of \hs{sprsh2} are determined by the contents of
-the store.
-\todo{slpj}{This track stuff is not well motivated. When do we want it?  How do we use it?}
+The implementation uses the standard Haskell \hs{Writer} monad: we instantiate
+\hs{f} to \hs{Writer}~\hs{[@@k]}~\hs{v}, and feed
+\hs{fetch}~\hs{k}~\hs{=}~\hs{writer}~\hs{(@@getValue}~\hs{store}~\hs{k,}~\hs{[@@k])}
+to the task, which in addition to fetching a value from the store, tracks the
+corresponding key. Here is an example of \hs{track}ing monadic tasks:
 
-\todo{slpj}{I think we should push 'inputs' into the appendix.}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store1 = initialise () (\key -> if key == "A1" then 10 else 1)
+@\ghci@ track sprsh2 store1 "B1"
+Just (1,["C1","B2"])
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ store2 = initialise () (\key -> if key == "A1" then 10 else 20)
+@\ghci@ track sprsh2 store2 "B2"
+Just (20,["C1","B1"])
+\end{minted}
+\vspace{1mm}
 
-Here is a draft implementation of \hs{inputs} used in the definition of
-build system correctness in \S\ref{sec-build-correctness}:
+\noindent
+As expected, the dependencies of \hs{sprsh2} (see the spreadsheet
+in~\S\ref{sec-task-monad}) are determined by the contents of the store.
+Furthermore, the above example highlights the static cycle between cells
+\cmd{B1} and \cmd{B2}: in \hs{store1}, we have \cmd{B1 = B2 = A1 = 10},
+whereas in \hs{store2}, we have \cmd{B2 = B1 = A2 = 20}.
 
-\begin{minted}[xleftmargin=10pt]{haskell}
-inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
-inputs task store key = filter (isInput task) (closure deps key)
-  where
-    deps k = maybe [] snd (track task store k)
+% \subsection{Free tasks}
+% \todo{AM}{Introduce free constructions \hs{TaskA} and \hs{TaskM} giving
+% intuitions on what tasks really are. This will take at least one page.}
+%
+% data TaskA k v a = InputA | TaskA [k] ([v] -> a)
+%
+% data TaskM k v a = InputM | DoneM a | TaskM [k] ([v] -> Task k v a)
 
-closure :: Eq a => (a -> [a]) -> a -> [a] -- Standard graph transitive closure
-
-data Proxy a = Proxy
-
-isInput :: Task Monad k v -> k -> Bool
-isInput task = isNothing . task (const Proxy)
-\end{minted}
-
-\subsection{Free tasks (?)}
-
-\todo{AM}{Introduce free constructions \hs{TaskA} and \hs{TaskM} giving
-intuitions on what tasks really are. Requires some further discussion with Neil.
-This will take at least one page.}
-
-% \subsection{Recognising an applicative compute wearing a monadic hat}
-
+% \subsection{Recognising an applicative task wearing a monadic hat}
 % Although dynamic dependencies are very useful, they typically represent only a
 % small fraction of all dependencies. One can exploit this in a build system by
-% recognising applicative computations even though they have the type
-% \hs{Compute}~\hs{Monad} and treating them specially.
-
-% Consider the following data type.
-
-% \subsection{Free build structures}\label{sec-free-build}
-
-% \begin{minted}{haskell}
-% data Depend k v r = Depend [k] ([v] -> r)
-
-% data Depends k v r = Depends [k] ([v] -> Depends k v r)
-%                    | Done r
-% \end{minted}
+% recognising applicative tasks even though they have the type
+% \hs{Task}~\hs{Monad} and treating them specially.

### 2018-03-15 paper/9-appendix.tex
Add description of track
Finalise S3

@@
 
-\todo{AM}{Explain \hs{trackDependencies}.}
+\todo{AM}{Explain \hs{track}.}
+
+Here is a draft implementation of \hs{inputs} used in the definition of
+build system correctness in \S\ref{sec-build-correctness}:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
+inputs task store key = filter (isInput task) (closure deps key)
+  where
+    deps k = maybe [] snd (track task store k)
+
+closure :: Eq a => (a -> [a]) -> a -> [a] -- Standard graph transitive closure
+
+data Proxy a = Proxy
+
+isInput :: Task Monad k v -> k -> Bool
+isInput task = isNothing . task (const Proxy)
+\end{minted}
 

### 2018-03-15 paper/6-engineering.tex

@@
-\clearpage
 \section{Engineering aspects}\label{sec-engineering}
@@
 We can model such failure conditions by instantiating \hs{v} to either \hs{Maybe v} (for missing values) or \hs{Either e v} (for exceptions of type \hs{e}).
-That can model the the values inside the store and the compute function, but because \hs{v} is polymorphic for builds, it does not let the build system apply special behaviour for errors - e.g.
+That can model the the values inside the store and the \hs{compute} function, but because \hs{v} is polymorphic for builds, it does not let the build system apply special behaviour for errors - e.g.
 early aborting.
-We consider such details an engineering concern, which has no real impact on the power or high-level implementation of the build system.
 
@@
 
-\todo{SLPJ}{Can we replace ``node'' with ``task''? Everywhere.}
 \begin{enumerate}
-\item The \hs{topological} function can instead retain the dependency graph, and whenever all the dependencies of a node are complete, the node itself can be started.
+\item The \hs{topological} function can build the full dependency graph, and whenever all dependencies of a task are complete, the task itself can be started.
 \item The \hs{reordering} function can be made parallel in a few ways, but the most direct is to have $n$ threads reading entries from the list of keys.
-As before, if a key requires something not yet built, it is added to the end -- the difference is that sometimes things will be moved back not because they are out of order but because of races with earlier nodes that had not yet finished.
+As before, if a key requires something not yet built, it is added to the end -- the difference is that sometimes things will be moved to the back of the queue not because they are out of order but because of races with earlier nodes that had not yet finished.
 As a consequence, over successive runs, potentially racey dependencies will be separated, giving better parallelism over time.
-\item The \hs{recursive} function can be made parallel by working more directly on the \hs{Depeneds} structure, which gives multiple dependency keys at a time (all those arising from \hs{Applicative} dependencies).
-All such dependencies can be executed simultaneously, with a slightly smarter check for whether something is already executing that also blocks until an existing execution has finished.
+\item The \hs{recursive} function can be made parallel by starting \hs{Applicative} dependencies of a \hs{Task} in parallel, using the strategy described by \citet{haxl}.
 \end{enumerate}
@@
 In our model we define \hs{Task} as a function - when give the same inputs it will always produce the same outputs.
-Alas, the real-world is not as obliging.
+Alas, the real-world is not so obliging.
 Some examples of impure \hs{Task} functions include:
@@
 \item[Non-determinism] Some actions are non-determinstic, producing one result from a possble set.
-As an example, GHC when compiled using parallelism can change the order in which unique variables are obtained from the supply, producing semantically identical results but with different hidden variables.
+As an example, GHC when compiled using parallelism can change the order in which unique variables are obtained from the supply, producing different but semantically identical results.
 \end{description}
@@
 
-\subsection{Cloud aspects}\label{sec-cloud-aspects}
+\subsection{Cloud implementations}\label{sec-cloud-aspects}
 
@@
 \begin{description}
-\item[Materialisation] Sometimes input files will go through many transformations before producing the end result, e.g.
+\item[Materialisation] Sometimes input files will involve many intermediate tasks before producing the end result, e.g.
 an installer package.
-These intermediate steps may be large, so some cloud builds can execute the graph \textit{without} downloading or \textit{materialising} the intermediate nodes - only the final one.
+These intermediate steps may be large, so some cloud builds can execute the graph \textit{without} downloading or \textit{materialising} the results of intermediate tasks - only the final result.
 Some build systems can go even further, integrating with the file system to only materialise the file when the user accesses it \cite{google_internal}.
-\item[Eviction] The store of traces as shown grows indefinitely.
-In many cases resource constraints mean that eventually items must be evicted from the store that are predicted not to be used again.
+\item[Eviction] The store of traces as shown grows indefinitely, but often resource constraints require evicting old items from the store.
 One option is to evict the contents and any trace that mentions the now-defunct \hs{Hash}.
@@
 \item[Frankenbuilds] A build is considered a franken-build \cite{cloudbuild} if a value is calculated locally, but something that depends on that key is pulled from the cache, and the value calculated locally does not match what was previously calculated by the cloud.
-Our implementations avoid this issue by storing complete traces, but if a cloud build system was to only reference input nodes it could be a problem.
-\item[Communication] With a cloud build communication with a central server can be a bottleneck, so it is important to send only the minimum amount of information, optimising with respect to build-system specific invariants - see \S\ref{sec-smart-traces} for some possible optimisations.
-\item[Offloading] Once the cloud is storing build products and traces, it is not infeasible for the cloud to also contain dedicated workers that can help with jobs run locally - offloading some of the computation and potentially running vastly more commands in parallel.
+Our implementations avoid this issue by storing complete traces, but if a cloud build system was to only reference input nodes this situation can arise.
+\item[Communication] When traces or contents are stored on a central server communication can become a bottleneck, so it is important to send only the minimum amount of information, optimising with respect to build-system specific invariants - see \S\ref{sec-smart-traces} for some possible optimisations.
+\item[Offloading] Once the cloud is storing build products and traces, it is possible for the cloud to also contain dedicated workers that can execute tasks remotely - offloading some of the computation and potentially running vastly more commands in parallel.
 \end{description}
@@
 
-Some build systems, for example \Excel and \Ninja are capable of recomputing a node if either its dependencies change, \textit{or} the rule itself changes. For example:
+Some build systems, for example \Excel and \Ninja, are capable of recomputing a task if either its dependencies change, \textit{or} the rule itself changes. For example:
 
-\begin{minted}[xleftmargin=10pt]{bash}
-A1 = 20
+\begin{minted}[xleftmargin=10pt]{text}
+A1 = 20            B1 = A1+A2
 A2 = 10
-B1 = A1+A2
 \end{minted}
@@
 to \textsf{A1-A2}.
-This pattern can be captured by describing the rule producing \textsf{B1} as also depending on the value \textsf{B1.formula}.
+This pattern can be captured by describing the rule producing \textsf{B1} as also depending on the value \textsf{B1-formula}.
 In a \hs{Monad} build system the implementation can be given very directly as:
@@
 \begin{minted}{haskell}
-compute ask "B1" = do
-    formula <- ask "B1-formula"
-    evalFormula ask formula
+task fetch "B1" = do
+    formula <- fetch "B1-formula"
+    evalFormula fetch formula
 \end{minted}
@@
 Namely, first look up the formula, then interpret it.
-In an \hs{Applicative} system it is not possible to change dependencies based on the formula, as would be required, so instead the formula can be captured as a dependency (but its value not used) and \textit{also} baked directly into the \hs{compute} function.
+In an \hs{Applicative} system it is not possible to change dependencies based on the formula, as would be required, so instead the formula can be captured as a dependency (but its value not used) and \textit{also} baked directly into the \hs{task} function.
 
@@
 
-Our build system abstraction assumes a \hs{k -> v} store, along with a build system that works directly on \hs{k} and \hs{v} values.
+Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build system that works directly on \hs{k} and \hs{v} values.
 However, certain build systems provide greater flexibility, e.g.
@@
 
-Another more compelling example is build rules that produce multiple keys -- for example \textsf{ghc Foo.hs} produces both \textsf{Foo.hi} and \textsf{Foo.o}.
+A more compelling example is build rules that produce multiple keys -- for example \textsf{ghc Foo.hs} produces both \textsf{Foo.hi} and \textsf{Foo.o}.
 That can be represented by having a key whose value is a pair of file names, and whose result is a pair of file contents.

### 2018-03-15 paper/8-conclusions.tex

@@
-\clearpage
 \section{Conclusions}\label{sec-conclusions}
 
-...
+We have investigated multiple build systems, showing how their properties are consequences of two implementation choices: what order you build in and whether you decide to rebuild.
+By first decomposing the pieces, we show how to recompose the pieces to find new points in the design space.
+In particular, a simple recombination leads to a design for a monadic cloud build system.
+Armed with that blueprint we hope to actually implement such a system as future work.

### 2018-03-15 paper/3-abstractions.tex
A few more tweaks

@@
 system~--~its ``memory''.
-% Given an \hs{i} and a key/value map \store, we can
-% \hs{initialise} a store.
+% Given an \hs{i} and a key/value map \store, we can \hs{initialise} a store.
 
@@
     \]
-    % where the implementation of \hs{inputs} is provided in the Appendix, which
-    % is dedicated to the task dependency analysis.
     This requirement asserts that (i) no inputs were corrupted during the
@@
 dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
-dependencies need to be tracked too. This cannot be done statically; note that
+dependencies need to be tracked too. This cannot be done statically; notice that
 the application of the function \hs{dependencies} to a \hs{Task}~\hs{Monad} will
@@
 \noindent
-The implementation uses the standard Haskell \hs{Writer} monad: we instantiate
-\hs{f} to \hs{Writer}~\hs{[@@k]}~\hs{v}, and feed
+The implementation uses the standard Haskell \hs{Writer} monad, which allows us
+to record addition information -- a list of keys of type \hs{[@@k]} -- when
+computing a task. We instantiate \hs{f} to \hs{Writer}~\hs{[@@k]}~\hs{v}, and feed
 \hs{fetch}~\hs{k}~\hs{=}~\hs{writer}~\hs{(@@getValue}~\hs{store}~\hs{k,}~\hs{[@@k])}

### 2018-03-16 paper/6-engineering.tex
Finalise S6

@@
 
-In the previous sections we have modelled the most critical subset of various build systems.
-However, like all real-world systems, there are many corners that obscure the essence.
-In this section we discuss some of those details, what would need to be done to capture them in our model, and what the impact would be.
+In the previous sections we have modelled the most critical subset of various
+build systems. However, like all real-world systems, there are many corners that
+obscure the essence. In this section we discuss some of those details, what
+would need to be done to capture them in our model, and what the impact would be.
 
-\subsection{Partial stores, exceptions}
+\subsection{Partial stores and exceptions}
 
-Our model assumes a world where the store is fully-defined, every \hs{k} is associated with a \hs{v}, and every compute successfully completes returning a valid value.
-In the real world, build systems frequently deal with errors - e.g.
-``file not found'', or ``compilation failed''.
-We can model such failure conditions by instantiating \hs{v} to either \hs{Maybe v} (for missing values) or \hs{Either e v} (for exceptions of type \hs{e}).
-That can model the the values inside the store and the \hs{compute} function, but because \hs{v} is polymorphic for builds, it does not let the build system apply special behaviour for errors - e.g.
-early aborting.
+Our model assumes a world where the store is fully-defined, every \hs{k} is
+associated with a \hs{v}, and every compute successfully completes returning a
+valid value. In the real world, build systems frequently deal with errors, e.g.
+``file not found'', or ``compilation failed''. We can model such failure
+conditions by instantiating \hs{v} to either \hs{Maybe}~\hs{v} (for missing
+values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
+model the values inside the store and the \hs{Task}, but because \hs{v} is
+polymorphic for builds, it does not let the build system apply special behaviour
+for errors, e.g. early aborting.
+
+\todo{NM}{Isn't ``early aborting'' automatically handled by short-circuiting
+withing monads, i.e. \hs{Nothing} or \hs{Left e} cancel out all subsequent
+computation?}
 
@@
 
-While we have given simple implementations assuming a single thread of execution, all the build systems we address are actually concurrent.
-While it complicates the model to add concurrency, the complications can be restricted exclusively to the dependency strategy:
+While we have given simple implementations assuming a single thread of execution,
+all the build systems we address are actually concurrent. While it complicates
+the model to add concurrency, the complications can be restricted exclusively
+to the dependency strategy:
 
 \begin{enumerate}
-\item The \hs{topological} function can build the full dependency graph, and whenever all dependencies of a task are complete, the task itself can be started.
-\item The \hs{reordering} function can be made parallel in a few ways, but the most direct is to have $n$ threads reading entries from the list of keys.
-As before, if a key requires something not yet built, it is added to the end -- the difference is that sometimes things will be moved to the back of the queue not because they are out of order but because of races with earlier nodes that had not yet finished.
-As a consequence, over successive runs, potentially racey dependencies will be separated, giving better parallelism over time.
-\item The \hs{recursive} function can be made parallel by starting \hs{Applicative} dependencies of a \hs{Task} in parallel, using the strategy described by \citet{haxl}.
+\item The \hs{topological} function can build the full dependency graph, and
+whenever all dependencies of a task are complete, the task itself can be started.
+
+\item The \hs{reordering} function can be made parallel in a few ways, but the
+most direct is to have $n$ threads reading entries from the list of keys. As
+before, if a key requires something not yet built, it is added to the end -- the
+difference is that sometimes things will be moved to the back of the queue not
+because they are out of order but because of races with earlier nodes that had
+not yet finished. As a consequence, over successive runs, potentially racey
+dependencies will be separated, giving better parallelism over time.
+
+\item The \hs{recursive} function can be made parallel by starting static
+dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
+resolved, using the strategy described by~\citet{marlow2014haxl}.
 \end{enumerate}
 
-The actual implementation of the parallel strategies is not overly onerous, but neither is it beautiful or informative.
+The actual implementation of the parallel strategies is not overly onerous,
+but neither is it beautiful or informative.
 
@@
 
-In our model we define \hs{Task} as a function - when give the same inputs it will always produce the same outputs.
-Alas, the real-world is not so obliging.
-Some examples of impure \hs{Task} functions include:
+In our model we define \hs{Task} as a function -- when given the same inputs it
+will always produce the same outputs. Alas, the real-world is not so obliging.
+Some examples of impure tasks include:
 
 \begin{description}
-\item[Untracked state] Some actions depend on untracked state - for example your C compilation may explicitly list the \texttt{source.c} file as a dependency, but it may not record that the version of \texttt{gcc} is also a dependency.
-\item[Volatility] Some actions are defined to change in every execution, for example \Excel has a function \textsf{RANDBETWEEN} which produces a random number - on each recalculation this function is defined to change.
-Similarly, build systems like \Make and \Shake provide \textsf{phony} rules which are also volatile.
-\item[Non-determinism] Some actions are non-determinstic, producing one result from a possble set.
-As an example, GHC when compiled using parallelism can change the order in which unique variables are obtained from the supply, producing different but semantically identical results.
+\item[Untracked state] Some tasks depend on untracked state -- for example
+your C compilation may explicitly list the \cmd{source.c} file as a dependency,
+but it may not record that the version of \cmd{gcc} is also a dependency.
+
+\item[Non-determinism] Some tasks are \emph{non-determinstic}, producing a
+result from a possible set. As an example, GHC when compiled using parallelism
+can change the order in which unique variables are obtained from the supply,
+producing different but semantically identical results.
+
+\item[Volatility] Some tasks are defined to change in every execution, for
+example \Excel provides a~``function'' \cmd{RANDBETWEEN} which produces a random
+number in a specified range~--~on each recalculation it is defined to change.
+Similarly, build systems like \Make and \Shake provide \emph{phony rules} which
+are also volatile.
 \end{description}
@@
 Interestingly, there is significant interplay between all three sources of impurity.
-Systems like \Bazel use various sandboxing techniques to guard against missing dependencies, but none are likely to capture all dependencies right down to CPU model and microcode version.
-Rules that do have untracked state can be marked as volatile, a technique \Excel takes with the \textsf{INDIRECT} function, removing the untracked-state at the cost of more recomputation.
+Systems like \Bazel use various sandboxing techniques to guard against missing
+dependencies, but none are likely to capture all dependencies right down to CPU
+model and microcode version. Rules that do have untracked state can be marked as
+volatile, a technique \Excel takes with the \cmd{INDIRECT} function, removing the
+untracked state at the cost of minimality.
 
-We note that while all the simple implementations presented in \S\ref{sec-implementations} can deal with non-determinism, some real build systems cannot, e.g. \Buck.
+We note that while all the simple implementations presented in
+\S\ref{sec-implementations} can deal with non-determinism, some real build
+systems, e.g. \Buck and \Bazel, rely on task determinism.
 
-One tempting way of modelling non-determinism is to enrich \hs{Task} from \hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus} respectively - both provide the underlying class with zero and choice.
-However, that seems to encode a different type of non-determinism - a kind of dependency-level non-determinism when most rules really experience a value-level non-determinism.
+One tempting way of modelling non-determinism is to enrich \hs{Task} from
+\hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus},
+respectively. More concretely, the following task description corresponds to
+a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}:
 
-Volatility can be modelled by introducing a special \hs{k} whose value is changed between every run.
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+sprsh3 :: Task MonadPlus String Integer
+sprsh3 fetch "B1" = Just $ (+) <$> fetch "A1" <*> pure 1 `mplus` pure 2
+sprsh3 _     _    = Nothing
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Handling such tasks is possible in our framework, but requires an adjustment of
+the correctness definition (\S\ref{sec-build-correctness}): instead of requiring
+that the result of recomputing the \hs{task} matches the value in the \hs{ideal}
+store:
+\[
+\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}\]
+\noindent
+we now require that \hs{ideal} contains \emph{one possible result of recomputing}
+the \hs{task}:
+\[
+\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{`elem`}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}
+\]
+
+Note that \hs{Task}~\hs{MonadPlus} is powerful enough to model dependency-level
+non-determinism, for example, \cmd{INDIRECT("A" \& RANDBETWEEN(1,2))}, whereas
+most build tasks in real-life build systems only experience a value-level
+non-determinism. \Excel handles this example simply by marking the cell
+volatile~--~an approach that can be readily adopted by any of our
+implementations by introducing a special key \cmd{RealWorld} whose value is
+changed between every run.
 
@@
 \begin{description}
-\item[Materialisation] Sometimes input files will involve many intermediate tasks before producing the end result, e.g.
-an installer package.
-These intermediate steps may be large, so some cloud builds can execute the graph \textit{without} downloading or \textit{materialising} the results of intermediate tasks - only the final result.
-Some build systems can go even further, integrating with the file system to only materialise the file when the user accesses it \cite{google_internal}.
-\item[Eviction] The store of traces as shown grows indefinitely, but often resource constraints require evicting old items from the store.
-One option is to evict the contents and any trace that mentions the now-defunct \hs{Hash}.
-However, if the build system can defer materialisation, it may be possible to only evict the contents, allowing builds to pass-through hashes of values where the underlying value is not known.
-If so, the build must be able to recreate the value if required, potentially dealing with a different result in a future run.
-\item[Frankenbuilds] A build is considered a franken-build \cite{cloudbuild} if a value is calculated locally, but something that depends on that key is pulled from the cache, and the value calculated locally does not match what was previously calculated by the cloud.
-Our implementations avoid this issue by storing complete traces, but if a cloud build system was to only reference input nodes this situation can arise.
-\item[Communication] When traces or contents are stored on a central server communication can become a bottleneck, so it is important to send only the minimum amount of information, optimising with respect to build-system specific invariants - see \S\ref{sec-smart-traces} for some possible optimisations.
-\item[Offloading] Once the cloud is storing build products and traces, it is possible for the cloud to also contain dedicated workers that can execute tasks remotely - offloading some of the computation and potentially running vastly more commands in parallel.
+\item[Materialisation] Sometimes input files will involve many intermediate tasks
+before producing the end result, e.g. an installer package. These intermediate
+steps may be large, so some cloud builds can execute the graph \emph{without}
+downloading or \emph{materialising} the results of intermediate tasks -- only
+the final result. Some build systems can go even further, integrating with the
+file system to only materialise the file when the user accesses it \cite{gvfs}.
+
+\todo{NM}{I cited GVFS above, and could not find a proper reference to Google's
+internal equivalent. There are some mentions online of Piper, but not citable.}
+
+\item[Eviction] The store of traces as shown grows indefinitely, but often
+resource constraints require evicting old items from the store. One option is
+to evict the contents and any trace that mentions the now-defunct
+\hs{Hash}~\hs{v}. However, if the build system can defer materialisation, it
+may be possible to only evict the contents, allowing builds to pass-through
+hashes of values where the underlying value is not known. If so, the build must
+be able to recreate the value if required, potentially dealing with a different
+result in a future run.
+
+\item[Frankenbuilds] A build is considered a
+\emph{frankenbuild}~\cite{esfahani2016cloudbuild} if a value is calculated
+locally, but something that depends on that key is pulled from the cache, and
+the value calculated locally does not match what was previously calculated and
+stored in the cloud. Our implementations avoid this issue by storing complete
+traces, but if a cloud build system was to only reference input nodes this
+situation can arise.
+
+\item[Communication] When traces or contents are stored on a central server
+communication can become a bottleneck, so it is important to send only the
+minimum amount of information, optimising with respect to build-system specific
+invariants~--~see~\S\ref{sec-smart-traces} for some possible optimisations.
+
+\item[Offloading] Once the cloud is storing build products and traces, it is
+possible for the cloud to also contain dedicated workers that can execute tasks
+remotely~--~offloading some of the computation and potentially running vastly
+more commands in parallel.
 \end{description}
@@
 
-Some build systems, for example \Excel and \Ninja, are capable of recomputing a task if either its dependencies change, \textit{or} the rule itself changes. For example:
+Some build systems, for example \Excel and \Ninja, are capable of recomputing a task if either its dependencies change, \emph{or} the rule itself changes. For example:
 
 \begin{minted}[xleftmargin=10pt]{text}
-A1 = 20            B1 = A1+A2
+A1 = 20      B1 = A1 + A2
 A2 = 10
@@
 
-In \Excel the user can alter the value produced by \textsf{B1} by either editing the inputs of \textsf{A1} or \textsf{A2}, \textit{or} editing the formula in \textsf{B1} - e.g.
-to \textsf{A1-A2}.
-This pattern can be captured by describing the rule producing \textsf{B1} as also depending on the value \textsf{B1-formula}.
-In a \hs{Monad} build system the implementation can be given very directly as:
+\noindent
+In \Excel the user can alter the value produced by \cmd{B1} by either editing
+the inputs of \cmd{A1} or \cmd{A2}, \emph{or} editing the formula in
+\cmd{B1}~--~e.g. to \cmd{A1 - A2}. This pattern can be captured by describing
+the rule producing \cmd{B1} as also depending on the value \cmd{B1-formula}.
+The implementation can be given very directly in a \hs{Task}~\hs{Monad} as:
 
@@
 
-Namely, first look up the formula, then interpret it.
-In an \hs{Applicative} system it is not possible to change dependencies based on the formula, as would be required, so instead the formula can be captured as a dependency (but its value not used) and \textit{also} baked directly into the \hs{task} function.
+\noindent
+Namely, first look up the formula, then interpret it. It is not possible to
+change dependencies based on the formula in a \hs{Task}~\hs{Applicative}, as
+would be required, so instead the formula can be captured as a dependency
+(but its value not used) and \emph{also} baked directly into the \hs{task}
+function.
 
-The build systems that have precise tracking of self-computation are all ones which use a non-Embedded Domain Specific Language to describe computations.
-Those which make use of a full programming language, e.g.
-\Shake, are much more difficult to implement equality on rules as they are arbitrary functions making use of opaque abstractions.
-For such build systems the incredibly pessimistic assuming of saying that any change to the build system potentially changes any rule can often be used -- the classic example being a \texttt{Makefile} depending on itself.
+The build systems that have precise self-tracking are all ones which use a
+\emph{non-embedded domain specific language} to describe computations.
+Those which make use of a full programming language, e.g. \Shake, are faced with
+the challenge of implementing equality on arbitrary task functions. For such
+build systems the incredibly pessimistic assumption of saying that any change to
+the build system potentially changes any build rule can often be used~--~the
+classic example being a makefile depending on itself.
 
@@
 
-Some computations are best described not by a chain of acyclic dependencies, but by a loop - for example \LaTeX requires repeated rebuilding until it reaches a fixed point - something that can be directly expressed in build systems such as Pluto \cite{pluto}.
-Another example of cyclic computations is \Excel, where a cell can depend on itself, for example:
+Some computations are best described not by a chain of acyclic dependencies,
+but by a loop~--~for example \LaTeX~requires repeated rebuilding until it
+reaches a fixed point~--~something that can be directly expressed in build
+systems, such as \Pluto~\cite{erdweg2015pluto}. Another example of cyclic
+computations is \Excel, where a cell can depend on itself, for example:
 
-\begin{minted}{bash}
+\begin{minted}[xleftmargin=10pt]{text}
 A1 = A1 + 1
@@
 
-In such cases \Excel will normally not execute anything, but if the ``Iterative Calculations'' feature is enabled will execute the formula a maximum of $n$ times per calculation (where $n$ is a setting that defaults to 100).
+\noindent
+In such cases \Excel will normally not execute anything, but if the
+``Iterative Calculations'' feature is enabled will execute the formula for a
+specified maximum number $N$ of times per calculation (where $N$ is a
+setting that defaults to 100).
 
-For examples like \LaTeX we consider the proper encoding to not be with circular rules, but with a series of iterative steps, as described in \S\ref{neil-iterative-on-stack-overflow}.
-It is important that the number of executions is bounded, otherwise the build system may not terminate (a legitimate concern with \LaTeX which can be put into a situation where it is bistable or diverging over multiple executions).
+For examples like \LaTeX~we consider the proper encoding to not be with circular
+tasks, but with a series of iterative steps, as described
+in~\cite{shake-fixed-point}. It is important that the number of executions is
+bounded, otherwise the build system may not terminate (a legitimate concern
+with \LaTeX, which can be put into a situation where it is bistable or diverging
+over multiple executions).
 
-The examples in \Excel tend to encode either mutable state, or recurrence relations.
-The former is only required because \Excel inherently lacks the ability to write mutable state, and the latter is probably better solved using explicit recurrence formulae.
+The examples in \Excel tend to encode either mutable state, or recurrence
+relations. The former is only required because \Excel inherently lacks the
+ability to write mutable state, and the latter is probably better solved using
+explicit recurrence formulae.
 
-Overall we choose not to deal with cyclic rules, a choice that most build systems also follow.
+Overall we choose not to deal with cyclic rules, a choice that most build
+systems also follow.
 
@@
 
-Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build system that works directly on \hs{k} and \hs{v} values.
-However, certain build systems provide greater flexibility, e.g.
-\Shake permits polymorphic keys and values, allowing types that are only stored in the Shake information, and never persisted to the Store.
+Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build
+system that works directly on \hs{k} and \hs{v} values. However, certain build
+systems provide greater flexibility, e.g. \Shake permits polymorphic keys and
+values, allowing types that are only stored in the \Shake information, and never
+persisted to the store.
 
-As one example of richer keys/values, consider the version of \textsf{gcc} - for many builds it should be a dependency.
-In \Shake we can define an \textit{oracle} rule as per \cite{mitchell2012shake} whose value is the result of running \hs{gcc --version} and which is volatile, making the \textsf{gcc} version something that can be depended upon.
-Of course, provided the build can express volatile dependencies and supports cutoff, the version number could equally be written to a file and used in a similar way.
+As one example of richer key/value types, consider the version of
+\cmd{gcc}~--~for many builds it should be a dependency. In \Shake it is possible
+to define an \emph{oracle} rule as per~\cite{mitchell2012shake} whose value is
+the result of running \cmd{gcc --version} and which is volatile, making the
+\cmd{gcc} version something that can be depended upon. Of course, provided the
+build can express volatile dependencies and supports cutoff, the version number
+could equally be written to a file and used in a similar way.
 
-A more compelling example is build rules that produce multiple keys -- for example \textsf{ghc Foo.hs} produces both \textsf{Foo.hi} and \textsf{Foo.o}.
-That can be represented by having a key whose value is a pair of file names, and whose result is a pair of file contents.
-From that, the rule for \textsf{Foo.hi} can be the first-component of the result of the pair.
-Again, such an operation can be encoded without polymorphic keys provided the pair of files (or a dummy file representing the pair) is marked as changed if either of the contained files change.
-Once again, polymorphic dependencies provide convenience rather than power.
+A more compelling example is build tasks that produce multiple output
+keys~--~for example, \cmd{ghc Foo.hs} produces both \cmd{Foo.hi} and \cmd{Foo.o}.
+That can be represented by having a key whose value is a pair of file names, and
+whose result is a pair of file contents. From that, the rule for \cmd{Foo.hi}
+can be the first component of the result of the pair. Again, such an operation
+can be encoded without polymorphic keys provided the pair of files (or a dummy
+file representing the pair) is marked as changed if either of the contained
+files change. Once again, polymorphic dependencies provide convenience rather
+than power.
 
-Authors have remarked that polymorphism provides a much easier expression of concepts, e.g. \citet{hadrian}, but it is not essential and thus not necessary to model.
+\Shake users have remarked that polymorphism provides a much easier expression
+of concepts, e.g.~\cite{hadrian}, but it is not essential and thus not necessary
+to model.

### 2018-03-16 paper/4-build.tex
Work on S4

@@
 
-The focus of this paper is on a variety of implementations of \hs{Build}, given
-a \emph{client-supplied} implementation of \hs{Task}. That is, we are going to take
-\hs{Task} as given from now on, and explore variants of \hs{Build}: first
-abstractly (in this section) and then concretely in~\S\ref{sec-implementations}.
+The focus of this paper is on a variety of implementations of
+\hs{Build}~\hs{c}~\hs{i}~\hs{k}~\hs{v}, given
+a \emph{client-supplied} implementation of \hs{Task}~\hs{c}~\hs{k}~\hs{v}. That
+is, we are going to take \hs{Task} as given from now on, and explore variants of
+\hs{Build}: first abstractly (in this section) and then concretely
+in~\S\ref{sec-implementations}.
 
-\todo{SLPJ}{What does this para add?}
-Let \hs{result}~\hs{=}~\hs{build}~\hs{task}~\hs{key}~\hs{store}. The build
-system \hs{build} is responsible for ensuring that \hs{key} and its dependencies
-are up to date in \hs{result} starting from \hs{store} (\S\ref{sec-build-correctness}). As part of that work,
-\hs{build} \emph{may} store auxiliary information \hs{info} of type \hs{i} in
-the \hs{store}, which will be provided to future calls of \hs{build}.
+As per the definition of minimality~\ref{def-minimal}, a minimal build
+system must \underline{rebuild only out-of-date keys} and at most once. The only
+way to achieve the ``at most once'' requirement while producing a correct build
+result (\S\ref{sec-build-correctness}) is to \underline{build all keys in the
+order that respects their dependencies}.
 
-\todo{SLPJ}{What does this para add?}
-All the build systems we define are polymorphic in both \hs{k} and \hs{v}
-(perhaps adding constraints to them), and have specific instantiations of \hs{c}
-and \hs{i}. As an example, recall that the \hs{busy} build system defined
-in~\S\ref{sec-general-build}, has \hs{c}~\hs{=}~\hs{Monad}, i.e. it supports
-monadic tasks, and \hs{i = ()}, i.e. it does not store any build information.
-
-According to the definition of minimality~\ref{def-minimal}, a minimal build
-system must \textbf{rebuild only out-of-date keys} and at most once. The only
-way to achieve the ``at most once'' requirement is to \textbf{build all keys in
-the order that respects their dependencies}.
-
-We have emphasised two different phrases above, and tackle each aspect separately.
+\vspace{1mm}
+We have underlined two different phrases above, and tackle each aspect separately.
 
@@
 An alternative approach, utilised by the \hs{busy} build system
-(\S\ref{sec-general-build}) is to simply build dependencies when they are
+(\S\ref{sec-general-build}), is to simply build dependencies when they are
 requested. By combining that with a transient log of which keys have already
@@
 \emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
-next build starts anything that changed between the two states is marked
+next build starts, anything that changed between the two states is marked
 dirty.
@@
 % what needs to rebuild.
-When reaching a key, if it and all its dependencies are
-clean, the key does not need recomputing.
+When reaching a key, if it and all its transitive dependencies are clean, the
+key does not need recomputing.
 
 \Excel models the dirty bit approach most directly, having an actual dirty bit
-associated with each cell, marking the cell it (and its dependencies) dirty
-if the user modifies it.
-When rebuilding,
-if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and
-marked dirty itself.
-\todo{SLPJ}{What does ``marked dirty itself'' mean?}
+associated with each cell, marking the cell dirty if the user modifies it.
+When rebuilding, if a cell only depends on clean cells it is skipped, otherwise
+it is rebuilt and is marked dirty so that the cells that depend on it are
+subsequently rebuilt too.
 
@@
 (\S\ref{sec-background-shake}) it would be important to not set the dirty bit
-after a computation that did not change the value. \Excel could model this
-approach, but does not. In contrast, \Make cannot model early cutoff -- to do so
-it would have to mark the node clean (so it didn't rebuild in the next run) and
-at the same time not mark the things it depends on dirty -- an impossible task
-with only the ability to update to the latest modification time.
+after a computation that did not change the value. \Excel could use this
+approach, but does not. In contrast, \Make cannot implement early cutoff -- to
+do so it would have to mark the node clean (so it would not rebuild in the next
+run) and at the same time not mark the things it depends on dirty -- an
+impossible task with only the ability to update to the latest modification time.
+
+\todo{NM}{I stopped here, as the traces seem not finalised yet.}
 
@@
 changed, the key is dirty and must be recomputed -- in essence a \emph{trace}
-which we can use to \textit{verify} existing values. We can describe a trace as:
-
-\todo{AM/NM}{Change from \hs{[Trace k v]} to \hs{Map k (Trave k v)} or something
-similar, so that it's obvious we have one trace per key .}
+which we can use to \emph{verify} existing values. We can describe a trace as:
 
@@
     { key          :: k
-    , dependencies :: [(k, Hash)]
-    , result       :: Hash }
+    , dependencies :: [(k, Hash v)]
+    , result       :: Hash v }
 \end{minted}
 
-We assume that \hs{Hash} is a small constant size, constructed from hashing the
+We assume that \hs{Hash v} is a small constant size, constructed from hashing the
 underlying \hs{v} rather than storing it directly. Checking a trace requires
@@
 same as the current value and that the result is the same. After computing a
-fresh value we add its \hs{Trace} to the list. Therefore the information stored
+fresh value we add its \hs{Trace k v} to the list. Therefore the information stored
 by a build system that verifies traces is \hs{[Trace k v]}.
+
+\todo{NM}{We need to emphasise that lists are convenient for expressing the semantics,
+but actual implementation uses a data structure optimised for a specific build system.}
 
@@
 A verifying trace allows us to mark a key dirty and rebuild it.
-Extending that information we can store a \textit{constructive} trace which is the trace plus the actual result.
+Extending that information we can store a \emph{constructive} trace which is the trace plus the actual result.
 Once we are storing the complete result it makes sense to record many constructive traces per key, and to share them with other users, providing cloud-build functionality.
@@
 
-\todo{AM/NM}{Traces -- why the indirection (first get the hash, then look it up)?
-Can we drop it? Can we switch to a more efficient/explicit data structure than a list?}
+\todo{NM}{Clarify why the indirection (first get the hash, then look it up)?}
 
@@
 We have a list of traces, plus a \hs{Map} from the hash to the actual contents.
-Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \textit{without} recomputing it.
+Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \emph{without} recomputing it.
 
@@
 
-\todo{AM}{AM: I would also like to see actual \hs{verify} and \hs{construct} functions.
+\subsection{Summary}\label{sec-design-space}
 
-NM: It actually gets a bit more confusing than that - since you end up needing to pass functions to get yourself, your dependencies etc.
-Unclear there is a win.
-The functions would occur in the next section anyway.
-
-AM: Now, after we've switched to \hs{Store i k v} it might work better?}
-
-\subsection{Summary} \label{sec-design-space}
-
-\todo{AM/NM/SLPJ}{Add a decription for the table.}
+\todo{NM}{Check the table and add a decription.}
 

### 2018-03-16 src/Development/Build/Task.hs
Add more examples of non-deterministic tasks

@@ -108,6 +108,10 @@ sprsh2 fetch "B2" = Just $ do c1 <- fetch "C1"
                               if c1 == 1 then fetch "A1" else fetch "B1"
 sprsh2 _     _    = Nothing
 
+sprsh3 :: Task MonadPlus String Integer
+sprsh3 fetch "B1" = Just $ (+) <$> fetch "A1" <*> pure 1 `mplus` pure 2
+sprsh3 _     _    = Nothing
+
 -- These type synonyms are not very useful, but enumerate all interesting cases.
 type FunctorialTask  k v = forall f. Functor     f => (k -> f v) -> k -> Maybe (f v)
 type ApplicativeTask k v = forall f. Applicative f => (k -> f v) -> k -> Maybe (f v)

### 2018-03-16 src/Development/Build/Task/Alternative.hs

@@ -4,7 +4,6 @@ module Development.Build.Task.Alternative (
     ) where
 
 import Control.Applicative
-import Data.Foldable
 import Data.Maybe
 
 import Development.Build.Task
@@ -19,7 +18,7 @@ failingTask _ _ = Just empty
 (|||) task1 task2 fetch key = task1 fetch key <|> task2 fetch key
 
 random :: (Int, Int) -> Task Alternative k Int
-random (low, high) _ _ = asum $ map (Just . pure) [low..high]
+random (low, high) _ _ = Just $ foldr (<|>) empty $ map pure [low..high]
 
 dependencies :: Task Alternative k v -> k -> [[k]]
 dependencies task = getAltConst . sequenceA . task (\k -> AltConst [[k]])

### 2018-03-16 src/Development/Build/Task/MonadPlus.hs

@@ -1,7 +1,7 @@
 {-# LANGUAGE ConstraintKinds, FlexibleContexts, RankNTypes #-}
 module Development.Build.Task.MonadPlus (
-    dependencies, transitiveDependencies, acyclic, inputs, isInput, consistent,
-    pure
+    random, dependencies, transitiveDependencies, acyclic, inputs, isInput,
+    consistent, pure
     ) where
 
 import Control.Monad
@@ -12,6 +12,9 @@ import Data.Maybe
 
 import Development.Build.Task
 import Development.Build.Utilities
+
+random :: (Int, Int) -> Task MonadPlus k Int
+random (low, high) _ _ = Just $ foldr mplus mzero $ map pure [low..high]
 
 -- Can we have something like this instead?
 -- dependencies :: c m => Task c k v -> (k -> m v) -> k -> m [k]

### 2018-03-16 paper/5-implementations.tex
Mostly TODOs for S5

@@
-\clearpage
 \section{Build systems, concretely}\label{sec-implementations}
 
-In the previous sections we discussed the types of build system, and how they broken down.
-But these divisions were not obvious to us, and only by concretely implementing and refactoring each build system did we determine the underlying commonalities.
-In this section we share some of the code that got us there, but full code is available in the supporting material and at \url{url-hidden-for-anonymous-submission}.
-Our code comprises a framework, utility functions implementing the various dependency orderings as per \S\ref{sec-dependency-orderings}, and actual implementations.
+In the previous sections we discussed the types of build system, and how they
+can be broken down. But these divisions were not obvious to us, and only by
+concretely implementing and refactoring each build system did we determine the
+underlying commonalities. In this section we share some of the code that got us
+there. Our code comprises a framework, utility functions implementing the
+various dependency orderings as per~\S\ref{sec-dependency-orderings}, and
+actual implementations.
 
-\subsection{Framework}
+% \subsection{Framework}
 
-todo{SLPJ}{I think this entire section, including the mysterious M, can be deleted.  Or is
-there anything that remains?}
+% todo{SLPJ}{I think this entire section, including the mysterious M, can be deleted.  Or is
+% there anything that remains?}
 
-As we saw in \S\ref{sec-build}, a build system can be defined as:
+% \begin{figure}
+% \begin{minted}{haskell}
+% type M i k v r = State (i, k -> v, ...) r
+% runM :: Default i => M i k v a -> Maybe i -> (k -> v) -> (i, k -> v)
 
-\begin{minted}{haskell}
-type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
-\end{minted}
+% getStore :: k -> M i k v v
+% putStore :: k -> v -> M i k v ()
+% getInfo :: M i k v i
+% modifyInfo :: (i -> i) -> M i k v ()
+% \end{minted}
+% Note: we have omitted the context on \hs{k}, things like \hs{Eq} (to implement the store).
+% \caption{API with which to implement build systems}
+% \label{fig-M-api}
+% \end{figure}
 
-Using the framework we build in this section, we can express a dumb build system as:
-
-\begin{minted}{haskell}
-dumb :: Build Monad () k v
-dumb compute k = runM (f k)
-    where
-        f k = case compute f k of
-            Nothing -> getStore k
-            Just act -> do
-                v <- act
-                putStore k v
-                return v
-\end{minted}
-
-The \hs{dumb} build system can build with monadic computes, using a function \hs{f} which given a key returns the associated value.
-To do so, \hs{f} calls \hs{compute} passing \hs{f} itself as an argument.
-If the result is \hs{Nothing} (this key is an input) it simply reads from the store.
-Otherwise it runs the action to produce a value, stores it and returns it.
-
-\begin{figure}
-\begin{minted}{haskell}
-type M i k v r = State (i, k -> v, ...) r
-runM :: Default i => M i k v a -> Maybe i -> (k -> v) -> (i, k -> v)
-
-getStore :: k -> M i k v v
-putStore :: k -> v -> M i k v ()
-getInfo :: M i k v i
-modifyInfo :: (i -> i) -> M i k v ()
-\end{minted}
-Note: we have omitted the context on \hs{k}, things like \hs{Eq} (to implement the store).
-\caption{API with which to implement build systems}
-\label{fig-M-api}
-\end{figure}
-
-We use \hs{runM} to switch between the pure world of \hs{Build} and a state monad \hs{M} which lets us more precisely express the ideas behind the build system without threading the state manually - the types of important operations are given in Figure~\ref{fig-M-api}.
+% We use \hs{runM} to switch between the pure world of \hs{Build} and a state monad \hs{M} which lets us more precisely express the ideas behind the build system without threading the state manually - the types of important operations are given in Figure~\ref{fig-M-api}.
 We provide functions like \hs{getStore} to load from the store and \hs{putStore} to write to the store.
@@
 
+\todo{AM}{Explain the new implementation. Most of the explanations below will
+need to be adapted.}
+
 The \Make build system builds files in a linear order based on a topological sort.
-(See \S\ref{sec-concurrency} for parallel \Make.)
-For each file, it builds it if the output file is older than any of the inputs. We can express this in our framework with:
+(See \S\ref{sec-concurrency} for parallel \Make.) For each file, it builds it if
+the output file is older than any of the inputs. We can express this in our
+framework with:
 
 \begin{minted}{haskell}
-make :: Eq v => Build Applicative (Changed k v, ()) k v
-make = withChanged $ topological $ \k dk act -> do
-    t <- getStoreTime k
-    dt <- mapM getStoreTime dk
-    let dirty = any (>= t) dt
-    when dirty (putStore k =<< act)
+type Time = Integer
+type MakeInfo k = (k -> Time, Time)
+
+make :: Eq k => Build Applicative (MakeInfo k) k v
+make = topological $ \key deps act -> do
+    (modTime, now) <- getInfo <$> get
+    let dirty = or [ modTime dep > modTime key | dep <- deps ]
+    when dirty $ do
+        v <- act
+        let newModTime k = if k == key then now else modTime k
+        modify $ \s -> putInfo (putValue s key v) (newModTime, now + 1)
 \end{minted}
@@
 \begin{minted}{haskell}
-topological
-    :: (m ~ M i k v, Default i)
-    => (k -> [k] -> m v -> m ())
+topological :: Eq k
+            => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
             -> Build Applicative i k v
-topological step compute k = runM $ do
-    let depends = dependencies compute
-    forM_ (topSort depends $ transitiveClosure depends k) $ \k ->
-        case compute getStore k of
+topological step task key = execState $ forM_ chain $ \k -> do
+    let fetch k = do { store <- get; return (getValue store k) }
+    case task fetch k of
         Nothing  -> return ()
-            Just act -> step k (depends k) act
+        Just act -> step k (deps k) act
+  where
+    deps  = dependencies task
+    chain = topSort deps (closure deps key)
 \end{minted}
@@
 
-\begin{minted}{haskell}
-getChanged :: Eq v => k -> M (Changed k v, i) k v Bool
-withChanged :: Build c (Changed k v, i) k v -> Build c (Changed k v, i) k v
-
-data Time = LastBuild | AfterLastBuild
-getStoreTime :: Eq v => k -> M (Changed k v, i) k v Time
-\end{minted}
-
 To model a dirty bit we extend the \hs{i} information with \hs{Changed} to capture the \hs{k -> v} store at the end of the previous run, and extend \hs{M} to record which files have been written to during this \hs{runM}.
 A file is considered dirty if either the current value as per \hs{getStore} has changed since the last run, or if it has had \hs{putStore} called on it\footnote{It would be reasonable to declare files unchanged if they have \hs{putStore} called at their previous value, but \Excel does not and \Make cannot using modification times, so we ignore this option.}.
-To track the values at the last run we use \hs{withChanged} to copy the resultant \hs{k -> v} store into the info after the build completes\footnote{For tedious reasons our implementation requires \hs{withChangedApplicative} and \hs{withChangedMonad} which are variants of \hs{withChanged} for concrete constraints \hs{c} - but with identical implementations.}
+To track the values at the last run we use \hs{withChanged} to copy the resultant \hs{k -> v} store into the info after the build completes.
 
@@
 \subsection{\Excel}\label{sec-implementation-excel}
+
+\todo{AM}{I might do some simplification of this tomorrow. Neil: I guess it's best
+for you to focus on nailing traces and associated \Shake and \Bazel sections.}
 
@@
 \subsection{\Shake}\label{sec-implementation-shake}
+
+\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
 
@@
 
+\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
+
 Now we have seen the three dependency schemes, we can directly reuse \hs{topological} to define \Bazel.
@@
 
+\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
+
 Using the abstractions and approaches built thus far, we have shown how to combine dependency scheme and change approach to reproduce existing build systems.
@@
 \subsection{Smarter \hs{[Trace]} Structures}\label{sec-smart-traces}
+\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
 

### 2018-03-16 paper/7-related.tex
Work on S7, add all references, drop Appendix

@@
 
-While there is research on individual build systems, there has been little research to date comparing different build systems.
-In \S\ref{sec-background} we cover several important build systems -- in this section we relate a few additional build systems to our abstractions, and discuss other places where similar abstractions arise.
+While there is research on individual build systems, there has been little
+research to date comparing different build systems. In~\S\ref{sec-background} we
+covered several important build systems~--~in this section we relate a few
+other build systems to our abstractions, and discuss other work where similar
+abstractions~arise.
 
@@
 \begin{itemize}
-\item \Ninja \cite{ninja} combines the dependency strategy of \Make with the validating traces of \Shake - our associated implementation provides such a combination.
-Ninja is also capable of modelling rules that produce multiple results, a limited form of polymorphism \S\ref{sec-polymorphism}.
-\item \Tup \cite{tup} functions much like \Make, but with a refined dirty-bit implementation that watches the file system for changes and can thus avoid rechecking the entire graph, and with automatic deleting of stale results.
-\item \Redo \cite{redo} almost exactly matches \Shake at the level of detail given here, differing only on aspects like polymorphic dependencies \S\ref{sec-polymorphism}.
-\item \Buck \cite{buck} and \CloudBuild \cite{cloudbuild} are both fundamentally similar to \Bazel, differing around areas such as non-determinism \S\ref{sec-non-determinism}.
-\item \Nix \cite{nix} has coarse-grained dependencies, with precise hashing of dependencies and downloading of precomputed build products.
-When combined with \hs{import-from-derivation} \Nix can also be considered monadic, making it similar to Cloud Shake from \S\ref{sec-implementation-cloud-shake}.
-However, \Nix is not intended as a build system, and the coarse grained nature (packages, not individual files) makes it targeted to a different purpose.
+\item \Ninja~\cite{ninja} combines the dependency strategy of \Make with the
+validating traces of \Shake~--~our associated implementation provides such a
+combination. Ninja is also capable of modelling rules that produce multiple
+results, a limited form of polymorphism \S\ref{sec-polymorphism}.
+
+\item \Tup~\cite{tup} functions much like \Make, but with a refined dirty-bit
+implementation that watches the file system for changes and can thus avoid
+rechecking the entire graph, and with automatic deleting of stale results.
+
+\item \Redo~\cite{redo} almost exactly matches \Shake at the level of detail
+given here, differing only on aspects like polymorphic
+dependencies~\S\ref{sec-polymorphism}.
+
+\item \Buck~\cite{buck} and \CloudBuild~\cite{esfahani2016cloudbuild} are both
+fundamentally similar to \Bazel, differing around areas such as
+non-determinism~\S\ref{sec-non-determinism}.
+
+\item \Nix~\cite{dolstra2004nix} has coarse-grained dependencies, with precise
+hashing of dependencies and downloading of precomputed build products. When
+combined with \cmd{import-from-derivation}, \Nix can also be considered monadic,
+making it similar to Cloud \Shake from~\S\ref{sec-implementation-cloud-shake}.
+However, \Nix is not intended as a build system, and the coarse grained nature
+(packages, not individual files) makes it targeted to a different purpose.
 % John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
 % \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.
-\item \Pluto \cite{pluto} is based on a similar model to \Shake, but additionally allows cyclic build rules combined with a user-specific resolution strategy.
-Often such a strategy can be unfolded into the user rules without loss of precision, but a fully general resolution handler extends the \hs{Task} abstraction with additional features.
+
+\item \Pluto~\cite{erdweg2015pluto} is based on a similar model to \Shake, but
+additionally allows cyclic build rules combined with a user-specific resolution
+strategy. Often such a strategy can be unfolded into the user rules without loss
+of precision, but a fully general resolution handler extends the \hs{Task}
+abstraction with additional features.
 \end{itemize}
 
-The one build system we are aware of that cannot be modelled in our framework is \Fabricate \cite{fabricate}.
-In \Fabricate a build system is a script which is run in-order, in the spirit of\footnote{\Fabricate requires scripts to be written in Python, but those details are not fundamental to what makes \Fabricate special.}:
+The one build system we are aware of that cannot be modelled in our framework is
+\Fabricate~\cite{fabricate}. In \Fabricate a build system is a script which is
+run in-order, in the spirit of\footnote{\Fabricate requires scripts to be
+written in Python, but those details are not fundamental to what makes
+\Fabricate special.}:
 
-\begin{minted}{bash}
+\begin{minted}[xleftmargin=10pt]{bash}
 gcc -c util.c
@@
 
-To achieve minimality, each separate command is traced at the OS-level, allowing \Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from \cmd{util.c}.
-In future runs \Fabricate runs the script from start to finish, skipping any commands where no inputs have changed.
+\noindent
+To achieve minimality, each separate command is traced at the OS-level, allowing
+\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
+\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
+skipping any commands where no inputs have changed.
 
-Taking our abstraction, it is possible to encode \Fabricate assuming that nodes like \cmd{gcc -c util.c} are keys, there is a linear dependency between each successive node, and that the OS-level tracing can be lifted back as a monadic \hs{Task} function\footnote{In fact, \Shake has an execution mode that can model \Fabricate{}-like build systems -- see \hs{Development.Shake.Forward} in the \Shake library.}.
-However, in our pure model the mapping is not perfect as \hs{gcc} writes to arbitrary files whose locations are not known in advance.
+Taking our abstraction, it is possible to encode \Fabricate assuming that
+commands like \cmd{gcc -c util.c} are keys, there is a linear dependency between
+each successive node, and that the OS-level tracing can be lifted back as a
+monadic \hs{Task} function\footnote{In fact, \Shake has an execution mode that
+can model \Fabricate{}-like build systems~--~see \hs{Development.Shake.Forward}
+in the \Shake library.}. However, in our pure model the mapping is not perfect
+as \cmd{gcc} writes to arbitrary files whose locations are not known in advance.
 
-% \todo{AM}{Not convined this is really related, but if you think it is, please write something.}
+\todo{AM}{Write a short section about memoization if enough time.}
 % \subsection{Memoization}\label{sec-related-memo}
@@
 
-While not typically considered build systems, self-adjusting computation is a well studied area, and in particular the contrast between different formulations has been well studied.
+While not typically considered build systems, self-adjusting computation is a
+well studied area, and in particular the contrast between different formulations
+has been well studied.
+
 \todo{AM}{Write this section.}
@@
 
-\begin{minted}{haskell}
+\begin{minted}[xleftmargin=10pt]{haskell}
 type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
@@
 
-Which looks tantalisingly close to the profunctor optics definition by \citet{gibbons-profunctor-optics}:
+\noindent
+Which looks tantalisingly close to the profunctor optics definition by
+\cite{pickering2017profunctor}:
 
@@
 
-Provided we instantiate \hs{p} to something like \hs{k -> f v} -- which many of the actual instances in that paper do.
-The properties of such optics are well studied, and the functions like \hs{dependencies} are very much based on observations from that field of work.
-Alas, we have been unable to remove the \hs{Maybe} used to encode whether a file is an input, without complicating other aspects of our definition.
-Furthermore, the \hs{Build} operation lacks any further such symmetry.
-
-% \subsection{Free applicative functors and monads}\label{sec-related-free}
-%
-% NM: Unconvinced it is worth it.
-% \todo{AM}{Mention that one could have chosen a free construction to create
-% a small DSL for build tasks. Although this simplifies some of the presented
-% combinators, in general this seems unnecessary.}
+Provided we instantiate \hs{p} to something like
+\hs{k}~\hs{->}~\hs{f}~\hs{v}~--~which many of the actual instances in that paper
+do. The properties of such optics are well studied, and the functions like
+\hs{dependencies} are very much based on observations from that field of work.
+Alas, we have been unable to remove the \hs{Maybe} used to encode whether a file
+is an input, without complicating other aspects of our definition. Furthermore,
+the \hs{Build} operation lacks any further such symmetry.

### 2018-03-16 paper/4-build.tex
4-build

@@
 
-\subsubsection{Topological}
+\subsubsection{Topological}\label{sec-topological}
 
@@
 
-\subsubsection{Reordering}
+\subsubsection{Reordering}\label{sec-reordering}
 
@@
 
-This requires a way to abort tasks that have failed due to out-of-date
+This approach requires a way to abort tasks that have failed due to out-of-date
 dependencies. It is also not minimal in the sense that a task may start, do some
@@
 
-\subsubsection{Recursive}
+\subsubsection{Recursive}\label{sec-recursive}
 
@@
 
-This requires a way to suspend a running task, which can be done with cheap
+This approach requires a way to suspend a running task, which can be done with cheap
 green threads and blocking (the original approach of \Shake) or with a
@@
 
-With a dirty bit it is possible to achieve minimality. To achieve early cutoff
+When using a dirty bit, it is necessary to check all the dependencies of a key.
+For applicative build systems that list is simple to obtain, but for monadic build systems there is no general way to get all dependencies.
+Instead \Excel computes a \textit{static approximation} of the dependencies.
+For applicative rules that approximation is correct.
+For functions such as \textsf{IF} it marks the cell dirty if \textit{any} potential dependency has changed, even on the untaken \textif{if} branch.
+For functions such as \textsf{INDIRECT} whose dependencies cannot be guessed, it conservatively assumes the dependencies have always changed.
+
+With a dirty bit it is simple to achieve minimality. However, to achieve early cutoff
 (\S\ref{sec-background-shake}) it would be important to not set the dirty bit
@@
 impossible task with only the ability to update to the latest modification time.
-
-\todo{NM}{I stopped here, as the traces seem not finalised yet.}
 
@@
 as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are
-same as the current value and that the result is the same. After computing a
-fresh value we add its \hs{Trace k v} to the list. Therefore the information stored
-by a build system that verifies traces is \hs{[Trace k v]}.
+same as the current value and that the result is the same.
 
-\todo{NM}{We need to emphasise that lists are convenient for expressing the semantics,
-but actual implementation uses a data structure optimised for a specific build system.}
+A verifying trace build system needs to maintain a set of traces. After computing a
+fresh value we add its \hs{Trace k v} to the set. Therefore the information stored
+by a build system that verifies traces can be stored as a list or set - we chose \hs{[Trace k v]} for simplicity.
+In practice, different build systems can optimise the data structures used by traces for their specific use cases, which we discuss in \S\ref{sec-smart-traces}.
 
@@
 
-\todo{NM}{Clarify why the indirection (first get the hash, then look it up)?}
-
 \begin{minted}{haskell}
@@
     { traces :: [Trace k v]
-    , cache  :: Map Hash v }
+    , cache  :: Map (Hash v) v }
 \end{minted}
@@
 Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \emph{without} recomputing it.
+We split the traces and caches as in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only at most one complete file.
 
-For both verifying and constructive choices it is possible to choose smarter data structures than simply a list of \hs{Trace}, which we discuss in \S\ref{sec-smart-traces}.
-
-\subsection{Summary}\label{sec-design-space}
-
-\todo{NM}{Check the table and add a decription.}
+\subsection{Build Systems \`a la Carte}\label{sec-design-space}
 
@@
 \hline
-Property           & Topological & Reordering & Recursive    \\\hline
+Property           & Topological \S\ref{sec-topological} & Reordering \S\ref{sec-reordering} & Recursive \S\ref{sec-recursive}    \\\hline
 \hline
-Dirty bit          & \Make                   & \Excel                   & \Shake without early cutoff \\\hline
-Verifying trace    & \Make with early cutoff & \Excel with early cutoff & \Shake                      \\\hline
-Constructive trace & \Bazel                  & Cloud \Excel             & Cloud \Shake                \\\hline
+Dirty bit \S\ref{sec-dirty-bit}                    & \Make    & \Excel                   & Approximate \Shake  \\\hline
+Verifying trace \S\ref{sec-verifying-traces}       & \Ninja   & Traced \Excel            & \Shake                      \\\hline
+Constructive trace \S\ref{sec-constructive-traces} & \Bazel   & Cloud \Excel             & Cloud \Shake                \\\hline
 \end{tabular}
 \vspace{0.5mm}
-\caption{Build systems \`a la carte.\label{tab-build-systems}}
+\caption{Build systems \`a la Carte.\label{tab-build-systems}}
 \end{table}
+
+With the information in this section we can build a table comparing the dependency order strategy with the out-of-date keys strategy, providing 9 possible build systems, 5 of which are actually inhabited by existing build systems (we discuss \Ninja \cite{ninja} in \S\ref{sec-related-build}).
+Of the remaining 4 spots, we believe neither Traced or Cloud \Excel make sense - the \Excel approach of reordering combined with static approximations reduces the memory usage significantly.
+However, as soon as you are paying the cost of storing traces, that benefit is gone.
+The advantage of Approximate \Shake over \Shake would be that it could avoid storing traces and having a separate information database, but that advantage is minor compared to the technical restrictions and approximations it would provide, so we consider it unlikely to be built.
+The Cloud \Shake system seems like an interesting and important point in the design space, which we explore further in \S\ref{sec-implementation-cloud-shake}.

### 2018-03-16 paper/4-build.tex
Fix textif

@@
 For applicative build systems that list is simple to obtain, but for monadic build systems there is no general way to get all dependencies.
-Instead \Excel computes a \textit{static approximation} of the dependencies.
+Instead \Excel computes a \emph{static approximation} of the dependencies.
 For applicative rules that approximation is correct.
-For functions such as \textsf{IF} it marks the cell dirty if \textit{any} potential dependency has changed, even on the untaken \textif{if} branch.
-For functions such as \textsf{INDIRECT} whose dependencies cannot be guessed, it conservatively assumes the dependencies have always changed.
+For functions such as \cmd{IF} it marks the cell dirty if \emph{any} potential dependency has changed, even on the untaken \emph{if} branch.
+For functions such as \cmd{INDIRECT} whose dependencies cannot be guessed, it conservatively assumes the dependencies have always changed.
 

### 2018-03-16 paper/1-intro.tex
More changes

@@
 a dozen lines of code or so (\S\ref{sec-implementations}).
-Doing this in a single setting allows
+Doing this modelling in a single setting allows
 the differences and similarities between these huge systems to be
@@
 
-\item Moreover, we can readily mix the ingredients differently to produce the first
+\item Moreover, we can readily remix the ingredients to produce the first
   build system that supports both \emph{dynamic dependencies}
@@
 abstractions are defined in Haskell, and we have (freely-available)
-executable models of all the build systems we describe.  An unusual
-feature is that we include \Excel in our line-up because, looked at
-in the right way, it certainly is a build system.
+executable models of all the build systems we describe.

### 2018-03-16 paper/2-background.tex
More changes

@@
 of build systems, and their relationships,
-in~\S\ref{sec-related}-\S\ref{sec-conclusions}.
+in~\S\ref{sec-related}.
 

### 2018-03-16 paper/4-build.tex
More changes

--- 2018-03-16 paper/4-build.tex
@@
 way to achieve the ``at most once'' requirement while producing a correct build
-result (\S\ref{sec-build-correctness}) is to \underline{build all keys in the
+result (\S\ref{sec-build-correctness}) is to \underline{build all keys in an
 order that respects their dependencies}.
@@
 
-This approach requires a way to suspend a running task, which can be done with cheap
+This approach requires that a task may be started, then during that execution
+another task will have to be run. Assuming a an IO-driven task structure,
+that requires suspend a running task, which can be done with cheap
 green threads and blocking (the original approach of \Shake) or with a
@@
 For applicative rules that approximation is correct.
-For functions such as \textsf{IF} it marks the cell dirty if \textit{any} potential dependency has changed, even on the untaken \textif{if} branch.
+For functions such as \textsf{IF} it marks the cell dirty if \textit{any} potential dependency has changed, even on the untaken \textsf{IF} branch.
 For functions such as \textsf{INDIRECT} whose dependencies cannot be guessed, it conservatively assumes the dependencies have always changed.
@@
 after a computation that did not change the value. \Excel could use this
-approach, but does not. In contrast, \Make cannot implement early cutoff -- to
+approach, but does not. In contrast, \Make cannot implement early cutoff nicely -- to
 do so it would have to mark the node clean (so it would not rebuild in the next
@@
 impossible task with only the ability to update to the latest modification time.
+It can approximate early cutoff by not touching the result file, and not marking it clean,
+but then it reruns in every subsequent build.
 

### 2018-03-16 paper/4-build.tex

@@
 For applicative build systems that list is simple to obtain, but for monadic build systems there is no general way to get all dependencies.
-Instead \Excel computes a \textit{static approximation} of the dependencies.
+Instead \Excel computes a \emph{static approximation} of the dependencies.
 For applicative rules that approximation is correct.
-For functions such as \textsf{IF} it marks the cell dirty if \textit{any} potential dependency has changed, even on the untaken \textsf{IF} branch.
-For functions such as \textsf{INDIRECT} whose dependencies cannot be guessed, it conservatively assumes the dependencies have always changed.
+For functions such as \cmd{IF} it marks the cell dirty if \emph{any} potential dependency has changed, even on the untaken \emph{if} branch.
+For functions such as \cmd{INDIRECT} whose dependencies cannot be guessed, it conservatively assumes the dependencies have always changed.
 

### 2018-03-16 src/Development/Build.hs
Mirror changes from the paper

@@ -62,11 +62,11 @@ memo task key store = fst $ execState (fetch key) (store, [])
 topological :: Eq k
             => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
             -> Build Applicative i k v
-topological step task key = execState $ forM_ chain $ \k -> do
-    let fetch k = do { store <- get; return (getValue store k) }
+topological process task key = execState $ forM_ chain $ \k -> do
+    let fetch k = do store <- get; return (getValue store k)
     case task fetch k of
         Nothing  -> return ()
-        Just act -> step k (deps k) act
+        Just act -> process k (deps k) act
   where
     deps  = dependencies task
     chain = topSort deps (closure deps key)
@@ -75,13 +75,15 @@ type Time = Integer
 type MakeInfo k = (k -> Time, Time)
 
 make :: Eq k => Build Applicative (MakeInfo k) k v
-make = topological $ \key deps act -> do
-    (modTime, now) <- getInfo <$> get
-    let dirty = or [ modTime dep > modTime key | dep <- deps ]
-    when dirty $ do
-        v <- act
-        let newModTime k = if k == key then now else modTime k
-        modify $ \s -> putInfo (putValue s key v) (newModTime, now + 1)
+make = topological process
+  where
+    process key deps act = do
+        (modTime, now) <- getInfo <$> get
+        let dirty = or [ modTime dep > modTime key | dep <- deps ]
+        when dirty $ do
+            v <- act
+            let newModTime k = if k == key then now else modTime k
+            modify $ \s -> putInfo (putValue s key v) (newModTime, now + 1)
 
 reordering :: forall i k v. Ord k
             => (k -> State (Store i k v) (Either k (v, [k])) -> State (Store i k v) (Maybe k))

### 2018-03-16 paper/2-background.tex

@@
 \Excel's build algorithm~\cite{excel_recalc} is significantly different from
-\Make. Since it is impossible to construct the full dependency graph upfront in
-the presence of dynamic dependencies, \Excel uses the \emph{calculation chain}
+\Make. \Excel uses the \emph{calculation chain}
 produced by the previous build as an approximation to the correct topological
@@
 \emph{defer recalculation of a cell} by moving it down the chain if a newly
-discovered dependency has not yet been rebuilt. We further refer to this
+discovered dependency has not yet been rebuilt. We refer to this
 algorithm as \emph{reordering}, and will discuss it in more detail
@@
 \Shake was developed to solve the issue of dynamic
-dependencies~\cite{mitchell2012shake} and it excels at them without sacrificing
-the minimality requirement. To demonstrate this, we elaborate the \Make example
-from~\S\ref{sec-background-make} by adding the following files whose
+dependencies~\cite{mitchell2012shake} without sacrificing
+the minimality requirement. Building on the \Make example
+from~\S\ref{sec-background-make}, we add the following files whose
 dependencies are shown in Fig.~\ref{fig-shake}(a):
@@
     \item \cmd{LICENSE} is an input text file containing the project license.
-    \item \cmd{release.txt} is a text file listing all release files. This file
+    \item \cmd{release.txt} is a text file listing all files that should be in the release. This file
     is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt}
@@
 determined by the content of \cmd{release.txt}, which might not even exist
-before the build. Makefiles cannot express such dependencies, requiring ad hoc
-workarounds such as \emph{build phases}, which are problematic~\cite{hadrian}.
-Here is how one can express the task for producing \cmd{release.tar} using \Shake.
+before the build. Makefiles cannot express such dependencies, requiring problematic
+workarounds such as \emph{build phases} \cite{hadrian}.
+In \Shake we can express the rule for \cmd{release.tar} as:
 
@@
 We first declare the static dependency on \cmd{release.txt}, then read its
-content (a list of files) and depend on it, dynamically. Finally, we specify the
+content (a list of files) and depend on each listed file, dynamically. Finally, we specify the
 command to produce the resulting archive. Crucially, the archive will only be
@@
 
-\Shake also supports the following \emph{early cutoff optimisation}. When it
+\Shake also supports the \emph{early cutoff optimisation}. When it
 executes a task and the result is unchanged from the previous build, it is

### 2018-03-16 paper/3-abstractions.tex
Work on S4

@@
 descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
-satisfies certain properties, such as \hs{Applicatove} or \hs{Monad}. That is
+satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
 why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
@@
     \[
-    \hs{compute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
+    \hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}.
     \]

### 2018-03-16 paper/5-implementations.tex
Simplify S5.1

@@
 
-In the previous sections we discussed the types of build system, and how they
+In the previous sections we discussed the types of build systems, and how they
 can be broken down. But these divisions were not obvious to us, and only by
@@
 underlying commonalities. In this section we share some of the code that got us
-there. Our code comprises a framework, utility functions implementing the
-various dependency orderings as per~\S\ref{sec-dependency-orderings}, and
-actual implementations.
+there.
+
+% Our code comprises a framework, utility functions implementing the
+% various dependency orderings as per~\S\ref{sec-dependency-orderings}, and
+% actual implementations.
 
@@
 
-\todo{AM}{Explain the new implementation. Most of the explanations below will
-need to be adapted.}
+\Make processes keys in a linear order based on a topological sort
+(see~\S\ref{sec-concurrency} for parallel \Make). For each key, it builds it if
+it is older than any of its dependencies. We can express this in our framework
+as follows. We capture the persistent information stored by \Make as a pair
+\hs{(@@modTime,}~\hs{now)} comprising the \emph{file modification time} function
+\hs{modTime}~\hs{::}~\hs{k}~\hs{->}~\hs{Time} and the current time \hs{now}.
 
-The \Make build system builds files in a linear order based on a topological sort.
-(See \S\ref{sec-concurrency} for parallel \Make.) For each file, it builds it if
-the output file is older than any of the inputs. We can express this in our
-framework with:
-
-\begin{minted}{haskell}
+\begin{minted}[xleftmargin=10pt]{haskell}
 type Time = Integer
 type MakeInfo k = (k -> Time, Time)
-
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 make :: Eq k => Build Applicative (MakeInfo k) k v
-make = topological $ \key deps act -> do
+make = topological process
+  where
+    process key deps act = do
         (modTime, now) <- getInfo <$> get
@@
 
-We'll come back to \hs{withChanged} afterwards, but ignoring that, reading the rest of the implementation you can see the underlying essence of \Make.
-We use the helper function \hs{topological} which gets called with each file in a dependency-respecting order given the file (\hs{k}), the dependencies (\hs{dk}) and an action to generate the result (\hs{act}).
-With those three things in hand, the implementation first gets the timestamp of the key being built, then the timestamps of the dependencies (which must have been rebuilt if necessary, thanks to the guarantees on \hs{topological}), and considers it dirty if any dependency was built after \hs{k}.
-If it is dirty, it rebuilds and stores the result.
+The helper function \hs{topological}, which we describe later, calls
+\hs{process} on every \hs{key} in the topological order, providing the list of
+its dependencies \hs{deps} and the action \hs{act} to compute the resulting
+value if it needs to be rebuilt. To determine if the \hs{key} is \hs{dirty},
+\hs{process} compares its modification time with those of its dependencies. If
+\hs{key} needs to be rebuilt, the associated action \hs{act} is executed and
+the obtained result is stored, along with an updated file modification
+timestamp. The \hs{make} build system, as described, captures the essence of
+\Make in a clear and precise manner.
 
-The implementation of \hs{topological} encodes the dependency strategy that \Make has chosen to use. The function itself is defined as:
+The implementation of \hs{topological} encodes the dependency strategy that
+\Make has chosen to use. The function itself is defined as:
 
+\vspace{1mm}
 \begin{minted}{haskell}
-topological :: Eq k
-            => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
+topological :: Eq@\,\blk{k}@ => (k -> [k] -> State@\,@(Store i k v)@\,\blk{v}@ -> State@\,@(Store i k v)@\,@(@@))
             -> Build Applicative i k v
-topological step task key = execState $ forM_ chain $ \k -> do
-    let fetch k = do { store <- get; return (getValue store k) }
+topological process task key = execState $ forM_ chain $ \k -> do
+    let fetch k = do store <- get; return (getValue store k)
     case task fetch k of
         Nothing  -> return ()
-        Just act -> step k (deps k) act
+        Just act -> process k (deps k) act
   where
     deps  = dependencies task
-    chain = topSort deps (closure deps key)
+    chain = topologicalSort deps (transitiveClosure deps key)
 \end{minted}
+\vspace{1mm}
 
-It uses a \hs{dependencies} on the computation function, takes the transitive-closure and topological-sort to provide an order in which to build the keys.
-For each key, if it is an input it does nothing, otherwise it calls the supplied \hs{step} function.
-Note that \hs{dependencies} is only defined for \hs{Applicative} computes, which is what restricts \Make to Applicative build systems.
-Moreover, anyone following the same topological-sort approach will also have the \hs{Applicative} restriction.
+\noindent
+The \hs{where} clause corresponds to the pre-processing step, which uses the
+function \hs{dependencies}, defined in~\S\ref{sec-deps}, to extract static
+dependencies from the provided applicative \hs{task}. We compute the linear
+processing \hs{chain} by taking the \hs{transitiveClosure} of \hs{key}'s
+dependencies, and performing the \hs{topologicalSort} of the result. We omit
+implementation of textbook graph algorithms \hs{transitiveClosure} and
+\hs{topologicalSort}~\cite{cormen2001introduction}.
 
-The greatest impedance mismatch between our framework and this implementation is the notion of \hs{Time} in \hs{getStoreTime}, which is missing from \hs{Build}.
-However, as we discussed in \S\ref{sec-dirty-bit}, \Make is really using modification-time to encode a dirty bit with specific properties.
-We model the dirty bit and time with the following API on \hs{M}:
+The \hs{chain} is processed in the \hs{State} monad, with each non-input key
+\hs{k} in the chain passed to the provided \hs{process} function, along with
+\hs{k}'s dependencies and the action \hs{act}, which when executed recomputes
+the \hs{k}'s value by fetching its dependencies from the store.
 
-To model a dirty bit we extend the \hs{i} information with \hs{Changed} to capture the \hs{k -> v} store at the end of the previous run, and extend \hs{M} to record which files have been written to during this \hs{runM}.
-A file is considered dirty if either the current value as per \hs{getStore} has changed since the last run, or if it has had \hs{putStore} called on it\footnote{It would be reasonable to declare files unchanged if they have \hs{putStore} called at their previous value, but \Excel does not and \Make cannot using modification times, so we ignore this option.}.
-To track the values at the last run we use \hs{withChanged} to copy the resultant \hs{k -> v} store into the info after the build completes.
-
-On top of a dirty bit we can encode a time with a very small domain, concretely if the file was last changed before the last build ended, or afterwards.
-That domain is sufficient for the \Make system.
-At first the fact that this domain is sufficient surprised us, until we realised \Make is only interested in the dirty bit anyway!
-If we strip back the modification time and directly implement \Make in terms of \hs{getChanged} we obtain:
-
-\begin{minted}{haskell}
-makeDirtyBit :: Eq v => Build Applicative (Changed k v, ()) k v
-makeDirtyBit = withChanged $ topological $ \k dk act -> do
-    dirty <- getChanged k ||^ anyM getChanged dk
-    when dirty (putStore k =<< act)
-\end{minted}
+Note that \hs{dependencies} is only defined for applicative tasks, which is what
+restricts \Make to static dependencies, as reflected in the
+type~\hs{Build}~\hs{Applicative}. Moreover, any other build system following
+the same \hs{topological} approach will also inherit the same restriction.
 

### 2018-03-16 paper/5-implementations.tex
Polish S5.1

@@
 (see~\S\ref{sec-concurrency} for parallel \Make). For each key, it builds it if
-it is older than any of its dependencies. We can express this in our framework
-as follows. We capture the persistent information stored by \Make as a pair
+it is older than any of its dependencies. To express \Make in our framework,
+we capture the persistent information it stores by a pair
 \hs{(@@modTime,}~\hs{now)} comprising the \emph{file modification time} function
-\hs{modTime}~\hs{::}~\hs{k}~\hs{->}~\hs{Time} and the current time \hs{now}.
+\hs{modTime}~\hs{::}~\hs{k}~\hs{->}~\hs{Time} and the \emph{current time} \hs{now}.
+Setting aside the explicit manipulation of file modification times, which in
+reality is taken care of by the file system, the \hs{make} build system defined
+below captures the essence of \Make in a clear and precise manner.
 
@@
 
-The helper function \hs{topological}, which we describe later, calls
+\noindent
+The helper function \hs{topological}, which we describe below, calls
 \hs{process} on every \hs{key} in the topological order, providing the list of
@@
 \hs{process} compares its modification time with those of its dependencies. If
-\hs{key} needs to be rebuilt, the associated action \hs{act} is executed and
-the obtained result is stored, along with an updated file modification
-timestamp. The \hs{make} build system, as described, captures the essence of
-\Make in a clear and precise manner.
+the \hs{key} needs to be rebuilt, the action \hs{act} is executed and the
+obtained result is stored, along with an updated file modification timestamp.
 

### 2018-03-16 src/Development/Build.hs
Polish excel build system

@@ -97,15 +97,17 @@ try task partialFetch = fmap (fmap toResult) . trackExceptions task partialFetch
     toResult (Left k       ) = MissingDependency k
     toResult (Right (v, ks)) = Result v ks
 
+type CalcChain k = [k]
+
 reordering :: forall i k v. Ord k
             => (k -> State (Store i k v) (Result k v) -> State (Store i k v) (Maybe (Result k v)))
-            -> Build Monad (i, [k]) k v
+            -> Build Monad (i, CalcChain k) k v
 reordering step task key = execState $ do
     chain    <- snd . getInfo <$> get
     newChain <- go Set.empty $ chain ++ [key | key `notElem` chain]
     modify $ \s -> putInfo s (fst (getInfo s), newChain)
   where
-    go :: Set k -> [k] -> State (Store (i, [k]) k v) [k]
+    go :: Set k -> CalcChain k -> State (Store (i, [k]) k v) (CalcChain k)
     go _    []     = return []
     go done (k:ks) = do
         case try task fetch k of
@@ -122,23 +124,25 @@ reordering step task key = execState $ do
         fetch k | k `Set.member` done = do s <- get; return (Just $ getValue s k)
                 | otherwise           = return Nothing
 
-type ExcelInfo k = ((k -> Bool, DependencyApproximation k), [k])
+type ExcelInfo k = ((k -> Bool, k -> DependencyApproximation k), CalcChain k)
 
 excel :: Ord k => Build Monad (ExcelInfo k) k v
-excel = reordering $ \key act -> do
-    (dirty, deps) <- getInfo <$> get
-    let rebuild = dirty key || case deps of SubsetOf ks -> any dirty ks
-                                            Unknown     -> True
-    if not rebuild
-        then return Nothing
-        else do
-            result <- act
-            case result of
-                MissingDependency _ -> return (Just result)
-                Result v _dynamicDependencies -> do
-                    let newDirty k = if k == key then True else dirty k
-                    modify $ \s -> putInfo (putValue s key v) (newDirty, deps)
-                    return (Just result)
+excel = reordering process
+  where
+    process key act = do
+        (dirty, deps) <- getInfo <$> get
+        let rebuild = dirty key || case deps key of SubsetOf ks -> any dirty ks
+                                                    Unknown     -> True
+        if not rebuild
+            then return Nothing
+            else do
+                result <- act
+                case result of
+                    MissingDependency _ -> return (Just result)
[…]

### 2018-03-16 paper/5-implementations.tex
Work on S5.1-S5.2

@@
 
-\Make processes keys in a linear order based on a topological sort
-(see~\S\ref{sec-concurrency} for parallel \Make). For each key, it builds it if
-it is older than any of its dependencies. To express \Make in our framework,
-we capture the persistent information it stores by a pair
+We provide the complete implementation of \Make in our framework in
+Fig.~\ref{fig-make-implementation}. \Make processes keys in a linear order based
+on the topological sort (see~\S\ref{sec-concurrency} for parallel \Make). For
+each key, it builds it if it is older than any of its dependencies. We capture
+the persistent build information that \Make stores by a pair
 \hs{(@@modTime,}~\hs{now)} comprising the \emph{file modification time} function
@@
 Setting aside the explicit manipulation of file modification times, which in
-reality is taken care of by the file system, the \hs{make} build system defined
-below captures the essence of \Make in a clear and precise manner.
+reality is taken care of by the file system, the function \hs{make} captures
+the essence of \Make in a clear and precise manner.
 
-\begin{minted}[xleftmargin=10pt]{haskell}
+\begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+-- Persistent build information
 type Time       = Integer
@@
 \vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
+\begin{minted}[fontsize=\small]{haskell}
+-- Make build system
 make :: Eq k => Build Applicative (MakeInfo k) k v
@@
 \end{minted}
-
-\noindent
-The helper function \hs{topological}, which we describe below, calls
-\hs{process} on every \hs{key} in the topological order, providing the list of
-its dependencies \hs{deps} and the action \hs{act} to compute the resulting
-value if it needs to be rebuilt. To determine if the \hs{key} is \hs{dirty},
-\hs{process} compares its modification time with those of its dependencies. If
-the \hs{key} needs to be rebuilt, the action \hs{act} is executed and the
-obtained result is stored, along with an updated file modification timestamp.
-
-The implementation of \hs{topological} encodes the dependency strategy that
-\Make has chosen to use. The function itself is defined as:
-
 \vspace{1mm}
-\begin{minted}{haskell}
-topological :: Eq@\,\blk{k}@ => (k -> [k] -> State@\,@(Store i k v)@\,\blk{v}@ -> State@\,@(Store i k v)@\,@(@@))
+\begin{minted}[fontsize=\small]{haskell}
+-- Standard graph algorithms (implementation omitted)
+transitiveClosure :: Eq a => (a -> [a]) ->  a  -> [a]
+topologicalSort   :: Eq a => (a -> [a]) -> [a] -> [a]
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Topological dependency strategy
+topological :: Eq k => (k -> [k] -> State (Store i k v) v -> State (Store i k v) (@@))
             -> Build Applicative i k v
@@
 \end{minted}
-\vspace{1mm}
+\vspace{-2mm}
+\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
+\vspace{-2mm}
+\end{figure}
 
-\noindent
-The \hs{where} clause corresponds to the pre-processing step, which uses the
-function \hs{dependencies}, defined in~\S\ref{sec-deps}, to extract static
-dependencies from the provided applicative \hs{task}. We compute the linear
-processing \hs{chain} by taking the \hs{transitiveClosure} of \hs{key}'s
-dependencies, and performing the \hs{topologicalSort} of the result. We omit
-implementation of textbook graph algorithms \hs{transitiveClosure} and
-\hs{topologicalSort}~\cite{cormen2001introduction}.
+The helper function \hs{topological} calls \hs{process} on every \hs{key} in
+the topological order, providing the list of its dependencies \hs{deps} and the
+action \hs{act} to compute the resulting value if it needs to be rebuilt. To
+determine if the \hs{key} is \hs{dirty}, \hs{process} compares its modification
+time with those of its dependencies. If the \hs{key} needs to be rebuilt, the
+action \hs{act} is executed and the obtained result is stored, along with an
+updated file modification timestamp.
+
+The implementation of \hs{topological} encodes the dependency strategy that
+\Make has chosen to use. The \hs{where} clause corresponds to the pre-processing
+stage, which uses the function \hs{dependencies}, defined in~\S\ref{sec-deps},
+to extract static dependencies from the provided applicative \hs{task}. We
+compute the linear processing \hs{chain} by taking the \hs{transitiveClosure}
+of \hs{key}'s dependencies, and performing the \hs{topologicalSort} of the
+result. We omit implementation of textbook graph algorithms
+\hs{transitiveClosure} and \hs{topologicalSort}, e.g.
+see~\cite{cormen2001introduction}.
 
@@
 
-\todo{AM}{I might do some simplification of this tomorrow. Neil: I guess it's best
-for you to focus on nailing traces and associated \Shake and \Bazel sections.}
+As a programming exercise, \Excel is the most challenging, due to the
+complicated reordering dependency strategy that requires retrying out-of-order
+operations, see Fig.~\ref{fig-excel-implementation}.
 
-As a programming exercise, \Excel is the most challenging, due to the complicated reordering dependency scheme, which requires retrying out-of-order operations. The complete implementation is:
+\Excel's persistently stored information is a triple: (i) the dirty bit
+function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
+\hs{k}~\hs{->}~\hs{DependencyApproximation}~\hs{k} that \Excel uses to handle
+dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
+previous build (\S\ref{sec-background-excel}).
 
-\begin{minted}{haskell}
-excel :: Eq v => Build Monad (Changed k v, [k]) k v
-excel = withChanged $ reordering $ \k dk _ act -> do
-    dirty <- getChanged k ||^ maybe (return True) (anyM getChanged) dk
-    if not dirty
+\begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+-- Approximation of task dependencies
+data DependencyApproximation k = SubsetOf [k] | Unknown
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Persistent build information
+type CalcChain k = [k]
+type ExcelInfo k = ((k -> Bool, k -> DependencyApproximation k), CalcChain k)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Result of speculative task execution
+data Result k v = MissingDependency k | Result v [k]
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Reordering dependency strategy (implementation omitted)
+reordering :: Ord k
+    => (k -> State (Store i k v) (Result k v) -> State (Store i k v) (Maybe (Result k v)))
+    -> Build Monad (i, CalcChain k) k v
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Excel build system
+excel :: Ord k => Build Monad (ExcelInfo k) k v
+excel = reordering process
+  where
+    process key act = do
+        (dirty, deps) <- getInfo <$> get
+        let rebuild = dirty key || case deps key of SubsetOf ks -> any dirty ks
+                                                    Unknown     -> True
+        if not rebuild
             then return Nothing
-       else act >>= \case
-              Left e -> return $ Just e
-              Right (_, v) -> do putStore k v
-                                 return Nothing
+            else do
+                result <- act
+                case result of
+                    MissingDependency _ -> return (Just result)
+                    Result v dynamicDependencies -> do
+                        let newDirty k = if k == key then True else dirty k
+                        modify $ \s -> putInfo (putValue s key v) (newDirty, deps)
+                        return (Just result)
 \end{minted}
+\vspace{-2mm}
+\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
+\vspace{-2mm}
+\end{figure}
 
-We use the helper function \hs{reordering} which provides the key to build (\hs{k}), the dependencies of \hs{k} (\hs{dk}) and an action to build \hs{k} (\hs{act}).
-The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Task} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway.
-If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possibilities:
+The helper function \hs{reordering}, whose implementation we omit since it is
+technical and not particularly enlightening, calls the function \hs{process} to
+\emph{try to build} a \hs{key} by executing the action \hs{act}, in the order
+determined by the calculation chain. To decide whether to \hs{rebuild} the
+\hs{key}, \hs{process} checks if the \hs{key} itself is marked dirty or the
+approximation of its dependencies contains a dirty key. Notice that if the
+dependencies of the \hs{key} are \hs{Unknown}, e.g. when it uses the
+\cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
+not needed, \hs{process} returns \hs{Nothing} to indicate this. Otherwise, it
+executes \hs{act} leading to two possible \hs{result}s:
 
-\begin{enumerate}
-\item The key cannot be built at this time because it depends on a key which has not yet been computed - and thus \hs{act} returns \hs{Left e}, where \hs{e} is the key that we are waiting for.
-In response we return \hs{e} as the result of step - asking \hs{reordering} to requeue this key after \hs{e}.
-\item The key builds successfully, so we store it and return \hs{Nothing} to indicate \hs{reordering} can move on to the next key.
-\end{enumerate}
+\begin{itemize}
+    \item \hs{MissingDependency}~\hs{k} indicates that the execution of \hs{act} has
+    failed, because one of its dependencies \hs{k} was out-of-date, i.e. the
+    calculation chain from the previous build was incorrect and therefore needs
+    to be reordered, deferring the \hs{key} to be rebuilt later.
+    \item \hs{Result}~\hs{v}~\hs{dynamicDependencies} indicates that the
+    execution has succeeded producing the value~\hs{v} and the list of the
+    \hs{key}'s dynamic dependencies. We store the value, and mark it dirty to
+    trigger the rebuilding of keys that depend on it.
+\end{itemize}
 
-The astute reader may notice that \hs{reordering} supplies extra information that is ignored with \hs{_} patterns.
-While not required for \Excel, we have implemented build systems using \hs{reordering} which use verifying and constructive traces, effectively turning \Excel into a cloud build system and ensuring \hs{reordering} is not overly fitted to \Excel alone (additional implementations available in the supplementary material).
-The definition of \hs{reordering} is:
-
-\begin{minted}{haskell}
-reordering
-    :: (m ~ M (i, [k]) k v, Default i)
-    => (k -> Maybe [k] -> (k -> m (Maybe v)) -> m (Either k ([k], v)) -> m (Maybe k))
-    -> Build Monad (i, [k]) k v
-\end{minted}
-
-Dynamic requires a function that takes takes 4 arguments (we use \hs{m} to elide \hs{M (i, [k]) k v} from all the signatures):
-
-\begin{enumerate}
-\item \hs{k} - the key that is being built.
-\item \hs{Maybe [k]} - the list of dependencies, or \hs{Nothing} if the step makes use of monadic dependencies.
-Note that above if a step is truly monadic it is always rebuilt.
-In the case of spreadsheets such steps are very rare, so a very conservative approximation can be used.\footnote{In truth, many spreadsheets take the alternative approximation of always assuming monadic steps \textit{did not} change - trading correctness for performance.}
-\item \hs{k -> m (Maybe v)} - a function that looks up a key - it returns \hs{Nothing} if that key has not yet been built (and thus there is a dependency violation).
-\item \hs{m (Either k ([k], v))} - an action to produce either the \hs{k} which is required but not yet computed (the reason you fail your dependency order check), or a list of the keys used and value produced.
-\item It returns \hs{m (Maybe k)}, with a \hs{Nothing} to indicate the the build succeeded, or \hs{Just k} to indicate that the given key \hs{k} has not been built, and thus this key should be moved after \hs{k}.
-\end{enumerate}
-
-The implementation of \hs{dynamic} is 15 lines, but is substantially complicated by the requirement to ensure newly discovered keys are added to the list - something an actual spreadsheet with a fixed universe of cells need not do.
+In both of the above cases, we notify the parent \hs{reordering} function about
+the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
+that \hs{process} ignores \hs{dynamicDependencies} available in the \hs{result}.
+While not required for \Excel, we have implemented build systems using
+\hs{reordering} which use verifying and constructive traces, effectively turning
+\Excel into a cloud build system and ensuring \hs{reordering} is not overly
+fitted to \Excel alone (additional implementations will be submitted to the
+Artefact Evaluation).
 

### 2018-03-16 paper/3-abstractions.tex
Tiny update to S3

@@
 argument, an application of \hs{busy} to a \hs{Task}~\hs{Functor} or
-a \hs{Task}~\hs{Applicative} will typecheck just fine. It feels a bit like
-sub-typing, but it is actually just ordinary higher-rank polymorphism at
+a \hs{Task}~\hs{Applicative} will typecheck and run just fine. It feels a bit like
+sub-typing, but is actually just ordinary higher-rank polymorphism at
 work~\cite{jones2007practical}.

### 2018-03-16 paper/5-implementations.tex
Tweak Excel implementation

@@
                 case result of
-                    MissingDependency _ -> return (Just result)
-                    Result v dynamicDependencies -> do
+                    MissingDependency _ -> return ()
+                    Result v _dynamicDependencies -> do
                         let newDirty k = if k == key then True else dirty k

### 2018-03-16 src/Development/Build.hs

@@ -138,11 +138,11 @@ excel = reordering process
             else do
                 result <- act
                 case result of
-                    MissingDependency _ -> return (Just result)
+                    MissingDependency _ -> return ()
                     Result v _dynamicDependencies -> do
                         let newDirty k = if k == key then True else dirty k
                         modify $ \s -> putInfo (putValue s key v) (newDirty, deps)
-                        return (Just result)
+                return (Just result)
 
 type MultiBuild c i k v = Task c k v -> [k] -> Store i k v -> Store i k v
 

### 2018-03-16 paper/1-intro.tex
Implement some Simon's changes

@@
 \begin{itemize}
-\item Build systems vary on many axes,
-including: static vs dynamic dependencies; cloud-build, including
-shallow vs deep; deterministic vs non-deterministic build rules;
-early cut-off; self-tracking build systems; and persistent metadata.
+\item Build systems vary on many axes, including:
+static vs dynamic dependencies; local vs cloud; deterministic vs
+non-deterministic build rules; support for early cutoff; self-tracking build
+systems; and the type of persistent build information.
 In \S\ref{sec-background} we identify some key properties, in the context of

### 2018-03-16 paper/3-abstractions.tex
Implement some Simon's changes

@@
 We  begin by establishing a common vocabulary for build systems.
-% that allows us to abstract away from specific application domains, such as
-% software development or spreadsheets.
-Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
-and gives types of the library functions we use in this paper.
-
-\vspace{1mm}
-A \emph{build system} takes a \emph{task description}, a target \emph{key},
-and a \emph{store}, and returns a new store in which the target key is
-up to date. The store is a key/value map, together with \emph{persistent build
-information} that is maintained by the build system itself.
 
 \vspace{-2mm}
-\paragraph{Keys and values.} In software build systems, keys are typically filenames,
-e.g.~\cmd{main.c}, and values are file contents (a C program source
-code in this case). In spreadsheets, keys are cell names,
-e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
-displayed inside cells. In our Haskell code, we will use type
-variables \hs{k} and \hs{v} to denote keys and values, respectively.
+\paragraph{Keys, values, and the store} The goal of any build system is to
+bring up to date a \emph{store} that implements a mapping from \emph{keys} to
+\emph{values}. In software build systems the store is the file system, the
+keys are filenames, and the values are file contents. In \Excel, the store is
+the worksheets, the keys are cell names (such as \cmd{A2}) and the values are
+numbers, strings etc, displayed as the cell contents. Many systems use
+\emph{hashes} of values as compact summaries with a fast equality check.
 
@@
 
-% \vspace{-2mm}
-% \paragraph{The store.}
-A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
-values. It is convenient to assume that a store is total, i.e. it contains a
-value for every possible key. We therefore also assume that the type of values
-is capable of encoding values corresponding to non-existent keys (missing files
-or empty cells). The store also holds a value of type \hs{i},
-representing the persistent build information maintained by the build
-system~--~its ``memory''.
-% Given an \hs{i} and a key/value map \store, we can \hs{initialise} a store.
+\vspace{-2mm}
+\paragraph{Persistent build information} As well as the key/value mapping, the store
+also contains information maintained by the build system itself, which persists
+from one invocation of the build system to the next -- its ``memory''.
+
+\paragraph{Task description} Any build system requires the user to specify how to
+compute the new value for one key, using the (up to date) values of its
+dependencies. We call this specification the \emph{task description}.  For
+example, in \Excel, the formulae of the spreadsheet constitute the task
+description; in \Make the rules in the makefile are the task description.
 
 \vspace{-2mm}
-\paragraph{Hashes.}
-We use a cryptographic \emph{hash function}
-\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash}~\hs{v}
-for efficient tracking and sharing of build results, where \hs{Hash}~\hs{v} is
-an abstract data type equipped with~the~equality test.
-In addition to usual \hs{getValue} and \hs{putValue} operations,
-some stores support a dedicated \hs{getHash} operation, which is implemented
-more efficiently than by hashing the result of \hs{getValue}\footnote{For
-example, Git Virtual File System~\cite{gvfs} stores file hashes and downloads
-file contents only on demand.}.
+\paragraph{Build system} A \emph{build system} takes a task description, a target
+key, and a store, and returns a new store in which the target key has an
+up to date value.
 
-% This, in particular, suggests that it is possible to \emph{depend on the
-% absence of a value}, which is a useful feature for build systems.
+\vspace{-2mm}
+\paragraph{Modelling in Haskell} We will model all our build systems concretely,
+as Haskell programs. To that end, Fig.~\ref{fig-types} gives the type
+declarations and function signatures of the library functions. For example,
+\hs{Store}~\hs{i}~\hs{k}~\hs{v} is the type of stores, with several associate
+functions (\hs{getValue}, etc.). We use \hs{k} as a type
+variable ranging over keys, \hs{v} for values, and \hs{i} for the persistent
+build information.
 

### 2018-03-16 paper/3-abstractions.tex
Redo the Shake code, flip some arguments

--- 2018-03-16 paper/3-abstractions.tex
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-track :: Task Monad k v -> Store i k v -> k -> Maybe (v, [k])
-track task store = runWriter . task (\k -> writer (getValue store k, [k]))
+track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
+track task fetch = fmap runWriter . task (\k -> writer (fetch k, [k]))
 \end{minted}
@@
 \noindent
+\todo{AM}{Type of track has changed - we no longer supply the Store, just an access fuction}
+\todo{AM}{trackM is essential for Shake, what do we do there?}
 The implementation uses the standard Haskell \hs{Writer} monad, which allows us

### 2018-03-16 paper/5-implementations.tex
Redo the Shake code, flip some arguments

@@
 
+\todo{AM}{Note that anything updating input files must also update Make's info.}
+\todo{AM}{Say that in reality that build rules mutate the state themselves.}
+
 \subsection{\Excel}\label{sec-implementation-excel}
@@
 
-\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
+The \Shake approach for dependency tracking involves recording traces and verifying them, for which we use the \hs{Trace} type defined in \ref{sec-verifying-traces}.
+Complete code is given in \S\ref{fig-shake-implementation}.
 
-The Shake approach for dependency tracking involves recording traces and verifying them. Using the \hs{Trace} type defined in \ref{sec-verifying-traces} we can implement:
+\begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+-- Determine whether a trace is relevant to the current state
+traceMatch :: (Monad m, Eq k) => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
+traceMatch check key ts = mapMaybeM f ts
+    where f (Trace k dkv v) = do
+                b <- return (key == k) &&^ allM (uncurry check) dkv
+                return $ if b then Just v else Nothing
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Recursive dependency strategy
+recursive
+    :: Eq k => (k -> (k -> State (Store i k v, [k]) v) -> State (Store i k v, [k]) (v, [k]) -> State (Store i k v, [k]) ())
+    -> Build Monad i k v
+recursive step task key store = fst $ execState (ensure key) (store, [])
+    where
+        ensure key = do
+            let fetch k = do ensure k; gets (getValue k . fst)
+            s <- get
+            when (key `notElem` snd s) $ do
+                modify $ \(s, done) -> (s, key:done)
+                case trackM task fetch key of
+                    Nothing -> return ()
+                    Just act -> step key fetch act
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Shake build system
+shake :: (Eq k, Hashable v) => Build Monad [Trace k v] k v
+shake = recursive $ \key fetch act -> do
+    traces <- gets (getInfo . fst)
+    poss <- traceMatch (\k v -> (==) v . hash <$> fetch k) key traces
+    current <- gets (getHash key . fst)
+    when (current `notElem` poss) $ do
+        (v, ds) <- act
+        modify $ \(s, done) ->
+            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
+            in (putInfo (t : getInfo s) (putValue key v s), done)
 
-\begin{minted}{haskell}
-shake :: Hashable v => Build Monad [Trace k v] k v
-shake = recursive $ \k _ fetch act -> do
-    poss <- traceMatch (\k v -> (==) v . getHash <$> fetch k) k =<< getInfo
-    h <- getStoreHash k
-    when (h `notElem` poss) $ do
-        (dk, v) <- act
-        putStore k v
-        dh <- mapM getStoreHash dk
-        modifyInfo (Trace k (zip dk dh) (getHash v) :)
 \end{minted}
+\vspace{-2mm}
+\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
+\vspace{-2mm}
+\end{figure}
 
@@
 
-\begin{minted}{haskell}
-traceMatch :: (Monad m, Eq k) => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
-traceMatch check k ts = mapMaybeM f ts
-    where f (Trace k2 dkv v) = do
-                b <- return (k == k2) &&^ allM (uncurry check) dkv
-                return $ if b then Just v else Nothing
-\end{minted}
-
 The function \hs{traceMatch} requires a way to check a dependency, and the key, and the traces.
@@
 We define the recursive combinator as:
-
-\begin{minted}{haskell}
-recursive
-    :: (m ~ M i k v, Default i)
-    => (k -> Maybe [k] -> (k -> m v) -> m ([k], v) -> m ())
-    -> Build Monad i k v
-recursive step compute = runM . ensure
-    where
-        ensure k = do
-            let fetch x = ensure x >> getStore x
-            done <- getTemp
-            when (k `Set.notMember` done) $ do
-                modifyTemp $ Set.insert k set
-                case track compute fetch k of
-                    Nothing -> return ()
-                    Just act -> step k (getDependenciesMaybe compute k) fetch act
-\end{minted}
 
@@
 As an aside we note that the optimal cloud structure for applicative and monadic constructive systems has a resemblance to the free dependency structures described in \S\ref{sec-free-build}.
+
+\subsection{Other build systems}
+
+We have implemented the complete product, reusing the \hs{topological} functions etc.

### 2018-03-16 src/Development/Build.hs

@@ -2,10 +2,10 @@
 {-# LANGUAGE ScopedTypeVariables, TupleSections #-}
 module Development.Build (
     -- * Build
-    Build, dumb, busy, memo, make, excel,
+    Build, dumb, busy, memo, make, excel, shake,
 
     -- * Algorithms
-    topological, reordering,
+    topological, reordering, recursive,
 
     -- * MultiBuild
     MultiBuild, sequentialMultiBuild,
@@ -16,6 +16,7 @@ module Development.Build (
 
 import Control.Monad.State
 import Data.Set (Set)
+import Control.Monad.Extra
 
 import Development.Build.Task
 import Development.Build.Task.Applicative hiding (exceptional)
@@ -33,37 +34,36 @@ import qualified Data.Set as Set
 type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
 
 dumb :: Eq k => Build Monad i k v
-dumb task key store = case compute task (getValue store) key of
+dumb task key store = case compute task (\k -> getValue k store) key of
     Nothing    -> store
-    Just value -> putValue store key value
+    Just value -> putValue key value store
 
 busy :: forall k v. Eq k => Build Monad () k v
 busy task key store = execState (fetch key) store
   where
     fetch :: k -> State (Store () k v) v
     fetch k = case task fetch k of
-        Nothing  -> do s <- get; return (getValue s k)
-        Just act -> do v <- act; modify (\s -> putValue s k v); return v
+        Nothing  -> gets (getValue k)
+        Just act -> do v <- act; modify (putValue k v); return v
 
 memo :: forall k v. Eq k => Build Monad () k v
 memo task key store = fst $ execState (fetch key) (store, [])
   where
     fetch :: k -> State (Store () k v, [k]) v
     fetch k = case task fetch k of
-        Nothing  -> do { s <- fst <$> get; return (getValue s k) }
+        Nothing  -> gets (getValue k . fst)
         Just act -> do
             built <- snd <$> get
             when (k `notElem` built) $ do
                 v <- act
-                modify $ \(s, built) -> (putValue s k v, k : built)
-            s <- fst <$> get
-            return (getValue s k)
+                modify $ \(s, built) -> (putValue k v s, k : built)
+            gets (getValue k . fst)
 
 topological :: Eq k
             => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
             -> Build Applicative i k v
[…]

### 2018-03-16 src/Development/Build/Store.hs

@@ -27,16 +27,22 @@ instance Hashable Int where
 instance Hashable Integer where
     hash = Hash . fromIntegral
 
-data Store i k v = Store { getInfo :: i, getValue :: k -> v }
+data Store i k v = Store { info :: i, values :: k -> v }
 
-getHash :: Hashable v => Store i k v -> k -> Hash v
-getHash s = hash . getValue s
+getInfo :: Store i k v -> i
+getInfo = info
 
-putInfo :: Store i k v -> i -> Store i k v
-putInfo s i = s { getInfo = i }
+getValue :: k -> Store i k v -> v
+getValue = flip values
 
-putValue :: Eq k => Store i k v -> k -> v -> Store i k v
-putValue s k v = s { getValue = \key -> if key == k then v else getValue s key }
+getHash :: Hashable v => k -> Store i k v -> Hash v
+getHash k = hash . getValue k
+
+putInfo :: i -> Store i k v -> Store i k v
+putInfo i s = s { info = i }
+
+putValue :: Eq k => k -> v -> Store i k v -> Store i k v
+putValue k v s = s { values = \key -> if key == k then v else values s key }
 
 initialise :: i -> (k -> v) -> Store i k v
 initialise = Store
@@ -45,12 +51,12 @@ mapInfo :: (i -> j) -> Store i k v -> Store j k v
 mapInfo f (Store i kv) = Store (f i) kv
 
 checkHashes :: Hashable v => Store m k v -> [(k, Hash v)] -> Bool
-checkHashes store = all (\(k, h) -> getHash store k == h)
+checkHashes store = all (\(k, h) -> getHash k store == h)
 
 agree :: Eq v => [Store i k v] -> [k] -> Bool
 agree ss = all same
   where
-    same k = let vs = [getValue s k | s <- ss] in and $ zipWith (==) vs (drop 1 vs)
+    same k = let vs = [getValue k s | s <- ss] in and $ zipWith (==) vs (drop 1 vs)
 
 
 -- | A key-value store monad that in addition to usual 'getValue' and 'putValue'

### 2018-03-16 src/Development/Build/Task/Monad.hs

@@ -25,8 +25,8 @@ dependencies task store = execWriterT . sequenceA . task fetch
   where
     fetch k = tell [k] >> lift (store k)
 
-track :: Task Monad k v -> Store i k v -> k -> Maybe (v, [k])
-track task store = fmap runWriter . task (\k -> writer (getValue store k, [k]))
+track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
+track task fetch = fmap runWriter . task (\k -> writer (fetch k, [k]))
 
 trackM :: Monad m => Task Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
 trackM task store = fmap runWriterT . task fetch
@@ -46,15 +46,15 @@ isInput task = isNothing . task (const Proxy)
 inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
 inputs task store key = filter (isInput task) (closure deps key)
   where
-    deps k = maybe [] snd (track task store k)
+    deps k = maybe [] snd (track task (\k -> getValue k store) k)
 
 -- | Check that a task is /consistent/ with a pure lookup function @f@, i.e.
 -- if it returns @Just v@ for some key @k@ then @f k == v@.
 consistent :: Eq v => Task Monad k v -> Store i k v -> Bool
 consistent task store =
-    forall $ \k -> case compute task (getValue store) k of
+    forall $ \k -> case compute task (flip getValue store) k of
         Nothing -> True
-        Just v  -> getValue store k == v
+        Just v  -> getValue k store == v
 
 -- | Given a @task@, a pair of key-value maps describing the contents of a
 -- store @store@ and @result@ a build system was executed to build a given @key@,
@@ -77,7 +77,7 @@ compute task fetch = fmap runIdentity . task (pure . fetch)
 
 -- The version used in the paper
 compute' :: Task Monad k v -> Store i k v -> k -> Maybe v
-compute' task store key = runIdentity <$> task (Identity . getValue store) key
+compute' task store key = runIdentity <$> task (Identity . flip getValue store) key
 
 -- | Run a task with a partial lookup function. The result @Left k@ indicates
 -- that the task failed due to a missing dependency @k@. Otherwise, the

### 2018-03-16 paper/2-background.tex

@@
 We summarise differences between four discussed build systems in
-Table~\ref{tab-summary}. The `Metadata' column refers to the \emph{auxiliary
-build information} persistently stored between builds by all build systems:
+Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
+to the information that build systems persistently store between builds:
 \begin{itemize}
@@
 \hline
-$\!$Build system$\!$& Metadata                       & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
+$\!$Build system$\!$& Persistent build information   & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
 $\!$\Make       $\!$& File modification times        & Topological  & Static       & Yes     & No     & No   $\!$\\

### 2018-03-16 paper/4-build.tex
Work on the implementation

@@
     { traces :: [Trace k v]
-    , cache  :: Map (Hash v) v }
+    , contents  :: Map (Hash v) v }
 \end{minted}
@@
 We have a list of traces, plus a \hs{Map} from the hash to the actual contents.
-Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \emph{without} recomputing it.
-We split the traces and caches as in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only at most one complete file.
+Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{contents} \emph{without} recomputing it.
+We split the traces and contents as in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only at most one complete file.
 

### 2018-03-16 paper/5-implementations.tex
Work on the implementation

@@
 
-The helper function \hs{reordering}, whose implementation we omit since it is
+The helper function \hs{reordering}, whose 21-line implementation we omit since it is
 technical and not particularly enlightening, calls the function \hs{process} to
@@
 
-The \Shake approach for dependency tracking involves recording traces and verifying them, for which we use the \hs{Trace} type defined in \ref{sec-verifying-traces}.
-Complete code is given in \S\ref{fig-shake-implementation}.
-
 \begin{figure}
@@
     -> Build Monad i k v
-recursive step task key store = fst $ execState (ensure key) (store, [])
+recursive process task key store = fst $ execState (ensure key) (store, [])
     where
@@
             let fetch k = do ensure k; gets (getValue k . fst)
-            s <- get
-            when (key `notElem` snd s) $ do
+            done <- gets snd
+            when (key `notElem` done) $ do
                 modify $ \(s, done) -> (s, key:done)
@@
                     Nothing -> return ()
-                    Just act -> step key fetch act
+                    Just act -> process key fetch act
 \end{minted}
@@
 
-This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to fetch the result of a dependency (\hs{fetch}), and a way to build \hs{k} (\hs{act}).
-The code first looks up the traces that match the current situation, creating a list of possible results as \hs{poss}, using the helper \hs{traceMatch}.
-We check if the current value for the key is one of the possible results, and it not, build the key and store it in the trace.
-We can implement \hs{traceMatch}, which is used in all build systems working with \hs{Trace}, as:
+The \Shake approach for dependency tracking involves recording traces and verifying them, for which we use the \hs{Trace} type defined in \ref{sec-verifying-traces}.
+Complete code is given in \S\ref{fig-shake-implementation}, split into three functions:
 
-The function \hs{traceMatch} requires a way to check a dependency, and the key, and the traces.
-We return the \hs{result} field from all traces where the key and dependencies match.
-We define the recursive combinator as:
-
-The required \hs{step} function gets given the \hs{k} to build, the dependencies as \hs{Maybe [k]} (where they are \hs{Nothing} if the rule is truly monadic), a function to demand the result of a dependent key, and a function to compute the current value and return the actual dependencies.
-To ensure each file is only visited once in a single execution a piece of temporary state is used with \hs{getTemp} and \hs{modifyTemp} to record which files were visited - this state is not persisted to the \hs{i} info parameter and is reset on each fresh execution of \hs{M}.
+\begin{description}
+\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the \hs{key} you are currently building, and a function \hs{check} which checks a specified dependency.
+It returns the \hs{result} field of all traces that match.
+Since \hs{check} is in an arbitrary monad, the function has to use \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&} functions\footnote{These functions are all available in the \cmd{extra} library on Hackage.}.
+\item[\hs{recursive}] defines the dependency ordering pattern.
+It requires a \hs{process} function that builds a key given a way to build a dependency, and a way to build the result.
+The main purpose of \hs{recursive} is to ensure that in a single run any key is not built twice -- for which it extends the \hs{State} monad with a list of \hs{done} keys.
+\item[\hs{shake}] ties everything together.
+First it tests if the traces allow the current state of the target key.
+If not, it builds the key and updates the store.
+The only subtlty is that Shake calls \hs{fetch} on the keys while checking them -- building the last-recorded dependencies before checking them.
+\end{description}
 
@@
 
-\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
+\begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+bazel :: (Eq k, Hashable v) => Build Applicative (Traces k v) k v
+bazel = topological $ \key ds act -> do
+    s <- get
+    let Traces traces contents = getInfo s
+    poss <- traceMatch (\k v -> return $ getHash k s == v) key traces
+    if null poss then do
+        v <- act
+        modify $ \s ->
+            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
+                ts = Traces (t : traces) (Map.insert (hash v) v contents)
+            in putInfo ts (putValue key v s)
+    else do
+        when (getHash key s `notElem` poss) $
+            modify $ putValue key (contents Map.! head poss)
+\end{minted}
+\vspace{-2mm}
+\caption{An implementation of \Bazel using our framework.}\label{fig-bazel-implementation}
+\vspace{-2mm}
+\end{figure}
 
-Now we have seen the three dependency schemes, we can directly reuse \hs{topological} to define \Bazel.
+Now we have seen all three dependency schemes, we can reuse , we can directly reuse \hs{topological} to define \Bazel.
 Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type.
 We define \Bazel as:
-
-\begin{minted}{haskell}
-bazel :: Hashable v => Build Applicative (Traces k v) k v
-bazel = topological $ \k dk act -> do
-    poss <- traceMatch (\k v -> (== v) <$> getStoreHash k) k . traces =<< getInfo
-    if null poss then do
-        v <- act
-        dh <- mapM getStoreHash dk
-        modifyInfo $ \i -> i
-            {traces = Trace k (zip dk dh) (getHash v) : traces i
-            ,contents = Map.insert (getHash v) v $ contents i}
-        putStore k v
-    else do
-        h <- getStoreHash k
-        when (h `notElem` poss) $
-            putStore k . (Map.! head poss) . contents =<< getInfo
-\end{minted}
 

### 2018-03-16 src/Development/Build.hs

@@ -16,6 +16,7 @@ module Development.Build (
 
 import Control.Monad.State
 import Data.Set (Set)
+import Data.Map (Map)
 import Control.Monad.Extra
 
 import Development.Build.Task
@@ -25,6 +26,7 @@ import Development.Build.Store
 import Development.Build.Utilities
 
 import qualified Data.Set as Set
+import qualified Data.Map as Map
 
 -- | A build system takes a 'Task', a key to build, some information from
 -- the previous build @i@ (which can be missing if this is the first build),
@@ -190,16 +192,16 @@ traceMatch check key ts = mapMaybeM f ts
 recursive
     :: Eq k => (k -> (k -> State (Store i k v, [k]) v) -> State (Store i k v, [k]) (v, [k]) -> State (Store i k v, [k]) ())
     -> Build Monad i k v
-recursive step task key store = fst $ execState (ensure key) (store, [])
+recursive process task key store = fst $ execState (ensure key) (store, [])
     where
         ensure key = do
             let fetch k = do ensure k; gets (getValue k . fst)
-            s <- get
-            when (key `notElem` snd s) $ do
+            done <- gets snd
+            when (key `notElem` done) $ do
                 modify $ \(s, done) -> (s, key:done)
                 case trackM task fetch key of
                     Nothing -> return ()
-                    Just act -> step key fetch act
+                    Just act -> process key fetch act
 
 -- Shake build system
 shake :: (Eq k, Hashable v) => Build Monad [Trace k v] k v
@@ -212,3 +214,22 @@ shake = recursive $ \key fetch act -> do
         modify $ \(s, done) ->
             let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
             in (putInfo (t : getInfo s) (putValue key v s), done)
+
+data Traces k v = Traces
+    { traces :: [Trace k v]
+    , contents  :: Map (Hash v) v }
+
+bazel :: (Eq k, Hashable v) => Build Applicative (Traces k v) k v
+bazel = topological $ \key ds act -> do
+    s <- get
+    let Traces traces contents = getInfo s
+    poss <- traceMatch (\k v -> return $ getHash k s == v) key traces
+    if null poss then do
+        v <- act
+        modify $ \s ->
+            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
+                ts = Traces (t : traces) (Map.insert (hash v) v contents)
+            in putInfo ts (putValue key v s)
+    else do
[…]

### 2018-03-16 src/Development/Build/Store.hs

@@ -14,7 +14,7 @@ module Development.Build.Store (
 
 -- | A 'Hash' is used for efficient tracking and sharing of build results. We
 -- use @newtype Hash a = Hash Int@ for prototyping.
-newtype Hash a = Hash Int deriving Eq
+newtype Hash a = Hash Int deriving (Eq,Ord)
 
 class Hashable a where
     -- | Compute the hash of a given value. We typically assume cryptographic

### 2018-03-16 src/Development/Build.hs
Simplify trackM, fix errors

@@ -1,5 +1,6 @@
 {-# LANGUAGE ConstraintKinds, FlexibleContexts, RankNTypes #-}
 {-# LANGUAGE ScopedTypeVariables, TupleSections #-}
+{-# OPTIONS_GHC -Wno-unused-top-binds #-}
 module Development.Build (
     -- * Build
     Build, dumb, busy, memo, make, excel, shake,

### 2018-03-16 src/Development/Build/Task/Monad.hs

@@ -29,9 +29,9 @@ track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
 track task fetch = fmap runWriter . task (\k -> writer (fetch k, [k]))
 
 trackM :: Monad m => Task Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
-trackM task store = fmap runWriterT . task fetch
+trackM task fetch key = runWriterT <$> task trackingFetch key
   where
-    fetch k = tell [k] >> lift (store k)
+    trackingFetch k = tell [k] >> lift (fetch k)
 
 transitiveDependencies :: (Eq k, Monad m) => Task Monad k v
                                           -> (k -> m v) -> k -> m (Maybe [k])

### 2018-03-16 test/Main.hs

@@ -52,7 +52,7 @@ memoResult = sequentialMultiBuild memo task outputs inputs
 
 printOutputs :: Store () Cell Int -> IO ()
 printOutputs store = forM_ outputs $
-    \key -> putStrLn (show (name key) ++ " = " ++ show (getValue store key))
+    \key -> putStrLn (show (name key) ++ " = " ++ show (getValue key store))
 
 main :: IO ()
 main = do

### 2018-03-16 paper/5-implementations.tex
More Bazel stuff

@@
                 modify $ \(s, done) -> (s, key:done)
-                case trackM task fetch key of
+                case track task fetch key of
                     Nothing -> return ()
@@
 
-Now we have seen all three dependency schemes, we can reuse , we can directly reuse \hs{topological} to define \Bazel.
-Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type.
-We define \Bazel as:
+Now we have seen all three dependency schemes, we can directly reuse \hs{topological} to define \Bazel.
+Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, along with the \hs{Traces} type from \S\ref{sec-constructive-traces}.
+With these pieces in place, the implementation of \Bazel is given in Figure~\ref{fig-bazel-implementation}.
+We first figure out the possible results given the current state.
+If there are no recorded traces for the current dependencies we run it and record the results, otherwise grab a result from the \hs{contents} cache.
+
+The program presented above captures certain aspects of \Bazel, but the real implementation makes one important additional assumption on \hs{Task} -- namely that it is \textit{deterministic}.
+With that assumption \Bazel is able to create the result hash in a trace by hashing together the result hashes of the dependencies and the key -- as the mapping between dependencies and results is deterministic.
+That change can be reflected in the code above by storing an additional \hs{Map k (Hash v)} which records the single most-recent
+
+ rather than adding \hs{getHash key s} to the trace, \Bazel adds a hash of the dependencies hashes
+
+more accurately describes \CloudBuild \cite{cloudbuild} - both \Bazel and the closely related \Buck make the assumption that builds are \textit{deterministic}, and thus that
 

### 2018-03-16 paper/3-abstractions.tex
Simplify correctness definition

@@
 getInfo    :: Store i k v -> i
-putInfo    :: Store i k v -> i -> Store i k v
-getValue   :: Store i k v -> k -> v
-putValue   :: Eq k => Store i k v -> k -> v -> Store i k v
-getHash    :: Hashable v => Store i k v -> k -> Hash v
+putInfo    :: i -> Store i k v -> Store i k v
+getValue   :: k -> Store i k v -> v
+putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
+getHash    :: Hashable v => k -> Store i k v -> Hash v
 initialise :: i -> (k -> v) -> Store i k v
@@
 get       :: State s s
-put       :: s -> State s ()
+gets      :: (s -> a) -> State s a
 modify    :: (s -> s) -> State s ()
@@
   where
-    fetch :: k -> State (Store i k v) v
+    fetch :: k -> State (Store () k v) v
     fetch k = case task fetch k of
-        Nothing  -> do s <- get; return (getValue s k)
-        Just act -> do v <- act; modify (\s -> putValue s k v); return v
+        Nothing  -> gets (getValue k)
+        Just act -> do v <- act; modify (putValue k v); return v
 \end{minted}
@@
 that is seldom stated formally. Our intuition is this: \emph{when the build
-system completes, the specified key, and all its dependencies, should be up to
+system completes, the target key, and all its dependencies, should be up to
 date}. What does ``up to date'' mean? It means that if we recompute the value of
@@
 any dependencies}:
-\begin{minted}[xleftmargin=10pt]{haskell}
+\begin{minted}[xleftmargin=5pt]{haskell}
 compute :: Task Monad k v -> Store i k v -> k -> Maybe v
-compute task store key = runIdentity <$> task (Identity . getValue store) key
+compute task store = fmap runIdentity@\,@.@\,\blk{task}@ (\k -> Identity (getValue k store))
 \end{minted}
@@
 we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
-The use of \hs{Identity} just fixes the `impedance mismatch' between the pure
-function \hs{getValue}~\hs{store}, whose type is \store, and the \hs{fetch}
-argument of the \hs{task}, whose type must be \hs{k}~\hs{->}~\hs{f}~\hs{v} for
-some \hs{f}. To fix the mismatch, we wrap the result of the pure function in the
-\hs{Identity} monad: the function \hs{Identity}~\hs{.}~\hs{getValue}~\hs{store}
+The use of \hs{Identity} just fixes the `impedance mismatch' between the
+function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
+argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
+To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
+monad: the function
+\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)}
 has the type \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a
@@
 
-\vspace{-1mm}
-\noindent\rule{\textwidth}{0.4pt}
-\vspace{-7mm}
+\vspace{1mm}
 \definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
-build task description, \hs{key} is an output key, \hs{store} is an initial
+build task description, \hs{key} is a target key, \hs{store} is an initial
 store, and \hs{result} is the store produced by running the build system with
@@
 \noindent
-Then the \hs{result} is~\emph{correct} if there exists an \hs{ideal} store, such
-that:
+The keys that belong to the transitive closure of \hs{key}'s dependencies fall
+into two classes: input keys and non-input keys, which we will denote by $I$ and
+$O$, respectively. Note that \hs{key} may be in either of these sets, although
+the case when \hs{key} is an input is degenerate: we have $I=\{$\hs{key}$\}$ and
+$O=\emptyset$.
+
+The build \hs{result} is~\emph{correct} if the following two conditions hold:
 
 \begin{itemize}
-    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
-    for all possible non-input keys \hs{k}, the result of recomputing the
-    \hs{task} matches the value in the \hs{ideal} store:
-    \vspace{-1mm}
+    \item \hs{store} and \hs{result} \emph{agree on inputs}, that is, for all
+    input keys \hs{k}~$\in$~$I$:
     \[
-    \hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}.
+    \hs{getValue}~\hs{k}~\hs{store}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}.
     \]
-    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
-    \vspace{-1mm}
+    In other words, no inputs were corrupted during the
+    build\footnote{\todo{SLPJ}{Add the joke that GHC used to delete inputs?}}.
+
+    \item The \hs{result} is \emph{consistent} with the \hs{task}, i.e.
+    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
+    \hs{task} matches the value stored in the \hs{result}:
     \[
-    \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
+    \hs{Just}~\hs{(}\hs{getValue}~\hs{result}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}.
     \]
-    In other words, we require that the output \hs{key} has the correct value,
-    but do not impose any restrictions on intermediate keys, therefore
-    permitting shallow cloud builds.
-    \item The \hs{store}, \hs{result} and \hs{ideal} agree on all input keys
-    \hs{k} that belong to the transitive closure of \hs{key}'s dependencies:
-    \[
-    \,\,\,\,\,\,\,\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{&&}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}
-    \]
-    This requirement asserts that (i) no inputs were corrupted during the
-    build, and (ii)~the \hs{ideal} store has constraints not only on the output,
-    but also on the inputs.
 \end{itemize}
@@
 \label{def-correct}
-\vspace{-2mm}
-\rule{\textwidth}{0.4pt}
+\vspace{3mm}
 
@@
 % \hs{Task}~\hs{Monad} and treating them specially.
+
+%------------------------- Old correctness definition -------------------------%
+
+% \definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
+% build task description, \hs{key} is an output key, \hs{store} is an initial
+% store, and \hs{result} is the store produced by running the build system with
+% parameters \hs{task}, \hs{key} and \hs{store}. Or, using the precise language of
+% our abstractions:
+% \vspace{1mm}
+% \begin{minted}[xleftmargin=10pt]{haskell}
+% build         :: Build c i k v
+% task          :: Task c k v
+% key           :: k
+% store, @@result :: Store i k v
+% result = @@build @@task @@key @@store
+% \end{minted}
+% \vspace{1mm}
+% \noindent
+% Then the \hs{result} is~\emph{correct} if there exists an \hs{ideal} store, such
+% that:
+% \begin{itemize}
+%     \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
+%     for all possible non-input keys \hs{k}, the result of recomputing the
+%     \hs{task} matches the value in the \hs{ideal} store:
+%     \vspace{-1mm}
+%     \[
+%     \hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}.
+%     \]
+%     \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
+%     \vspace{-1mm}
+%     \[
+%     \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
+%     \]
+%     In other words, we require that the output \hs{key} has the correct value,
+%     but do not impose any restrictions on intermediate keys, therefore
+%     permitting shallow cloud builds.
+%     \item The \hs{store}, \hs{result} and \hs{ideal} agree on all input keys
+%     \hs{k} that belong to the transitive closure of \hs{key}'s dependencies:
+%     \[
+%     \,\,\,\,\,\,\,\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{&&}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}
+%     \]
+%     This requirement asserts that (i) no inputs were corrupted during the
+%     build, and (ii)~the \hs{ideal} store has constraints not only on the output,
+%     but also on the inputs.
+% \end{itemize}
+% A build system is \emph{correct} if it produces a correct \hs{result} for any
+% given \hs{task}, \hs{key} and \hs{store}.
+% }
+% \label{def-correct}

### 2018-03-16 paper/5-implementations.tex

@@
 The only subtlty is that Shake calls \hs{fetch} on the keys while checking them -- building the last-recorded dependencies before checking them.
+One minor annoyance is that the \hs{State} has been extended and thus needs to be projected using \hs{fst} before it is used.
 \end{description}
@@
 We first figure out the possible results given the current state.
-If there are no recorded traces for the current dependencies we run it and record the results, otherwise grab a result from the \hs{contents} cache.
+If there are no recorded traces for the current dependencies we run it and record the results, otherwise grab a suitable result from the \hs{contents} cache.
 
@@
 With that assumption \Bazel is able to create the result hash in a trace by hashing together the result hashes of the dependencies and the key -- as the mapping between dependencies and results is deterministic.
-That change can be reflected in the code above by storing an additional \hs{Map k (Hash v)} which records the single most-recent
+As a consequence \Bazel can compute the results of traces locally, without looking at \hs{Traces} (potentially saving a roundtrip to the server).
+To model this change in the code would require storing an additional transient piece of information \hs{done} of type \hs{Map k (Hash v)}, storing the computed hashes so far this run.
+With that available, \hs{getHash key s} would become:
 
- rather than adding \hs{getHash key s} to the trace, \Bazel adds a hash of the dependencies hashes
+\begin{minted}{haskell}
+hash (key, [ done Map.! d | d <- ds])
+\end{minted}
 
-more accurately describes \CloudBuild \cite{cloudbuild} - both \Bazel and the closely related \Buck make the assumption that builds are \textit{deterministic}, and thus that
-
-We execute keys in a topological order.
-For each key we first get a list of the traces that match.
-If there are none we rebuild and record the trace, much like \Shake, but also recording the value of the result.
-If there are some matching traces, and the current value of \hs{k} is not compatible, we grab the first possibility and grab its value from \hs{contents}.
+And that new hash would have to be stored in \hs{done}.
 
 \subsection{Cloud \Shake}\label{sec-implementation-cloud-shake}
-
-\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
 
@@
 To us, the most interesting build system as yet unavailable would matching recursive ordering with constructive traces - providing a cloud-capable build system with minimality, cut-off and monadic dependencies.
-Using our framework it is possible to define and test such a system:
+Using our framework it is possible to define and test such a system as per Figure~\ref{fig-cloudshake-implementation}.
 
-\begin{minted}{haskell}
-cloudShake :: Hashable v => Build Monad (Traces k v) k v
-cloudShake = recursive $ \k _ fetch act -> do
-    poss <- traceMatch (\k v -> (== v) . getHash <$> fetch k) k . traces =<< getInfo
+\begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+cloudShake :: (Eq k, Hashable v) => Build Monad (Traces k v) k v
+cloudShake = recursive $ \key fetch act -> do
+    s <- gets fst
+    let Traces traces contents = getInfo s
+    poss <- traceMatch (\k v -> (==) v . hash <$> fetch k) key traces
     if null poss then do
-        (dk, v) <- act
-        dh <- mapM getStoreHash dk
-        modifyInfo $ \i -> i
-            {traces = Trace k (zip dk dh) (getHash v) : traces i
-            ,contents = Map.insert (getHash v) v $ contents i}
-        putStore k v
+        (v, ds) <- act
+        modify $ \(s,done) ->
+            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
+                ts = Traces (t : traces) (Map.insert (hash v) v contents)
+            in (putInfo ts (putValue key v s), done)
     else do
-        now <- getStoreHash k
-        when (now `notElem` poss) $
-            putStore k . (Map.! head poss) . contents =<< getInfo
+        s <- gets fst
+        when (getHash key s `notElem` poss) $
+            modify $ \(s, done) -> (putValue key (contents Map.! head poss) s, done)
 \end{minted}
+\vspace{-2mm}
+\caption{An implementation of Cloud \Shake using our framework.}\label{fig-cloudshake-implementation}
+\vspace{-2mm}
+\end{figure}
 
-The differences from \hs{bazel} are minor - the dependency scheme has changed from \hs{topological} to \hs{recursive}, and thus the dependency keys \hs{dk} are captured from the action rather than in advance, and the checking calls \hs{fetch} to get the result instead of accessing the store directly.
-We believe this accurately captures and can be used to reason about how to make \Shake work with a cloud-based cache -- a challenge we intend to attempt in the future.
+The differences from \hs{bazel} are minor - the dependency scheme has changed from \hs{topological} to \hs{recursive}, and thus the dependency keys \hs{dk} are captured from the action rather than in advance, the transient state has gained a list of keys, and the checking calls \hs{fetch} to get the result instead of accessing the store directly.
 
 \subsection{Smarter \hs{[Trace]} Structures}\label{sec-smart-traces}
-\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
 
-In the examples above, we have used \hs{[Trace]} to capture a list of traces stored in the info field, which accurately models the data and operations.
-However, using a list necessarily means that finding the right trace takes $O(n)$.
+In the examples above, we have used \hs{[Trace k v]} to capture a list of traces -- however, using a list necessarily means that finding the right trace takes $O(n)$.
 For each of the \hs{Trace} based systems it is possible to devise a smarter representation, which we sketch below.
@@
 \end{enumerate}
-
-As an aside we note that the optimal cloud structure for applicative and monadic constructive systems has a resemblance to the free dependency structures described in \S\ref{sec-free-build}.
-
-\subsection{Other build systems}
-
-We have implemented the complete product, reusing the \hs{topological} functions etc.

### 2018-03-16 src/Development/Build/Task/Monad.hs
Mirror paper

@@ -77,7 +77,7 @@ compute task fetch = fmap runIdentity . task (pure . fetch)
 
 -- The version used in the paper
 compute' :: Task Monad k v -> Store i k v -> k -> Maybe v
-compute' task store key = runIdentity <$> task (Identity . flip getValue store) key
+compute' task store = fmap runIdentity . task (\k -> Identity (getValue k store))
 
 -- | Run a task with a partial lookup function. The result @Left k@ indicates
 -- that the task failed due to a missing dependency @k@. Otherwise, the

### 2018-03-17 paper/5-implementations.tex
More work on S5

@@
 
-We provide the complete implementation of \Make in our framework in
+We provide an implementation of \Make using our framework in
 Fig.~\ref{fig-make-implementation}. \Make processes keys in a linear order based
-on the topological sort (see~\S\ref{sec-concurrency} for parallel \Make). For
+on a topological sort (see~\S\ref{sec-concurrency} for parallel \Make). For
 each key, it builds it if it is older than any of its dependencies. We capture
@@
     process key deps act = do
-        (modTime, now) <- getInfo <$> get
+        (modTime, now) <- gets getInfo
         let dirty = or [ modTime dep > modTime key | dep <- deps ]
@@
             let newModTime k = if k == key then now else modTime k
-            modify $ \s -> putInfo (putValue s key v) (newModTime, now + 1)
+            modify $ putInfo (newModTime, now + 1) . putValue key v
 \end{minted}
@@
 -- Topological dependency strategy
-topological :: Eq k => (k -> [k] -> State (Store i k v) v -> State (Store i k v) (@@))
+topological :: Eq k => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
             -> Build Applicative i k v
 topological process task key = execState $ forM_ chain $ \k -> do
-    let fetch k = do store <- get; return (getValue store k)
+    let fetch k = gets (getValue k)
     case task fetch k of
@@
   where
-    deps  = dependencies task
-    chain = topologicalSort deps (transitiveClosure deps key)
+    deps  = dependencies task -- dependencies is defined in @\S\ref{secdeps}@
+    chain = topSort deps (reachable deps key)
 \end{minted}
@@
 The helper function \hs{topological} calls \hs{process} on every \hs{key} in
-the topological order, providing the list of its dependencies \hs{deps} and the
+a topological order, providing the list of its dependencies \hs{deps} and the
 action \hs{act} to compute the resulting value if it needs to be rebuilt. To
@@
 to extract static dependencies from the provided applicative \hs{task}. We
-compute the linear processing \hs{chain} by taking the \hs{transitiveClosure}
-of \hs{key}'s dependencies, and performing the \hs{topologicalSort} of the
+compute the linear processing \hs{chain} by taking the keys \hs{reachable}
+from \hs{key} via dependencies, and performing the topological sort of the
 result. We omit implementation of textbook graph algorithms
-\hs{transitiveClosure} and \hs{topologicalSort}, e.g.
+\hs{reachable} and \hs{topSort}, e.g.
 see~\cite{cormen2001introduction}.
@@
 
-\todo{AM}{Note that anything updating input files must also update Make's info.}
-\todo{AM}{Say that in reality that build rules mutate the state themselves.}
-
 \subsection{\Excel}\label{sec-implementation-excel}
 
-As a programming exercise, \Excel is the most challenging, due to the
-complicated reordering dependency strategy that requires retrying out-of-order
-operations, see Fig.~\ref{fig-excel-implementation}.
-
+We define \Excel, with it's reordering dependency strategy, in  Fig.~\ref{fig-excel-implementation}.
 \Excel's persistently stored information is a triple: (i) the dirty bit
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Reordering dependency strategy (implementation omitted)
+-- Reordering dependency strategy (implementation omitted, 21 lines)
 reordering :: Ord k
@@
     process key act = do
-        (dirty, deps) <- getInfo <$> get
+        (dirty, deps) <- gets getInfo
         let rebuild = dirty key || case deps key of SubsetOf ks -> any dirty ks
@@
                         let newDirty k = if k == key then True else dirty k
-                        modify $ \s -> putInfo (putValue s key v) (newDirty, deps)
+                        modify $ putInfo (newDirty, deps) . putValue key v
                 return (Just result)
@@
 
-The helper function \hs{reordering}, whose 21-line implementation we omit since it is
+The helper function \hs{reordering}, whose implementation we omit since it is
 technical and not particularly enlightening, calls the function \hs{process} to
@@
 \cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
-not needed, \hs{process} returns \hs{Nothing} to indicate this. Otherwise, it
-executes \hs{act} leading to two possible \hs{result}s:
+not needed, \hs{process} returns \hs{Nothing} to indicate that. Otherwise, it
+executes \hs{act} leading to one of two possible \hs{result}s:
 
@@
 
-In both of the above cases, we notify the parent \hs{reordering} function about
+In both of the above cases, we notify the parent \hs{reordering} function of
 the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
@@
 
+In reality \Excel works slightly differently (as far as we are able to ascertain) --
+on a change it propagates the dirty bits forward using the \hs{dynamicDependencies}, then only checks if the current \hs{key} is dirty.
+While both methods are equivalent, merely changing the interleaving, our approach allows us to model more of the behaviour of \Excel.
+
 \subsection{\Shake}\label{sec-implementation-shake}
@@
 -- Determine whether a trace is relevant to the current state
-traceMatch :: (Monad m, Eq k) => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
+traceMatch :: (Monad m, Eq k) =>
+    (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
 traceMatch check key ts = mapMaybeM f ts
@@
 -- Recursive dependency strategy
-recursive
-    :: Eq k => (k -> (k -> State (Store i k v, [k]) v) -> State (Store i k v, [k]) (v, [k]) -> State (Store i k v, [k]) ())
+recursive :: Eq k =>
+    (k -> (k -> State (Store i k v, [k]) v)
+       -> State (Store i k v, [k]) (v, [k])
+       -> State (Store i k v, [k]) ())
     -> Build Monad i k v
@@
 \item[\hs{recursive}] defines the dependency ordering pattern.
-It requires a \hs{process} function that builds a key given a way to build a dependency, and a way to build the result.
+It requires a \hs{process} function that builds a key given a way to recursively build a dependency, and a way to run task to produce result.
 The main purpose of \hs{recursive} is to ensure that in a single run any key is not built twice -- for which it extends the \hs{State} monad with a list of \hs{done} keys.

### 2018-03-17 src/Development/Build/Utilities.hs

@@ -1,6 +1,6 @@
 module Development.Build.Utilities (
     -- * Graph operations
-    closure, topSort, reach, reachM,
+    reachable, topSort, reach, reachM,
 
     -- * Transformers
     AltConst (..),
@@ -13,8 +13,8 @@ import Control.Applicative
 import Data.Functor.Identity
 
 -- TODO: Implement
-closure :: Eq a => (a -> [a]) -> a -> [a]
-closure = undefined
+reachable :: Eq a => (a -> [a]) -> a -> [a]
+reachable = undefined
 
 topSort :: Eq a => (a -> [a]) -> [a] -> [a]
 topSort = undefined

### 2018-03-17 paper/3-abstractions.tex
More work on S5
Work on S3

@@
 \vspace{-2mm}
-\paragraph{Build system} A \emph{build system} takes a task description, a target
-key, and a store, and returns a new store in which the target key has an
-up to date value.
+\paragraph{Build system} A \emph{build system} takes a task description, a
+\emph{target} key, and a store, and returns a new store in which the target key
+and all its dependencies have an up to date value.
 
@@
 \noindent
-The keys that belong to the transitive closure of \hs{key}'s dependencies fall
+The keys that are reachable from the target \hs{key} via dependencies fall
 into two classes: input keys and non-input keys, which we will denote by $I$ and
@@
 support for cyclic tasks, see~\S\ref{sec-iterative-compute}.
+
+The above definition of correctness needs to be adjusted to accommodate
+non-deterministic tasks and shallow cloud builds; we address this
+in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects}, respectively.
 
@@
 the function \hs{track}: a combination of \hs{compute} and \hs{dependencies}
-that computes both the resulting value and the list of its dependencies:
+that computes both the resulting value and the list of its dependencies in an
+arbitrary monadic context~\hs{m}:
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
-track task fetch = fmap runWriter . task (\k -> writer (fetch k, [k]))
+track :: Monad m => Task Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
+track task fetch = fmap runWriterT . task trackingFetch
+  where
+    trackingFetch :: k -> WriterT [k] m v
+    trackingFetch k = tell [k] >> lift (fetch k)
 \end{minted}
@@
 \noindent
-\todo{AM}{Type of track has changed - we no longer supply the Store, just an access fuction}
-\todo{AM}{trackM is essential for Shake, what do we do there?}
-The implementation uses the standard Haskell \hs{Writer} monad, which allows us
+The implementation uses the standard Haskell \hs{WriterT} \emph{monad
+transformer}~\cite{liang1995monad}, which allows us
 to record addition information -- a list of keys of type \hs{[@@k]} -- when
-computing a task. We instantiate \hs{f} to \hs{Writer}~\hs{[@@k]}~\hs{v}, and feed
-\hs{fetch}~\hs{k}~\hs{=}~\hs{writer}~\hs{(@@getValue}~\hs{store}~\hs{k,}~\hs{[@@k])}
-to the task, which in addition to fetching a value from the store, tracks the
-corresponding key. Here is an example of \hs{track}ing monadic tasks:
+computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
+with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
+corresponding key. The \hs{task} returns the value of type
+\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
+\hs{runWriterT} to the contents of \hs{Maybe}.
 
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store1 = initialise () (\key -> if key == "A1" then 10 else 1)
-@\ghci@ track sprsh2 store1 "B1"
-Just (1,["C1","B2"])
-\end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ store2 = initialise () (\key -> if key == "A1" then 10 else 20)
-@\ghci@ track sprsh2 store2 "B2"
-Just (20,["C1","B1"])
-\end{minted}
-\vspace{1mm}
+% Here is an example of \hs{track}ing monadic tasks:
 
-\noindent
-As expected, the dependencies of \hs{sprsh2} (see the spreadsheet
-in~\S\ref{sec-task-monad}) are determined by the contents of the store.
-Furthermore, the above example highlights the static cycle between cells
-\cmd{B1} and \cmd{B2}: in \hs{store1}, we have \cmd{B1 = B2 = A1 = 10},
-whereas in \hs{store2}, we have \cmd{B2 = B1 = A2 = 20}.
+% \vspace{1mm}
+% \begin{minted}[xleftmargin=10pt]{haskell}
+% @\ghci@ store1 = initialise () (\key -> if key == "A1" then 10 else 1)
+% @\ghci@ track sprsh2 store1 "B1"
+% Just (1,["C1","B2"])
+% \end{minted}
+% \vspace{1mm}
+% \begin{minted}[xleftmargin=10pt]{haskell}
+% @\ghci@ store2 = initialise () (\key -> if key == "A1" then 10 else 20)
+% @\ghci@ track sprsh2 store2 "B2"
+% Just (20,["C1","B1"])
+% \end{minted}
+% \vspace{1mm}
+
+% \noindent
+% As expected, the dependencies of \hs{sprsh2} (see the spreadsheet
+% in~\S\ref{sec-task-monad}) are determined by the contents of the store.
+% Furthermore, the above example highlights the static cycle between cells
+% \cmd{B1} and \cmd{B2}: in \hs{store1}, we have \cmd{B1 = B2 = A1 = 10},
+% whereas in \hs{store2}, we have \cmd{B2 = B1 = A2 = 20}.
 

### 2018-03-17 src/Development/Build/Task/Monad.hs
Sync with paper

@@ -28,9 +28,10 @@ dependencies task store = execWriterT . sequenceA . task fetch
 track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
 track task fetch = fmap runWriter . task (\k -> writer (fetch k, [k]))
 
-trackM :: Monad m => Task Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
-trackM task fetch key = runWriterT <$> task trackingFetch key
+trackM :: forall m k v. Monad m => Task Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
+trackM task fetch = fmap runWriterT . task trackingFetch
   where
+    trackingFetch :: k -> WriterT [k] m v
     trackingFetch k = tell [k] >> lift (fetch k)
 
 transitiveDependencies :: (Eq k, Monad m) => Task Monad k v

### 2018-03-17 paper/4-build.tex

@@
 \hline
-Dirty bit \S\ref{sec-dirty-bit}                    & \Make    & \Excel                   & Approximate \Shake  \\\hline
-Verifying trace \S\ref{sec-verifying-traces}       & \Ninja   & Traced \Excel            & \Shake                      \\\hline
-Constructive trace \S\ref{sec-constructive-traces} & \Bazel   & Cloud \Excel             & Cloud \Shake                \\\hline
+Dirty bit \S\ref{sec-dirty-bit}                    & \Make    & \Excel                   & Approximate \Shake*  \\\hline
+Verifying trace \S\ref{sec-verifying-traces}       & \Ninja   & Traced \Excel*            & \Shake                      \\\hline
+Constructive trace \S\ref{sec-constructive-traces} & \Bazel   & Cloud \Excel*             & Cloud \Shake*                \\\hline
 \end{tabular}
 \vspace{0.5mm}
-\caption{Build systems \`a la Carte.\label{tab-build-systems}}
+\caption{Build systems \`a la Carte, systems marked * are hypothetical systems that do not currently exist.\label{tab-build-systems}}
 \end{table}

### 2018-03-17 paper/5-implementations.tex
Delete some old commented out stuff

@@
 there.
-
-% Our code comprises a framework, utility functions implementing the
-% various dependency orderings as per~\S\ref{sec-dependency-orderings}, and
-% actual implementations.
-
-% \subsection{Framework}
-
-% todo{SLPJ}{I think this entire section, including the mysterious M, can be deleted.  Or is
-% there anything that remains?}
-
-% \begin{figure}
-% \begin{minted}{haskell}
-% type M i k v r = State (i, k -> v, ...) r
-% runM :: Default i => M i k v a -> Maybe i -> (k -> v) -> (i, k -> v)
-
-% getStore :: k -> M i k v v
-% putStore :: k -> v -> M i k v ()
-% getInfo :: M i k v i
-% modifyInfo :: (i -> i) -> M i k v ()
-% \end{minted}
-% Note: we have omitted the context on \hs{k}, things like \hs{Eq} (to implement the store).
-% \caption{API with which to implement build systems}
-% \label{fig-M-api}
-% \end{figure}
-
-% We use \hs{runM} to switch between the pure world of \hs{Build} and a state monad \hs{M} which lets us more precisely express the ideas behind the build system without threading the state manually - the types of important operations are given in Figure~\ref{fig-M-api}.
-We provide functions like \hs{getStore} to load from the store and \hs{putStore} to write to the store.
-We'll introduce the other functions in in the monad \hs{M} as we need them while exploring the build systems.
 
@@
 on a non-deterministic build - something \Bazel already prohibits which is discussed more in \S\ref{sec-non-determinism}.
-\item Finally, a \hs{Monad} system with constructive traces can be efficiently encoded as \hs{Map k (Choice k v)}, assuming a definition of \hs{Choice} as:
+\item Finally, a \hs{Monad} system with constructive traces can be represented as \hs{Map k (Choice k v)}, assuming a definition of \hs{Choice} as:
 \begin{minted}{haskell}

### 2018-03-17 paper/6-engineering.tex
4-build

@@
 
-\todo{NM}{Isn't ``early aborting'' automatically handled by short-circuiting
-withing monads, i.e. \hs{Nothing} or \hs{Left e} cancel out all subsequent
-computation?}
+% Q: {Isn't ``early aborting'' automatically handled by short-circuiting
+% withing monads, i.e. \hs{Nothing} or \hs{Left e} cancel out all subsequent
+% computation?}
+% NM says: The build system could chose an early abort monad, but the tasks are free in the monad, so you can't get the encoding to cross between them.
 
@@
 
-\item[Non-determinism] Some tasks are \emph{non-determinstic}, producing a
+\item[Non-determinism] Some tasks are \emph{non-deterministic}, producing a
 result from a possible set. As an example, GHC when compiled using parallelism
@@
 file system to only materialise the file when the user accesses it \cite{gvfs}.
-
-\todo{NM}{I cited GVFS above, and could not find a proper reference to Google's
-internal equivalent. There are some mentions online of Piper, but not citable.}
 

### 2018-03-17 src/Development/Build/Task/Monad.hs
Rename closure -> reachable

@@ -45,7 +45,7 @@ isInput :: Task Monad k v -> k -> Bool
 isInput task = isNothing . task (const Proxy)
 
 inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
-inputs task store key = filter (isInput task) (closure deps key)
+inputs task store key = filter (isInput task) (reachable deps key)
   where
     deps k = maybe [] snd (track task (\k -> getValue k store) k)
 

### 2018-03-17 paper/7-related.tex
Minor tweaks around the bazel/buck non-det story

@@
 
-\item \Buck~\cite{buck} and \CloudBuild~\cite{esfahani2016cloudbuild} are both
-fundamentally similar to \Bazel, differing around areas such as
-non-determinism~\S\ref{sec-non-determinism}.
+\item \Buck~\cite{buck} is very similar to \Bazel. \CloudBuild~\cite{esfahani2016cloudbuild}
+differs by allowing non-determinism~\S\ref{sec-non-determinism}, thus more closely modelling
+our original definition of \Bazel from \S\ref{sec-implementation-bazel}.
 

### 2018-03-17 paper/6-engineering.tex

@@
 
-We note that while all the simple implementations presented in
-\S\ref{sec-implementations} can deal with non-determinism, some real build
-systems, e.g. \Buck and \Bazel, rely on task determinism.
+Most of the implementations in \S\ref{sec-implementations} can deal with non-determinism,
+apart from \Bazel and \Buck, which require deterministic execution, and in turn can
+optimise the number of roundtrips required to the server.
 

### 2018-03-17 src/Development/Build/Task.hs
Add fetchIO

@@ -112,6 +112,9 @@ sprsh3 :: Task MonadPlus String Integer
 sprsh3 fetch "B1" = Just $ (+) <$> fetch "A1" <*> pure 1 `mplus` pure 2
 sprsh3 _     _    = Nothing
 
+fetchIO :: String -> IO Integer
+fetchIO k = do putStr (k ++ ": "); read <$> getLine
+
 -- These type synonyms are not very useful, but enumerate all interesting cases.
 type FunctorialTask  k v = forall f. Functor     f => (k -> f v) -> k -> Maybe (f v)
 type ApplicativeTask k v = forall f. Applicative f => (k -> f v) -> k -> Maybe (f v)

### 2018-03-17 paper/3-abstractions.tex
Add example for track

@@
 So the type \hs{Task} emerges naturally, almost inevitably.  But now that
-it \emph{has} emgerged, we find the that constraints \hs{c} classify task descriptions
+it \emph{has} emerged, we find the that constraints \hs{c} classify task descriptions
 in a very interesting way:
@@
 
-\vspace{1mm}
+\vspace{-1mm}
 \definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
@@
     In other words, no inputs were corrupted during the
-    build\footnote{\todo{SLPJ}{Add the joke that GHC used to delete inputs?}}.
+    build.
+    %\footnote{\todo{SLPJ}{Add the joke that GHC used to delete inputs?}}.
 
@@
 \label{def-correct}
-\vspace{3mm}
+\vspace{2mm}
 
@@
 
-The above definition of correctness needs to be adjusted to accommodate
-non-deterministic tasks and shallow cloud builds; we address this
-in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects}, respectively.
+% The above definition of correctness needs to be adjusted to accommodate
+% non-deterministic tasks and shallow cloud builds; we address this
+% in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects}, respectively.
 
-\subsection{Computing dependencies}\label{sec-deps}
+\subsection{Computing dependencies}\label{sec-deps}\label{secdeps}
 
@@
 the actual work}. Here is the code:
-\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-\vspace{1mm}
+
 Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
@@
 transformer}~\cite{liang1995monad}, which allows us
-to record addition information -- a list of keys of type \hs{[@@k]} -- when
+to record additional information -- a list of keys of type \hs{[@@k]} -- when
 computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
@@
 \hs{runWriterT} to the contents of \hs{Maybe}.
+Here is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}:
 
-% Here is an example of \hs{track}ing monadic tasks:
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
+@\ghci@ fromJust $ track sprsh2 fetchIO "B1"
+C1: 1
+B2: 10
+(10,["C1","B2"])
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ fromJust $ track sprsh2 fetchIO "B1"
+C1: 2
+A2: 20
+(20,["C1","A2"])
+\end{minted}
+\vspace{1mm}
 
-% \vspace{1mm}
-% \begin{minted}[xleftmargin=10pt]{haskell}
-% @\ghci@ store1 = initialise () (\key -> if key == "A1" then 10 else 1)
-% @\ghci@ track sprsh2 store1 "B1"
-% Just (1,["C1","B2"])
-% \end{minted}
-% \vspace{1mm}
-% \begin{minted}[xleftmargin=10pt]{haskell}
-% @\ghci@ store2 = initialise () (\key -> if key == "A1" then 10 else 20)
-% @\ghci@ track sprsh2 store2 "B2"
-% Just (20,["C1","B1"])
-% \end{minted}
-% \vspace{1mm}
+\noindent
+As expected, the dependencies of the cell \cmd{B1} from \hs{sprsh2} (see the
+spreadsheet in~\S\ref{sec-task-monad}) are determined by the value of \cmd{C1},
+which in this case is obtained by reading from the standard input.
 
-% \noindent
-% As expected, the dependencies of \hs{sprsh2} (see the spreadsheet
-% in~\S\ref{sec-task-monad}) are determined by the contents of the store.
 % Furthermore, the above example highlights the static cycle between cells

### 2018-03-17 paper/4-build.tex
Minor tweaks to S4

--- 2018-03-16 paper/4-build.tex
@@
 When using a dirty bit, it is necessary to check all the dependencies of a key.
-For applicative build systems that list is simple to obtain, but for monadic build systems there is no general way to get all dependencies.
-Instead \Excel computes a \emph{static approximation} of the dependencies.
-For applicative rules that approximation is correct.
-For functions such as \cmd{IF} it marks the cell dirty if \emph{any} potential dependency has changed, even on the untaken \emph{if} branch.
-For functions such as \cmd{INDIRECT} whose dependencies cannot be guessed, it conservatively assumes the dependencies have always changed.
+For applicative build systems that list is easy to obtain, but for monadic
+build systems there is no general way to get all dependencies. Instead \Excel
+computes a \emph{static approximation} of the dependencies. For applicative
+tasks that approximation is correct. For functions such as \cmd{IF} it marks the
+cell dirty if \emph{any} potential dependency has changed, even on the untaken
+\emph{if} branch. For functions such as \cmd{INDIRECT} whose dependencies cannot
+be guessed, it conservatively assumes the dependencies have always changed.
 
@@
 
-We assume that \hs{Hash v} is a small constant size, constructed from hashing the
+We assume that \hs{Hash}~\hs{v} is a small constant size, constructed from hashing the
 underlying \hs{v} rather than storing it directly. Checking a trace requires
@@
 
-A verifying trace build system needs to maintain a set of traces. After computing a
-fresh value we add its \hs{Trace k v} to the set. Therefore the information stored
-by a build system that verifies traces can be stored as a list or set - we chose \hs{[Trace k v]} for simplicity.
-In practice, different build systems can optimise the data structures used by traces for their specific use cases, which we discuss in \S\ref{sec-smart-traces}.
+A build system that uses verifying traces needs to persistently maintain a set
+of traces. After computing a fresh value we add its \hs{Trace}~\hs{k}~\hs{v} to
+the set. Therefore the information stored by a build system that verifies
+traces can be stored as a list or set~--~we chose \hs{[Trace}~\hs{k}~\hs{v]}
+for simplicity. In practice, different build systems can optimise the data
+structures used by traces for their specific use cases, which we discuss
+in~\S\ref{sec-smart-traces}.
 
@@
 \hline
-Property           & Topological \S\ref{sec-topological} & Reordering \S\ref{sec-reordering} & Recursive \S\ref{sec-recursive}    \\\hline
+Property           & Topological\hspace{2mm}\S\ref{sec-topological} & Reordering\hspace{2mm}\S\ref{sec-reordering} & Recursive\hspace{2mm}\S\ref{sec-recursive}    \\\hline
 \hline
-Dirty bit \S\ref{sec-dirty-bit}                    & \Make    & \Excel                   & Approximate \Shake  \\\hline
-Verifying trace \S\ref{sec-verifying-traces}       & \Ninja   & Traced \Excel            & \Shake                      \\\hline
-Constructive trace \S\ref{sec-constructive-traces} & \Bazel   & Cloud \Excel             & Cloud \Shake                \\\hline
+Dirty bit\hfill\S\ref{sec-dirty-bit}                    & \Make    & \Excel                   & Approximate \Shake*  \\\hline
+Verifying trace\hfill\S\ref{sec-verifying-traces}       & \Ninja   & Traced \Excel*            & \Shake                      \\\hline
+Constructive trace\hspace{2mm}\hfill\S\ref{sec-constructive-traces} & \Bazel   & Cloud \Excel*             & Cloud \Shake*                \\\hline
 \end{tabular}
-\vspace{0.5mm}
-\caption{Build systems \`a la Carte.\label{tab-build-systems}}
+\vspace{2mm}
+\caption{Build systems \`a la carte. Systems marked * are hypothetical systems that do not currently exist.\label{tab-build-systems}}
+\vspace{-2mm}
 \end{table}
 
-With the information in this section we can build a table comparing the dependency order strategy with the out-of-date keys strategy, providing 9 possible build systems, 5 of which are actually inhabited by existing build systems (we discuss \Ninja \cite{ninja} in \S\ref{sec-related-build}).
-Of the remaining 4 spots, we believe neither Traced or Cloud \Excel make sense - the \Excel approach of reordering combined with static approximations reduces the memory usage significantly.
-However, as soon as you are paying the cost of storing traces, that benefit is gone.
-The advantage of Approximate \Shake over \Shake would be that it could avoid storing traces and having a separate information database, but that advantage is minor compared to the technical restrictions and approximations it would provide, so we consider it unlikely to be built.
-The Cloud \Shake system seems like an interesting and important point in the design space, which we explore further in \S\ref{sec-implementation-cloud-shake}.
+With the information in this section we can build a table comparing the
+dependency order strategy with the out-of-date keys strategy, providing 9~possible
+build systems, 5~of which are actually inhabited by existing build systems
+(we discuss \Ninja \cite{ninja} in \S\ref{sec-related-build}). Of the
+remaining 4~spots, we believe neither Traced or Cloud \Excel make sense~--~the
+\Excel approach of reordering combined with static approximations reduces
+the memory usage significantly. However, as soon as you are paying the cost of
+storing traces, that benefit is gone. The advantage of Approximate \Shake over
+\Shake would be that it could avoid storing traces and having a separate
+information database, but that advantage is minor compared to the technical
+restrictions and approximations it would provide, so we consider it unlikely
+to be built. The Cloud \Shake system is an interesting and important point in
+the design space, which we explore further in~\S\ref{sec-implementation-cloud-shake}.

### 2018-03-17 src/Development/Build.hs
Add type signature

@@ -77,9 +77,10 @@ topological process task key = execState $ forM_ chain $ \k -> do
 type Time = Integer
 type MakeInfo k = (k -> Time, Time)
 
-make :: Eq k => Build Applicative (MakeInfo k) k v
+make :: forall k v. Eq k => Build Applicative (MakeInfo k) k v
 make = topological process
   where
+    process :: k -> [k] -> State (Store (MakeInfo k) k v) v -> State (Store (MakeInfo k) k v) ()
     process key deps act = do
         (modTime, now) <- gets getInfo
         let dirty = or [ modTime dep > modTime key | dep <- deps ]

### 2018-03-17 paper/5-implementations.tex
Minor changes

@@
 Fig.~\ref{fig-make-implementation}. \Make processes keys in a linear order based
-on a topological sort (see~\S\ref{sec-concurrency} for parallel \Make). For
+on a topological sort (see~\S\ref{sec-parallelism} for parallel \Make). For
 each key, it builds it if it is older than any of its dependencies. We capture
@@
 
+% Type signature for process. Not sure we really want this, as it's quite long
+% and is repeated in topological.
+% process :: k -> [k] -> State@\,@(Store@\,@(MakeInfo@\,\blk{k}@)@\,\blk{k}@@\,\blk{v}@)@\,\blk{v}@ -> State@\,@(Store@\,@(MakeInfo@\,\blk{k}@)@\,\blk{k}@@\,\blk{v}@)@\,@(@@)
 \begin{figure}
@@
 -- Standard graph algorithms (implementation omitted)
-transitiveClosure :: Eq a => (a -> [a]) ->  a  -> [a]
-topologicalSort   :: Eq a => (a -> [a]) -> [a] -> [a]
+reachable :: Eq a => (a -> [a]) ->  a  -> [a]
+topSort   :: Eq a => (a -> [a]) -> [a] -> [a]
 \end{minted}
@@
                     MissingDependency _ -> return ()
-                    Result v _dynamicDependencies -> do
+                    Result v dynamicDependencies -> do
                         let newDirty k = if k == key then True else dirty k
@@
 
+The \Shake approach for dependency tracking involves recording traces and verifying them, for which we use the \hs{Trace} type defined in \ref{sec-verifying-traces}.
+Complete code is given in \S\ref{fig-shake-implementation}, split into three functions:
+
+\begin{description}
+\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the \hs{key} you are currently building, and a function \hs{check} which checks a specified dependency.
+It returns the \hs{result} field of all traces that match.
+Since \hs{check} is in an arbitrary monad, the function has to use \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&} functions\footnote{These functions are all available in the \cmd{extra} library on Hackage.}.
+\item[\hs{recursive}] defines the dependency ordering pattern.
+It requires a \hs{process} function that builds a key given a way to recursively build a dependency, and a way to run task to produce result.
+The main purpose of \hs{recursive} is to ensure that in a single run no key is built twice~--~for which it extends the \hs{State} monad with a list of \hs{done} keys.
+\item[\hs{shake}] ties everything together.
+First it tests if the traces allow the current state of the target key.
+If not, it builds the key and updates the store.
+The only subtlety is that Shake calls \hs{fetch} on the keys while checking them~--~building the last-recorded dependencies before checking them.
+One minor annoyance is that the \hs{State} has been extended and thus needs to be projected using \hs{fst} before it is used.
+\end{description}
+
 \begin{figure}
@@
 -- Determine whether a trace is relevant to the current state
-traceMatch :: (Monad m, Eq k) =>
-    (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
+traceMatch :: (Monad m, Eq k)
+    => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
 traceMatch check key ts = mapMaybeM f ts
@@
 -- Recursive dependency strategy
-recursive :: Eq k =>
-    (k -> (k -> State (Store i k v, [k]) v)
+recursive :: Eq k => (k -> (k -> State (Store i k v, [k]) v)
                   -> State (Store i k v, [k]) (v, [k])
@@
                   -> Build Monad i k v
-recursive process task key store = fst $ execState (ensure key) (store, [])
+recursive process task key store = fst $ execState (ensure key) (store, [@@])
     where
@@
 \end{minted}
-\vspace{-2mm}
+\vspace{-3mm}
 \caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
-\vspace{-2mm}
+\vspace{1mm}
 \end{figure}
-
-The \Shake approach for dependency tracking involves recording traces and verifying them, for which we use the \hs{Trace} type defined in \ref{sec-verifying-traces}.
-Complete code is given in \S\ref{fig-shake-implementation}, split into three functions:
-
-\begin{description}
-\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the \hs{key} you are currently building, and a function \hs{check} which checks a specified dependency.
-It returns the \hs{result} field of all traces that match.
-Since \hs{check} is in an arbitrary monad, the function has to use \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&} functions\footnote{These functions are all available in the \cmd{extra} library on Hackage.}.
-\item[\hs{recursive}] defines the dependency ordering pattern.
-It requires a \hs{process} function that builds a key given a way to recursively build a dependency, and a way to run task to produce result.
-The main purpose of \hs{recursive} is to ensure that in a single run any key is not built twice -- for which it extends the \hs{State} monad with a list of \hs{done} keys.
-\item[\hs{shake}] ties everything together.
-First it tests if the traces allow the current state of the target key.
-If not, it builds the key and updates the store.
-The only subtlty is that Shake calls \hs{fetch} on the keys while checking them -- building the last-recorded dependencies before checking them.
-One minor annoyance is that the \hs{State} has been extended and thus needs to be projected using \hs{fst} before it is used.
-\end{description}
 
@@
 \vspace{-2mm}
-\caption{An implementation of \Bazel using our framework.}\label{fig-bazel-implementation}
+\caption{An implementation of \Bazel using our framework; \hs{topological} is
+defined in Fig.~\ref{fig-make-implementation}.}\label{fig-bazel-implementation}
 \vspace{-2mm}
@@
 Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, along with the \hs{Traces} type from \S\ref{sec-constructive-traces}.
-With these pieces in place, the implementation of \Bazel is given in Figure~\ref{fig-bazel-implementation}.
+With these pieces in place, the implementation of \Bazel is given in Fig.~\ref{fig-bazel-implementation}.
 We first figure out the possible results given the current state.
@@
 As a consequence \Bazel can compute the results of traces locally, without looking at \hs{Traces} (potentially saving a roundtrip to the server).
-To model this change in the code would require storing an additional transient piece of information \hs{done} of type \hs{Map k (Hash v)}, storing the computed hashes so far this run.
-With that available, \hs{getHash key s} would become:
+To model this change in the code would require storing an additional transient piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)}, storing the computed hashes so far this run.
+With that available, \hs{getHash}~\hs{key}~\hs{s} would become:
 
-\begin{minted}{haskell}
+\begin{minted}[xleftmargin=10pt]{haskell}
 hash (key, [ done Map.! d | d <- ds])
@@
 
+\noindent
 And that new hash would have to be stored in \hs{done}.
@@
 Using the abstractions and approaches built thus far, we have shown how to combine dependency scheme and change approach to reproduce existing build systems.
-In the attached materials we have implemented 9 build systems corresponding to all three dependency schemes, matched with all three change approaches.
-To us, the most interesting build system as yet unavailable would matching recursive ordering with constructive traces - providing a cloud-capable build system with minimality, cut-off and monadic dependencies.
-Using our framework it is possible to define and test such a system as per Figure~\ref{fig-cloudshake-implementation}.
+In the attached materials we have implemented 9~build systems corresponding to all three dependency schemes, matched with all three change approaches.
+To us, the most interesting build system as yet unavailable would matching recursive ordering with constructive traces~--~providing a cloud-capable build system with minimality, cutoff and monadic dependencies.
+Using our framework it is possible to define and test such a system as per Fig.~\ref{fig-cloudshake-implementation}.
 
@@
 
-\subsection{Smarter \hs{[Trace]} Structures}\label{sec-smart-traces}
+\subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}
 
-In the examples above, we have used \hs{[Trace k v]} to capture a list of traces -- however, using a list necessarily means that finding the right trace takes $O(n)$.
+In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list of traces -- however, using a list necessarily means that finding the right trace takes $O(n)$.
 For each of the \hs{Trace} based systems it is possible to devise a smarter representation, which we sketch below.
@@
 \Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}.
-Therefore, such systems can store \hs{Map k (Trace k v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
+Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
 \item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
-\item Any \hs{Applicative} system storing constructive traces, e.g.
-\Bazel, can index directly from the key and results to the output result - i.e.
-\hs{Map (k, [Hash v]) (Hash v)}.
+\item Any \hs{Applicative} build system storing constructive traces, e.g.
+\Bazel, can index directly from the key and results to the output result~--~i.e.
+\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{(Hash}~\hs{v)}.
 Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash.
 In this formulation we have removed the possibility for a single key/dependency state to map to multiple different hashes, e.g.
-on a non-deterministic build - something \Bazel already prohibits which is discussed more in \S\ref{sec-non-determinism}.
-\item Finally, a \hs{Monad} system with constructive traces can be represented as \hs{Map k (Choice k v)}, assuming a definition of \hs{Choice} as:
-\begin{minted}{haskell}
+on a non-deterministic build~--~something \Bazel already prohibits which is discussed more in~\S\ref{sec-non-determinism}.
+\item Finally, a \hs{Monad} build system with constructive traces can be represented as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
+\begin{minted}[xleftmargin=10pt]{haskell}
 data Choice k v = Choice k (Map (Hash v) Choice)
@@
 \end{minted}
-Here the \hs{Choice} effectively encodes a tree, asking successive questions about keys, and taking different branches based on the resulting value, until it reaches a final result.
+Here the \hs{Choice} encodes a tree, asking successive questions about keys, and taking different branches based on the answers, until it reaches a final result.
 Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.

### 2018-03-17 paper/6-engineering.tex
Minor changes

@@
 % computation?}
-% NM says: The build system could chose an early abort monad, but the tasks are free in the monad, so you can't get the encoding to cross between them.
+% NM says: The build system could chose an early abort monad, but the tasks are
+% free in the monad, so you can't get the encoding to cross between them.
 
-\subsection{Concurrency}\label{sec-concurrency}
+\subsection{Parallelism}\label{sec-parallelism}
 
 While we have given simple implementations assuming a single thread of execution,
-all the build systems we address are actually concurrent. While it complicates
-the model to add concurrency, the complications can be restricted exclusively
+all the build systems we address can actually build independent keys in parallel.
+While it complicates the model, the complications can be restricted exclusively
 to the dependency strategy:
@@
 the correctness definition (\S\ref{sec-build-correctness}): instead of requiring
-that the result of recomputing the \hs{task} matches the value in the \hs{ideal}
-store:
+that the result of recomputing the \hs{task} matches the value stored in the \hs{result}:
 \[
-\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}\]
+\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}\]
 \noindent
-we now require that \hs{ideal} contains \emph{one possible result of recomputing}
+we now require that \hs{result} contains \emph{one possible result of recomputing}
 the \hs{task}:
 \[
-\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{`elem`}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}
+\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{`elem`}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}
 \]

### 2018-03-17 paper/7-related.tex
Minor changes
Add some text on self-adjusting computation

@@
 
-\todo{AM}{Write a short section about memoization if enough time.}
+% \todo{AM}{Write a short section about memoization if enough time.}
 % \subsection{Memoization}\label{sec-related-memo}
@@
 well studied area, and in particular the contrast between different formulations
-has been well studied.
+has been well studied~\cite{acar2007selfadjusting}.
 
-\todo{AM}{Write this section.}
+Self-adjusting computations can automatically adjust to an external change
+to their inputs. A classic example is a self-adjusting sorting algorithm, which
+can efficiently (in $O(\log{n})$ time where $n$ is the length of the input)
+recalculate the result given an incremental change of the input. While very
+close to build systems in spirit, self-adjusting computations are mostly used
+for in-memory computation and rely on the ability to dynamically allocate new
+keys in the store for sharing intermediate computations~--~an intriguing feature
+rarely seen in build systems (\Shake's oracles~\S\ref{sec-polymorphism} can be
+used to model this to a limited degree).
+
+A lot of research has been dedicated to finding efficient data structures and
+algorithms for self-adjusting computations~--~we plan to explore this
+untapped (by build systems) resource in~future.
 
@@
 
-\begin{minted}{haskell}
+\begin{minted}[xleftmargin=10pt]{haskell}
 type Optic p a b s t = p a b -> p s t

### 2018-03-17 paper/5-implementations.tex
Minor changes

@@
                 modify $ \(s, done) -> (s, key:done)
-                case track task fetch key of
+                case track task fetch key of -- track is defined in @\S\ref{secdeps}@
                     Nothing -> return ()
@@
 
-The differences from \hs{bazel} are minor - the dependency scheme has changed from \hs{topological} to \hs{recursive}, and thus the dependency keys \hs{dk} are captured from the action rather than in advance, the transient state has gained a list of keys, and the checking calls \hs{fetch} to get the result instead of accessing the store directly.
+The differences from \hs{bazel} are minor~--~the dependency scheme has changed
+from \hs{topological} to \hs{recursive}, and thus the dependency keys \hs{dk}
+are captured from the action rather than in advance, the transient state has
+gained a list of keys, and the checking calls \hs{fetch} to get the result
+instead of accessing the store directly.
 
@@
 
-In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list of traces -- however, using a list necessarily means that finding the right trace takes $O(n)$.
-For each of the \hs{Trace} based systems it is possible to devise a smarter representation, which we sketch below.
-Note that these implementations do not avoid calls to \hs{compute}, merely overheads in the build system itself.
+In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list
+of traces~--~however, using a list necessarily means that finding the right trace
+takes $O(n)$. For each of the \hs{Trace} based systems it is possible to devise
+a smarter representation, which we sketch below. Note that these implementations
+do not avoid calls to \hs{compute}, merely overheads in the build system itself.
 

### 2018-03-17 paper/6-engineering.tex
Minor changes

@@
 \[
-\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{`elem`}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}
+\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}~\hs{k}
 \]
+where
+\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{k}~\hs{->}~\hs{Maybe}~\hs{[@@v]}
+returns the list of all possible results of the \hs{task} instead of just one
+value (`ND' stands for `non-deterministic').
+% (a possible implementation chooses
+% \hs{f}~\hs{=}~\hs{ListT}~\hs{Identity}~\hs{[@@v]}).
 

### 2018-03-17 paper/6-engineering.tex
Add a relaxed correctness definition for shallow builds

@@
 \begin{description}
-\item[Materialisation] Sometimes input files will involve many intermediate tasks
-before producing the end result, e.g. an installer package. These intermediate
-steps may be large, so some cloud builds can execute the graph \emph{without}
-downloading or \emph{materialising} the results of intermediate tasks -- only
-the final result. Some build systems can go even further, integrating with the
-file system to only materialise the file when the user accesses it \cite{gvfs}.
-
 \item[Eviction] The store of traces as shown grows indefinitely, but often
@@
 more commands in parallel.
+
+\item[Shallow builds] Sometimes input files will involve many intermediate tasks
+before producing the end result, e.g. an installer package. These intermediate
+steps may be large, so some cloud build systems are designed to build end
+products \emph{without} downloading or \emph{materialising} the results of
+intermediate tasks~--~only the final result~--~a so-called \emph{shallow build}.
+Some build systems can go even further, integrating with the file system to only
+materialise the file when the user accesses it \cite{gvfs}.
 \end{description}
+
+To legitimise shallow builds, we need to relax the correctness
+Definition~\ref{def-correct} as follows. Let the \hs{shallow} store correspond
+to the result of a shallow build. Then \hs{shallow} is correct, if \emph{there
+exists} \hs{result} which satisfies all requirements of
+Definition~\ref{def-correct}, \emph{such that} \hs{shallow} agrees with
+\hs{result} on the input keys \hs{k}~$\in$~$I$:
+\[
+\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}
+\]
+and on the target \hs{key}:
+\[
+\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result}.
+\]
+
+\noindent
+This relaxes the requirements on shallow builds by dropping the constraints on
+the \hs{shallow} store for all intermediate keys \hs{k}~$\in$~$O\setminus \{\hs{key}\}$.
 

### 2018-03-17 paper/5-implementations.tex

@@
 on a non-deterministic build~--~something \Bazel already prohibits which is discussed more in~\S\ref{sec-non-determinism}.
-\item Finally, a \hs{Monad} build system with constructive traces can be represented as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
+\item Finally, a \hs{Monad} build system with constructive traces can be stored as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
 \begin{minted}[xleftmargin=10pt]{haskell}

### 2018-03-17 paper/7-related.tex

@@
 
-% \todo{AM}{Write a short section about memoization if enough time.}
-% \subsection{Memoization}\label{sec-related-memo}
-
 \subsection{Self-adjusting computation}
@@
 untapped (by build systems) resource in~future.
+
+\subsection{Memoization}\label{sec-related-memo}
+
+\emph{Memoization} is a classic optimisation technique for storing values of a
+function instead of recomputing them each time the function is called. Minimal
+build systems (see the Definition~\ref{def-minimal}) certainly perform
+memoization: they \emph{store values instead of recomputing them each time}.
+Memoization can therefore be reduced to a minimal build system (as we
+demonstrate below), but not vice versa, since minimal build systems solve a more
+complex optimisation problem.
+
+As a simple example of using a build system for memoization, we solve a textbook
+dynamic programming problem~--~Levenshtein's \emph{edit
+distance}~\cite{levenshtein1966binary}: given two input strings $a$ and
+$b$, find the minimum-weight series of edit operations that transforms $a$
+to $b$. The edit operations are typically \emph{inserting}, \emph{deleting} or
+\emph{replacing} a symbol. The dynamic programming solution of this problem is
+widely known (e.g., see~\cite{cormen2001introduction}) that we provide its
+encoding in our \hs{Task} abstraction without further explanation.
+We address elements of strings $a_i$ and $b_i$ by keys \hs{A}~$i$ and \hs{B}~$i$,
+respectively, while the cost of a subproblem $d_{ij}$ is identified by
+\hs{D}~$i$~$j$.
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Key = A Integer | B Integer | D Integer Integer deriving Ord
+\end{minted}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+editDistance :: Task Monad Key Integer
+editDistance _     (D i 0) = Just $ pure i
+editDistance _     (D 0 j) = Just $ pure j
+editDistance fetch (D i j) = Just $ do
+    ai <- fetch (A i)
+    bj <- fetch (B i)
+    if ai == bj
+        then fetch (D (i - 1) (j - 1))
+        else do
+            x <- fetch (D (i - 1)  j     )
+            y <- fetch (D  i      (j - 1))
+            z <- fetch (D (i - 1) (j - 1))
+            return (1 + minimum [x, y, z])
+editDistance _ _ = Nothing
+\end{minted}
+
+\noindent
+When asked to build \hs{D}~$n$~$m$, a minimal build system will calculate the
+result using memoization. Furthermore, when an input symbol $a_i$ is changed,
+only necessary, incremental recomputation will be performed.
 

### 2018-03-17 paper/7-related.tex
Tweak memoization

@@
 \emph{replacing} a symbol. The dynamic programming solution of this problem is
-widely known (e.g., see~\cite{cormen2001introduction}) that we provide its
+so widely known (e.g., see~\cite{cormen2001introduction}) that we provide its
 encoding in our \hs{Task} abstraction without further explanation.
@@
         else do
-            x <- fetch (D (i - 1)  j     )
-            y <- fetch (D  i      (j - 1))
-            z <- fetch (D (i - 1) (j - 1))
-            return (1 + minimum [x, y, z])
+            insert  <- fetch (D  i      (j - 1))
+            delete  <- fetch (D (i - 1)  j     )
+            replace <- fetch (D (i - 1) (j - 1))
+            return (1 + minimum [insert, delete, replace])
 editDistance _ _ = Nothing
@@
 result using memoization. Furthermore, when an input symbol $a_i$ is changed,
-only necessary, incremental recomputation will be performed.
+only necessary, incremental recomputation will be performed~--~an optimisation
+that cannot be achieved just with memoization.
 

### 2018-03-17 src/Development/Build/Task.hs
Sync with paper

@@ -126,10 +126,10 @@ editDistance fetch (D i j) = Just $ do
     if ai == bj
         then fetch (D (i - 1) (j - 1))
         else do
-            x <- fetch (D (i - 1)  j     )
-            y <- fetch (D  i      (j - 1))
-            z <- fetch (D (i - 1) (j - 1))
-            return (1 + minimum [x, y, z])
+            insert  <- fetch (D  i      (j - 1))
+            delete  <- fetch (D (i - 1)  j     )
+            replace <- fetch (D (i - 1) (j - 1))
+            return (1 + minimum [insert, delete, replace])
 editDistance _ _ = Nothing
 
 -- These type synonyms are not very useful, but enumerate all interesting cases.

### 2018-03-17 paper/1-intro.tex

@@
 focus of attention.
-Seldom do people ask questions like ``What does it mean for my build
+Rarely do people ask questions like ``What does it mean for my build
 system to be correct?'' or ``What are the trade-offs between different
@@
 are often hidden away, and not the object of study.
-
 Recently, the challenges of scale have driven large software firms
 like Microsoft, Facebook, and Google to develop their own build
-systems, each with its own choices and idiosyncrasies.
+systems, exploring new points in the design space.
+
 In this paper we offer a general framework in which to understand
@@
 \item We identify two key design choices
-  that are typically deeply built into any build system:
+  that are typically deeply wired into any build system:
   \emph{the order in which dependencies are built} (\S\ref{sec-dependency-orderings})
@@
 
-\item Moreover, we can readily remix the ingredients to produce the first
+\item Moreover, we can readily remix the ingredients to describe the first
   build system that supports both \emph{dynamic dependencies}
@@
 points in space, we now see them as locations in a landscape,
-leading to better understanding of what they do and how they compare,
+leading to a better understanding of what they do and how they compare,
 and suggesting exploration of other (as yet unoccupied points) in the

### 2018-03-17 paper/2-background.tex
Proof S2

@@
 If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
-will first build \cmd{util.o} and \cmd{main.o}, in any order since the
-corresponding tasks are independent, and then build \cmd{main.exe}. If the
+will first build \cmd{util.o} and \cmd{main.o}, in any order since these
+tasks are independent, and then build \cmd{main.exe}. If the
 user modifies the sources of \cmd{util.h} and runs \Make again, it will
@@
 to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
-$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and which is
+$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, which is
 not known statically\footnote{One might say that the value of \cmd{C1} is statically
@@
 \Excel handles this example correctly: if \cmd{C1=1} and the user changes
-\cmd{A2}, \Excel will not recompute the result \cmd{B1}. Alas, other forms of
+\cmd{A2}, \Excel will not require \cmd{B1} to be computed in advance. Alas, other forms of
 dynamic dependencies can force \Excel to perform unnecessary computation.
@@
 current implementation of \Excel recomputes indirect references in every
-build~\cite{excel_recalc}. This clearly violates the minimality
+build~\cite{excel_recalc}. This approach clearly violates the minimality
 property~\ref{def-minimal}: if \cmd{C1=1} and the user modifies \cmd{A2}, \Excel
@@
 software build systems, where one often needs to manually initiate a full
-rebuild even if just a single build task has changed. We further discuss
-self-tracking in~\S\ref{sec-tracking-aspects}.
+rebuild even if just a single build task has changed. We discuss
+self-tracking further in~\S\ref{sec-tracking-aspects}.
 
@@
 executing exactly the same tasks on their local machines. A \emph{cloud build
-system} can speed up builds dramatically by transparently sharing build results
+system} can speed up builds dramatically by sharing build results
 among team members. Furthermore, cloud build systems allow one to perform
@@
 all intermediates in the cloud. We illustrate shallow cloud builds by an example
-(see~Fig.~\ref{fig-bazel}).
+in Fig.~\ref{fig-bazel}.
 
@@
 combination and must therefore build \cmd{util.o}, which results in changing its
-hash from~4~to~8. The obtained combination of hashes of \cmd{util.o} and
+hash from~4~to~8. The combination of hashes of \cmd{util.o} and
 \cmd{main.o} has not been encountered before either, therefore the build system
 first downloads \cmd{main.o} from the cloud and then builds \cmd{main.exe} by
-linking the two object files. When the build is complete, the obtained results
-may be uploaded to the cloud for future reuse by other team members.
+linking the two object files. When the build is complete, the results
+can be uploaded to the cloud for future reuse by other team members.
 
@@
 \emph{dependency graphs from all previous builds}, annotated with observed file
-hashes. The latter allows to bypass the execution of a task, by predicting the
+hashes. The latter allows builds to bypass the execution of a task, by predicting the
 hash of the result from the hashes of its dependencies, and subsequently
@@
     \item \Make stores file modification times, or rather, it relies on the file
-    system to do that. In principle a single \emph{dirty bit} per file is
-    sufficient, as will be demonstrated in~\S\ref{sec-implementations}.
+    system to do that.
     \item \Excel stores one dirty bit per cell and the calculation chain from
@@
 new build systems with desired properties by recombining parts of existing
-implementations. As a compelling example, we demonstrate how to combine
+implementations. As a compelling example, we demonstrate how to combine the
 advantages of \Shake and \Bazel in a cloud build system with dynamic

### 2018-03-17 paper/3-abstractions.tex
Minor changes
Add a relaxed correctness definition for shallow builds
Proof S3

@@
 This section presents purely functional abstractions that allow us to express
-all the intricacies of build systems discussed in the previous
-section~\S\ref{sec-background} and design complex build systems from simple
+all the intricacies of build systems discussed previously in \S\ref{sec-background},
+and design complex build systems from simple
 primitives. Specifically, we present the \emph{task} and \emph{build}
@@
 \end{figure}
-We  begin by establishing a common vocabulary for build systems.
+We begin by establishing a common vocabulary for build systems:
 
@@
 the worksheets, the keys are cell names (such as \cmd{A2}) and the values are
-numbers, strings etc, displayed as the cell contents. Many systems use
+numbers, strings etc, displayed as the cell contents. Many build systems use
 \emph{hashes} of values as compact summaries with a fast equality check.
@@
 \vspace{1mm}
-Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
+Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains
 the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
@@
 \vspace{1mm}
-Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
+We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
 (Real spreadsheet cells would contain a wider range of values, of course.)
@@
 \noindent
-As one can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
+As we can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
 right order (if it had built \cmd{B2} before building \cmd{B1}, the result would
@@
 \cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
-\hs{fetch}~\hs{"B1"} twice -- once per each occurrence of \cmd{B1} in the
+\hs{fetch}~\hs{"B1"} twice -- once per occurrence of \cmd{B1} in the
 formula.
@@
 
-Now we can see why the \hs{Task} abstraction is polymorphic in \hs{f}:
+The previous example shows why the \hs{Task} abstraction is polymorphic in \hs{f}, recall the definition:
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \noindent
-Here, \cmd{B1} and \cmd{B2} statically depend on each other, but \Excel (which
-uses dynamic dependencies) is perfectly happy. We can express this using
-our task abstraction as follows:
+Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel (which
+uses dynamic dependencies) is perfectly happy. We can express these rules using
+our task abstraction as:
 
@@
 The big difference compared to \hs{sprsh1} is that the computation now takes
-place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
-\hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.
+place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
+\hs{fetch} \emph{different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.
 
@@
 dependencies. For minimal build systems, however, dynamic dependencies, and hence
-monadic tasks, are much more challenging, as we will see
+monadic tasks, are much more challenging, as we shall see
 in~\S\ref{sec-implementations}.
@@
     \[
-    \hs{Just}~\hs{(}\hs{getValue}~\hs{result}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}.
+    \hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}.
     \]
@@
 \noindent
-The implementation uses the standard Haskell \hs{WriterT} \emph{monad
+This implementation uses the standard Haskell \hs{WriterT} \emph{monad
 transformer}~\cite{liang1995monad}, which allows us

### 2018-03-17 paper/7-related.tex

@@
 distance}~\cite{levenshtein1966binary}: given two input strings $a$ and
-$b$, find the minimum-weight series of edit operations that transforms $a$
+$b$, find the shortest series of edit operations that transforms $a$
 to $b$. The edit operations are typically \emph{inserting}, \emph{deleting} or

### 2018-03-17 paper/4-build.tex
S4 and S7 tweaks

@@
 As per the definition of minimality~\ref{def-minimal}, a minimal build
-system must \underline{rebuild only out-of-date keys} and at most once. The only
+system must \textbf{rebuild only out-of-date keys} and at most once. The only
 way to achieve the ``at most once'' requirement while producing a correct build
-result (\S\ref{sec-build-correctness}) is to \underline{build all keys in an
+result (\S\ref{sec-build-correctness}) is to \textbf{build all keys in an
 order that respects their dependencies}.
@@
 \vspace{1mm}
-We have underlined two different phrases above, and tackle each aspect separately.
+We have bolded two different phrases above, and tackle each aspect separately.
 
@@
 from a key to its dependencies, and the output \hs{key}, you can compute the
-linear order by first taking the transitive closure of \hs{key}'s dependencies,
-and then computing its topological sort. However, as we have seen
+linear order by first finding the reachable dependencies of \hs{key},
+and then computing a topological sort. However, as we have seen
 in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task,
@@
 actions themselves may be incremental (i.e. unnecessary tasks will not be performed),
-the pre-processing is not. We may try to incrementalise the topological sort by
+the pre-processing is not. We can try to incrementalise the topological sort by
 storing the topological order between build runs and assume it to
-be correct, but if the build determines it is wrong, fix it up.
+be correct, but if the build discovers it is wrong, fix it up.
 
@@
 (\S\ref{sec-general-build}), is to simply build dependencies when they are
-requested. By combining that with a transient log of which keys have already
+requested. By combining that with a transient set of which keys have already
 been built, you can obtain a minimal build system.
@@
 another task will have to be run. Assuming a an IO-driven task structure,
-that requires suspend a running task, which can be done with cheap
-green threads and blocking (the original approach of \Shake) or with a
-continuation-passing style (what \Shake does now). An alternative approach to
+that requires suspending a running task, which can be done with cheap
+green threads and blocking (the original approach of \Shake) or using
+continuation-passing style (what \Shake does currently). An alternative approach to
 suspending a task is to abort it and restart it again later, at the cost
@@
 When rebuilding, if a cell only depends on clean cells it is skipped, otherwise
-it is rebuilt and is marked dirty so that the cells that depend on it are
+it is rebuilt and marked dirty so that the cells that depend on it are
 subsequently rebuilt too.
@@
 a file is newer than its dependencies. The interesting property of
-this dirty bit is that it is not under control of \Make; rather it is
-existing file-system information that is repurposed. In particular,
-rebuilding a file automatically clears its dirty bit, and
+this dirty bit is that it is not under the control of \Make; rather it is
+existing file-system information that has been repurposed. In particular,
+modifying a file automatically clears its dirty bit, and
 automatically sets the dirty bit of the nodes depending on it. One
@@
 impossible task with only the ability to update to the latest modification time.
-It can approximate early cutoff by not touching the result file, and not marking it clean,
-but then it reruns in every subsequent build.
+\Make can approximate early cutoff by not modifying the result file, and not marking it clean,
+but then it will rerun in every subsequent build.
 

### 2018-03-17 paper/7-related.tex
S4 and S7 tweaks

@@
 well studied area, and in particular the contrast between different formulations
-has been well studied~\cite{acar2007selfadjusting}.
+has been thoroughly investigated~\cite{acar2007selfadjusting}.
 
@@
 rarely seen in build systems (\Shake's oracles~\S\ref{sec-polymorphism} can be
-used to model this to a limited degree).
+used to model this feature to a limited degree).
 
 A lot of research has been dedicated to finding efficient data structures and
-algorithms for self-adjusting computations~--~we plan to explore this
-untapped (by build systems) resource in~future.
+algorithms for self-adjusting computations~--~we plan to investigate how these
+insights can be utilised by build systems as future work.
 

### 2018-03-17 paper/4-build.tex
Final tweaks to S4

@@
 the set. Therefore the information stored by a build system that verifies
-traces can be stored as a list or set~--~we chose \hs{[Trace}~\hs{k}~\hs{v]}
+traces can be modelled as a list or set~--~we chose \hs{[Trace}~\hs{k}~\hs{v]}
 for simplicity. In practice, different build systems can optimise the data
@@
 Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{contents} \emph{without} recomputing it.
-We split the traces and contents as in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only at most one complete file.
+We split the traces and contents because in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only retrieve at most one complete file per key.
 

### 2018-03-17 paper/6-engineering.tex
Minor tweaks to 6 and 7

@@
 \item[Untracked state] Some tasks depend on untracked state -- for example
-your C compilation may explicitly list the \cmd{source.c} file as a dependency,
+C compilation will explicitly list the \cmd{source.c} file as a dependency,
 but it may not record that the version of \cmd{gcc} is also a dependency.
@@
 Most of the implementations in \S\ref{sec-implementations} can deal with non-determinism,
-apart from \Bazel and \Buck, which require deterministic execution, and in turn can
+apart from \Bazel, which requires deterministic execution, and in turn can
 optimise the number of roundtrips required to the server.
@@
 tasks, but with a series of iterative steps, as described
-in~\cite{shake-fixed-point}. It is important that the number of executions is
+by~\citet{shake-fixed-point}. It is important that the number of executions is
 bounded, otherwise the build system may not terminate (a legitimate concern

### 2018-03-17 paper/7-related.tex
Link self-adjusting computation, memoization and build systems
Minor tweaks to 6 and 7

@@
 
-\item \Buck~\cite{buck} is very similar to \Bazel. \CloudBuild~\cite{esfahani2016cloudbuild}
-differs by allowing non-determinism~\S\ref{sec-non-determinism}, thus more closely modelling
+\item \Buck~\cite{buck} is very similar to \Bazel at the level of abstraction presented here.
+
+\item \CloudBuild~\cite{esfahani2016cloudbuild}
+differs from \Bazel by allowing non-determinism~\S\ref{sec-non-determinism}, thus more closely modelling
 our original definition of \Bazel from \S\ref{sec-implementation-bazel}.

### 2018-03-17 paper/7-related.tex
Fix typo

@@
     ai <- fetch (A i)
-    bj <- fetch (B i)
+    bj <- fetch (B j)
     if ai == bj
@@
 that cannot be achieved just with memoization.
+
+Self-adjusting computation, memoization and build systems are inherently related
+topics, which poses the question of whether there is an underlying common
+abstraction waiting to be discovered.
 

### 2018-03-17 src/Development/Build/Task.hs

@@ -122,7 +122,7 @@ editDistance _     (D i 0) = Just $ pure i
 editDistance _     (D 0 j) = Just $ pure j
 editDistance fetch (D i j) = Just $ do
     ai <- fetch (A i)
-    bj <- fetch (B i)
+    bj <- fetch (B j)
     if ai == bj
         then fetch (D (i - 1) (j - 1))
         else do

### 2018-03-17 paper/3-abstractions.tex
Add authors, minor changes

@@
 
+\vspace{-2mm}
 \paragraph{Task description} Any build system requires the user to specify how to
@@
 target key is up to date. What exactly does ``up to date'' mean?  We answer
-that precisely in \S\ref{sec-build-correctness}.
+that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:
 
-\noindent
-Here is a simple build system:
 \vspace{1mm}
@@
 Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel (which
-uses dynamic dependencies) is perfectly happy. We can express these rules using
-our task abstraction as:
+uses dynamic dependencies) is perfectly happy. We can express this spreadsheet
+using our task abstraction as:
 
@@
 
-% The above definition of correctness needs to be adjusted to accommodate
-% non-deterministic tasks and shallow cloud builds; we address this
-% in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects}, respectively.
+The presented definition of correctness needs to be adjusted to accommodate
+non-deterministic tasks and shallow cloud builds, as will be discussed
+in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects},
+respectively.
 
@@
 the actual work}. Here is the code:
+
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \hs{runWriterT} to the contents of \hs{Maybe}.
-Here is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}:
+Below we give an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
 

### 2018-03-17 paper/4-build.tex
Add authors, minor changes

@@
 
+\vspace{-2mm}
 \subsubsection{Topological}\label{sec-topological}
@@
 
+\vspace{-2mm}
 \subsubsection{Reordering}\label{sec-reordering}
@@
 
+\vspace{-2mm}
 \subsubsection{Recursive}\label{sec-recursive}
@@
 This approach requires that a task may be started, then during that execution
-another task will have to be run. Assuming a an IO-driven task structure,
+another task will have to be run. Assuming an IO-driven task structure,
 that requires suspending a running task, which can be done with cheap
@@
 
+\vspace{-2mm}
 \subsubsection{A dirty bit}\label{sec-dirty-bit}
@@
 
+\vspace{-2mm}
 \subsubsection{Verifying traces}\label{sec-verifying-traces}

### 2018-03-17 paper/5-implementations.tex
Minor tweaks
Proof S5
Add authors, minor changes

@@
 \Excel into a cloud build system and ensuring \hs{reordering} is not overly
-fitted to \Excel alone (additional implementations will be submitted to the
-Artefact Evaluation).
+fitted to \Excel alone.
+%  (additional implementations will be submitted to the
+% Artefact Evaluation).
 
@@
 
+\vspace{-2mm}
 \subsection{\Shake}\label{sec-implementation-shake}
+\vspace{-1mm}
 
@@
 
+\vspace{-2mm}
 \subsection{\Bazel}\label{sec-implementation-bazel}
+\vspace{-1mm}
 
@@
 With that available, \hs{getHash}~\hs{key}~\hs{s} would become:
-
 \begin{minted}[xleftmargin=10pt]{haskell}
-hash (key, [ done Map.! d | d <- ds])
+hash (key, [ done Map.! d | d <- ds ])
 \end{minted}
-
-\noindent
 And that new hash would have to be stored in \hs{done}.
@@
 \caption{An implementation of Cloud \Shake using our framework.}\label{fig-cloudshake-implementation}
-\vspace{-2mm}
+\vspace{-4mm}
 \end{figure}
@@
 
+\vspace{-1mm}
 \subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}
+\vspace{-1mm}
 

### 2018-03-28 paper/3-abstractions.tex
Tweak wording (thanks Zhen!)

@@
 To bring a key up to date, \hs{fetch} asks the task description \hs{task} how
-to compute \hs{k}. If \hs{task} returns \hs{Nothing} the key is an input, so
-\hs{fetch} simply reads the result from the store. Otherwise \hs{fetch} runs
-the action \hs{act} returned by the \hs{task} to produce a resulting
+to compute the value of \hs{k}. If \hs{task} returns \hs{Nothing} the key is an
+input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
+runs the action \hs{act} returned by the \hs{task} to produce a resulting
 value~\hs{v}, records the new key/value mapping in the store, and returns \hs{v}.

### 2018-03-28 src/Build/Task.hs
Add MonadFail example

@@ -4,6 +4,7 @@ module Build.Task (Task, inputTask, isInput, compose, sprsh1, sprsh2, sprsh3) wh
 
 import Control.Applicative
 import Control.Monad
+import Control.Monad.Fail (MonadFail)
 import Control.Monad.Reader
 import Data.Maybe
 import Data.Proxy
@@ -126,6 +127,13 @@ sprsh2 _     _    = Nothing
 sprsh3 :: Task Alternative String Integer
 sprsh3 fetch "B1" = Just $ (+) <$> fetch "A1" <*> (pure 1 <|> pure 2)
 sprsh3 _     _    = Nothing
+
+sprsh4 :: Task MonadFail String Integer
+sprsh4 fetch "B1" = Just $ do
+    a1 <- fetch "A1"
+    a2 <- fetch "A2"
+    if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
+sprsh4 _ _ = Nothing
 
 fetchIO :: String -> IO Integer
 fetchIO k = do putStr (k ++ ": "); read <$> getLine

### 2018-03-30 paper/3-abstractions.tex
Fix compile errors (thanks Irakli!)

@@
 \begin{itemize}
-    \item \hs{store} and \hs{result} \emph{agree on inputs}, that is, for all
+    \item \hs{result} and \hs{store} \emph{agree on inputs}, that is, for all
     input keys \hs{k}~$\in$~$I$:
     \[
-    \hs{getValue}~\hs{k}~\hs{store}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}.
+    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{getValue}~\hs{k}~\hs{store}.
     \]
@@
     \[
-    \hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}.
+    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{@\std{fromJust}@}~\hs{(}\hs{compute}~\hs{task}~\hs{result}~\hs{k)}
     \]

### 2018-03-30 paper/5-implementations.tex
Fix a typo (thanks Evan!)
Fix compile errors (thanks Irakli!)

@@
 
-We define \Excel, with it's reordering dependency strategy, in  Fig.~\ref{fig-excel-implementation}.
+We define \Excel, with its reordering dependency strategy, in  Fig.~\ref{fig-excel-implementation}.
 \Excel's persistently stored information is a triple: (i) the dirty bit
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Choice k v = Choice k (Map (Hash v) Choice)
+data Choice k v = Choice k (Map (Hash v) (Choice k v))
                 | Result (Hash v)

### 2018-03-30 paper/6-engineering.tex
Fix compile errors (thanks Irakli!)

@@
 \[
-\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}\]
+\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{@\std{fromJust}@}~\hs{(}\hs{compute}~\hs{task}~\hs{result}~\hs{k)}\]
 \noindent
@@
 \[
-\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}~\hs{k}
+\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{@\std{fromJust}@}~\hs{(}\hs{computeND}~\hs{task}~\hs{result}~\hs{k)}
 \]

### 2018-04-05 src/Build/Trace.hs
Changes

@@ -19,16 +19,20 @@ import Data.Semigroup
 
 import qualified Data.Map as Map
 
+data Trace k v = Trace
+    { key     :: k
+    , depends :: [(k, Hash v)]
+    , result  :: Hash v }
+
+
 -- | An abstract data type for a set of verifying traces equipped with 'record',
 -- 'verify' and a 'Monoid' instance.
 newtype VT k v = VT [Trace k v] deriving (Monoid, Semigroup)
 
 -- | Record a new trace for building a @key@ with dependencies @deps@, obtaining
 -- the hashes of up-to-date values from the given @store@.
-recordVT :: Hashable v => Store i k v -> k -> [k] -> VT k v -> VT k v
-recordVT store key deps (VT ts) = VT (t : ts)
-  where
-    t = Trace key [ (k, getHash k store) | k <- deps ] (getHash key store)
+recordVT :: Hashable v => Store i k v -> k -> [k] -> VT k v
+recordVT store key deps = VT [Trace key [ (k, getHash k store) | k <- deps ] (getHash key store)]
 
 -- | Given a function to compute the hash of a key's current value,
 -- a @key@, and a set of verifying traces, return 'True' if the @key@ is
@@ -40,42 +44,32 @@ verifyVT fetchHash key (VT ts) = anyM match ts
         | k /= key  = return False
         | otherwise = andM [ (h==) <$> fetchHash k | (k, h) <- (key, result) : deps ]
 
-newtype CT k v = CT (Traces k v)
-
-instance Semigroup (CT k v) where
-    CT (Traces t1 c1) <> CT (Traces t2 c2) = CT (Traces (t1 ++ t2) (Map.union c1 c2))
-
-instance Monoid (CT k v) where
-    mempty  = CT (Traces [] Map.empty)
-    mappend = (<>)
-
-recordCT :: Hashable v => Store i k v -> k -> [k] -> CT k v -> CT k v
-recordCT store key deps (CT (Traces ts c)) = CT (Traces (t : ts) (Map.insert h v c))
-  where
-    h = getHash key store
-    v = getValue key store
-    t = Trace key [ (k, getHash k store) | k <- deps ] h
-
-verifyCT :: (Monad m, Eq k) => (k -> m (Hash v)) -> k -> CT k v -> m Bool
-verifyCT fetchHash key (CT (Traces ts _)) = verifyVT fetchHash key (VT ts)
-
[…]

### 2018-04-05 test/Main.hs

@@ -46,7 +46,7 @@ acyclicSpreadsheet cell = case name cell of
     _     -> Nothing
 
 targets :: [Cell]
-targets = [ "A1", "A2", "A3", "B1", "B2", "B3", "C1", "C2", "F0", "F1", "F30" ]
+targets = [ "A1", "A2", "A3", "B1", "B2", "B3", "C1", "C2", "F0", "F1", "F4" ]
 
 task :: Task Monad Cell Int
 task = spreadsheetTask spreadsheet

### 2018-04-05 paper/2-background.tex
Drop hanging word, tweak whitespace
Tweak whitespace
Add authors, minor changes
Improve Excel wording in section 2

@@
-\clearpage
 \section{Background}\label{sec-background}
@@
 
-The following property is essential for build systems, it is their raison d'\^etre:
+The following property is essential for build systems; indeed, it is their raison d'\^etre:
 
@@
 \begin{minted}[xleftmargin=10pt]{text}
-A1: 10      B1: IF(C1=1,A1,A2)      C1: 1
-A2: 20
-\end{minted}
-\vspace{1mm}
-
-\noindent
-Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
-to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
-$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, which is
-not known statically\footnote{One might say that the value of \cmd{C1} is statically
-known in this particular example, but imagine that it is the result of a long computation
-chain -- its value will only become available during the build.}.
-
-\Excel handles this example correctly: if \cmd{C1=1} and the user changes
-\cmd{A2}, \Excel will not require \cmd{B1} to be computed in advance. Alas, other forms of
-dynamic dependencies can force \Excel to perform unnecessary computation.
-Consider the following modification of the above example:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{text}
 A1: 10      B1: INDIRECT("A" & C1)      C1: 1
@@
 \noindent
-The new version uses the \cmd{INDIRECT} function, which allows us to reference a
-cell indirectly by a text string that is not necessarily known in advance. The
-current implementation of \Excel recomputes indirect references in every
-build~\cite{excel_recalc}. This approach clearly violates the minimality
-property~\ref{def-minimal}: if \cmd{C1=1} and the user modifies \cmd{A2}, \Excel
-will recompute \cmd{B1}, potentially triggering further unnecessary
-recalculation even though \cmd{B1} does not transitively depend on \cmd{A2}.
+The formula in \cmd{B1} uses the \cmd{INDIRECT} function, which takes a string and
+returns the value of the cell with that name.  The string evaluates
+to \cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
+formula in \cmd{B1} are determined by the value of \cmd{C1}, so there is no
+possibility\footnote{In this particular example one might say that
+the value of \cmd{C1} is statically known, but imagine that it is the result of a long computation
+chain -- its value will only become available during the build.}
+of computing the dependency graph prior to calc.
 
+To support dynamic dependencies,
 \Excel's build algorithm~\cite{excel_recalc} is significantly different from
@@
 in~\S\ref{sec-implementation-excel}.
+
+Dynamic dependencies complicate minimality.  In the above example,
+\cmd{B1} should only be recomputed if \cmd{A1} or \cmd{C1} change, but
+not if (say) \cmd{A2} changes; but these facts are not statically apparent.
+In pracice Excel implements a conservative approximation to
+minimality: roughly, it recomputes a formula if (a) the formula
+text mentions a changed cell, or (b) the formula uses a function like \cmd{INDIRECT}
+whose dependencies are not statically visible.
 
@@
 \begin{figure}[h]
+\vspace{-2mm}
 \begin{subfigure}[b]{0.90\linewidth}
@@
 list of release documents \cmd{docs.txt}.\label{fig-shake}}
-\vspace{-2mm}
+\vspace{-4mm}
 \end{figure}
@@
 \cmd{docs.txt}, \Shake will appropriately rebuild \cmd{release.txt} and
-\cmd{release.tar}, discovering the new dependency, as illustrated in
-Fig.~\ref{fig-shake}(b).
+\cmd{release.tar}, discovering the new dependency, see Fig.~\ref{fig-shake}(b).
 
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
-\vspace{-1mm}
 \caption{Checkout source files}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
-\vspace{-1mm}
 \caption{Build \cmd{main.exe}}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
-\vspace{-1mm}
 \caption{Modify \cmd{util.c} and rebuild}
 \end{subfigure}
-\vspace{-2mm}
 \caption{A cloud build example: (a)~checkout sources, (b)~download \cmd{main.exe}
@@
 \label{fig-bazel}}
-\vspace{-1mm}
 \end{figure}
@@
 \begin{table}[h]
-\vspace{-2mm}
 \smaller
@@
 \caption{Summary of build system differences.\label{tab-summary}}
-\vspace{-8mm}
+\vspace{-6mm}
 \end{table}

### 2018-05-02 src/Build/SelfTracking.hs
Drop TaskParser type synonym

@@ -36,11 +36,8 @@ fetchValueTask fetch key = extract <$> fetch (KeyTask key)
     extract (ValueTask t) = t
     extract _ = error "Inconsistent fetch"
 
--- For simplicity our task description parsers are total.
-type TaskParser c k v t = t -> Task c k v
-
 -- A model using Monad, works beautifully and allows storing the key on the disk
-selfTrackingM :: TaskParser Monad k v t -> Tasks Monad k t -> Tasks Monad (Key k) (Value v t)
+selfTrackingM :: (t -> Task Monad k v) -> Tasks Monad k t -> Tasks Monad (Key k) (Value v t)
 selfTrackingM _      _     (KeyTask _) = Nothing -- Task keys are inputs
 selfTrackingM parser tasks (Key     k) = runTask <$> tasks k
   where
@@ -51,7 +48,7 @@ selfTrackingM parser tasks (Key     k) = runTask <$> tasks k
 
 -- | The Applicative model requires every key to be able to associate with its
 -- environment (e.g. a reader somewhere). Does not support cutoff if a key changes
-selfTrackingA :: TaskParser Applicative k v t -> (k -> t) -> Tasks Applicative (Key k) (Value v t)
+selfTrackingA :: (t -> Task Applicative k v) -> (k -> t) -> Tasks Applicative (Key k) (Value v t)
 selfTrackingA _      _   (KeyTask _) = Nothing -- Task keys are inputs
 selfTrackingA parser ask (Key k) = Just $ \fetch ->
     fetch (KeyTask k) *> (Value <$> parser (ask k) (fetchValue fetch))

### 2018-05-02 paper/response.md
Draft a response

# Reviewer A

Thank you!

# Reviewer B

We disagree about correctness vs minimality: our main focus is correctness.

# Reviewer C

> the code of each build system is quite different from each other

The framework does provide reusable "schedulers" (topological, recursive, etc.)
And in the final version we will also provide reusable "rebuilders".

> Build abstraction is trivial

The whole point of the paper is to ignore many crucially important and
substantial engineering aspects of build systems and bring out more subtle
design choices.

Parallelism is an example.

### 2018-05-03 paper/response.md
Expand the response

@@
+We thank all reviewers for their comments and suggestions and will address them
+in the revision. We also respond to some of them below.
+
 # Reviewer A
 
-Thank you!
+> you refer to a "static" dependency on release.txt,
+> but I can't see how the build system could distinguish this
+> dependency from those arising from parsing release.txt.
+
+Indeed, Shake does not distinguish static dependencies from dynamic: it
+treats them uniformly as if they all were dynamic.
+
+> how does the file system metadata fit into this model? Is it persistent
+> build information?
+
+We consider file system metadata to be a part of persistent build information.
+One could refine the model by introducing a separate notion for the file system
+metadata, but we decided against this, since it makes the model more complex.
+
+> Presumably you could call `dependencies` on monadic tasks --- `Monad`s are
+> `Applicative` after all --- but just wouldn't get any useful information back?
+
+No, we cannot call `dependencies` on monadic tasks. Dependencies of a monadic
+task cannot be determined without providing actual values: for `sprsh2`, for
+example, we need to know the value of the cell `C1` to determine the exact set
+of dependencies; `track` obtains values using the `fetch :: k -> m v` argument.
 
@@
 
-We disagree about correctness vs minimality: our main focus is correctness.
+> The emphasis of the minimality property over correctness seems somewhat off.
+
+We disagree that our emphasis is on minimality. On the contrary, a larger part
+of the paper is dedicated to the precise formulation of what is means for a
+build system to be correct -- see S3.6 (correctness), S6.3 (correctness in
+presence of non-determinism) and S6.4 (correctness of shallow cloud builds).
+
+> The property most build system authors chase after is that an incremental
+> build should be equivalent to a clean build, up to non-determinism - if there
+> is any in your build!
+
+We fully agree that correctness must take precedence. Our definition of
+correctness allows the designers of build systems to verify that even with
+incrementality and non-determinism, their build systems produce results that are
+equivalent to clean builds.
+
+> How do these design decisions affect the developer experience?
+> Are there any similarly interesting points in the design space for more
+> build+package systems
+> Are there any interesting insights from the ML Compilation Manager?
+
+These questions are outside the scope of this paper, but are on our list of
+further research.
 
@@
 
-> the code of each build system is quite different from each other
+> this paper seems to suppose that the toplevel target is single and the changes
+> are passively propagated. If this is true, this must be mentioned earlier.
 
-The framework does provide reusable "schedulers" (topological, recursive, etc.)
-And in the final version we will also provide reusable "rebuilders".
+Excel does not have the toplevel target, but one could introduce it by adding a
+cell that depends on all other cells. We will clarify the correspondence between
+real Excel and our model.
 
-> Build abstraction is trivial
+> It would be more interesting if the framework could provide a set of building
+> blocks of build systems. In the present form, the code of each build system is
+> quite different from each other
+
+The framework does provide reusable 'schedulers' (topological, recursive, etc.),
+which are reused in the definitions of build systems. One can also factor out
+'rebuilders', e.g. based on verifying/constructive traces. We will address this
+in the revision.
+
+> the Build abstraction itself is trivial and therefore does not help us
+> understand some interesting aspects of the build systems.
 
@@
 substantial engineering aspects of build systems and bring out more subtle
-design choices.
-
-Parallelism is an example.
+design choices. The Build abstraction is simple, but not too simple: it does
+allow us to distinguish between many existing build systems in a meaningful way.
+Adding more details to the Build abstraction, e.g. introducing parallelism,
+would obscure the essence.

### 2018-05-03 paper/response.md
Update response.md

@@
 We thank all reviewers for their comments and suggestions and will address them
-in the revision. We also respond to some of them below.
+in the revision. We also respond to some of them below, starting with the most
+important.
 
-# Reviewer A
+Referee B (an expert) makes a single criticism of our paper:
+
+> The emphasis of the minimality property over correctness seems somewhat off.
+
+We agree that correctness is more important, but respectfully *disagree* that our 
+emphasis is on minimality. On the contrary, a larger part
+of the paper is dedicated to the precise formulation of what it means for a
+build system to be correct -- see S3.6 (correctness), S6.3 (correctness in
+presence of non-determinism) and S6.4 (correctness of shallow cloud builds).
+
+> The property most build system authors chase after is that an incremental
+> build should be equivalent to a clean build, up to non-determinism - if there
+> is any in your build!
+
+We fully agree that correctness must take precedence. Our definition of
+correctness allows the designers of build systems to verify that even with
+incrementality and non-determinism, their build systems produce results that are
+equivalent to clean builds.
+
+*** Simon says: what does this para mean?  Are you saying "yes, your statement
+"an incremental build should be equivalent to a clean build" is what correctness
+means, and is what our correctness property states?  Or what? ***
+
+# Other more detailed points
+
+## Reviewer A
 
@@
 
-# Reviewer B
+## Reviewer B
 
-> The emphasis of the minimality property over correctness seems somewhat off.
+> most large projects are today moving towards reproducible/deterministic builds.
 
-We disagree that our emphasis is on minimality. On the contrary, a larger part
-of the paper is dedicated to the precise formulation of what is means for a
-build system to be correct -- see S3.6 (correctness), S6.3 (correctness in
-presence of non-determinism) and S6.4 (correctness of shallow cloud builds).
-
-> The property most build system authors chase after is that an incremental
-> build should be equivalent to a clean build, up to non-determinism - if there
-> is any in your build!
-
-We fully agree that correctness must take precedence. Our definition of
-correctness allows the designers of build systems to verify that even with
-incrementality and non-determinism, their build systems produce results that are
-equivalent to clean builds.
+A strength of our abstractions is that they allow us to discuss the the tradeoffs
+involved in requiring full determinism.  We did not do so in the submission, but
+we will address the determinism/non-determinism tradeoff in our revision.
 
 > How do these design decisions affect the developer experience?
+
+Our abstractions are, by design, rather high level.  We identify and study a few key
+(and rather subtle) features, while carefully abstracting away from the details
+(such as exploiting parallelism, caching, and so on) .  
+But for the developer experience, those engineering details are super-important,
+and they absorb 90%+ of the code that implements a build system.
+In short, our abstractions allow us to study, classify, compare, and contrast
+the landscape of build systems, but they abstract away too much to say much
+about the developer experience.
+
 > Are there any similarly interesting points in the design space for more
-> build+package systems
+> build+package systems?
+
+Packaging systems like cabal or npm bring in a huge new collection of
+considerations (such as constraint-solving to find consistent package sets)
+which are way beyond our scope here.  Perhaps there is another similar paper
+to be written about packaging systems!
+
 > Are there any interesting insights from the ML Compilation Manager?
 
-These questions are outside the scope of this paper, but are on our list of
-further research.
+Many languages have in-built build systems, including ML (as you mention),
+Haskell's `ghc --make`, and OCamlMake. In the interests of space, and the
+reader's bandwidth, and breadth of impact, we chose to focus on big, corpoprate-scale,
+language-independent build sytems, rather than language-specific ones.
 
-# Reviewer C
+## Reviewer C
 
@@
 The framework does provide reusable 'schedulers' (topological, recursive, etc.),
-which are reused in the definitions of build systems. One can also factor out
+which are re-used in the definitions of build systems. One can also factor out
 'rebuilders', e.g. based on verifying/constructive traces. We will address this
@@
 substantial engineering aspects of build systems and bring out more subtle
-design choices. The Build abstraction is simple, but not too simple: it does
+design choices. The Build abstraction is simple, but not *too* simple: it does
 allow us to distinguish between many existing build systems in a meaningful way.

### 2018-05-03 paper/response.md
Update response.md

@@
 important.
+
+# Key points
 
@@
 > The property most build system authors chase after is that an incremental
-> build should be equivalent to a clean build, up to non-determinism - if there
-> is any in your build!
+> build should be equivalent to a clean build, up to non-determinism
 
-We fully agree that correctness must take precedence. Our definition of
-correctness allows the designers of build systems to verify that even with
-incrementality and non-determinism, their build systems produce results that are
-equivalent to clean builds.
+Yes indeed. That statement is a consequence of our defintion of correctness
+on page 11.   (Our statement is a bit stronger, because it does not rely
+on some earlier notion of a "clean build".   For example, presumably
+a "clean build" means a build in which all inputs are treated as out
+of date; and also that the results of that build are "correct".  But
+what does "correct" mean?  Our definition avoids these difficulties.)
 
-*** Simon says: what does this para mean?  Are you saying "yes, your statement
-"an incremental build should be equivalent to a clean build" is what correctness
-means, and is what our correctness property states?  Or what? ***
-
-# Other more detailed points
+# Details
 

### 2018-05-03 paper/response.md
Respose edits (Simon & Andrey)

@@
-We thank all reviewers for their comments and suggestions and will address them
-in the revision. We also respond to some of them below, starting with the most
-important.
+We thank the reviewers for their comments and suggestions.
 
@@
 
-Referee B (an expert) makes a single criticism of our paper:
+Referee B makes a single criticism of our paper:
 
@@
 
+> B: How do these design decisions affect the developer experience?
+> C: the Build abstraction itself is trivial and therefore does not help us understand some interesting aspects of the build systems.
+
+Our abstractions are simple, but not *too* simple. They allow us to
+identify and study a few key (but subtle) features, while carefully
+abstracting away from the details (such as exploiting parallelism,
+caching, and so on).  But for the developer experience, those
+engineering details are super-important, and they absorb 90%+ of the
+code that implements a build system.  wo while our abstractions allow
+us to study, classify, compare, and contrast the landscape of build
+systems, they abstract away too much to say much about the
+developer experience.
+
 # Details
@@
 
-> you refer to a "static" dependency on release.txt,
-> but I can't see how the build system could distinguish this
-> dependency from those arising from parsing release.txt.
+> Presumably you could call `dependencies` on monadic tasks?
+
+No, we cannot call `dependencies` on monadic tasks. Dependencies of a monadic
+task cannot be determined without providing actual values; see e.g. line 576ff.
+
+## Reviewer B
+
+> reproducible/deterministic builds.
+
+A strength of our abstractions is that they allow us to discuss the the tradeoffs
+involved in requiring full determinism.  We did not do so in the submission, but
+we will address the determinism/non-determinism tradeoff in our revision.
+
+> npm, yarn, cabal, and cargo, etc?
+
+Packaging systems like cabal or npm bring in a huge new collection of
+considerations (such as constraint-solving to find consistent package sets)
+which are way beyond our scope here.  Perhaps there is another similar paper
+to be written about packaging systems!
+
+> ML Compilation Manager?
+
+Many languages have in-built build systems, including ML (as you mention),
+Haskell's `ghc --make`, and OCamlMake. In the interests of space, and the
+reader's bandwidth, and breadth of impact, we chose to focus on big, corpoprate-scale,
+language-independent build systems, rather than language-specific ones.
+
+## Reviewer C
+
+> In the present form, the code of each build system is quite different from each other
+
+In fact, our framework does provide reusable 'schedulers' (topological, recursive, etc.),
+which are re-used in the definitions of build systems. One can also factor out
+'rebuilders', e.g. based on verifying/constructive traces. We will address this
+in the revision.
+
+-------------------
+> I can't see how the build system could distinguish this
+> [static] dependency from those arising from parsing release.txt.
 
@@
 
-> Presumably you could call `dependencies` on monadic tasks --- `Monad`s are
-> `Applicative` after all --- but just wouldn't get any useful information back?
-
-No, we cannot call `dependencies` on monadic tasks. Dependencies of a monadic
-task cannot be determined without providing actual values: for `sprsh2`, for
-example, we need to know the value of the cell `C1` to determine the exact set
-of dependencies; `track` obtains values using the `fetch :: k -> m v` argument.
-
-## Reviewer B
-
-> most large projects are today moving towards reproducible/deterministic builds.
-
-A strength of our abstractions is that they allow us to discuss the the tradeoffs
-involved in requiring full determinism.  We did not do so in the submission, but
-we will address the determinism/non-determinism tradeoff in our revision.
-
-> How do these design decisions affect the developer experience?
-
-Our abstractions are, by design, rather high level.  We identify and study a few key
-(and rather subtle) features, while carefully abstracting away from the details
-(such as exploiting parallelism, caching, and so on) .  
-But for the developer experience, those engineering details are super-important,
-and they absorb 90%+ of the code that implements a build system.
-In short, our abstractions allow us to study, classify, compare, and contrast
-the landscape of build systems, but they abstract away too much to say much
-about the developer experience.
-
-> Are there any similarly interesting points in the design space for more
-> build+package systems?
-
-Packaging systems like cabal or npm bring in a huge new collection of
-considerations (such as constraint-solving to find consistent package sets)
-which are way beyond our scope here.  Perhaps there is another similar paper
-to be written about packaging systems!
-
-> Are there any interesting insights from the ML Compilation Manager?
-
-Many languages have in-built build systems, including ML (as you mention),
-Haskell's `ghc --make`, and OCamlMake. In the interests of space, and the
-reader's bandwidth, and breadth of impact, we chose to focus on big, corpoprate-scale,
-language-independent build sytems, rather than language-specific ones.
-
-## Reviewer C
-
 > this paper seems to suppose that the toplevel target is single and the changes
@@
 
-> It would be more interesting if the framework could provide a set of building
-> blocks of build systems. In the present form, the code of each build system is
-> quite different from each other
-
-The framework does provide reusable 'schedulers' (topological, recursive, etc.),
-which are re-used in the definitions of build systems. One can also factor out
-'rebuilders', e.g. based on verifying/constructive traces. We will address this
-in the revision.
-
-> the Build abstraction itself is trivial and therefore does not help us
-> understand some interesting aspects of the build systems.
-
-The whole point of the paper is to ignore many crucially important and
-substantial engineering aspects of build systems and bring out more subtle
-design choices. The Build abstraction is simple, but not *too* simple: it does
-allow us to distinguish between many existing build systems in a meaningful way.
-Adding more details to the Build abstraction, e.g. introducing parallelism,
-would obscure the essence.
+: for `sprsh2`, for
+example, we need to know the value of the cell `C1` to determine the exact set
+of dependencies; `track` obtains values using the `fetch :: k -> m v` argument.

### 2018-05-03 paper/1-intro.tex
Small redrafts AWF

@@
 focus of attention.
-Rarely do people ask questions like ``What does it mean for my build
-system to be correct?'' or ``What are the trade-offs between different
-approaches?''.  Moreover, complex build systems use subtle algorithms, but they
+% Rarely do people ask questions like ``What does it mean for my build
+% system to be correct?'' or ``What are the trade-offs between different
+% approaches?''.  Moreover,
+Complex build systems use subtle algorithms, but they
 are often hidden away, and not the object of study.

### 2018-05-03 paper/response.md
Work on response

@@
 
-Referee B makes a single criticism of our paper:
+> **B:** The emphasis of the minimality property over correctness seems somewhat
+off.
 
-> The emphasis of the minimality property over correctness seems somewhat off.
+We agree that correctness is more important, but respectfully *disagree* that
+our emphasis is on minimality. On the contrary, a larger part of the paper is
+dedicated to the precise formulation of build system correctness (S3.6),
+including non-deterministic (S6.3) and shallow build (S6.4) cases.
 
-We agree that correctness is more important, but respectfully *disagree* that our
-emphasis is on minimality. On the contrary, a larger part
-of the paper is dedicated to the precise formulation of what it means for a
-build system to be correct -- see S3.6 (correctness), S6.3 (correctness in
-presence of non-determinism) and S6.4 (correctness of shallow cloud builds).
+> **B:** The property most build system authors chase after is that an
+incremental build should be equivalent to a clean build, up to non-determinism
 
-> The property most build system authors chase after is that an incremental
-> build should be equivalent to a clean build, up to non-determinism
+Yes indeed. Our definition of correctness is stronger than "equivalent to a
+clean build", because it does not rely on some earlier notion of a "clean build".
+Presumably a "clean build" means a build in which all inputs are treated as
+out-of-date; and also that the results of that build are "correct". But what
+does "correct" mean? Our definition is self-contained.
 
-Yes indeed. That statement is a consequence of our defintion of correctness
-on page 11.   (Our statement is a bit stronger, because it does not rely
-on some earlier notion of a "clean build".   For example, presumably
-a "clean build" means a build in which all inputs are treated as out
-of date; and also that the results of that build are "correct".  But
-what does "correct" mean?  Our definition avoids these difficulties.)
+> **B:** How do these design decisions affect the developer experience?
 
-> B: How do these design decisions affect the developer experience?
-> C: the Build abstraction itself is trivial and therefore does not help us understand some interesting aspects of the build systems.
+> **C:** the Build abstraction itself is trivial and therefore does not help
+us understand some interesting aspects
 
-Our abstractions are simple, but not *too* simple. They allow us to
-identify and study a few key (but subtle) features, while carefully
-abstracting away from the details (such as exploiting parallelism,
-caching, and so on).  But for the developer experience, those
-engineering details are super-important, and they absorb 90%+ of the
-code that implements a build system.  wo while our abstractions allow
-us to study, classify, compare, and contrast the landscape of build
-systems, they abstract away too much to say much about the
-developer experience.
+Our abstractions are simple, but not *too* simple. They allow us to identify and
+study a few key (but subtle) features, while carefully abstracting away from the
+details (such as exploiting parallelism, caching, and so on). But for the
+developer experience, those engineering details are super-important, and they
+absorb 90%+ of the code that implements a build system. So while our
+abstractions allow us to study, classify and compare build systems, they
+abstract away too much to say much about the developer experience.
 
@@
 
-## Reviewer A
+> **A:** I can't see how the build system could distinguish this
+> [static] dependency from those arising from parsing `release.txt`.
 
-> Presumably you could call `dependencies` on monadic tasks?
+Indeed, Shake does not distinguish static dependencies from dynamic: it treats
+them uniformly as if they all were dynamic.
 
-No, we cannot call `dependencies` on monadic tasks. Dependencies of a monadic
-task cannot be determined without providing actual values; see e.g. line 576ff.
+> **A:** Presumably you could call `dependencies` on monadic tasks?
 
-## Reviewer B
+No. Dependencies of a monadic task cannot be determined without providing actual
+values; see line 577.
 
-> reproducible/deterministic builds.
+> **B:** reproducible/deterministic builds
 
-A strength of our abstractions is that they allow us to discuss the the tradeoffs
-involved in requiring full determinism.  We did not do so in the submission, but
-we will address the determinism/non-determinism tradeoff in our revision.
+A strength of our abstractions is that they allow us to discuss the tradeoffs
+involved in requiring full determinism. We will add such a discussion in our
+revision.
 
-> npm, yarn, cabal, and cargo, etc?
+> **B**: npm, yarn, cabal, and cargo, etc?
 
-Packaging systems like cabal or npm bring in a huge new collection of
-considerations (such as constraint-solving to find consistent package sets)
-which are way beyond our scope here.  Perhaps there is another similar paper
-to be written about packaging systems!
+Packaging systems bring in a new collection of considerations (such as
+constraint-solving to find consistent package sets) which are beyond our scope
+here. Perhaps there is another similar paper to be written about packaging systems!
 
-> ML Compilation Manager?
+> **B:** ML Compilation Manager?
 
-Many languages have in-built build systems, including ML (as you mention),
-Haskell's `ghc --make`, and OCamlMake. In the interests of space, and the
-reader's bandwidth, and breadth of impact, we chose to focus on big, corpoprate-scale,
-language-independent build systems, rather than language-specific ones.
+Many languages have built-in build systems, including ML (as you mention),
+Haskell's `ghc --make`, and OCamlMake. In the interests of space, we chose to
+focus on language-independent build systems, rather than language-specific ones.
 
-## Reviewer C
+> **C:** In the present form, the code of each build system is quite different
+from each other
 
-> In the present form, the code of each build system is quite different from each other
-
-In fact, our framework does provide reusable 'schedulers' (topological, recursive, etc.),
-which are re-used in the definitions of build systems. One can also factor out
+In fact, our framework does provide reusable 'schedulers' (`topological`,
+`recursive`, etc.) and defines build systems using them. One can also factor out
 'rebuilders', e.g. based on verifying/constructive traces. We will address this
@@
 
--------------------
-> I can't see how the build system could distinguish this
-> [static] dependency from those arising from parsing release.txt.
-
-Indeed, Shake does not distinguish static dependencies from dynamic: it
-treats them uniformly as if they all were dynamic.
-
+The following didn't fit the limit:
+-----------------------------------
 > how does the file system metadata fit into this model? Is it persistent
@@
 metadata, but we decided against this, since it makes the model more complex.
-
-> this paper seems to suppose that the toplevel target is single and the changes
-> are passively propagated. If this is true, this must be mentioned earlier.
-
-Excel does not have the toplevel target, but one could introduce it by adding a
-cell that depends on all other cells. We will clarify the correspondence between
-real Excel and our model.
-
-: for `sprsh2`, for
-example, we need to know the value of the cell `C1` to determine the exact set
-of dependencies; `track` obtains values using the `fetch :: k -> m v` argument.

### 2018-05-03 paper/2-background.tex

@@
 referred to as \emph{build rules}) and their dependencies in a simple textual form.
-For example:
+For example\footnote{Do not worry about the dependency on \cmd{stdio.h}, and its transitive (perhaps dynamic) dependencies.  We will deal with these in Section \simon{xxx}.}:
 
@@
     need files
-    system "tar" $ ["-cf", "result.tar"] ++ files
+    system "tar" $ ["-cf", "release.tar"] ++ files
 \end{minted}

### 2018-05-04 paper/3-abstractions.tex
Fix paper build

@@
 -- An abstract store
-data Store i k v
+data Store i k v -- i = info, k = key, v = value
 getInfo    :: Store i k v -> i

### 2018-06-11 src/Build.hs
Minor changes for the paper

@@ -29,7 +29,7 @@ correctBuild tasks store result = all correct . reachable deps
     deps = maybe [] (\task -> snd $ track task (flip getValue result)) . tasks
     correct k = case tasks k of
         Nothing   -> getValue k result == getValue k store
-        Just task -> getValue k result == compute task (flip getValue result)
+        Just task -> getValue k result == compute task result
 
 -- | Given a @build@ and @tasks@, check that @build@ produces a correct result
 -- for any initial store and a target key.

### 2018-06-11 src/Build/Task/Monad.hs

@@ -3,7 +3,9 @@
 -- | Monadic tasks, as used by Excel, Shake and other build systems.
 -- Dependencies of monadic tasks can only be discovered dynamically, i.e. during
 -- their execution.
-module Build.Task.Monad (track, trackM, isInput, compute, partial, exceptional) where
+module Build.Task.Monad (
+    track, trackM, isInput, computePure, compute, partial, exceptional
+    ) where
 
 import Control.Monad.Trans
 import Control.Monad.Trans.Except
@@ -12,6 +14,7 @@ import Control.Monad.Writer
 import Data.Functor.Identity
 import Data.Maybe
 
+import Build.Store
 import Build.Task
 
 -- | Execute a monadic task on a pure store @k -> v@, tracking the dependencies.
@@ -31,8 +34,12 @@ isInput :: Tasks Monad k v -> k -> Bool
 isInput tasks = isNothing . tasks
 
 -- | Run a task with a pure lookup function.
-compute :: Task Monad k v -> (k -> v) -> v
-compute task store = runIdentity $ run task (Identity . store)
+computePure :: Task Monad k v -> (k -> v) -> v
+computePure task store = runIdentity $ run task (Identity . store)
+
+-- | Run a task in a given store.
+compute :: Task Monad k v -> Store i k v -> v
+compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
 
 -- | Convert a task with a total lookup function @k -> m v@ into a task with a
 -- partial lookup function @k -> m (Maybe v)@. This essentially lifts the task

### 2018-06-11 test/Examples.hs

@@ -72,10 +72,12 @@ sprsh1 "B2" = Just $ Task $ \fetch -> ((*2) <$> fetch "B1")
 sprsh1 _    = Nothing
 
 sprsh2 :: Tasks Monad String Integer
-sprsh2 "B1" = Just $ Task $ \fetch -> do c1 <- fetch "C1"
-                                         if c1 == 1 then fetch "B2" else fetch "A2"
-sprsh2 "B2" = Just $ Task $ \fetch -> do c1 <- fetch "C1"
-                                         if c1 == 1 then fetch "A1" else fetch "B1"
+sprsh2 "B1" = Just $ Task $ \fetch -> do
+    c1 <- fetch "C1"
+    if c1 == 1 then fetch "B2" else fetch "A2"
+sprsh2 "B2" = Just $ Task $ \fetch -> do
+    c1 <- fetch "C1"
+    if c1 == 1 then fetch "A1" else fetch "B1"
 sprsh2 _ = Nothing
 
 sprsh3 :: Tasks Alternative String Integer

### 2018-06-11 paper/3-abstractions.tex
Work on S3

@@
 
+% Old type:
+% type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+
 Our first main abstraction is for \emph{task descriptions}:
 \begin{minted}[xleftmargin=10pt]{haskell}
-type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
+type    Tasks c k v = k -> Maybe (Task c k v)
 \end{minted}
-This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
-\emph{profunctor optics} might recognise a familiar pattern. We discuss this
-in~\S\ref{sec-related-optics}.} is best introduced by an example.
-Consider this \Excel spreadsheet:
+
+\noindent
+Here \hs{Task} describes a single build task, while \hs{Tasks} associates a
+task description to every non-input key; input keys are associated with
+\hs{Nothing}. The highly-abstracted type\footnote{Readers familiar with
+\emph{lenses} or \emph{profunctor optics} might recognise a familiar pattern.
+We discuss this in~\S\ref{sec-related-optics}.} \hs{Task} is best introduced by
+an example. Consider this \Excel spreadsheet:
+
 \vspace{1mm}
@@
 \vspace{1mm}
+
+\noindent
 Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains
@@
 with the following task description:
+
 \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh1 :: Task Applicative String Integer
-sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
-sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
-sprsh1 _     _    = Nothing
+sprsh1 :: Tasks Applicative String Integer
+sprsh1 "B1" = Just $ Task $ \fetch -> ((+)  <$> fetch "A1" <*> fetch "A2")
+sprsh1 "B2" = Just $ Task $ \fetch -> ((*2) <$> fetch "B1")
+sprsh1 _    = Nothing
 \end{minted}
 \vspace{1mm}
+
+\noindent
 We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
-(Real spreadsheet cells would contain a wider range of values, of course.)
-The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
-but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
-\emph{callback} \hs{fetch} and a key. It pattern-matches on the key to see if it
-has a task description (in the \Excel case, a formula) for it. If not, it returns
-\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
-it computes the value of the formula, using \hs{fetch} to find the value of any
-keys on which it depends.
+(Real spreadsheet cells would contain a wider range of values, of course.) The
+task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
+but not the input values. It pattern-matches on the key to see if it has a task
+description (in the \Excel case, a formula) for it. If not, it returns
+\hs{Nothing}, indicating that the key is an input. If there is a formula in the
+cell, it returns the \hs{Task} to compute the value of the formula. Every
+\hs{Task} is given a \emph{callback} \hs{fetch} to find the value of any keys on
+which it depends.
 
-The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
-because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
--- the relevant type signatures are given in Fig.~\ref{fig-types}. We will
-explain why in subsection~\S\ref{sec-general-build}.
+The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit
+mysterious because it takes place in an \hs{Applicative}
+computation~\cite{mcbride2008applicative} -- the relevant type signatures are
+given in Fig.~\ref{fig-types}. We will explain why in~\S\ref{sec-general-build}.
 
-For now, we content ourselves with observing that a task description,
-of type \hs{Task}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
+For now, we content ourselves with observing that a task description, of type
+\hs{Tasks}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
 compilers, calculation chains, file systems, caches, and all other
@@
 Next comes our second main abstraction -- a build system:
+
 \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
+type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
 \end{minted}
 \vspace{1mm}
+
+\noindent
 The signature is very straightforward. Given a task description, a target key,
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-busy :: Eq k => Build Monad () k v
-busy task key store = execState (fetch key) store
+busy :: Eq k => Build Applicative () k v
+busy tasks key store = execState (fetch key) store
   where
     fetch :: k -> State (Store () k v) v
-    fetch k = case task fetch k of
+    fetch k = case tasks k of
         Nothing   -> gets (getValue k)
-        Just act -> do v <- act; modify (putValue k v); return v
+        Just task -> do v <- run task fetch; modify (putValue k v); return v
 \end{minted}
@@
 Fig.~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
-To bring a key up to date, \hs{fetch} asks the task description \hs{task} how
-to compute the value of \hs{k}. If \hs{task} returns \hs{Nothing} the key is an
+To bring a key up to date, \hs{fetch} asks the task description \hs{tasks} how
+to compute the value of \hs{k}. If \hs{tasks} returns \hs{Nothing} the key is an
 input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
-runs the action \hs{act} returned by the \hs{task} to produce a resulting
-value~\hs{v}, records the new key/value mapping in the store, and returns \hs{v}.
-Notice that \hs{fetch} passes itself to \hs{task} as an argument, so that the
-latter can use \hs{fetch} to recursively find the values of \hs{k}'s dependencies.
+runs the obtained \hs{task} to produce a resulting value~\hs{v}, records the new
+key/value mapping in the store, and returns \hs{v}. Notice that \hs{fetch}
+passes itself to \hs{task} as an argument, so that the latter can use \hs{fetch}
+to recursively find the values of \hs{k}'s dependencies.
 
@@
 @\ghci@ result = busy sprsh1 "B2" store
-@\ghci@ getValue result "B1"
+@\ghci@ getValue "B1" result
 30
-@\ghci@ getValue result "B2"
+@\ghci@ getValue "B2" result
 60
@@
 The previous example shows why the \hs{Task} abstraction is polymorphic in \hs{f}, recall the definition:
+
 \begin{minted}[xleftmargin=10pt]{haskell}
-type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+newtype Task c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
 \end{minted}
+
+\noindent
 The \hs{busy} build system instantiates \hs{f} to
@@
 So the type \hs{Task} emerges naturally, almost inevitably.  But now that
-it \emph{has} emerged, we find the that constraints \hs{c} classify task descriptions
+it \emph{has} emerged, we find that constraints \hs{c} classify task descriptions
 in a very interesting way:
@@
   value of another cell \cmd{D}.
+%TODO: Add a reference for Docker?
 \item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
   even use the application operator \hs{<*>}, which limits dependencies to a
-  single linear chain. It is interesting to note that, when run on a
-  \hs{Task}~\hs{Functor}, the \hs{busy} build system will build each key at most
-  once, thus partially fulfilling the minimality requirement~\ref{def-minimal}.
-  Alas, it still has no mechanism to decide which input keys changed since the
-  previous build.
+  single linear chain, as e.g. in the Docker build system. It is interesting to
+  note that, when run on such a task description, the \hs{busy} build system
+  will build each key at most once, thus partially fulfilling the minimality
+  requirement~\ref{def-minimal}. Alas, it still has no mechanism to decide which
+  input keys changed since the previous build.
 \item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
@@
   non-determinism, as discussed in~\S\ref{sec-non-determinism}.
+\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
+  in~\S\ref{sec-implementations} to describe tasks that have read and write
+  access to the persistently stored build information~\hs{i}.
 \end{itemize}
 
-Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
-argument, an application of \hs{busy} to a \hs{Task}~\hs{Functor} or
-a \hs{Task}~\hs{Applicative} will typecheck and run just fine. It feels a bit like
-sub-typing, but is actually just ordinary higher-rank polymorphism at
-work~\cite{jones2007practical}.
+% TODO: Shall we describe the lifTask and liftTasks combinators? They aren't pretty.
+%
+%   liftTask :: Task Applicative k v -> Task Monad k v
+%   liftTask (Task task) = Task task
+%
+%   liftTasks :: Tasks Applicative k v -> Tasks Monad k v
+%   liftTasks = fmap (fmap liftTask)
+%
+% Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
+% argument, an application of \hs{busy} to a \hs{Task}~\hs{Functor} or
+% a \hs{Task}~\hs{Applicative} will typecheck and run just fine. It feels a bit like
+% sub-typing, but is actually just ordinary higher-rank polymorphism at
+% work~\cite{jones2007practical}.
 
@@
 
-As explained in~\S\ref{sec-background-excel}, some task descriptions have dynamic
-dependencies, which are determined by values of intermediate computations. In
-our framework, such task descriptions correspond to the type
-\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Consider this spreadsheet example:
+As explained in~\S\ref{sec-background-excel}, some task descriptions have
+dynamic dependencies, which are determined by values of intermediate
+computations. In our framework, such task descriptions correspond to the type
+\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{v}. Consider this spreadsheet example:
 
@@
 \noindent
-Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel (which
-uses dynamic dependencies) is perfectly happy. We can express this spreadsheet
-using our task abstraction as:
+Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel
+(which uses dynamic dependencies) is perfectly happy. We can express this
+spreadsheet using our task abstraction as:
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh2 :: Task Monad String Integer
-sprsh2 fetch "B1" = Just $ do c1 <- fetch "C1"
+sprsh2 :: Tasks Monad String Integer
+sprsh2 "B1" = Just $ Task $ \fetch -> do
+    c1 <- fetch "C1"
     if c1 == 1 then fetch "B2" else fetch "A2"
-sprsh2 fetch "B2" = Just $ do c1 <- fetch "C1"
+sprsh2 "B2" = Just $ Task $ \fetch -> do
+    c1 <- fetch "C1"
     if c1 == 1 then fetch "A1" else fetch "B1"
-sprsh2 _     _    = Nothing
+sprsh2 _ = Nothing
 \end{minted}
@@
 any dependencies}:
+
 \begin{minted}[xleftmargin=5pt]{haskell}
-compute :: Task Monad k v -> Store i k v -> k -> Maybe v
-compute task store = fmap runIdentity@\,@.@\,\blk{task}@ (\k -> Identity (getValue k store))
+compute :: Task Monad k v -> Store i k v -> v
+compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
 \end{minted}
@@
 monad: the function
-\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)}
-has the type \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a
-\hs{task}. The result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now
-need to get rid of the \hs{Identity} wrapper by applying \hs{runIdentity} to the
-contents of \hs{Maybe}.
+\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
+\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
+result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.
 
 \vspace{-1mm}
-\definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
+\definition[Correctness]{Suppose \hs{build} is a build system, \hs{tasks} is a
 build task description, \hs{key} is a target key, \hs{store} is an initial
 store, and \hs{result} is the store produced by running the build system with
-parameters \hs{task}, \hs{key} and \hs{store}. Or, using the precise language of
+parameters \hs{tasks}, \hs{key} and \hs{store}. Or, using the precise language of
 our abstractions:
@@
 build         :: Build c i k v
-task          :: Task c k v
+tasks         :: Tasks c k v
 key           :: k
 store, @@result :: Store i k v
-result = @@build @@task @@key @@store
+result = @@build @@tasks @@key @@store
 \end{minted}
@@
 
-    \item The \hs{result} is \emph{consistent} with the \hs{task}, i.e.
+    \item The \hs{result} is \emph{consistent} with the \hs{tasks}, i.e.
     for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
-    \hs{task} matches the value stored in the \hs{result}:
+    corresponding \hs{task} matches the value stored in the \hs{result}:
     \[
-    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{@\std{fromJust}@}~\hs{(}\hs{compute}~\hs{task}~\hs{result}~\hs{k)}
+    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
     \]
@@
 A build system is \emph{correct} if it produces a correct \hs{result} for any
-given \hs{task}, \hs{key} and \hs{store}.
+given \hs{tasks}, \hs{key} and \hs{store}.
 }
@@
 includes the \hs{busy} build system introduced earlier: it will loop
-indefinitely given a cyclic \hs{task}. Some build systems provide a limited
+indefinitely given a cyclic \hs{tasks}. Some build systems provide a limited
 support for cyclic tasks, see~\S\ref{sec-iterative-compute}.

### 2018-06-11 paper/3-abstractions.tex
Finish S3

@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-dependencies :: Task Applicative k v -> k -> [k]
-dependencies task key = case task (\k -> Const [k]) key of
-                            Nothing         -> []
-                            Just (Const ks) -> ks
+dependencies :: Task Applicative k v -> [k]
+dependencies task = getConst $ run task (\k -> Const [k])
 \end{minted}
@@
 
+\noindent
 Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ dependencies sprsh1 "A1"
+@\ghci@ dependencies $ fromJust $ sprsh1 "A1"
 [@@]
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ dependencies sprsh1 "B1"
+@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
 ["A1", "A2"]
@@
 dependencies need to be tracked too. This cannot be done statically; notice that
-the application of the function \hs{dependencies} to a \hs{Task}~\hs{Monad} will
-not typecheck. We need to run a monadic task on a store with concrete values,
-which will determine the discovered dependencies. Accordingly, we introduce
-the function \hs{track}: a combination of \hs{compute} and \hs{dependencies}
-that computes both the resulting value and the list of its dependencies in an
-arbitrary monadic context~\hs{m}:
+we cannot apply the function \hs{dependencies} to a \hs{Task}~\hs{Monad} because
+the \hs{Const} functor has no \hs{Monad} instance. We need to run a monadic task
+on a store with concrete values, which will determine the discovered
+dependencies. Accordingly, we introduce the function \hs{track}: a combination
+of \hs{compute} and \hs{dependencies} that computes both the resulting value and
+the list of its dependencies in an arbitrary monadic context~\hs{m}:
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-track :: Monad m => Task Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
-track task fetch = fmap runWriterT . task trackingFetch
+track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [k])
+track task fetch = runWriterT $ run task trackingFetch
   where
@@
 corresponding key. The \hs{task} returns the value of type
-\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
-\hs{runWriterT} to the contents of \hs{Maybe}.
-Below we give an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
+\hs{WriterT}~\hs{[@@k]}~\hs{m}~\hs{v}, which we unwrap by \hs{runWriterT}. Here
+is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
 
@@
 @\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
-@\ghci@ fromJust $ track sprsh2 fetchIO "B1"
+@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
 C1: 1
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ fromJust $ track sprsh2 fetchIO "B1"
+@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
 C1: 2

### 2018-06-12 src/Build/Rebuilder.hs
Minor tweaks

@@ -5,7 +5,7 @@
 module Build.Rebuilder (
     Rebuilder, unliftRebuilder, perpetualRebuilder,
     modTimeRebuilder, Time, MakeInfo,
-    approximationRebuilder, DependencyApproximation, ApproximationInfo,
+    approximateRebuilder, ApproximateDependencies, ApproximationInfo,
     vtRebuilder, stRebuilder, ctRebuilder, dctRebuilder
     ) where
 
@@ -37,35 +37,35 @@ perpetualRebuilder _key _value task = Task $ run task
 
 ------------------------------------- Make -------------------------------------
 type Time = Integer
-type MakeInfo k = (Map k Time, Time)
+type MakeInfo k = (Time, Map k Time)
 
 -- | This rebuilder uses modification time to decide whether a key is dirty and
 -- needs to be rebuilt. Used by Make.
 modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
 modTimeRebuilder key value task = Task $ \fetch -> do
-    (modTime, now) <- get
+    (now, modTime) <- get
     let dirty = case Map.lookup key modTime of
             Nothing -> True
             time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
     if not dirty
     then return value
     else do
-        put (Map.insert key now modTime, now + 1)
+        put (now + 1, Map.insert key now modTime)
         run task fetch
 
---------------------------- Dependency approximation ---------------------------
+--------------------------- Approximate dependencies ---------------------------
 -- | If there is an entry for a key, it is an conservative approximation of its
 -- dependencies. Otherwise, we have no reasonable approximation and assume the
 -- key is always dirty (e.g. it uses an INDIRECT reference).
-type DependencyApproximation k = Map k [k]
+type ApproximateDependencies k = Map k [k]
 
 -- | A set of dirty keys and information about dependencies.
-type ApproximationInfo k = (Set k, DependencyApproximation k)
+type ApproximationInfo k = (Set k, ApproximateDependencies k)
 
 -- | This rebuilders uses approximate dependencies to decide whether a key
 -- needs to be rebuilt. Used by Excel.
-approximationRebuilder :: Ord k => Rebuilder Monad (ApproximationInfo k) k v
-approximationRebuilder key value task = Task $ \fetch -> do
+approximateRebuilder :: Ord k => Rebuilder Monad (ApproximationInfo k) k v
+approximateRebuilder key value task = Task $ \fetch -> do
[…]

### 2018-06-12 src/Build/Scheduler.hs

@@ -3,8 +3,8 @@
 -- | Build schedulers execute task rebuilders in the right order.
 module Build.Scheduler (
     topological,
-    reordering, Chain,
-    restarting,
+    restarting, Chain,
+    restartingB,
     recursive,
     independent
     ) where
@@ -15,13 +15,13 @@ import Data.Set (Set)
 
 import Build
 import Build.Task
+import Build.Task.Applicative
 import Build.Task.Monad
 import Build.Store
 import Build.Rebuilder
 import Build.Utilities
 
 import qualified Data.Set               as Set
-import qualified Build.Task.Applicative as A
 
 -- | Update the value of a key in the store. The function takes both the current
 -- value (the first parameter of type @v@) and the new value (the second
@@ -35,21 +35,21 @@ updateValue key _value newValue = putValue key newValue
 -- | This scheduler constructs the dependency graph of the target key by
 -- extracting all (static) dependencies upfront, and then traversing the graph
 -- in the topological order, rebuilding keys using the supplied rebuilder.
-topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
-topological rebuilder tasks key = execState $ forM_ chain $ \k ->
-    case tasks k of
-        Nothing   -> return ()
-        Just task -> do
-            value <- gets (getValue k)
-            let newTask = rebuilder k value task
-                newFetch :: k -> StateT i (State (Store i k v)) v
-                newFetch = lift . gets . getValue
-            info <- gets getInfo
-            (newValue, newInfo) <- runStateT (run newTask newFetch) info
-            modify $ putInfo newInfo . updateValue k value newValue
+topological :: forall i k v. Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
+topological rebuilder tasks key = execState $ forM_ chain $ \k -> case tasks k of
+    Nothing   -> return ()
+    Just task -> do
+        value <- gets (getValue k)
+        let newTask :: Task (MonadState i) k v
+            newTask = rebuilder k value task
+            newFetch :: k -> StateT i (State (Store i k v)) v
+            newFetch = lift . gets . getValue
+        info <- gets getInfo
+        (newValue, newInfo) <- runStateT (run newTask newFetch) info
+        modify $ putInfo newInfo . updateValue k value newValue
   where
-    deps  = maybe [] A.dependencies . tasks
-    chain = case topSort (graph deps key) of
+    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
[…]

### 2018-06-12 src/Build/System.hs

@@ -45,7 +45,7 @@ memo = recursive perpetualRebuilder
 
 -- | A model of Make: an applicative build system that uses file modification
 -- times to check if a key is up to date.
-make :: forall k v. Ord k => Build Applicative (MakeInfo k) k v
+make :: Ord k => Build Applicative (MakeInfo k) k v
 make = topological modTimeRebuilder
 
 -- | A model of Ninja: an applicative build system that uses verifying traces
@@ -58,7 +58,7 @@ type ExcelInfo k = (ApproximationInfo k, Chain k)
 -- | A model of Excel: a monadic build system that stores the calculation chain
 -- from the previuos build and approximate dependencies.
 excel :: Ord k => Build Monad (ExcelInfo k) k v
-excel = reordering approximationRebuilder
+excel = restarting approximateRebuilder
 
 -- | A model of Shake: a monadic build system that uses verifying traces to
 -- check if a key is up to date.
@@ -70,7 +70,7 @@ shake = recursive vtRebuilder
 -- that Bazel currently does not allow users to write monadic build rules: only
 -- built-in rules have access to dynamic dependencies.
 bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
-bazel = restarting isDirtyCT ctRebuilder
+bazel = restartingB isDirtyCT ctRebuilder
 
 -- | A model of Cloud Shake: a monadic build system that uses constructive
 -- traces to check if a key is up to date as well as for caching build results.

### 2018-06-12 test/Main.hs

@@ -98,7 +98,7 @@ testSuite = and <$> sequence
     [ test  "dumb      " dumb       ()
     , test  "busy      " busy       ()
     , test  "memo      " memo       ()
-    , testA "make      " make       (Map.empty, 0)
+    , testA "make      " make       (0, Map.empty)
     , testA "ninja     " ninja      mempty
     , test  "excel     " excel      ((Set.fromList inputCells, Map.empty), mempty)
     , test  "shake     " shake      mempty

### 2018-06-12 paper/5-implementations.tex
Work on Make and Excel implementations

@@
 
-% Type signature for process. Not sure we really want this, as it's quite long
-% and is repeated in topological.
-% process :: k -> [k] -> State@\,@(Store@\,@(MakeInfo@\,\blk{k}@)@\,\blk{k}@@\,\blk{v}@)@\,\blk{v}@ -> State@\,@(Store@\,@(MakeInfo@\,\blk{k}@)@\,\blk{k}@@\,\blk{v}@)@\,@(@@)
 \begin{figure}
 \begin{minted}[fontsize=\small]{haskell}
--- Persistent build information
+-- Build information: current time and file modification times
 type Time       = Integer
-type MakeInfo k = (k -> Time, Time)
+type MakeInfo k = (Time, Map k Time)
 \end{minted}
@@
 -- Make build system
-make :: Eq k => Build Applicative (MakeInfo k) k v
-make = topological process
+make :: Ord k => Build Applicative (MakeInfo k) k v
+make = topological modTimeRebuilder
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Topological scheduler
+topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
+topological rebuilder tasks key = execState $ forM_ chain $ \k -> case tasks k of
+    Nothing   -> return ()
+    Just task -> do
+        value <- gets (getValue k)
+        let newTask :: Task (MonadState i) k v
+            newTask = rebuilder k value task
+            newFetch :: k -> StateT i (State (Store i k v)) v
+            newFetch = lift . gets . getValue
+        info <- gets getInfo
+        (newValue, newInfo) <- runStateT (run newTask newFetch) info
+        modify $ putInfo newInfo . putValue k newValue
   where
-    process key deps act = do
-        (modTime, now) <- gets getInfo
-        let dirty = or [ modTime dep > modTime key | dep <- deps ]
-        when dirty $ do
-            v <- act
-            let newModTime k = if k == key then now else modTime k
-            modify $ putInfo (newModTime, now + 1) . putValue key v
+    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
+    chain  = case topSort (reachable deps key) of
+        Nothing -> error "Cannot build tasks with cyclic dependencies"
+        Just xs -> xs
 \end{minted}
@@
 -- Standard graph algorithms (implementation omitted)
-reachable :: Eq a => (a -> [a]) ->  a  -> [a]
-topSort   :: Eq a => (a -> [a]) -> [a] -> [a]
+reachable :: Ord k => (k -> [k]) -> k -> Graph k
+topSort   :: Ord k => Graph k -> Maybe [k]
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Topological dependency strategy
-topological :: Eq k => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
-            -> Build Applicative i k v
-topological process task key = execState $ forM_ chain $ \k -> do
-    let fetch k = gets (getValue k)
-    case task fetch k of
-        Nothing  -> return ()
-        Just act -> process k (deps k) act
-  where
-    deps  = dependencies task -- dependencies is defined in @\S\ref{secdeps}@
-    chain = topSort deps (reachable deps key)
+-- A task rebuilder based on file modification times
+modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
+modTimeRebuilder key value task = Task $ \fetch -> do
+    (now, modTime) <- get
+    let dirty = case Map.lookup key modTime of
+            Nothing -> True
+            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
+    if not dirty
+    then return value
+    else do
+        put (now + 1, Map.insert key now modTime)
+        run task fetch
 \end{minted}
@@
 function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
-\hs{k}~\hs{->}~\hs{DependencyApproximation}~\hs{k} that \Excel uses to handle
+\hs{k}~\hs{->}~\hs{ApproximateDependencies}~\hs{k} that \Excel uses to handle
 dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Approximation of task dependencies
-data DependencyApproximation k = SubsetOf [k] | Unknown
+-- Approximation of task dependencies: a map of known upper bounds of dependencies
+type ApproximateDependencies k = Map k [k]
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Persistent build information
+-- Build information: dirty keys, dependency approximation, and calculation chain
 type CalcChain k = [k]
-type ExcelInfo k = ((k -> Bool, k -> DependencyApproximation k), CalcChain k)
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Result of speculative task execution
-data Result k v = MissingDependency k | Result v [k]
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Reordering dependency strategy (implementation omitted, 21 lines)
-reordering :: Ord k
-    => (k -> State (Store i k v) (Result k v) -> State (Store i k v) (Maybe (Result k v)))
-    -> Build Monad (i, CalcChain k) k v
+type ExcelInfo k = ((Set k, ApproximateDependencies k), CalcChain k)
 \end{minted}
@@
 excel :: Ord k => Build Monad (ExcelInfo k) k v
-excel = reordering process
+excel = restarting approximateRebuilder
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Restarting scheduler
+restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
+restarting rebuilder tasks key = execState $ do
+    chain    <- gets (snd . getInfo)
+    newChain <- go Set.empty $ chain ++ [key | key `notElem` chain]
+    modify . mapInfo $ \(i, _) -> (i, newChain)
   where
-    process key act = do
-        (dirty, deps) <- gets getInfo
-        let rebuild = dirty key || case deps key of SubsetOf ks -> any dirty ks
-                                                    Unknown     -> True
-        if not rebuild
-            then return Nothing
+    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
+    go _    []     = return []
+    go done (k:ks) = case tasks k of
+        Nothing -> (k :) <$> go (Set.insert k done) ks
+        Just task -> do
+            store <- get
+            let value = getValue k store
+                newTask :: Task (MonadState i) k (Either k v)
+                newTask = try $ rebuilder k value task -- `try' is defined below
+                newFetch :: k -> State i (Either k v)
+                newFetch k | k `Set.member` done = return $ Right (getValue k store)
+                           | otherwise           = return $ Left k
+            case runState (run newTask newFetch) (fst $ getInfo store) of
+                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [k]
+                (Right newValue, newInfo) -> do
+                    modify $ putInfo (newInfo, []) . updateValue k value newValue
+                    (k :) <$> go (Set.insert k done) ks
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Convert a total task into a task accepting a partial fetch callback
+try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
+try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Rebuilder based on approximate dependencies
+approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
+approximateRebuilder key value task = Task $ \fetch -> do
+    (isDirty, deps) <- get
+    let dirty = key `Set.member` isDirty || case Map.lookup key deps of
+                                                Nothing -> True
+                                                Just ks -> any (`Set.member` isDirty) ks
+    if not dirty
+    then return value
     else do
-                result <- act
-                case result of
-                    MissingDependency _ -> return ()
-                    Result v dynamicDependencies -> do
-                        let newDirty k = if k == key then True else dirty k
-                        modify $ putInfo (newDirty, deps) . putValue key v
-                return (Just result)
+        put (Set.insert key isDirty, deps)
+        run task fetch
 \end{minted}

### 2018-06-12 test/Main.hs
Fix recursive scheduler (#14)
See #13

@@ -75,8 +75,7 @@ test name build i = do
     let store   = inputs i
         result  = sequentialMultiBuild build tasks targets store
         correct = all (correctBuild tasks store result) targets
-        info    = show (getInfo result)
-    unless correct $ putStrLn $ "============\n" ++ info ++ "\n============"
+    when False $ putStrLn $ "========\n" ++ show (getInfo result) ++ "\n========"
     putStr $ name ++ " is "
     case (trim name, correct) of
         ("dumb", False) -> do putStr "incorrect, which is [OK]\n"; return True
@@ -88,8 +87,7 @@ testA name build i = do
     let store   = inputs i
         result  = sequentialMultiBuildA build tasksA targets store
         correct = all (correctBuild tasks store result) targets
-        info    = show (getInfo result)
-    unless correct $ putStrLn $ "============\n" ++ info ++ "\n============"
+    when False $ putStrLn $ "========\n" ++ show (getInfo result) ++ "\n========"
     putStrLn $ name ++ " is " ++ bool "incorrect: [FAIL]" "correct: [OK]" correct
     return correct
 

### 2018-06-13 src/Build/Scheduler.hs
Tweak recursive builder

@@ -163,15 +163,15 @@ recursive rebuilder tasks key store = fst $ execState (fetch key) (store, Set.em
             done <- gets snd
             when (key `Set.notMember` done) $ do
                 value <- gets (getValue key . fst)
-                let newTask = rebuilder key value task
+                let newTask :: Task (MonadState i) k v
+                    newTask = rebuilder key value task
                     newFetch :: k -> StateT i (State (Store i k v, Set k)) v
                     newFetch k = do v <- lift $ fetch k
-                                    i <- lift $ gets (getInfo . fst)
-                                    put i
+                                    put =<< lift (gets (getInfo . fst))
                                     return v
                 info <- gets (getInfo . fst)
                 (newValue, newInfo) <- runStateT (run newTask newFetch) info
-                modify $ \(s, done) ->
+                modify $ \(s, _) ->
                     ( putInfo newInfo $ updateValue key value newValue s
                     , Set.insert key done )
             gets (getValue key . fst)

### 2018-06-13 paper/5-implementations.tex
Work on S5

@@
 \begin{minted}[fontsize=\small]{haskell}
--- Approximation of task dependencies: a map of known upper bounds of dependencies
+-- Approximate task dependencies: a map of known upper bounds of dependencies
 type ApproximateDependencies k = Map k [k]
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Build information: dirty keys, dependency approximation, and calculation chain
+-- Build information: dirty keys, approximate dependencies, and calculation chain
 type CalcChain k = [k]
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Determine whether a trace is relevant to the current state
-traceMatch :: (Monad m, Eq k)
-    => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
-traceMatch check key ts = mapMaybeM f ts
-    where f (Trace k dkv v) = do
-                b <- return (key == k) &&^ allM (uncurry check) dkv
-                return $ if b then Just v else Nothing
+-- Shake build system
+shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
+shake = recursive vtRebuilder
 \end{minted}
@@
 -- Recursive dependency strategy
-recursive :: Eq k => (k -> (k -> State (Store i k v, [k]) v)
-                  -> State (Store i k v, [k]) (v, [k])
-                  -> State (Store i k v, [k]) ())
-                  -> Build Monad i k v
-recursive process task key store = fst $ execState (ensure key) (store, [@@])
+recursive :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
+recursive rebuilder tasks key store = fst $ execState (fetch key) (store, Set.empty)
   where
-        ensure key = do
-            let fetch k = do ensure k; gets (getValue k . fst)
+    fetch :: k -> State (Store i k v, Set k) v
+    fetch key = case tasks key of
+        Nothing -> gets (getValue key . fst)
+        Just task -> do
             done <- gets snd
-            when (key `notElem` done) $ do
-                modify $ \(s, done) -> (s, key:done)
-                case track task fetch key of -- track is defined in @\S\ref{secdeps}@
-                    Nothing -> return ()
-                    Just act -> process key fetch act
+            when (key `Set.notMember` done) $ do
+                value <- gets (getValue key . fst)
+                let newTask :: Task (MonadState i) k v
+                    newTask = rebuilder key value task
+                    newFetch :: k -> StateT i (State (Store i k v, Set k)) v
+                    newFetch k = do v <- lift $ fetch k
+                                    put =<< lift (gets (getInfo . fst))
+                                    return v
+                info <- gets (getInfo . fst)
+                (newValue, newInfo) <- runStateT (run newTask newFetch) info
+                modify $ \(s, _) ->
+                    (putInfo newInfo $ putValue key newValue s, Set.insert key done)
+            gets (getValue key . fst)
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Shake build system
-shake :: (Eq k, Hashable v) => Build Monad [Trace k v] k v
-shake = recursive $ \key fetch act -> do
-    traces <- gets (getInfo . fst)
-    poss <- traceMatch (\k v -> (==) v . hash <$> fetch k) key traces
-    current <- gets (getHash key . fst)
-    when (current `notElem` poss) $ do
-        (v, ds) <- act
-        modify $ \(s, done) ->
-            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
-            in (putInfo (t : getInfo s) (putValue key v s), done)
-
+-- Rebuilder based on verifying traces
+vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
+vtRebuilder key value task = Task $ \fetch -> do
+    vt <- get
+    dirty <- not <$> verifyVT key value (fmap hash . fetch) vt
+    if not dirty
+    then return value
+    else do
+        (newValue, deps) <- trackM task fetch
+        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
+        return newValue
 \end{minted}

### 2018-06-15 paper/5-implementations.tex
Work on S5

@@
 
-In reality \Excel works slightly differently (as far as we are able to ascertain) --
-on a change it propagates the dirty bits forward using the \hs{dynamicDependencies}, then only checks if the current \hs{key} is dirty.
-While both methods are equivalent, merely changing the interleaving, our approach allows us to model more of the behaviour of \Excel.
+In reality \Excel works slightly differently (as far as we are able to
+ascertain)~--~on a change it propagates the dirty bits forward using the
+\hs{dynamicDependencies}, then only checks if the current \hs{key} is dirty.
+While both methods are equivalent, merely changing the interleaving, our
+approach allows us to model more of the behaviour of \Excel.
 
@@
 
-The \Shake approach for dependency tracking involves recording traces and verifying them, for which we use the \hs{Trace} type defined in \ref{sec-verifying-traces}.
-Complete code is given in \S\ref{fig-shake-implementation}, split into three functions:
+The \Shake approach for dependency tracking involves recording traces and
+verifying them, for which we use the \hs{Trace} type defined
+in~\ref{sec-verifying-traces}. Complete code is given
+in~\S\ref{fig-shake-implementation}, split into three functions:
 
 \begin{description}
-\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the \hs{key} you are currently building, and a function \hs{check} which checks a specified dependency.
-It returns the \hs{result} field of all traces that match.
-Since \hs{check} is in an arbitrary monad, the function has to use \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&} functions\footnote{These functions are all available in the \cmd{extra} library on Hackage.}.
-\item[\hs{recursive}] defines the dependency ordering pattern.
-It requires a \hs{process} function that builds a key given a way to recursively build a dependency, and a way to run task to produce result.
-The main purpose of \hs{recursive} is to ensure that in a single run no key is built twice~--~for which it extends the \hs{State} monad with a list of \hs{done} keys.
-\item[\hs{shake}] ties everything together.
-First it tests if the traces allow the current state of the target key.
-If not, it builds the key and updates the store.
-The only subtlety is that Shake calls \hs{fetch} on the keys while checking them~--~building the last-recorded dependencies before checking them.
-One minor annoyance is that the \hs{State} has been extended and thus needs to be projected using \hs{fst} before it is used.
+\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the
+    \hs{key} you are currently building, and a function \hs{check} which checks
+    a specified dependency. It returns the \hs{result} field of all traces that
+    match. Since \hs{check} is in an arbitrary monad, the function has to use
+    \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&}
+    functions\footnote{These functions are all available in the \cmd{extra}
+    library on Hackage.}.
+\item[\hs{recursive}] defines the dependency ordering pattern. It requires a
+    \hs{process} function that builds a key given a way to recursively build a
+    dependency, and a way to run task to produce result. The main purpose of
+    \hs{recursive} is to ensure that in a single run no key is built twice~--~for
+    which it extends the \hs{State} monad with a list of \hs{done} keys.
+\item[\hs{shake}] ties everything together. First it tests if the traces allow
+the current state of the target key. If not, it builds the key and updates the
+store. The only subtlety is that Shake calls \hs{fetch} on the keys while
+checking them~--~building the last-recorded dependencies before checking them.
+One minor annoyance is that the \hs{State} has been extended and thus needs to
+be projected using \hs{fst} before it is used.
 \end{description}
@@
 \begin{minted}[fontsize=\small]{haskell}
-bazel :: (Eq k, Hashable v) => Build Applicative (Traces k v) k v
-bazel = topological $ \key ds act -> do
-    s <- get
-    let Traces traces contents = getInfo s
-    poss <- traceMatch (\k v -> return $ getHash k s == v) key traces
-    if null poss then do
-        v <- act
-        modify $ \s ->
-            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
-                ts = Traces (t : traces) (Map.insert (hash v) v contents)
-            in putInfo ts (putValue key v s)
-    else do
-        when (getHash key s `notElem` poss) $
-            modify $ putValue key (contents Map.! head poss)
+-- Bazel build system, implementation of 'restarting2' is omitted (22 lines)
+bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
+bazel = restarting2 ctRebuilder
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Cloud Shake build system, implementation of 'recursive' is given in Fig. 8
+cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
+cloudShake = recursive ctRebuilder
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Rebuilder based on constructive traces
+ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
+ctRebuilder key value task = Task $ \fetch -> do
+    ct <- get
+    maybeCachedValue <- constructCT key value (fmap hash . fetch) ct
+    case maybeCachedValue of
+        Just cachedValue -> return cachedValue
+        Nothing -> do
+            (newValue, deps) <- track task fetch
+            put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
+            return newValue
 \end{minted}
 \vspace{-2mm}
-\caption{An implementation of \Bazel using our framework; \hs{topological} is
-defined in Fig.~\ref{fig-make-implementation}.}\label{fig-bazel-implementation}
+\caption{Implementations of \Bazel and Cloud \Shake using our framework;
+\hs{restarting2} differs from the restarting scheduler used by Excel
+(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the
+calculation chain.}
+% \hs{topological} is defined in Fig.~\ref{fig-make-implementation}.
+\label{fig-bazel-cloud-shake-implementation}
 \vspace{-2mm}
@@
 
-Now we have seen all three dependency schemes, we can directly reuse \hs{topological} to define \Bazel.
-Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, along with the \hs{Traces} type from \S\ref{sec-constructive-traces}.
-With these pieces in place, the implementation of \Bazel is given in Fig.~\ref{fig-bazel-implementation}.
-We first figure out the possible results given the current state.
-If there are no recorded traces for the current dependencies we run it and record the results, otherwise grab a suitable result from the \hs{contents} cache.
+Now we have seen all three dependency schemes, we can directly reuse
+\hs{topological} to define \Bazel. Furthermore, as \Bazel is a tracing build
+system, we can reuse \hs{Trace} and \hs{traceMatch}, along with the \hs{Traces}
+type from \S\ref{sec-constructive-traces}. With these pieces in place, the
+implementation of \Bazel is given in
+Fig.~\ref{fig-bazel-cloud-shake-implementation}. We first figure out the
+possible results given the current state. If there are no recorded traces for
+the current dependencies we run it and record the results, otherwise grab a
+suitable result from the \hs{contents} cache.
 
-The program presented above captures certain aspects of \Bazel, but the real implementation makes one important additional assumption on \hs{Task} -- namely that it is \textit{deterministic}.
-With that assumption \Bazel is able to create the result hash in a trace by hashing together the result hashes of the dependencies and the key -- as the mapping between dependencies and results is deterministic.
-As a consequence \Bazel can compute the results of traces locally, without looking at \hs{Traces} (potentially saving a roundtrip to the server).
-To model this change in the code would require storing an additional transient piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)}, storing the computed hashes so far this run.
-With that available, \hs{getHash}~\hs{key}~\hs{s} would become:
+The program presented above captures certain aspects of \Bazel, but the real
+implementation makes one important additional assumption on \hs{Task} -- namely
+that it is \textit{deterministic}. With that assumption \Bazel is able to create
+the result hash in a trace by hashing together the result hashes of the
+dependencies and the key -- as the mapping between dependencies and results is
+deterministic. As a consequence \Bazel can compute the results of traces locally,
+without looking at \hs{Traces} (potentially saving a roundtrip to the server).
+To model this change in the code would require storing an additional transient
+piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)},
+storing the computed hashes so far this run. With that available,
+\hs{getHash}~\hs{key}~\hs{s} would become:
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 
-Using the abstractions and approaches built thus far, we have shown how to combine dependency scheme and change approach to reproduce existing build systems.
-In the attached materials we have implemented 9~build systems corresponding to all three dependency schemes, matched with all three change approaches.
-To us, the most interesting build system as yet unavailable would matching recursive ordering with constructive traces~--~providing a cloud-capable build system with minimality, cutoff and monadic dependencies.
-Using our framework it is possible to define and test such a system as per Fig.~\ref{fig-cloudshake-implementation}.
-
-\begin{figure}
-\begin{minted}[fontsize=\small]{haskell}
-cloudShake :: (Eq k, Hashable v) => Build Monad (Traces k v) k v
-cloudShake = recursive $ \key fetch act -> do
-    s <- gets fst
-    let Traces traces contents = getInfo s
-    poss <- traceMatch (\k v -> (==) v . hash <$> fetch k) key traces
-    if null poss then do
-        (v, ds) <- act
-        modify $ \(s,done) ->
-            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
-                ts = Traces (t : traces) (Map.insert (hash v) v contents)
-            in (putInfo ts (putValue key v s), done)
-    else do
-        s <- gets fst
-        when (getHash key s `notElem` poss) $
-            modify $ \(s, done) -> (putValue key (contents Map.! head poss) s, done)
-\end{minted}
-\vspace{-2mm}
-\caption{An implementation of Cloud \Shake using our framework.}\label{fig-cloudshake-implementation}
-\vspace{-4mm}
-\end{figure}
+Using the abstractions and approaches built thus far, we have shown how to
+combine dependency scheme and change approach to reproduce existing build
+systems. In the attached materials we have implemented 9~build systems
+corresponding to all three dependency schemes, matched with all three change
+approaches. To us, the most interesting build system as yet unavailable would
+matching recursive ordering with constructive traces~--~providing a cloud-capable
+build system with minimality, cutoff and monadic dependencies. Using our
+framework it is possible to define and test such a system as per
+Fig.~\ref{fig-bazel-cloud-shake-implementation}.
 
@@
 instead of accessing the store directly.
+
+\begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+-- Buck build system, implementation of 'topological' is given in Fig. 6
+buck :: (Hashable k, Hashable v) => Build Applicative (DCT k v) k v
+buck = topological (unliftRebuilder dctRebuilder)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Convert a monadic rebuilder to the corresponding applicative one
+unliftRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
+unliftRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Nix build system, implementation of 'recursive' is given in Fig. 8
+nix :: (Hashable k, Hashable v) => Build Monad (DCT k v) k v
+nix = recursive dctRebuilder
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Rebuilder based on deterministic constructive traces
+dctRebuilder :: (Hashable k, Hashable v) => Rebuilder Monad (DCT k v) k v
+dctRebuilder key _value task = Task $ \fetch -> do
+    dct <- get
+    maybeCachedValue <- constructDCT key (fmap hash . fetch) dct
+    case maybeCachedValue of
+        Just cachedValue -> return cachedValue
+        Nothing -> do
+            (newValue, deps) <- trackM task fetch
+            put =<< recordDCT key newValue deps (fmap hash . fetch) =<< get
+            return newValue
+\end{minted}
+\vspace{-2mm}
+\caption{Implementations of \Buck and \Nix using our framework.}
+\label{fig-buck-nix-implementation}
+\vspace{-2mm}
+\end{figure}
 

### 2018-06-15 paper/4-build.tex
Work on S5

@@
 \vspace{-2mm}
-\subsubsection{Reordering}\label{sec-reordering}
+\subsubsection{Reordering}\label{sec-restarting}
 
@@
 \vspace{-2mm}
-\subsubsection{Recursive}\label{sec-recursive}
+\subsubsection{Recursive}\label{sec-suspending}
 
@@
 \hline
-Property           & Topological\hspace{2mm}\S\ref{sec-topological} & Reordering\hspace{2mm}\S\ref{sec-reordering} & Recursive\hspace{2mm}\S\ref{sec-recursive}    \\\hline
+Property           & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
 \hline

### 2018-06-15 paper/5-implementations.tex
Work on S5

@@
 In the previous sections we discussed the types of build systems, and how they
-can be broken down. But these divisions were not obvious to us, and only by
-concretely implementing and refactoring each build system did we determine the
-underlying commonalities. In this section we share some of the code that got us
-there.
+can be broken down into two main components: a scheduler and a rebuilder. This
+decomposition was not obvious to us, and only by concretely implementing and
+refactoring each build system did we determine the underlying commonalities.
+In this section we share some of the code that got us there.
 
@@
 the persistent build information that \Make stores by a pair
-\hs{(@@modTime,}~\hs{now)} comprising the \emph{file modification time} function
-\hs{modTime}~\hs{::}~\hs{k}~\hs{->}~\hs{Time} and the \emph{current time} \hs{now}.
+\hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
+dictionary \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
+modification times}. The top-level definition
+\hs{make}~\hs{=}~\hs{topological}~\hs{modTimeRebuilder} is a direct translation
+of the scheduler-rebuilder decomposition, illustrated in Table~\ref{tab-build-systems}.
+
 Setting aside the explicit manipulation of file modification times, which in
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Topological scheduler
+-- A topological task scheduler
 topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Restarting scheduler
+-- A restarting task scheduler
 restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Rebuilder based on approximate dependencies
+-- A task rebuilder based on approximate dependencies
 approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
@@
 shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
-shake = recursive vtRebuilder
+shake = suspending vtRebuilder
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Recursive dependency strategy
-recursive :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
-recursive rebuilder tasks key store = fst $ execState (fetch key) (store, Set.empty)
+-- A suspending task scheduler
+suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
+suspending rebuilder tasks key store = fst $ execState (fetch key) (store, Set.empty)
   where
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Cloud Shake build system, implementation of 'recursive' is given in Fig. 8
+-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 8
 cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
-cloudShake = recursive ctRebuilder
+cloudShake = suspending ctRebuilder
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Nix build system, implementation of 'recursive' is given in Fig. 8
+-- Nix build system, implementation of 'suspending' is given in Fig. 8
 nix :: (Hashable k, Hashable v) => Build Monad (DCT k v) k v
-nix = recursive dctRebuilder
+nix = suspending dctRebuilder
 \end{minted}

### 2018-06-16 src/Build/Scheduler.hs
Minor tweaks

@@ -37,8 +37,8 @@ updateValue key _value newValue = putValue key newValue
 -- extracting all (static) dependencies upfront, and then traversing the graph
 -- in the topological order, rebuilding keys using the supplied rebuilder.
 topological :: forall i k v. Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
-topological rebuilder tasks target = execState $ forM_ chain $ \key -> case tasks key of
-    Nothing   -> return ()
+topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
+    Nothing -> return ()
     Just task -> do
         value <- gets (getValue key)
         let newTask :: Task (MonadState i) k v
@@ -50,7 +50,7 @@ topological rebuilder tasks target = execState $ forM_ chain $ \key -> case task
         modify $ putInfo newInfo . updateValue key value newValue
   where
     deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
-    chain  = case topSort (graph deps target) of
+    order  = case topSort (graph deps target) of
         Nothing -> error "Cannot build tasks with cyclic dependencies"
         Just xs -> xs
 

### 2018-06-16 paper/5-implementations.tex
Finish S5.1

@@
 
-We provide an implementation of \Make using our framework in
-Fig.~\ref{fig-make-implementation}. \Make processes keys in a linear order based
-on a topological sort (see~\S\ref{sec-parallelism} for parallel \Make). For
-each key, it builds it if it is older than any of its dependencies. We capture
-the persistent build information that \Make stores by a pair
-\hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
-dictionary \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
-modification times}. The top-level definition
-\hs{make}~\hs{=}~\hs{topological}~\hs{modTimeRebuilder} is a direct translation
-of the scheduler-rebuilder decomposition, illustrated in Table~\ref{tab-build-systems}.
-
-Setting aside the explicit manipulation of file modification times, which in
-reality is taken care of by the file system, the function \hs{make} captures
-the essence of \Make in a clear and precise manner.
-
 \begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+-- Make build system
+make :: Ord k => Build Applicative (MakeInfo k) k v
+make = topological modTimeRebuilder
+\end{minted}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Make build system
-make :: Ord k => Build Applicative (MakeInfo k) k v
-make = topological modTimeRebuilder
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- A topological task scheduler
-topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
-topological rebuilder tasks key = execState $ forM_ chain $ \k -> case tasks k of
-    Nothing   -> return ()
-    Just task -> do
-        value <- gets (getValue k)
-        let newTask :: Task (MonadState i) k v
-            newTask = rebuilder k value task
-            newFetch :: k -> StateT i (State (Store i k v)) v
-            newFetch = lift . gets . getValue
-        info <- gets getInfo
-        (newValue, newInfo) <- runStateT (run newTask newFetch) info
-        modify $ putInfo newInfo . putValue k newValue
-  where
-    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
-    chain  = case topSort (reachable deps key) of
-        Nothing -> error "Cannot build tasks with cyclic dependencies"
-        Just xs -> xs
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Standard graph algorithms (implementation omitted)
-reachable :: Ord k => (k -> [k]) -> k -> Graph k
-topSort   :: Ord k => Graph k -> Maybe [k]
-\end{minted}
-\vspace{1mm}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- A topological task scheduler
+topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
+topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
+    Nothing   -> return ()
+    Just task -> do
+        value <- gets (getValue key)
+        let newTask :: Task (MonadState i) k v
+            newTask = rebuilder key value task
+            newFetch :: k -> StateT i (State (Store i k v)) v
+            newFetch = lift . gets . getValue
+        info <- gets getInfo
+        (newValue, newInfo) <- runStateT (run newTask newFetch) info
+        modify $ putInfo newInfo . putValue key newValue
+  where
+    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
+    order  = case topSort (reachable deps target) of
+        Nothing -> error "Cannot build tasks with cyclic dependencies"
+        Just xs -> xs
+\end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Standard graph algorithms (implementation omitted)
+reachable :: Ord k => (k -> [k]) -> k -> Graph k
+topSort   :: Ord k => Graph k -> Maybe [k]
+\end{minted}
 \vspace{-2mm}
 \caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
-\vspace{-2mm}
+\vspace{-4mm}
 \end{figure}
 
-The helper function \hs{topological} calls \hs{process} on every \hs{key} in
-a topological order, providing the list of its dependencies \hs{deps} and the
-action \hs{act} to compute the resulting value if it needs to be rebuilt. To
-determine if the \hs{key} is \hs{dirty}, \hs{process} compares its modification
-time with those of its dependencies. If the \hs{key} needs to be rebuilt, the
-action \hs{act} is executed and the obtained result is stored, along with an
-updated file modification timestamp.
+An implementation of \Make using our framework is shown in
+Fig.~\ref{fig-make-implementation}. The top-level definition
+\hs{make}~\hs{=}~\hs{topological}~\hs{modTimeRebuilder} is a direct translation
+of the scheduler-rebuilder decomposition in Table~\ref{tab-build-systems}. Below
+we describe both components, starting with \hs{modTimeRebuilder}.
 
-The implementation of \hs{topological} encodes the dependency strategy that
-\Make has chosen to use. The \hs{where} clause corresponds to the pre-processing
-stage, which uses the function \hs{dependencies}, defined in~\S\ref{sec-deps},
-to extract static dependencies from the provided applicative \hs{task}. We
-compute the linear processing \hs{chain} by taking the keys \hs{reachable}
-from \hs{key} via dependencies, and performing the topological sort of the
-result. We omit implementation of textbook graph algorithms
-\hs{reachable} and \hs{topSort}, e.g.
-see~\cite{cormen2001introduction}.
+We capture persistent build information that \Make stores by the pair
+\hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
+dictionary \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
+modification times}. The rebuilder receives three arguments: a \hs{key}, its
+current \hs{value}, and the \hs{task} of type
+\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild the
+\hs{key} if necessary. The rebuilder returns a modified monadic task of type
+\hs{Task}~\hs{(@@MonadState}~\hs{(@@MakeInfo}~\hs{k))}~\hs{k}~\hs{v} that has
+access to the build information and uses it to decide whether to return the
+current \hs{value} as is or rebuild the \hs{key} if it is \hs{dirty}, i.e.
+if it is older than any of its dependencies\footnote{Since the \hs{task} is
+applicative we can statically find its dependencies using the function
+\hs{dependencies} defined in~\S\ref{sec-deps}.} or has never been built before
+(in which case \hs{modTime} has no entry for it). Note that in our model we
+update the current time and file modification time whenever the \hs{key} is
+rebuilt, which in reality is taken care of by the file system.
 
-The \hs{chain} is processed in the \hs{State} monad, with each non-input key
-\hs{k} in the chain passed to the provided \hs{process} function, along with
-\hs{k}'s dependencies and the action \hs{act}, which when executed recomputes
-the \hs{k}'s value by fetching its dependencies from the store.
+\Make's scheduler processes keys in a linear \hs{order} based on a
+\hs{topological} sort of the statically known dependency graph
+(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
+Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
+build information \hs{i} and is therefore compatible with any applicative
+\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
+\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
+\hs{newValue} and updated build information \hs{newInfo}, which are subsequently
+written to the store. The computation runs in the
+\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
+build system (\S\ref{sec-general-build}), but we also need to use an additional
+\hs{StateT}~\hs{i} wrapper to run the \hs{newTask} since it only has access to
+the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}.
+
+The \hs{where} clause corresponds to the pre-processing stage, which uses the
+function \hs{dependencies}, defined in~\S\ref{sec-deps}, to extract static
+dependencies from the provided applicative \hs{task}. We compute the linear
+processing \hs{order} by constructing the graph of keys \hs{reachable} from the
+\hs{target} via dependencies, and performing the topological sort of the result.
+We omit implementation of textbook graph algorithms
+\hs{reachable} and \hs{topSort}, e.g. see~\cite{cormen2001introduction}.
 
@@
 restricts \Make to static dependencies, as reflected in the
-type~\hs{Build}~\hs{Applicative}. Moreover, any other build system following
-the same \hs{topological} approach will also inherit the same restriction.
+type~\hs{Build}~\hs{Applicative}. Moreover, any other build system using
+the same \hs{topological} scheduler will also inherit the same restriction.
 

### 2018-06-16 paper/5-implementations.tex
Work on S5

@@
             time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
-    if not dirty
-    then return value
-    else do
+    if not dirty then return value else do
         put (now + 1, Map.insert key now modTime)
@@
             newTask = rebuilder key value task
-            newFetch :: k -> StateT i (State (Store i k v)) v
-            newFetch = lift . gets . getValue
+            fetch :: k -> StateT i (State (Store i k v)) v
+            fetch = lift . gets . getValue
         info <- gets getInfo
-        (newValue, newInfo) <- runStateT (run newTask newFetch) info
+        (newValue, newInfo) <- runStateT (run newTask fetch) info
         modify $ putInfo newInfo . putValue key newValue
@@
 \hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
-dictionary \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
+map \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
 modification times}. The rebuilder receives three arguments: a \hs{key}, its
@@
 \hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
-build system (\S\ref{sec-general-build}), but we also need to use an additional
+build system (\S\ref{sec-general-build}), but we need to use an additional
 \hs{StateT}~\hs{i} wrapper to run the \hs{newTask} since it only has access to
-the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}.
+the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}. The \hs{newTask} can
+find values of the \hs{key}'s dependencies via the \hs{fetch} callback that
+reads the store by executing \hs{getValue} in the inner state monad.
 
@@
 
-We define \Excel, with its reordering dependency strategy, in  Fig.~\ref{fig-excel-implementation}.
-\Excel's persistently stored information is a triple: (i) the dirty bit
-function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
-\hs{k}~\hs{->}~\hs{ApproximateDependencies}~\hs{k} that \Excel uses to handle
-dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
-previous build (\S\ref{sec-background-excel}).
+Fig.~\ref{fig-excel-implementation} shows an implementation of \Excel in our
+framework as a combination of the \hs{restarting} scheduler and an
+\hs{approximateRebuilder}.
+
+\Excel's persistently stored information is a triple: (i) the set of dirty keys
+\hs{Set}~\hs{k}, (ii) a map storing known upper bounds of dependencies
+\hs{Map}~\hs{k}~\hs{[@@k]} that \Excel uses to handle dynamic dependencies, and
+(iii) the calculation chain \hs{[@@k]} recorded in the previous build
+(\S\ref{sec-background-excel}).
 
 \begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+-- Excel build system
+excel :: Ord k => Build Monad (ExcelInfo k) k v
+excel = restarting approximateRebuilder
+\end{minted}
+\vspace{1mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Build information: dirty keys, approximate dependencies, and calculation chain
-type CalcChain k = [k]
-type ExcelInfo k = ((Set k, ApproximateDependencies k), CalcChain k)
+-- A task rebuilder based on the set of dirty keys and approximate dependencies
+approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
+approximateRebuilder key value task = Task $ \fetch -> do
+    (isDirty, deps) <- get
+    let dirty = key `Set.member` isDirty || case Map.lookup key deps of
+                                                Nothing -> True
+                                                Just ks -> any (`Set.member` isDirty) ks
+    if not dirty then return value else do
+        put (Set.insert key isDirty, deps)
+        run task fetch
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Excel build system
-excel :: Ord k => Build Monad (ExcelInfo k) k v
-excel = restarting approximateRebuilder
+-- Build information: dirty keys, approximate dependencies, and calculation chain
+type Chain k = [k]
+type ExcelInfo k = ((Set k, ApproximateDependencies k), Chain k)
 \end{minted}
@@
                 newTask = try $ rebuilder k value task -- `try' is defined below
-                newFetch :: k -> State i (Either k v)
-                newFetch k | k `Set.member` done = return $ Right (getValue k store)
+                fetch :: k -> State i (Either k v)
+                fetch k | k `Set.member` done = return $ Right (getValue k store)
                         | otherwise           = return $ Left k
-            case runState (run newTask newFetch) (fst $ getInfo store) of
+            case runState (run newTask fetch) (fst $ getInfo store) of
                 (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [k]
@@
 try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- A task rebuilder based on approximate dependencies
-approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
-approximateRebuilder key value task = Task $ \fetch -> do
-    (isDirty, deps) <- get
-    let dirty = key `Set.member` isDirty || case Map.lookup key deps of
-                                                Nothing -> True
-                                                Just ks -> any (`Set.member` isDirty) ks
-    if not dirty
-    then return value
-    else do
-        put (Set.insert key isDirty, deps)
-        run task fetch
 \end{minted}
@@
 suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
-suspending rebuilder tasks key store = fst $ execState (fetch key) (store, Set.empty)
+suspending rebuilder tasks key store = fst $ execState (build key) (store, Set.empty)
   where
-    fetch :: k -> State (Store i k v, Set k) v
-    fetch key = case tasks key of
+    build :: k -> State (Store i k v, Set k) v
+    build key = case tasks key of
         Nothing -> gets (getValue key . fst)
@@
                     newTask = rebuilder key value task
-                    newFetch :: k -> StateT i (State (Store i k v, Set k)) v
-                    newFetch k = do v <- lift $ fetch k
+                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
+                    fetch k = do v <- lift (build k)
                                  put =<< lift (gets (getInfo . fst))
@@
                 info <- gets (getInfo . fst)
-                (newValue, newInfo) <- runStateT (run newTask newFetch) info
+                (newValue, newInfo) <- runStateT (run newTask fetch) info
                 modify $ \(s, _) ->

### 2018-06-16 paper/2-background.tex
Fix paper build
Finish S5.2

@@
 referred to as \emph{build rules}) and their dependencies in a simple textual form.
-For example\footnote{Do not worry about the dependency on \cmd{stdio.h}, and its transitive (perhaps dynamic) dependencies.  We will deal with these in Section \simon{xxx}.}:
+For example\footnote{Do not worry about the dependency on \cmd{stdio.h}, and
+its transitive (perhaps dynamic) dependencies.
+We will deal with these in Section~\S\ref{sec-task-monad}.}:
 
@@
 discovered dependency has not yet been rebuilt. We refer to this
-algorithm as \emph{reordering}, and will discuss it in more detail
+algorithm as \emph{restarting}, and will discuss it in more detail
 in~\S\ref{sec-implementation-excel}.
 
-Dynamic dependencies complicate minimality.  In the above example,
-\cmd{B1} should only be recomputed if \cmd{A1} or \cmd{C1} change, but
-not if (say) \cmd{A2} changes; but these facts are not statically apparent.
-In pracice Excel implements a conservative approximation to
-minimality: roughly, it recomputes a formula if (a) the formula
-text mentions a changed cell, or (b) the formula uses a function like \cmd{INDIRECT}
-whose dependencies are not statically visible.
+Dynamic dependencies complicate minimality.  In the above example, \cmd{B1}
+should only be recomputed if \cmd{A1} or \cmd{C1} change, but not if (say)
+\cmd{A2} changes; but these facts are not statically apparent. In practice
+\Excel implements a conservative approximation to minimality: roughly, it
+recomputes a formula if (i) the formula text mentions a changed cell, or (ii)
+the formula uses a function like \cmd{INDIRECT} whose dependencies are not
+statically visible.
 

### 2018-06-16 paper/5-implementations.tex
Finish S5.2

@@
     Just task -> do
-        value <- gets (getValue key)
+        store <- get
         let newTask :: Task (MonadState i) k v
-            newTask = rebuilder key value task
-            fetch :: k -> StateT i (State (Store i k v)) v
-            fetch = lift . gets . getValue
-        info <- gets getInfo
-        (newValue, newInfo) <- runStateT (run newTask fetch) info
+            newTask = rebuilder key (getValue key store) task
+            fetch :: k -> State i v
+            fetch k = pure (getValue k store)
+            (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
         modify $ putInfo newInfo . putValue key newValue
@@
 \hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
-build system (\S\ref{sec-general-build}), but we need to use an additional
-\hs{StateT}~\hs{i} wrapper to run the \hs{newTask} since it only has access to
-the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}. The \hs{newTask} can
-find values of the \hs{key}'s dependencies via the \hs{fetch} callback that
-reads the store by executing \hs{getValue} in the inner state monad.
+build system (\S\ref{sec-general-build}), however since \hs{newTask} has access
+only to the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}, we need to use
+\hs{runState} to execute it. The \hs{newTask} finds values of the \hs{key}'s
+dependencies via the \hs{fetch} callback that directly reads the \hs{store}.
 
@@
 Fig.~\ref{fig-excel-implementation} shows an implementation of \Excel in our
-framework as a combination of the \hs{restarting} scheduler and an
+framework as a composition of the \hs{restarting} scheduler and an
 \hs{approximateRebuilder}.
 
-\Excel's persistently stored information is a triple: (i) the set of dirty keys
-\hs{Set}~\hs{k}, (ii) a map storing known upper bounds of dependencies
-\hs{Map}~\hs{k}~\hs{[@@k]} that \Excel uses to handle dynamic dependencies, and
-(iii) the calculation chain \hs{[@@k]} recorded in the previous build
-(\S\ref{sec-background-excel}).
+\Excel's persistently stored build information is a triple:
+
+\begin{itemize}
+    \item A set of dirty keys \hs{Set}~\hs{k}, i.e. all cells whose contents
+          has changed since the previous build, including cells that contain
+          input values as well as formulas.
+    \item A map \hs{Map}~\hs{k}~\hs{[@@k]} that stores known upper bounds of
+          dependencies. If there is an entry for a key, it is a conservative
+          approximation of its dependencies (e.g. the dependencies of the
+          formula \cmd{IF(C1=1,B2,A2)} can be approximated by the list
+          \hs{[}\cmd{C1}\hs{,}\cmd{B2}\hs{,}\cmd{A2}\hs{]}). Otherwise, there is
+          no upper bound (e.g. the key uses an \cmd{INDIRECT} reference) and
+          the key is assumed to be dirty.
+    \item A calculation \hs{Chain} recorded in the previous build
+          (\S\ref{sec-background-excel}), represented by a list of keys \hs{[@@k]}.
+\end{itemize}
+
+The \hs{approximateRebuilder} uses the first two elements of the above triple
+to decide whether a given \hs{key} is \hs{dirty} and hence needs to be rebuilt.
+The logic is similar to that of \Make's \hs{modTimeRebuilder}, but instead of
+relying on file modification times, \Excel stores the information about dirty
+keys more directly. Furthermore, \Excel's dependency approximation makes it
+possible to build tasks with dynamic dependencies: a \hs{key} is declared
+\hs{dirty} if any of its approximate dependencies is in the \hs{dirtyKeys} set,
+or if there is no known upper bound on dependencies, i.e. if the lookup in the
+map \hs{deps}~\hs{::}~\hs{Map}~\hs{k}~\hs{[@@k]} fails. Note that when a
+\hs{key} is rebuilt it is added to the \hs{dirtyKeys} set to trigger further
+rebuilds.
+
+\Excel's \hs{restarting} scheduler processes keys in the order specified by the
+calculation \hs{chain}, constructing a \hs{newChain} for the next build and
+maintaining a set of keys \hs{done} that have already been processed. For each
+non-input \hs{key}, the scheduler tries to rebuild it using a partial \hs{fetch}
+callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}. The callback
+is defined to fail with \hs{Left}~\hs{dep} when asked for the value of a
+dependency \hs{dep} that has not yet been processed (and hence may potentially
+be dirty); otherwise it returns the current value of the dependency in the
+\hs{store}.
+
+After the \hs{newTask} is executed with \hs{runState} there are two cases to
+consider:
+
+\begin{itemize}
+    \item The \hs{newTask} has failed, because one of its dependencies \hs{dep}
+    has not yet been processed. This indicates that the calculation \hs{chain}
+    from the previous build is incorrect and needs to be adjusted by making sure
+    it contains \hs{dep} and moving the \hs{key} to the end, so that its build
+    can be restarted after the \hs{dep} is ready.
+    \item The \hs{newTask} succeeded. The resulting \hs{newValue} and
+    \hs{newInto} are written to the store, and \Excel continues to build the
+    \hs{chain} after marking the \hs{key} as \hs{done}.
+\end{itemize}
+
+Note that the task returned by the \hs{rebuilder} expects a total callback
+function and cannot be directly executed with the partial callback \hs{fetch}.
+We fix the mismatch with the function \hs{try} that relies on the standard
+monad transformer \hs{ExceptT} from the \cmd{transformers} library.
 
@@
 \end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Approximate task dependencies: a map of known upper bounds of dependencies
-type ApproximateDependencies k = Map k [k]
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- A task rebuilder based on the set of dirty keys and approximate dependencies
-approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
-approximateRebuilder key value task = Task $ \fetch -> do
-    (isDirty, deps) <- get
-    let dirty = key `Set.member` isDirty || case Map.lookup key deps of
-                                                Nothing -> True
-                                                Just ks -> any (`Set.member` isDirty) ks
-    if not dirty then return value else do
-        put (Set.insert key isDirty, deps)
-        run task fetch
-\end{minted}
-\vspace{1mm}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 type Chain k = [k]
-type ExcelInfo k = ((Set k, ApproximateDependencies k), Chain k)
+type ExcelInfo k = ((Set k, Map k [k]), Chain k)
 \end{minted}
-\vspace{1mm}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- A task rebuilder based on the set of dirty keys and approximate dependencies
+approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
+approximateRebuilder key value task = Task $ \fetch -> do
+    (dirtyKeys, deps) <- get
+    let dirty = key `Set.member` dirtyKeys ||
+                case Map.lookup key deps of Nothing -> True
+                                            Just ks -> any (`Set.member` dirtyKeys) ks
+    if not dirty then return value else do
+        put (Set.insert key dirtyKeys, deps)
+        run task fetch
+\end{minted}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
-restarting rebuilder tasks key = execState $ do
+restarting rebuilder tasks target = execState $ do
     chain    <- gets (snd . getInfo)
-    newChain <- go Set.empty $ chain ++ [key | key `notElem` chain]
+    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
     modify . mapInfo $ \(i, _) -> (i, newChain)
@@
     go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
-    go _    []     = return []
-    go done (k:ks) = case tasks k of
-        Nothing -> (k :) <$> go (Set.insert k done) ks
+    go _    [@@]       = return [@@]
+    go done (key:ks) = case tasks key of
+        Nothing -> (key :) <$> go (Set.insert key done) ks
         Just task -> do
             store <- get
-            let value = getValue k store
-                newTask :: Task (MonadState i) k (Either k v)
-                newTask = try $ rebuilder k value task -- `try' is defined below
+            let newTask :: Task (MonadState i) k (Either k v)
+                newTask = try $ rebuilder key (getValue key store) task
                 fetch :: k -> State i (Either k v)
@@
             case runState (run newTask fetch) (fst $ getInfo store) of
-                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [k]
+                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
                 (Right newValue, newInfo) -> do
-                    modify $ putInfo (newInfo, []) . updateValue k value newValue
-                    (k :) <$> go (Set.insert k done) ks
+                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
+                    (key :) <$> go (Set.insert key done) ks
 \end{minted}
-\vspace{1mm}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
-\vspace{-2mm}
+\vspace{-4mm}
 \end{figure}
-
-The helper function \hs{reordering}, whose implementation we omit since it is
-technical and not particularly enlightening, calls the function \hs{process} to
-\emph{try to build} a \hs{key} by executing the action \hs{act}, in the order
-determined by the calculation chain. To decide whether to \hs{rebuild} the
-\hs{key}, \hs{process} checks if the \hs{key} itself is marked dirty or the
-approximation of its dependencies contains a dirty key. Notice that if the
-dependencies of the \hs{key} are \hs{Unknown}, e.g. when it uses the
-\cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
-not needed, \hs{process} returns \hs{Nothing} to indicate that. Otherwise, it
-executes \hs{act} leading to one of two possible \hs{result}s:
-
-\begin{itemize}
-    \item \hs{MissingDependency}~\hs{k} indicates that the execution of \hs{act} has
-    failed, because one of its dependencies \hs{k} was out-of-date, i.e. the
-    calculation chain from the previous build was incorrect and therefore needs
-    to be reordered, deferring the \hs{key} to be rebuilt later.
-    \item \hs{Result}~\hs{v}~\hs{dynamicDependencies} indicates that the
-    execution has succeeded producing the value~\hs{v} and the list of the
-    \hs{key}'s dynamic dependencies. We store the value, and mark it dirty to
-    trigger the rebuilding of keys that depend on it.
-\end{itemize}
-
-In both of the above cases, we notify the parent \hs{reordering} function of
-the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
-that \hs{process} ignores \hs{dynamicDependencies} available in the \hs{result}.
-While not required for \Excel, we have implemented build systems using
-\hs{reordering} which use verifying and constructive traces, effectively turning
-\Excel into a cloud build system and ensuring \hs{reordering} is not overly
-fitted to \Excel alone.
-%  (additional implementations will be submitted to the
-% Artefact Evaluation).
-
-In reality \Excel works slightly differently (as far as we are able to
-ascertain)~--~on a change it propagates the dirty bits forward using the
-\hs{dynamicDependencies}, then only checks if the current \hs{key} is dirty.
-While both methods are equivalent, merely changing the interleaving, our
-approach allows us to model more of the behaviour of \Excel.
 

### 2018-06-17 src/Build/Scheduler.hs
Stick to return everywhere

@@ -45,7 +45,7 @@ topological rebuilder tasks target = execState $ forM_ order $ \key -> case task
             newTask :: Task (MonadState i) k v
             newTask = rebuilder key value task
             fetch :: k -> State i v
-            fetch k = pure (getValue k store)
+            fetch k = return (getValue k store)
             (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
         modify $ putInfo newInfo . updateValue key value newValue
   where

### 2018-06-17 paper/5-implementations.tex
Finish S5.3

@@
             fetch :: k -> State i v
-            fetch k = pure (getValue k store)
+            fetch k = return (getValue k store)
             (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Convert a total task into a task accepting a partial fetch callback
+-- Convert a total task into a task that accepts a partial fetch callback
 try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
@@
 \caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
-\vspace{-4mm}
+\vspace{-5mm}
 \end{figure}
 
-\vspace{-2mm}
+\vspace{-1mm}
 \subsection{\Shake}\label{sec-implementation-shake}
@@
 
-The \Shake approach for dependency tracking involves recording traces and
-verifying them, for which we use the \hs{Trace} type defined
-in~\ref{sec-verifying-traces}. Complete code is given
-in~\S\ref{fig-shake-implementation}, split into three functions:
-
-\begin{description}
-\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the
-    \hs{key} you are currently building, and a function \hs{check} which checks
-    a specified dependency. It returns the \hs{result} field of all traces that
-    match. Since \hs{check} is in an arbitrary monad, the function has to use
-    \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&}
-    functions\footnote{These functions are all available in the \cmd{extra}
-    library on Hackage.}.
-\item[\hs{recursive}] defines the dependency ordering pattern. It requires a
-    \hs{process} function that builds a key given a way to recursively build a
-    dependency, and a way to run task to produce result. The main purpose of
-    \hs{recursive} is to ensure that in a single run no key is built twice~--~for
-    which it extends the \hs{State} monad with a list of \hs{done} keys.
-\item[\hs{shake}] ties everything together. First it tests if the traces allow
-the current state of the target key. If not, it builds the key and updates the
-store. The only subtlety is that Shake calls \hs{fetch} on the keys while
-checking them~--~building the last-recorded dependencies before checking them.
-One minor annoyance is that the \hs{State} has been extended and thus needs to
-be projected using \hs{fst} before it is used.
-\end{description}
+Our model of \Shake is shown in Fig.~\ref{fig-shake-implementation}. It stores
+verifying traces \hs{VT}~\hs{k}~\hs{v} defined in~\S\ref{sec-verifying-traces}
+as persistent build information and is composed of the \hs{suspending} scheduler
+and \hs{vtRebuilder}.
 
@@
 \end{minted}
-\vspace{1mm}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- A task rebuilder based on verifying traces
+vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
+vtRebuilder key value task = Task $ \fetch -> do
+    upToDate <- verifyVT key value (fmap hash . fetch) =<< get
+    if upToDate then return value else do
+        (newValue, deps) <- track task fetch
+        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
+        return newValue
+\end{minted}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
-suspending rebuilder tasks key store = fst $ execState (build key) (store, Set.empty)
+suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
   where
-    build :: k -> State (Store i k v, Set k) v
+    build :: k -> State (Store i k v, Set k) (@@)
     build key = case tasks key of
-        Nothing -> gets (getValue key . fst)
+        Nothing -> return (@@)
         Just task -> do
@@
                     fetch :: k -> StateT i (State (Store i k v, Set k)) v
-                    fetch k = do v <- lift (build k)
-                                 put =<< lift (gets (getInfo . fst))
-                                 return v
+                    fetch k = do lift (build k)                      -- build the key
+                                 lift (gets (getInfo . fst)) >>= put -- save new traces
+                                 lift (gets (getValue k . fst))      -- fetch the value
                 info <- gets (getInfo . fst)
@@
                     (putInfo newInfo $ putValue key newValue s, Set.insert key done)
-            gets (getValue key . fst)
 \end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Rebuilder based on verifying traces
-vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
-vtRebuilder key value task = Task $ \fetch -> do
-    vt <- get
-    dirty <- not <$> verifyVT key value (fmap hash . fetch) vt
-    if not dirty
-    then return value
-    else do
-        (newValue, deps) <- trackM task fetch
-        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
-        return newValue
-\end{minted}
-\vspace{-3mm}
+\vspace{-2mm}
 \caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
-\vspace{1mm}
+\vspace{-6mm}
 \end{figure}
+
+The rebuilder executes the verification query \hs{verifyVT} to determine if the
+\hs{key} is \hs{upToDate}. If it is, the rebuilder simply returns the \hs{key}'s
+current \hs{value}. Otherwise it executes the \hs{task}, obtaining both a
+\hs{newValue} and the \hs{key}'s dynamic dependencies \hs{deps} (see the
+definition of \hs{track} in~\S\ref{sec-deps}), which are subsequently recorded
+in a new verification trace using \hs{recordVT}.
+
+The \hs{suspending} scheduler is based on a recursive function \hs{build} that
+builds a given \hs{key}, making sure not to duplicate work when called on the
+same \hs{key} again in future. To achieve that, it keeps track of keys that have
+already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}, running the
+computation in the
+\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
+
+Given a non-input \hs{key} that has not yet been built, we use the supplied
+\hs{rebuilder} to embed the build information~\hs{i} into the \hs{task}.
+We then execute the obtained \hs{newTask} with the \hs{fetch} callback, which
+performs the following three steps when given a dependency \hs{k}:
+
+\begin{itemize}
+    \item Build \hs{k} by recursively calling the \hs{build} function. Note
+          that if \hs{k} is dirty, this will effectively suspend the build of
+          the current \hs{key} until \hs{k} is brought up to date.
+    \item Save any new traces that have been generated when building \hs{k}.
+    \item Read the up to date value of \hs{k} from the store and return it.
+\end{itemize}
+
+These steps are run in the
+\hs{StateT}~\hs{i}~\hs{(@@State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k))}
+monad, because the \hs{newTask} has access only to the~\hs{i} part of the state;
+we use \hs{lift} to execute commands in the inner monad.
+
+Finally, the \hs{newValue} and \hs{newInfo} obtained by running the \hs{newTask}
+are stored, and \hs{key} is added to the set \hs{done}.
 

### 2018-06-18 paper/5-implementations.tex
Work on S5.4

@@
 
-\vspace{-2mm}
 \subsection{\Bazel}\label{sec-implementation-bazel}
-\vspace{-1mm}
 
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Rebuilder based on constructive traces
+-- A rebuilder based on constructive traces
 ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
 ctRebuilder key value task = Task $ \fetch -> do
-    ct <- get
-    maybeCachedValue <- constructCT key value (fmap hash . fetch) ct
+    maybeCachedValue <- constructCT key value (fmap hash . fetch) =<< get
     case maybeCachedValue of
@@
 
-Now we have seen all three dependency schemes, we can directly reuse
-\hs{topological} to define \Bazel. Furthermore, as \Bazel is a tracing build
-system, we can reuse \hs{Trace} and \hs{traceMatch}, along with the \hs{Traces}
-type from \S\ref{sec-constructive-traces}. With these pieces in place, the
-implementation of \Bazel is given in
-Fig.~\ref{fig-bazel-cloud-shake-implementation}. We first figure out the
-possible results given the current state. If there are no recorded traces for
-the current dependencies we run it and record the results, otherwise grab a
-suitable result from the \hs{contents} cache.
+In this section we show how constructive traces can be used to implement cloud
+build systems. Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models
+of \Bazel and Cloud \Shake, which both rely on \hs{ctRebuilder} -- a
+rebuilder based on constructive traces (\S\ref{sec-constructive-traces}).
 
-The program presented above captures certain aspects of \Bazel, but the real
-implementation makes one important additional assumption on \hs{Task} -- namely
-that it is \textit{deterministic}. With that assumption \Bazel is able to create
-the result hash in a trace by hashing together the result hashes of the
-dependencies and the key -- as the mapping between dependencies and results is
-deterministic. As a consequence \Bazel can compute the results of traces locally,
-without looking at \hs{Traces} (potentially saving a roundtrip to the server).
-To model this change in the code would require storing an additional transient
-piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)},
-storing the computed hashes so far this run. With that available,
-\hs{getHash}~\hs{key}~\hs{s} would become:
-\begin{minted}[xleftmargin=10pt]{haskell}
-hash (key, [ done Map.! d | d <- ds ])
-\end{minted}
-And that new hash would have to be stored in \hs{done}.
+The implementation of \hs{ctRebuilder} is analogous to that of \hs{vtRebuilder}
+in Fig.~\ref{fig-shake-implementation}, but the \hs{verifyVT} query is replaced
+with a more powerful query to \hs{constructCT} that can construct a resulting
+\hs{cachedValue} by looking it up the cloud cache. If the cache lookup fails, we
+fall back to executing the \hs{task}. The obtained \hs{newValue} and the
+\hs{task}'s dependencies are recorded as a new constructive trace for future
+use.
+
+The \Bazel build system uses a restarting scheduler whose implementation we
+omit. It is similar to \Excel's \hs{restarting} scheduler defined in
+Fig.~\ref{fig-excel-implementation}, but instead of building keys in the order
+specified by the persistently stored calculation chain, \Bazel uses a
+\emph{build queue}. The build starts with the queue containing all dirty keys.
+Similar to \Excel, the rebuilding of a key extracted from the queue may fail
+because one of its dynamic dependencies is dirty. In this case the key is marked
+as \emph{blocked} and its rebuilding is deferred. Whenever a key is successfully
+rebuilt, all keys that were previously blocked on it are added back to the
+queue, and their build is eventually restarted.
+
+Note that although both our model and \Bazel's actual implementation supports
+dynamic dependencies, it is currently not possible to define new monadic build
+rules in the language available to users. Instead, users have to rely on a
+(rich) collection of predefined built-in rules, which cover most but not all
+practically important cases.
 
 \subsection{Cloud \Shake}\label{sec-implementation-cloud-shake}
+
+By changing to \Shake's \hs{suspending} scheduler, we obtain an implementation
+of Cloud \Shake.
+
+\textbf{TODO:} A real implementation of Cloud \Shake is now available... It was
+obtained by directly applying the presented ideas.
 
@@
 Fig.~\ref{fig-bazel-cloud-shake-implementation}.
-
-The differences from \hs{bazel} are minor~--~the dependency scheme has changed
-from \hs{topological} to \hs{recursive}, and thus the dependency keys \hs{dk}
-are captured from the action rather than in advance, the transient state has
-gained a list of keys, and the checking calls \hs{fetch} to get the result
-instead of accessing the store directly.
 
@@
 
+\subsection{Buck and Nix}
+
+The program presented above captures certain aspects of \Bazel, but the real
+implementation makes one important additional assumption on \hs{Task} -- namely
+that it is \textit{deterministic}. With that assumption \Bazel is able to create
+the result hash in a trace by hashing together the result hashes of the
+dependencies and the key -- as the mapping between dependencies and results is
+deterministic. As a consequence \Bazel can compute the results of traces locally,
+without looking at \hs{Traces} (potentially saving a roundtrip to the server).
+To model this change in the code would require storing an additional transient
+piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)},
+storing the computed hashes so far this run. With that available,
+\hs{getHash}~\hs{key}~\hs{s} would become:
+\begin{minted}[xleftmargin=10pt]{haskell}
+hash (key, [ done Map.! d | d <- ds ])
+\end{minted}
+And that new hash would have to be stored in \hs{done}.
+
 \vspace{-1mm}

### 2018-06-18 paper/1-intro.tex
Work on S5

@@
   build system that supports both \emph{dynamic dependencies}
-  and \emph{cloud build} (\S\ref{sec-implementation-cloud-shake}).
+  and \emph{cloud build} (\S\ref{sec-implementation-bazel-cloud-shake}).
 

### 2018-06-18 paper/2-background.tex
Work on S5

@@
 advantages of \Shake and \Bazel in a cloud build system with dynamic
-dependencies, see~\S\ref{sec-implementation-cloud-shake}.
+dependencies, see~\S\ref{sec-implementation-bazel-cloud-shake}.

### 2018-06-18 paper/4-build.tex
Work on S5

@@
 to be built. The Cloud \Shake system is an interesting and important point in
-the design space, which we explore further in~\S\ref{sec-implementation-cloud-shake}.
+the design space, which we explore further
+in~\S\ref{sec-implementation-bazel-cloud-shake}.

### 2018-06-18 paper/5-implementations.tex
Work on S5

@@
     deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
-    order  = case topSort (reachable deps target) of
-        Nothing -> error "Cannot build tasks with cyclic dependencies"
-        Just xs -> xs
+    order  = fromJust $ topSort (reachable deps target) -- Assumes acyclic dependencies
 \end{minted}
@@
 reachable :: Ord k => (k -> [k]) -> k -> Graph k
-topSort   :: Ord k => Graph k -> Maybe [k]
+topSort   :: Ord k => Graph k -> Maybe [k] -- Returns Nothing on a cyclic graph
 \end{minted}
-\vspace{-2mm}
+\vspace{-3mm}
 \caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
-\vspace{-4mm}
+\vspace{-3mm}
 \end{figure}
@@
 
-We capture persistent build information that \Make stores by the pair
-\hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
-map \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
+\Make stores the pair \hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTime)}
+as persistent build information. It comprises the \emph{current time} \hs{now}
+and the map \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
 modification times}. The rebuilder receives three arguments: a \hs{key}, its
@@
 \hs{Task}~\hs{(@@MonadState}~\hs{(@@MakeInfo}~\hs{k))}~\hs{k}~\hs{v} that has
-access to the build information and uses it to decide whether to return the
-current \hs{value} as is or rebuild the \hs{key} if it is \hs{dirty}, i.e.
+access to \Make's build information and uses it to decide whether to return the
+current \hs{value} as is, or rebuild the \hs{key} if it is \hs{dirty}, i.e.
 if it is older than any of its dependencies\footnote{Since the \hs{task} is
@@
 \hs{dependencies} defined in~\S\ref{sec-deps}.} or has never been built before
-(in which case \hs{modTime} has no entry for it). Note that in our model we
-update the current time and file modification time whenever the \hs{key} is
+(in which case \hs{modTime} has no entry for it). Note that in our model we need
+to update the current time and file modification time whenever the \hs{key} is
 rebuilt, which in reality is taken care of by the file system.
@@
 \hs{runState} to execute it. The \hs{newTask} finds values of the \hs{key}'s
-dependencies via the \hs{fetch} callback that directly reads the \hs{store}.
+dependencies via the \hs{fetch} callback, which is defined to directly read the
+\hs{store}.
 
@@
 
-Note that \hs{dependencies} is only defined for applicative tasks, which is what
-restricts \Make to static dependencies, as reflected in the
-type~\hs{Build}~\hs{Applicative}. Moreover, any other build system using
-the same \hs{topological} scheduler will also inherit the same restriction.
+Note that the function \hs{dependencies} can only be applied to applicative
+tasks, which restricts \Make to static dependencies, as reflected in the
+type~\hs{Build}~\hs{Applicative}. Moreover, any other build system that uses
+the \hs{topological} scheduler will also inherit the same restriction.
 
@@
 
-Fig.~\ref{fig-excel-implementation} shows an implementation of \Excel in our
-framework as a composition of the \hs{restarting} scheduler and an
-\hs{approximateRebuilder}.
-
-\Excel's persistently stored build information is a triple:
+Our \Excel's model is a composition of the \hs{restarting} scheduler and the
+\hs{approximateRebuilder}, as shown in Fig.~\ref{fig-excel-implementation}.
+Persistently stored build information \hs{ExcelInfo}~\hs{k} is a triple:
 
@@
     \item A set of dirty keys \hs{Set}~\hs{k}, i.e. all cells whose contents
-          has changed since the previous build, including cells that contain
-          input values as well as formulas.
-    \item A map \hs{Map}~\hs{k}~\hs{[@@k]} that stores known upper bounds of
+          has changed since the previous build, including input cells and
+          formulas.
+    \item A map \hs{Map}~\hs{k}~\hs{[@@k]} of known upper bounds of
           dependencies. If there is an entry for a key, it is a conservative
           approximation of its dependencies (e.g. the dependencies of the
-          formula \cmd{IF(C1=1,B2,A2)} can be approximated by the list
+          formula \cmd{IF(C1=1,B2,A2)} are approximated by the list
           \hs{[}\cmd{C1}\hs{,}\cmd{B2}\hs{,}\cmd{A2}\hs{]}). Otherwise, there is
-          no upper bound (e.g. the key uses an \cmd{INDIRECT} reference) and
-          the key is assumed to be dirty.
+          no upper bound (e.g. because the key uses an \cmd{INDIRECT} reference)
+          and the key is assumed to be dirty.
     \item A calculation \hs{Chain} recorded in the previous build
@@
 \hs{dirty} if any of its approximate dependencies is in the \hs{dirtyKeys} set,
-or if there is no known upper bound on dependencies, i.e. if the lookup in the
-map \hs{deps}~\hs{::}~\hs{Map}~\hs{k}~\hs{[@@k]} fails. Note that when a
-\hs{key} is rebuilt it is added to the \hs{dirtyKeys} set to trigger further
+or if there is no known upper bound on its dependencies, i.e. if the lookup in
+the map \hs{deps}~\hs{::}~\hs{Map}~\hs{k}~\hs{[@@k]} fails. Note that when a
+\hs{key} is rebuilt it is added to the set \hs{dirtyKeys} to trigger further
 rebuilds.
@@
 \Excel's \hs{restarting} scheduler processes keys in the order specified by the
-calculation \hs{chain}, constructing a \hs{newChain} for the next build and
-maintaining a set of keys \hs{done} that have already been processed. For each
-non-input \hs{key}, the scheduler tries to rebuild it using a partial \hs{fetch}
-callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}. The callback
-is defined to fail with \hs{Left}~\hs{dep} when asked for the value of a
-dependency \hs{dep} that has not yet been processed (and hence may potentially
-be dirty); otherwise it returns the current value of the dependency in the
-\hs{store}.
+calculation \hs{chain}. During the build, it constructs a \hs{newChain} for the
+next build and maintains a set of keys \hs{done} that have been processed. For
+each non-input \hs{key}, the scheduler tries to rebuild it using a partial
+\hs{fetch} callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}.
+The callback is defined to fail with \hs{Left}~\hs{dep} when asked for the value
+of a dependency \hs{dep} that has not yet been processed (and hence may
+potentially be dirty); otherwise it returns the current value of the dependency
+by looking it up in the \hs{store}.
 
@@
     \item The \hs{newTask} succeeded. The resulting \hs{newValue} and
-    \hs{newInto} are written to the store, and \Excel continues to build the
-    \hs{chain} after marking the \hs{key} as \hs{done}.
+    \hs{newInfo} are written to the store, the \hs{key} is marked as \hs{done},
+    and \Excel continues to build the \hs{chain}.
 \end{itemize}
@@
 \end{minted}
-\vspace{-2mm}
+\vspace{-3mm}
 \caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
@@
 as persistent build information and is composed of the \hs{suspending} scheduler
-and \hs{vtRebuilder}.
+and the \hs{vtRebuilder}.
 
@@
                     fetch :: k -> StateT i (State (Store i k v, Set k)) v
-                    fetch k = do lift (build k)                      -- build the key
-                                 lift (gets (getInfo . fst)) >>= put -- save new traces
-                                 lift (gets (getValue k . fst))      -- fetch the value
+                    fetch k = do lift (build k)                      -- Build the key
+                                 lift (gets (getInfo . fst)) >>= put -- Save new traces
+                                 lift (gets (getValue k . fst))      -- Fetch the value
                 info <- gets (getInfo . fst)
@@
 \end{minted}
-\vspace{-2mm}
+\vspace{-3mm}
 \caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
@@
     \item Save any new traces that have been generated when building \hs{k}.
-    \item Read the up to date value of \hs{k} from the store and return it.
+    \item Read the (up to date) value of \hs{k} from the store and return it.
 \end{itemize}
@@
 monad, because the \hs{newTask} has access only to the~\hs{i} part of the state;
-we use \hs{lift} to execute commands in the inner monad.
+we therefore need to use \hs{lift} when executing commands in the inner
+\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
+The pair \hs{(@@newValue,}~\hs{newInfo)} obtained by running the \hs{newTask} is
+stored, and the \hs{key} is added to the set \hs{done}.
 
-Finally, the \hs{newValue} and \hs{newInfo} obtained by running the \hs{newTask}
-are stored, and \hs{key} is added to the set \hs{done}.
-
-\subsection{\Bazel}\label{sec-implementation-bazel}
+\subsection{Cloud build systems: \Bazel and \Cloud \Shake}\label{sec-implementation-bazel-cloud-shake}
 
@@
 \vspace{-2mm}
-\caption{Implementations of \Bazel and Cloud \Shake using our framework;
+\caption{Implementations of \Bazel and \Cloud \Shake using our framework;
 \hs{restarting2} differs from the restarting scheduler used by Excel
@@
 
-In this section we show how constructive traces can be used to implement cloud
-build systems. Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models
-of \Bazel and Cloud \Shake, which both rely on \hs{ctRebuilder} -- a
-rebuilder based on constructive traces (\S\ref{sec-constructive-traces}).
+In this section we use constructive traces to implement cloud build systems.
+Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models of \Bazel and
+\Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a rebuilder based on
+constructive traces (\S\ref{sec-constructive-traces}).
 
@@
 (rich) collection of predefined built-in rules, which cover most but not all
-practically important cases.
+practically important cases of dynamic dependencies.
 
-\subsection{Cloud \Shake}\label{sec-implementation-cloud-shake}
-
-By changing to \Shake's \hs{suspending} scheduler, we obtain an implementation
-of Cloud \Shake.
+Using the abstractions built thus far, we have shown how to combine schedulers
+with rebuilders to reproduce existing build systems. To us, the most interesting
+build system as yet unavailable would combine a suspending scheduler with
+constructive traces~--~providing a cloud-capable build system that is minimal,
+and supports both early cutoff and monadic dependencies. Using our framework it
+is possible to define and test such a system, which we call \Cloud \Shake. All
+we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
+of \Bazel, as shown in Fig.~\ref{fig-bazel-cloud-shake-implementation}.
 
@@
 obtained by directly applying the presented ideas.
-
-Using the abstractions and approaches built thus far, we have shown how to
-combine dependency scheme and change approach to reproduce existing build
-systems. In the attached materials we have implemented 9~build systems
-corresponding to all three dependency schemes, matched with all three change
-approaches. To us, the most interesting build system as yet unavailable would
-matching recursive ordering with constructive traces~--~providing a cloud-capable
-build system with minimality, cutoff and monadic dependencies. Using our
-framework it is possible to define and test such a system as per
-Fig.~\ref{fig-bazel-cloud-shake-implementation}.
 

### 2018-06-18 paper/7-related.tex
Add authors, minor changes
Fix paper build
Work on S3
Work on S5

@@
 differs from \Bazel by allowing non-determinism~\S\ref{sec-non-determinism}, thus more closely modelling
-our original definition of \Bazel from \S\ref{sec-implementation-bazel}.
+our original definition of \Bazel from \S\ref{sec-implementation-bazel-cloud-shake}.
 
@@
 combined with \cmd{import-from-derivation}, \Nix can also be considered monadic,
-making it similar to Cloud \Shake from~\S\ref{sec-implementation-cloud-shake}.
+making it similar to \Cloud \Shake
+from~\S\ref{sec-implementation-bazel-cloud-shake}.
 However, \Nix is not intended as a build system, and the coarse grained nature
@@
 
+%TODO: Add a reference to the paper about forall f. c f => (k -> f v) -> f a
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-
-Provided we instantiate \hs{p} to something like
+\noindent
+provided we instantiate \hs{p} to something like
 \hs{k}~\hs{->}~\hs{f}~\hs{v}~--~which many of the actual instances in that paper
@@
 is an input, without complicating other aspects of our definition. Furthermore,
-the \hs{Build} operation lacks any further such symmetry.
+the \hs{Build} abstraction lacks any further such symmetry.

### 2018-06-19 paper/1-intro.tex
Finish S5

--- 2018-05-03 paper/1-intro.tex
@@
   build system that supports both \emph{dynamic dependencies}
-  and \emph{cloud build} (\S\ref{sec-implementation-cloud-shake}).
+  and \emph{cloud build} (\S\ref{sec-implementation-cloud}).
 

### 2018-06-19 paper/2-background.tex
Finish S5

--- 2018-06-16 paper/2-background.tex
@@
 advantages of \Shake and \Bazel in a cloud build system with dynamic
-dependencies, see~\S\ref{sec-implementation-cloud-shake}.
+dependencies, see~\S\ref{sec-implementation-cloud}.

### 2018-06-19 paper/4-build.tex
Finish S5

--- 2018-06-15 paper/4-build.tex
@@
 
+\subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}
+
+In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list
+of traces~--~however, using a list necessarily means that finding the right trace
+takes $O(n)$. For each of the \hs{Trace} based systems it is possible to devise
+a smarter representation, which we sketch below. Note that these implementations
+do not avoid calls to \hs{compute}, merely overheads in the build system itself.
+
+\begin{enumerate}
+\item Any system using verifying traces, e.g.
+\Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}.
+Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
+\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
+\item Any \hs{Applicative} build system storing constructive traces, e.g.
+\Bazel, can index directly from the key and results to the output result~--~i.e.
+\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{(Hash}~\hs{v)}.
+Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash.
+In this formulation we have removed the possibility for a single key/dependency state to map to multiple different hashes, e.g.
+on a non-deterministic build~--~something \Bazel already prohibits which is discussed more in~\S\ref{sec-non-determinism}.
+\item Finally, a \hs{Monad} build system with constructive traces can be stored as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Choice k v = Choice k (Map (Hash v) (Choice k v))
+                | Result (Hash v)
+\end{minted}
+Here the \hs{Choice} encodes a tree, asking successive questions about keys, and taking different branches based on the answers, until it reaches a final result.
+Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
+\end{enumerate}
+
 \subsection{Build Systems \`a la Carte}\label{sec-design-space}
@@
 to be built. The Cloud \Shake system is an interesting and important point in
-the design space, which we explore further in~\S\ref{sec-implementation-cloud-shake}.
+the design space, which we explore further
+in~\S\ref{sec-implementation-cloud}.

### 2018-06-19 paper/5-implementations.tex
Finish S5

@@
 
-\subsection{Cloud build systems: \Bazel and \Cloud \Shake}\label{sec-implementation-bazel-cloud-shake}
+\subsection{Cloud build systems: \Bazel, \CloudBuild and \Cloud \Shake}\label{sec-implementation-cloud}
 
@@
 bazel = restarting2 ctRebuilder
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Cloud Shake build system, implementation of 'suspending' is given in Fig. 8
-cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
-cloudShake = suspending ctRebuilder
 \end{minted}
@@
 \end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- CloudBuild build system, implementation of 'topological' is given in Fig. 6
+cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
+cloudBuild = topological (adaptRebuilder ctRebuilder)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Convert a monadic rebuilder to the corresponding applicative one
+adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
+adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 8
+cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
+cloudShake = suspending ctRebuilder
+\end{minted}
 \vspace{-2mm}
-\caption{Implementations of \Bazel and \Cloud \Shake using our framework;
-\hs{restarting2} differs from the restarting scheduler used by Excel
+\caption{Implementations of \Bazel, \CloudBuild and \Cloud \Shake using our
+framework; \hs{restarting2} differs from the restarting scheduler used by \Excel
 (Fig.~\ref{fig-excel-implementation}) in that it does not rely on the
 calculation chain.}
-% \hs{topological} is defined in Fig.~\ref{fig-make-implementation}.
 \label{fig-bazel-cloud-shake-implementation}
-\vspace{-2mm}
+\vspace{-4mm}
 \end{figure}
@@
 In this section we use constructive traces to implement cloud build systems.
-Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models of \Bazel and
-\Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a rebuilder based on
-constructive traces (\S\ref{sec-constructive-traces}).
+Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models of \Bazel,
+\CloudBuild and \Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a
+rebuilder based on constructive traces (\S\ref{sec-constructive-traces}).
 
@@
 rules in the language available to users. Instead, users have to rely on a
-(rich) collection of predefined built-in rules, which cover most but not all
+(rich) collection of predefined built-in rules, which cover many but not all
 practically important cases of dynamic dependencies.
+
+By switching to the \hs{topological} scheduler, we obtain a model of
+Microsoft's \CloudBuild~-- an applicative build system that combines
+conventional scheduling of statically known directed acyclic graphs of
+dependencies with constructive traces~\cite{esfahani2016cloudbuild}. Note that
+we need to convert a monadic \hs{ctRebuilder} into an applicative one by
+applying an adapter function \hs{adaptRebuilder}, which unwraps a given
+\hs{Task Applicative} and wraps it into \hs{Task Monad}.
 
@@
 
-\textbf{TODO:} A real implementation of Cloud \Shake is now available... It was
+\textbf{TODO:} A real implementation of \Cloud \Shake is now available... It was
 obtained by directly applying the presented ideas.
@@
 buck :: (Hashable k, Hashable v) => Build Applicative (DCT k v) k v
-buck = topological (unliftRebuilder dctRebuilder)
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Convert a monadic rebuilder to the corresponding applicative one
-unliftRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
-unliftRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
+buck = topological (adaptRebuilder dctRebuilder)
 \end{minted}
@@
         Nothing -> do
-            (newValue, deps) <- trackM task fetch
+            (newValue, deps) <- track task fetch
             put =<< recordDCT key newValue deps (fmap hash . fetch) =<< get
@@
 
-The program presented above captures certain aspects of \Bazel, but the real
-implementation makes one important additional assumption on \hs{Task} -- namely
-that it is \textit{deterministic}. With that assumption \Bazel is able to create
-the result hash in a trace by hashing together the result hashes of the
-dependencies and the key -- as the mapping between dependencies and results is
-deterministic. As a consequence \Bazel can compute the results of traces locally,
-without looking at \hs{Traces} (potentially saving a roundtrip to the server).
-To model this change in the code would require storing an additional transient
-piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)},
-storing the computed hashes so far this run. With that available,
-\hs{getHash}~\hs{key}~\hs{s} would become:
-\begin{minted}[xleftmargin=10pt]{haskell}
-hash (key, [ done Map.! d | d <- ds ])
-\end{minted}
-And that new hash would have to be stored in \hs{done}.
+Some build systems allow to exploit \emph{build task determinism} for efficient
+fetching of build results from the cache: one can simply hash the values of all
+transitive inputs of a key (which can be done efficiently by storing all hashes
+in a Merkle tree) and download the matching value from the cache, without the
+need to build any intermediate values. This can bring significant performance
+benefits in practice, but sacrifices the early cutoff property: indeed, with
+this approach addition of a new comment to one of the source files will
+invalidate all dependent keys.
 
-\vspace{-1mm}
-\subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}
-\vspace{-1mm}
-
-In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list
-of traces~--~however, using a list necessarily means that finding the right trace
-takes $O(n)$. For each of the \hs{Trace} based systems it is possible to devise
-a smarter representation, which we sketch below. Note that these implementations
-do not avoid calls to \hs{compute}, merely overheads in the build system itself.
-
-\begin{enumerate}
-\item Any system using verifying traces, e.g.
-\Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}.
-Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
-\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
-\item Any \hs{Applicative} build system storing constructive traces, e.g.
-\Bazel, can index directly from the key and results to the output result~--~i.e.
-\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{(Hash}~\hs{v)}.
-Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash.
-In this formulation we have removed the possibility for a single key/dependency state to map to multiple different hashes, e.g.
-on a non-deterministic build~--~something \Bazel already prohibits which is discussed more in~\S\ref{sec-non-determinism}.
-\item Finally, a \hs{Monad} build system with constructive traces can be stored as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
-\begin{minted}[xleftmargin=10pt]{haskell}
-data Choice k v = Choice k (Map (Hash v) (Choice k v))
-                | Result (Hash v)
-\end{minted}
-Here the \hs{Choice} encodes a tree, asking successive questions about keys, and taking different branches based on the answers, until it reaches a final result.
-Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
-\end{enumerate}
+Fig.~\ref{fig-buck-nix-implementation} shows our models of \Buck and \Nix build
+systems that use the \hs{dctRebuilder} -- a rebuilder based on deterministic
+constructive traces. As one can see, \hs{dctRebuilder} is almost identical to
+\hs{ctRebuilder}, but it ignores the current \hs{value}, since the
+\hs{constructDCT} query only needs to examine the transitive inputs of the given
+\hs{key}. \Buck uses the \hs{topological} scheduler and is therefore an
+applicative build system, whereas \Nix uses the \hs{suspending} scheduler and is
+monadic.

### 2018-06-19 paper/7-related.tex
Finish S5

@@
 differs from \Bazel by allowing non-determinism~\S\ref{sec-non-determinism}, thus more closely modelling
-our original definition of \Bazel from \S\ref{sec-implementation-bazel-cloud-shake}.
+our original definition of \Bazel from \S\ref{sec-implementation-cloud}.
 
@@
 making it similar to \Cloud \Shake
-from~\S\ref{sec-implementation-bazel-cloud-shake}.
+from~\S\ref{sec-implementation-cloud}.
 However, \Nix is not intended as a build system, and the coarse grained nature

### 2018-06-19 paper/4-build.tex
Expand the table

@@
 
+\subsubsection{Deterministic constructive traces}\label{sec-deterministic-constructive-traces}
+
+\textbf{TODO}
+
 \subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}
@@
 \hline
-Dirty bit\hfill\S\ref{sec-dirty-bit}                    & \Make    & \Excel                   & Approximate \Shake*  \\\hline
-Verifying trace\hfill\S\ref{sec-verifying-traces}       & \Ninja   & Traced \Excel*            & \Shake                      \\\hline
-Constructive trace\hspace{2mm}\hfill\S\ref{sec-constructive-traces} & \Bazel   & Cloud \Excel*             & Cloud \Shake*                \\\hline
+Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
+Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
+Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & \Cloud \Shake  \\\hline
+Deterministic constructive traces\hspace{2mm}\hfill\S\ref{sec-deterministic-constructive-traces} & \Buck       & -      & \Nix           \\\hline
 \end{tabular}
 \vspace{2mm}
-\caption{Build systems \`a la carte. Systems marked * are hypothetical systems that do not currently exist.\label{tab-build-systems}}
+\caption{Build systems \`a la carte.\label{tab-build-systems}}
 \vspace{-2mm}

### 2018-06-19 paper/5-implementations.tex
Expand the table

@@
 systems that use the \hs{dctRebuilder} -- a rebuilder based on deterministic
-constructive traces. As one can see, \hs{dctRebuilder} is almost identical to
-\hs{ctRebuilder}, but it ignores the current \hs{value}, since the
-\hs{constructDCT} query only needs to examine the transitive inputs of the given
-\hs{key}. \Buck uses the \hs{topological} scheduler and is therefore an
-applicative build system, whereas \Nix uses the \hs{suspending} scheduler and is
-monadic.
+constructive traces (\S\ref{sec-deterministic-constructive-traces}). As one can
+see, \hs{dctRebuilder} is almost identical to \hs{ctRebuilder}, but it ignores
+the current \hs{value}, since the \hs{constructDCT} query only needs to examine
+the transitive inputs of the given \hs{key}. \Buck uses the \hs{topological}
+scheduler and is therefore an applicative build system, whereas \Nix uses the
+\hs{suspending} scheduler and is monadic.

### 2018-06-20 paper/5-implementations.tex
Bit at the start of 5

@@
 In the previous sections we discussed the types of build systems, and how they
-can be broken down into two main components: a scheduler and a rebuilder. This
-decomposition was not obvious to us, and only by concretely implementing and
-refactoring each build system did we determine the underlying commonalities.
-In this section we share some of the code that got us there.
+can be broken down into two main components: a scheduler and a rebuilder.
+In this section we make this abstract distinction concrete, by actually
+implementing a number of build systems as a composition of a schedulear
+and a rebuilder.  Our design is expressed by these types:
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Scheduler c i k v = Rebuilder c i k v -> Build c k v
+type Rebuilder c i k v = k -> v -> Task c k v -> Task (MonadState i) k v
+\end{minted}
+That is:
+\begin{itemize}
+\item A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
+  it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
+\item A \hs{Rebuilder} is a function that is given a key, the key's ``old'' value,
+  and a \hs{Task} that can compute the value of the key if necessary. It uses
+  the persistent build state \hs{i} (carried by the state monad) to decide
+  whether to re-build the value.  If doing so is unnecesary, it returns the
+  old value; otherwise it runs the supplied \hs{Task} to rebuild it.  In both
+  cases it updates the persistent build information \hs{i} to reflect what happened.
+\end{itemize}
 

### 2018-06-20 paper/4-build.tex
Wibbles to start of 5

@@
 dependency order strategy with the out-of-date keys strategy, providing 9~possible
-build systems, 5~of which are actually inhabited by existing build systems
+build systems \simon{Or perhaps 12 now?}, 5~of which are actually inhabited by existing build systems
 (we discuss \Ninja \cite{ninja} in \S\ref{sec-related-build}). Of the

### 2018-06-20 paper/5-implementations.tex
Wibbles to start of 5

@@
 can be broken down into two main components: a scheduler and a rebuilder.
-In this section we make this abstract distinction concrete, by actually
-implementing a number of build systems as a composition of a schedulear
+In this section we make this abstract distinction concrete, by
+implementing a number of build systems as a composition of a scheduler
 and a rebuilder.  Our design is expressed by these types:
@@
 \end{itemize}
+We write a scheduler for each column of Table~\ref{tab-build-systems}, and a
+rebuilder for each row; then applying the former to the latter, in any combination,
+gives a correct build system.
 
@@
 
-\Make's scheduler processes keys in a linear \hs{order} based on a
-\hs{topological} sort of the statically known dependency graph
+\Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based on a
+topological sort of the statically known dependency graph
 (see~\S\ref{sec-parallelism} for parallel \Make). Our definition in

### 2018-06-21 paper/2-background.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

@@
 
-To support dynamic dependencies,
-\Excel's build algorithm~\cite{excel_recalc} is significantly different from
-\Make. \Excel uses the \emph{calculation chain}
-produced by the previous build as an approximation to the correct topological
-order. During recalculation, \Excel processes cells in this order, but can
-\emph{defer recalculation of a cell} by moving it down the chain if a newly
-discovered dependency has not yet been rebuilt. We refer to this
-algorithm as \emph{restarting}, and will discuss it in more detail
-in~\S\ref{sec-implementation-excel}.
+To support dynamic dependencies, \Excel's build
+algorithm~\cite{excel_recalc} is significantly different from
+\Make. \Excel{} arranges the cells into a linear sequence, called the
+\emph{calc chain}.  During recalculation, \Excel{} processes
+cells in the calc-chain sequence, but if computing a cell C requires the
+value of a cell D that has not yet beeen computed, \Excel{}
+\emph{aborts} computation of C, moves D before C in the calc chain,
+and resumes recalculation starting with D.  When a calc cycle is
+complete, the calc chain respects all the dynamic dependencies of the
+spreadsheet.  When an input value (or formula) is changed, \Excel{}
+uses the final calc chain from the \emph{previous} calc cycle as its
+starting point so that, in the common case where changing an input
+value does not change dependencies, there are no aborts.  Notice
+that calc always succeeds regardless of the initial calc chain
+(barring truly circular dependencies); the calc chain is just an optimisation.
+We refer to this algorithm as \emph{restarting}, and will discuss it
+in more detail in~\S\ref{sec-implementation-excel}.
 
@@
 Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
-systems only track changes of inputs and intermediate results, but \Excel can
-also track changes in the tasks themselves: if a formula is modified, \Excel
+systems only track changes of inputs and intermediate results, but \Excel
+also tracks changes in the tasks themselves: if a formula is modified, \Excel
 will recompute it and propagate the changes. Self-tracking is uncommon in

### 2018-06-21 paper/5-implementations.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-type Scheduler c i k v = Rebuilder c i k v -> Build c k v
+type Scheduler c is ib k v = Rebuilder c ib k v -> Build c is k v
 type Rebuilder c i k v = k -> v -> Task c k v -> Task (MonadState i) k v
 \end{minted}
-That is:
-\begin{itemize}
-\item A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
+So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
 it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
-\item A \hs{Rebuilder} is a function that is given a key, the key's ``old'' value,
+The \hs{Rebuilder} makes uses of the persistent build information \hs{ib}, while
+the scheduler might augment that with further persistent information of its own,
+yielding \hs{is}.
+
+A \hs{Rebuilder} is a function which the \hs{Scheduler} uses to rebuild
+a particular key, passing to it the key, the key's ``old'' value,
 and a \hs{Task} that can compute the value of the key if necessary. It uses
@@
 cases it updates the persistent build information \hs{i} to reflect what happened.
-\end{itemize}
-We write a scheduler for each column of Table~\ref{tab-build-systems}, and a
-rebuilder for each row; then applying the former to the latter, in any combination,
-gives a correct build system.
+So a \hs{Rebuilder} wraps a \hs{Task c k v}, which unconditionally rebulds the key,
+to make a \hs{Task (Monadstate i) k v}, which rebuilds the key only if necessary, and
+does the necessary book-keeping. \simon{I'm puzzled by the absence of 'c' in the return type.}
+
+These two abstractions are the key to modularity: \emph{we can combine any scheduler
+  with any rebuilder, and obtain a correct build system}.
+In this section we will write a scheduler for each column of Table~\ref{tab-build-systems},
+and a rebuilder for each row; then combine them to obtain the build systems in the
+body of the Table.
 
@@
 modTimeRebuilder key value task = Task $ \fetch -> do
-    (now, modTime) <- get
-    let dirty = case Map.lookup key modTime of
+    (now, modTimes) <- get
+    let dirty = case Map.lookup key modTimes of
             Nothing -> True
-            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
+            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
     if not dirty then return value else do
-        put (now + 1, Map.insert key now modTime)
+        put (now + 1, Map.insert key now modTimes)
         run task fetch
@@
 -- A topological task scheduler
-topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
+topological :: Ord k => Scheduler Applicative i i k v
 topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
@@
 An implementation of \Make using our framework is shown in
-Fig.~\ref{fig-make-implementation}. The top-level definition
-\hs{make}~\hs{=}~\hs{topological}~\hs{modTimeRebuilder} is a direct translation
-of the scheduler-rebuilder decomposition in Table~\ref{tab-build-systems}. Below
-we describe both components, starting with \hs{modTimeRebuilder}.
+Fig.~\ref{fig-make-implementation}. As promised, its definition
+is just the application of a \hs{Scheduler}, \hs{topological},
+to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
+We discuss each component in turn, starting with the rebuilder.
 
-\Make stores the pair \hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTime)}
-as persistent build information. It comprises the \emph{current time} \hs{now}
-and the map \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
-modification times}. The rebuilder receives three arguments: a \hs{key}, its
-current \hs{value}, and the \hs{task} of type
-\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild the
-\hs{key} if necessary. The rebuilder returns a modified monadic task of type
-\hs{Task}~\hs{(@@MonadState}~\hs{(@@MakeInfo}~\hs{k))}~\hs{k}~\hs{v} that has
-access to \Make's build information and uses it to decide whether to return the
-current \hs{value} as is, or rebuild the \hs{key} if it is \hs{dirty}, i.e.
-if it is older than any of its dependencies\footnote{Since the \hs{task} is
-applicative we can statically find its dependencies using the function
-\hs{dependencies} defined in~\S\ref{sec-deps}.} or has never been built before
-(in which case \hs{modTime} has no entry for it). Note that in our model we need
-to update the current time and file modification time whenever the \hs{key} is
-rebuilt, which in reality is taken care of by the file system.
+The \hs{modTimeRebuilder} uses the pair
+\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
+build information, carried by a \hs{MonadState} monad. This
+\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
+\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
+modification times}. The rebuilder receives three arguments: a
+\hs{key}, its current \hs{value}, and the \hs{task} of type
+\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
+the \hs{key} if necessary.  The rebuilder first decides if the
+\hs{key} is \hs{dirty} by consulting \hs{modTime}: if the key is not
+found, that must mean it has never been built before; otherwise
+\hs{modTimeRebuilder} can see if any of the task's dependencies
+(computed by \hs{dependencies}) are out of date.  If the key is
+\hs{dirty}, we use `run task` to rebuild it, and update the state with
+the new modification time of the key\footnote{The real \hs{make}
+relies on the file system to track modification times, but we prefer
+to make it explicit here.}; othewise we can just return the old vale.
 
@@
 \hs{approximateRebuilder}, as shown in Fig.~\ref{fig-excel-implementation}.
-Persistently stored build information \hs{ExcelInfo}~\hs{k} is a triple:
+The persistent build information \hs{ExcelInfo}~\hs{k} is a triple:
 
@@
 -- A restarting task scheduler
-restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
+restarting :: Ord k => Scheduler Monad i (i, Chain k) k v
 restarting rebuilder tasks target = execState $ do
@@
 -- A suspending task scheduler
-suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
+suspending :: Ord k => Scheduler Monad i i k v
 suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)

### 2018-06-21 paper/2-background.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

@@
 \Make. \Excel{} arranges the cells into a linear sequence, called the
-\emph{calc chain}.  During recalculation, \Excel{} processes
-cells in the calc-chain sequence, but if computing a cell C requires the
+\emph{calculation chain}.  During recalculation, \Excel{} processes
+cells the calc-chain sequence, but if computing a cell C requires the
 value of a cell D that has not yet beeen computed, \Excel{}
@@
 uses the final calc chain from the \emph{previous} calc cycle as its
-starting point so that, in the common case where changing an input
-value does not change dependencies, there are no aborts.  Notice
+starting point so that, in the common case where chaning an input
+value does not change dependencies, there are zero aborts.  Notice
 that calc always succeeds regardless of the initial calc chain
-(barring truly circular dependencies); the calc chain is just an optimisation.
+(barring truly circular dependencies); the calc chain is ``just'' an optimisation.
 We refer to this algorithm as \emph{restarting}, and will discuss it

### 2018-06-21 paper/3-abstractions.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

@@
 modify    :: (s -> s) -> State s ()
+runState  :: State s a -> s -> (a, s)
 execState :: State s a -> s -> s
@@
 running its code in a different \hs{Applicative}! This is not new, for example
-see~\cite{free-applicatives}, but it is cool.
+see~\cite{free-applicatives}, but it is etremely cool.  We will see a practical
+use for \hs{dependencies} in \S\ref{sec-implementation-make}.
 
@@
 which in this case is obtained by reading from the standard input.
+We will see a use for \hs{track} in \S\ref{sec-implementation-shake}.
 

### 2018-06-21 paper/5-implementations.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

@@
 
-These two abstractions are the key to modularity: \emph{we can combine any scheduler
-  with any rebuilder, and obtain a correct build system}.
-In this section we will write a scheduler for each column of Table~\ref{tab-build-systems},
-and a rebuilder for each row; then combine them to obtain the build systems in the
-body of the Table.
+We write a scheduler for each column of Table~\ref{tab-build-systems}, and a
+rebuilder for each row; then applying the former to the latter, in any combination,
+gives a correct build system.
 
@@
 to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
-We discuss each component in turn, starting with the rebuilder.
+of the scheduler-rebuilder decomposition in Table~\ref{tab-build-systems}. We discuss
+each component in turn, starting with the rebuilder.
 
-The \hs{modTimeRebuilder} uses the pair
+The rebuilder \hs{modTimeRebuilder} uses the pair
 \hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent

### 2018-06-23 src/Build/Rebuilder.hs
Minor revision

@@ -43,14 +43,14 @@ type MakeInfo k = (Time, Map k Time)
 -- needs to be rebuilt. Used by Make.
 modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
 modTimeRebuilder key value task = Task $ \fetch -> do
-    (now, modTime) <- get
-    let dirty = case Map.lookup key modTime of
+    (now, modTimes) <- get
+    let dirty = case Map.lookup key modTimes of
             Nothing -> True
-            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
+            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
     if not dirty
     then return value
     else do
-        put (now + 1, Map.insert key now modTime)
+        put (now + 1, Map.insert key now modTimes)
         run task fetch
 
 --------------------------- Approximate dependencies ---------------------------

### 2018-06-23 src/Build/Scheduler.hs

@@ -75,22 +75,22 @@ type Chain k = [k]
 -- changed and a new dependency is still dirty, the corresponding build task is
 -- abandoned and the key is moved at the end of the calculation chain, so it can
 -- be restarted when all its dependencies are up to date.
-restarting :: forall i k v. Ord k => Scheduler Monad (i, Chain k) i k v
+restarting :: forall ir k v. Ord k => Scheduler Monad (ir, Chain k) ir k v
 restarting rebuilder tasks target = execState $ do
     chain    <- gets (snd . getInfo)
     newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
-    modify . mapInfo $ \(i, _) -> (i, newChain)
+    modify . mapInfo $ \(ir, _) -> (ir, newChain)
   where
-    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
+    go :: Set k -> Chain k -> State (Store (ir, Chain k) k v) (Chain k)
     go _    []       = return []
     go done (key:ks) = case tasks key of
         Nothing -> (key :) <$> go (Set.insert key done) ks
         Just task -> do
             store <- get
             let value = getValue key store
-                newTask :: Task (MonadState i) k (Either k v)
+                newTask :: Task (MonadState ir) k (Either k v)
                 newTask = try $ rebuilder key value task
-                fetch :: k -> State i (Either k v)
+                fetch :: k -> State ir (Either k v)
                 fetch k | k `Set.member` done = return $ Right (getValue k store)
                         | otherwise           = return $ Left k
             case runState (run newTask fetch) (fst $ getInfo store) of

### 2018-06-23 paper/5-implementations.tex
Work on S5

--- 2018-06-21 paper/5-implementations.tex
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-type Scheduler c is ib k v = Rebuilder c ib k v -> Build c is k v
-type Rebuilder c i k v = k -> v -> Task c k v -> Task (MonadState i) k v
+type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
+type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
 \end{minted}
@@
 it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
-The \hs{Rebuilder} makes uses of the persistent build information \hs{ib}, while
+The \hs{Rebuilder} makes uses of the persistent build information \hs{ir}, while
 the scheduler might augment that with further persistent information of its own,
-yielding \hs{is}.
+yielding \hs{i}.
 
-A \hs{Rebuilder} is a function which the \hs{Scheduler} uses to rebuild
-a particular key, passing to it the key, the key's ``old'' value,
-and a \hs{Task} that can compute the value of the key if necessary. It uses
-the persistent build state \hs{i} (carried by the state monad) to decide
-whether to re-build the value.  If doing so is unnecesary, it returns the
-old value; otherwise it runs the supplied \hs{Task} to rebuild it.  In both
-cases it updates the persistent build information \hs{i} to reflect what happened.
-So a \hs{Rebuilder} wraps a \hs{Task c k v}, which unconditionally rebulds the key,
-to make a \hs{Task (Monadstate i) k v}, which rebuilds the key only if necessary, and
-does the necessary book-keeping. \simon{I'm puzzled by the absence of 'c' in the return type.}
+A \hs{Rebuilder} is a function which the \hs{Scheduler} uses to rebuild a
+particular key, passing to it the key, the key's current value, and a \hs{Task}
+that can compute the value of the key if necessary. It uses the persistent build
+information \hs{ir} (carried by the state monad) to decide whether to rebuild
+the value. If doing so is unnecessary, it returns the current value; otherwise
+it runs the supplied \hs{Task} to rebuild it. In both cases it can choose to
+update the persistent build information \hs{ir} to reflect what happened. So a
+\hs{Rebuilder} wraps a \hs{Task}~\hs{c}~\hs{k}~\hs{v}, which unconditionally
+rebuilds the key, to make a \hs{Task}~\hs{(MonadState}~\hs{ir)}~\hs{k}~\hs{v},
+which rebuilds the key only if necessary, and does the necessary book-keeping.
+Note that the resulting \hs{Task} is always monadic; static dependency analysis
+can be performed on the original \hs{Task}~\hs{Applicative} if need be.
 
-These two abstractions are the key to modularity: \emph{we can combine any scheduler
-  with any rebuilder, and obtain a correct build system}.
-In this section we will write a scheduler for each column of Table~\ref{tab-build-systems},
-and a rebuilder for each row; then combine them to obtain the build systems in the
-body of the Table.
+These two abstractions are the key to modularity: \emph{we can combine any
+scheduler with any rebuilder, and obtain a correct build system}.
+In this section we will write a scheduler for each column of
+Table~\ref{tab-build-systems}, and a rebuilder for each row; then combine them
+to obtain the build systems in the table's body.
 
@@
 the \hs{key} if necessary.  The rebuilder first decides if the
-\hs{key} is \hs{dirty} by consulting \hs{modTime}: if the key is not
+\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the key is not
 found, that must mean it has never been built before; otherwise
-\hs{modTimeRebuilder} can see if any of the task's dependencies
+\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
 (computed by \hs{dependencies}) are out of date. If the key is
-\hs{dirty}, we use `run task` to rebuild it, and update the state with
-the new modification time of the key\footnote{The real \hs{make}
+\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
+the new modification time of the key\footnote{The real \Make
 relies on the file system to track modification times, but we prefer
-to make it explicit here.}; othewise we can just return the old vale.
+to make it explicit here.}; otherwise we can just return the current value.
+We assume that the external system, which invokes the build system, updates
+\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.
 
@@
 -- A restarting task scheduler
-restarting :: Ord k => Scheduler Monad i (i, Chain k) k v
+restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
 restarting rebuilder tasks target = execState $ do
@@
     newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
-    modify . mapInfo $ \(i, _) -> (i, newChain)
+    modify . mapInfo $ \(ir, _) -> (ir, newChain)
   where
-    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
+    go :: Set k -> Chain k -> State (Store (ir, Chain k) k v) (Chain k)
     go _    [@@]       = return [@@]
@@
             store <- get
-            let newTask :: Task (MonadState i) k (Either k v)
+            let newTask :: Task (MonadState ir) k (Either k v)
                 newTask = try $ rebuilder key (getValue key store) task
-                fetch :: k -> State i (Either k v)
+                fetch :: k -> State ir (Either k v)
                 fetch k | k `Set.member` done = return $ Right (getValue k store)

### 2018-06-28 paper/2-background.tex
Work on S2

--- 2018-06-19 paper/2-background.tex
@@
 
-Build systems automate the execution of simple repeatable tasks for individual
-users, as well as for large organisations. In this section we explore the design
-space of build systems, using four concrete examples:
-\Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake},
-\Bazel~\cite{bazel}, and \Excel~\cite{advanced_excel}\footnote{\Excel appears
-very different to the others but, seen through the lens of this paper, it is
-very close indeed.}.
-We have carefully chosen these four to illustrate the various axes on which build systems
-differ; we discuss many other notable examples
-of build systems, and their relationships,
-in~\S\ref{sec-related}.
+Build systems automate the execution of repeatable tasks for individual users
+and large organisations. In this section we explore the design space of build
+systems, using four concrete examples: \Make~\cite{feldman1979make},
+\Shake~\cite{mitchell2012shake}, \Bazel~\cite{bazel}, and
+\Excel~\cite{advanced_excel}\footnote{\Excel appears very different to the
+others but, seen through the lens of this paper, it is very close indeed.}.
+We have carefully chosen these four to illustrate the various axes on which
+build systems differ; we discuss many other notable examples of build systems,
+and their relationships, in~\S\ref{sec-related}.
 
@@
 referred to as \emph{build rules}) and their dependencies in a simple textual form.
-For example\footnote{Do not worry about the dependency on \cmd{stdio.h}, and
-its transitive (perhaps dynamic) dependencies.
-We will deal with these in Section~\S\ref{sec-task-monad}.}:
+For example:
 
@@
 files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
-executing\footnote{In this example we pretend \cmd{gcc} is a pure function for the
-sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and the
-actual binary that is used to compile and link files is often also listed as a task
-dependency.} the command \cmd{gcc -c util.c}, (ii) compile the main source file
-\cmd{main.c} into \cmd{main.o}, and (iii) link object files \cmd{util.o} and
-\cmd{main.o} into the executable \cmd{main.exe}. The makefile contains the
-complete information about the \emph{task dependency graph}, which is shown in
-Fig.~\ref{fig-make}(a).
+executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
+the sake of simplicity. In reality, there are multiple versions of \cmd{gcc}
+and the actual binary used to compile and link files, and standard libraries
+such that \cmd{stdio.h}, are often also listed as dependencies.} the command
+\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
+\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
+executable \cmd{main.exe}. The makefile contains the complete information about
+the \emph{task dependency graph}, which is shown in Fig.~\ref{fig-make}(a).
 
@@
 To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
-modification time} to detect which files changed\footnote{Technically, you
+modification times} to detect which files changed\footnote{Technically, you
 can fool \Make by altering the modification time of a file without changing its
@@
 \noindent
-The formula in \cmd{B1} uses the \cmd{INDIRECT} function, which takes a string and
-returns the value of the cell with that name.  The string evaluates
-to \cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
+The formula in \cmd{B1} uses the \cmd{INDIRECT} function, which takes a string
+and returns the value of the cell with that name.  The string evaluates to
+\cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
 formula in \cmd{B1} are determined by the value of \cmd{C1}, so there is no
-possibility\footnote{In this particular example one might say that
-the value of \cmd{C1} is statically known, but imagine that it is the result of a long computation
-chain -- its value will only become available during the build.}
-of computing the dependency graph prior to calc.
+possibility\footnote{In this particular example one might say that the value of
+\cmd{C1} is statically known, but imagine that it is the result of a long
+computation chain -- its value will only become available during the build.}
+of computing the dependency graph before the build starts.
 
-To support dynamic dependencies,
-\Excel's build algorithm~\cite{excel_recalc} is significantly different from
-\Make. \Excel uses the \emph{calculation chain}
-produced by the previous build as an approximation to the correct topological
-order. During recalculation, \Excel processes cells in this order, but can
-\emph{defer recalculation of a cell} by moving it down the chain if a newly
-discovered dependency has not yet been rebuilt. We refer to this
-algorithm as \emph{restarting}, and will discuss it in more detail
-in~\S\ref{sec-implementation-excel}.
+To support dynamic dependencies, \Excel's build algorithm~\cite{excel_recalc} is
+significantly different from \Make. \Excel arranges the cells into a linear
+sequence, called the \emph{calc chain}.  During the build, \Excel processes
+cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
+value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
+computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
+resumes the build starting with \cmd{D}. When a build is complete, the resulting
+calc chain respects all the dynamic dependencies of the spreadsheet. When an
+input value (or formula) is changed, \Excel uses the final calc chain from the
+\emph{previous} build as its starting point so that, in the common case where
+changing an input value does not change dependencies, there are no aborts.
+Notice that build always succeeds regardless of the initial calc chain (barring
+truly circular dependencies); the calc chain is just an optimisation.
+We refer to this algorithm as \emph{restarting}, and will discuss it
+in more detail in~\S\ref{sec-implementation-excel}.
 
@@
 \cmd{A2} changes; but these facts are not statically apparent. In practice
-\Excel implements a conservative approximation to minimality: roughly, it
-recomputes a formula if (i) the formula text mentions a changed cell, or (ii)
-the formula uses a function like \cmd{INDIRECT} whose dependencies are not
-statically visible.
+\Excel implements a conservative approximation to minimality: it recomputes a
+formula if~(i)~the formula mentions a changed cell, or~(ii)~the formula uses a
+function like \cmd{INDIRECT} whose dependencies are not statically visible,
+or~(iii)~the formula itself has changed.
 
-Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
-systems only track changes of inputs and intermediate results, but \Excel can
-also track changes in the tasks themselves: if a formula is modified, \Excel
-will recompute it and propagate the changes. Self-tracking is uncommon in
-software build systems, where one often needs to manually initiate a full
-rebuild even if just a single build task has changed. We discuss
-self-tracking further in~\S\ref{sec-tracking-aspects}.
+Item (iii) in the above list highlights a distinguishing feature of \Excel: it
+is \emph{self-tracking}. Most build systems only track changes of inputs and
+intermediate results, but \Excel also tracks changes in the tasks themselves: if
+a formula is modified, \Excel will recompute it and propagate the changes.
+Self-tracking is uncommon in software build systems, where one often needs to
+manually initiate a full rebuild even if just a single build task has changed.
+We discuss self-tracking further in~\S\ref{sec-tracking-aspects}.
 
@@
     \item \cmd{release.txt} is a text file listing all files that should be in the release. This file
-    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt}
-    that list all binary and documentation files of the project.
+    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt},
+    which list all binary and documentation files of the project.
     \item \cmd{release.tar} is the release archive built by executing the
@@
 \emph{adaptive}~\cite{acar2002adaptive}, and
-\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and \S\ref{sec-related}).
-Second, instead of abandoning
-and deferring the execution of tasks whose newly discovered dependencies have
-not yet been built (as \Excel does), \Shake \emph{pauses} their execution until
-the dependencies are brought up to date. We refer to this build algorithm as
-\emph{recursive}.
+\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and
+\S\ref{sec-related}). Second, instead of aborting and deferring the execution of
+tasks whose newly discovered dependencies have not yet been built (as \Excel
+does), \Shake \emph{suspends} their execution until the dependencies are brought
+up to date. We refer to this build algorithm as \emph{suspending}.
 

### 2018-06-28 paper/4-build.tex
Redo S4

--- 2018-06-19 paper/4-build.tex
@@
 \hs{Build}~\hs{c}~\hs{i}~\hs{k}~\hs{v}, given
-a \emph{client-supplied} implementation of \hs{Task}~\hs{c}~\hs{k}~\hs{v}. That
-is, we are going to take \hs{Task} as given from now on, and explore variants of
+a \emph{client-supplied} implementation of \hs{Tasks}~\hs{c}~\hs{k}~\hs{v}. That
+is, we are going to take \hs{Tasks} as given from now on, and explore variants of
 \hs{Build}: first abstractly (in this section) and then concretely
@@
 
-The second aspect, determining what to rebuild, can be addressed in one of three
+The second aspect, determining what to rebuild, can be addressed in one of four
 fundamental ways, with a number of tweaks and variations within them.
@@
     { key     :: k
-    , dependencies :: [(k, Hash v)]
+    , depends :: [(k, Hash v)]
     , result  :: Hash v }
@@
 Once we are storing the complete result it makes sense to record many constructive traces per key, and to share them with other users, providing cloud-build functionality.
-We can represent that as:
+We can represent that by changing \hs{Trace} to:
 
 \begin{minted}{haskell}
-data Traces k v = Traces
-    { traces    :: [Trace k v]
-    , contents  :: Map (Hash v) v }
+data Trace k v = Trace
+    { key     :: k
+    , depends :: [(k, Hash v)]
+    , result  :: v }
 \end{minted}
 
-We have a list of traces, plus a \hs{Map} from the hash to the actual contents.
-Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{contents} \emph{without} recomputing it.
-We split the traces and contents because in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only retrieve at most one complete file per key.
+The change from~\S\ref{sec-verifying-traces} is that instead of storing the \hs{Hash}~\hs{v} for \hs{result}, we now store the full value \hs{v}.
+Checking a trace follows the same pattern as before, but if the result is the only thing that is different we can simply use the result \emph{without} recomputing it.
 
@@
 
-\textbf{TODO}
+If we assume all actions are \emph{deterministic} (specifically that given the same inputs they produce the same output) then we can make a further optimisation to arrive at deterministic constructive traces:
+
+\begin{minted}{haskell}
+data Trace k v = Trace
+    { key     :: k
+    , depends :: [(k, Hash (Tree (Hash v)))]
+    , result  :: v
+    }
+\end{minted}
+
+Instead of storing the hash of the immediate dependencies, a determinstic constructive trace only ever takes the hash of initial inputs.
+To form the hash of a dependency, it forms the tree of dependencies all the way to the inputs, and hashes that.
+As an example, if \hs{a} depends on the input \hs{b}, then the dependency value used for \hs{a} uses \hs{hash}~\hs{b}, \emph{but not} \hs{hash}~\hs{a}.
+As a consequence, given the hashes of all inputs, and dependency information (e.g.
+from an \hs{Applicative} constraint) it is possible to compute the dependency hash of \emph{any} node, without having built any intermediate nodes.
+This property is particularly useful for ``shallow builds'', as discussed in~\S\ref{sec-cloud-aspects}.
 
@@
 \Bazel, can index directly from the key and results to the output result~--~i.e.
-\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{(Hash}~\hs{v)}.
-Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash.
-In this formulation we have removed the possibility for a single key/dependency state to map to multiple different hashes, e.g.
-on a non-deterministic build~--~something \Bazel already prohibits which is discussed more in~\S\ref{sec-non-determinism}.
+\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}.
+Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result.
+\item Many cloud build systems store hashes of values in the trace information, then have a separate ``blob server'' which associates hashes with their actual contents.
 \item Finally, a \hs{Monad} build system with constructive traces can be stored as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
@@
 Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
-Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & \Cloud \Shake  \\\hline
+Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
 Deterministic constructive traces\hspace{2mm}\hfill\S\ref{sec-deterministic-constructive-traces} & \Buck       & -      & \Nix           \\\hline
@@
 With the information in this section we can build a table comparing the
-dependency order strategy with the out-of-date keys strategy, providing 9~possible
-build systems, 5~of which are actually inhabited by existing build systems
-(we discuss \Ninja \cite{ninja} in \S\ref{sec-related-build}). Of the
-remaining 4~spots, we believe neither Traced or Cloud \Excel make sense~--~the
-\Excel approach of reordering combined with static approximations reduces
-the memory usage significantly. However, as soon as you are paying the cost of
-storing traces, that benefit is gone. The advantage of Approximate \Shake over
-\Shake would be that it could avoid storing traces and having a separate
-information database, but that advantage is minor compared to the technical
-restrictions and approximations it would provide, so we consider it unlikely
-to be built. The Cloud \Shake system is an interesting and important point in
-the design space, which we explore further
-in~\S\ref{sec-implementation-cloud}.
+dependency order strategy with the out-of-date keys strategy, providing 12~possible
+build systems, 8~of which are actually inhabited by existing build systems
+(we discuss all these systems in \S\ref{sec-related-build}). Of the
+remaining 4~spots, all result in workable build systems. The most interesting
+unfilled spot in the table is suspending constructive traces, which would provide
+many benefits, and which we title \Cloud \Shake (as we plan on extending \Shake to occupy that spot)
+and further explore further in~\S\ref{sec-implementation-cloud}.

### 2018-06-28 paper/2-background.tex
Work on S2

@@
 \caption{A task dependency graph and two build scenarios. Input files are shown
-as rectangles, intermediate and output files are shown as rounded rectangles.
+in rectangles, intermediate and output files are shown in rounded rectangles.
 Modified inputs and files that are rebuilt are highlighted.
@@
 
-To support dynamic dependencies, \Excel's build algorithm~\cite{excel_recalc} is
+To support dynamic dependencies, \Excel's build engine~\cite{excel_recalc} is
 significantly different from \Make. \Excel arranges the cells into a linear
@@
 does), \Shake \emph{suspends} their execution until the dependencies are brought
-up to date. We refer to this build algorithm as \emph{suspending}.
+up to date. We refer to this task scheduling algorithm as \emph{suspending}.
 
@@
 
-The user starts by checking out the project sources, whose hashes are (for
-simplicity)~1,~2~and~3, and requests to build \cmd{main.exe}, see
+The user starts by downloading the project sources, whose content hashes are
+(for simplicity)~1,~2 and~3, and requests to build \cmd{main.exe}, see
 Fig.~\ref{fig-bazel}(a,b). By looking up the global history of all previous
@@
 the build system finds that someone has already compiled these exact sources
-before and their resulting files \cmd{util.o} and \cmd{main.o} had hashes 4~and~5.
-Similarly, the build system finds that the hash of the resulting \cmd{main.exe}
-should be 6 and downloads the actual binary from the shared cloud storage, since
-it is the end build product and must therefore be materialised.
+before and the resulting files \cmd{util.o} and \cmd{main.o} had hashes
+4~and~5, respectively. Similarly, the build system finds that the hash of the
+resulting \cmd{main.exe} was 6 and downloads the actual binary from the shared
+cloud storage -- it must be materialised, because it is the end build product.
 
 In the second scenario, shown in Fig.~\ref{fig-bazel}(c), the user modifies the
-source \cmd{util.c}, thereby changing its hash from~1~to~7. The build system
-finds that nobody has ever compiled the new $\{\cmd{util.c}, \cmd{util.h}\}$
-combination and must therefore build \cmd{util.o}, which results in changing its
-hash from~4~to~8. The combination of hashes of \cmd{util.o} and
-\cmd{main.o} has not been encountered before either, therefore the build system
-first downloads \cmd{main.o} from the cloud and then builds \cmd{main.exe} by
-linking the two object files. When the build is complete, the results
-can be uploaded to the cloud for future reuse by other team members.
+source file \cmd{util.c}, thereby changing its hash from~1~to~7. The cloud
+lookup of the new combination $\{\cmd{util.c}, \cmd{util.h}\}$ fails,
+which means that nobody has ever compiled it. The build system must therefore
+build \cmd{util.o}, which results in materialising it with a new hash~8. The
+combination of hashes of \cmd{util.o} and \cmd{main.o} has not been encountered
+before either, therefore the build system first downloads \cmd{main.o} from the
+cloud and then builds \cmd{main.exe} by linking the two object files. When the
+build is complete, the results can be uploaded to the cloud for future reuse by
+other team members.
 
 \Bazel is one of the first examples of openly available cloud build systems.
-Like \Make, it does not support dynamic dependencies and can therefore benefit
-from the simplicity of building tasks in a statically known topological order.
-It is minimal and supports the early cutoff optimisation. To support cloud
-builds, \Bazel maintains a \emph{content-addressable cache} that can be used to
-fetch a previously built file given the hash of its content, and
-\emph{dependency graphs from all previous builds}, annotated with observed file
-hashes. The latter allows builds to bypass the execution of a task, by predicting the
-hash of the result from the hashes of its dependencies, and subsequently
-fetch the result from the cache. A concrete implementation is provided
-in~\S\ref{sec-implementations}.
+As of writing, it is not possible to express dynamic dependencies in
+user-defined build rules, however some of the predefined build rules require
+dynamic dependencies and the internal build engine can cope with them by using
+a \emph{restarting} task scheduler, which is similar to that of \Excel but does
+not use the calc chain. \Bazel is minimal and supports the early cutoff
+optimisation.
+
+To support cloud builds, \Bazel maintains (i) a \emph{content-addressable cache}
+that can be used to download a previously built file given the hash of its
+content, and (ii) the history of all executed build commands annotated with
+observed file hashes. The latter allows the build engine to bypass the execution
+of a task, by predicting the hash of the result from the hashes of its
+dependencies, and subsequently download the result from the cache. A concrete
+implementation is provided in~\S\ref{sec-implementations}.
 
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
-\caption{Checkout source files}
+\caption{Download source files}
 \end{subfigure}
@@
 \end{subfigure}
-\caption{A cloud build example: (a)~checkout sources, (b)~download \cmd{main.exe}
-from the cloud and skip intermediate files (only their hashes are needed),
-(c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which requires building
-\cmd{util.o} (since nobody has compiled \cmd{util.c} before) and downloading
-\cmd{main.o}. File hashes are shown inside circles.
+\caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
+by downloading it from the cloud and skipping intermediate files (only their
+hashes are needed), (c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which
+requires building \cmd{util.o} (nobody has compiled \cmd{util.c} before) and
+downloading \cmd{main.o} (it is needed for linking \cmd{main.exe}). File hashes
+are shown in circles, and non-materialised intermediates in dashed rounded
+rectangles.
 \label{fig-bazel}}
@@
     system to do that.
-    \item \Excel stores one dirty bit per cell and the calculation chain from
-    the previous build.
+    \item \Excel stores one dirty bit per cell and the calc chain from the
+    previous build.
     \item \Shake stores the dependency graph discovered in the previous build,
     annotated with file content hashes for efficient checking of file changes.
-    \item \Bazel stores all dependency graphs discovered in previous builds
-    annotated with file hashes, and the content-addressable cache.
+    \item \Bazel stores the content-addressable cache and the history of all
+    previous build commands annotated with file hashes. This information is
+    shared among all users of the build system.
 \end{itemize}
@@
 $\!$\Make       $\!$& File modification times      & Topological & Static         & Yes     & No     & No   $\!$\\
-$\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
-$\!$\Shake      $\!$& Previous dependency graph      & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
-$\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
+$\!$\Excel      $\!$& Dirty cells, calc chain      & Restarting  & Dynamic        & No      & No     & No   $\!$\\
+$\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic        & Yes     & Yes    & No   $\!$\\
+$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Static$^{(*)}$ & Yes     & Yes    & Yes  $\!$\\\hline
 \hline
@@
 \vspace{0.5mm}
-\caption{Summary of build system differences.\label{tab-summary}}
+\caption{Summary of build system differences; $^{(*)}$predefined build rules can
+have dynamic dependencies.\label{tab-summary}}
 \vspace{-6mm}
@@
 In this paper we elucidate which build system properties are consequences of
-specific implementation choices (metadata and algorithm), and how one can obtain
-new build systems with desired properties by recombining parts of existing
-implementations. As a compelling example, we demonstrate how to combine the
-advantages of \Shake and \Bazel in a cloud build system with dynamic
-dependencies, see~\S\ref{sec-implementation-cloud}.
+specific implementation choices (stored metadata and task scheduling algorithm),
+and how one can obtain new build systems with desired properties by recombining
+parts of existing implementations. As a compelling example, we demonstrate how
+to combine the advantages of \Shake and \Bazel in a cloud build system with
+dynamic dependencies, see~\S\ref{sec-implementation-cloud}.

### 2018-06-28 paper/2-background.tex
Restarting schedulers are not minimal

@@
 a \emph{restarting} task scheduler, which is similar to that of \Excel but does
-not use the calc chain. \Bazel is minimal and supports the early cutoff
-optimisation.
+not use the calc chain. \Bazel is not minimal in the sense that it may restart a
+task multiple times as new dependencies are discovered and rebuilt, but it
+supports the early cutoff optimisation.
 
@@
 $\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic        & Yes     & Yes    & No   $\!$\\
-$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Static$^{(*)}$ & Yes     & Yes    & Yes  $\!$\\\hline
+$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Static$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
 \hline

### 2018-06-28 paper/2-background.tex
Label Bazel as dynamic

@@
 $\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   $\!$\\
-$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Static$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
+$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
 \hline
@@
 \vspace{0.5mm}
-\caption{Summary of build system differences; $^{(*)}$predefined build rules can
-have dynamic dependencies.\label{tab-summary}}
+\caption{Summary of build system differences; $^{(*)}$used-defined build rules
+cannot have dynamic dependencies.\label{tab-summary}}
 \vspace{-6mm}
@@
 and how one can obtain new build systems with desired properties by recombining
-parts of existing implementations. As a compelling example, we demonstrate how
-to combine the advantages of \Shake and \Bazel in a cloud build system with
-dynamic dependencies, see~\S\ref{sec-implementation-cloud}.
+parts of existing implementations. As a compelling example,
+in~\S\ref{sec-implementation-cloud} we demonstrate how to combine the advantages
+of \Shake and \Bazel.

### 2018-06-29 paper/4-build.tex
Work on S4.1

@@
 The build systems overview (\S\ref{sec-background-summary}) highlighted three
-distinct approaches to respecting the dependency order. This subsection explores
-their properties and possible implementations.
+distinct task schedulers respecting the dependency order. This subsection
+explores their properties and possible implementations.
 
@@
 
-The topological approach pre-computes a linear order, which when followed, ensures
-the \hs{build} is correct regardless of the initial \hs{store}. Given a function
-from a key to its dependencies, and the output \hs{key}, you can compute the
-linear order by first finding the reachable dependencies of \hs{key},
-and then computing a topological sort. However, as we have seen
-in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task,
-which requires the build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling
-out dynamic dependencies.
+The topological scheduler pre-computes a linear order, which when followed,
+ensures the build result is correct regardless of the initial store. Given a
+task description and the output \hs{key}, you can compute the linear order by
+first finding the (acyclic) graph of the \hs{key}'s reachable dependencies, and
+then computing a topological sort. However, as we have seen in~\S\ref{sec-deps},
+we can only extract dependencies from an applicative task, which requires the
+build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic
+dependencies.
 
 \vspace{-2mm}
-\subsubsection{Reordering}\label{sec-restarting}
+\subsubsection{Restarting}\label{sec-restarting}
 
-The topological approach has two downsides: it is limited to \hs{Applicative}
+The topological scheduler has two downsides: it is limited to \hs{Applicative}
 build systems and requires a fresh topological sort each time. So, while the
-actions themselves may be incremental (i.e. unnecessary tasks will not be performed),
-the pre-processing is not. We can try to incrementalise the topological sort by
-storing the topological order between build runs and assume it to
-be correct, but if the build discovers it is wrong, fix it up.
+actions themselves may be incremental (i.e. unnecessary tasks will not be
+performed), the pre-processing is not. We can try to incrementalise the
+topological sort by storing the topological order between build runs and assume
+it to be correct, fixing it up if the build discovers any inconsistency, as
+\Excel does.
 
@@
 dependencies. It is also not minimal in the sense that a task may start, do some
-meaningful work, then abort. However, in the case of an \hs{Applicative} system,
-that work is zero.
+meaningful work, and then abort. However, in the case of an \hs{Applicative}
+system, that work is zero.
+
+\Bazel's restarting scheduler does not store the topological order between build
+runs; instead, it stores the most recent task dependency information. Since this
+information may become outdated, \Bazel may also need to abort a task if a
+newly discovered dependency is out-of-date.
 
 \vspace{-2mm}
-\subsubsection{Recursive}\label{sec-suspending}
+\subsubsection{Suspending}\label{sec-suspending}
 
 An alternative approach, utilised by the \hs{busy} build system
-(\S\ref{sec-general-build}), is to simply build dependencies when they are
-requested. By combining that with a transient set of which keys have already
-been built, you can obtain a minimal build system.
+(\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
+they are requested, suspending the currently running task. By combining that
+with tracking the keys that have already been built, one can obtain a minimal
+build system with dynamic dependencies.
 
-This approach requires that a task may be started, then during that execution
-another task will have to be run. Assuming an IO-driven task structure,
-that requires suspending a running task, which can be done with cheap
-green threads and blocking (the original approach of \Shake) or using
-continuation-passing style (what \Shake does currently). An alternative approach to
-suspending a task is to abort it and restart it again later, at the cost
-of doing additional work.
+This approach requires that a task may be started and then suspended until
+another task is complete. This can be done with cheap green threads and blocking
+(the original approach of \Shake) or using continuation-passing style (what
+\Shake does currently).
+
+% An alternative approach to suspending a task is to abort
+% it and restart it again later, at the cost of doing additional work.
 

### 2018-06-29 paper/2-background.tex
Fix typo
Address Simon's comments

@@
 \hline
-$\!$Build system$\!$& Persistent build information & Algorithm   & Dependencies    & Minimal & Cutoff & Cloud$\!$\\\hline
+$\!$Build system$\!$& Persistent build information & Scheduler   & Dependencies    & Minimal & Cutoff & Cloud$\!$\\\hline
 $\!$\Make       $\!$& File modification times      & Topological & Static          & Yes     & No     & No   $\!$\\
@@
 \vspace{0.5mm}
-\caption{Summary of build system differences; $^{(*)}$used-defined build rules
+\caption{Summary of build system differences; $^{(*)}$user-defined build rules
 cannot have dynamic dependencies.\label{tab-summary}}

### 2018-06-29 paper/4-build.tex
Address Simon's comments

@@
 \vspace{1mm}
-We have bolded two different phrases above, and tackle each aspect separately.
+We have bolded two different phrases above and, as we will see, the part of the
+build system responsible for scheduling tasks in the dependency order
+(a `scheduler') can be cleanly separated from the part responsible for deciding
+whether a key needs to be rebuilt (a `rebuilder'). We therefore tackle each
+aspect separately in subsections~\S\ref{sec-dependency-orderings}
+and~\S\ref{sec-out-of-date}.
 

### 2018-06-29 paper/1-intro.tex
Work on abstract and intro

@@
 
-\item We show that we can instantiate
-our abstractions to describe the essence of a variety of different
-real-life build systems, including \Make, \Shake, \Bazel, and \Excel, each in
-a dozen lines of code or so (\S\ref{sec-implementations}).
-Doing this modelling in a single setting allows
-the differences and similarities between these huge systems to be
+\item We show that we can instantiate our abstractions to describe the essence
+of a variety of different real-life build systems, including \Make, \Shake,
+\Bazel, \CloudBuild, \Buck, \Nix, and \Excel, each in a dozen lines of code
+or so (\S\ref{sec-implementations}). Doing this modelling in a single setting
+allows the differences and similarities between these huge systems to be
 brought out clearly.
 
-\item Moreover, we can readily remix the ingredients to describe the first
-  build system that supports both \emph{dynamic dependencies}
-  and \emph{cloud build} (\S\ref{sec-implementation-cloud}).
+\item Moreover, we can readily remix the ingredients to design new build systems
+with desired properties, for example, to combine the advantages of \Shake and
+\Bazel (\S\ref{sec-implementation-cloud}).
 

### 2018-06-30 paper/1-intro.tex
Tweak S1 and S2

@@
 in a way that is both abstract (omitting incidental detail)
-and yet precise (implemented as Haskell code).  Specifically we make
+and yet precise (implemented as Haskell code:
+\url{https://hackage.haskell.org/package/build-0.0.1.1}). Specifically we make
 these contributions:
 \begin{itemize}
-\item Build systems vary on many axes, including:
-static vs dynamic dependencies; local vs cloud; deterministic vs
-non-deterministic build rules; support for early cutoff; self-tracking build
-systems; and the type of persistent build information.
-In \S\ref{sec-background} we identify some key properties, in the context of
-four carefully-chosen build systems.
+\item Build systems vary on many axes, including: static vs dynamic
+  dependencies; local vs cloud; deterministic vs non-deterministic build rules;
+  support for early cutoff; self-tracking build systems; and the type of
+  persistent build information. In~\S\ref{sec-background} we identify some key
+  properties, in the context of four carefully-chosen build systems.
+
 \item We describe some simple but novel abstractions that
@@
 
-\item We identify two key design choices
-  that are typically deeply wired into any build system:
-  \emph{the order in which dependencies are built} (\S\ref{sec-dependency-orderings})
-  and \emph{whether or not a dependency is (re-)built} (\S\ref{sec-out-of-date}).
-  These choices turn out to be orthogonal, which leads us to a new
-  classification of the design space (\S\ref{sec-design-space}).
+\item We identify two key design choices that are typically deeply wired into
+  any build system: \emph{the order in which dependencies are
+  built}~(\S\ref{sec-dependency-orderings}) and \emph{whether or not a
+  dependency is (re-)built}~(\S\ref{sec-out-of-date}). These choices turn out to
+  be orthogonal, which leads us to a new classification of the design
+  space~(\S\ref{sec-design-space}).
 

### 2018-06-30 paper/2-background.tex
Tweak S1 and S2

@@
 If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
-will first build \cmd{util.o} and \cmd{main.o}, in any order since these
-tasks are independent, and then build \cmd{main.exe}. If the
-user modifies the sources of \cmd{util.h} and runs \Make again, it will
-perform a \emph{full rebuild}, because all three tasks transitively depend on
-\cmd{util.h}, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if
-the user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient:
-the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
-changed, see Fig.~\ref{fig-make}(c). Note that if the dependency graph is
-\emph{acyclic} then each task needs to be executed at most once. Cyclic task
-dependencies are typically not allowed in build systems although there are rare
+will build~\cmd{util.o}~and~\cmd{main.o}, in any order since these tasks are
+independent, and then \cmd{main.exe}. If the user modifies \cmd{util.h} and runs
+\Make again, it will perform a \emph{full rebuild}, because all three tasks
+transitively depend on \cmd{util.h}, as illustrated in Fig.~\ref{fig-make}(b).
+On the other hand, if the user modifies \cmd{main.c} then a \emph{partial
+rebuild} is sufficient: \cmd{util.o} does not need to be rebuilt, since its
+inputs have not changed, see Fig.~\ref{fig-make}(c). Note that if the dependency
+graph is \emph{acyclic} then each task needs to be executed at most once. Cyclic
+task dependencies are typically not allowed in build systems but there are rare
 exceptions, see~\S\ref{sec-iterative-compute}.
 
-The following property is essential for build systems; indeed, it is their raison d'\^etre:
+\noindent
+The following property is essential for build systems; indeed, it is their
+raison d'\^etre:
 
@@
 
-\begin{figure}[h]
-\vspace{-2mm}
+\begin{figure}
 \begin{subfigure}[b]{0.90\linewidth}
@@
 does), \Shake \emph{suspends} their execution until the dependencies are brought
-up to date. We refer to this task scheduling algorithm as \emph{suspending}.
+up to date. We refer to this task scheduling algorithm as \emph{suspending},
+see a concrete implementation in~\S\ref{sec-implementation-shake}.
 
-\begin{figure}[h]
-\vspace{-5mm}
+\begin{figure}
+\vspace{-2mm}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example-cutoff.pdf}}
@@
 rebuilt.\label{fig-cutoff}}
-\vspace{-2mm}
+\vspace{-4mm}
 \end{figure}
@@
 
-\begin{figure}[t]
+\begin{figure}
 \begin{subfigure}[b]{0.25\linewidth}
@@
 \end{subfigure}
+\vspace{-2mm}
 \caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
@@
 are shown in circles, and non-materialised intermediates in dashed rounded
-rectangles.
-\label{fig-bazel}}
+rectangles.\label{fig-bazel}}
+\vspace{-4mm}
 \end{figure}
@@
 
-\begin{table}[h]
+In this paper we elucidate which build system properties are consequences of
+specific implementation choices (stored metadata and task scheduling algorithm),
+and how one can obtain new build systems with desired properties by recombining
+parts of existing implementations. As a compelling example,
+in~\S\ref{sec-implementation-cloud} we demonstrate how to combine the advantages
+of \Shake and \Bazel.
+
+\begin{table}
 \smaller
@@
 \vspace{0.5mm}
-\caption{Summary of build system differences; $^{(*)}$user-defined build rules
+\caption{Build system differences; $^{(*)}$at present, user-defined build rules
 cannot have dynamic dependencies.\label{tab-summary}}
@@
 \end{table}
-
-In this paper we elucidate which build system properties are consequences of
-specific implementation choices (stored metadata and task scheduling algorithm),
-and how one can obtain new build systems with desired properties by recombining
-parts of existing implementations. As a compelling example,
-in~\S\ref{sec-implementation-cloud} we demonstrate how to combine the advantages
-of \Shake and \Bazel.

### 2018-06-30 paper/2-background.tex
Work on S2-S3, add a figure with key abstractions

@@
 \begin{figure}
-\vspace{-2mm}
+\vspace{-3mm}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example-cutoff.pdf}}
-\vspace{-3mm}
+\vspace{-4mm}
 \caption{An early cutoff example: if a comment is added to \cmd{main.c}, the
@@
 
-\subsection{\Bazel: a cloud build system}
-\label{sec-background-bazel}
+\vspace{-0.5mm}
+\subsection{\Bazel: a cloud build system}\label{sec-background-bazel}
+\vspace{-0.5mm}
 
@@
 system} can speed up builds dramatically by sharing build results
-among team members. Furthermore, cloud build systems allow one to perform
-\emph{shallow builds} that materialise only end build products locally, leaving
-all intermediates in the cloud. We illustrate shallow cloud builds by an example
-in Fig.~\ref{fig-bazel}.
+among team members. Furthermore, cloud build systems can support \emph{shallow
+builds} that materialise only end build products locally, leaving all
+intermediates in the cloud.
 
-The user starts by downloading the project sources, whose content hashes are
-(for simplicity)~1,~2 and~3, and requests to build \cmd{main.exe}, see
-Fig.~\ref{fig-bazel}(a,b). By looking up the global history of all previous
-builds of the project\footnote{Here we ignore the issue of limited cloud storage
-resources for the sake of simplicity; in practice, old entries are regularly
-evicted from the storage, as further discussed in~\S\ref{sec-cloud-aspects}.},
-the build system finds that someone has already compiled these exact sources
-before and the resulting files \cmd{util.o} and \cmd{main.o} had hashes
-4~and~5, respectively. Similarly, the build system finds that the hash of the
-resulting \cmd{main.exe} was 6 and downloads the actual binary from the shared
-cloud storage -- it must be materialised, because it is the end build product.
+Consider an example in Fig.~\ref{fig-bazel}. The user starts by downloading the
+sources, whose content hashes are (for simplicity)~1,~2 and~3, and requests to
+build \cmd{main.exe}, see Fig.~\ref{fig-bazel}(a,b). By looking up the global
+history of all previous builds\footnote{In practice, old entries are regularly
+evicted from the cloud storage, as further discussed
+in~\S\ref{sec-cloud-aspects}.}, the build system finds that someone has already
+compiled these exact sources before and the resulting files \cmd{util.o} and
+\cmd{main.o} had hashes 4~and~5, respectively. Similarly, the build system finds
+that the hash of the resulting \cmd{main.exe} was 6 and downloads the actual
+binary from the cloud storage -- it must be materialised, because it is the end
+build product.
 
-In the second scenario, shown in Fig.~\ref{fig-bazel}(c), the user modifies the
-source file \cmd{util.c}, thereby changing its hash from~1~to~7. The cloud
-lookup of the new combination $\{\cmd{util.c}, \cmd{util.h}\}$ fails,
-which means that nobody has ever compiled it. The build system must therefore
-build \cmd{util.o}, which results in materialising it with a new hash~8. The
-combination of hashes of \cmd{util.o} and \cmd{main.o} has not been encountered
-before either, therefore the build system first downloads \cmd{main.o} from the
-cloud and then builds \cmd{main.exe} by linking the two object files. When the
-build is complete, the results can be uploaded to the cloud for future reuse by
-other team members.
+In Fig.~\ref{fig-bazel}(c), the user modifies the source file \cmd{util.c},
+thereby changing its hash from~1~to~7. The cloud lookup of the new combination
+$\{\cmd{util.c}, \cmd{util.h}\}$ fails, which means that nobody has ever
+compiled it. The build system must therefore build \cmd{util.o}, materialising
+it with the new hash~8. The combination of hashes of \cmd{util.o} and
+\cmd{main.o} has not been encountered before either, thus the build system first
+downloads \cmd{main.o} from the cloud and then builds \cmd{main.exe} by linking
+the two object files. When the build is complete, the results can be uploaded to
+the cloud for future reuse by other team members.
 
@@
 dependencies, and subsequently download the result from the cache. A concrete
-implementation is provided in~\S\ref{sec-implementations}.
+implementation is provided in~\S\ref{sec-implementation-cloud}.
 
 \begin{figure}
+\vspace{-2mm}
 \begin{subfigure}[b]{0.25\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
+\vspace{-1mm}
 \caption{Download source files}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
+\vspace{-1mm}
 \caption{Modify \cmd{util.c} and rebuild}
 \end{subfigure}
-\vspace{-2mm}
+\vspace{-2.5mm}
 \caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
@@
 rectangles.\label{fig-bazel}}
-\vspace{-4mm}
+\vspace{-3mm}
 \end{figure}
 
-\subsection{Summary}
-\label{sec-background-summary}
+\begin{table}[h]
+\vspace{-1mm}
+\smaller
+\centering
+\begin{tabular}{l||l|l||l|c|c|c}
+\hline
+$\!$Build system$\!$& Persistent build information & Scheduler   & Dependencies    & Minimal & Cutoff & Cloud$\!$\\\hline
+$\!$\Make       $\!$& File modification times      & Topological & Static          & Yes     & No     & No   $\!$\\
+$\!$\Excel      $\!$& Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   $\!$\\
+$\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   $\!$\\
+$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
+\hline
+\end{tabular}
+\caption{Build system differences; $^{(*)}$at present, user-defined build rules
+cannot have dynamic dependencies.\label{tab-summary}}
+\vspace{-8mm}
+\end{table}
+
+\vspace{-1.5mm}
+\subsection{Summary}\label{sec-background-summary}
+\vspace{-0.5mm}
 
@@
 of \Shake and \Bazel.
-
-\begin{table}
-\smaller
-\centering
-\begin{tabular}{l||l|l||l|c|c|c}
-\hline
-$\!$Build system$\!$& Persistent build information & Scheduler   & Dependencies    & Minimal & Cutoff & Cloud$\!$\\\hline
-$\!$\Make       $\!$& File modification times      & Topological & Static          & Yes     & No     & No   $\!$\\
-$\!$\Excel      $\!$& Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   $\!$\\
-$\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   $\!$\\
-$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
-\hline
-\end{tabular}
-\vspace{0.5mm}
-\caption{Build system differences; $^{(*)}$at present, user-defined build rules
-cannot have dynamic dependencies.\label{tab-summary}}
-\vspace{-6mm}
-\end{table}

### 2018-06-30 paper/3-abstractions.tex
Work on S2-S3, add a figure with key abstractions

--- 2018-06-11 paper/3-abstractions.tex
@@
 
-This section presents purely functional abstractions that allow us to express
-all the intricacies of build systems discussed previously in \S\ref{sec-background},
-and design complex build systems from simple
-primitives. Specifically, we present the \emph{task} and \emph{build}
-abstractions in~\S\ref{sec-task} and~\S\ref{sec-general-build}, respectively.
-Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the
-abstractions further and provide concrete implementations for several build
-systems.
-
-\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
 \begin{figure}
+\vspace{-1mm}
 \begin{minted}[fontsize=\small]{haskell}
--- An abstract store
+-- An abstract store containing a key/value map and persistent build information
 data Store i k v -- i = info, k = key, v = value
+initialise :: i -> (k -> v) -> Store i k v
 getInfo    :: Store i k v -> i
@@
 getHash    :: Hashable v => k -> Store i k v -> Hash v
-initialise :: i -> (k -> v) -> Store i k v
+hash       :: Hashable v => v -> Hash v -- provides a fast equality check
 \end{minted}
-\vspace{1mm}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
--- Hashing
-hash :: Hashable a => a -> Hash a
+-- Build tasks (see §3.2)
+newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
+type    Tasks c k v = k -> Maybe (Task c k v)
 \end{minted}
-\vspace{1mm}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
--- Applicative functors
-pure  :: Applicative f => a -> f a
-(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
-(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
+-- Build system (see §3.3)
+type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
 \end{minted}
-\vspace{1mm}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
--- Standard State monad from Control.Monad.State
-data State s a
-instance Applicative (State s)
-instance Monad       (State s)
-get       :: State s s
-gets      :: (s -> a) -> State s a
-modify    :: (s -> s) -> State s ()
-execState :: State s a -> s -> s
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Standard types from Data.Functor.Identity and Data.Functor.Const
-newtype Identity a = Identity { runIdentity :: a }
-newtype Const m a  = Const    { getConst    :: m }
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[fontsize=\small]{haskell}
-instance Functor (Const m) where
-    fmap _ (Const m) = Const m
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[fontsize=\small]{haskell}
-instance Monoid m => Applicative (Const m) where
-    pure _              = Const mempty
-    Const x <*> Const y = Const (x <> y)
+-- Build system components: a scheduler and a rebuilder (see §5)
+type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
+type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
 \end{minted}
 \vspace{-3mm}
-\caption{Signatures of main data types and library functions.}\label{fig-types}
-\vspace{-4mm}
+\caption{Type signatures of key build systems abstractions.}\label{fig-types}
+\vspace{-3mm}
 \end{figure}
-We begin by establishing a common vocabulary for build systems:
 
-\vspace{-2mm}
-\paragraph{Keys, values, and the store} The goal of any build system is to
+This section presents purely functional abstractions that allow us to express
+all the intricacies of build systems discussed in~\S\ref{sec-background},
+and design complex build systems from simple primitives. Specifically, we
+present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
+and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-build}
+and~\S\ref{sec-implementations} scrutinise the abstractions further and provide
+concrete implementations for several build systems.
+
+\vspace{-1.5mm}
+\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
+\vspace{-0.5mm}
+% We begin by establishing a common vocabulary for build systems:
+
+\emph{Keys, values, and the store.} The goal of any build system is to
 bring up to date a \emph{store} that implements a mapping from \emph{keys} to
@@
 keys are filenames, and the values are file contents. In \Excel, the store is
-the worksheets, the keys are cell names (such as \cmd{A2}) and the values are
-numbers, strings etc, displayed as the cell contents. Many build systems use
+the worksheets, the keys are cell names (such as \cmd{A1}) and the values are
+numbers, strings, etc., displayed as the cell contents. Many build systems use
 \emph{hashes} of values as compact summaries with a fast equality check.
 
-\vspace{-2mm}
-\paragraph{Input, output, and intermediate values}
-Some values must be provided by the user as \emph{input}. For example,
-\cmd{main.c} can be edited by the user who relies on the build system to
-compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
-such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
-\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
-but are produced in the process of turning inputs into outputs.
+\emph{Input, output, and intermediate values.} Some values must be provided by
+the user as \emph{input}. For example, \cmd{main.c} can be edited by the user
+who relies on the build system to compile it into \cmd{main.o} and subsequently
+\cmd{main.exe}. End build products, such as \cmd{main.exe}, are \emph{output}
+values. All other values (in this case \cmd{main.o}) are \emph{intermediate};
+they are not interesting for the user but are produced in the process of turning
+inputs into outputs.
 
-\vspace{-2mm}
-\paragraph{Persistent build information} As well as the key/value mapping, the store
-also contains information maintained by the build system itself, which persists
-from one invocation of the build system to the next -- its ``memory''.
+\emph{Persistent build information.} As well as the key/value mapping, the
+store also contains information maintained by the build system itself, which
+persists from one invocation of the build system to the next -- its ``memory''.
 
-\vspace{-2mm}
-\paragraph{Task description} Any build system requires the user to specify how to
-compute the new value for one key, using the (up to date) values of its
+\emph{Task description.} Any build system requires the user to specify how
+to compute the new value for one key, using the (up to date) values of its
 dependencies. We call this specification the \emph{task description}. For
@@
 
-\vspace{-2mm}
-\paragraph{Build system} A \emph{build system} takes a task description, a
+\emph{Build system.} A \emph{build system} takes a task description, a
 \emph{target} key, and a store, and returns a new store in which the target key
@@
 
-\vspace{-2mm}
-\paragraph{Modelling in Haskell} We will model all our build systems concretely,
-as Haskell programs. To that end, Fig.~\ref{fig-types} gives the type
-declarations and function signatures of the library functions. For example,
-\hs{Store}~\hs{i}~\hs{k}~\hs{v} is the type of stores, with several associate
-functions (\hs{getValue}, etc.). We use \hs{k} as a type
-variable ranging over keys, \hs{v} for values, and \hs{i} for the persistent
-build information.
+We will model build systems concretely, as Haskell programs. To that end,
+Fig.~\ref{fig-types} provides the type signatures for all key abstractions
+introduced in the paper. For example, \hs{Store}~\hs{i}~\hs{k}~\hs{v} is the
+type of stores, with several associate functions (\hs{getValue}, etc.). We use
+\hs{k} as a type variable ranging over keys, \hs{v} for values, and \hs{i} for
+the persistent build information. Fig.~\ref{fig-stdlib} lists standard library
+definitions.
 
 \subsection{The Task abstraction}\label{sec-task}
-
-% Old type:
-% type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
 
@@
 
-\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{text}
@@
 \end{minted}
-\vspace{1mm}
 
 \noindent
-Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains
-the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
-with the following task description:
+Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains the
+formula \cmd{A1+A2}, etc.
+
+% \newpage
+
+\noindent
+We can represent the formulae of this spreadsheet with the following task
+description:
 
@@
 
+\begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+-- Applicative functors
+pure  :: Applicative f => a -> f a
+(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
+(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Standard State monad from Control.Monad.State
+data State s a
+instance Monad (State s)
+get       :: State s s
+gets      :: (s -> a) -> State s a
+put       :: s -> State s ()
+modify    :: (s -> s) -> State s ()
+runState  :: State s a -> s -> (a, s)
+execState :: State s a -> s -> s
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Standard types from Data.Functor.Identity and Data.Functor.Const
+newtype Identity a = Identity { runIdentity :: a }
+newtype Const m a  = Const    { getConst    :: m }
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+instance Functor (Const m) where
+    fmap _ (Const m) = Const m
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+instance Monoid m => Applicative (Const m) where
+    pure _              = Const mempty
+    Const x <*> Const y = Const (x <> y)
+\end{minted}
+\vspace{-3mm}
+\caption{Standard library definitions.}\label{fig-stdlib}
+\vspace{-4mm}
+\end{figure}
+
 The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit
@@
 computation~\cite{mcbride2008applicative} -- the relevant type signatures are
-given in Fig.~\ref{fig-types}. We will explain why in~\S\ref{sec-general-build}.
+given in Fig.~\ref{fig-stdlib}. We will explain why in~\S\ref{sec-general-build}.
 
@@
 \hs{Tasks}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
-compilers, calculation chains, file systems, caches, and all other
-complexities of real build systems.  It just computes a single output, in
-a side-effect-free way, using a callback (\hs{fetch}) to find the values
-of its dependencies.
+compilers, calc chains, file systems, caches, and all other complexities of real
+build systems. It just computes a single output, in a side-effect-free way,
+using a callback (\hs{fetch}) to find the values of its dependencies.
 
@@
 The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
-Fig.~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
+Fig.~\ref{fig-stdlib} -- initialised with the incoming \hs{store} by \hs{execState}.
 To bring a key up to date, \hs{fetch} asks the task description \hs{tasks} how
@@
 \subsection{The need for polymorphism in \hs{Task}}\label{sec-why-polymorphism}
+\vspace{-0.5mm}
 
@@
 expressing that \hs{f} can only be instantiated by types that satisfy the
-constraint \hs{c}.
-
-So the type \hs{Task} emerges naturally, almost inevitably.  But now that
-it \emph{has} emerged, we find that constraints \hs{c} classify task descriptions
-in a very interesting way:
+constraint \hs{c}. So the type \hs{Task} emerges naturally, almost inevitably.
+But now that it \emph{has} emerged, we find that constraints \hs{c} classify
+task descriptions in a very interesting way:
 \begin{itemize}
@@
   ``computing'' them (see \S\ref{sec-deps}).
-\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
-  description allows \emph{dynamic} dependencies, in which a formula may depend
-  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
-  value of another cell \cmd{D}.
+\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
+  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
+  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
+  \cmd{D}.
 %TODO: Add a reference for Docker?
@@
 
+\vspace{-0.5mm}
 \subsection{Monadic tasks}\label{sec-task-monad}
@@
 
-\vspace{1mm}
+\vspace{0.5mm}
 \begin{minted}[xleftmargin=10pt]{text}
@@
 \end{minted}
+\vspace{0.5mm}
 
@@
 
-\vspace{1mm}
+\vspace{0.5mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-\vspace{1mm}
+\vspace{0.5mm}
 
@@
 Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
-we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
+we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
 The use of \hs{Identity} just fixes the `impedance mismatch' between the
@@
 \noindent
-Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
+Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-stdlib}. We
 instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
@@
 and the standard definition of \hs{Applicative} for \hs{Const} (which we give
-in Fig.~\ref{fig-types}) combines the dependencies from different parts of the
+in Fig.~\ref{fig-stdlib}) combines the dependencies from different parts of the
 task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
@@
 running its code in a different \hs{Applicative}! This is not new, for example
-see~\cite{free-applicatives}, but it is cool.
+see~\cite{free-applicatives}, but it is extremely cool. We will see a practical
+use for \hs{dependencies} when implementing applicative build systems,
+see~\S\ref{sec-implementation-make}.
 
@@
 corresponding key. The \hs{task} returns the value of type
-\hs{WriterT}~\hs{[@@k]}~\hs{m}~\hs{v}, which we unwrap by \hs{runWriterT}. Here
-is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
+\hs{WriterT}~\hs{[@@k]}~\hs{m}~\hs{v}, which we unwrap by \hs{runWriterT}. We
+will use \hs{track} when implementing monadic build systems,
+see~\S\ref{sec-implementation-shake}; here we show an example of \hs{track}ing
+monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
 
-\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 which in this case is obtained by reading from the standard input.
-
-% Furthermore, the above example highlights the static cycle between cells
-% \cmd{B1} and \cmd{B2}: in \hs{store1}, we have \cmd{B1 = B2 = A1 = 10},
-% whereas in \hs{store2}, we have \cmd{B2 = B1 = A2 = 20}.
-
-% \subsection{Free tasks}
-% \todo{AM}{Introduce free constructions \hs{TaskA} and \hs{TaskM} giving
-% intuitions on what tasks really are. This will take at least one page.}
-%
-% data TaskA k v a = InputA | TaskA [k] ([v] -> a)
-%
-% data TaskM k v a = InputM | DoneM a | TaskM [k] ([v] -> Task k v a)
-
-% \subsection{Recognising an applicative task wearing a monadic hat}
-% Although dynamic dependencies are very useful, they typically represent only a
-% small fraction of all dependencies. One can exploit this in a build system by
-% recognising applicative tasks even though they have the type
-% \hs{Task}~\hs{Monad} and treating them specially.
-
-%------------------------- Old correctness definition -------------------------%
-
-% \definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
-% build task description, \hs{key} is an output key, \hs{store} is an initial
-% store, and \hs{result} is the store produced by running the build system with
-% parameters \hs{task}, \hs{key} and \hs{store}. Or, using the precise language of
-% our abstractions:
-% \vspace{1mm}
-% \begin{minted}[xleftmargin=10pt]{haskell}
-% build         :: Build c i k v
-% task          :: Task c k v
-% key           :: k
-% store, @@result :: Store i k v
-% result = @@build @@task @@key @@store
-% \end{minted}
-% \vspace{1mm}
-% \noindent
-% Then the \hs{result} is~\emph{correct} if there exists an \hs{ideal} store, such
-% that:
-% \begin{itemize}
-%     \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
-%     for all possible non-input keys \hs{k}, the result of recomputing the
-%     \hs{task} matches the value in the \hs{ideal} store:
-%     \vspace{-1mm}
-%     \[
-%     \hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}.
-%     \]
-%     \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
-%     \vspace{-1mm}
-%     \[
-%     \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
-%     \]
-%     In other words, we require that the output \hs{key} has the correct value,
-%     but do not impose any restrictions on intermediate keys, therefore
-%     permitting shallow cloud builds.
-%     \item The \hs{store}, \hs{result} and \hs{ideal} agree on all input keys
-%     \hs{k} that belong to the transitive closure of \hs{key}'s dependencies:
-%     \[
-%     \,\,\,\,\,\,\,\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{&&}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}
-%     \]
-%     This requirement asserts that (i) no inputs were corrupted during the
-%     build, and (ii)~the \hs{ideal} store has constraints not only on the output,
-%     but also on the inputs.
-% \end{itemize}
-% A build system is \emph{correct} if it produces a correct \hs{result} for any
-% given \hs{task}, \hs{key} and \hs{store}.
-% }
-% \label{def-correct}

### 2018-06-30 src/Build/Scheduler.hs
Tweak naming

@@ -42,9 +42,9 @@ liftInfo x = do
     put $ mapInfo (, snd $ getInfo $ store) newStore
     return a
 
--- | Remove one layer of computation.
-lowerInfo :: StateT i (State (Store i k v, j)) a -> State (Store i k v, j) a
-lowerInfo x = do
+-- | Collapse two layers of stateful computations into one.
+flattenInfo :: StateT i (State (Store i k v, j)) a -> State (Store i k v, j) a
+flattenInfo x = do
     info <- gets (getInfo . fst)
     (a, newInfo) <- runStateT x info
     modify $ \(s, j) -> (putInfo newInfo s, j)
@@ -199,7 +199,7 @@ suspending rebuilder tasks target store = fst $ execState (build target) (store,
                     fetch k = do lift (build k)                      -- build the key
                                  lift (gets (getInfo . fst)) >>= put -- save new traces
                                  lift (gets (getValue k . fst))      -- fetch the value
-                newValue <- lowerInfo (run newTask fetch)
+                newValue <- flattenInfo (run newTask fetch)
                 modify $ \(s, d) -> (updateValue key value newValue s, Set.insert key d)
 
 -- | An incorrect scheduler that builds the target key without respecting its

### 2018-06-30 paper/5-implementations.tex
Simplify build system implementations

@@
 
+An implementation of \Make using our framework is shown in
+Fig.~\ref{fig-make-implementation}. As promised, its definition
+is just the application of a \hs{Scheduler}, \hs{topological},
+to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
+We discuss each component in turn, starting with the rebuilder.
+
+The \hs{modTimeRebuilder} uses the pair
+\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
+build information, carried by a \hs{MonadState} monad. This
+\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
+\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
+modification times}. The rebuilder receives three arguments: a
+\hs{key}, its current \hs{value}, and the \hs{task} of type
+\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
+the \hs{key} if necessary.  The rebuilder first decides if the
+\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the key is not
+found, that must mean it has never been built before; otherwise
+\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
+(computed by \hs{dependencies}) are out of date. If the key is
+\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
+the new modification time of the key\footnote{The real \Make
+relies on the file system to track modification times, but we prefer
+to make it explicit here.}; otherwise we can just return the current value.
+We assume that the external system, which invokes the build system, updates
+\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.
+
 \begin{figure}
 \begin{minted}[fontsize=\small]{haskell}
--- Make build system
-make :: Ord k => Build Applicative (MakeInfo k) k v
-make = topological modTimeRebuilder
+-- Make build system; stores current time and file modification times
+type Time       = Integer
+type MakeInfo k = (Time, Map k Time)
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Build information: current time and file modification times
-type Time       = Integer
-type MakeInfo k = (Time, Map k Time)
+make :: Ord k => Build Applicative (MakeInfo k) k v
+make = topological modTimeRebuilder
 \end{minted}
@@
             fetch k = return (getValue k store)
-            (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
-        modify $ putInfo newInfo . putValue key newValue
+        newValue <- liftStore (run newTask fetch)
+        modify $ putValue key newValue
   where
@@
 \end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Expand the scope of visibility of a stateful computation (implementation omitted)
+liftStore :: State i a -> State (Store i k v) a
+\end{minted}
 \vspace{-3mm}
@@
 \end{figure}
-
-An implementation of \Make using our framework is shown in
-Fig.~\ref{fig-make-implementation}. As promised, its definition
-is just the application of a \hs{Scheduler}, \hs{topological},
-to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
-We discuss each component in turn, starting with the rebuilder.
-
-The \hs{modTimeRebuilder} uses the pair
-\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
-build information, carried by a \hs{MonadState} monad. This
-\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
-\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
-modification times}. The rebuilder receives three arguments: a
-\hs{key}, its current \hs{value}, and the \hs{task} of type
-\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
-the \hs{key} if necessary.  The rebuilder first decides if the
-\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the key is not
-found, that must mean it has never been built before; otherwise
-\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
-(computed by \hs{dependencies}) are out of date. If the key is
-\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
-the new modification time of the key\footnote{The real \Make
-relies on the file system to track modification times, but we prefer
-to make it explicit here.}; otherwise we can just return the current value.
-We assume that the external system, which invokes the build system, updates
-\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.
 
@@
 
+\begin{figure}
+\begin{minted}[fontsize=\small]{haskell}
+-- Excel build system; stores a dirty bit per key and calc chain
+type Chain k = [k]
+type ExcelInfo k = (k -> Bool, Chain k)
+\end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+excel :: Ord k => Build Monad (ExcelInfo k) k v
+excel = restarting dirtyBitRebuilder
+\end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- A task rebuilder based on dirty bits
+dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
+dirtyBitRebuilder key value task = Task $ \fetch -> do
+    isDirty <- get
+    if isDirty key then run task fetch else return value
+\end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- A restarting task scheduler
+restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
+restarting rebuilder tasks target = execState $ do
+    chain    <- gets (snd . getInfo)
+    newChain <- liftChain $ go Set.empty $ chain ++ [target | target `notElem` chain]
+    modify . mapInfo $ \(ir, _) -> (ir, newChain)
+  where
+    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
+    go _    []       = return []
+    go done (key:ks) = case tasks key of
+        Nothing -> (key :) <$> go (Set.insert key done) ks
+        Just task -> do
+            store <- get
+            let value = getValue key store
+                newTask :: Task (MonadState ir) k (Either k v)
+                newTask = try $ rebuilder key value task
+                fetch :: k -> State ir (Either k v)
+                fetch k | k `Set.member` done = return $ Right (getValue k store)
+                        | otherwise           = return $ Left k
+            result <- liftStore (run newTask fetch)
+            case result of
+                Left dep -> go done $ dep : filter (/= dep) ks ++ [key]
+                Right newValue -> do
+                    modify $ updateValue key value newValue
+                    (key :) <$> go (Set.insert key done) ks
+\end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Convert a total task into a task that accepts a partial fetch callback
+try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
+try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
+\end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Expand the scope of visibility of a stateful computation (implementation omitted)
+liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
+\end{minted}
+\vspace{-3mm}
+\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
+\vspace{-5mm}
+\end{figure}
+
 \Excel's \hs{restarting} scheduler processes keys in the order specified by the
@@
 
-\begin{figure}
-\begin{minted}[fontsize=\small]{haskell}
--- Excel build system
-excel :: Ord k => Build Monad (ExcelInfo k) k v
-excel = restarting approximateRebuilder
-\end{minted}
-\vspace{0mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Build information: dirty keys, approximate dependencies, and calculation chain
-type Chain k = [k]
-type ExcelInfo k = ((Set k, Map k [k]), Chain k)
-\end{minted}
-\vspace{0mm}
-\begin{minted}[fontsize=\small]{haskell}
--- A task rebuilder based on the set of dirty keys and approximate dependencies
-approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
-approximateRebuilder key value task = Task $ \fetch -> do
-    (dirtyKeys, deps) <- get
-    let dirty = key `Set.member` dirtyKeys ||
-                case Map.lookup key deps of Nothing -> True
-                                            Just ks -> any (`Set.member` dirtyKeys) ks
-    if not dirty then return value else do
-        put (Set.insert key dirtyKeys, deps)
-        run task fetch
-\end{minted}
-\vspace{0mm}
-\begin{minted}[fontsize=\small]{haskell}
--- A restarting task scheduler
-restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
-restarting rebuilder tasks target = execState $ do
-    chain    <- gets (snd . getInfo)
-    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
-    modify . mapInfo $ \(ir, _) -> (ir, newChain)
-  where
-    go :: Set k -> Chain k -> State (Store (ir, Chain k) k v) (Chain k)
-    go _    [@@]       = return [@@]
-    go done (key:ks) = case tasks key of
-        Nothing -> (key :) <$> go (Set.insert key done) ks
-        Just task -> do
-            store <- get
-            let newTask :: Task (MonadState ir) k (Either k v)
-                newTask = try $ rebuilder key (getValue key store) task
-                fetch :: k -> State ir (Either k v)
-                fetch k | k `Set.member` done = return $ Right (getValue k store)
-                        | otherwise           = return $ Left k
-            case runState (run newTask fetch) (fst $ getInfo store) of
-                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
-                (Right newValue, newInfo) -> do
-                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
-                    (key :) <$> go (Set.insert key done) ks
-\end{minted}
-\vspace{0mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Convert a total task into a task that accepts a partial fetch callback
-try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
-try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
-\end{minted}
-\vspace{-3mm}
-\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
-\vspace{-5mm}
-\end{figure}
-
 \vspace{-1mm}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Shake build system
+-- Shake build system; stores verifying traces
 shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
@@
                     fetch :: k -> StateT i (State (Store i k v, Set k)) v
-                    fetch k = do lift (build k)                      -- Build the key
-                                 lift (gets (getInfo . fst)) >>= put -- Save new traces
-                                 lift (gets (getValue k . fst))      -- Fetch the value
-                info <- gets (getInfo . fst)
-                (newValue, newInfo) <- runStateT (run newTask fetch) info
-                modify $ \(s, _) ->
-                    (putInfo newInfo $ putValue key newValue s, Set.insert key done)
+                    fetch k = do lift (build k)                      -- build the key
+                                 lift (gets (getInfo . fst)) >>= put -- save new traces
+                                 lift (gets (getValue k . fst))      -- fetch the value
+                newValue <- flattenInfo (run newTask fetch)
+                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
+\end{minted}
+\begin{minted}[fontsize=\small]{haskell}
+-- Collapse two layers of stateful computations into one (implementation omitted)
+flattenInfo :: StateT i (State (Store i k v, Set k)) a -> State (Store i k v, Set k) a
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Bazel build system, implementation of 'restarting2' is omitted (22 lines)
+-- Bazel build system; stores constructive traces
 bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
-bazel = restarting2 ctRebuilder
+bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- CloudBuild build system, implementation of 'topological' is given in Fig. 6
+-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
 cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Cloud Shake build system, implementation of 'suspending' is given in Fig. 8
+-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
 cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Buck build system, implementation of 'topological' is given in Fig. 6
+-- Buck build system, implementation of 'topological' is given in Fig. 7
 buck :: (Hashable k, Hashable v) => Build Applicative (DCT k v) k v
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Nix build system, implementation of 'suspending' is given in Fig. 8
+-- Nix build system, implementation of 'suspending' is given in Fig. 9
 nix :: (Hashable k, Hashable v) => Build Monad (DCT k v) k v

### 2018-07-01 paper/5-implementations.tex
Switch to deep constructive traces

@@
 \end{minted}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 ctRebuilder key value task = Task $ \fetch -> do
-    maybeCachedValue <- constructCT key value (fmap hash . fetch) =<< get
-    case maybeCachedValue of
-        Just cachedValue -> return cachedValue
-        Nothing -> do
-            (newValue, deps) <- track task fetch
+    cachedValues <- constructCT key (fmap hash . fetch) =<< get
+    if value `elem` cachedValues then return value else case cachedValues of
+        (cachedValue:_) -> return cachedValue
+        _ -> do (newValue, deps) <- trackM task fetch
                 put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
                 return newValue
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
+cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
+cloudShake = suspending ctRebuilder
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
-cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
-cloudShake = suspending ctRebuilder
+-- Buck build system, implementation of 'topological' is given in Fig. 7
+buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
+buck = topological (adaptRebuilder dctRebuilder)
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Rebuilder based on deep constructive traces, analogous to ctRebuilder
+dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
+\end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Nix build system, implementation of 'suspending' is given in Fig. 9
+nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
+nix = suspending dctRebuilder
 \end{minted}
 \vspace{-2mm}
-\caption{Implementations of \Bazel, \CloudBuild and \Cloud \Shake using our
-framework; \hs{restarting2} differs from the restarting scheduler used by \Excel
-(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the
-calculation chain.}
-\label{fig-bazel-cloud-shake-implementation}
+\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
+our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
+(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
+chain.}
+\label{fig-cloud-implementations}
 \vspace{-4mm}
@@
 In this section we use constructive traces to implement cloud build systems.
-Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models of \Bazel,
+Fig.~\ref{fig-cloud-implementations} shows our models of \Bazel,
 \CloudBuild and \Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a
@@
 we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
-of \Bazel, as shown in Fig.~\ref{fig-bazel-cloud-shake-implementation}.
+of \Bazel, as shown in Fig.~\ref{fig-cloud-implementations}.
 
@@
 obtained by directly applying the presented ideas.
-
-\begin{figure}
-\begin{minted}[fontsize=\small]{haskell}
--- Buck build system, implementation of 'topological' is given in Fig. 7
-buck :: (Hashable k, Hashable v) => Build Applicative (DCT k v) k v
-buck = topological (adaptRebuilder dctRebuilder)
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Nix build system, implementation of 'suspending' is given in Fig. 9
-nix :: (Hashable k, Hashable v) => Build Monad (DCT k v) k v
-nix = suspending dctRebuilder
-\end{minted}
-\vspace{1mm}
-\begin{minted}[fontsize=\small]{haskell}
--- Rebuilder based on deterministic constructive traces
-dctRebuilder :: (Hashable k, Hashable v) => Rebuilder Monad (DCT k v) k v
-dctRebuilder key _value task = Task $ \fetch -> do
-    dct <- get
-    maybeCachedValue <- constructDCT key (fmap hash . fetch) dct
-    case maybeCachedValue of
-        Just cachedValue -> return cachedValue
-        Nothing -> do
-            (newValue, deps) <- track task fetch
-            put =<< recordDCT key newValue deps (fmap hash . fetch) =<< get
-            return newValue
-\end{minted}
-\vspace{-2mm}
-\caption{Implementations of \Buck and \Nix using our framework.}
-\label{fig-buck-nix-implementation}
-\vspace{-2mm}
-\end{figure}
 
@@
 
-Fig.~\ref{fig-buck-nix-implementation} shows our models of \Buck and \Nix build
+Fig.~\ref{fig-cloud-implementations} shows our models of \Buck and \Nix build
 systems that use the \hs{dctRebuilder} -- a rebuilder based on deterministic

### 2018-07-01 src/Build/Scheduler.hs
Simplify liftStore

@@ -28,11 +28,9 @@ type Scheduler c i j k v = Rebuilder c j k v -> Build c i k v
 
 -- | Lift a computation operating on @i@ to @Store i k v@.
 liftStore :: State i a -> State (Store i k v) a
-liftStore x = do
-    store <- get
-    let (a, newInfo) = runState x (getInfo store)
-    modify $ putInfo newInfo
-    return a
+liftStore x = do (a, newInfo) <- gets (runState x . getInfo)
+                 modify (putInfo newInfo)
+                 return a
 
 -- | Lift a computation operating on @Store i k v@ to @Store (i, j) k v@.
 liftInfo :: State (Store i k v) a -> State (Store (i, j) k v) a

### 2018-07-01 paper/5-implementations.tex
Work on S5.1

@@
 So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
-it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
-The \hs{Rebuilder} makes uses of the persistent build information \hs{ir}, while
-the scheduler might augment that with further persistent information of its own,
-yielding \hs{i}.
+it to construct a \hs{Build} system, by choosing which keys to rebuild in which
+order. The \hs{Rebuilder} makes uses of the persistent build information
+\hs{ir}, while the scheduler might augment that with further persistent
+information of its own, yielding \hs{i}.
 
@@
 the \hs{key} if necessary.  The rebuilder first decides if the
-\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the key is not
+\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the \hs{key} is not
 found, that must mean it has never been built before; otherwise
 \hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
-(computed by \hs{dependencies}) are out of date. If the key is
+(computed by \hs{dependencies}) are out of date. If the \hs{key} is
 \hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
-the new modification time of the key\footnote{The real \Make
+the new modification time of the \hs{key}\footnote{The real \Make
 relies on the file system to track modification times, but we prefer
-to make it explicit here.}; otherwise we can just return the current value.
+to make it explicit here.}; otherwise we can just return the current \hs{value}.
 We assume that the external system, which invokes the build system, updates
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Expand the scope of visibility of a stateful computation (implementation omitted)
+-- Expand the scope of visibility of a stateful computation
 liftStore :: State i a -> State (Store i k v) a
+liftStore x = do (a, newInfo) <- gets (runState x . getInfo)
+                 modify (putInfo newInfo)
+                 return a
 \end{minted}
@@
 \hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
-\hs{newValue} and updated build information \hs{newInfo}, which are subsequently
-written to the store. The computation runs in the
-\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
-build system (\S\ref{sec-general-build}), however since \hs{newTask} has access
-only to the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}, we need to use
-\hs{runState} to execute it. The \hs{newTask} finds values of the \hs{key}'s
-dependencies via the \hs{fetch} callback, which is defined to directly read the
-\hs{store}.
+\hs{newValue}. Note that \hs{newTask} has access only to the \hs{i} part of the
+\hs{Store}~\hs{i}~\hs{k}~\hs{v}, but the rest of the \hs{do} block runs in the
+\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad; we use the (unremarkable)
+helper function \hs{liftStore} to fix the mismatch. The \hs{newTask} finds
+values of the \hs{key}'s dependencies via the \hs{fetch} callback, which is
+defined to directly read the \hs{store}.
 
@@
 \end{minted}
-\vspace{1mm}
+\vspace{0mm}
 \begin{minted}[fontsize=\small]{haskell}

### 2018-07-01 paper/5-implementations.tex
Work on S5.2: simplify the model of Excel

@@
 can be performed on the original \hs{Task}~\hs{Applicative} if need be.
-
-These two abstractions are the key to modularity: \emph{we can combine any
-scheduler with any rebuilder, and obtain a correct build system}.
-In this section we will write a scheduler for each column of
-Table~\ref{tab-build-systems}, and a rebuilder for each row; then combine them
-to obtain the build systems in the table's body.
-
-\subsection{\Make}\label{sec-implementation-make}
-
-An implementation of \Make using our framework is shown in
-Fig.~\ref{fig-make-implementation}. As promised, its definition
-is just the application of a \hs{Scheduler}, \hs{topological},
-to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
-We discuss each component in turn, starting with the rebuilder.
-
-The \hs{modTimeRebuilder} uses the pair
-\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
-build information, carried by a \hs{MonadState} monad. This
-\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
-\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
-modification times}. The rebuilder receives three arguments: a
-\hs{key}, its current \hs{value}, and the \hs{task} of type
-\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
-the \hs{key} if necessary.  The rebuilder first decides if the
-\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the \hs{key} is not
-found, that must mean it has never been built before; otherwise
-\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
-(computed by \hs{dependencies}) are out of date. If the \hs{key} is
-\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
-the new modification time of the \hs{key}\footnote{The real \Make
-relies on the file system to track modification times, but we prefer
-to make it explicit here.}; otherwise we can just return the current \hs{value}.
-We assume that the external system, which invokes the build system, updates
-\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.
 
@@
 
+These two abstractions are the key to modularity: \emph{we can combine any
+scheduler with any rebuilder, and obtain a correct build system}.
+In this section we will write a scheduler for each column of
+Table~\ref{tab-build-systems}, and a rebuilder for each row; then combine them
+to obtain the build systems in the table's body.
+
+\subsection{\Make}\label{sec-implementation-make}
+
+An implementation of \Make using our framework is shown in
+Fig.~\ref{fig-make-implementation}. As promised, its definition
+is just the application of a \hs{Scheduler}, \hs{topological},
+to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
+We discuss each component in turn, starting with the rebuilder.
+
+The \hs{modTimeRebuilder} uses the pair
+\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
+build information, carried by a \hs{MonadState} monad. This
+\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
+\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
+modification times}. The rebuilder receives three arguments: a
+\hs{key}, its current \hs{value}, and the \hs{task} of type
+\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
+the \hs{key} if necessary.  The rebuilder first decides if the
+\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the \hs{key} is not
+found, that must mean it has never been built before; otherwise
+\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
+(computed by \hs{dependencies}) are out of date. If the \hs{key} is
+\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
+the new modification time of the \hs{key}\footnote{The real \Make
+relies on the file system to track modification times, but we prefer
+to make it explicit here.}; otherwise we can just return the current \hs{value}.
+We assume that the external system, which invokes the build system, updates
+\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.
+
 \Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based on a
@@
 
-Our \Excel's model is a composition of the \hs{restarting} scheduler and the
-\hs{approximateRebuilder}, as shown in Fig.~\ref{fig-excel-implementation}.
-The persistent build information \hs{ExcelInfo}~\hs{k} is a triple:
+Our model of \Excel uses the \hs{restarting} scheduler and the
+\hs{dirtyBitRebuilder}, see Fig.~\ref{fig-excel-implementation}. The persistent
+build information \hs{ExcelInfo}~\hs{k} is a pair: (i) a map
+\hs{k}~\hs{->}~\hs{Bool} associating a dirty bit with every key, and (ii) a
+calc \hs{Chain}~\hs{::}~\hs{[@@k]} recorded in the previous build
+(\S\ref{sec-background-excel}).
 
-\begin{itemize}
-    \item A set of dirty keys \hs{Set}~\hs{k}, i.e. all cells whose contents
-          has changed since the previous build, including input cells and
-          formulas.
-    \item A map \hs{Map}~\hs{k}~\hs{[@@k]} of known upper bounds of
-          dependencies. If there is an entry for a key, it is a conservative
-          approximation of its dependencies (e.g. the dependencies of the
-          formula \cmd{IF(C1=1,B2,A2)} are approximated by the list
-          \hs{[}\cmd{C1}\hs{,}\cmd{B2}\hs{,}\cmd{A2}\hs{]}). Otherwise, there is
-          no upper bound (e.g. because the key uses an \cmd{INDIRECT} reference)
-          and the key is assumed to be dirty.
-    \item A calculation \hs{Chain} recorded in the previous build
-          (\S\ref{sec-background-excel}), represented by a list of keys \hs{[@@k]}.
-\end{itemize}
-
-The \hs{approximateRebuilder} uses the first two elements of the above triple
-to decide whether a given \hs{key} is \hs{dirty} and hence needs to be rebuilt.
-The logic is similar to that of \Make's \hs{modTimeRebuilder}, but instead of
-relying on file modification times, \Excel stores the information about dirty
-keys more directly. Furthermore, \Excel's dependency approximation makes it
-possible to build tasks with dynamic dependencies: a \hs{key} is declared
-\hs{dirty} if any of its approximate dependencies is in the \hs{dirtyKeys} set,
-or if there is no known upper bound on its dependencies, i.e. if the lookup in
-the map \hs{deps}~\hs{::}~\hs{Map}~\hs{k}~\hs{[@@k]} fails. Note that when a
-\hs{key} is rebuilt it is added to the set \hs{dirtyKeys} to trigger further
-rebuilds.
+The \hs{dirtyBitRebuilder} is very simple: if the \hs{key}'s dirty bit is set,
+we \hs{run} the \hs{task} to rebuild the \hs{key}; otherwise we return the
+current \hs{value} as is. The external system, which invokes \Excel's build
+engine, is required to provide a transitively closed set of dirty bits.
+Therefore, unlike \Make's \hs{modTimeRebuilder}, the \hs{dirtyBitRebuilder} does
+not need to modify \hs{i} to trigger rebuilds of dependent keys.
 
@@
             store <- get
-            let value = getValue key store
-                newTask :: Task (MonadState ir) k (Either k v)
-                newTask = try $ rebuilder key value task
+            let newTask :: Task (MonadState ir) k (Either k v)
+                newTask = try $ rebuilder key (getValue key store) task
                 fetch :: k -> State ir (Either k v)
@@
                         | otherwise           = return $ Left k
-            result <- liftStore (run newTask fetch)
+            result <- liftStore (run newTask fetch) -- liftStore is defined in Fig. 7
             case result of
@@
                 Right newValue -> do
-                    modify $ updateValue key value newValue
+                    modify $ putValue key newValue
                     (key :) <$> go (Set.insert key done) ks
@@
 \Excel's \hs{restarting} scheduler processes keys in the order specified by the
-calculation \hs{chain}. During the build, it constructs a \hs{newChain} for the
-next build and maintains a set of keys \hs{done} that have been processed. For
-each non-input \hs{key}, the scheduler tries to rebuild it using a partial
-\hs{fetch} callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}.
-The callback is defined to fail with \hs{Left}~\hs{dep} when asked for the value
-of a dependency \hs{dep} that has not yet been processed (and hence may
-potentially be dirty); otherwise it returns the current value of the dependency
-by looking it up in the \hs{store}.
+calc \hs{chain}. During the build, it constructs a \hs{newChain} for the next
+build and maintains a set of keys \hs{done} that have been processed. For each
+non-input \hs{key}, the scheduler tries to rebuild it using a partial \hs{fetch}
+callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}. The callback
+is defined to fail with \hs{Left}~\hs{dep} when asked for the value of a
+dependency \hs{dep} that has not yet been processed (and hence may potentially
+be dirty); otherwise it returns the current value of the dependency by looking
+it up in the \hs{store}.
 
-After the \hs{newTask} is executed with \hs{runState} there are two cases to
-consider:
+After the \hs{newTask} is executed (with the help of \hs{liftStore}) there are
+two cases to consider:
 
@@
     has not yet been processed. This indicates that the calculation \hs{chain}
-    from the previous build is incorrect and needs to be adjusted by making sure
-    it contains \hs{dep} and moving the \hs{key} to the end, so that its build
-    can be restarted after the \hs{dep} is ready.
-    \item The \hs{newTask} succeeded. The resulting \hs{newValue} and
-    \hs{newInfo} are written to the store, the \hs{key} is marked as \hs{done},
-    and \Excel continues to build the \hs{chain}.
+    from the previous build is incorrect and needs to be adjusted by moving the
+    \hs{dep} in front of the \hs{key}, so that we can restart building the
+    \hs{key} after the \hs{dep} is ready.
+    \item The \hs{newTask} succeeded. The resulting \hs{newValue} is written to
+    the store, the \hs{key} is marked as \hs{done}, and \Excel continues to
+    build the rest of the \hs{chain}.
 \end{itemize}
@@
 We fix the mismatch with the function \hs{try} that relies on the standard
-monad transformer \hs{ExceptT} from the \cmd{transformers} library.
+monad transformer \hs{ExceptT} from the \cmd{transformers} library. We also
+need the \hs{liftChain} helper function, whose implementation we omit since it
+is analogous to \hs{liftStore} defined in Fig.~\ref{fig-make-implementation}.
 

### 2018-07-01 paper/4-build.tex
Work on S4

@@
 changed, the key is dirty and must be recomputed -- in essence a \emph{trace}
-which we can use to \emph{verify} existing values. We can describe a trace as:
+which we can use to \emph{verify} existing values. For traces, there are two essential
+operations -- adding a new value to the trace store, and using the traces to determine if something needs
+rebuilding. Assuming a store of verifying traces \hs{VT}~\hs{k}~\hs{v}, the operations are:
 
 \begin{minted}{haskell}
-data Trace k v = Trace
-    { key     :: k
-    , depends :: [(k, Hash v)]
-    , result  :: Hash v }
+recordVT :: k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
+verifyVT :: (Monad m, Eq k, Eq v)
+         => k -> Hash v -> (k -> m (Hash v)) -> VT k v -> m Bool
 \end{minted}
@@
 We assume that \hs{Hash}~\hs{v} is a small constant size, constructed from hashing the
-underlying \hs{v} rather than storing it directly. Checking a trace requires
-ensuring all the dependencies are up to date (using whatever ordering strategy
-as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are
-same as the current value and that the result is the same.
+underlying \hs{v} rather than storing it directly.
+After successfully building a key, we call \hs{recordVT} passing the key, the hash of its value, and the list of hashes and dependencies.
+Combined with the existing set of verifying traces, that provides a new set of verifying traces.
 
-A build system that uses verifying traces needs to persistently maintain a set
-of traces. After computing a fresh value we add its \hs{Trace}~\hs{k}~\hs{v} to
-the set. Therefore the information stored by a build system that verifies
-traces can be modelled as a list or set~--~we chose \hs{[Trace}~\hs{k}~\hs{v]}
-for simplicity. In practice, different build systems can optimise the data
-structures used by traces for their specific use cases, which we discuss
-in~\S\ref{sec-smart-traces}.
+More interestingly, to \emph{verify} whether a key needs rebuilding we supply the key and the hash of the current result in the store, along with a function for obtaining the post-build result of any key (using whatever ordering strategy
+as per \S\ref{sec-dependency-orderings}), and the existing \hs{VT} information.
+The result will be a \hs{Bool} where \hs{True} indicates that the trace could be verified, or \hs{False} to indicate the key should be rebuilt.
+
+One potential implementation would be to record all arguments passed to \hs{recordVT}, and verify by simply checking if any argument matches the information passed by \hs{verifyVT} (we discuss smarter implementations in~\S\ref{sec-smart-traces}).
 
@@
 Once we are storing the complete result it makes sense to record many constructive traces per key, and to share them with other users, providing cloud-build functionality.
-We can represent that by changing \hs{Trace} to:
+We can represent this additional information by providing the operations:
 
 \begin{minted}{haskell}
-data Trace k v = Trace
-    { key     :: k
-    , depends :: [(k, Hash v)]
-    , result  :: v }
+recordCT :: k -> v -> [(k, Hash v)] -> CT k v -> CT k v
+constructCT :: (Monad m, Eq k, Eq v) => k -> (k -> m (Hash v)) -> CT k v -> m [v]
 \end{minted}
 
-The change from~\S\ref{sec-verifying-traces} is that instead of storing the \hs{Hash}~\hs{v} for \hs{result}, we now store the full value \hs{v}.
-Checking a trace follows the same pattern as before, but if the result is the only thing that is different we can simply use the result \emph{without} recomputing it.
+The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing the hash of the result, we require the whole result.
+The \hs{verifyVT} has been replaced with \hs{constructCT}, which instead of taking the hash of the value, instead returns a list of suitable values.
+If the current value in the store matches one of the possible values, the build can skip this rule.
+If the resulting list is empty, the build must rerun.
+However, if the current value does not match the store, and there is a possible value, we can use any value from the constructive list \emph{without} doing any work to compute it.
 
@@
 do not avoid calls to \hs{compute}, merely overheads in the build system itself.
+
+\begin{minted}{haskell}
+data Trace k v = Trace
+    { key     :: k
+    , depends :: [(k, Hash v)]
+    , result  :: Hash v }
+\end{minted}
 

### 2018-07-01 paper/4-build.tex
More work on S4

@@
 
-More interestingly, to \emph{verify} whether a key needs rebuilding we supply the key and the hash of the current result in the store, along with a function for obtaining the post-build result of any key (using whatever ordering strategy
+More interestingly, to \emph{verify} whether a key needs rebuilding we supply the key and the hash of the current result in the store, along with a function for obtaining the post-build result of any key (using an ordering strategy
 as per \S\ref{sec-dependency-orderings}), and the existing \hs{VT} information.
@@
 recordCT :: k -> v -> [(k, Hash v)] -> CT k v -> CT k v
-constructCT :: (Monad m, Eq k, Eq v) => k -> (k -> m (Hash v)) -> CT k v -> m [v]
+constructCT :: (Monad m, Eq k, Eq v)
+            => k -> (k -> m (Hash v)) -> CT k v -> m [v]
 \end{minted}
@@
 
-\subsubsection{Deterministic constructive traces}\label{sec-deterministic-constructive-traces}
+\subsubsection{Deep constructive traces}\label{sec-deep-constructive-traces}
 
-If we assume all actions are \emph{deterministic} (specifically that given the same inputs they produce the same output) then we can make a further optimisation to arrive at deterministic constructive traces:
+Constructive traces always verify keys by looking at their immediate dependencies, which must have first been brought up to date, meaning that the time to verify a key is based on the number of transitive dependencies.
+A \emph{deep} constructive trace optimises this process by only looking at the terminal input keys, ignoring any intermediate dependencies.
+The operations capturing this approach are the same as for constructive traces in~\S\ref{sec-constructive-traces}, but we use the names \hs{constructDCT} and \hs{recordDCT}, where the underlying \hs{DCT} state need only record information about hashes of inputs, not intermediate dependencies.
 
-\begin{minted}{haskell}
-data Trace k v = Trace
-    { key     :: k
-    , depends :: [(k, Hash (Tree (Hash v)))]
-    , result  :: v
-    }
-\end{minted}
+The deep constructive trace approach only works if the operations are \emph{deterministic}, otherwise inconsistencies arise (see~\S\ref{sec-non-determinism}).
+Concretely, if the result of building a key given a set of immediate dependencies can vary, then it is possible that key \hs{k} will have been computed locally, but something \emph{using} \hs{k} will have been taken as a constructive input using a different value for \hs{k}.
+If another rule combines those two values then two different values of \hs{k} will have been used in one rule, leading to problems such as link-errors for typical compilers.
 
-Instead of storing the hash of the immediate dependencies, a determinstic constructive trace only ever takes the hash of initial inputs.
-To form the hash of a dependency, it forms the tree of dependencies all the way to the inputs, and hashes that.
-As an example, if \hs{a} depends on the input \hs{b}, then the dependency value used for \hs{a} uses \hs{hash}~\hs{b}, \emph{but not} \hs{hash}~\hs{a}.
-As a consequence, given the hashes of all inputs, and dependency information (e.g.
-from an \hs{Applicative} constraint) it is possible to compute the dependency hash of \emph{any} node, without having built any intermediate nodes.
-This property is particularly useful for ``shallow builds'', as discussed in~\S\ref{sec-cloud-aspects}.
+While current build systems using deep constructive traces store only hashes of inputs, that is not a strict requirement, as the technique works equally well by skipping $n$ levels of dependency - where only inputs is the special case of $\infty$.
+Similarly, constructive traces can be described as taking the direct dependencies, and thus being deep constructive traces of $1$.
+A downside of deep constructive traces is that they can't support cutoff (\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies.
+On the otherhand, these traces are particularly useful for ``shallow builds'', as discussed in~\S\ref{sec-cloud-aspects}.
 
@@
 
-In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list
-of traces~--~however, using a list necessarily means that finding the right trace
-takes $O(n)$. For each of the \hs{Trace} based systems it is possible to devise
-a smarter representation, which we sketch below. Note that these implementations
-do not avoid calls to \hs{compute}, merely overheads in the build system itself.
+In the examples above, we have used abstract types for the traces. Concretely, in our example code, they are all recorded as lists of:
 
 \begin{minted}{haskell}
-data Trace k v = Trace
+data Trace k h r = Trace
     { key     :: k
-    , depends :: [(k, Hash v)]
-    , result  :: Hash v }
+    , depends :: [(k, h)]
+    , result  :: r }
 \end{minted}
+
+Where \hs{k} is the key type, \hs{h} is always \hs{Hash}~\hs{v} and \hs{r} is either \hs{Hash}~\hs{v} (verify traces) or \hs{v} (constructive traces).
+A real system is highly likey to use a more optimised implementation.
+Some of the most obvious optimisations are:
 
@@
 \item Many cloud build systems store hashes of values in the trace information, then have a separate ``blob server'' which associates hashes with their actual contents.
-\item Finally, a \hs{Monad} build system with constructive traces can be stored as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
-\begin{minted}[xleftmargin=10pt]{haskell}
-data Choice k v = Choice k (Map (Hash v) (Choice k v))
-                | Result (Hash v)
-\end{minted}
-Here the \hs{Choice} encodes a tree, asking successive questions about keys, and taking different branches based on the answers, until it reaches a final result.
-Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
 \end{enumerate}
@@
 Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
-Deterministic constructive traces\hspace{2mm}\hfill\S\ref{sec-deterministic-constructive-traces} & \Buck       & -      & \Nix           \\\hline
+Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
 \end{tabular}

### 2018-07-01 paper/5-implementations.tex
More work on S4

@@
 Fig.~\ref{fig-cloud-implementations} shows our models of \Buck and \Nix build
-systems that use the \hs{dctRebuilder} -- a rebuilder based on deterministic
-constructive traces (\S\ref{sec-deterministic-constructive-traces}). As one can
+systems that use the \hs{dctRebuilder} -- a rebuilder based on deep
+constructive traces (\S\ref{sec-deep-constructive-traces}). As one can
 see, \hs{dctRebuilder} is almost identical to \hs{ctRebuilder}, but it ignores

### 2018-07-01 paper/3-abstractions.tex
Switch to new track function

@@
 of \hs{compute} and \hs{dependencies} that computes both the resulting value and
-the list of its dependencies in an arbitrary monadic context~\hs{m}:
+the list of its dependencies (key/value pairs) in an arbitrary monadic
+context~\hs{m}:
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [k])
+track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [(k, v)])
 track task fetch = runWriterT $ run task trackingFetch
   where
-    trackingFetch :: k -> WriterT [k] m v
-    trackingFetch k = tell [k] >> lift (fetch k)
+    trackingFetch :: k -> WriterT [(k, v)] m v
+    trackingFetch k = do v <- lift (fetch k); tell [(k, v)]; return v
 \end{minted}
@@
 This implementation uses the standard Haskell \hs{WriterT} \emph{monad
-transformer}~\cite{liang1995monad}, which allows us
-to record additional information -- a list of keys of type \hs{[@@k]} -- when
-computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
-with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
-corresponding key. The \hs{task} returns the value of type
-\hs{WriterT}~\hs{[@@k]}~\hs{m}~\hs{v}, which we unwrap by \hs{runWriterT}. We
-will use \hs{track} when implementing monadic build systems,
-see~\S\ref{sec-implementation-shake}; here we show an example of \hs{track}ing
-monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
+transformer}~\cite{liang1995monad}, which allows us to record additional
+information -- a list of key/value pairs \hs{[(@@k,}~\hs{v)]} -- when computing
+a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
+\hs{trackingFetch} that, in addition to fetching a value, tracks the
+corresponding key/value pair. The \hs{task} returns the value of type
+\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap by
+\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
+with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
+example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
 
@@
 B2: 10
-(10,["C1","B2"])
+(10,[("C1",1),("B2",10)])
 \end{minted}
@@
 A2: 20
-(20,["C1","A2"])
+(20,[("C1",2),("A2",20)])
 \end{minted}

### 2018-07-02 paper/5-implementations.tex
Work on S5

@@
 vtRebuilder key value task = Task $ \fetch -> do
-    upToDate <- verifyVT key value (fmap hash . fetch) =<< get
+    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
     if upToDate then return value else do
         (newValue, deps) <- track task fetch
-        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
+        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
         return newValue
@@
 \hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
-The pair \hs{(@@newValue,}~\hs{newInfo)} obtained by running the \hs{newTask} is
+We use the helper function \hs{flattenInfo} to collapse two layers of stateful
+computations into one. The \hs{newValue} obtained by running the \hs{newTask} is
 stored, and the \hs{key} is added to the set \hs{done}.
 
-\subsection{Cloud build systems: \Bazel, \CloudBuild and \Cloud \Shake}\label{sec-implementation-cloud}
+\subsection{Cloud build systems: \Bazel, \CloudBuild, \Cloud \Shake, \Buck and \Nix}\label{sec-implementation-cloud}
 
@@
         (cachedValue:_) -> return cachedValue
-        _ -> do (newValue, deps) <- trackM task fetch
-                put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
+        _ -> do (newValue, deps) <- track task fetch
+                modify $ recordCT key newValue [ (k, hash v) | (k, v) <- deps ]
                 return newValue
@@
 
-In this section we use constructive traces to implement cloud build systems.
-Fig.~\ref{fig-cloud-implementations} shows our models of \Bazel,
-\CloudBuild and \Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a
-rebuilder based on constructive traces (\S\ref{sec-constructive-traces}).
+Fig.~\ref{fig-cloud-implementations} shows our models of several cloud build
+systems. \Bazel, \CloudBuild and \Cloud \Shake are based on constructive traces
+(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
+version (\S\ref{sec-deep-constructive-traces}).
 
@@
 in Fig.~\ref{fig-shake-implementation}, but the \hs{verifyVT} query is replaced
-with a more powerful query to \hs{constructCT} that can construct a resulting
-\hs{cachedValue} by looking it up the cloud cache. If the cache lookup fails, we
-fall back to executing the \hs{task}. The obtained \hs{newValue} and the
+with a more powerful query to \hs{constructCT} that returns a list of suitable
+\hs{cachedValues} by looking them up the cloud cache. If the current \hs{value}
+is in the list, we can use it as is. Otherwise, if the list is non-empty, we can
+use an arbitrary \hs{cachedValue}. Finally, if the cache has no suitable values,
+we fall back to executing the \hs{task}. The obtained \hs{newValue} and the
 \hs{task}'s dependencies are recorded as a new constructive trace for future
@@
 Fig.~\ref{fig-excel-implementation}, but instead of building keys in the order
-specified by the persistently stored calculation chain, \Bazel uses a
-\emph{build queue}. The build starts with the queue containing all dirty keys.
-Similar to \Excel, the rebuilding of a key extracted from the queue may fail
-because one of its dynamic dependencies is dirty. In this case the key is marked
-as \emph{blocked} and its rebuilding is deferred. Whenever a key is successfully
+specified by the persistently stored calc chain, \Bazel uses a \emph{build
+queue}. The build starts with the queue containing all dirty keys. Similar to
+\Excel, the rebuilding of a key extracted from the queue may fail because one of
+its dynamic dependencies is dirty. In this case the key is marked as
+\emph{blocked} and its rebuilding is deferred. Whenever a key is successfully
 rebuilt, all keys that were previously blocked on it are added back to the
@@
 
-\textbf{TODO:} A real implementation of \Cloud \Shake is now available... It was
-obtained by directly applying the presented ideas.
+Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the
+\hs{dctRebuilder}, a rebuilder based on deep constructive traces
+(\S\ref{sec-deep-constructive-traces}), whose implementation we omit since it
+is very similar to that of \hs{ctRebuilder}. \Buck uses the \hs{topological}
+scheduler and is therefore an applicative build system, whereas \Nix uses the
+\hs{suspending} scheduler and is monadic. Deep constructive traces allow these
+build systems to efficiently support shallow builds: the value of the target key
+can be downloaded by examining only the hashes of all transitive inputs,
+skipping the intermediate keys. The downside of this approach is the loss of the
+early cutoff property.
 
-\subsection{Buck and Nix}
-
-Some build systems allow to exploit \emph{build task determinism} for efficient
-fetching of build results from the cache: one can simply hash the values of all
-transitive inputs of a key (which can be done efficiently by storing all hashes
-in a Merkle tree) and download the matching value from the cache, without the
-need to build any intermediate values. This can bring significant performance
-benefits in practice, but sacrifices the early cutoff property: indeed, with
-this approach addition of a new comment to one of the source files will
-invalidate all dependent keys.
-
-Fig.~\ref{fig-cloud-implementations} shows our models of \Buck and \Nix build
-systems that use the \hs{dctRebuilder} -- a rebuilder based on deep
-constructive traces (\S\ref{sec-deep-constructive-traces}). As one can
-see, \hs{dctRebuilder} is almost identical to \hs{ctRebuilder}, but it ignores
-the current \hs{value}, since the \hs{constructDCT} query only needs to examine
-the transitive inputs of the given \hs{key}. \Buck uses the \hs{topological}
-scheduler and is therefore an applicative build system, whereas \Nix uses the
-\hs{suspending} scheduler and is monadic.
+% Some build systems allow to exploit \emph{build task determinism} for efficient
+% fetching of build results from the cache: one can simply hash the values of all
+% transitive inputs of a key (which can be done efficiently by storing all hashes
+% in a Merkle tree) and download the matching value from the cache, without the
+% need to build any intermediate values. This can bring significant performance
+% benefits in practice, but sacrifices the early cutoff property: indeed, with
+% this approach addition of a new comment to one of the source files will
+% invalidate all dependent keys.

### 2018-07-02 paper/4-build.tex
Work on S4

@@
 
-The idea of a dirty bit is to have one piece of persistent
-information per key, saying whether the key is
-\emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
-next build starts, anything that changed between the two states is marked
-dirty.
+The idea of a dirty bit is to have one piece of persistent information per key,
+saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
+are set to clean. When the next build starts, anything that changed between the
+two states is marked dirty.
 % ; and by marking additional things dirty/clean the build system can track
 % what needs to rebuild.
-When reaching a key, if it and all its transitive dependencies are clean, the
-key does not need recomputing.
+If a key and all its transitive dependencies are clean, the key does not need
+to be rebuild.
 
 \Excel models the dirty bit approach most directly, having an actual dirty bit
-associated with each cell, marking the cell dirty if the user modifies it.
-When rebuilding, if a cell only depends on clean cells it is skipped, otherwise
-it is rebuilt and marked dirty so that the cells that depend on it are
-subsequently rebuilt too.
+associated with each cell, marking the cell dirty if the user modifies it, and
+also making sure all cells that depend on it are recursively marked dirty too.
+When rebuilding, if a clean cell only depends on clean cells it is skipped,
+otherwise it is rebuilt.
 
@@
 cell dirty if \emph{any} potential dependency has changed, even on the untaken
-\emph{if} branch. For functions such as \cmd{INDIRECT} whose dependencies cannot
+\cmd{IF} branch. For functions such as \cmd{INDIRECT} whose dependencies cannot
 be guessed, it conservatively assumes the dependencies have always changed.
 
-With a dirty bit it is simple to achieve minimality. However, to achieve early cutoff
-(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
-after a computation that did not change the value. \Excel could use this
-approach, but does not. In contrast, \Make cannot implement early cutoff nicely -- to
-do so it would have to mark the node clean (so it would not rebuild in the next
-run) and at the same time not mark the things it depends on dirty -- an
+With a dirty bit it is simple to achieve minimality. However, to achieve early
+cutoff (\S\ref{sec-background-shake}) it would be important to not set the dirty
+bit after a computation that did not change the value. \Excel could use this
+approach, but does not. In contrast, \Make cannot implement early cutoff nicely
+-- to do so it would have to mark the key clean (so it would not rebuild in the next
+run) and at the same time not mark the keys that depend on it dirty -- an
 impossible task with only the ability to update to the latest modification time.
-\Make can approximate early cutoff by not modifying the result file, and not marking it clean,
-but then it will rerun in every subsequent build.
+\Make can approximate early cutoff by not modifying the result file, and not
+marking it clean, but then it will rerun in every subsequent build.
 
@@
 
-An alternative way to determine if a key is dirty is to record what state the
-values/hashes of dependencies were used at last time, and if something has
-changed, the key is dirty and must be recomputed -- in essence a \emph{trace}
-which we can use to \emph{verify} existing values. For traces, there are two essential
-operations -- adding a new value to the trace store, and using the traces to determine if something needs
-rebuilding. Assuming a store of verifying traces \hs{VT}~\hs{k}~\hs{v}, the operations are:
+An alternative way to determine if a key is dirty is to record the
+values/hashes of dependencies used last time, and if something has changed, the
+key is dirty and must be rebuilt~--~in essence a \emph{trace} which we can use
+to \emph{verify} existing values. For traces, there are two essential
+operations~--~adding a new value to the trace store, and using the traces to
+determine if a key needs rebuilding. Assuming a store of verifying traces
+\hs{VT}~\hs{k}~\hs{v}, the operations are:
 
-\begin{minted}{haskell}
-recordVT :: k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
-verifyVT :: (Monad m, Eq k, Eq v)
-         => k -> Hash v -> (k -> m (Hash v)) -> VT k v -> m Bool
+\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
+recordVT@\,@::@\,@k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
+verifyVT@\,@::@\,@(Monad m,@\,@Eq k,@\,@Eq v) => k -> Hash v -> (k -> m (Hash v)) -> VT k v -> m Bool
 \end{minted}
 
-We assume that \hs{Hash}~\hs{v} is a small constant size, constructed from hashing the
-underlying \hs{v} rather than storing it directly.
-After successfully building a key, we call \hs{recordVT} passing the key, the hash of its value, and the list of hashes and dependencies.
-Combined with the existing set of verifying traces, that provides a new set of verifying traces.
+\noindent
+We assume that \hs{Hash}~\hs{v} is a small constant size, constructed from
+hashing the underlying \hs{v} rather than storing it directly. After
+successfully building a key, we call \hs{recordVT} passing the key, the hash of
+its value, and the list of hashes and dependencies. Combined with the existing
+set of verifying traces, that provides a new set of verifying traces.
 
-More interestingly, to \emph{verify} whether a key needs rebuilding we supply the key and the hash of the current result in the store, along with a function for obtaining the post-build result of any key (using an ordering strategy
-as per \S\ref{sec-dependency-orderings}), and the existing \hs{VT} information.
-The result will be a \hs{Bool} where \hs{True} indicates that the trace could be verified, or \hs{False} to indicate the key should be rebuilt.
+More interestingly, to \emph{verify} whether a key needs rebuilding we supply
+the key, the hash of its current value, a function for obtaining the post-build
+value of any key (using a scheduling strategy as per
+\S\ref{sec-dependency-orderings}), and the existing \hs{VT} information. The
+result will be a \hs{Bool} where \hs{True} indicates that the current value is
+up to date, and \hs{False} indicates that it should be rebuilt.
 
-One potential implementation would be to record all arguments passed to \hs{recordVT}, and verify by simply checking if any argument matches the information passed by \hs{verifyVT} (we discuss smarter implementations in~\S\ref{sec-smart-traces}).
+One potential implementation would be to record all arguments passed to
+\hs{recordVT} in a list, and verify by simply checking if any list item matches
+the information passed by \hs{verifyVT} (we discuss smarter implementations
+in~\S\ref{sec-smart-traces}).
 
@@
 
-A verifying trace allows us to mark a key dirty and rebuild it.
-Extending that information we can store a \emph{constructive} trace which is the trace plus the actual result.
-Once we are storing the complete result it makes sense to record many constructive traces per key, and to share them with other users, providing cloud-build functionality.
-We can represent this additional information by providing the operations:
+A verifying trace allows us to mark a key dirty and rebuild it. Extending that
+information we can store a \emph{constructive} trace which is the trace plus the
+actual resulting value. Once we are storing the complete result it makes sense
+to record many constructive traces per key, and to share them with other users,
+providing cloud-build functionality. We can represent this additional
+information by providing the operations:
 
-\begin{minted}{haskell}
+\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
 recordCT    :: k -> v -> [(k, Hash v)] -> CT k v -> CT k v
-constructCT :: (Monad m, Eq k, Eq v)
-            => k -> (k -> m (Hash v)) -> CT k v -> m [v]
+constructCT :: (Monad m, Eq k, Eq v) => k -> (k -> m (Hash v)) -> CT k v -> m [v]
 \end{minted}
 
-The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing the hash of the result, we require the whole result.
-The \hs{verifyVT} has been replaced with \hs{constructCT}, which instead of taking the hash of the value, instead returns a list of suitable values.
-If the current value in the store matches one of the possible values, the build can skip this rule.
-If the resulting list is empty, the build must rerun.
-However, if the current value does not match the store, and there is a possible value, we can use any value from the constructive list \emph{without} doing any work to compute it.
+\noindent
+The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
+the hash of the resulting value, we require the whole value. The \hs{verifyVT}
+has been replaced with \hs{constructCT}, which instead of taking the hash of the
+value, returns a list of suitable values. If the current value in the store
+matches one of the possible values, the build can skip this rule. If the
+resulting list is empty, the key must be rebuilt. However, if the current value
+does not match the store, and there is a possible value, we can use any value
+from the constructive list \emph{without} doing any work to build it.
 
@@
 
-Constructive traces always verify keys by looking at their immediate dependencies, which must have first been brought up to date, meaning that the time to verify a key is based on the number of transitive dependencies.
-A \emph{deep} constructive trace optimises this process by only looking at the terminal input keys, ignoring any intermediate dependencies.
-The operations capturing this approach are the same as for constructive traces in~\S\ref{sec-constructive-traces}, but we use the names \hs{constructDCT} and \hs{recordDCT}, where the underlying \hs{DCT} state need only record information about hashes of inputs, not intermediate dependencies.
+Constructive traces always verify keys by looking at their immediate
+dependencies, which must have first been brought up to date, meaning that the
+time to verify a key is based on the number of transitive dependencies. A
+\emph{deep} constructive trace optimises this process by only looking at the
+terminal input keys, ignoring any intermediate dependencies. The operations
+capturing this approach are the same as for constructive traces
+in~\S\ref{sec-constructive-traces}, but we use the names \hs{recordDCT} and
+\hs{constructDCT}, where the underlying \hs{DCT} representation need only record
+information about hashes of inputs, not intermediate dependencies.
 
-The deep constructive trace approach only works if the operations are \emph{deterministic}, otherwise inconsistencies arise (see~\S\ref{sec-non-determinism}).
-Concretely, if the result of building a key given a set of immediate dependencies can vary, then it is possible that key \hs{k} will have been computed locally, but something \emph{using} \hs{k} will have been taken as a constructive input using a different value for \hs{k}.
-If another rule combines those two values then two different values of \hs{k} will have been used in one rule, leading to problems such as link-errors for typical compilers.
+The deep constructive trace approach only works if the operations are
+\emph{deterministic}, otherwise inconsistencies arise
+(see~\S\ref{sec-non-determinism}). Concretely, if the result of building a key
+given a set of immediate dependencies can vary, then it is possible that key
+\hs{k} will have been computed locally, but something \emph{using} \hs{k} will
+have been taken as a constructive input using a different value for \hs{k}. If
+another rule combines those two values then two different values of \hs{k} will
+have been used in one rule, leading to problems such as link-errors for typical
+compilers.
 
-While current build systems using deep constructive traces store only hashes of inputs, that is not a strict requirement, as the technique works equally well by skipping $n$ levels of dependency - where only inputs is the special case of $\infty$.
-Similarly, constructive traces can be described as taking the direct dependencies, and thus being deep constructive traces of $1$.
-A downside of deep constructive traces is that they can't support cutoff (\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies.
-On the otherhand, these traces are particularly useful for ``shallow builds'', as discussed in~\S\ref{sec-cloud-aspects}.
+While current build systems using deep constructive traces store only hashes of
+inputs, that is not a strict requirement, as the technique works equally well by
+skipping $n$ dependency levels, where the input-only approach is the special
+case of $n = \infty$. Similarly, constructive traces can be described as taking
+the direct dependencies, and thus having depth $n = 1$. A downside of deep
+constructive traces is that they cannot support cutoff
+(\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies. On
+the other hand, these traces are particularly useful for ``shallow builds'', as
+discussed in~\S\ref{sec-cloud-aspects}.
 
@@
 
-\begin{minted}{haskell}
-data Trace k h r = Trace
-    { key     :: k
-    , depends :: [(k, h)]
-    , result  :: r }
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Trace k v r = Trace { key :: k, depends :: [(k, Hash v)], result :: r }
 \end{minted}
 
-Where \hs{k} is the key type, \hs{h} is always \hs{Hash}~\hs{v} and \hs{r} is either \hs{Hash}~\hs{v} (verify traces) or \hs{v} (constructive traces).
-A real system is highly likey to use a more optimised implementation.
-Some of the most obvious optimisations are:
+\noindent
+Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
+\hs{v} (constructive traces). A real system is highly likely to use a more
+optimised implementation. Some of the most obvious optimisations are:
 
 \begin{enumerate}
-\item Any system using verifying traces, e.g.
-\Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}.
-Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
-\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
+\item Any system using verifying traces, e.g. \Shake, is unlikely to see
+significant benefit from storing more than one \hs{Trace} per key\footnote{There
+is a small chance of a benefit if the dependencies change but the result does
+not, and then the dependencies change back to what they were before.}.
+Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)},
+where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
+
+\item Any system using \hs{Applicative} dependencies can omit the dependency
+keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
+
 \item Any \hs{Applicative} build system storing constructive traces, e.g.
 \Bazel, can index directly from the key and results to the output result~--~i.e.
-\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}.
-Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result.
-\item Many cloud build systems store hashes of values in the trace information, then have a separate ``blob server'' which associates hashes with their actual contents.
+\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
+are stored on a central server, the client can compute the key and the hashes of
+its dependencies, then make a single call to the server to retrieve the result.
+
+\item Many cloud build systems store hashes of values in the trace information,
+then have a separate content-addressable cache which associates hashes with
+their actual contents.
 \end{enumerate}
@@
 
-\begin{table}[h]
+\begin{table}
+\vspace{-1mm}
 \smaller
@@
 \end{tabular}
-\vspace{2mm}
+\vspace{1mm}
 \caption{Build systems \`a la carte.\label{tab-build-systems}}
-\vspace{-2mm}
+\vspace{-8mm}
 \end{table}

### 2018-07-02 paper/5-implementations.tex
Minor revision of S5

@@
 
-A \hs{Rebuilder} is a function which the \hs{Scheduler} uses to rebuild a
-particular key, passing to it the key, the key's current value, and a \hs{Task}
-that can compute the value of the key if necessary. It uses the persistent build
-information \hs{ir} (carried by the state monad) to decide whether to rebuild
-the value. If doing so is unnecessary, it returns the current value; otherwise
-it runs the supplied \hs{Task} to rebuild it. In both cases it can choose to
-update the persistent build information \hs{ir} to reflect what happened. So a
-\hs{Rebuilder} wraps a \hs{Task}~\hs{c}~\hs{k}~\hs{v}, which unconditionally
-rebuilds the key, to make a \hs{Task}~\hs{(MonadState}~\hs{ir)}~\hs{k}~\hs{v},
-which rebuilds the key only if necessary, and does the necessary book-keeping.
-Note that the resulting \hs{Task} is always monadic; static dependency analysis
-can be performed on the original \hs{Task}~\hs{Applicative} if need be.
+A \hs{Rebuilder} takes three arguments: a key, its current value, and a
+\hs{Task} that can (re)compute the value of the key if necessary. It uses the
+persistent build information \hs{ir} (carried by the state monad) to decide
+whether to rebuild the value. If doing so is unnecessary, it returns the current
+value; otherwise it runs the supplied \hs{Task} to rebuild it. In both cases it
+can choose to update the persistent build information \hs{ir} to reflect what
+happened. So a \hs{Rebuilder} wraps a \hs{Task}~\hs{c}~\hs{k}~\hs{v}, which
+unconditionally rebuilds the key, to make a
+\hs{Task}~\hs{(MonadState}~\hs{ir)}~\hs{k}~\hs{v}, which rebuilds the key only
+if necessary, and does the necessary book-keeping. Note that the resulting
+\hs{Task} is always monadic; static dependency analysis can be performed on the
+original \hs{Task}~\hs{Applicative} if needed.
 
@@
 monad transformer \hs{ExceptT} from the \cmd{transformers} library. We also
-need the \hs{liftChain} helper function, whose implementation we omit since it
-is analogous to \hs{liftStore} defined in Fig.~\ref{fig-make-implementation}.
+need the helper \hs{liftChain}, whose implementation we omit since it
+is analogous to \hs{liftStore} in Fig.~\ref{fig-make-implementation}.
 

### 2018-07-02 src/Build/Task/MonadPlus.hs
Simplify computeND

@@ -13,8 +13,12 @@ random (low, high) = Task $ const $ foldr mplus mzero $ map pure [low..high]
 
 -- | Run a non-deterministic task with a pure lookup function, listing all
 -- possible results.
-computeND :: Task MonadPlus k v -> (k -> v) -> [v]
-computeND task store = run task (return . store)
+computePureND :: Task MonadPlus k v -> (k -> v) -> [v]
+computePureND task store = run task (return . store)
+
+-- | Run a task in a given store.
+computeND :: Task MonadPlus k v -> Store i k v -> [v]
+computeND task store = computePureND task (flip getValue store)
 
 -- | Given a description of @tasks@, an initial @store@, and a @result@ produced
 -- by running a build system on a target @key@, this function returns 'True' if
@@ -22,4 +26,4 @@ computeND task store = run task (return . store)
 correctBuildValue :: Eq v => Tasks MonadPlus k v -> Store i k v -> Store i k v -> k -> Bool
 correctBuildValue tasks store result k = case tasks k of
     Nothing   -> getValue k result == getValue k store
-    Just task -> getValue k result `elem` computeND task (flip getValue store)
+    Just task -> getValue k result `elem` computeND task store

### 2018-07-02 test/Examples.hs
Add sprsh5 example

@@ -80,6 +80,14 @@ sprsh2 "B2" = Just $ Task $ \fetch -> do
     if c1 == 1 then fetch "A1" else fetch "B1"
 sprsh2 _ = Nothing
 
+sprsh5 :: Tasks Monad String String
+sprsh5 "B1" = Just $ Task $ \fetch -> do
+    formula <- fetch "B1-formula"
+    evalFormula fetch formula
+  where
+    evalFormula = undefined
+sprsh5 _ = Nothing
+
 sprsh3 :: Tasks Alternative String Integer
 sprsh3 "B1" = Just $ Task $ \fetch -> (+) <$> fetch "A1" <*> (pure 1 <|> pure 2)
 sprsh3 _    = Nothing

### 2018-07-02 paper/6-engineering.tex
Work on S6

--- 2018-03-17 paper/6-engineering.tex
@@
 While it complicates the model, the complications can be restricted exclusively
-to the dependency strategy:
+to the scheduler:
 
 \begin{enumerate}
-\item The \hs{topological} function can build the full dependency graph, and
-whenever all dependencies of a task are complete, the task itself can be started.
+\item The \hs{topological} scheduler can build the full dependency graph, and
+whenever all dependencies of a task are complete, the task itself can be
+started.
 
-\item The \hs{reordering} function can be made parallel in a few ways, but the
+\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
 most direct is to have $n$ threads reading entries from the list of keys. As
@@
 
-\item The \hs{recursive} function can be made parallel by starting static
+\item The \hs{suspending} function can be made parallel by starting static
 dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
@@
 
-The actual implementation of the parallel strategies is not overly onerous,
+The actual implementation of the parallel schedulers is not overly onerous,
 but neither is it beautiful or informative.
@@
 
-In our model we define \hs{Task} as a function -- when given the same inputs it
-will always produce the same outputs. Alas, the real-world is not so obliging.
+In our model we define a \hs{Task} as a function -- when given the same inputs
+it will always produce the same output. Alas, the real-world is not so obliging.
 Some examples of impure tasks include:
@@
 
-Interestingly, there is significant interplay between all three sources of impurity.
-Systems like \Bazel use various sandboxing techniques to guard against missing
-dependencies, but none are likely to capture all dependencies right down to CPU
-model and microcode version. Rules that do have untracked state can be marked as
-volatile, a technique \Excel takes with the \cmd{INDIRECT} function, removing the
-untracked state at the cost of minimality.
+Interestingly, there is significant interplay between all three sources of
+impurity. Systems like \Bazel use various sandboxing techniques to guard against
+missing dependencies, but none are likely to capture all dependencies right down
+to CPU model and microcode version. Rules that do have untracked state can be
+marked as volatile, a technique \Excel takes with the \cmd{INDIRECT} function,
+removing the untracked state at the cost of minimality.
 
-Most of the implementations in \S\ref{sec-implementations} can deal with non-determinism,
-apart from \Bazel, which requires deterministic execution, and in turn can
-optimise the number of roundtrips required to the server.
+Most of the implementations in \S\ref{sec-implementations} can deal with
+non-determinism, apart from \Buck, which relies on deterministic tasks, and
+in turn can optimise the number of roundtrips required to the server.
 
-One tempting way of modelling non-determinism is to enrich \hs{Task} from
+One tempting way of modelling non-determinism is to enrich \hs{Tasks} from
 \hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus},
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh3 :: Task MonadPlus String Integer
-sprsh3 fetch "B1" = Just $ (+) <$> fetch "A1" <*> pure 1 `mplus` pure 2
-sprsh3 _     _    = Nothing
+sprsh3 :: Tasks MonadPlus String Integer
+sprsh3 "B1"@\,@=@\,@Just $ Task $ \fetch -> (+) <$> fetch "A1" <*> (pure 1 <|> pure 2)
+sprsh3 _   @\,@=@\,@Nothing
 \end{minted}
@@
 the correctness definition (\S\ref{sec-build-correctness}): instead of requiring
-that the result of recomputing the \hs{task} matches the value stored in the \hs{result}:
+that the result of recomputing the \hs{task} matches the value stored in the
+\hs{result}:
 \[
-\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}\]
+\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
+\]
 \noindent
@@
 \[
-\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}~\hs{k}
+\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}
 \]
 where
-\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{k}~\hs{->}~\hs{Maybe}~\hs{[@@v]}
+\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{[@@v]}
 returns the list of all possible results of the \hs{task} instead of just one
 value (`ND' stands for `non-deterministic').
-% (a possible implementation chooses
-% \hs{f}~\hs{=}~\hs{ListT}~\hs{Identity}~\hs{[@@v]}).
 
@@
 
-Our model of cloud builds provides a basic framework to discuss and reason about them, but lacks a number of important engineering corners:
+Our model of cloud builds provides a basic framework to discuss and reason
+about them, but lacks a number of important engineering corners:
 
@@
 minimum amount of information, optimising with respect to build-system specific
-invariants~--~see~\S\ref{sec-smart-traces} for some possible optimisations.
+invariants.
 
@@
 to the result of a shallow build. Then \hs{shallow} is correct, if \emph{there
-exists} \hs{result} which satisfies all requirements of
-Definition~\ref{def-correct}, \emph{such that} \hs{shallow} agrees with
-\hs{result} on the input keys \hs{k}~$\in$~$I$:
+exists} a \hs{result} which satisfies all requirements of the
+Definition~\ref{def-correct}, \emph{such that} \hs{shallow} agrees with the
+\hs{result} on all the input keys \hs{k}~$\in$~$I$:
 \[
@@
 the rule producing \cmd{B1} as also depending on the value \cmd{B1-formula}.
-The implementation can be given very directly in a \hs{Task}~\hs{Monad} as:
+The implementation can be given very directly in a \hs{Tasks}~\hs{Monad} as:
 
 \begin{minted}{haskell}
-task fetch "B1" = do
+sprsh5 "B1" = Just $ Task $ \fetch -> do
     formula <- fetch "B1-formula"
@@
 Namely, first look up the formula, then interpret it. It is not possible to
-change dependencies based on the formula in a \hs{Task}~\hs{Applicative}, as
+change dependencies based on the formula in a \hs{Tasks}~\hs{Applicative}, as
 would be required, so instead the formula can be captured as a dependency
-(but its value not used) and \emph{also} baked directly into the \hs{task}
+(but its value not used) and \emph{also} baked directly into the \hs{sprsh5}
 function.

### 2018-07-02 paper/7-related.tex
Work on S7
More tweaks to S7

@@
 validating traces of \Shake~--~our associated implementation provides such a
-combination. Ninja is also capable of modelling rules that produce multiple
+combination. \Ninja~is also capable of modelling rules that produce multiple
 results, a limited form of polymorphism \S\ref{sec-polymorphism}.
@@
 
-\item \Buck~\cite{buck} is very similar to \Bazel at the level of abstraction presented here.
+\item \CloudBuild~\cite{esfahani2016cloudbuild} uses constructive traces with a topological sort.
 
-\item \CloudBuild~\cite{esfahani2016cloudbuild}
-differs from \Bazel by allowing non-determinism~\S\ref{sec-non-determinism}, thus more closely modelling
-our original definition of \Bazel from \S\ref{sec-implementation-cloud}.
+\item \Buck~\cite{buck} uses deep constructive traces with a topological sort,
+meaning that it disallows non-determinism~\S\ref{sec-non-determinism}.
 
@@
 hashing of dependencies and downloading of precomputed build products. When
-combined with \cmd{import-from-derivation}, \Nix can also be considered monadic,
-making it similar to \Cloud \Shake
-from~\S\ref{sec-implementation-cloud}.
-However, \Nix is not intended as a build system, and the coarse grained nature
+combined with \cmd{import-from-derivation}, \Nix provides monadic dependencies.
+\Nix is not primarily intended as a build system, and the coarse grained nature
 (packages, not individual files) makes it targeted to a different purpose.

### 2018-07-02 paper/7-related.tex
S7 adjust fabricate
Work on S7

@@
 
-Most build systems, when viewed at the level we talk, can be captured with minor variations on the code presented in \S\ref{sec-implementations}. As some examples:
+Most build systems, when viewed at the level we talk, can be captured with minor
+variations on the code presented in \S\ref{sec-implementations}. As some
+examples:
 
 \begin{itemize}
-\item \Ninja~\cite{ninja} combines the dependency strategy of \Make with the
-validating traces of \Shake~--~our associated implementation provides such a
-combination. \Ninja~is also capable of modelling rules that produce multiple
-results, a limited form of polymorphism \S\ref{sec-polymorphism}.
+\item \Ninja~\cite{ninja} combines the \hs{topological} scheduler of \Make with
+the verifying traces of \Shake~--~our associated implementation provides such a
+combination. \Ninja~is also capable of modelling build rules that produce
+multiple results, a limited form of polymorphism \S\ref{sec-polymorphism}.
 
@@
 
-\item \CloudBuild~\cite{esfahani2016cloudbuild} uses constructive traces with a topological sort.
-
-\item \Buck~\cite{buck} uses deep constructive traces with a topological sort,
-meaning that it disallows non-determinism~\S\ref{sec-non-determinism}.
-
 \item \Nix~\cite{dolstra2004nix} has coarse-grained dependencies, with precise
-hashing of dependencies and downloading of precomputed build products. When
-combined with \cmd{import-from-derivation}, \Nix provides monadic dependencies.
-\Nix is not primarily intended as a build system, and the coarse grained nature
-(packages, not individual files) makes it targeted to a different purpose.
-% John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
-% \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.
+hashing of dependencies and downloading of precomputed build products. We
+provided a model of \Nix in \S\ref{sec-implementation-cloud}, although it is
+worth noting that \Nix is not primarily intended as a build system, and the
+coarse grained nature (packages, not individual files) makes it targeted to a
+different purpose.
 
@@
 The one build system we are aware of that cannot be modelled in our framework is
-\Fabricate~\cite{fabricate}. In \Fabricate a build system is a script which is
-run in-order, in the spirit of\footnote{\Fabricate requires scripts to be
-written in Python, but those details are not fundamental to what makes
-\Fabricate special.}:
+\Fabricate by \citet{fabricate}. In \Fabricate a build system is a script that is
+run in-order, in the spirit of:
+% I'm not sure the footnote adds much, let's save space?
+% \footnote{\Fabricate requires scripts to be
+% written in Python, but those details are not fundamental to what makes
+% \Fabricate special.}
 
@@
 
-\noindent
 To achieve minimality, each separate command is traced at the OS-level, allowing
@@
 commands like \cmd{gcc -c util.c} are keys, there is a linear dependency between
-each successive node, and that the OS-level tracing can be lifted back as a
-monadic \hs{Task} function\footnote{In fact, \Shake has an execution mode that
-can model \Fabricate{}-like build systems~--~see \hs{Development.Shake.Forward}
+each successive key, and that the OS-level tracing can be lifted back as a
+monadic \hs{Task} function\footnote{\Shake provides support for
+\Fabricate{}-like build systems~--~see \cmd{Development.Shake.Forward}
 in the \Shake library.}. However, in our pure model the mapping is not perfect
@@
 well studied area, and in particular the contrast between different formulations
-has been thoroughly investigated~\cite{acar2007selfadjusting}.
-
+has been thoroughly investigated, e.g. see~\cite{acar2007selfadjusting}.
 Self-adjusting computations can automatically adjust to an external change
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
+newtype Task c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
 \end{minted}

### 2018-07-02 paper/1-intro.tex
Small edits from simon

@@
 % approaches?''.  Moreover,
-Complex build systems use subtle algorithms, but they
-are often hidden away, and not the object of study.
+For years \Make{} dominated, but more recently
 Recently, the challenges of scale have driven large software firms
@@
 systems, exploring new points in the design space.
+These complex build systems use subtle algorithms, but they
+are often hidden away, and not the object of study.
 
@@
   persistent build information. In~\S\ref{sec-background} we identify some key
-  properties, in the context of four carefully-chosen build systems.
+  properties, illustrated by four carefully-chosen build systems.
 
@@
 In short, instead of seeing build systems as unrelated
-points in space, we now see them as locations in a landscape,
+points in space, we now see them as locations in a connected landscape,
 leading to a better understanding of what they do and how they compare,

### 2018-07-02 paper/2-background.tex
Small edits from simon

@@
 
-Unlike \Make, \Excel does not need to know all task dependencies upfront. Some
+Unlike \Make, \Excel does not need to know all task dependencies upfront. Indeed, some
 dependencies may change \emph{dynamically} according to computation results. For
@@
 
-To support dynamic dependencies, \Excel's build engine~\cite{excel_recalc} is
+To support dynamic dependencies, \Excel's calc engine~\cite{excel_recalc} is
 significantly different from \Make. \Excel arranges the cells into a linear
@@
 \Excel implements a conservative approximation to minimality: it recomputes a
-formula if~(i)~the formula mentions a changed cell, or~(ii)~the formula uses a
+formula if~(i)~the formula statically mentions a changed cell, or~(ii)~the formula uses a
 function like \cmd{INDIRECT} whose dependencies are not statically visible,
@@
 
-Item (iii) in the above list highlights a distinguishing feature of \Excel: it
+Item (iii) in the above list highlights another distinguishing feature of \Excel: it
 is \emph{self-tracking}. Most build systems only track changes of inputs and
@@
 
-\Bazel is one of the first examples of openly available cloud build systems.
+\Bazel is one of the first openly-available cloud build systems.
 As of writing, it is not possible to express dynamic dependencies in
-user-defined build rules, however some of the predefined build rules require
+user-defined build rules; however some of the pre-defined build rules require
 dynamic dependencies and the internal build engine can cope with them by using

### 2018-07-02 paper/3-abstractions.tex
Small edits from simon

@@
 \noindent
-Here \hs{Task} describes a single build task, while \hs{Tasks} associates a
-task description to every non-input key; input keys are associated with
+A \hs{Task} describes a single build task, while \hs{Tasks} associates a
+\hs{Task} to every non-input key; input keys are associated with
 \hs{Nothing}. The highly-abstracted type\footnote{Readers familiar with

### 2018-07-02 paper/4-build.tex
Add an extra line
Deep traces don't support early cutoff
Small edits from simon

@@
 
-The build systems overview (\S\ref{sec-background-summary}) highlighted three
-distinct task schedulers respecting the dependency order. This subsection
-explores their properties and possible implementations.
+Section \S\ref{sec-background} introduced three different \emph{task scehedulers}
+that decide which tasks to execute and in what order; see the ``Scheduler'' column
+of Table~\ref{tab-summary} in \S\ref{sec-background-summary}.
+This subsection explores the properties of the three schedulers, and
+possible implementations.
 
@@
 
-The topological scheduler pre-computes a linear order, which when followed,
+The topological scheduler pre-computes a linear order of tasks, which when followed,
 ensures the build result is correct regardless of the initial store. Given a
@@
 performed), the pre-processing is not. We can try to incrementalise the
-topological sort by storing the topological order between build runs and assume
-it to be correct, fixing it up if the build discovers any inconsistency, as
+topological sort by storing the topological order between build runs,
+using it as a starting point in the next run, 
+fixing it up if the build discovers any inconsistency, as
 \Excel does.
@@
 
-The second aspect, determining what to rebuild, can be addressed in one of four
+Suppose the scheduler decides that that a key \hs{k} should be
+brought up to date.  The next question is: does any work need to be done,
+or is it already up to date?   Or, in a cloud-build system, do we have
+a cached copy of the value we need?
+These questions can be addressed in one of four
 fundamental ways, with a number of tweaks and variations within them.
@@
 \noindent
-We assume that \hs{Hash}~\hs{v} is a small constant size, constructed from
-hashing the underlying \hs{v} rather than storing it directly. After
-successfully building a key, we call \hs{recordVT} passing the key, the hash of
-its value, and the list of hashes and dependencies. Combined with the existing
-set of verifying traces, that provides a new set of verifying traces.
+Rather than storing (large) values \hs{v}, the verifying trace \hs{VT}
+stores only hashes, of type \hs{Hash}~\hs{v}, of those values.  Since the
+verfying trace persists from one build to the next -- it constitutes the build system's ``memory''
+-- it is helpful for it to be of modest size.
+After successfully building a key, we call \hs{recordVT} to add a record
+to the current \hs{VT}, passing the key, the hash of
+its value, and the list of hashes and dependencies.
 
@@
 result will be a \hs{Bool} where \hs{True} indicates that the current value is
-up to date, and \hs{False} indicates that it should be rebuilt.
+already up to date, and \hs{False} indicates that it should be rebuilt.
 
@@
 \hs{recordVT} in a list, and verify by simply checking if any list item matches
-the information passed by \hs{verifyVT} (we discuss smarter implementations
-in~\S\ref{sec-smart-traces}).
+the information passed by \hs{verifyVT}.  We discuss smarter implementations
+in~\S\ref{sec-smart-traces}.
+
+A verifying trace, and other types of traces discussed in this section, support
+dynamic dependencies and minimality; furthermore, all traces except for deep
+traces~(\S\ref{sec-deep-constructive-traces}) support early cutoff.
 
@@
 
-A verifying trace allows us to mark a key dirty and rebuild it. Extending that
-information we can store a \emph{constructive} trace which is the trace plus the
-actual resulting value. Once we are storing the complete result it makes sense
+A verifying trace deliberately records only small hashes, so that it can be small.
+A \emph{constructive} trace additionally stores the resulting value.
+Once we are storing the complete result it makes sense
 to record many constructive traces per key, and to share them with other users,
@@
 has been replaced with \hs{constructCT}, which instead of taking the hash of the
-value, returns a list of suitable values. If the current value in the store
+current value as \emph{input}, returns a list of suitable values as \emph{output}.
+If the current value in the store
 matches one of the possible values, the build can skip this rule. If the
@@
 does not match the store, and there is a possible value, we can use any value
-from the constructive list \emph{without} doing any work to build it.
+from the constructive list \emph{without} doing any work to build it, and copy
+it into the store from the cloud.
 
@@
 \emph{deep} constructive trace optimises this process by only looking at the
-terminal input keys, ignoring any intermediate dependencies. The operations
+terminal \emph{input keys}, ignoring any intermediate dependencies. The operations
 capturing this approach are the same as for constructive traces
@@
 
+Current build systems using deep constructive traces always record hashes of
+terminal \emph{input keys}, but the technique works equally well
+if we skip any number dependency levels (say $n$ levels).
+The input-only approach is the special
+case of $n = \infty$, and constructive trances are the special case of 
+$n = 1$.
+
 The deep constructive trace approach only works if the operations are
 \emph{deterministic}, otherwise inconsistencies arise
-(see~\S\ref{sec-non-determinism}). Concretely, if the result of building a key
+(see~\S\ref{sec-non-determinism}). \simon{That section doesn't actually
+  mention ``inconsistencies''.
+  Also I think the inconsistencies are called Frankenbuilds, a term we might
+  want to introduce.}
+Concretely, if the result of building a key
 given a set of immediate dependencies can vary, then it is possible that key
@@
 
-While current build systems using deep constructive traces store only hashes of
-inputs, that is not a strict requirement, as the technique works equally well by
-skipping $n$ dependency levels, where the input-only approach is the special
-case of $n = \infty$. Similarly, constructive traces can be described as taking
-the direct dependencies, and thus having depth $n = 1$. A downside of deep
-constructive traces is that they cannot support cutoff
+A downside of deep constructive traces is that they cannot support cutoff
 (\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies. On
@@
 where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
+\simon{The data type decl had three args, k, v, r.}
 
 \item Any system using \hs{Applicative} dependencies can omit the dependency
-keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
+keys from the \hs{Trace} since they can be recovered from the \hs{key} field.
 
-\item Any \hs{Applicative} build system storing constructive traces, e.g.
+\item Any \hs{Applicative} build system using constructive traces, e.g.
 \Bazel, can index directly from the key and results to the output result~--~i.e.
@@
 \hline
-Property           & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
+ & \multicolumn{3}{c}{Scheduling strategy} \\
+Out-of-date key strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
 \hline
@@
 
-With the information in this section we can build a table comparing the
-dependency order strategy with the out-of-date keys strategy, providing 12~possible
-build systems, 8~of which are actually inhabited by existing build systems
+With the information in this section we can build Table~\ref{tab-build-systems}, which
+tabulates combinations of the schedling strategy and the out-of-date keys strategy,
+providing 12~possible build systems, 8~of which are actually inhabited by existing build systems
 (we discuss all these systems in \S\ref{sec-related-build}). Of the
@@
 unfilled spot in the table is suspending constructive traces, which would provide
-many benefits, and which we title \Cloud \Shake (as we plan on extending \Shake to occupy that spot)
+many benefits, and which we title \Cloud \Shake
+% (as we plan on extending \Shake to occupy that spot)
 and further explore further in~\S\ref{sec-implementation-cloud}.

### 2018-07-02 paper/5-implementations.tex
Small edits from simon

@@
 implementing a number of build systems as a composition of a scheduler
-and a rebuilder.  Our design is expressed by these types:
+and a rebuilder.  Our design is expressed by these types (Figure~\ref{fig-types})
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
+A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
 it to construct a \hs{Build} system, by choosing which keys to rebuild in which
@@
 \hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
-modification times}. The rebuilder receives three arguments: a
+modification times}.
+We assume that the external system, which invokes the build system, updates
+\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.
+
+The rebuilder receives three arguments: a
 \hs{key}, its current \hs{value}, and the \hs{task} of type
@@
 to make it explicit here.}; otherwise we can just return the current \hs{value}.
-We assume that the external system, which invokes the build system, updates
-\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.
 
@@
 
+The external system, which invokes \Excel's build engine, is required
+to provide a transitively closed set of dirty bits.  That is, if a
+cell is changed, its dirty bit is set, and the dirty bit of any cell
+whose value might perhaps change as a result.  It is OK to mark too many
+cells as dirty; but not OK to mark too few.
+
 The \hs{dirtyBitRebuilder} is very simple: if the \hs{key}'s dirty bit is set,
 we \hs{run} the \hs{task} to rebuild the \hs{key}; otherwise we return the
-current \hs{value} as is. The external system, which invokes \Excel's build
-engine, is required to provide a transitively closed set of dirty bits.
-Therefore, unlike \Make's \hs{modTimeRebuilder}, the \hs{dirtyBitRebuilder} does
+current \hs{value} as is.
+Because the dirty cells are transitively closed,
+unlike \Make's \hs{modTimeRebuilder}, the \hs{dirtyBitRebuilder} does
 not need to modify \hs{i} to trigger rebuilds of dependent keys.

### 2018-07-02 paper/6-engineering.tex
Small edits from simon

@@
 
-While we have given simple implementations assuming a single thread of execution,
-all the build systems we address can actually build independent keys in parallel.
+We have given simple implementations assuming a single thread of execution,
+but all the build systems we address can actually build independent keys in parallel.
 While it complicates the model, the complications can be restricted exclusively
@@
 \begin{description}
-\item[Untracked state] Some tasks depend on untracked state -- for example
+\item[Untracked dependencies] Some tasks depend on untracked values -- for example
 C compilation will explicitly list the \cmd{source.c} file as a dependency,
@@
 \end{description}
+\simon{Is there a difference between non-determinism and volatility?  Perhaps
+  one is accidental non-det while volatility is by-design non-det? But does the
+  distinction matter?   Maybe for accidental non-det you are free to use an
+  earlier cached result, whereas for volatity you should always recompute?
+  Volatility is best handled by a dependency on RealWorld, adn we coudl say that right here.}
 
@@
 missing dependencies, but none are likely to capture all dependencies right down
-to CPU model and microcode version. Rules that do have untracked state can be
+to CPU model and microcode version. Rules that do have untracked dependencies can be
 marked as volatile, a technique \Excel takes with the \cmd{INDIRECT} function,
-removing the untracked state at the cost of minimality.
+removing the untracked dependency at the cost of minimality.
 
@@
 the value calculated locally does not match what was previously calculated and
-stored in the cloud. Our implementations avoid this issue by storing complete
+stored in the cloud. \simon{Don't understand. The explanation in 4.2.4 is better -- and I think describes
+the same thing?}
+Our implementations avoid this issue by storing complete
 traces, but if a cloud build system was to only reference input nodes this
@@
 (but its value not used) and \emph{also} baked directly into the \hs{sprsh5}
-function.
+function. \simon{Don't understand.}
 
@@
 \emph{non-embedded domain specific language} to describe computations.
+\simon{I think you mean that you can find dependencies with executing the function,
+  which seems quite difference from ``backing it into'' the Haskell code.}
 Those which make use of a full programming language, e.g. \Shake, are faced with

### 2018-07-03 paper/1-intro.tex
Final tweaks

@@
 % approaches?''.  Moreover,
-For years \Make{} dominated, but more recently
-Recently, the challenges of scale have driven large software firms
-like Microsoft, Facebook, and Google to develop their own build
-systems, exploring new points in the design space.
-These complex build systems use subtle algorithms, but they
-are often hidden away, and not the object of study.
+For years \Make dominated, but more recently the challenges of scale have driven
+large software firms like Microsoft, Facebook, and Google to develop their own
+build systems, exploring new points in the design space. These complex build
+systems use subtle algorithms, but they are often hidden away, and not the
+object of study.
 
-In this paper we offer a general framework in which to understand
-and compare build systems,
-in a way that is both abstract (omitting incidental detail)
+In this paper we offer a general framework in which to understand and compare
+build systems, in a way that is both abstract (omitting incidental detail)
 and yet precise (implemented as Haskell code:

### 2018-07-03 paper/2-background.tex
Final tweaks

@@
 Self-tracking is uncommon in software build systems, where one often needs to
-manually initiate a full rebuild even if just a single build task has changed.
+manually initiate a full rebuild even if just a single task has changed.
 We discuss self-tracking further in~\S\ref{sec-tracking-aspects}.

### 2018-07-03 paper/4-build.tex
Final tweaks

@@
 actions themselves may be incremental (i.e. unnecessary tasks will not be
-performed), the pre-processing is not. We can try to incrementalise the
-topological sort by storing the topological order between build runs,
-using it as a starting point in the next run, 
-fixing it up if the build discovers any inconsistency, as
+performed), the pre-processing is not. We can incrementalise the topological
+sort by storing the topological order between build runs, using it as a starting
+point in the next run, fixing it up if the build discovers any inconsistency, as
 \Excel does.
@@
 \noindent
-Rather than storing (large) values \hs{v}, the verifying trace \hs{VT}
-stores only hashes, of type \hs{Hash}~\hs{v}, of those values.  Since the
-verfying trace persists from one build to the next -- it constitutes the build system's ``memory''
--- it is helpful for it to be of modest size.
-After successfully building a key, we call \hs{recordVT} to add a record
-to the current \hs{VT}, passing the key, the hash of
-its value, and the list of hashes and dependencies.
+Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} stores
+only hashes, of type \hs{Hash}~\hs{v}, of those values. Since the verifying
+trace persists from one build to the next -- it constitutes the build system's
+``memory'' -- it is helpful for it to be of modest size. After successfully
+building a key, we call \hs{recordVT} to add a record to the current \hs{VT},
+passing the key, the hash of its value, and the list of hashes and dependencies.
 
@@
 Current build systems using deep constructive traces always record hashes of
-terminal \emph{input keys}, but the technique works equally well
-if we skip any number dependency levels (say $n$ levels).
-The input-only approach is the special
-case of $n = \infty$, and constructive trances are the special case of 
+terminal \emph{input keys}, but the technique works equally well if we skip any
+number $n$ of dependency levels (say $n$ levels). The input-only approach is the
+special case of $n = \infty$, and constructive traces are the special case of
 $n = 1$.
@@
 
-With the information in this section we can build Table~\ref{tab-build-systems}, which
-tabulates combinations of the schedling strategy and the out-of-date keys strategy,
-providing 12~possible build systems, 8~of which are actually inhabited by existing build systems
-(we discuss all these systems in \S\ref{sec-related-build}). Of the
-remaining 4~spots, all result in workable build systems. The most interesting
-unfilled spot in the table is suspending constructive traces, which would provide
-many benefits, and which we title \Cloud \Shake
+With the information in this section we can build Table~\ref{tab-build-systems},
+which tabulates combinations of the scheduling strategy and the out-of-date keys
+strategy, providing 12~possible build systems, 8~of which are actually inhabited
+by existing build systems (we discuss all these systems in
+\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
+build systems. The most interesting unfilled spot in the table is suspending
+constructive traces, which would provide many benefits, and which we title
+\Cloud \Shake and further explore further in~\S\ref{sec-implementation-cloud}.
 % (as we plan on extending \Shake to occupy that spot)
-and further explore further in~\S\ref{sec-implementation-cloud}.

### 2018-07-03 paper/5-implementations.tex
Final tweaks

@@
 implementing a number of build systems as a composition of a scheduler
-and a rebuilder.  Our design is expressed by these types (Figure~\ref{fig-types})
+and a rebuilder.  Our design is expressed by these types (Fig.~\ref{fig-types}):
+
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
+
+\noindent
 A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
@@
 
-The rebuilder receives three arguments: a
-\hs{key}, its current \hs{value}, and the \hs{task} of type
-\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
-the \hs{key} if necessary.  The rebuilder first decides if the
-\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the \hs{key} is not
-found, that must mean it has never been built before; otherwise
-\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
-(computed by \hs{dependencies}) are out of date. If the \hs{key} is
+The rebuilder receives three arguments: a \hs{key}, its current \hs{value}, and
+the applicative \hs{task} that can be used to rebuild the \hs{key} if necessary.
+The rebuilder first decides if the \hs{key} is \hs{dirty} by consulting
+\hs{modTimes}: if the \hs{key} is not found, that must mean it has never been
+built before; otherwise \hs{modTimeRebuilder} can see if any of the \hs{task}'s
+dependencies (computed by \hs{dependencies}) are out of date. If the \hs{key} is
 \hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
@@
                 Left dep -> go done $ dep : filter (/= dep) ks ++ [key]
-                Right newValue -> do
-                    modify $ putValue key newValue
+                Right newValue -> do modify $ putValue key newValue
                                      (key :) <$> go (Set.insert key done) ks
@@
 
-The \hs{suspending} scheduler is based on a recursive function \hs{build} that
-builds a given \hs{key}, making sure not to duplicate work when called on the
-same \hs{key} again in future. To achieve that, it keeps track of keys that have
-already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}, running the
-computation in the
-\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
+The \hs{suspending} scheduler uses a recursive function \hs{build} that builds
+a given \hs{key}, making sure not to duplicate work when called on the same
+\hs{key} again in future. To achieve that, it keeps track of keys that have
+already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}.
+% , running the computation in the
+% \hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
 

### 2018-07-03 paper/7-related.tex
Final tweaks

@@
 \cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
-skipping any commands where no inputs have changed.
+skipping any commands where no inputs have changed. The key difference from our
+\hs{Tasks} abstraction is that instead of supplying a mapping from output to action,
+\Fabricate supplies a list of statements, in an order, without declaring what each line produces.
+There is no need to schedule the statements, and not enough information to do so.
 
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-editDistance :: Task Monad Key Integer
-editDistance _     (D i 0) = Just $ pure i
-editDistance _     (D 0 j) = Just $ pure j
-editDistance fetch (D i j) = Just $ do
+editDistance :: Tasks Monad Key Integer
+editDistance (D i 0) = Just $ Task $ const $ pure i
+editDistance (D 0 j) = Just $ Task $ const $ pure j
+editDistance (D i j) = Just $ Task $ \fetch -> do
     ai <- fetch (A i)
     bj <- fetch (B j)
-    if ai == bj
-        then fetch (D (i - 1) (j - 1))
-        else do
-            insert  <- fetch (D  i      (j - 1))
+    if ai == bj then fetch (D (i - 1) (j - 1))
+                else do insert  <- fetch (D  i      (j - 1))
                         delete  <- fetch (D (i - 1)  j     )
@@
                         return (1 + minimum [insert, delete, replace])
-editDistance _ _ = Nothing
+editDistance _ = Nothing
 \end{minted}

### 2018-07-03 paper/1-intro.tex
Address suggestions of reviewers

@@
 of a variety of different real-life build systems, including \Make, \Shake,
-\Bazel, \CloudBuild, \Buck, \Nix, and \Excel, each in a dozen lines of code
-or so (\S\ref{sec-implementations}). Doing this modelling in a single setting
+\Bazel, \CloudBuild, \Buck, \Nix, and \Excel\footnote{\Excel appears very
+different to the others but, seen through the lens of this paper, it is very
+close indeed.}, each in a dozen lines of code or so
+(\S\ref{sec-implementations}). Doing this modelling in a single setting
 allows the differences and similarities between these huge systems to be

### 2018-07-03 paper/4-build.tex
Wibbles

@@
 \vspace{1mm}
-We have bolded two different phrases above and, as we will see, the part of the
+We have emboldened two different aspects above: the part of the
 build system responsible for scheduling tasks in the dependency order
 (a `scheduler') can be cleanly separated from the part responsible for deciding
-whether a key needs to be rebuilt (a `rebuilder'). We therefore tackle each
+whether a key needs to be rebuilt (a `rebuilder'). We tackle each
 aspect separately in subsections~\S\ref{sec-dependency-orderings}
@@
 
-\subsection{Respecting the dependency order}
+\subsection{The scheduler: respecting the dependency order}
 \label{sec-dependency-orderings}
 
-Section \S\ref{sec-background} introduced three different \emph{task scehedulers}
+Section \S\ref{sec-background} introduced three different \emph{task schedulers}
 that decide which tasks to execute and in what order; see the ``Scheduler'' column
@@
 
-The topological scheduler has two downsides: it is limited to \hs{Applicative}
-build systems and requires a fresh topological sort each time. So, while the
-actions themselves may be incremental (i.e. unnecessary tasks will not be
-performed), the pre-processing is not. We can incrementalise the topological
-sort by storing the topological order between build runs, using it as a starting
-point in the next run, fixing it up if the build discovers any inconsistency, as
-\Excel does.
+To handle dynamic dependencies we can use the following speculative approach:
+build tasks in an arbitrary order, discovering their dependencies on the fly;
+whenever a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task,
+and switch to building the dependency \hs{dep}; eventually the previously
+aborted task is restarted and makes further progress thanks to \hs{dep} now
+being up to date. When a build completes, we can record the resulting order,
+using it as a starting point for the next build and fixing it up if a task's
+dependencies have changed, as \Excel does.
 
@@
 dependencies. It is also not minimal in the sense that a task may start, do some
-meaningful work, and then abort. However, in the case of an \hs{Applicative}
-system, that work is zero.
+meaningful work, and then abort.
 
-\Bazel's restarting scheduler does not store the topological order between build
+\Bazel's restarting scheduler does not store the discovered order between build
 runs; instead, it stores the most recent task dependency information. Since this
@@
 
-% An alternative approach to suspending a task is to abort
-% it and restart it again later, at the cost of doing additional work.
+\subsection{The rebuilder: determining out-of-date keys} \label{sec-out-of-date}
 
-\subsection{Determining out-of-date keys} \label{sec-out-of-date}
-
-Suppose the scheduler decides that that a key \hs{k} should be
-brought up to date.  The next question is: does any work need to be done,
-or is it already up to date?   Or, in a cloud-build system, do we have
-a cached copy of the value we need?
-These questions can be addressed in one of four
-fundamental ways, with a number of tweaks and variations within them.
+Suppose the scheduler decides that that a key should be brought up to date. The
+next question is: does any work need to be done, or is it already up to date?
+Or, in a cloud-build system, do we have a cached copy of the value we need?
+These questions can be addressed in one of four fundamental ways, with a number
+of tweaks and variations within them.
 
@@
 are set to clean. When the next build starts, anything that changed between the
-two states is marked dirty.
-% ; and by marking additional things dirty/clean the build system can track
-% what needs to rebuild.
-If a key and all its transitive dependencies are clean, the key does not need
-to be rebuild.
+two builds is marked dirty. If a key and all its transitive dependencies are
+clean, the key does not need to be rebuild.
 
@@
 also making sure all cells that depend on it are recursively marked dirty too.
-When rebuilding, if a clean cell only depends on clean cells it is skipped,
-otherwise it is rebuilt.
+To mark dependent cells dirty, it is necessary to statically know task
+dependencies. This is possible only for applicative tasks (see \S\ref{sec-deps}).
+For monadic tasks, \Excel computes a \emph{static approximation} of task
+dependencies: for functions such as \cmd{IF}, the approximation includes all
+cells that are mentioned in the formula (the \cmd{IF}'s condition and both
+branches), whereas functions such as \cmd{INDIRECT} whose dependencies cannot be
+guessed, are conservatively assumed to depend on all cells and are therefore
+always marked dirty.
 
-% AM: I didn't understand this bit, so commented it out.
-% The only wrinkle in this scheme is that \Excel supports
-% monadic tasks, and does not separately record the rebuilt, so it has
-% to approximate in this case.
-
-\Make uses file modification times, and compares files to their
-dependencies, which can be thought of as a dirty bit which is set when
-a file is newer than its dependencies. The interesting property of
-this dirty bit is that it is not under the control of \Make; rather it is
-existing file-system information that has been repurposed. In particular,
-modifying a file automatically clears its dirty bit, and
-automatically sets the dirty bit of the nodes depending on it. One
-thing \Make does require is that file timestamps only go forward in
-time -- something that can be violated by backup software.
-
-When using a dirty bit, it is necessary to check all the dependencies of a key.
-For applicative build systems that list is easy to obtain, but for monadic
-build systems there is no general way to get all dependencies. Instead \Excel
-computes a \emph{static approximation} of the dependencies. For applicative
-tasks that approximation is correct. For functions such as \cmd{IF} it marks the
-cell dirty if \emph{any} potential dependency has changed, even on the untaken
-\cmd{IF} branch. For functions such as \cmd{INDIRECT} whose dependencies cannot
-be guessed, it conservatively assumes the dependencies have always changed.
+\Make uses file modification times, and compares files to their dependencies,
+which can be thought of as a dirty bit which is set when a file is older than
+its dependencies. The interesting property of this dirty bit is that it is not
+under the control of \Make; rather it is existing file-system information that
+has been repurposed. Modifying a file automatically clears its dirty bit, and
+automatically sets the dirty bit of the files depending on it (but not
+recursively). Note that \Make requires that file timestamps only go forward in
+time, which can be violated by backup software.
 
 With a dirty bit it is simple to achieve minimality. However, to achieve early
-cutoff (\S\ref{sec-background-shake}) it would be important to not set the dirty
-bit after a computation that did not change the value. \Excel could use this
-approach, but does not. In contrast, \Make cannot implement early cutoff nicely
--- to do so it would have to mark the key clean (so it would not rebuild in the next
-run) and at the same time not mark the keys that depend on it dirty -- an
-impossible task with only the ability to update to the latest modification time.
-\Make can approximate early cutoff by not modifying the result file, and not
-marking it clean, but then it will rerun in every subsequent build.
+cutoff (\S\ref{sec-background-shake}) it would be important to reset the dirty
+bit after a computation that did not change the value and make sure that cells
+that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
+because the dependent cells have already been recursively marked dirty. For
+\Make, it is impossible to mark a file clean and at the same time not mark the
+keys that depend on it dirty. \Make can approximate early cutoff by not
+modifying the result file, and not marking it clean, but then it will be rebuilt
+in every subsequent build.
 
@@
 terminal \emph{input keys}, but the technique works equally well if we skip any
-number $n$ of dependency levels (say $n$ levels). The input-only approach is the
+number of dependency levels (say $n$ levels). The input-only approach is the
 special case of $n = \infty$, and constructive traces are the special case of
@@
 The deep constructive trace approach only works if the operations are
-\emph{deterministic}, otherwise inconsistencies arise
-(see~\S\ref{sec-non-determinism}). \simon{That section doesn't actually
-  mention ``inconsistencies''.
-  Also I think the inconsistencies are called Frankenbuilds, a term we might
-  want to introduce.}
-Concretely, if the result of building a key
-given a set of immediate dependencies can vary, then it is possible that key
-\hs{k} will have been computed locally, but something \emph{using} \hs{k} will
-have been taken as a constructive input using a different value for \hs{k}. If
-another rule combines those two values then two different values of \hs{k} will
-have been used in one rule, leading to problems such as link-errors for typical
-compilers.
+\emph{deterministic}, otherwise it is possible to violate correctness resulting
+in a so-called ``Frankenbuild''~\cite{esfahani2016cloudbuild}. Concretely, if
+the result of building a key given a set of immediate dependencies can vary,
+then it is possible that key \hs{k} will have been computed locally, but
+something \emph{using} \hs{k} will have been taken as a constructive input using
+a different value for \hs{k}. If another rule combines those two values then two
+different values of \hs{k} will have been used in one rule, leading to problems
+such as link-errors for typical compilers.
 
@@
 not, and then the dependencies change back to what they were before.}.
-Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)},
+Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash v))},
 where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
-\simon{The data type decl had three args, k, v, r.}
 

### 2018-07-03 paper/4-build.tex
Comments from Simon

@@
 
-Section \S\ref{sec-background} introduced three different \emph{task scehedulers}
+Section \S\ref{sec-background} introduced three different \emph{task schedulers}
 that decide which tasks to execute and in what order; see the ``Scheduler'' column
@@
 The topological scheduler has two downsides: it is limited to \hs{Applicative}
-build systems and requires a fresh topological sort each time. So, while the
+build systems and requires a fresh topological sort each time. \simon{If the tasks don't change,
+  the dependencies are unchanged, so the topological sort from the last time
+  will do fine, won't it?}  So, while the
 actions themselves may be incremental (i.e. unnecessary tasks will not be
@@
 also making sure all cells that depend on it are recursively marked dirty too.
+\simon{Not always (see make) and only possible if we have static dependencies
+  or have recorded dynamic dependencies.}
 When rebuilding, if a clean cell only depends on clean cells it is skipped,
-otherwise it is rebuilt.
+otherwise it is rebuilt. \simon{But if all the dependencies of a dirty cell
+  are marked dirty, we rebuild only dirty cells not clean ones, contradicting this statement.}
 
@@
 thing \Make does require is that file timestamps only go forward in
-time -- something that can be violated by backup software.
+time -- something that can be violated by backup software. \simon{Also transitive dependencies
+  of dirty files are not made dirty.}
 
@@
 cutoff (\S\ref{sec-background-shake}) it would be important to not set the dirty
-bit after a computation that did not change the value. \Excel could use this
-approach, but does not. In contrast, \Make cannot implement early cutoff nicely
+bit after a computation that did not change the value. \simon{And if we'd recursively
+made the transitive dependencies dirty, we could not do early cutoff.}
+\Excel could use this
+approach, but does not. \simon{It'd be tricky becase of setting dirtiness recursively.}
+In contrast, \Make cannot implement early cutoff nicely
 -- to do so it would have to mark the key clean (so it would not rebuild in the next

### 2018-07-03 paper/5-implementations.tex
Comments from Simon

@@
 original \hs{Task}~\hs{Applicative} if needed.
+
+The scheduler calls the rebuilder, but passes to it a \hs{fetch} function
+that that latter calls when it needs the value of a dependent key.  This
+callback returns control to the scheduler, which may in turn call the
+rebuilder to bring the dependent key up to date, and so on.
 
@@
 
+\simon{topological is hard to read because of the execState and forM stuff.
+  Could we have a local defn, with a type signature, for the $\lambda key. blah$?
+  If necessary we could omit the code for liftStore, leaving only its signature}
 \Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based on a
@@
 \hs{k}~\hs{->}~\hs{Bool} associating a dirty bit with every key, and (ii) a
-calc \hs{Chain}~\hs{::}~\hs{[@@k]} recorded in the previous build
+calc chain of type \hs{[@@k]} recorded in the previous build
 (\S\ref{sec-background-excel}).
@@
 
+\simon{New subsection here? Or at least put this para after the next one.}
 Using the abstractions built thus far, we have shown how to combine schedulers

### 2018-07-04 paper/4-build.tex
Revise S4.1.2 Restarting

@@
 
-The topological scheduler has two downsides: it is limited to \hs{Applicative}
-build systems and requires a fresh topological sort each time. \simon{If the tasks don't change,
-  the dependencies are unchanged, so the topological sort from the last time
-  will do fine, won't it?}  So, while the
-actions themselves may be incremental (i.e. unnecessary tasks will not be
-performed), the pre-processing is not. We can incrementalise the topological
-sort by storing the topological order between build runs, using it as a starting
-point in the next run, fixing it up if the build discovers any inconsistency, as
-\Excel does.
+To handle dynamic dependencies we can use the following speculative approach:
+build tasks in an arbitrary order, discovering their dependencies on the fly;
+whenever a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task,
+and switch to building the dependency \hs{dep}; eventually the previously
+aborted task is restarted and makes further progress thanks to \hs{dep} now
+being up to date. When a build completes, we can record the resulting order,
+using it as a starting point for the next build and fixing it up if a task's
+dependencies have changed, as \Excel does.
 

### 2018-07-05 paper/4-build.tex
Revise S4.2.1: A dirty bit

@@
 
-% An alternative approach to suspending a task is to abort
-% it and restart it again later, at the cost of doing additional work.
-
 \subsection{Determining out-of-date keys} \label{sec-out-of-date}
 
-Suppose the scheduler decides that that a key \hs{k} should be
-brought up to date.  The next question is: does any work need to be done,
-or is it already up to date?   Or, in a cloud-build system, do we have
-a cached copy of the value we need?
-These questions can be addressed in one of four
-fundamental ways, with a number of tweaks and variations within them.
+Suppose the scheduler decides that that a key should be brought up to date. The
+next question is: does any work need to be done, or is it already up to date?
+Or, in a cloud-build system, do we have a cached copy of the value we need?
+These questions can be addressed in one of four fundamental ways, with a number
+of tweaks and variations within them.
 
@@
 also making sure all cells that depend on it are recursively marked dirty too.
-\simon{Not always (see make) and only possible if we have static dependencies
-  or have recorded dynamic dependencies.}
-When rebuilding, if a clean cell only depends on clean cells it is skipped,
-otherwise it is rebuilt. \simon{But if all the dependencies of a dirty cell
-  are marked dirty, we rebuild only dirty cells not clean ones, contradicting this statement.}
+To mark dependent cells dirty, it is necessary to statically know task
+dependencies. This is possible only for applicative tasks (see \S\ref{sec-deps}).
+For monadic tasks, \Excel computes a \emph{static approximation} of task
+dependencies: for functions such as \cmd{IF}, the approximation includes all
+cells that are mentioned in the formula (the \cmd{IF}'s condition and both
+branches), whereas functions such as \cmd{INDIRECT} whose dependencies cannot be
+guessed, are conservatively assumed to depend on all cells and are therefore
+always marked dirty.
 
-% AM: I didn't understand this bit, so commented it out.
-% The only wrinkle in this scheme is that \Excel supports
-% monadic tasks, and does not separately record the rebuilt, so it has
-% to approximate in this case.
-
-\Make uses file modification times, and compares files to their
-dependencies, which can be thought of as a dirty bit which is set when
-a file is newer than its dependencies. The interesting property of
-this dirty bit is that it is not under the control of \Make; rather it is
-existing file-system information that has been repurposed. In particular,
-modifying a file automatically clears its dirty bit, and
-automatically sets the dirty bit of the nodes depending on it. One
-thing \Make does require is that file timestamps only go forward in
-time -- something that can be violated by backup software. \simon{Also transitive dependencies
-  of dirty files are not made dirty.}
-
-When using a dirty bit, it is necessary to check all the dependencies of a key.
-For applicative build systems that list is easy to obtain, but for monadic
-build systems there is no general way to get all dependencies. Instead \Excel
-computes a \emph{static approximation} of the dependencies. For applicative
-tasks that approximation is correct. For functions such as \cmd{IF} it marks the
-cell dirty if \emph{any} potential dependency has changed, even on the untaken
-\cmd{IF} branch. For functions such as \cmd{INDIRECT} whose dependencies cannot
-be guessed, it conservatively assumes the dependencies have always changed.
+\Make uses file modification times, and compares files to their dependencies,
+which can be thought of as a dirty bit which is set when a file is older than
+its dependencies. The interesting property of this dirty bit is that it is not
+under the control of \Make; rather it is existing file-system information that
+has been repurposed. Modifying a file automatically clears its dirty bit, and
+automatically sets the dirty bit of the files depending on it (but not
+recursively). Note that \Make requires that file timestamps only go forward in
+time, which can be violated by backup software.
 
 With a dirty bit it is simple to achieve minimality. However, to achieve early
-cutoff (\S\ref{sec-background-shake}) it would be important to not set the dirty
-bit after a computation that did not change the value. \simon{And if we'd recursively
-made the transitive dependencies dirty, we could not do early cutoff.}
-\Excel could use this
-approach, but does not. \simon{It'd be tricky becase of setting dirtiness recursively.}
-In contrast, \Make cannot implement early cutoff nicely
--- to do so it would have to mark the key clean (so it would not rebuild in the next
-run) and at the same time not mark the keys that depend on it dirty -- an
-impossible task with only the ability to update to the latest modification time.
-\Make can approximate early cutoff by not modifying the result file, and not
-marking it clean, but then it will rerun in every subsequent build.
+cutoff (\S\ref{sec-background-shake}) it would be important to reset the dirty
+bit after a computation that did not change the value and make sure that cells
+that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
+because the dependent cells have already been recursively marked dirty. For
+\Make, it is impossible to mark a file clean and at the same time not mark the
+keys that depend on it dirty. \Make can approximate early cutoff by not
+modifying the result file, and not marking it clean, but then it will be rebuilt
+in every subsequent build.
 

### 2018-07-05 src/Build/Scheduler.hs
Make build function explicit in the topological scheduler

@@ -61,22 +61,24 @@ updateValue key _value newValue = putValue key newValue
 -- extracting all (static) dependencies upfront, and then traversing the graph
 -- in the topological order, rebuilding keys using the supplied rebuilder.
 topological :: forall i k v. Ord k => Scheduler Applicative i i k v
-topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
-    Nothing -> return ()
-    Just task -> do
-        store <- get
-        let value = getValue key store
-            newTask :: Task (MonadState i) k v
-            newTask = rebuilder key value task
-            fetch :: k -> State i v
-            fetch k = return (getValue k store)
-        newValue <- liftStore (run newTask fetch)
-        modify $ updateValue key value newValue
+topological rebuilder tasks target = execState $ mapM_ build order
   where
-    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
-    order  = case topSort (graph deps target) of
+    build :: k -> State (Store i k v) ()
+    build key = case tasks key of
+        Nothing -> return ()
+        Just task -> do
+            store <- get
+            let value = getValue key store
+                newTask :: Task (MonadState i) k v
+                newTask = rebuilder key value task
+                fetch :: k -> State i v
+                fetch k = return (getValue k store)
+            newValue <- liftStore (run newTask fetch)
+            modify $ updateValue key value newValue
+    order = case topSort (graph deps target) of
         Nothing -> error "Cannot build tasks with cyclic dependencies"
         Just xs -> xs
+    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
 
 ---------------------------------- Restarting ----------------------------------
 -- | Convert a task with a total lookup function @k -> m v@ into a task

### 2018-07-05 paper/5-implementations.tex
Clarify topological implementation

@@
 topological :: Ord k => Scheduler Applicative i i k v
-topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
+topological rebuilder tasks target = execState $ mapM_ build order
+  where
+    build :: k -> State (Store i k v) ()
+    build key = case tasks key of
         Nothing -> return ()
@@
             store <- get
-        let newTask :: Task (MonadState i) k v
-            newTask = rebuilder key (getValue key store) task
+            let value = getValue key store
+                newTask :: Task (MonadState i) k v
+                newTask = rebuilder key value task
                 fetch :: k -> State i v
@@
             newValue <- liftStore (run newTask fetch)
-        modify $ putValue key newValue
-  where
-    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
-    order  = fromJust $ topSort (reachable deps target) -- Assumes acyclic dependencies
+            modify $ updateValue key value newValue
+    order = case topSort (graph deps target) of
+        Nothing -> error "Cannot build tasks with cyclic dependencies"
+        Just xs -> xs
+    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
 \end{minted}
@@
 
-\simon{topological is hard to read because of the execState and forM stuff.
-  Could we have a local defn, with a type signature, for the $\lambda key. blah$?
-  If necessary we could omit the code for liftStore, leaving only its signature}
-\Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based on a
-topological sort of the statically known dependency graph
+\Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based
+on a topological sort of the statically known dependency graph
 (see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
@@
 
-The \hs{where} clause corresponds to the pre-processing stage, which uses the
-function \hs{dependencies}, defined in~\S\ref{sec-deps}, to extract static
-dependencies from the provided applicative \hs{task}. We compute the linear
-processing \hs{order} by constructing the graph of keys \hs{reachable} from the
-\hs{target} via dependencies, and performing the topological sort of the result.
-We omit implementation of textbook graph algorithms
-\hs{reachable} and \hs{topSort}, e.g. see~\cite{cormen2001introduction}.
+The pre-processing stage uses the function \hs{dependencies}, defined
+in~\S\ref{sec-deps}, to extract static dependencies from the provided
+applicative \hs{task}. We compute the linear processing \hs{order} by
+constructing the graph of keys \hs{reachable} from the \hs{target} via
+dependencies, and performing the topological sort of the result. We omit
+implementation of textbook graph algorithms \hs{reachable} and \hs{topSort},
+e.g. see~\cite{cormen2001introduction}.
 

### 2018-07-05 paper/5-implementations.tex
Minor revision

@@
 
-\simon{New subsection here? Or at least put this para after the next one.}
-Using the abstractions built thus far, we have shown how to combine schedulers
-with rebuilders to reproduce existing build systems. To us, the most interesting
-build system as yet unavailable would combine a suspending scheduler with
-constructive traces~--~providing a cloud-capable build system that is minimal,
-and supports both early cutoff and monadic dependencies. Using our framework it
-is possible to define and test such a system, which we call \Cloud \Shake. All
-we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
-of \Bazel, as shown in Fig.~\ref{fig-cloud-implementations}.
-
 Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the
@@
 
-% Some build systems allow to exploit \emph{build task determinism} for efficient
-% fetching of build results from the cache: one can simply hash the values of all
-% transitive inputs of a key (which can be done efficiently by storing all hashes
-% in a Merkle tree) and download the matching value from the cache, without the
-% need to build any intermediate values. This can bring significant performance
-% benefits in practice, but sacrifices the early cutoff property: indeed, with
-% this approach addition of a new comment to one of the source files will
-% invalidate all dependent keys.
+Using the abstractions built thus far, we have shown how to combine schedulers
+with rebuilders to reproduce existing build systems. To us, the most interesting
+build system as yet unavailable would combine a suspending scheduler with
+constructive traces~--~providing a cloud-capable build system that is minimal,
+and supports both early cutoff and monadic dependencies. Using our framework it
+is possible to define and test such a system, which we call \Cloud \Shake. All
+we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
+of \Bazel, as shown in Fig.~\ref{fig-cloud-implementations}.

### 2018-07-05 paper/6-engineering.tex
Final tweaks
Minor revision

@@
 for errors, e.g. early aborting.
-
-% Q: {Isn't ``early aborting'' automatically handled by short-circuiting
-% withing monads, i.e. \hs{Nothing} or \hs{Left e} cancel out all subsequent
-% computation?}
-% NM says: The build system could chose an early abort monad, but the tasks are
-% free in the monad, so you can't get the encoding to cross between them.
 
@@
 Similarly, build systems like \Make and \Shake provide \emph{phony rules} which
-are also volatile.
+are also volatile. The main difference from non-deterministic tasks, is that
+volatile tasks cannot be cached. They are best handled by introducing a special
+key \cmd{RealWorld} whose value is changed between every run.
 \end{description}
-\simon{Is there a difference between non-determinism and volatility?  Perhaps
-  one is accidental non-det while volatility is by-design non-det? But does the
-  distinction matter?   Maybe for accidental non-det you are free to use an
-  earlier cached result, whereas for volatity you should always recompute?
-  Volatility is best handled by a dependency on RealWorld, adn we coudl say that right here.}
 
@@
 volatile~--~an approach that can be readily adopted by any of our
-implementations by introducing a special key \cmd{RealWorld} whose value is
-changed between every run.
+implementations.
 
@@
 
-\item[Frankenbuilds] A build is considered a
-\emph{frankenbuild}~\cite{esfahani2016cloudbuild} if a value is calculated
-locally, but something that depends on that key is pulled from the cache, and
-the value calculated locally does not match what was previously calculated and
-stored in the cloud. \simon{Don't understand. The explanation in 4.2.4 is better -- and I think describes
-the same thing?}
-Our implementations avoid this issue by storing complete
-traces, but if a cloud build system was to only reference input nodes this
-situation can arise.
+% \item[Frankenbuilds] A build is considered a
+% \emph{frankenbuild}~\cite{esfahani2016cloudbuild} if a value is calculated
+% locally, but something that depends on that key is pulled from the cache, and
+% the value calculated locally does not match what was previously calculated and
+% stored in the cloud. \simon{Don't understand. The explanation in 4.2.4 is better -- and I think describes
+% the same thing?}
+% Our implementations avoid this issue by storing complete
+% traces, but if a cloud build system was to only reference input nodes this
+% situation can arise.
 
@@
 the inputs of \cmd{A1} or \cmd{A2}, \emph{or} editing the formula in
-\cmd{B1}~--~e.g. to \cmd{A1 - A2}. This pattern can be captured by describing
+\cmd{B1}~--~e.g. to \cmd{A1 * A2}. This pattern can be captured by describing
 the rule producing \cmd{B1} as also depending on the value \cmd{B1-formula}.
-The implementation can be given very directly in a \hs{Tasks}~\hs{Monad} as:
+The implementation can be given very directly in a
+\hs{Tasks}~\hs{Monad}~--~concretely, first look up the formula, then interpret
+it:
 
@@
 
+% It is not possible to
+% change dependencies based on the formula in a \hs{Tasks}~\hs{Applicative}, as
+% would be required, so instead the formula can be captured as a dependency
+% (but its value not used) and \emph{also} baked directly into the \hs{sprsh5}
+% function. \simon{Don't understand.}
+% \simon{I think you mean that you can find dependencies with executing the function,
+%   which seems quite difference from ``backing it into'' the Haskell code.}
+
 \noindent
-Namely, first look up the formula, then interpret it. It is not possible to
-change dependencies based on the formula in a \hs{Tasks}~\hs{Applicative}, as
-would be required, so instead the formula can be captured as a dependency
-(but its value not used) and \emph{also} baked directly into the \hs{sprsh5}
-function. \simon{Don't understand.}
-
 The build systems that have precise self-tracking are all ones which use a
 \emph{non-embedded domain specific language} to describe computations.
-\simon{I think you mean that you can find dependencies with executing the function,
-  which seems quite difference from ``backing it into'' the Haskell code.}
 Those which make use of a full programming language, e.g. \Shake, are faced with
@@
 
-\noindent
 In such cases \Excel will normally not execute anything, but if the

### 2018-07-05 paper/7-related.tex
Address suggestions of reviewers
Add Dune

@@
 abstraction with additional features.
+
+\item \Dune~\cite{dune} is a build system designed for OCaml/Reason projects.
+Its distinguishing feature is that it uses
+\emph{arrows}~\cite{hughes2000generalising} rather than monads to model
+dynamic dependencies, which simplifies static dependency approximation.
 \end{itemize}
@@
 \noindent
-provided we instantiate \hs{p} to something like
+provided we instantiate \hs{p}~\hs{a}~\hs{b} to something like
 \hs{k}~\hs{->}~\hs{f}~\hs{v}~--~which many of the actual instances in that paper

### 2018-07-05 paper/6-engineering.tex
Add notes on Reflow and Incremental

@@
 communication can become a bottleneck, so it is important to send only the
-minimum amount of information, optimising with respect to build-system specific
-invariants.
+minimum amount of information, optimising with respect to build system
+invariants. For example, incremental data processing systems in the cloud, such
+as \Reflow~\cite{reflow}, need to orchestrate terabytes of data.
 

### 2018-07-05 paper/7-related.tex
Add notes on Reflow and Incremental

@@
 A lot of research has been dedicated to finding efficient data structures and
-algorithms for self-adjusting computations~--~we plan to investigate how these
-insights can be utilised by build systems as future work.
+algorithms for self-adjusting computations (with a few open-source
+implementations, e.g. \Incremental~\cite{incremental}). We plan to investigate
+how these insights can be utilised by build systems as future work.
 

### 2018-07-05 paper/6-engineering.tex
Minor fixes

--- 2018-07-05 paper/6-engineering.tex
@@
 
-\item[Communication] When traces or contents are stored on a central server
+\item[Communication] When traces or contents are stored in the cloud
 communication can become a bottleneck, so it is important to send only the
-minimum amount of information, optimising with respect to build-system specific
-invariants.
+minimum amount of information, optimising with respect to build system
+invariants. For example, incremental data processing systems in the cloud, such
+as \Reflow~\cite{reflow}, need to efficiently orchestrate terabytes of data.
 

### 2018-07-05 paper/3-abstractions.tex
Address suggestions of reviewers
Polishing (thanks, Andrew!)

@@
 \noindent
+% Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative} or \hs{Monad}.
 A \hs{Task} describes a single build task, while \hs{Tasks} associates a
 \hs{Task} to every non-input key; input keys are associated with
-\hs{Nothing}. The highly-abstracted type\footnote{Readers familiar with
-\emph{lenses} or \emph{profunctor optics} might recognise a familiar pattern.
-We discuss this in~\S\ref{sec-related-optics}.} \hs{Task} is best introduced by
-an example. Consider this \Excel spreadsheet:
+\hs{Nothing}. The highly-abstracted type\footnote{Readers
+familiar with \emph{lenses} or \emph{profunctor optics} might recognise a
+familiar pattern. We discuss this in~\S\ref{sec-related-optics}.} \hs{Task} is
+best introduced by an example. Consider this \Excel spreadsheet:
 
@@
 formula \cmd{A1+A2}, etc.
-
-% \newpage
-
-\noindent
-We can represent the formulae of this spreadsheet with the following task
-description:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-sprsh1 :: Tasks Applicative String Integer
-sprsh1 "B1" = Just $ Task $ \fetch -> ((+)  <$> fetch "A1" <*> fetch "A2")
-sprsh1 "B2" = Just $ Task $ \fetch -> ((*2) <$> fetch "B1")
-sprsh1 _    = Nothing
-\end{minted}
-\vspace{1mm}
-
-\noindent
-We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
-(Real spreadsheet cells would contain a wider range of values, of course.) The
-task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
-but not the input values. It pattern-matches on the key to see if it has a task
-description (in the \Excel case, a formula) for it. If not, it returns
-\hs{Nothing}, indicating that the key is an input. If there is a formula in the
-cell, it returns the \hs{Task} to compute the value of the formula. Every
-\hs{Task} is given a \emph{callback} \hs{fetch} to find the value of any keys on
-which it depends.
 
@@
 instance Monoid m => Applicative (Const m) where
-    pure _              = Const mempty
-    Const x <*> Const y = Const (x <> y)
+    pure _              = Const mempty   -- mempty is the identity of the monoid m
+    Const x <*> Const y = Const (x <> y) -- <> is the binary operation of the monoid m
 \end{minted}
@@
 \end{figure}
+
+\noindent
+We can represent the formulae of this spreadsheet with the following task
+description:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+sprsh1 :: Tasks Applicative String Integer
+sprsh1 "B1" = Just $ Task $ \fetch -> ((+)  <$> fetch "A1" <*> fetch "A2")
+sprsh1 "B2" = Just $ Task $ \fetch -> ((*2) <$> fetch "B1")
+sprsh1 _    = Nothing
+\end{minted}
+\vspace{1mm}
+
+\noindent
+We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
+(Real spreadsheet cells would contain a wider range of values, of course.) The
+task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
+but not the input values. It pattern-matches on the key to see if it has a task
+description (in the \Excel case, a formula) for it. If not, it returns
+\hs{Nothing}, indicating that the key is an input. If there is a formula in the
+cell, it returns the \hs{Task} to compute the value of the formula. Every
+\hs{Task} is given a \emph{callback} \hs{fetch} to find the value of any keys on
+which it depends.
 
@@
 running its code in a different \hs{Applicative}! This is not new, for example
-see~\cite{free-applicatives}, but it is extremely cool. We will see a practical
+see~\citet{free-applicatives}, but it is extremely cool. We will see a practical
 use for \hs{dependencies} when implementing applicative build systems,

### 2018-07-05 paper/7-related.tex
Polishing (thanks, Andrew!)

@@
 
+\noindent
 To achieve minimality, each separate command is traced at the OS-level, allowing
@@
 \noindent
-provided we instantiate \hs{p}~\hs{a}~\hs{b} to something like
+Provided we instantiate \hs{p}~\hs{a}~\hs{b} to something like
 \hs{k}~\hs{->}~\hs{f}~\hs{v}~--~which many of the actual instances in that paper

### 2018-07-05 paper/4-build.tex
More from Simon

@@
 
-\subsection{Respecting the dependency order}
+\subsection{The scheduler: respecting the dependency order}
 \label{sec-dependency-orderings}
@@
 
-To handle dynamic dependencies we can use the following speculative approach:
-build tasks in an arbitrary order, discovering their dependencies on the fly;
+To handle dynamic dependencies we can use the following approach:
+build tasks in an arbitrary initial order, discovering their dependencies on the fly;
 whenever a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task,
@@
 aborted task is restarted and makes further progress thanks to \hs{dep} now
-being up to date. When a build completes, we can record the resulting order,
-using it as a starting point for the next build and fixing it up if a task's
-dependencies have changed, as \Excel does.
-
+being up to date.
 This approach requires a way to abort tasks that have failed due to out-of-date
@@
 
-\Bazel's restarting scheduler does not store the discovered order between build
-runs; instead, it stores the most recent task dependency information. Since this
-information may become outdated, \Bazel may also need to abort a task if a
-newly discovered dependency is out-of-date.
+To reduce the number of aborts (often to zero) Excel records the
+discovered task order in its \emph{calc chain}, and uses it as the
+starting point for the next build (\S\ref{sec-background-excel}).
+\Bazel's restarting scheduler does not store the discovered order
+between build runs; instead, it stores the most recent task dependency
+information. Since this information may become outdated, \Bazel may
+also need to abort a task if a newly discovered dependency is
+out-of-date.
 
@@
 
-\subsection{Determining out-of-date keys} \label{sec-out-of-date}
+\subsection{The rebuilder: determining out-of-date keys} \label{sec-out-of-date}
 
@@
 two builds is marked dirty. If a key and all its transitive dependencies are
-clean, the key does not need to be rebuild.
+clean, the key does not need to be rebuilt.
 
 \Excel models the dirty bit approach most directly, having an actual dirty bit
-associated with each cell, marking the cell dirty if the user modifies it, and
-also making sure all cells that depend on it are recursively marked dirty too.
-To mark dependent cells dirty, it is necessary to statically know task
-dependencies. This is possible only for applicative tasks (see \S\ref{sec-deps}).
-For monadic tasks, \Excel computes a \emph{static approximation} of task
-dependencies: for functions such as \cmd{IF}, the approximation includes all
-cells that are mentioned in the formula (the \cmd{IF}'s condition and both
-branches), whereas functions such as \cmd{INDIRECT} whose dependencies cannot be
-guessed, are conservatively assumed to depend on all cells and are therefore
-always marked dirty.
+associated with each cell, marking the cell dirty if the user modifies it.
+It also marks dirty all cells that depend (transitively) on the modified cell.
+Excel does not record the dynamic dependencies of each cell,
+so instead it computes a \emph{static over-approximation} -- it is always
+safe for it to make more cells dirty than is strictly necessary, but not
+vice versa.  The over-approximation is as follows: a cell is marked dirty
+if its formula statically refers to a dirty cell, or if
+the formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
+guessed from the formula alone.  The over-approximation is clear for \cmd{INDIRECT},
+but it is also present for \cmd{IF}, where both branches are followed even though
+dynamically only one is used.
 

### 2018-07-05 paper/6-engineering.tex
Polishing (thanks, Andrew!)
Add references on fixed-point/cyclic tasks

@@
 \begin{minted}{haskell}
-sprsh5 "B1" = Just $ Task $ \fetch -> do
+sprsh4 "B1" = Just $ Task $ \fetch -> do
     formula <- fetch "B1-formula"
@@
 % would be required, so instead the formula can be captured as a dependency
-% (but its value not used) and \emph{also} baked directly into the \hs{sprsh5}
+% (but its value not used) and \emph{also} baked directly into the \hs{sprsh4}
 % function. \simon{Don't understand.}
@@
 
-Overall we choose not to deal with cyclic rules, a choice that most build
-systems also follow.
+Overall we choose not to deal with cyclic dependencies, a choice that most build
+systems also follow. There are computation frameworks that support tasks with
+cyclic dependencies under the assumption that tasks are \emph{monotonic} in a
+certain sense, e.g. see~\cite{pottier2009lazy} and~\cite{radul2009propagation}.
 

### 2018-07-05 paper/7-related.tex
Refer to earlier redo papers

@@
 
-\item \Redo~\cite{redo} almost exactly matches \Shake at the level of detail
-given here, differing only on aspects like polymorphic
-dependencies~\S\ref{sec-polymorphism}.
+\item \Redo~\cite{redo-idea}\cite{grosskurth2007redo}\cite{redo} almost exactly
+matches \Shake at the level of detail given here, differing only on aspects like
+polymorphic dependencies~\S\ref{sec-polymorphism}.
 

### 2018-07-05 paper/4-build.tex
Another wibble in Section 4

@@
 The deep constructive trace approach only works if the operations are
-\emph{deterministic}, otherwise it is possible to violate correctness resulting
-in a so-called ``Frankenbuild''~\cite{esfahani2016cloudbuild}. Concretely, if
+\emph{deterministic}, otherwise it is possible to violate correctness.
+Concretely, if
 the result of building a key given a set of immediate dependencies can vary,
@@
 different values of \hs{k} will have been used in one rule, leading to problems
-such as link-errors for typical compilers.
+such as link-errors for typical compilers -- a phenomenon known as a
+\emph{Frankenbuild}~\cite{esfahani2016cloudbuild}.
 

### 2018-07-05 paper/3-abstractions.tex
Clarify what c stands for in Task c k v
Add a ref to Docker containers

@@
 \noindent
-% Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative} or \hs{Monad}.
+Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative} or \hs{Monad}.
 A \hs{Task} describes a single build task, while \hs{Tasks} associates a
@@
   even use the application operator \hs{<*>}, which limits dependencies to a
-  single linear chain, as e.g. in the Docker build system. It is interesting to
-  note that, when run on such a task description, the \hs{busy} build system
-  will build each key at most once, thus partially fulfilling the minimality
-  requirement~\ref{def-minimal}. Alas, it still has no mechanism to decide which
-  input keys changed since the previous build.
+  linear chain, as e.g. in Docker containers~\cite{docker}. It is
+  interesting to note that, when run on such a task description, the \hs{busy}
+  build system will build each key at most once, thus partially fulfilling the
+  minimality requirement~\ref{def-minimal}. Alas, it still has no mechanism to
+  decide which input keys changed since the previous build.
 \item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their

### 2018-07-05 paper/6-engineering.tex
Polish S6.3

@@
 
-\item[Volatility] Some tasks are defined to change in every execution, for
-example \Excel provides a~``function'' \cmd{RANDBETWEEN} which produces a random
-number in a specified range~--~on each recalculation it is defined to change.
-Similarly, build systems like \Make and \Shake provide \emph{phony rules} which
-are also volatile. The main difference from non-deterministic tasks, is that
-volatile tasks cannot be cached. They are best handled by introducing a special
-key \cmd{RealWorld} whose value is changed between every run.
+\item[Volatility] Some tasks are defined to change in every build, e.g. \Excel
+provides a~``function'' \cmd{RANDBETWEEN} producing a fresh random number in a
+specified range on each recalculation. Similarly, build systems like \Make and
+\Shake provide volatile \emph{phony rules}. The main difference from
+non-deterministic tasks is that volatile tasks cannot be cached. They are best
+modelled as depending on a special key \cmd{RealWorld} whose value is changed
+in every build.
 \end{description}
@@
 \noindent
-Handling such tasks is possible in our framework, but requires an adjustment of
-the correctness definition (\S\ref{sec-build-correctness}): instead of requiring
-that the result of recomputing the \hs{task} matches the value stored in the
-\hs{result}:
+Such tasks can be modelled in our framework by adjusting the correctness
+definition (\S\ref{sec-build-correctness}): instead of requiring that the
+\hs{result}ing value \emph{equals} the one obtained by recomputing the \hs{task}:
 \[
-\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
+\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}\text{,}
 \]
 \noindent
-we now require that \hs{result} contains \emph{one possible result of recomputing}
-the \hs{task}:
+we now require that \hs{result} contains \emph{one possible result of
+recomputing} the \hs{task}:
 \[
-\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}
+\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}\text{,}
 \]

### 2018-07-05 paper/6-engineering.tex
Polish S6.4

@@
 \begin{description}
-\item[Eviction] The store of traces as shown grows indefinitely, but often
-resource constraints require evicting old items from the store. One option is
-to evict the contents and any trace that mentions the now-defunct
-\hs{Hash}~\hs{v}. However, if the build system can defer materialisation, it
-may be possible to only evict the contents, allowing builds to pass-through
-hashes of values where the underlying value is not known. If so, the build must
-be able to recreate the value if required, potentially dealing with a different
-result in a future run.
-
-% \item[Frankenbuilds] A build is considered a
-% \emph{frankenbuild}~\cite{esfahani2016cloudbuild} if a value is calculated
-% locally, but something that depends on that key is pulled from the cache, and
-% the value calculated locally does not match what was previously calculated and
-% stored in the cloud. \simon{Don't understand. The explanation in 4.2.4 is better -- and I think describes
-% the same thing?}
-% Our implementations avoid this issue by storing complete
-% traces, but if a cloud build system was to only reference input nodes this
-% situation can arise.
-
-\item[Communication] When traces or contents are stored in the cloud
+\item[Communication] When traces or contents are stored in the cloud,
 communication can become a bottleneck, so it is important to send only the
@@
 
-\item[Shallow builds] Sometimes input files will involve many intermediate tasks
-before producing the end result, e.g. an installer package. These intermediate
-steps may be large, so some cloud build systems are designed to build end
-products \emph{without} downloading or \emph{materialising} the results of
-intermediate tasks~--~only the final result~--~a so-called \emph{shallow build}.
-Some build systems can go even further, integrating with the file system to only
-materialise the file when the user accesses it \cite{gvfs}.
+\item[Eviction] The cloud storage, as modelled
+in~\S\ref{sec-constructive-traces}, grows indefinitely, but often resource
+constraints require evicting old items from the store. When evicting an old
+value \hs{v}, one can also evict all traces mentioning the now-defunct
+\hs{hash}~\hs{v}. However, for shallow builds (see below), it is beneficial to
+keep these traces, allowing builds to ``pass-through'' hashes whose underlying
+values are not known, recreating them only when they must be materialised.
+
+\item[Shallow builds] Building the end result, e.g. an installer package, often
+involves many intermediate tasks. The intermediate results may be large, so some
+cloud build systems are designed to build end products \emph{without
+materialising} intermediate results, producing a so-called \emph{shallow build}
+(see an example in~\S\ref{sec-background-bazel}). Some build systems can go even
+further, integrating with the file system to only materialise the file when the
+user accesses it \cite{gvfs}.
 \end{description}

### 2018-07-05 paper/6-engineering.tex
Polish S6.5-6.6

@@
 \[
-\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}
+\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}\text{,}
 \]
@@
 \[
-\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result}.
+\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result}\text{.}
 \]
@@
 
-\subsection{Tracking and self-tracking}\label{sec-tracking-aspects}
+\subsection{Self-tracking}\label{sec-tracking-aspects}
 
-Some build systems, for example \Excel and \Ninja, are capable of recomputing a task if either its dependencies change, \emph{or} the rule itself changes. For example:
+Some build systems, for example \Excel and \Ninja, are capable of recomputing a
+task if either its dependencies change, \emph{or} the rule itself changes. For
+example:
 
+\vspace{0.5mm}
 \begin{minted}[xleftmargin=10pt]{text}
@@
 
-\begin{minted}{haskell}
+\vspace{0.5mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 sprsh4 "B1" = Just $ Task $ \fetch -> do
@@
 \end{minted}
-
-% It is not possible to
-% change dependencies based on the formula in a \hs{Tasks}~\hs{Applicative}, as
-% would be required, so instead the formula can be captured as a dependency
-% (but its value not used) and \emph{also} baked directly into the \hs{sprsh4}
-% function. \simon{Don't understand.}
-% \simon{I think you mean that you can find dependencies with executing the function,
-%   which seems quite difference from ``backing it into'' the Haskell code.}
+\vspace{0.5mm}
 
@@
 The build systems that have precise self-tracking are all ones which use a
-\emph{non-embedded domain specific language} to describe computations.
-Those which make use of a full programming language, e.g. \Shake, are faced with
-the challenge of implementing equality on arbitrary task functions. For such
-build systems the incredibly pessimistic assumption of saying that any change to
-the build system potentially changes any build rule can often be used~--~the
-classic example being a makefile depending on itself.
+\emph{non-embedded domain specific language} to describe build tasks. Those
+which use a full programming language, e.g. \Shake, are faced with the challenge
+of implementing equality on arbitrary task functions. For such build systems the
+pessimistic assumption of saying that any change to the build system potentially
+changes any build rule can often be used~--~the classic example being a makefile
+depending on itself.
 
@@
 Some computations are best described not by a chain of acyclic dependencies,
-but by a loop~--~for example \LaTeX~requires repeated rebuilding until it
-reaches a fixed point~--~something that can be directly expressed in build
-systems, such as \Pluto~\cite{erdweg2015pluto}. Another example of cyclic
-computations is \Excel, where a cell can depend on itself, for example:
-
-\begin{minted}[xleftmargin=10pt]{text}
-A1 = A1 + 1
-\end{minted}
-
-In such cases \Excel will normally not execute anything, but if the
-``Iterative Calculations'' feature is enabled will execute the formula for a
-specified maximum number $N$ of times per calculation (where $N$ is a
-setting that defaults to 100).
+but by a loop. For example, \LaTeX~requires repeated rebuilding until it
+reaches a fixed point, which can be directly expressed in build systems, such as
+\Pluto~\cite{erdweg2015pluto}. Another example is \Excel, where a cell can
+depend on itself, for example: \cmd{A1 = A1 + 1}. In such cases \Excel will
+normally not execute anything, but if the ``Iterative Calculations'' feature is
+enabled will execute the formula for a specified maximum number $N$ of times per
+calculation (where $N$ is a setting that defaults to 100).
 
@@
 with \LaTeX, which can be put into a situation where it is bistable or diverging
-over multiple executions).
-
-The examples in \Excel tend to encode either mutable state, or recurrence
-relations. The former is only required because \Excel inherently lacks the
-ability to write mutable state, and the latter is probably better solved using
-explicit recurrence formulae.
+over multiple executions). The examples in \Excel tend to encode either mutable
+state, or recurrence relations. The former is only required because \Excel
+inherently lacks the ability to write mutable state, and the latter is probably
+better solved using explicit recurrence formulae.
 

### 2018-07-05 paper/6-engineering.tex
Polish S6.7-S7

@@
 Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build
-system that works directly on \hs{k} and \hs{v} values. However, certain build
+system that works directly on \hs{k} and \hs{v} values. However, some build
 systems provide greater flexibility, e.g. \Shake permits polymorphic keys and
-values, allowing types that are only stored in the \Shake information, and never
-persisted to the store.
+values, allowing types that are stored only in the \Shake's build information.
 
@@
 to define an \emph{oracle} rule as per~\cite{mitchell2012shake} whose value is
-the result of running \cmd{gcc --version} and which is volatile, making the
+the result of running \cmd{gcc -}\cmd{-version} and which is volatile, making the
 \cmd{gcc} version something that can be depended upon. Of course, provided the

### 2018-07-05 paper/7-related.tex
Polish S6.7-S7

@@
 Most build systems, when viewed at the level we talk, can be captured with minor
-variations on the code presented in \S\ref{sec-implementations}. As some
-examples:
+variations on the code presented in \S\ref{sec-implementations}. Below we list
+some notable examples:
 
 \begin{itemize}
+\item \Dune~\cite{dune} is a build system designed for OCaml/Reason projects.
+Its distinguishing feature is that it uses
+\emph{arrows}~\cite{hughes2000generalising} rather than monads to model
+dynamic dependencies, which simplifies static dependency approximation.
+
 \item \Ninja~\cite{ninja} combines the \hs{topological} scheduler of \Make with
@@
 multiple results, a limited form of polymorphism \S\ref{sec-polymorphism}.
-
-\item \Tup~\cite{tup} functions much like \Make, but with a refined dirty-bit
-implementation that watches the file system for changes and can thus avoid
-rechecking the entire graph, and with automatic deleting of stale results.
-
-\item \Redo~\cite{redo-idea}\cite{grosskurth2007redo}\cite{redo} almost exactly
-matches \Shake at the level of detail given here, differing only on aspects like
-polymorphic dependencies~\S\ref{sec-polymorphism}.
 
@@
 
-\item \Dune~\cite{dune} is a build system designed for OCaml/Reason projects.
-Its distinguishing feature is that it uses
-\emph{arrows}~\cite{hughes2000generalising} rather than monads to model
-dynamic dependencies, which simplifies static dependency approximation.
+\item \Redo~\cite{redo-idea}\cite{grosskurth2007redo}\cite{redo} almost exactly
+matches \Shake at the level of detail given here, differing only on aspects like
+polymorphic dependencies~\S\ref{sec-polymorphism}.
+
+\item \Tup~\cite{tup} functions much like \Make, but with a refined dirty-bit
+implementation that watches the file system for changes and can thus avoid
+rechecking the entire graph, and with automatic deleting of stale results.
 \end{itemize}
@@
 
-\begin{minted}[xleftmargin=10pt]{haskell}
-data Key = A Integer | B Integer | D Integer Integer deriving Ord
+\begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
+data Key = A Integer | B Integer | D Integer Integer
 \end{minted}
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
+\vspace{0mm}
+\begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
 editDistance :: Tasks Monad Key Integer

### 2018-07-05 paper/1-intro.tex
Changes to intro

--- 2018-07-03 paper/1-intro.tex
@@
 of a variety of different real-life build systems, including \Make, \Shake,
-\Bazel, \CloudBuild, \Buck, \Nix, and \Excel, each in a dozen lines of code
-or so (\S\ref{sec-implementations}). Doing this modelling in a single setting
+\Bazel, \CloudBuild, \Buck, \Nix, and \Excel\footnote{\Excel appears very
+different to the others but, seen through the lens of this paper, it is very
+close indeed.}, each by the composition of the two design choices
+(\S\ref{sec-implementations}). Doing this modelling in a single setting
 allows the differences and similarities between these huge systems to be
@@
 
-Papers about ``frameworks'' are often fuzzy.  This one is not: all our
-abstractions are defined in Haskell, and we have (freely-available)
-executable models of all the build systems we describe.
+% Papers about ``frameworks'' are often fuzzy.  This one is not: all our
+% abstractions are defined in Haskell, and we have (freely-available)
+% executable models of all the build systems we describe.

### 2018-07-05 paper/2-background.tex
Address suggestions of reviewers
Tweaks to S2

@@
 \Shake~\cite{mitchell2012shake}, \Bazel~\cite{bazel}, and
-\Excel~\cite{advanced_excel}\footnote{\Excel appears very different to the
-others but, seen through the lens of this paper, it is very close indeed.}.
+\Excel~\cite{advanced_excel}.
 We have carefully chosen these four to illustrate the various axes on which
@@
 There are two input cells \cmd{A1} and \cmd{A2}, and a single task that computes
-the sum of their values, writing the result into the cell \cmd{B1}. If either of
+the sum of their values, producing the result in cell \cmd{B1}. If either of
 the inputs change, \Excel will recompute the result.
@@
 \cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
-formula in \cmd{B1} are determined by the value of \cmd{C1}, so there is no
-possibility\footnote{In this particular example one might say that the value of
+formula in \cmd{B1} are determined by the value of \cmd{C1}, so it is impossible
+to compute the dependency graph before the build starts\footnote{In
+this particular example one might say that the value of
 \cmd{C1} is statically known, but imagine that it is the result of a long
-computation chain -- its value will only become available during the build.}
-of computing the dependency graph before the build starts.
+computation chain -- its value will only become available during the build.}.
 
@@
 truly circular dependencies); the calc chain is just an optimisation.
-We refer to this algorithm as \emph{restarting}, and will discuss it
+We refer to this algorithm as \emph{restarting}, and discuss it
 in more detail in~\S\ref{sec-implementation-excel}.

### 2018-07-05 paper/3-abstractions.tex
S3 updates

@@
 putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
+data Hash v
 getHash    :: Hashable v => k -> Store i k v -> Hash v
@@
 familiar with \emph{lenses} or \emph{profunctor optics} might recognise a
-familiar pattern. We discuss this in~\S\ref{sec-related-optics}.} \hs{Task} is
-best introduced by an example. Consider this \Excel spreadsheet:
+familiar pattern. We discuss this in~\S\ref{sec-related-optics}.} \hs{Task}
+describes how to build a value when given a way to build its dependencies, and is
+best explained by an example. Consider this \Excel spreadsheet:
 
@@
 pure  :: Applicative f => a -> f a
-(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
+(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
 (<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
@@
 compute :: Task Monad k v -> Store i k v -> v
-compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
+compute task store = runIdentity $ run task (\k -> return (getValue k store))
 \end{minted}
@@
 argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
-To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
-monad: the function
-\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
-\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
-result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.
+To fix the mismatch, we unwrap using the function \hs{runIdentity} of type \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
+which both forces the monad to \hs{Identity} and strips it away.
 
@@
 task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
-accumulate a list of the task's dependencies -- and that is just what
+accumulate a list of the task's dependencies -- and that is what
 \hs{dependencies} does:
@@
 \hs{trackingFetch} that, in addition to fetching a value, tracks the
-corresponding key/value pair. The \hs{task} returns the value of type
-\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap by
+corresponding key/value pair. The \hs{task} returns a value of type
+\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap with
 \hs{runWriterT}. We will use \hs{track} when implementing monadic build systems

### 2018-07-05 src/Build/Task/Monad.hs

@@ -38,11 +38,11 @@ isInput tasks = isNothing . tasks
 
 -- | Run a task with a pure lookup function.
 computePure :: Task Monad k v -> (k -> v) -> v
-computePure task store = runIdentity $ run task (Identity . store)
+computePure task store = runIdentity $ run task (return . store)
 
 -- | Run a task in a given store.
 compute :: Task Monad k v -> Store i k v -> v
-compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
+compute task store = runIdentity $ run task (\k -> return (getValue k store))
 
 -- | Convert a task with a total lookup function @k -> m v@ into a task with a
 -- partial lookup function @k -> m (Maybe v)@. This essentially lifts the task

### 2018-07-05 paper/4-build.tex
Work on S4

@@
 This approach requires that a task may be started and then suspended until
-another task is complete. This can be done with cheap green threads and blocking
+another task is complete. Suspending can be done with cheap green threads and blocking
 (the original approach of \Shake) or using continuation-passing style (what
-\Shake does currently).
+\Shake currently does).
 
@@
 
-Suppose the scheduler decides that that a key should be brought up to date. The
+Suppose the scheduler decides that a key should be brought up to date. The
 next question is: does any work need to be done, or is it already up to date?
@@
 
-With a dirty bit it is simple to achieve minimality. However, to achieve early
+With a dirty bit it is possible to achieve minimality. However, to achieve early
 cutoff (\S\ref{sec-background-shake}) it would be important to reset the dirty
@@
 The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
-the hash of the resulting value, we require the whole value. The \hs{verifyVT}
+the hash of the resulting value, we require the actual value. The \hs{verifyVT}
 has been replaced with \hs{constructCT}, which instead of taking the hash of the
@@
 from the constructive list \emph{without} doing any work to build it, and copy
-it into the store from the cloud.
+it into the store.
 
@@
 \item Any \hs{Applicative} build system using constructive traces, e.g.
-\Bazel, can index directly from the key and results to the output result~--~i.e.
+\CloudBuild, can index directly from the key and results to the output result~--~i.e.
 \hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
@@
 With the information in this section we can build Table~\ref{tab-build-systems},
-which tabulates combinations of the scheduling strategy and the out-of-date keys
-strategy, providing 12~possible build systems, 8~of which are actually inhabited
-by existing build systems (we discuss all these systems in
+which tabulates combinations of the scheduling strategy and the out-of-date key
+strategy, providing 12~possible build systems, 8~of which are inhabited
+by existing build systems (we discuss these systems in~\S\ref{sec-background} and
 \S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
@@
 constructive traces, which would provide many benefits, and which we title
-\Cloud \Shake and further explore further in~\S\ref{sec-implementation-cloud}.
+\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
 % (as we plan on extending \Shake to occupy that spot)

### 2018-07-05 paper/5-implementations.tex
Work on S5

@@
 \noindent
-A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
+A \hs{Scheduler} is a function that takes a \hs{Rebuilder} and uses
 it to construct a \hs{Build} system, by choosing which keys to rebuild in which
@@
 
-The scheduler calls the rebuilder, but passes to it a \hs{fetch} function
+The scheduler calls the rebuilder, but passes it a \hs{fetch} function
 that that latter calls when it needs the value of a dependent key.  This
@@
             newValue <- liftStore (run newTask fetch)
-            modify $ updateValue key value newValue
-    order = case topSort (graph deps target) of
-        Nothing -> error "Cannot build tasks with cyclic dependencies"
-        Just xs -> xs
+            modify $ putValue key newValue
+    order = topSort (graph deps target)
     deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
@@
 reachable :: Ord k => (k -> [k]) -> k -> Graph k
-topSort   :: Ord k => Graph k -> Maybe [k] -- Returns Nothing on a cyclic graph
+topSort   :: Ord k => Graph k -> Maybe [k] -- Throws error on a cyclic graph
 \end{minted}
@@
 build information, carried by a \hs{MonadState} monad. This
-\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
+\hs{MakeInfo} comprises the \emph{current time} \hs{now}~\hs{::}~\hs{Time} and the map
 \hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
@@
     newChain <- liftChain $ go Set.empty $ chain ++ [target | target `notElem` chain]
-    modify . mapInfo $ \(ir, _) -> (ir, newChain)
+    modify $ mapInfo $ \(ir, _) -> (ir, newChain)
   where
@@
     go _    []       = return []
-    go done (key:ks) = case tasks key of
-        Nothing -> (key :) <$> go (Set.insert key done) ks
+    go done (key:keys) = case tasks key of
+        Nothing -> (key :) <$> go (Set.insert key done) keys
         Just task -> do
@@
             case result of
-                Left dep -> go done $ dep : filter (/= dep) ks ++ [key]
+                Left dep -> go done $ dep : filter (/= dep) keys ++ [key]
                 Right newValue -> do modify $ putValue key newValue
-                                     (key :) <$> go (Set.insert key done) ks
+                                     (key :) <$> go (Set.insert key done) keys
 \end{minted}
@@
     cachedValues <- constructCT key (fmap hash . fetch) =<< get
-    if value `elem` cachedValues then return value else case cachedValues of
-        (cachedValue:_) -> return cachedValue
-        _ -> do (newValue, deps) <- track task fetch
+    case cachedValues of
+        _ | value `elem` cachedValues -> return value
+        cachedValue:_ -> return cachedValue
+        [] -> do (newValue, deps) <- track task fetch
                 modify $ recordCT key newValue [ (k, hash v) | (k, v) <- deps ]
@@
 rules in the language available to users. Instead, users have to rely on a
-(rich) collection of predefined built-in rules, which cover many but not all
-practically important cases of dynamic dependencies.
+(rich) collection of predefined built-in rules, which cover many
+important instances of dynamic dependencies.
 
@@
 scheduler and is therefore an applicative build system, whereas \Nix uses the
-\hs{suspending} scheduler and is monadic. Deep constructive traces allow these
-build systems to efficiently support shallow builds: the value of the target key
-can be downloaded by examining only the hashes of all transitive inputs,
-skipping the intermediate keys. The downside of this approach is the loss of the
-early cutoff property.
+\hs{suspending} scheduler and is monadic.
 
@@
 is possible to define and test such a system, which we call \Cloud \Shake. All
-we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
-of \Bazel, as shown in Fig.~\ref{fig-cloud-implementations}.
+we need to do is combine \hs{suspending} with \hs{ctRebuilder},
+as shown in Fig.~\ref{fig-cloud-implementations}.

### 2018-07-05 src/Build/Scheduler.hs

@@ -74,7 +74,7 @@ topological rebuilder tasks target = execState $ mapM_ build order
                 fetch :: k -> State i v
                 fetch k = return (getValue k store)
             newValue <- liftStore (run newTask fetch)
-            modify $ updateValue key value newValue
+            modify $ putValue key newValue
     order = case topSort (graph deps target) of
         Nothing -> error "Cannot build tasks with cyclic dependencies"
         Just xs -> xs

### 2018-07-05 paper/5-implementations.tex
Fix topological

@@
             modify $ putValue key newValue
-    order = topSort (graph deps target)
-    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
+    order = topSort (graph dep target)
+    dep k = case tasks k of { Nothing -> []; Just task -> dependencies task }
 \end{minted}
@@
 reachable :: Ord k => (k -> [k]) -> k -> Graph k
-topSort   :: Ord k => Graph k -> Maybe [k] -- Throws error on a cyclic graph
+topSort   :: Ord k => Graph k -> [k] -- Throws error on a cyclic graph
 \end{minted}

### 2018-07-05 paper/6-engineering.tex
S6

@@
 
-In our model we define a \hs{Task} as a function -- when given the same inputs
+In our model we define \hs{Task} as a function -- when given the same inputs
 it will always produce the same output. Alas, the real-world is not so obliging.
@@
 
-For examples like \LaTeX~we consider the proper encoding to not be with circular
-tasks, but with a series of iterative steps, as described
+For examples like \LaTeX~we consider the proper encoding to not be circular
+tasks, but a series of iterative steps, as described
 by~\citet{shake-fixed-point}. It is important that the number of executions is

### 2018-07-05 paper/7-related.tex
S7

@@
 implementation that watches the file system for changes and can thus avoid
-rechecking the entire graph, and with automatic deleting of stale results.
+rechecking the entire graph. \Tup also automatically deletes stale results.
 \end{itemize}
@@
 
+\begin{comment}
 \subsection{Profunctor Optics}\label{sec-related-optics}
@@
 the \hs{Build} abstraction lacks any further such symmetry.
+\end{comment}

### 2018-07-05 paper/3-abstractions.tex
Drop forward reference to profunctor optics

@@
 A \hs{Task} describes a single build task, while \hs{Tasks} associates a
-\hs{Task} to every non-input key; input keys are associated with
-\hs{Nothing}. The highly-abstracted type\footnote{Readers
-familiar with \emph{lenses} or \emph{profunctor optics} might recognise a
-familiar pattern. We discuss this in~\S\ref{sec-related-optics}.} \hs{Task}
-describes how to build a value when given a way to build its dependencies, and is
-best explained by an example. Consider this \Excel spreadsheet:
+\hs{Task} to every non-input key; input keys are associated with \hs{Nothing}.
+The highly-abstracted type \hs{Task} describes how to build a value when given
+a way to build its dependencies, and is best explained by an example.
+Consider this \Excel spreadsheet:
+% \footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics} might
+% recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}
 

### 2018-07-05 paper/1-intro.tex
Minor revision

@@
 build systems, in a way that is both abstract (omitting incidental detail)
-and yet precise (implemented as Haskell code:
-\url{https://hackage.haskell.org/package/build-0.0.1.1}). Specifically we make
-these contributions:
+and yet precise (implemented as Haskell code). Specifically we make these
+contributions:
 \begin{itemize}
@@
 allows the differences and similarities between these huge systems to be
-brought out clearly.
+brought out clearly\footnote{All our models are executable and are available on
+Hackage: \url{https://hackage.haskell.org/package/build-0.0.1.1}.}.
 

### 2018-07-05 paper/3-abstractions.tex
Minor revision

@@
 putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
-data Hash v
+\end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+data Hash v -- a compact summary of a value with a fast equality check
+hash    :: Hashable v => v -> Hash v
 getHash :: Hashable v => k -> Store i k v -> Hash v
-hash       :: Hashable v => v -> Hash v -- provides a fast equality check
 \end{minted}

### 2018-07-06 paper/3-abstractions.tex
Minor revision

@@
 \section{Build systems, abstractly}\label{sec-abstractions}
+\vspace{-0.5mm}
 
@@
 \end{minted}
-\vspace{0mm}
+\vspace{-1mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \end{minted}
-\vspace{-3mm}
+\vspace{-3.5mm}
 \caption{Type signatures of key build systems abstractions.}\label{fig-types}
-\vspace{-3mm}
+\vspace{-4.5mm}
 \end{figure}
@@
 \vspace{-0.5mm}
-% We begin by establishing a common vocabulary for build systems:
 
@@
 \noindent
-Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative} or \hs{Monad}.
+Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative}
+(\S\ref{sec-why-polymorphism} explains why we need it).
 A \hs{Task} describes a single build task, while \hs{Tasks} associates a
@@
 Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains the
-formula \cmd{A1+A2}, etc.
+formula \cmd{A1 + A2}, etc. We can represent the formulae of this spreadsheet
+with the following task description:
 
@@
 \end{figure}
-
-\noindent
-We can represent the formulae of this spreadsheet with the following task
-description:
 
@@
 compute :: Task Monad k v -> Store i k v -> v
-compute task store = runIdentity $ run task (\k -> return (getValue k store))
+compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
 \end{minted}
@@
 argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
-To fix the mismatch, we unwrap using the function \hs{runIdentity} of type \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
-which both forces the monad to \hs{Identity} and strips it away.
+To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
+monad: the function
+\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
+\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
+result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.
+
+% To fix the mismatch, we unwrap using the function \hs{runIdentity} of type \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
+% which both forces the monad to \hs{Identity} and strips it away.
 

### 2018-07-06 paper/4-build.tex
Minor revision

@@
 associated with each cell, marking the cell dirty if the user modifies it.
-It also marks dirty all cells that depend (transitively) on the modified cell.
-Excel does not record the dynamic dependencies of each cell,
-so instead it computes a \emph{static over-approximation} -- it is always
-safe for it to make more cells dirty than is strictly necessary, but not
-vice versa.  The over-approximation is as follows: a cell is marked dirty
-if its formula statically refers to a dirty cell, or if
-the formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
-guessed from the formula alone.  The over-approximation is clear for \cmd{INDIRECT},
-but it is also present for \cmd{IF}, where both branches are followed even though
-dynamically only one is used.
+It also marks dirty all cells that (transitively) depend on the modified cell.
+Excel does not record dynamic dependencies of each cell; instead it computes a
+\emph{static over-approximation} -- it is safe for it to make more cells dirty
+than necessary, but not vice versa. The over-approximation is as follows: a cell
+is marked dirty if its formula statically refers to a dirty cell, or if the
+formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
+guessed from the formula alone. The over-approximation is clear for
+\cmd{INDIRECT}, but it is also present for \cmd{IF}, where both branches are
+followed even though dynamically only one is used.
 
@@
 special case of $n = \infty$, and constructive traces are the special case of
-$n = 1$.
+$n = 1$. When $n \ge 2$, deep constructive traces require the tasks to be
+\emph{deterministic}, as otherwise it is possible to violate correctness, as
+illustrated by an example in~\S\ref{sec-cloud-aspects}.
 
-The deep constructive trace approach only works if the operations are
-\emph{deterministic}, otherwise it is possible to violate correctness.
-Concretely, if
-the result of building a key given a set of immediate dependencies can vary,
-then it is possible that key \hs{k} will have been computed locally, but
-something \emph{using} \hs{k} will have been taken as a constructive input using
-a different value for \hs{k}. If another rule combines those two values then two
-different values of \hs{k} will have been used in one rule, leading to problems
-such as link-errors for typical compilers -- a phenomenon known as a
-\emph{Frankenbuild}~\cite{esfahani2016cloudbuild}.
+% Concretely, if
+% the result of building a key given a set of immediate dependencies can vary,
+% then it is possible that key \hs{k} will have been computed locally, but
+% something \emph{using} \hs{k} will have been taken as a constructive input using
+% a different value for \hs{k}. If another rule combines those two values then two
+% different values of \hs{k} will have been used in one rule, leading to problems
+% such as link-errors for typical compilers -- a phenomenon known as a
+% \emph{Frankenbuild}~\cite{esfahani2016cloudbuild}.
 
-A downside of deep constructive traces is that they cannot support cutoff
-(\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies. On
-the other hand, these traces are particularly useful for ``shallow builds'', as
+A downside of deep constructive traces is that they cannot support early cutoff
+(\S\ref{sec-background-shake}), other than at $n$ levels of dependencies. On the
+other hand, these traces are particularly useful for \emph{shallow builds}, as
 discussed in~\S\ref{sec-cloud-aspects}.
@@
 
-\noindent
 Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
@@
 \begin{enumerate}
-\item Any system using verifying traces, e.g. \Shake, is unlikely to see
-significant benefit from storing more than one \hs{Trace} per key\footnote{There
-is a small chance of a benefit if the dependencies change but the result does
-not, and then the dependencies change back to what they were before.}.
-Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash v))},
-where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
+\item Any system using verifying traces is unlikely to see significant benefit
+from storing more than one \hs{Trace} per key\footnote{There is a small chance
+of a benefit if the dependencies change but the result does not, and then the
+dependencies change back to what they were before.}. Therefore, such systems can
+store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
+initial \hs{k} is the \hs{key} field of \hs{Trace}.
 

### 2018-07-06 paper/5-implementations.tex
Minor revision

@@
 \hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
-build information, carried by a \hs{MonadState} monad. This
-\hs{MakeInfo} comprises the \emph{current time} \hs{now}~\hs{::}~\hs{Time} and the map
-\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
-modification times}.
-We assume that the external system, which invokes the build system, updates
-\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.
+build information, carried by a state monad. This \hs{MakeInfo} comprises
+the \emph{current time} \hs{now}~\hs{::}~\hs{Time} and the map
+\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file modification
+times}. We assume that the external system, which invokes the build system,
+updates \hs{MakeInfo} reflecting any file changes between successive builds.
 

### 2018-07-06 src/Build/Task/Monad.hs

@@ -38,11 +38,11 @@ isInput tasks = isNothing . tasks
 
 -- | Run a task with a pure lookup function.
 computePure :: Task Monad k v -> (k -> v) -> v
-computePure task store = runIdentity $ run task (return . store)
+computePure task store = runIdentity $ run task (Identity . store)
 
 -- | Run a task in a given store.
 compute :: Task Monad k v -> Store i k v -> v
-compute task store = runIdentity $ run task (\k -> return (getValue k store))
+compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
 
 -- | Convert a task with a total lookup function @k -> m v@ into a task with a
 -- partial lookup function @k -> m (Maybe v)@. This essentially lifts the task

### 2018-07-06 paper/4-build.tex
Add reference to continuations

@@
 another task is complete. Suspending can be done with cheap green threads and blocking
-(the original approach of \Shake) or using continuation-passing style (what
+(the original approach of \Shake) or using continuation-passing style \citep{claessen_continuations} (what
 \Shake currently does).

### 2018-07-06 src/Build/Scheduler.hs
Make suspending recursive

@@ -179,20 +179,18 @@ restarting2 rebuilder tasks target = execState $ go (enqueue target [] mempty)
 suspending :: forall i k v. Ord k => Scheduler Monad i i k v
 suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
   where
-    build :: k -> State (Store i k v, Set k) ()
-    build key = case tasks key of
-        Nothing -> return ()
-        Just task -> do
-            done <- gets snd
-            when (key `Set.notMember` done) $ do
+    build :: k -> State (Store i k v, Set k) v
+    build key = do
+        done <- gets snd
+        case tasks key of
+            Just task | key `Set.notMember` done -> do
                 value <- gets (getValue key . fst)
                 let newTask :: Task (MonadState i) k v
                     newTask = rebuilder key value task
-                    fetch :: k -> State (Store i k v, Set k) v
-                    fetch k = do build k                      -- build the key
-                                 gets (getValue k . fst)      -- fetch the value
-                newValue <- liftRun newTask fetch
+                newValue <- liftRun newTask build
                 modify $ \(s, d) -> (updateValue key value newValue s, Set.insert key d)
+                return newValue
+            _ -> gets (getValue key . fst) -- fetch the existing value
 
 
 newtype Wrap i k v a = Wrap {unwrap :: State (Store i k v, Set k) a}

### 2018-07-06 paper/5-implementations.tex
S5 update suspending to use liftRun and be actually recursive

@@
   where
-    build :: k -> State (Store i k v, Set k) (@@)
-    build key = case tasks key of
-        Nothing -> return (@@)
-        Just task -> do
+    build :: k -> State (Store i k v, Set k) v
+    build key = do
         done <- gets snd
-            when (key `Set.notMember` done) $ do
+        case tasks key of
+            Just task | key `Set.notMember` done -> do
                 value <- gets (getValue key . fst)
@@
                     newTask = rebuilder key value task
-                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
-                    fetch k = do lift (build k)                      -- build the key
-                                 lift (gets (getInfo . fst)) >>= put -- save new traces
-                                 lift (gets (getValue k . fst))      -- fetch the value
-                newValue <- flattenInfo (run newTask fetch)
-                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
+                newValue <- liftRun newTask build
+                modify $ \(s, d) -> (updateValue key value newValue s, Set.insert key d)
+                return newValue
+            _ -> gets (getValue key . fst) -- fetch the existing value
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Collapse two layers of stateful computations into one (implementation omitted)
-flattenInfo :: StateT i (State (Store i k v, Set k)) a -> State (Store i k v, Set k) a
+-- Run a Task using Store to obtain the i
+liftRun :: Task (MonadState i) k v
+        -> (k -> State (Store i k v, Set k) v) -> State (Store i k v, Set k) v
 \end{minted}
@@
 These steps are run in the
-\hs{StateT}~\hs{i}~\hs{(@@State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k))}
-monad, because the \hs{newTask} has access only to the~\hs{i} part of the state;
-we therefore need to use \hs{lift} when executing commands in the inner
-\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
-We use the helper function \hs{flattenInfo} to collapse two layers of stateful
-computations into one. The \hs{newValue} obtained by running the \hs{newTask} is
+\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)}
+monad. To make \hs{MonadState}~\hs{i} access the \hs{i} inside the \hs{Store}
+we use the helper function \hs{liftRun} (which uses a \hs{newtype} to provide a \hs{MonadState}
+instance that sees through into the \hs{Store}).
+The \hs{newValue} obtained by running the \hs{newTask} is
 stored, and the \hs{key} is added to the set \hs{done}.

### 2018-07-06 paper/6-engineering.tex
Minor revision

@@
 
+\textbf{TODO:} Explain the Frankenbuild example.
+
+\begin{figure}
+\vspace{-2mm}
+\begin{subfigure}[b]{0.25\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-build.pdf}}
+\vspace{-1mm}
+\caption{Build \cmd{report.txt}}
+\end{subfigure}
+\begin{subfigure}[b]{0.40\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-clean.pdf}}
+\vspace{-1mm}
+\caption{Clean up and evict \cmd{main.prof}}
+\end{subfigure}
+\begin{subfigure}[b]{0.31\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-rebuild.pdf}}
+\vspace{-1mm}
+\caption{Full rebuild}
+\end{subfigure}
+\vspace{-2.5mm}
+\caption{A Frankenbuild example: (a)~build a profiling report for \cmd{main.exe}
+and record deep constructive traces in the cloud, (b)~clean built files locally
+and evict \cmd{main.prof} from the cloud storage, (c)~rebuild \cmd{main.prof} by
+executing the build task, rebuild \cmd{report.txt} by downloading it from the
+cloud, resulting in a Frankenbuild. New and evicted cloud storage entries are
+highlighted.\label{fig-frankenbuild}}
+\vspace{-3mm}
+\end{figure}
+
 \subsection{Self-tracking}\label{sec-tracking-aspects}

### 2018-07-06 paper/6-engineering.tex
Tweak Frankenbuild figure caption

@@
 and evict \cmd{main.prof} from the cloud storage, (c)~rebuild \cmd{main.prof} by
-executing the build task, rebuild \cmd{report.txt} by downloading it from the
-cloud, resulting in a Frankenbuild. New and evicted cloud storage entries are
-highlighted.\label{fig-frankenbuild}}
+executing the build task (profiling is non-deterministic, hence the new hash),
+rebuild \cmd{report.txt} by downloading it from the cloud, resulting in a
+Frankenbuild. New and evicted cloud storage entries are highlighted.
+\label{fig-frankenbuild}}
 \vspace{-3mm}

### 2018-07-06 paper/5-implementations.tex
Rename build to fetch in suspending

@@
 suspending :: Ord k => Scheduler Monad i i k v
-suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
+suspending rebuilder tasks target store = fst $ execState (fetch target) (store, Set.empty)
   where
-    build :: k -> State (Store i k v, Set k) v
-    build key = do
+    fetch :: k -> State (Store i k v, Set k) v
+    fetch key = do
         done <- gets snd
@@
                     newTask = rebuilder key value task
-                newValue <- liftRun newTask build
+                newValue <- liftRun newTask fetch
                 modify $ \(s, d) -> (updateValue key value newValue s, Set.insert key d)

### 2018-07-06 src/Build/Scheduler.hs

@@ -177,17 +177,17 @@ restarting2 rebuilder tasks target = execState $ go (enqueue target [] mempty)
 -- It stores the set of keys that have already been built as part of the state
 -- to avoid executing the same task twice.
 suspending :: forall i k v. Ord k => Scheduler Monad i i k v
-suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
+suspending rebuilder tasks target store = fst $ execState (fetch target) (store, Set.empty)
   where
-    build :: k -> State (Store i k v, Set k) v
-    build key = do
+    fetch :: k -> State (Store i k v, Set k) v
+    fetch key = do
         done <- gets snd
         case tasks key of
             Just task | key `Set.notMember` done -> do
                 value <- gets (getValue key . fst)
                 let newTask :: Task (MonadState i) k v
                     newTask = rebuilder key value task
-                newValue <- liftRun newTask build
+                newValue <- liftRun newTask fetch
                 modify $ \(s, d) -> (updateValue key value newValue s, Set.insert key d)
                 return newValue
             _ -> gets (getValue key . fst) -- fetch the existing value

### 2018-07-06 paper/6-engineering.tex
Explain what makes it a frankenbuild

@@
 rebuild \cmd{report.txt} by downloading it from the cloud, resulting in a
-Frankenbuild. New and evicted cloud storage entries are highlighted.
+Frankenbuild because \cmd{report.txt} and \cmd{main.prof} are inconsistent.
+New and evicted cloud storage entries are highlighted.
 \label{fig-frankenbuild}}

### 2018-07-06 paper/5-implementations.tex
Simplify the description of suspending

@@
 suspending :: Ord k => Scheduler Monad i i k v
-suspending rebuilder tasks target store = fst $ execState (fetch target) (store, Set.empty)
+suspending rebuilder tasks target store = fst $ execState@\,@(fetch target)@\,@(store,@\,@Set.empty)
   where
@@
                 newValue <- liftRun newTask fetch
-                modify $ \(s, d) -> (updateValue key value newValue s, Set.insert key d)
+                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
                 return newValue
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Run a Task using Store to obtain the i
+-- Run a task using a callback that operates on a larger state (implementation omitted)
 liftRun :: Task (MonadState i) k v
@@
 
-The \hs{suspending} scheduler uses a recursive function \hs{build} that builds
+The \hs{suspending} scheduler uses a recursive \hs{fetch} callback, defined
+similarly to the \hs{busy} build system (\S\ref{sec-general-build}), that builds
 a given \hs{key}, making sure not to duplicate work when called on the same
 \hs{key} again in future. To achieve that, it keeps track of keys that have
-already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}.
-% , running the computation in the
-% \hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.
+already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}. Given a non-input
+\hs{key} that has not yet been built, we use the supplied \hs{rebuilder} to
+embed the build information~\hs{i} into the \hs{task}. We then execute the
+obtained \hs{newTask} by passing it the \hs{fetch} function as a callback for
+building dependencies: the \hs{newTask} will therefore be suspended while its
+dependencies are being brought up to date. The \hs{newValue} obtained by running
+the \hs{newTask} is stored, and the \hs{key} is added to the set \hs{done}.
 
-Given a non-input \hs{key} that has not yet been built, we use the supplied
-\hs{rebuilder} to embed the build information~\hs{i} into the \hs{task}.
-We then execute the obtained \hs{newTask} with the \hs{fetch} callback, which
-performs the following three steps when given a dependency \hs{k}:
+The \hs{fetch} computation runs in the
+\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad. To make
+\hs{MonadState}~\hs{i} access the \hs{i} inside the \hs{Store} we use the helper
+function \hs{liftRun} (which uses a \hs{newtype} to provide a \hs{MonadState}
+instance that sees through into the \hs{Store}).
 
-\begin{itemize}
-    \item Build \hs{k} by recursively calling the \hs{build} function. Note
-          that if \hs{k} is dirty, this will effectively suspend the build of
-          the current \hs{key} until \hs{k} is brought up to date.
-    \item Save any new traces that have been generated when building \hs{k}.
-    \item Read the (up to date) value of \hs{k} from the store and return it.
-\end{itemize}
-
-These steps are run in the
-\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)}
-monad. To make \hs{MonadState}~\hs{i} access the \hs{i} inside the \hs{Store}
-we use the helper function \hs{liftRun} (which uses a \hs{newtype} to provide a \hs{MonadState}
-instance that sees through into the \hs{Store}).
-The \hs{newValue} obtained by running the \hs{newTask} is
-stored, and the \hs{key} is added to the set \hs{done}.
-
-\subsection{Cloud build systems: \Bazel, \CloudBuild, \Cloud \Shake, \Buck and \Nix}\label{sec-implementation-cloud}
+\subsection{Cloud build systems: \Bazel, \CloudBuild, \Cloud \Shake, \Buck and \Nix}
+\label{sec-implementation-cloud}
 

### 2018-07-06 paper/6-engineering.tex
Explain the Frankenbuild example

--- 2018-07-06 paper/6-engineering.tex
@@
 
-\textbf{TODO:} Explain the Frankenbuild example.
-
 \begin{figure}
@@
 \vspace{-1mm}
-\caption{Build \cmd{report.txt}}
+\caption{Initial build}
 \end{subfigure}
@@
 \vspace{-1mm}
-\caption{Full rebuild}
+\caption{Build \cmd{main.prof} and \cmd{report.txt}}
 \end{subfigure}
-\vspace{-2.5mm}
-\caption{A Frankenbuild example: (a)~build a profiling report for \cmd{main.exe}
-and record deep constructive traces in the cloud, (b)~clean built files locally
-and evict \cmd{main.prof} from the cloud storage, (c)~rebuild \cmd{main.prof} by
-executing the build task, rebuild \cmd{report.txt} by downloading it from the
-cloud, resulting in a Frankenbuild. New and evicted cloud storage entries are
-highlighted.\label{fig-frankenbuild}}
+\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
+\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
+tool and record deep constructive traces in the cloud, (b)~remove built files
+locally and evict \cmd{main.prof} from the cloud storage, (c)~build
+\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
+the new hash), then build \cmd{report.txt} by downloading it from the matching
+deep constructive trace in the cloud, resulting in a Frankenbuild because
+\cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted cloud
+storage entries are highlighted; file hashes are shown in circles.
+\label{fig-frankenbuild}}
 \vspace{-3mm}
 \end{figure}
+
+Deep constructive traces (\S\ref{sec-deep-constructive-traces}) combined with
+task non-determinism (\S\ref{sec-non-determinism}) can lead to very subtle bugs,
+especially in the cloud setting. Fig.~\ref{fig-frankenbuild} shows a
+\emph{Frankenbuild}~\cite{esfahani2016cloudbuild} example, where the target
+\cmd{report.txt}, which is downloaded from the cloud, is inconsistent with its
+immediate dependency \cmd{main.prof}. This is caused by (i)~inherent
+non-determinism of profiling: running a profiling tool on the very same
+\cmd{main.exe} will produce different \cmd{main.prof} results every time; and
+(ii)~relying on deep constructive traces, which cache build results based only
+on the hashes of terminal task inputs.
 

### 2018-07-06 paper/3-abstractions.tex
Simplify the description of suspending
Expand the Frankenbuild explanation

@@
 
-The previous example shows why the \hs{Task} abstraction is polymorphic in \hs{f}, recall the definition:
+The previous example shows why the \hs{Task} abstraction is polymorphic in
+\hs{f}, recall the definition:
 
@@
   \cmd{D}.
-%TODO: Add a reference for Docker?
 \item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
@@
 
-\subsection{Correctness of a build system} \label{sec-build-correctness}
+\subsection{Correctness of a build system}\label{sec-build-correctness}
 

### 2018-07-06 paper/6-engineering.tex
Expand the Frankenbuild explanation

@@
 \end{subfigure}
+\vspace{-1mm}
 \caption{A Frankenbuild example: (a)~build a human-readable profiling report for
@@
 \label{fig-frankenbuild}}
-\vspace{-3mm}
+\vspace{-4mm}
 \end{figure}
@@
 \cmd{report.txt}, which is downloaded from the cloud, is inconsistent with its
-immediate dependency \cmd{main.prof}. This is caused by (i)~inherent
-non-determinism of profiling: running a profiling tool on the very same
-\cmd{main.exe} will produce different \cmd{main.prof} results every time; and
-(ii)~relying on deep constructive traces, which cache build results based only
-on the hashes of terminal task inputs.
+immediate dependency \cmd{main.prof}. This is caused by two factors:
+(i)~inherent non-determinism of profiling: running a profiling tool on the very
+same \cmd{main.exe} will produce different \cmd{main.prof} results every time;
+and (ii)~relying on deep constructive traces, which cache build results based
+only on the hashes of terminal task inputs (in this case \cmd{main.exe}). Note
+that the resulting store is incorrect according to all three definitions of
+correctness: the main definition (\S\ref{sec-build-correctness}), the variant
+for non-deterministic tasks (\S\ref{sec-non-determinism}) and the variant for
+shallow builds (this section).
 

### 2018-07-06 paper/6-engineering.tex
Two minor rewordings

@@
 \cmd{report.txt}, which is downloaded from the cloud, is inconsistent with its
-immediate dependency \cmd{main.prof}. This is caused by two factors:
+immediate dependency \cmd{main.prof}. This inconsistency is caused by two factors:
 (i)~inherent non-determinism of profiling: running a profiling tool on the very

### 2018-07-06 test/Examples.hs
Minor revision

@@ -111,19 +111,19 @@ staticIF _ _    = Nothing
 
 -------------------------- Dynamic programming example -------------------------
 
-data Key = A Integer | B Integer | D Integer Integer
+data Key = A Int | B Int | C Int Int
 
-editDistance :: Tasks Monad Key Integer
-editDistance (D i 0) = Just $ Task $ const $ pure i
-editDistance (D 0 j) = Just $ Task $ const $ pure j
-editDistance (D i j) = Just $ Task $ \fetch -> do
+editDistance :: Tasks Monad Key Int
+editDistance (C i 0) = Just $ Task $ const $ pure i
+editDistance (C 0 j) = Just $ Task $ const $ pure j
+editDistance (C i j) = Just $ Task $ \fetch -> do
     ai <- fetch (A i)
     bj <- fetch (B j)
     if ai == bj
-        then fetch (D (i - 1) (j - 1))
+        then fetch (C (i - 1) (j - 1))
         else do
-            insert  <- fetch (D  i      (j - 1))
-            delete  <- fetch (D (i - 1)  j     )
-            replace <- fetch (D (i - 1) (j - 1))
+            insert  <- fetch (C  i      (j - 1))
+            delete  <- fetch (C (i - 1)  j     )
+            replace <- fetch (C (i - 1) (j - 1))
             return (1 + minimum [insert, delete, replace])
 editDistance _ = Nothing

### 2018-07-06 paper/5-implementations.tex
Give a bit more space to S5
Mostly whitespace tweaks

@@
 
+\vspace{0.5mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 liftStore :: State i a -> State (Store i k v) a
-liftStore x = do (a, newInfo) <- gets (runState x . getInfo)
+liftStore x = do
+    (a, newInfo) <- gets (runState x . getInfo)
     modify (putInfo newInfo)
@@
 \caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
-\vspace{-3mm}
+\vspace{-5mm}
 \end{figure}
@@
 \end{minted}
-\vspace{0mm}
+\vspace{1mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \end{minted}
-\vspace{-3mm}
+\vspace{-2.5mm}
 \caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
-\vspace{-5mm}
+\vspace{-2.5mm}
 \end{figure}
@@
 
-\vspace{-1mm}
 \subsection{\Shake}\label{sec-implementation-shake}
-\vspace{-1mm}
 
@@
 \end{minted}
+\vspace{-1mm}
+\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
 \vspace{-3mm}
-\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
-\vspace{-6mm}
 \end{figure}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Rebuilder based on deep constructive traces, analogous to ctRebuilder
+-- Rebuilder based on deep constructive traces, analogous to 'ctRebuilder'
 dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
@@
 \label{fig-cloud-implementations}
-\vspace{-4mm}
+\vspace{-2mm}
 \end{figure}

### 2018-07-06 paper/6-engineering.tex
Mostly whitespace tweaks

@@
 respectively. More concretely, the following task description corresponds to
-a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}:
+a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}.
 
-\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
 sprsh3 :: Tasks MonadPlus String Integer
-sprsh3 "B1"@\,@=@\,@Just $ Task $ \fetch -> (+) <$> fetch "A1" <*> (pure 1 <|> pure 2)
+sprsh3 "B1"@\,@=@\,@Just $ Task $ \fetch -> (+) <$> fetch "A1" <*> (pure@\,@1 <|> pure@\,@2)
 sprsh3 _   @\,@=@\,@Nothing
@@
 \centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-build.pdf}}
-\vspace{-1mm}
 \caption{Initial build}
@@
 \centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-clean.pdf}}
-\vspace{-1mm}
 \caption{Clean up and evict \cmd{main.prof}}
@@
 \centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-rebuild.pdf}}
-\vspace{-1mm}
 \caption{Build \cmd{main.prof} and \cmd{report.txt}}
@@
 that the resulting store is incorrect according to all three definitions of
-correctness: the main definition (\S\ref{sec-build-correctness}), the variant
-for non-deterministic tasks (\S\ref{sec-non-determinism}) and the variant for
-shallow builds (this section).
+correctness: the main definition~(\S\ref{sec-build-correctness}), the variant
+for non-deterministic tasks~(\S\ref{sec-non-determinism}) and the variant for
+shallow builds~(this section).
 

### 2018-07-06 paper/7-related.tex
Mostly whitespace tweaks

@@
 We address elements of strings $a_i$ and $b_i$ by keys \hs{A}~$i$ and \hs{B}~$i$,
-respectively, while the cost of a subproblem $d_{ij}$ is identified by
-\hs{D}~$i$~$j$.
+respectively, while the cost of a subproblem $c_{ij}$ is identified by
+\hs{C}~$i$~$j$.
 
 \begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
-data Key = A Integer | B Integer | D Integer Integer
+data Key = A Int | B Int | C Int Int
 \end{minted}
@@
 \begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
-editDistance :: Tasks Monad Key Integer
-editDistance (D i 0) = Just $ Task $ const $ pure i
-editDistance (D 0 j) = Just $ Task $ const $ pure j
-editDistance (D i j) = Just $ Task $ \fetch -> do
+editDistance :: Tasks Monad Key Int
+editDistance (C i 0) = Just $ Task $ const $ pure i
+editDistance (C 0 j) = Just $ Task $ const $ pure j
+editDistance (C i j) = Just $ Task $ \fetch -> do
     ai <- fetch (A i)
     bj <- fetch (B j)
-    if ai == bj then fetch (D (i - 1) (j - 1))
-                else do insert  <- fetch (D  i      (j - 1))
-                        delete  <- fetch (D (i - 1)  j     )
-                        replace <- fetch (D (i - 1) (j - 1))
+    if ai == bj
+        then fetch (C (i - 1) (j - 1))
+        else do insert  <- fetch (C  i      (j - 1))
+                delete  <- fetch (C (i - 1)  j     )
+                replace <- fetch (C (i - 1) (j - 1))
                 return (1 + minimum [insert, delete, replace])
@@
 \noindent
-When asked to build \hs{D}~$n$~$m$, a minimal build system will calculate the
+When asked to build \hs{C}~$n$~$m$, a minimal build system will calculate the
 result using memoization. Furthermore, when an input symbol $a_i$ is changed,

### 2018-07-06 test/Examples.hs
Add missing Eq in editDistance example

@@ -111,7 +111,7 @@ staticIF _ _    = Nothing
 
 -------------------------- Dynamic programming example -------------------------
 
-data Key = A Int | B Int | C Int Int
+data Key = A Int | B Int | C Int Int deriving Eq
 
 editDistance :: Tasks Monad Key Int
 editDistance (C i 0) = Just $ Task $ const $ pure i

### 2018-07-06 paper/7-related.tex
Add missing Eq in editDistance example
Two more tweaks

@@
 so widely known (e.g., see~\cite{cormen2001introduction}) that we provide its
-encoding in our \hs{Task} abstraction without further explanation.
-We address elements of strings $a_i$ and $b_i$ by keys \hs{A}~$i$ and \hs{B}~$i$,
+encoding in our \hs{Tasks} abstraction without further explanation. We address
+elements of strings $a_i$ and $b_i$ by keys \hs{A}~$i$ and \hs{B}~$i$,
 respectively, while the cost of a subproblem $c_{ij}$ is identified by
@@
 
+\vspace{0.5mm}
 \begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
-data Key = A Int | B Int | C Int Int
+data Key = A Int | B Int | C Int Int deriving Eq
 \end{minted}

### 2018-07-06 paper/3-abstractions.tex
Two minor rewordings
Drop TODOs and old comments

@@
 
-% TODO: Shall we describe the lifTask and liftTasks combinators? They aren't pretty.
-%
-%   liftTask :: Task Applicative k v -> Task Monad k v
-%   liftTask (Task task) = Task task
-%
-%   liftTasks :: Tasks Applicative k v -> Tasks Monad k v
-%   liftTasks = fmap (fmap liftTask)
-%
-% Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
-% argument, an application of \hs{busy} to a \hs{Task}~\hs{Functor} or
-% a \hs{Task}~\hs{Applicative} will typecheck and run just fine. It feels a bit like
-% sub-typing, but is actually just ordinary higher-rank polymorphism at
-% work~\cite{jones2007practical}.
-
 \vspace{-0.5mm}
@@
 compute :: Task Monad k v -> Store i k v -> v
-compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
+compute task store = runIdentity $ run task (\k -> Identity@\,@(getValue k store))
 \end{minted}
@@
 \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
-result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.
+result has type \hs{Identity}~\hs{v}, which we unwrap with \hs{runIdentity}.
 
-% To fix the mismatch, we unwrap using the function \hs{runIdentity} of type \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
+% To fix the mismatch, we unwrap using the function \hs{runIdentity} of type
+% \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
 % which both forces the monad to \hs{Identity} and strips it away.
@@
     build.
-    %\footnote{\todo{SLPJ}{Add the joke that GHC used to delete inputs?}}.
 

### 2018-07-06 paper/7-related.tex
Drop TODOs and old comments

@@
 abstraction waiting to be discovered.
-
-\begin{comment}
-\subsection{Profunctor Optics}\label{sec-related-optics}
-
-The definition of \hs{Task} is:
-
-\begin{minted}[xleftmargin=10pt]{haskell}
-newtype Task c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
-\end{minted}
-
-\noindent
-Which looks tantalisingly close to the profunctor optics definition by
-\cite{pickering2017profunctor}:
-
-%TODO: Add a reference to the paper about forall f. c f => (k -> f v) -> f a
-\begin{minted}[xleftmargin=10pt]{haskell}
-type Optic p a b s t = p a b -> p s t
-\end{minted}
-\noindent
-Provided we instantiate \hs{p}~\hs{a}~\hs{b} to something like
-\hs{k}~\hs{->}~\hs{f}~\hs{v}~--~which many of the actual instances in that paper
-do. The properties of such optics are well studied, and the functions like
-\hs{dependencies} are very much based on observations from that field of work.
-Alas, we have been unable to remove the \hs{Maybe} used to encode whether a file
-is an input, without complicating other aspects of our definition. Furthermore,
-the \hs{Build} abstraction lacks any further such symmetry.
-\end{comment}

### 2018-07-06 src/Build/Scheduler.hs
Drop old TODOs and fix minor inconsistencies

@@ -29,9 +29,10 @@ type Scheduler c i j k v = Rebuilder c j k v -> Build c i k v
 
 -- | Lift a computation operating on @i@ to @Store i k v@.
 liftStore :: State i a -> State (Store i k v) a
-liftStore x = do (a, newInfo) <- gets (runState x . getInfo)
-                 modify (putInfo newInfo)
-                 return a
+liftStore x = do
+    (a, newInfo) <- gets (runState x . getInfo)
+    modify (putInfo newInfo)
+    return a
 
 -- | Lift a computation operating on @Store i k v@ to @Store (i, j) k v@.
 liftInfo :: State (Store i k v) a -> State (Store (i, j) k v) a
@@ -67,7 +68,7 @@ topological rebuilder tasks target = execState $ mapM_ build order
                 fetch :: k -> State i v
                 fetch k = return (getValue k store)
             newValue <- liftStore (run newTask fetch)
-            modify $ putValue key newValue
+            modify $ updateValue key value newValue
     order = case topSort (graph deps target) of
         Nothing -> error "Cannot build tasks with cyclic dependencies"
         Just xs -> xs
@@ -135,10 +136,6 @@ enqueue key blocked ((k, bs):q)
 dequeue :: Queue k -> Maybe (k, [k], Queue k)
 dequeue []          = Nothing
 dequeue ((k, bs):q) = Just (k, bs, q)
-
--- TODO: The implementation below is specialised to constructive traces @CT@.
--- Can we make it polymorphic over the type of build information @i@ like other
--- schedulers?
 
 -- | A model of the scheduler used by Bazel. We extract a key K from the queue
 -- and try to build it. There are now two cases:

### 2018-07-06 test/Spreadsheet.hs

@@ -78,7 +78,6 @@ rel = RelativeReference
 -- the mapping returns @Nothing@ are inputs.
 type Spreadsheet = Cell -> Maybe Formula
 
--- TODO: Implement 'Random'.
 -- | Monadic spreadsheet computation.
 spreadsheetTask :: Spreadsheet -> Tasks Monad Cell Int
 spreadsheetTask spreadsheet cell@(Cell r c) = case spreadsheet cell of
@@ -95,7 +94,7 @@ spreadsheetTask spreadsheet cell@(Cell r c) = case spreadsheet cell of
                 IfZero fx fy fz         -> do
                     x <- go fx
                     if x == 0 then go fy else go fz
-                Random _ _      -> error "Random not implemented"
+                Random _ _      -> error "Not supported by monadic tasks"
 
 -- | Applicative spreadsheet computation.
 spreadsheetTaskA :: Spreadsheet -> Tasks Applicative Cell Int

### 2018-07-06 src/Build/Task/Depend.hs
Remove unused module

@@ -1,44 +1,0 @@
-{-# LANGUAGE DeriveFunctor #-}
-{-# OPTIONS_GHC -Wno-incomplete-uni-patterns #-}
-
--- | The \"free\" structures for dependencies, providing either an applicative
--- interface (for 'Depend') or a monadic interface (for 'Depends'). By passing
--- them to a suitable 'Task' you can reconstruct all necessary dependencies.
-module Build.Task.Depend (toDepend, Depend (..), toDepends, Depends (..)) where
-
-import Build.Task
-
------------------------------ Free Task Applicative ----------------------------
-
--- | A list of dependencies, and a function that when applied to those
--- dependencies produces the result.
-data Depend k v r = Depend [k] ([v] -> r)
-    deriving Functor
-
-instance Applicative (Depend k v) where
-    pure v = Depend [] (\[] -> v)
-    Depend d1 f1 <*> Depend d2 f2 = Depend (d1++d2) $
-        \vs -> let (v1,v2) = splitAt (length d1) vs in f1 v1 $ f2 v2
-
-toDepend :: Task Applicative k v -> Depend k v v
-toDepend f = run f $ \k -> Depend [k] $ \[v] -> v
-
--------------------------------- Free Task Monad -------------------------------
-
--- | A list of dependencies, and a function that when applied to those
--- dependencies either the result or more dependencies.
-data Depends k v r = Depends [k] ([v] -> Depends k v r)
-                   | Done r
-    deriving Functor
-
-instance Applicative (Depends k v) where
-    pure = return
-    f1 <*> f2 = f2 >>= \v -> ($ v) <$> f1
-
-instance Monad (Depends k v) where
-    return = Done
-    Done x >>= f = f x
-    Depends ds op >>= f = Depends ds $ \vs -> f =<< op vs
-
-toDepends :: Task Monad k v -> Depends k v v
-toDepends f = run f $ \k -> Depends [k] $ \[v] -> Done v

### 2018-07-06 src/Build/Multi.hs
Tweak comments

@@ -1,6 +1,5 @@
--- | Given a build system that can work with single keys, generalise that to one
--- that deals with multiple keys at a time.
-module Build.Multi (multi) where
+-- | Support for multiple-output tasks.
+module Build.Multi (Partition, multi) where
 
 import Data.Maybe
 import Build.Task
@@ -13,8 +12,8 @@ import Build.Task
 -- * @forall i \in ks . f i == ks@
 type Partition k = k -> [k]
 
--- | Given a build rule where you can build some combinations of multiple rules,
--- use a partition to enable building lots of multiple rule subsets.
+-- | Given a task description with individual multiple-output keys, compute its
+-- "closure" supporting all possible combinations of keys.
 multi :: Eq k => Partition k -> Tasks Applicative [k] [v] -> Tasks Applicative [k] [v]
 multi partition tasks keys
     | k:_ <- keys, partition k == keys = tasks keys

### 2018-07-06 src/Build/Task/Typed.hs

@@ -2,7 +2,7 @@
 {-# OPTIONS_GHC -Wno-unused-top-binds #-}
 {-# OPTIONS_GHC -Wno-unticked-promoted-constructors #-}
 
--- | A Typed version of dependencies where the value type depends on the key.
+-- | A model of polymorphic tasks, where the value type depends on the key.
 -- See the source for an example.
 module Build.Task.Typed (Task, Key (..), showDependencies) where
 

### 2018-07-06 paper/6-engineering.tex
Refer to examples of polymorphism in the framework

@@
 \Shake users have remarked that polymorphism provides a much easier expression
-of concepts, e.g.~\cite{hadrian}, but it is not essential and thus not necessary
-to model.
+of concepts, e.g.~\cite{hadrian}, but it is not essential and we therefore do
+not model it in this paper\footnote{See modules \cmd{Build.Multi} and
+\cmd{Build.Task.Typed} in our framework for models of multiple-output and
+polymorphic tasks.}.

### 2018-07-06 paper/6-engineering.tex
Final tweaks to Fig. 11

@@
 \begin{figure}
-\vspace{-2mm}
-\begin{subfigure}[b]{0.25\linewidth}
+\begin{subfigure}[b]{0.32\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-build.pdf}}
+\vspace{-1.5mm}
 \caption{Initial build}
 \end{subfigure}
-\begin{subfigure}[b]{0.40\linewidth}
+\begin{subfigure}[b]{0.32\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-clean.pdf}}
+\vspace{-1.5mm}
 \caption{Clean up and evict \cmd{main.prof}}
 \end{subfigure}
-\begin{subfigure}[b]{0.31\linewidth}
+\begin{subfigure}[b]{0.34\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-rebuild.pdf}}
+\vspace{-1.5mm}
 \caption{Build \cmd{main.prof} and \cmd{report.txt}}
 \end{subfigure}
-\vspace{-1mm}
+\vspace{-1.5mm}
 \caption{A Frankenbuild example: (a)~build a human-readable profiling report for
 \cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
-tool and record deep constructive traces in the cloud, (b)~remove built files
+tool, then record deep constructive traces in the cloud, (b)~remove built files
 locally and evict \cmd{main.prof} from the cloud storage, (c)~build

### 2018-07-06 paper/1-intro.tex
Link to build-1.0 on Hackage

@@
 brought out clearly\footnote{All our models are executable and are available on
-Hackage: \url{https://hackage.haskell.org/package/build-0.0.1.1}.}.
+Hackage: \url{https://hackage.haskell.org/package/build-1.0}.}.
 

### 2018-07-07 paper/2-background.tex
Final proofread

@@
 build systems differ; we discuss many other notable examples of build systems,
-and their relationships, in~\S\ref{sec-related}.
+and their relationships, in~\S\ref{sec-implementations} and~\S\ref{sec-related}.
 
@@
 executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
-the sake of simplicity. In reality, there are multiple versions of \cmd{gcc}
-and the actual binary used to compile and link files, and standard libraries
+the sake of simplicity. In reality, there are multiple versions of \cmd{gcc};
+the actual binary used to compile and link files, and standard libraries
 such that \cmd{stdio.h}, are often also listed as dependencies.} the command

### 2018-07-07 paper/3-abstractions.tex
Final proofread

@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ dependencies $ fromJust $ sprsh1 "A1"
-[@@]
+@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
+["A1","A2"]
 \end{minted}
 \begin{minted}[xleftmargin=10pt]{haskell}
-@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
-["A1", "A2"]
+@\ghci@ dependencies $ fromJust $ sprsh1 "B2"
+["B1"]
 \end{minted}

### 2018-07-07 paper/4-build.tex
Explain the Frankenbuild example
Mostly whitespace tweaks
Final proofread

@@
 
-To handle dynamic dependencies we can use the following approach:
-build tasks in an arbitrary initial order, discovering their dependencies on the fly;
-whenever a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task,
-and switch to building the dependency \hs{dep}; eventually the previously
-aborted task is restarted and makes further progress thanks to \hs{dep} now
-being up to date.
-This approach requires a way to abort tasks that have failed due to out-of-date
-dependencies. It is also not minimal in the sense that a task may start, do some
-meaningful work, and then abort.
+To handle dynamic dependencies we can use the following approach: build tasks in
+an arbitrary initial order, discovering their dependencies on the fly; whenever
+a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task, and
+switch to building the dependency \hs{dep}; eventually the previously aborted
+task is restarted and makes further progress thanks to \hs{dep} now being up to
+date. This approach requires a way to abort tasks that have failed due to
+out-of-date dependencies. It is also not minimal in the sense that a task may
+start, do some meaningful work, and then abort.
 
-To reduce the number of aborts (often to zero) Excel records the
+To reduce the number of aborts (often to zero) \Excel records the
 discovered task order in its \emph{calc chain}, and uses it as the
@@
 information. Since this information may become outdated, \Bazel may
-also need to abort a task if a newly discovered dependency is
-out-of-date.
+also need to abort a task if a newly discovered dependency is out of date.
 
@@
 
-Suppose the scheduler decides that a key should be brought up to date. The
-next question is: does any work need to be done, or is it already up to date?
+Suppose the scheduler decides that a key should be brought up to date. The next
+question is: does any work need to be done, or is the key already up to date?
 Or, in a cloud-build system, do we have a cached copy of the value we need?
@@
 It also marks dirty all cells that (transitively) depend on the modified cell.
-Excel does not record dynamic dependencies of each cell; instead it computes a
+\Excel does not record dynamic dependencies of each cell; instead it computes a
 \emph{static over-approximation} -- it is safe for it to make more cells dirty
@@
 current value as \emph{input}, returns a list of suitable values as \emph{output}.
-If the current value in the store
-matches one of the possible values, the build can skip this rule. If the
-resulting list is empty, the key must be rebuilt. However, if the current value
-does not match the store, and there is a possible value, we can use any value
-from the constructive list \emph{without} doing any work to build it, and copy
-it into the store.
+If the current value in the store matches one of the possible values, the build
+can skip this key. If the resulting list is empty, the key must be rebuilt.
+However, if the current value does not match the store, and there is a possible
+value, we can use any value from the constructive list \emph{without} doing any
+work to build it, and copy it into the store.
 
@@
 dependencies, which must have first been brought up to date, meaning that the
-time to verify a key is based on the number of transitive dependencies. A
+time to verify a key depends on the number of transitive dependencies. A
 \emph{deep} constructive trace optimises this process by only looking at the
@@
 illustrated by an example in~\S\ref{sec-cloud-aspects}.
-
-% Concretely, if
-% the result of building a key given a set of immediate dependencies can vary,
-% then it is possible that key \hs{k} will have been computed locally, but
-% something \emph{using} \hs{k} will have been taken as a constructive input using
-% a different value for \hs{k}. If another rule combines those two values then two
-% different values of \hs{k} will have been used in one rule, leading to problems
-% such as link-errors for typical compilers -- a phenomenon known as a
-% \emph{Frankenbuild}~\cite{esfahani2016cloudbuild}.
 
@@
 
-\begin{enumerate}
+\begin{itemize}
 \item Any system using verifying traces is unlikely to see significant benefit
@@
 their actual contents.
-\end{enumerate}
+\vspace{-1mm}
+\end{itemize}
 
@@
 \hline
- & \multicolumn{3}{c}{Scheduling strategy} \\
-Out-of-date key strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
+ & \multicolumn{3}{c}{Scheduling algorithm} \\
+Rebuilding strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
 \hline
@@
 With the information in this section we can build Table~\ref{tab-build-systems},
-which tabulates combinations of the scheduling strategy and the out-of-date key
+which tabulates combinations of the scheduling algorithm and the rebuilding
 strategy, providing 12~possible build systems, 8~of which are inhabited

### 2018-07-07 paper/5-implementations.tex
Final proofread

@@
 it to construct a \hs{Build} system, by choosing which keys to rebuild in which
-order. The \hs{Rebuilder} makes uses of the persistent build information
+order. The \hs{Rebuilder} makes use of the persistent build information
 \hs{ir}, while the scheduler might augment that with further persistent
@@
 The scheduler calls the rebuilder, but passes it a \hs{fetch} function
-that that latter calls when it needs the value of a dependent key.  This
+that the latter calls when it needs the value of a dependent key.  This
 callback returns control to the scheduler, which may in turn call the
@@
 the new modification time of the \hs{key}\footnote{The real \Make
-relies on the file system to track modification times, but we prefer
+relies on the file system to track file modification times, but we prefer
 to make it explicit here.}; otherwise we can just return the current \hs{value}.
@@
 to provide a transitively closed set of dirty bits. That is, if a
-cell is changed, its dirty bit is set, and the dirty bit of any cell
-whose value might perhaps change as a result.  It is OK to mark too many
+cell is changed, its dirty bit is set, as well as the dirty bit of any other
+cell whose value might perhaps change as a result. It is OK to mark too many
 cells as dirty; but not OK to mark too few.
@@
 (\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
-version (\S\ref{sec-deep-constructive-traces}).
+variant (\S\ref{sec-deep-constructive-traces}).
 
@@
 applying an adapter function \hs{adaptRebuilder}, which unwraps a given
-\hs{Task Applicative} and wraps it into \hs{Task Monad}.
+\hs{Task}~\hs{Applicative} and wraps it into \hs{Task}~\hs{Monad}.
 

### 2018-07-07 paper/6-engineering.tex
Final proofread

@@
 valid value. In the real world, build systems frequently deal with errors, e.g.
-``file not found'', or ``compilation failed''. We can model such failure
+``file not found'' or ``compilation failed''. We can model such failure
 conditions by instantiating \hs{v} to either \hs{Maybe}~\hs{v} (for missing
 values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
-model the values inside the store and the \hs{Task}, but because \hs{v} is
+model the errors inside the store and the \hs{Task}, but because \hs{v} is
 polymorphic for builds, it does not let the build system apply special behaviour
@@
 \item The \hs{restarting} scheduler can be made parallel in a few ways, but the
-most direct is to have $n$ threads reading entries from the list of keys. As
-before, if a key requires something not yet built, it is added to the end -- the
-difference is that sometimes things will be moved to the back of the queue not
-because they are out of order but because of races with earlier nodes that had
+most direct is to have $n$ threads reading keys from the build queue. As before,
+if a key requires a dependency not yet built, it is moved to the end~--~the
+difference is that sometimes keys will be moved to the back of the queue not
+because they are out of date but because of races with earlier tasks that had
 not yet finished. As a consequence, over successive runs, potentially racey
@@
 
-\item The \hs{suspending} function can be made parallel by starting static
+\item The \hs{suspending} scheduler can be made parallel by starting static
 dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
@@
 missing dependencies, but none are likely to capture all dependencies right down
-to CPU model and microcode version. Rules that do have untracked dependencies can be
-marked as volatile, a technique \Excel takes with the \cmd{INDIRECT} function,
-removing the untracked dependency at the cost of minimality.
+to the CPU model and microcode version. Tasks that do have untracked
+dependencies can be marked as volatile, a technique \Excel takes with the
+\cmd{INDIRECT} function, removing the untracked dependency at the cost of
+minimality.
 
@@
 materialising} intermediate results, producing a so-called \emph{shallow build}
-(see an example in~\S\ref{sec-background-bazel}). Some build systems can go even
+(see an example in~\S\ref{sec-background-bazel}). Some build systems go even
 further, integrating with the file system to only materialise the file when the
@@
 \cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
-the new hash), then build \cmd{report.txt} by downloading it from the matching
-deep constructive trace in the cloud, resulting in a Frankenbuild because
-\cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted cloud
-storage entries are highlighted; file hashes are shown in circles.
+the new hash value), then build \cmd{report.txt} by downloading it from the
+matching deep constructive trace in the cloud, resulting in a Frankenbuild
+because \cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted
+cloud storage entries are highlighted; file hashes are shown in circles.
 \label{fig-frankenbuild}}
@@
 Some build systems, for example \Excel and \Ninja, are capable of recomputing a
-task if either its dependencies change, \emph{or} the rule itself changes. For
+task if either its dependencies change, \emph{or} the task itself changes. For
 example:
@@
 which use a full programming language, e.g. \Shake, are faced with the challenge
-of implementing equality on arbitrary task functions. For such build systems the
-pessimistic assumption of saying that any change to the build system potentially
-changes any build rule can often be used~--~the classic example being a makefile
+of implementing equality on arbitrary task functions. For such build systems,
+the pessimistic assumption that any change to the build system potentially
+changes any build task can often be used~--~the classic example being a makefile
 depending on itself.
@@
 normally not execute anything, but if the ``Iterative Calculations'' feature is
-enabled will execute the formula for a specified maximum number $N$ of times per
-calculation (where $N$ is a setting that defaults to 100).
+enabled \Excel will execute the formula for a specified maximum number $N$ of
+times per calculation (where $N$ is a setting that defaults to 100).
 
@@
 That can be represented by having a key whose value is a pair of file names, and
-whose result is a pair of file contents. From that, the rule for \cmd{Foo.hi}
+whose result is a pair of file contents. From that, the task for \cmd{Foo.hi}
 can be the first component of the result of the pair. Again, such an operation

### 2018-07-07 paper/7-related.tex
Final proofread

@@
 Most build systems, when viewed at the level we talk, can be captured with minor
-variations on the code presented in \S\ref{sec-implementations}. Below we list
+variations of the code presented in \S\ref{sec-implementations}. Below we list
 some notable examples:
@@
 skipping any commands where no inputs have changed. The key difference from our
-\hs{Tasks} abstraction is that instead of supplying a mapping from output to action,
-\Fabricate supplies a list of statements, in an order, without declaring what each line produces.
-There is no need to schedule the statements, and not enough information to do so.
+\hs{Tasks} abstraction is that instead of supplying a mapping from outputs to
+tasks, \Fabricate supplies a list of statements, in an order, without declaring
+what each line produces. There is no need to schedule the statements, and not
+enough information to do so.
 
@@
 well studied area, and in particular the contrast between different formulations
-has been thoroughly investigated, e.g. see~\cite{acar2007selfadjusting}.
+has been thoroughly investigated, e.g. see~\citet{acar2007selfadjusting}.
 Self-adjusting computations can automatically adjust to an external change
@@
 algorithms for self-adjusting computations (with a few open-source
-implementations, e.g. \Incremental~\cite{incremental}). We plan to investigate
-how these insights can be utilised by build systems as future work.
+implementations, e.g. \Incremental by~\citet{incremental}). We plan to
+investigate how these insights can be utilised by build systems as future work.
 

### 2018-07-12 paper/2-background.tex
Comply with requirements for section heading capitalisation

@@
 
-\subsection{The venerable \Make: static dependencies and file modification times}
+\subsection{The Venerable \Make: Static Dependencies and File Modification Times}
 \label{sec-background-make}
@@
 
-\subsection{\Excel: dynamic dependencies at the cost of minimality}
+\subsection{\Excel: Dynamic Dependencies at the Cost of Minimality}
 \label{sec-background-excel}
@@
 
-\subsection{\Shake: dynamic dependencies with no remorse}
+\subsection{\Shake: Dynamic Dependencies with No Remorse}
 \label{sec-background-shake}
@@
 \vspace{-0.5mm}
-\subsection{\Bazel: a cloud build system}\label{sec-background-bazel}
+\subsection{\Bazel: A Cloud Build System}\label{sec-background-bazel}
 \vspace{-0.5mm}

### 2018-07-12 paper/4-build.tex
Comply with requirements for section heading capitalisation

@@
 
-\subsection{The scheduler: respecting the dependency order}
+\subsection{The Scheduler: Respecting the Dependency Order}
 \label{sec-dependency-orderings}
@@
 
-\subsection{The rebuilder: determining out-of-date keys} \label{sec-out-of-date}
+\subsection{The Rebuilder: Determining Out-of-date Keys}\label{sec-out-of-date}
 
@@
 \emph{deterministic}, as otherwise it is possible to violate correctness, as
-illustrated by an example in~\S\ref{sec-cloud-aspects}.
+illustrated by an example in~\S\ref{sec-cloud-aspects}
+(see Fig.~\ref{fig-frankenbuild}).
 
@@
 
-\subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}
+\subsection{Efficient Data Structures for Traces}\label{sec-smart-traces}
 
-In the examples above, we have used abstract types for the traces. Concretely, in our example code, they are all recorded as lists of:
+In the examples above, we have used abstract types for the traces. Concretely,
+in our example code, they are all recorded as lists of:
 

### 2019-03-05 papers/icfp/3-abstractions.tex
Move the ICFP paper into a separate folder

--- 2018-07-07 paper/3-abstractions.tex
@@
 \vspace{-1.5mm}
-\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
+\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
 \vspace{-0.5mm}
@@
 
-\subsection{The Task abstraction}\label{sec-task}
+\subsection{The Task Abstraction}\label{sec-task}
 
@@
 
-\subsection{The Build abstraction}\label{sec-general-build}
+\subsection{The Build Abstraction}\label{sec-general-build}
 
@@
 
-\subsection{The need for polymorphism in \hs{Task}}\label{sec-why-polymorphism}
+\subsection{The Need for Polymorphism in Task}\label{sec-why-polymorphism}
 \vspace{-0.5mm}
@@
 \vspace{-0.5mm}
-\subsection{Monadic tasks}\label{sec-task-monad}
+\subsection{Monadic Tasks}\label{sec-task-monad}
 
@@
 
-\subsection{Correctness of a build system}\label{sec-build-correctness}
+\subsection{Correctness of a Build System}\label{sec-build-correctness}
 
@@
 
-\subsection{Computing dependencies}\label{sec-deps}\label{secdeps}
+\subsection{Computing Dependencies}\label{sec-deps}\label{secdeps}
 

### 2019-03-05 papers/icfp/5-implementations.tex
Move the ICFP paper into a separate folder

--- 2018-07-07 paper/5-implementations.tex
@@
             modify $ putValue key newValue
-    order = topSort (graph dep target)
+    order = topSort (reachable dep target)
     dep k = case tasks k of { Nothing -> []; Just task -> dependencies task }
@@
 
-\subsection{Cloud build systems: \Bazel, \CloudBuild, \Cloud \Shake, \Buck and \Nix}
+\subsection{Cloud Build Systems: \Bazel, \CloudBuild, \Cloud \Shake, \Buck and \Nix}
 \label{sec-implementation-cloud}

### 2019-03-05 papers/jfp/1-intro.tex
Add first JFP version

--- 2018-07-06 paper/1-intro.tex
@@
 % system to be correct?'' or ``What are the trade-offs between different
-% approaches?''.  Moreover,
+% approaches?''.
 For years \Make dominated, but more recently the challenges of scale have driven
-large software firms like Microsoft, Facebook, and Google to develop their own
+large software firms like Microsoft, Facebook and Google to develop their own
 build systems, exploring new points in the design space. These complex build
@@
 \item Build systems vary on many axes, including: static vs dynamic
-  dependencies; local vs cloud; deterministic vs non-deterministic build rules;
+  dependencies; local vs cloud; deterministic vs non-deterministic build tasks;
   support for early cutoff; self-tracking build systems; and the type of
@@
 \item We identify two key design choices that are typically deeply wired into
-  any build system: \emph{the order in which dependencies are
+  any build system: \emph{the order in which tasks are
   built}~(\S\ref{sec-dependency-orderings}) and \emph{whether or not a
-  dependency is (re-)built}~(\S\ref{sec-out-of-date}). These choices turn out to
+  task is (re-)built}~(\S\ref{sec-out-of-date}). These choices turn out to
   be orthogonal, which leads us to a new classification of the design

### 2019-03-05 papers/jfp/2-background.tex
Add first JFP version

--- 2018-07-12 paper/2-background.tex
@@
 
-\definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
-most once per build and only if they transitively depend on inputs that changed
-since the previous build.}\label{def-minimal}
+\textbf{Definition} (Minimality). A build system is \emph{minimal} if it
+executes tasks at most once per build and only if they transitively depend on
+inputs that changed since the previous build.\label{def-minimal}
 \vspace{2mm}
@@
 \begin{table}[h]
-\vspace{-1mm}
-\smaller
+\vspace{-2mm}
+\caption{Build system differences.\label{tab-summary}}
+\vspace{-4mm}
+% \smaller
 \centering
@@
 $\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
-\hline
 \end{tabular}
-\caption{Build system differences; $^{(*)}$at present, user-defined build rules
-cannot have dynamic dependencies.\label{tab-summary}}
+\vspace{2mm}
+$^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
 \vspace{-8mm}

### 2019-03-05 papers/jfp/3-abstractions.tex
Add first JFP version

--- 2019-03-05 papers/icfp/3-abstractions.tex
@@
 
-\vspace{-1mm}
-\definition[Correctness]{Suppose \hs{build} is a build system, \hs{tasks} is a
-build task description, \hs{key} is a target key, \hs{store} is an initial
-store, and \hs{result} is the store produced by running the build system with
-parameters \hs{tasks}, \hs{key} and \hs{store}. Or, using the precise language of
-our abstractions:
+% \vspace{-1mm}
+\textbf{Definition}(Correctness). Suppose \hs{build} is a build system,
+\hs{tasks} is a build task description, \hs{key} is a target key, \hs{store} is
+an initial store, and \hs{result} is the store produced by running the build
+system with parameters \hs{tasks}, \hs{key} and \hs{store}. Or, using the
+precise language of our abstractions:
 
@@
 given \hs{tasks}, \hs{key} and \hs{store}.
-}
+
 \label{def-correct}
@@
 running its code in a different \hs{Applicative}! This is not new, for example
-see~\citet{free-applicatives}, but it is extremely cool. We will see a practical
+see~\cite{free-applicatives}, but it is extremely cool. We will see a practical
 use for \hs{dependencies} when implementing applicative build systems,

### 2019-03-05 papers/jfp/4-build.tex
Add first JFP version

--- 2018-07-12 paper/4-build.tex
@@
 This approach requires that a task may be started and then suspended until
-another task is complete. Suspending can be done with cheap green threads and blocking
-(the original approach of \Shake) or using continuation-passing style \citep{claessen_continuations} (what
-\Shake currently does).
+another task is complete. Suspending can be done with cheap green threads and
+blocking (the original approach of \Shake) or using continuation-passing style
+\cite{claessen_continuations} (what \Shake currently does).
 
@@
 \vspace{-2mm}
-\subsubsection{A dirty bit}\label{sec-dirty-bit}
+\subsubsection{A Dirty Bit}\label{sec-dirty-bit}
 
@@
 \vspace{-2mm}
-\subsubsection{Verifying traces}\label{sec-verifying-traces}
+\subsubsection{Verifying Traces}\label{sec-verifying-traces}
 
@@
 
-\subsubsection{Constructive traces}\label{sec-constructive-traces}
+\subsubsection{Constructive Traces}\label{sec-constructive-traces}
 
@@
 
-\subsubsection{Deep constructive traces}\label{sec-deep-constructive-traces}
+\subsubsection{Deep Constructive Traces}\label{sec-deep-constructive-traces}
 
@@
 \begin{table}
-\vspace{-1mm}
-\smaller
+\caption{Build systems \`a la carte.\label{tab-build-systems}}
+\vspace{-4mm}
+% \smaller
 \centering
@@
 \end{tabular}
-\vspace{1mm}
-\caption{Build systems \`a la carte.\label{tab-build-systems}}
-\vspace{-8mm}
+\vspace{-4mm}
 \end{table}

### 2019-03-05 papers/jfp/6-engineering.tex
Add first JFP version

--- 2018-07-07 paper/6-engineering.tex
@@
 
-\subsection{Partial stores and exceptions}
+\subsection{Partial Stores and Exceptions}
 
@@
 dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
-resolved, using the strategy described by~\citet{marlow2014haxl}.
+resolved, using the strategy described by~\cite{marlow2014haxl}.
 \end{enumerate}
@@
 
-\subsection{Impure computations}\label{sec-non-determinism}
+\subsection{Impure Computations}\label{sec-non-determinism}
 
@@
 \[
-\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}\text{,}
+\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result} %\text{,}
 \]
@@
 \[
-\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}\text{,}
+\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result} %\text{,}
 \]
@@
 
-\subsection{Cloud implementations}\label{sec-cloud-aspects}
+\subsection{Cloud Implementation}\label{sec-cloud-aspects}
 
@@
 \[
-\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}\text{,}
+\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result} % \text{,}
 \]
@@
 \[
-\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result}\text{.}
+\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result} % \text{.}
 \]
@@
 
-\subsection{Iterative computations}\label{sec-iterative-compute}
+\subsection{Iterative Computations}\label{sec-iterative-compute}
 
@@
 tasks, but a series of iterative steps, as described
-by~\citet{shake-fixed-point}. It is important that the number of executions is
+by~\cite{shake-fixed-point}. It is important that the number of executions is
 bounded, otherwise the build system may not terminate (a legitimate concern

### 2019-03-05 papers/jfp/8-conclusions.tex
Add first JFP version

--- 2018-03-10 paper/8-conclusions.tex
@@
-\clearpage
 \section{Conclusions}\label{sec-conclusions}
 
-...
+We have investigated multiple build systems, showing how their properties are consequences of two implementation choices: what order you build in and how you decide whether to rebuild.
+By first decomposing the pieces, we show how to recompose the pieces to find new points in the design space.
+In particular, a simple recombination leads to a design for a monadic suspending cloud build system.
+Armed with that blueprint we hope to actually implement such a system as future work.

### 2019-03-05 papers/jfp/9-appendix.tex
Add first JFP version

--- 2018-03-15 paper/9-appendix.tex
@@
   where
-    deps k = maybe [] snd (track task store k)
+    deps k = maybe [] snd (track task (\k -> getValue k store) k)
 

### 2019-03-05 papers/jfp/jfp2egui.tex
Add first JFP version

% This is file JFP2egui.tex
% release v1.02, 27th September 2001
%   (based on JFPguide.tex v1.11 for LaTeX 2.09)
% Copyright (C) 2001 Cambridge University Press

\NeedsTeXFormat{LaTeX2e}

\documentclass{jfp1}

%%% Macros for the guide only %%%
\providecommand\AMSLaTeX{AMS\,\LaTeX}
\newcommand\eg{\emph{e.g.}\ }
\newcommand\etc{\emph{etc.}}
\newcommand\bcmdtab{\noindent\bgroup\tabcolsep=0pt%
  \begin{tabular}{@{}p{10pc}@{}p{20pc}@{}}}
\newcommand\ecmdtab{\end{tabular}\egroup}
\newcommand\rch[1]{$\longrightarrow\rlap{$#1$}$\hspace{1em}}
\newcommand\lra{\ensuremath{\quad\longrightarrow\quad}}

\title[Journal of Functional Programming]
      {\LaTeXe\ guide for the Journal of Functional Programming}

 \author[M. A. Reed]
        {MARK A. REED\\
         Cambridge University Press, Cambridge CB2 2BS, UK\\
         \email{texline@cambridge.org}}

\jdate{September 2001, update April 2007}
\pubyear{2001}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\doi{S0956796801004857}

\newtheorem{lemma}{Lemma}[section]

\begin{document}

\label{firstpage}

\maketitle

\begin{abstract}
This guide is for authors who are preparing papers for the \emph{Journal of
Functional Programming} using the \LaTeXe\ document-preparation system
and the Functional Programming class file (\texttt{jfp1.cls}).
\end{abstract}

\tableofcontents

\section{Introduction}

In addition to the standard submission of hardcopy from authors, the
journal now accepts machine-readable forms of papers
in \LaTeXe. The layout design for the \emph{Functional Programming} journal
has been implemented as a \LaTeXe\ class file, based on the \verb"article"
class as discussed in the \LaTeX\ manual (2nd edition) \cite{LaTeX}.
Commands which differ from the standard \LaTeXe\ interface, or which are
provided in addition to the standard interface, are explained in this
guide (which is \emph{not} a substitute for the \LaTeXe\ manual itself).

Note that the final printed version of papers will use the Monotype Times
typeface rather than the Computer Modern typeface available to authors. For
this reason line and page breaks will change and authors should not insert
hard breaks in their text.

Authors planning to submit their papers in \LaTeXe\ are advised to use
\verb"jfp1.cls" as early as possible in the creation of their files.

The older \LaTeX\ style file is no longer supported, but authors should
find it easy to convert to \LaTeXe\ (see Section~\ref{usingjfpclass}).

\subsection{Introduction to \LaTeX}

\LaTeX\ is constructed as a series of macros on top of the \TeX\ typesetting
program. \LaTeX\ adds to \TeX\ a collection of facilities which simplify
typesetting for authors by allowing them to concentrate on the logical
structure of the document rather than its visual layout. Careful use of the
\LaTeX\ mark-up philosophy results in uniform layout rather than the
\emph{ad hoc} results of some word-processing systems. Authors are advised to
let the defaults control font selection etc., rather than tinker themselves.

The \LaTeX\ system provides a consistent and comprehensive document preparation
interface. Among other things, \LaTeX\ can automatically number list
entries, equations, figures, tables and footnotes, as well as sections and
subsections. Using this numbering system, bibliographic citations, page
references and cross references to any other numbered entity (e.g. sections,
equations, figures) are straightforward.

\subsection{The JFP document class}

The use of document classes allows a simple change of style (or style option)
to transform the appearance of your document. The JFP class preserves the
standard \LaTeX\ interface such that any document which can be produced
using the standard \LaTeX\ \verb"article" class can also be produced with
the JFP class.
However, the measure (or width of text) is different from that
for ARTICLE; therefore line breaks will change and it is possible that
longer equations may need re-setting.
Authors are urged to use \verb"jfp1.cls" from the beginning of their document
preparation, otherwise longer lines may require re-formatting at a later stage.

\subsection{General style issues}

Use of \LaTeX\ defaults will result in a pleasing uniformity of layout
and font selection. Authors should resist the temptation to make
\emph{ad hoc} changes to these. Also avoid use of direct formatting unless
really necessary. Papers will be edited as usual, and this process may be
obstructed by the use of inserted line breaks, etc.

For general style issues, authors are referred to the `Preparation of
manuscripts' in the back cover of the journal. Authors who are interested in
the details of style are referred to \cite{Butcher} and \cite{Chicago}. The
language used in the journal is British English, and spelling should conform
to this.

Use should be made of symbolic references (\verb"\ref") in order to
protect against late changes of order, etc.

\subsection{Submission of \LaTeX\ articles}

Authors who intend to submit a \LaTeX\ article should obtain a copy of the
JFP class file. This is available by anonymous FTP from
%
\begin{verbatim}
  ftp.cup.cam.ac.uk
\end{verbatim}
%
You will find the class file and instructions contained in a
in the directory
%
\begin{verbatim}
  pub/texarchive/journals/latex/jfp-cls
\end{verbatim}
%
The \verb"readme.txt" (which is the same directory) gives more details.

When submitting the final article, ensure that  the following are included and
are clearly labelled.
\begin{enumerate}
  \item A hardcopy printout of the article.
  \item The input file (exactly matching the hardcopy).
  \item A copy of any user-defined macros.
  \item If you have used \textsc{Bib}\TeX, the \verb".bib", \verb".bbl"
        and \verb".bst" files that were used.
  \item Any other files necessary to prepare the article for typesetting.
\end{enumerate}
The source files for the \emph{final} article should be text-only with no
system-dependent control codes, via email as an attachment, along with
a pdf file that matches the source files exactly.

\section{Using the JFP1 class file}
\label{usingjfpclass}

First, copy the file \verb"jfp1.cls" (and \verb"jfp.bst" if you use Bib\TeX)
into an appropriate subdirectory on your system. The JFP class is implemented
as a complete document class, and \emph{not} as an class option.
In order to use the JFP class, replace \verb"article" by \verb"jfp1" in the
\verb"\documentclass" command at the beginning of your document: that is,
%
\begin{verbatim}
  \documentclass{article}
\end{verbatim}
%
is replaced by
%
\begin{verbatim}
  \documentclass{jfp1}
\end{verbatim}
%
Author-defined macros should be inserted before \verb"\begin{document}",
or in a separate file and should be included with the submission.
Authors must not change any of the macro definitions or parameters
in \verb"jfp1.cls".

If you have a document prepared using an old \verb"jfp.sty" file,
just change the line
\begin{verbatim}
  \documentstyle{jfp}
\end{verbatim}
to
\begin{verbatim}
  \documentclass{jfp1}
\end{verbatim}
The latter form uses the \verb"jfp1.cls" file.

\subsection{Document class options}\label{sec:ClassOp}

In general, the following standard document class options should \emph{not} be
used with the JFP class file:
%
\begin{itemize}
  \item \texttt{10pt}, \texttt{11pt} and \texttt{12pt} -- unavailable;
  \item \texttt{twoside} is the default (\texttt{oneside} is disabled);
  \item \texttt{onecolumn} is the default (\texttt{twocolumn} is disabled);
  \item \texttt{titlepage} is not required and is disabled;
  \item \texttt{fleqn} and \texttt{leqno} should not be used, and are disabled.
\end{itemize}
%
\ifprodtf
The following new class options are provided:
\begin{itemize}
  \item \texttt{prodtf} -- tells the class file that we want to use the
    production typeface. This automatically sets the odd, even and top
    margins.
\end{itemize}
\fi

\section{Additional facilities}

In addition to all the standard \LaTeX\ design elements, the JFP class
includes the following features.
%
\begin{itemize}
  \item Additional commands for typesetting the title page. Extended
        commands for specifying a short version of the title and author(s)
        for the running headlines.
  \item New options to the \verb"\maketitle" command to create Functional and
        Theoretical Pearl(s).
  \item A \verb"proof" environment.
  \item A \verb"capsule" environment for typesetting Capsule Reviews.
  \item Control of enumerated lists.
\end{itemize}
%
Once you have used these additional facilities in your document,
it can be processed only with \verb"jfp1.cls".

\subsection{Titles, authors' names and running headlines}

At the beginning of your article, the title should be generated in the
usual way using the \verb"\maketitle" command. Immediately following the
title you may include an abstract and/or capsule review. For example, the titles
for this guide were produced by the following source.
%
\begin{verbatim}
  \title[Journal of Functional Programming]
        {\LaTeXe\ guide for the Journal of Functional Programming}
  \author[M. A. Reed]
         {MARK A. REED\\
          Cambridge University Press, Cambridge CB2 2BS, UK\\
          \email{texline@cup.cam.ac.uk}}
\end{verbatim}

\verb"\begin{document}"\\
\indent\verb"\maketitle"

\begin{verbatim}
  \begin{abstract}
  This guide is for authors who are preparing papers...
  \end{abstract}
\end{verbatim}

In the JFP1 class, the title of the article and the author's name (or
authors' names) are used both at the beginning of the article for the main
title and throughout the article as running headlines at the top of every
page. The title is used on odd-numbered pages (rectos) and the author's name
appears on even-numbered pages (versos). The \verb"\pagestyle" and
\verb"\thispagestyle" commands should \emph{not} be used.  Similarly, the
commands \verb"\markright" and \verb"\markboth" should not be necessary.

Although the article title can run to several lines of text, the running
headline must be a single line. Moreover, the title can incorporate new-line
commands (e.g. \verb"\\"), but these are not acceptable in a running headline.
To enable you to specify an alternative short title, and an alternative short
author's name, the standard \verb"\title" and \verb"\author" commands have
been extended to take an optional argument to be used as the running headline.
%
\begin{verbatim}
  \title[Short title]
        {Full title which can be as long as necessary}
  \author[Author name]
         {AUTHOR NAME \\ Affiliation}
\end{verbatim}
%
Notice that the author name in the argument for the running head should
be in mixed case, and the author name for the title should be in upper
case only. The author affiliation is set in the normal way, after a
\verb"\\" in the argument to the \verb"\author" command.

Any `work supported by' or `authors current address' information should
be inserted via \verb"\thanks" commands, which should be positioned
after the appropriate `AUTHOR NAME' in the \verb"\author" command.

If there are four (or more) authors for the article, the author running head
should contain the first author name followed by `et~al.' only. e.g.
%
\begin{verbatim}
  \author[Author1 et al.]
         {AUTHOR1...}
\end{verbatim}

The previous examples show an article with one author, the normal
\LaTeX\ conventions have been extended to allow the author names and
their affiliations to be typeset in the correct JFP style. The following
examples should cover most possibilities:

\textit{Case 1.} Two authors with the same affiliation:
%
\begin{verbatim}
  \author[Author1 and Author2]
         {AUTHOR1 and AUTHOR2\\
          Affiliation for both authors}
\end{verbatim}
%
If the author names are too long to fit onto one line, it should be
broken into two or more lines using the \verb"\authorbreak" command.
Don't use \verb"\\" to linebreak the author names -- as this will not do what
you expect.

\textit{Case 2.} Two authors with different affiliations:
%
\begin{verbatim}
  \author[Author1 and Author2]
         {AUTHOR1\\
          Affiliation for Author1
          \and AUTHOR2\\
          Affiliation for Author2}
\end{verbatim}

\textit{Case 3.} Three (or more) authors, two with the same affiliation:
%
\begin{verbatim}
  \author[Author1, Author2 and Author3]
         {AUTHOR1, AUTHOR2\\
            Affiliation for Author1 and Author2
          \and AUTHOR3\\
            Affiliation for Author3}
\end{verbatim}

\subsection{Pearls}

To provide for the Theoretical Pearl and Functional Pearl styles,
\verb"\maketitle" can now take an optional argument, which can take the
values \verb"t" and \verb"f". Thus, \verb"\maketitle[t]" will make a
title page, and set up the running headline, for a \linebreak Theoretical
Pearls article; \verb"\maketitle[f]" will do the same for a Functional Pearls
article.
If you need to have a `Theoretical Pearl' or `Functional Pearl' (singular)
you can use the \verb"T" and \verb"F" options to obtain these.

\subsubsection{Other `Pearl' styles}

If you need to have a `Pearl' style which cannot be generated by the
standard \verb"t", \verb"f", \verb"T" or \verb"F" options, then the \verb"o"
option can be used.

The \verb"o" option allows you to define a custom `Pearl' style. \eg for
`CUSTOM PEARL' you would use:
%
\begin{verbatim}
  \begin{document}
  \renewcommand\otherpearl{C\ls U\ls S\ls T\ls O\ls M\ns
                        P\ls E\ls A\ls R\ls L}
  \maketitle[o]
  \shorttitle{Custom pearl}
\end{verbatim}
%
Notice the use of the \verb"\ls" and \verb"\ns" commands to letter-space the
words in the title.
You can use the \verb"\\" command to line-break the title as normal.
The \verb"\shorttitle" command should appear \emph{after} the
\verb"\maketitle" command (as above).

\subsection{Proofs}

A new environment exists for creating Proofs, e.g.
%
\begin{proof}
  Use $K_\lambda$ and $S_\lambda$ to translate combinators
  into $\lambda$-terms. For the converse, translate
  $\lambda x$ \ldots by [$x$] \ldots and use induction
  and the lemma.
\end{proof}
%
This was produced by the following code:
%
\begin{verbatim}
  \begin{proof}
    Use $K_\lambda$ and $S_\lambda$ to...
  \end{proof}
\end{verbatim}
%
The end of proof marker \proofbox\ is produced automatically. If you wish
to omit this, use the \verb"proof*" environment instead.

If a proof ends with a display equation, then it is customary for the proofbox
to be positioned at the end of equation finishing the proof. \eg
%
\begin{proof*}
  Use $K_\lambda$ and $S_\lambda$ to translate combinators
  into $\lambda$-terms. For the converse, translate
  $\lambda x$ \ldots by [$x$] \ldots and use induction
  and the lemma.
  \[ a_1 \equiv (2\Omega M^2/x) \mathproofbox \]
\end{proof*}
%
Was produced with:
%
\begin{verbatim}
  \begin{proof*}
    Use $K_\lambda$ and $S_\lambda$ to...
    \[ a_1 \equiv (2\Omega M^2/x) \mathproofbox \]
  \end{proof*}
\end{verbatim}
%
Notice the use of \verb"proof*" to turn off the automatic proofbox.

The proof environment will also take an optional argument which allows you
to produce `special' proofs. e.g.
%
\begin{proof}[Proof of Theorem 27]
  We define a linear isometry $A:H\rightarrow H$ by the nonlinear
  Schr\"{o}dinger equation. It would not be hard to modify the
  proof to obtain an analogous result for ellipsoids rather than
  spheres.
\end{proof}
%
Which was produced like this:
%
\begin{verbatim}
  \begin{proof}[Proof of Theorem 27]
    We define a linear isometry...
  \end{proof}
\end{verbatim}
%
Notice that once the optional argument is used, you have to type all of
the text which is to appear as the heading.

\subsection{Programs}\label{sec-programs}

JFP encourages authors to use one of two styles for typesetting programs,
\emph{mathematical} and \emph{verbatim}.

A program typeset in the mathematical style is shown in
Figure~\ref{mathfigure}, and the commands used to typeset this program
are shown in Figure~\ref{mathtypeset}.  This uses the ordinary
mathematics mode of \LaTeX: displayed programs are surrounded by
\verb|\[| and \verb|\]|, and the \verb|array| command is used for
alignment.  However, there are two important differences.  First, the
\verb|\programmath| command appears before the program text; this
causes math mode to use ordinary spacing for italic identifiers,
rather than math spacing.  The \verb|\unprogrammath| command returns
to normal math spacing.  Second, in math mode spaces are ignored, so a
tilde \verb|~| is used instead.  (In \LaTeX, a tilde generates a
``hard space'' that is never replaced by a line break.)  To include
program text in mathematics style inline, surround it with dollar
\verb|$| signs.  For example, the input
\begin{verbatim}
  See how \programmath $differ~x$
  differs from \unprogrammath $differ x$.
\end{verbatim}
produces the output
\begin{center}
  See how \programmath $differ~x$
  differs from \unprogrammath $differ x$.
\end{center}

A program typeset in the verbatim style is shown in
Figure~\ref{verbfigure}, and the commands used to typeset this program
are shown in Figure~\ref{verbtypeset}.  This uses the ordinary
verbatim mode of \LaTeX: displayed programs are surrounded by
\verb|\begin{verbatim}| and \verb|\end{verbatim}|, and alignment is
indicated with spaces in the source file (don't use tabs, which may
not be processed properly).  To include program
text in verbatim style inline, use the \verb|\verb| command.
For example, the input
\begin{center}
\begin{verbatim}
  On a terminal, this looks like \verb"differ x".
\end{verbatim}
\end{center}
produces the output
\begin{center}
  On a terminal, this looks like \verb"differ x".
\end{center}

It is recommended that programs in figures be offset from the text
using the \verb|\figrule| command, as shown in Figures~\ref{mathfigure}%
--\ref{verbtypeset}.

\begin{figure}
\figrule
\programmath
\[
\begin{array}{lcl}
exp                   & :: & Exp \rightarrow Arr \rightarrow Val \\
exp~(Var~i)~a         & =  & index~i~a \\
exp~(Const~v)~a       & =  & v \\
exp~(Plus~e_1~e_2)~a  & =  & exp~e_1~a+exp~e_2~a \\
\\
com                   & :: & Com \rightarrow Arr \rightarrow Arr \\
com~(Asgn~i~e)~a      & =  & update~i~(exp~e~a)~a \\
com~(Seq~c_1~c_2)~a   & =  & com~c_2~(com~c_1~a) \\
com~(If~e~c_1~c_2)~a  & =  & \textrm{if } exp~e~a \dequals 0 \textrm{ then }
                                      com~c_1~a \textrm{ else } com~c_2~a \\
\\
prog                  & :: & Prog \rightarrow Val\\
prog~(Prog~c~e)       & =  & exp~e~(com~c~(newarray~0))
\end{array}
\]
\unprogrammath
\caption{Example program in mathematical style.}\label{mathfigure}
\figrule
\begin{center}
\begin{verbatim}
\begin{figure}
\figrule
\programmath
\[
\begin{array}{lcl}
exp                   & :: & Exp \rightarrow Arr \rightarrow Val \\
exp~(Var~i)~a         & =  & index~i~a \\
exp~(Const~v)~a       & =  & v \\
exp~(Plus~e_1~e_2)~a  & =  & exp~e_1~a+exp~e_2~a \\
\\
com                   & :: & Com \rightarrow Arr \rightarrow Arr \\
com~(Asgn~i~e)~a      & =  & update~i~(exp~e~a)~a \\
com~(Seq~c_1~c_2)~a   & =  & com~c_2~(com~c_1~a) \\
com~(If~e~c_1~c_2)~a  & =  & \textrm{if } exp~e~a \dequals 0 \textrm{ then }
                                      com~c_1~a \textrm{ else } com~c_2~a \\
\\
prog                  & :: & Prog \rightarrow Val\\
prog~(Prog~c~e)       & =  & exp~e~(com~c~(newarray~0))
\end{array}
\]
\unprogrammath
\caption{Example program in mathematical style.}\label{mathfigure}
\figrule
\end{verbatim}
\end{center}
\caption{Typesetting the example program in mathematical style.}\label{mathtypeset}
\figrule
\end{figure}

\begin{figure}
\figrule
\begin{verbatim}
  exp                 :: Exp -> Arr -> Val
  exp (Var i) a       =  index i a
  exp (Const v) a     =  v
  exp (Plus e1 e2) a  =  exp e1 a + exp e2 a

  com                 :: Com -> Arr -> Arr
  com (Asgn i e) a    =  update i (exp e a) a
  com (Seq c1 c2) a   =  com c2 (com c1 a)
  com (If e c1 c2) a  =  if  exp e a == 0  then  com c1 a  else  com c2 a

  prog                :: Prog -> Val
  prog (Prog c e)     =  exp e (com c (newarray 0))
\end{verbatim}
\caption{Example program in verbatim style.}\label{verbfigure}
\figrule
\begin{tabular}{l}
\verb|\begin{figure}| \\
\verb|\figrule| \\
\verb|\begin{center}| \\
\verb|\begin{verbatim}| \\
\verb|exp                 :: Exp -> Arr -> Val| \\
\verb|exp (Var i) a       =  index i a| \\
\verb|exp (Const v) a     =  v| \\
\verb|exp (Plus e1 e2) a  =  exp e1 a + exp e2 a| \\
\\
\verb|com                 :: Com -> Arr -> Arr| \\
\verb|com (Asgn i e) a    =  update i (exp e a) a| \\
\verb|com (Seq c1 c2) a   =  com c2 (com c1 a)| \\
\verb|com (If e c1 c2) a  =  if  exp e a == 0  then  com c1 a  else  com c2 a| \\
\\
\verb|prog                :: Prog -> Val| \\
\verb|prog (Prog c e)     =  exp e (com c (newarray 0))| \\
\verb|\end{verbatim}| \\
\verb|\end{center}| \\
\verb|\caption{Example program in verbatim style.}\label{verbfigure}| \\
\verb|\figrule| \\
\verb|\end{figure}|
\end{tabular}
\caption{Typesetting the example program in verbatim style.}\label{verbtypeset}
\figrule
\end{figure}

Some new macros have been provided for a few convenient symbols in
math mode.  These are illustrated in Table~\ref{newcommands}.
\ifprodtf\newpage\fi

\begin{table}
\caption{New symbol macros}\label{newcommands}
\programmath
\begin{tabular}{lll}
\hline\hline
Symbol                & Usage                  & Keyed as\\
\hline
\verb"\dplus"         & $abc \dplus xyz$         & \verb"$abc \dplus xyz$"\\
\verb"\dequals"       & $abc \dequals xyz$       & \verb"$abc \dequals xyz$"\\
\verb"\dcolon"        & $abc \dcolon xyz$        & \verb"$abc \dcolon xyz$"\\
\verb"\dcolonequals"  & $abc \dcolonequals xyz$  & \verb"$abc \dcolonequals xyz$"\\
\hline\hline
\end{tabular}
\unprogrammath
\end{table}

\subsection{Abstract and Capsule reviews}

The JFP class provides for an abstract and/or a capsule review; the
abstract is produced by the following commands:
%
\begin{verbatim}
  \begin{abstract}
       :
  \end{abstract}
\end{verbatim}
%
whereas the capsule review is produced by:
%
\begin{verbatim}
  \begin{capsule}
       :
  \end{capsule}
\end{verbatim}
%
Either or both of these may be used, in either order, but it is assumed
that, if both are used, there will be no other material between them.
In JFP the abstract should precede any capsule review.

\subsection{Lists}

The JFP class provides the three standard list environments.
\begin{itemize}
  \item Numbered lists, created using the \verb"enumerate" environment;
  \item Bulleted lists, created using the \verb"itemize" environment;
  \item Labelled lists, created using the \verb"description" environment.
\end{itemize}
The \verb"enumerate" environment numbers each list item with an arabic numeral;
alternative styles can be achieved by inserting a redefinition of the
number labelling command after the \verb"\begin{enumerate}". For example, a
list numbered with roman numerals inside parentheses can be produced by the
following commands:
%
\begin{verbatim}
  \begin{enumerate}[(iii).]
    \renewcommand{\theenumi}{(\roman{enumi})}
    \item first item
          :
  \end{enumerate}
\end{verbatim}
%
This produces the following list:
%
\begin{enumerate}[(iii).]
  \renewcommand{\theenumi}{(\roman{enumi})}
  \item first item
  \item second item
  \item \etc
\end{enumerate}
%
Notice that an optional argument ``\verb"(iii)."'' has been given to the
\verb"enumerate" environment, specifying the \emph{widest label} used in the
list. This is because roman numerals are wider than the arabic numerals
normally used by \verb"enumerate", and so the labels would otherwise have been
pushed out into the margin.

\section{User-defined macros}

If you define your own macros, you must ensure that their names do not
conflict with any existing macros in \LaTeX\ (or \AMSLaTeX\ %
if you are using this). You should also place them in the preamble of
your input file, between the \verb"\documentclass" (but after any
\verb"\usepackage" commands) and before the \verb"\begin{document}" command.

Apart from scanning the indexes of the relevant manuals, you can check
whether a macro name is already used by using \verb"\newcommand", which
will check for the existence of the macro you are trying to define.
If the macro exists \LaTeX\ will respond with:
%
\begin{verbatim}
  ! LaTeX Error: Command ... already defined.
\end{verbatim}
%
In this case you should choose another name, and try again.

Such macros must be in a place where they can easily be found and
modified by the journal's editors or typesetter. They must be gathered
together in the preamble of your input file, or in a separate
\verb"macros.tex" file with the command \verb"\input{macros}" in the
preamble. Macro definitions must not be scattered about your document
where they are likely to be completely overlooked by the typesetter.

The same applies to font definitions that are based on Computer Modern
fonts. These must be changed by the typesetter to use the journal's
correct typeface. In this case, you should draw
attention to these font definitions on the hard copy that you submit for
publication and by placing a comment in your input file just before the
relevant definitions, for example \verb"% replace font!"

\section{Some guidelines for using standard facilities}

The following notes may help you achieve the best effects with the JFP class file.

\subsection{Sections}

\LaTeX\ provides five levels of section headings and they are all
defined in the JFP class file:
\begin{itemize}
  \item[] Heading A -- \verb"\section{...}"
  \item[] Heading B -- \verb"\subsection{...}"
  \item[] Heading C -- \verb"\subsubsection{...}"
  \item[] Heading D -- \verb"\paragraph{...}"
  \item[] Heading E -- \verb"\subparagraph{...}"
\end{itemize}
Section numbers are given for sections, subsection and subsubsection headings.

\subsection{Figures and tables}

The \texttt{figure} and \texttt{table} environments are implemented as described in
the \LaTeX\ Manual to
provide consecutively numbered floating inserts for illustrations and tables
respectively.
The standard inserts and their captions are formatted centred.
Line breaks in captions can be inserted as required using \verb"\\".

\subsubsection{Illustrations (or figures)}

The JFP class will cope with most positioning of your illustrations
and you should not normally use the optional positional qualifiers on
the \verb"figure" environment which would override these decisions.
Figure captions should be below the figure itself, therefore the
\verb"\caption" command should appear after the figure or space left
for an illustration.

Figures in JFP will frequently illustrate programs, as shown in
section~\ref{sec-programs} of this guide. Figure~\ref{sample-figure}
shows an example of space left above a caption for artwork to be pasted in.
This was produced with the following commands:
%
\begin{verbatim}
  \begin{figure}
    \vspace{5cm} % the vertical depth of the artwork
    \caption{An example figure with space for artwork.}
    \label{sample-figure}
  \end{figure}
\end{verbatim}
%
\begin{figure}
  \vspace{5cm} % the vertical depth of the artwork
  \caption{An example figure with space for artwork.}
  \label{sample-figure}
\end{figure}
%
The vertical depth should correspond roughly to the artwork you will submit;
it will be adjusted to fit the final artwork exactly.

If your illustration extends over two~pages, you can use the
\verb"\continuedfigure" facility. To use this, you key the figure
caption for the second figure as follows:
%
\begin{verbatim}
  \begin{figure}
    \continuedfigure
    \vspace{80pt}
    \caption{First figure, continued.}
    \label{continued}
  \end{figure}
\end{verbatim}
%
This ensures that the figure counter does not get incremented, and at the
same time adds the word (cont.) to the caption. You may still use labels
and references for this figure.

\subsubsection{Tables}

The JFP class file will cope with most positioning of your tables
and you should not normally use the optional positional qualifiers on the
\verb"table" environment which would override these decisions.
Normal journal style sets the table caption first, followed by a double
rule, the table body and a double rule at the bottom. Single rules and
spanner rules (\verb"\cline") can be used to separate headings from the
columns. For example, Table~\ref{sample-table} is produced using the
following commands:\par
%
{\small
\begin{verbatim}
\begin{table}
  \caption{Results of Overloading for 3 Experimental Setups}
  \label{sample-table}
  \begin{minipage}{\textwidth}
    \begin{tabular}{lcrrrrr}
      \hline\hline
      Program& Expt.&
      CPU\footnote{Seconds of elapsed time on an unloaded Sun 3/50.}&
      RelCPU\footnote{CPU Time relative to experiment (a).}& GC&
      Mem\footnote{Bytes of heap used over the duration of the program.}&
      RelMem\footnote{Memory usage relative to experient (a).}\\
      \hline
      8 Queens& (a)&   2.88&  1.00&    6&   1.7M&  1.00\\
      &         (b)&  32.51& 11.29&  193&  48.9M& 28.76\\
      &         (c)&   7.90&  2.74&   42&  11.3M&  6.65\\
      \noalign{\vspace {.5cm}}
      Primes&   (a)&   4.89&  1.00&   19&   5.3M&  1.00\\
      &         (b)&  47.54&  9.72&  204&  54.5M& 10.28\\
      &         (c)&  10.08&  2.06&   47&  13.0M&  2.45\\
      \noalign{\vspace {.5cm}}
      Nfib&     (a)&  21.65&  1.00&  161&  40.4M&  1.00\\
      &         (b)& 221.65& 10.24& 1382& 349.0M&  8.64\\
      &         (c)&  21.30&  0.98&  161&  42.0M&  1.03\\
      \noalign{\vspace {.5cm}}
      KWIC&     (a)&   7.07&  1.00&   15&   6.3M&  1.00\\
      &         (b)&  34.55&  4.89&  109&  47.8M&  7.59\\
      &         (c)&  31.62&  4.47&   53&  45.0M&  7.14\\
      \hline\hline
    \end{tabular}
    \vspace{-2\baselineskip}
  \end{minipage}
\end{table}
\end{verbatim}}
%
\begin{table}
  \caption{Results of Overloading for 3 Experimental Setups}
  \label{sample-table}
  \begin{minipage}{\textwidth}
    \begin{tabular}{lcrrrrr}
      \hline\hline
      Program& Expt.&
      CPU\footnote{Seconds of elapsed time on an unloaded Sun 3/50.}&
      RelCPU\footnote{CPU Time relative to experiment (a).}& GC&
      Mem\footnote{Bytes of heap used over the duration of the program.}&
      RelMem\footnote{Memory usage relative to experient (a).}\\
      \hline
      8 Queens& (a)&   2.88&  1.00&    6&   1.7M&  1.00\\
      &         (b)&  32.51& 11.29&  193&  48.9M& 28.76\\
      &         (c)&   7.90&  2.74&   42&  11.3M&  6.65\\
      \noalign{\vspace {.5cm}}
      Primes&   (a)&   4.89&  1.00&   19&   5.3M&  1.00\\
      &         (b)&  47.54&  9.72&  204&  54.5M& 10.28\\
      &         (c)&  10.08&  2.06&   47&  13.0M&  2.45\\
      \noalign{\vspace {.5cm}}
      Nfib&     (a)&  21.65&  1.00&  161&  40.4M&  1.00\\
      &         (b)& 221.65& 10.24& 1382& 349.0M&  8.64\\
      &         (c)&  21.30&  0.98&  161&  42.0M&  1.03\\
      \noalign{\vspace {.5cm}}
      KWIC&     (a)&   7.07&  1.00&   15&   6.3M&  1.00\\
      &         (b)&  34.55&  4.89&  109&  47.8M&  7.59\\
      &         (c)&  31.62&  4.47&   53&  45.0M&  7.14\\
      \hline\hline
    \end{tabular}
    \vspace{-2\baselineskip}
  \end{minipage}
\end{table}

Notice the use of the `\verb"\vspace{-2\baselineskip}"' command to remove the
unwanted vertical space from above the table footnotes in this example.

Captions for `continued' tables can be generated (in the same way as for figures)
using the \verb"\continuedtable" command. These should be positioned just before
the \verb"\caption" command in the appropriate table environment.

The \verb"tabular" environment should be used to produce ruled tables;
it has been modified for the JFP class in the following ways:
\begin{enumerate}
  \item Additional vertical space is inserted above and below a horizontal rule
        (produced by \verb"\hline");
  \item Tables are centred, and span the full width of the page; that is,
  they are similar to the tables that would be produced by
  \verb"\begin{minipage}{\textwidth}".
\end{enumerate}
Because of this reformatting, vertical rules should not be used;
furthermore, commands to
redefine quantities such as \verb"\arraystretch" should be omitted. If
the old tabular facilities are needed, there is a new environment,
\verb"oldtabular", which has none of the reformatting; it should be used
in exactly the same way.

\subsection{Appendices}

You should use the standard \LaTeX\ \verb"\appendix" command to place any
Appendices, normally, just before any references.
From that point on \verb"\section" will produce an appendix, which are
numbered A, B etc., equations as (A1), (B1) etc.\ Figures and
tables also number A\,1, B\,1 etc.

\subsection{References}

As with standard \LaTeX, there are two ways of producing a list of references;
either by using Bib\TeX\ with the JFP bibliography style \verb"jfp.bst",
or by compiling a list of references by hand (using a
\verb"thebibliography" environment).

\subsubsection{Using Bib\TeX}

If you have Bib\TeX\ installed on your system, the following is a brief description
on how to automatically generate a bibliography (\verb".bbl" file) for your
article. Your article should contain at least the following elements:\par
\vspace{6pt}\noindent
%
\verb"  % sample.tex"\\[3pt]
\verb"  \documentclass{jfp}"\\[3pt]
\verb"  \bibliographystyle{jfp}"\\[3pt]
\verb"  \begin{document}"\\[3pt]
\verb"    \cite{"\textit{citations}\verb"}"\\[3pt]
\verb"    \bibliography{"\textit{biblio database files}\verb"}"\\[3pt]
\verb"  \end{document}"\\[6pt]
%
Where `\emph{biblio database files}' may be one or more filenames of bibliographic
database files (without the \verb".bib" extension) separated by commas.
First, \LaTeX\ the file \verb"sample.tex". Second, run Bib\TeX\ by typing:
%
\begin{verbatim}
  bibtex sample
\end{verbatim}
%
This creates the file \verb"sample.bbl". Third, re-\LaTeX\ your
document, and the newly-created \verb"sample.bbl" will be read in
and typeset. You will then need to \LaTeX\ the document once more to resolve any
unresolved citation references.

\subsubsection{Typesetting the references by hand}

The following listing shows some references prepared in the style of the
journal; this code produces the references at the end of this guide.
%
\begin{verbatim}
\begin{thebibliography}{}
 \bibitem[\protect\citename{Augustsson and Johnsson, }1987]{AJ187}
   Augustsson,~L. and Johnsson,~T. (1987) LML users' manual. PMG
   Report, Department of Computer Science, Chalmers University of
   Technology, Goteborg, Sweden.
 \bibitem[\protect\citename{Butcher, }1981]{Butcher}
   Butcher,~J. (1981) Copy-editing: the Cambridge handbook.
   Cambridge University Press.
 \bibitem[\protect\citename{Chicago, }1982]{Chicago}
   The Chicago manual of style. University of Chicago Press.
 \bibitem[\protect\citename{Conklin, }1987]{JC87}
   Conklin,~J. (1987) Hypertext: an introduction and survey.
   \emph{IEEE Computer}, 20~(9): pp.~17--41.
 \bibitem[\protect\citename{Dijkstra, }1976]{EWD76}
   Dijkstra,~E.~W. (1976) \emph{A Discipline of Programming}.
   Prentice-Hall.
 \bibitem[\protect\citename{Knuth, }1984]{DEK84}
   Knuth,~D.~E. (1984) Literate programming. \emph{BCS Comput. J.}
   27~(2): 97--111 (May).
 \bibitem[\protect\citename{Lamport, }1986]{LaTeX}
   Lamport,~L. (1986) \LaTeX: a document preparation system
   (2nd edition). Addison-Wesley, New York.
 \bibitem[\protect\citename{Reynolds, }1969]{JCR69}
   Reynolds,~J.~C. (1969) Transformation systems and the
   algebraic structure of atomic formulas. In B. Meltzer and
   D. Michie (editors), \emph{Machine Intelligence 5}, pp.~135--151.
   Edinburgh University Press.
 \bibitem[\protect\citename{Toyn \emph{et al.}, }1987]{TDR87}
   Toyn,~I., Dix,~A. and Runciman,~C. (1987) Performance
   polymorphism. In \emph{Functional Programming Languages and
   Computer Architecture, Lecture Notes in Computer Science, 274},
   pp.~325--346. Springer-Verlag.
\end{thebibliography}
\end{verbatim}
%
The above list is typeset at the end of this guide.
Each entry takes the form
%
\begin{verbatim}
\bibitem[\protect\citename{Author(s), }Date]{tag}
  Bibliography entry
\end{verbatim}
%
where \verb"Author(s)"\ should be the author names as they are cited in
the text (note the space before the closing \verb"}" of the \verb"\citename"
command is vital),
\verb"Date" is the date to be cited in the text, and \verb"tag"
is the tag that is to be used as an argument for the \verb"\cite{}" and
\verb"\shortcite{}" commands. \verb"Bibliography entry" should be the
material that is to appear in the bibliography, suitably formatted. This
rather unwieldy scheme makes up for the lack of an author-date system in
\LaTeX.

\subsubsection{Multiple references}

References should be listed alphabetically by author name(s) and
then by year if the same author has several papers. If some papers by
the same author(s) also fall in the same year, their dates should be
in the form (1993a), (1993b), \etc

Formatting for italic etc.\ should be avoided unless you are sure you
understand the style of references; please concentrate on giving full and
clear information.

\subsubsection{References in the text}

References in the text are given by author and date. Whichever method
is used to produce the bibliography, the references in
the text are done in the same way. Each bibliographical entry has a key,
which is assigned by the author and used to refer to that entry in the
text.
There is one form of citation -- \verb"\cite{key}" -- to produce the
author and date, and another form -- \verb"\shortcite{key}" -- which
produces the date only. Thus,
Augustsson and Johnsson \shortcite{AJ187} is produced by
%
\begin{verbatim}
  Augustsson and Johnsson \shortcite{AJ187},
\end{verbatim}
%
while \cite{TDR87} is produced by
%
\begin{verbatim}
  \cite{TDR87}.
\end{verbatim}

To allow further flexibility of the \verb"\cite" (and \verb"\shortcite") commands
the function of the optional argument has been changed. For example, the output of
the command \verb"\cite{TDR87}" gives \cite{TDR87}, but you may want to list all
of the author names (instead of just the first author and \emph{et~al.}).
In this case we would use:
\[ \hbox{\verb"\cite[(Toyn, Dix and Runciman, 1987)]{TDR87}"} \]
in the text to produce the desired effect.

\newpage
\appendix
\section{Special commands in {\mdseries\texttt{jfp1.cls}}}

The following is a summary of the new commands, optional arguments and
environments which have been added to the standard \LaTeX\ user-interface
in creating the JFP class file.\par
\vspace{8pt}

\bcmdtab
\emph{New commands}    & \\[3pt]
\verb"\authorbreak"    & allows a list of authors to be broken in to separate
                         lines without starting an affiliation.\\
\verb"\continuedfigure" & adds the word (cont.) to the next figure caption, and
                          also stops the figure counter from being stepped.\\
\verb"\continuedtable"  & as for \verb"\continuedfigure", except this command
                          achieves the same effect for tables.\\
\verb"\email"          & used to typeset an authors e-mail address (should only be
                         used in the \verb"\author" command).\\
\verb"\figrule"        & adds a rule, used around programs set in figure environments.\\
\verb"\ls", \verb"\ns" & used to letter-space the heading in a alternate `Pearl'
                         style (when used with the \verb"\maketitle[o]" argument).\\
\verb"\programmath"    & gives normal spacing for verbatim math mode.\\
\verb"\proofbox"       & typesets a proof box \proofbox\ (this is normally
                         put in automatically at the end of the \verb"proof"
                         environment). If you need to insert a \verb"\proofbox"
                         manually, you should add a `\verb"\quad"' of
                         space before it in the output.\\
\verb"\removebrackets" & removes the `(\ )' brackets from the optional
                          argument of environments created by the
                          \verb"\newtheorem" command. Should be placed
                          just before the appropriate environment.\\
\verb"\unprogrammath"  & reverts to math spacing.\\
\ifprodtf
\verb"\removefullpoint" & removes the full point from the next \verb"\caption"
                          command, usually used for figures when they are split
                          from their captions.\\
\fi
\verb"\shortcite"      &  typesets the `year' part of the bibliographic entry only.
                          \eg (1987).\\
\verb"\mathproofbox"   & as \verb"\proofbox", except this version is intended
                         for use in equations ending proof's (it typesets the
                         proof box using \verb"\rlap" with 1em of extra space).\\[6pt]
\emph{New environments} & \\[3pt]
\ifprodtf
\verb"bottomfigure"     & for split figures and captions (on facing page).\\
\fi
\verb"capsule"          & used to typeset an articles Capsule Review.\\
\verb"oldtabular"       & preserves the orginal tabular environment, which
                          has been modified to insert additional space above
                          and below an \verb"\hrule". The body of the environment
                          is centred with rules full out across the text measure.\\
\verb"proof"            & to typeset mathematical proofs, the \verb"*"-form omits
                          the proof box.
\ecmdtab \newpage \bcmdtab
\emph{New optional arguments} & \\[3pt]
\verb"[<short title>]"  & in the \verb"\title" command: to define a right
                          running headline that is different from the article
                          title. The \verb"\shorttitle" command also
                          achieves the same effect.\\
\verb"[<short author>]" & in the \verb"\author" command: to define a left
                          running headline that is different from the
                          authors' names as typeset at the article opening.
                          The \verb"\shortauthor" command also
                          achieves the same effect.\\
\verb"[<pearl style>]"  & Optional argument to the \verb"\maketitle" command
                          which allows Functional or Theoretical pearls to be
                          typeset.\\
\verb"[<widest label>]" & in \verb"\begin{enumerate}": to ensure the correct
                          alignment of numbered lists.
\ecmdtab

\ifprodtf
%
\section{Notes for editors}

This appendix contains additional information which may be useful to
those who are involved with the final production stages of an article.
Authors, who are generally not typesetting the final pages in the
journal's typeface (Monotype Times), do not need this information.

\subsection{Setting the production typeface}

The global \verb"\documentclass" option `\verb"prodtf"' sets up
\verb"jfp1.cls" to typeset in the production typeface -- Monotype Times.
e.g.
%
\begin{verbatim}
  \documentclass[prodtf]{jfp1}
\end{verbatim}

\subsection{Catchline commands}

To be placed in the preamble:
\begin{itemize}
  \item \verb"\jdate{September 2001}"
  \item \verb"\pagerange{35--48}"
  \item \verb"\pubyear{2001}"
  \item \verb"\volume{\textbf{8} (1):}"
  \item \verb"\doi{S0956796801004857}"
\end{itemize}

\subsection{Landscape material}

The add on package \verb"JFPland" provides macros for landscape figures and
tables. See the \verb"JFPland" guide for further information.

\subsection{Figures with split artwork/captions}

When a figure is too large to fit on a page with its caption, you can use
the following procedure to place the figure, and then its caption at the foot
of the facing page. First set the figure with a short caption
using the normal \verb"figure" environment, e.g.
\begin{verbatim}
  \begin{figure}
    ...
    \caption{For caption see facing page.}
  \end{figure}
\end{verbatim}
Then set the correct (long) caption, so that it appears on the facing page:
\begin{verbatim}
  \begin{bottomfigure}
    \addtocounter{figure}{-1}
    \caption{This is the long caption...}
  \end{bottomfigure}
\end{verbatim}
If the figure falls on a recto, you may have to move the \verb"bottomfigure"
environment to before the \verb"figure" environment. In this case you need to
move the \verb"\addtocounter" command into the \verb"figure" environment
instead.

The \verb"bottomfigure" environment places a full measure rule above the
bottom-caption automatically.

\subsection{Editing citations (when the author has used the
 {\mdseries\textup{\texttt{cite}}} command)}

In the past when an automatic \verb"\cite" command produced text in the output
which needed to be changed, the argument (in [ ]) from the bibliography entry
was copied to the location of the \verb"\cite" command and then modified.
The \verb"\cite" command would then be removed as part of this process.

In the near future, we will probably have to supply \TeX\ output which will
need to contain `PDF marks' for interactive browsing.  Clearly by removing
the automatic link to the bibliographic entry (referenced by the \verb"\cite"),
we are making extra work for ourselves later on.

To avoid this, the function of the \verb"\cite" command's optional argument
has been changed. For example, the \verb"\cite" command for the
`\verb"JC87"' entry gives:
\[ \hbox{\cite{JC87}} \]
but you want the following to appear in the text:
\[ \hbox{\cite[(Conklin, 1987 -- see p.~22)]{JC87}} \]
you would then use:
\[ \hbox{\verb"\cite[(Conklin, 1987 -- see p.~22)]{JC87}"} \]
to obtain the desired result. Notice that you have to supply
the round brackets as well in the optional argument.
This modification also works for the \verb"\shortcite" command.

\section{Macros provided by {\mdseries\texttt{JFP2esym.sty}}}

\subsection{Automatic font/character changes}

\begin{itemize}\itemsep=6pt
\item The \verb|\le|, \verb|\leq|, \verb|\ge|, \verb|\geq| commands
use the equivalent AMS slanted symbols:
\[
\oldle \oldleq \oldge \oldgeq
 \lra
\le \leq \ge \geq
\]
The normal characters can be obtained by using the \verb|\old| form
(\eg \verb|\oldge|).
\end{itemize}

\subsection{Additional fonts}

\begin{itemize}\itemsep=6pt
\item The complete (v1) AMS symbols are available using the normal names:
\[
\hbox{\verb"\boxdot \boxplus \boxtimes"} \lra
  \boxdot \boxplus \boxtimes
\]

\item Blackboard bold:
\[
\hbox{\verb"$\mathbb{ABC}$"} \lra \mathbb{ABC}
\]

\item Fraktur/Gothic (bold math version available):
\[
   \hbox{\verb"$\mathfrak{ABC}$"} \lra \mathfrak{ABC}
\]

\item Bold math italic/symbols are provided by the \verb"\boldsymbol" macro
(from the \verb"amsbsy" package). The \verb"\bmath" macro is provided as
an alias.

\strut\hspace*{11pt}The \verb"JFP2esym" package also defines most of the
symbols from Appendix F of the \TeX book. These can be obtained by using
their normal (unbold) symbol name prefixed with a `b'. \eg \verb|\nabla|
becomes \verb|\bnabla|. The only exception to this rule is \verb|\eta|,
which whould lead to a clash with \verb|\beta|. In this case use
\verb|\boldeta| for bold eta.

\strut\hspace*{11pt}The problem with space disappearing around certain
bold math symbols (\verb"\bcdot") does not happen, when using the
\verb"amsbsy" package.

\item Sans serif symbols\\[6pt]
%
\verb"  \textsf{text}  " \lra \textsf{text}
  \qquad \verb"\mathsf{math}  " \lra $\mathsf{math}$\\
\verb"  \textsfi{text} " \lra \textsfi{text}
  \qquad \verb"\mathsfi{math} " \lra $\mathsfi{math}$\\
\verb"  \textsfb{text} " \lra \textsfb{text}
  \qquad \verb"\mathsfb{math} " \lra $\mathsfb{math}$\\
\verb"  \textsfbi{text}" \lra \textsfbi{text}
  \qquad \verb"\mathsfbi{math}" \lra $\mathsfbi{math}$
\end{itemize}

\subsection{Additional symbols}

\verb"  \colonequal   " \rch \colonequal \qquad \verb"\equalcolon     " \rch \equalcolon\\
\verb"  \colonequiv   " \rch \colonequiv \qquad \verb"\equivcolon     " \rch \equivcolon
%
\fi

\begin{thebibliography}{}
 \bibitem[\protect\citename{Augustsson and Johnsson, }1987]{AJ187}
   Augustsson,~L. and Johnsson,~T. (1987) LML users' manual. PMG
   Report, Department of Computer Science, Chalmers University of
   Technology, Goteborg, Sweden.
 \bibitem[\protect\citename{Butcher, }1981]{Butcher}
   Butcher,~J. (1981) Copy-editing: the Cambridge handbook.
   Cambridge University Press.
 \bibitem[\protect\citename{Chicago, }1982]{Chicago}
   The Chicago manual of style. University of Chicago Press.
 \bibitem[\protect\citename{Conklin, }1987]{JC87}
   Conklin,~J. (1987) Hypertext: an introduction and survey.
   \emph{IEEE Computer}, 20~(9): pp.~17--41.
 \bibitem[\protect\citename{Dijkstra, }1976]{EWD76}
   Dijkstra,~E.~W. (1976) \emph{A Discipline of Programming}.
   Prentice-Hall.
 \bibitem[\protect\citename{Knuth, }1984]{DEK84}
   Knuth,~D.~E. (1984) Literate programming. \emph{BCS Comput. J.}
   27~(2): 97--111 (May).
 \bibitem[\protect\citename{Lamport, }1986]{LaTeX}
   Lamport,~L. (1986) \LaTeX: a document preparation system
   (2nd edition). Addison-Wesley, New York.
 \bibitem[\protect\citename{Reynolds, }1969]{JCR69}
   Reynolds,~J.~C. (1969) Transformation systems and the
   algebraic structure of atomic formulas. In B. Meltzer and
   D. Michie (editors), \emph{Machine Intelligence 5}, pp.~135--151.
   Edinburgh University Press.
 \bibitem[\protect\citename{Toyn \emph{et al.}, }1987]{TDR87}
   Toyn,~I., Dix,~A. and Runciman,~C. (1987) Performance
   polymorphism. In \emph{Functional Programming Languages and
   Computer Architecture, Lecture Notes in Computer Science, 274},
   pp.~325--346. Springer-Verlag.
\end{thebibliography}

\label{lastpage}

\end{document}

% end of JFP2egui.tex

### 2019-04-14 papers/jfp/1-intro.tex
Minor revision to fit JFP style

@@
   built}~(\S\ref{sec-dependency-orderings}) and \emph{whether or not a
-  task is (re-)built}~(\S\ref{sec-out-of-date}). These choices turn out to
+  task is rebuilt}~(\S\ref{sec-out-of-date}). These choices turn out to
   be orthogonal, which leads us to a new classification of the design

### 2019-04-14 papers/jfp/2-background.tex
Minor revision to fit JFP style

@@
 
-\begin{figure}[h]
+\begin{figure}%[h]
 \begin{subfigure}[b]{0.32\linewidth}
@@
 \end{subfigure}
-\vspace{-2mm}
+% \vspace{-2mm}
 \caption{A task dependency graph and two build scenarios. Input files are shown
@@
 \label{fig-make}}
-\vspace{-2mm}
+% \vspace{-2mm}
 \end{figure}
@@
 If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
-will build~\cmd{util.o}~and~\cmd{main.o}, in any order since these tasks are
+will build~\cmd{util.o} and~\cmd{main.o}, in any order since these tasks are
 independent, and then \cmd{main.exe}. If the user modifies \cmd{util.h} and runs
@@
 \end{subfigure}
-\vspace{-2mm}
+% \vspace{-2mm}
 \caption{Dynamic dependencies example: create \cmd{README} and add it to the
 list of release documents \cmd{docs.txt}.\label{fig-shake}}
-\vspace{-4mm}
+% \vspace{-4mm}
 \end{figure}
@@
 \begin{figure}
-\vspace{-3mm}
+% \vspace{-3mm}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example-cutoff.pdf}}
-\vspace{-4mm}
+% \vspace{-4mm}
 \caption{An early cutoff example: if a comment is added to \cmd{main.c}, the
@@
 rebuilt.\label{fig-cutoff}}
-\vspace{-4mm}
+% \vspace{-4mm}
 \end{figure}
@@
 
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 \subsection{\Bazel: A Cloud Build System}\label{sec-background-bazel}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 
@@
 \begin{figure}
-\vspace{-2mm}
+% \vspace{-2mm}
 \begin{subfigure}[b]{0.25\linewidth}
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
-\vspace{-1mm}
-\caption{Download source files}
+% \vspace{-1mm}
+\caption{Download sources}
 \end{subfigure}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
-\vspace{-1mm}
+% \vspace{-1mm}
 \caption{Build \cmd{main.exe}}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
-\vspace{-1mm}
+% \vspace{-1mm}
 \caption{Modify \cmd{util.c} and rebuild}
 \end{subfigure}
-\vspace{-2.5mm}
+% \vspace{-2.5mm}
 \caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
@@
 rectangles.\label{fig-bazel}}
-\vspace{-3mm}
+% \vspace{-3mm}
 \end{figure}
@@
 \begin{table}[h]
-\vspace{-2mm}
+% \vspace{-2mm}
 \caption{Build system differences.\label{tab-summary}}
-\vspace{-4mm}
+% \vspace{-4mm}
 % \smaller
@@
 \end{tabular}
-\vspace{2mm}
+% \vspace{2mm}
 $^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
-\vspace{-8mm}
+% \vspace{-8mm}
 \end{table}
 
-\vspace{-1.5mm}
+% \vspace{-1.5mm}
 \subsection{Summary}\label{sec-background-summary}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 

### 2019-04-14 papers/jfp/6-engineering.tex
Minor revision to fit JFP style

@@
 \begin{subfigure}[b]{0.32\linewidth}
-\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-build.pdf}}
-\vspace{-1.5mm}
+\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-build.pdf}}
+% \vspace{-1.5mm}
 \caption{Initial build}
@@
 \begin{subfigure}[b]{0.32\linewidth}
-\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-clean.pdf}}
-\vspace{-1.5mm}
-\caption{Clean up and evict \cmd{main.prof}}
+\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-clean.pdf}}
+% \vspace{-1.5mm}
+\caption{Clean up, evict \cmd{main.prof}}
 \end{subfigure}
 \begin{subfigure}[b]{0.34\linewidth}
-\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-rebuild.pdf}}
-\vspace{-1.5mm}
-\caption{Build \cmd{main.prof} and \cmd{report.txt}}
+\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-rebuild.pdf}}
+% \vspace{-1.5mm}
+\caption{Build \cmd{main.prof} \& \cmd{report.txt}}
 \end{subfigure}
-\vspace{-1.5mm}
+% \vspace{-1.5mm}
 \caption{A Frankenbuild example: (a)~build a human-readable profiling report for
@@
 \label{fig-frankenbuild}}
-\vspace{-4mm}
+% \vspace{-4mm}
 \end{figure}

### 2019-04-18 papers/jfp/3-abstractions.tex
Minor revision to fit JFP style
Lift the contents of todo.md into the .tex and start to do the proposed splits

@@
 \end{figure}
+
+\todo{AM: A section with examples of functorial, applicative and monadic tasks based on
+  Collatz, Fibonacci and Ackermann sequences:
+  https://blogs.ncl.ac.uk/andreymokhov/the-task-abstraction/}
+
+\todo{AM: Maybe combine with the Edit Distance example from S7.3 Memoization?}
+
+\todo{Compute Functor = tail recursion
+Compute Applicative = primitive recursion
+Compute Monad = general recursion
+}
 
@@
 % \vspace{-1mm}
-\textbf{Definition}(Correctness). Suppose \hs{build} is a build system,
+\textbf{Definition}~(Correctness). Suppose \hs{build} is a build system,
 \hs{tasks} is a build task description, \hs{key} is a target key, \hs{store} is

### 2019-04-18 papers/jfp/4-schedulers.tex
Lift the contents of todo.md into the .tex and start to do the proposed splits

--- 2019-03-05 papers/jfp/4-build.tex
@@
-\section{Build Systems \`a la Carte}\label{sec-build}
+\section{Build Systems \`a la Carte Schedulers}\label{sec-build}
+
+\todo{Elaborate (it's very terse right now) with examples}
 
@@
 \cite{claessen_continuations} (what \Shake currently does).
-
-\subsection{The Rebuilder: Determining Out-of-date Keys}\label{sec-out-of-date}
-
-Suppose the scheduler decides that a key should be brought up to date. The next
-question is: does any work need to be done, or is the key already up to date?
-Or, in a cloud-build system, do we have a cached copy of the value we need?
-These questions can be addressed in one of four fundamental ways, with a number
-of tweaks and variations within them.
-
-\vspace{-2mm}
-\subsubsection{A Dirty Bit}\label{sec-dirty-bit}
-
-The idea of a dirty bit is to have one piece of persistent information per key,
-saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
-are set to clean. When the next build starts, anything that changed between the
-two builds is marked dirty. If a key and all its transitive dependencies are
-clean, the key does not need to be rebuilt.
-
-\Excel models the dirty bit approach most directly, having an actual dirty bit
-associated with each cell, marking the cell dirty if the user modifies it.
-It also marks dirty all cells that (transitively) depend on the modified cell.
-\Excel does not record dynamic dependencies of each cell; instead it computes a
-\emph{static over-approximation} -- it is safe for it to make more cells dirty
-than necessary, but not vice versa. The over-approximation is as follows: a cell
-is marked dirty if its formula statically refers to a dirty cell, or if the
-formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
-guessed from the formula alone. The over-approximation is clear for
-\cmd{INDIRECT}, but it is also present for \cmd{IF}, where both branches are
-followed even though dynamically only one is used.
-
-\Make uses file modification times, and compares files to their dependencies,
-which can be thought of as a dirty bit which is set when a file is older than
-its dependencies. The interesting property of this dirty bit is that it is not
-under the control of \Make; rather it is existing file-system information that
-has been repurposed. Modifying a file automatically clears its dirty bit, and
-automatically sets the dirty bit of the files depending on it (but not
-recursively). Note that \Make requires that file timestamps only go forward in
-time, which can be violated by backup software.
-
-With a dirty bit it is possible to achieve minimality. However, to achieve early
-cutoff (\S\ref{sec-background-shake}) it would be important to reset the dirty
-bit after a computation that did not change the value and make sure that cells
-that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
-because the dependent cells have already been recursively marked dirty. For
-\Make, it is impossible to mark a file clean and at the same time not mark the
-keys that depend on it dirty. \Make can approximate early cutoff by not
-modifying the result file, and not marking it clean, but then it will be rebuilt
-in every subsequent build.
-
-\vspace{-2mm}
-\subsubsection{Verifying Traces}\label{sec-verifying-traces}
-
-An alternative way to determine if a key is dirty is to record the
-values/hashes of dependencies used last time, and if something has changed, the
-key is dirty and must be rebuilt~--~in essence a \emph{trace} which we can use
-to \emph{verify} existing values. For traces, there are two essential
-operations~--~adding a new value to the trace store, and using the traces to
-determine if a key needs rebuilding. Assuming a store of verifying traces
-\hs{VT}~\hs{k}~\hs{v}, the operations are:
-
-\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
-recordVT@\,@::@\,@k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
-verifyVT@\,@::@\,@(Monad m,@\,@Eq k,@\,@Eq v) => k -> Hash v -> (k -> m (Hash v)) -> VT k v -> m Bool
-\end{minted}
-
-\noindent
-Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} stores
-only hashes, of type \hs{Hash}~\hs{v}, of those values. Since the verifying
-trace persists from one build to the next -- it constitutes the build system's
-``memory'' -- it is helpful for it to be of modest size. After successfully
-building a key, we call \hs{recordVT} to add a record to the current \hs{VT},
-passing the key, the hash of its value, and the list of hashes and dependencies.
-
-More interestingly, to \emph{verify} whether a key needs rebuilding we supply
-the key, the hash of its current value, a function for obtaining the post-build
-value of any key (using a scheduling strategy as per
-\S\ref{sec-dependency-orderings}), and the existing \hs{VT} information. The
-result will be a \hs{Bool} where \hs{True} indicates that the current value is
-already up to date, and \hs{False} indicates that it should be rebuilt.
-
-One potential implementation would be to record all arguments passed to
-\hs{recordVT} in a list, and verify by simply checking if any list item matches
-the information passed by \hs{verifyVT}.  We discuss smarter implementations
-in~\S\ref{sec-smart-traces}.
-
-A verifying trace, and other types of traces discussed in this section, support
-dynamic dependencies and minimality; furthermore, all traces except for deep
-traces~(\S\ref{sec-deep-constructive-traces}) support early cutoff.
-
-\subsubsection{Constructive Traces}\label{sec-constructive-traces}
-
-A verifying trace deliberately records only small hashes, so that it can be small.
-A \emph{constructive} trace additionally stores the resulting value.
-Once we are storing the complete result it makes sense
-to record many constructive traces per key, and to share them with other users,
-providing cloud-build functionality. We can represent this additional
-information by providing the operations:
-
-\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
-recordCT    :: k -> v -> [(k, Hash v)] -> CT k v -> CT k v
-constructCT :: (Monad m, Eq k, Eq v) => k -> (k -> m (Hash v)) -> CT k v -> m [v]
-\end{minted}
-
-\noindent
-The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
-the hash of the resulting value, we require the actual value. The \hs{verifyVT}
-has been replaced with \hs{constructCT}, which instead of taking the hash of the
-current value as \emph{input}, returns a list of suitable values as \emph{output}.
-If the current value in the store matches one of the possible values, the build
-can skip this key. If the resulting list is empty, the key must be rebuilt.
-However, if the current value does not match the store, and there is a possible
-value, we can use any value from the constructive list \emph{without} doing any
-work to build it, and copy it into the store.
-
-\subsubsection{Deep Constructive Traces}\label{sec-deep-constructive-traces}
-
-Constructive traces always verify keys by looking at their immediate
-dependencies, which must have first been brought up to date, meaning that the
-time to verify a key depends on the number of transitive dependencies. A
-\emph{deep} constructive trace optimises this process by only looking at the
-terminal \emph{input keys}, ignoring any intermediate dependencies. The operations
-capturing this approach are the same as for constructive traces
-in~\S\ref{sec-constructive-traces}, but we use the names \hs{recordDCT} and
-\hs{constructDCT}, where the underlying \hs{DCT} representation need only record
-information about hashes of inputs, not intermediate dependencies.
-
-Current build systems using deep constructive traces always record hashes of
-terminal \emph{input keys}, but the technique works equally well if we skip any
-number of dependency levels (say $n$ levels). The input-only approach is the
-special case of $n = \infty$, and constructive traces are the special case of
-$n = 1$. When $n \ge 2$, deep constructive traces require the tasks to be
-\emph{deterministic}, as otherwise it is possible to violate correctness, as
-illustrated by an example in~\S\ref{sec-cloud-aspects}
-(see Fig.~\ref{fig-frankenbuild}).
-
-A downside of deep constructive traces is that they cannot support early cutoff
-(\S\ref{sec-background-shake}), other than at $n$ levels of dependencies. On the
-other hand, these traces are particularly useful for \emph{shallow builds}, as
-discussed in~\S\ref{sec-cloud-aspects}.
-
-\subsection{Efficient Data Structures for Traces}\label{sec-smart-traces}
-
-In the examples above, we have used abstract types for the traces. Concretely,
-in our example code, they are all recorded as lists of:
-
-\begin{minted}[xleftmargin=10pt]{haskell}
-data Trace k v r = Trace { key :: k, depends :: [(k, Hash v)], result :: r }
-\end{minted}
-
-Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
-\hs{v} (constructive traces). A real system is highly likely to use a more
-optimised implementation. Some of the most obvious optimisations are:
-
-\begin{itemize}
-\item Any system using verifying traces is unlikely to see significant benefit
-from storing more than one \hs{Trace} per key\footnote{There is a small chance
-of a benefit if the dependencies change but the result does not, and then the
-dependencies change back to what they were before.}. Therefore, such systems can
-store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
-initial \hs{k} is the \hs{key} field of \hs{Trace}.
-
-\item Any system using \hs{Applicative} dependencies can omit the dependency
-keys from the \hs{Trace} since they can be recovered from the \hs{key} field.
-
-\item Any \hs{Applicative} build system using constructive traces, e.g.
-\CloudBuild, can index directly from the key and results to the output result~--~i.e.
-\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
-are stored on a central server, the client can compute the key and the hashes of
-its dependencies, then make a single call to the server to retrieve the result.
-
-\item Many cloud build systems store hashes of values in the trace information,
-then have a separate content-addressable cache which associates hashes with
-their actual contents.
-\vspace{-1mm}
-\end{itemize}
-
-\subsection{Build Systems \`a la Carte}\label{sec-design-space}
-
-\begin{table}
-\caption{Build systems \`a la carte.\label{tab-build-systems}}
-\vspace{-4mm}
-% \smaller
-\centering
-\begin{tabular}{l||c|c|c}
-\hline
- & \multicolumn{3}{c}{Scheduling algorithm} \\
-Rebuilding strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
-\hline
-Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
-Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
-Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
-Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
-\end{tabular}
-\vspace{-4mm}
-\end{table}
-
-With the information in this section we can build Table~\ref{tab-build-systems},
-which tabulates combinations of the scheduling algorithm and the rebuilding
-strategy, providing 12~possible build systems, 8~of which are inhabited
-by existing build systems (we discuss these systems in~\S\ref{sec-background} and
-\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
-build systems. The most interesting unfilled spot in the table is suspending
-constructive traces, which would provide many benefits, and which we title
-\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
-% (as we plan on extending \Shake to occupy that spot)

### 2019-04-18 papers/jfp/5-rebuilders.tex
Lift the contents of todo.md into the .tex and start to do the proposed splits

--- 2019-03-05 papers/jfp/4-build.tex
@@
-\section{Build Systems \`a la Carte}\label{sec-build}
+\section{Build Systems \`a la Carte Rebuilders}\label{sec-build}
 
-The focus of this paper is on a variety of implementations of
-\hs{Build}~\hs{c}~\hs{i}~\hs{k}~\hs{v}, given
-a \emph{client-supplied} implementation of \hs{Tasks}~\hs{c}~\hs{k}~\hs{v}. That
-is, we are going to take \hs{Tasks} as given from now on, and explore variants of
-\hs{Build}: first abstractly (in this section) and then concretely
-in~\S\ref{sec-implementations}.
+\todo{Elaborate (it's very terse right now) with examples}
 
-As per the definition of minimality~\ref{def-minimal}, a minimal build
-system must \textbf{rebuild only out-of-date keys} and at most once. The only
-way to achieve the ``at most once'' requirement while producing a correct build
-result (\S\ref{sec-build-correctness}) is to \textbf{build all keys in an
-order that respects their dependencies}.
+\todo{Add Shake's step traces: Optimisation, Correctness}
 
-\vspace{1mm}
-We have emboldened two different aspects above: the part of the
-build system responsible for scheduling tasks in the dependency order
-(a `scheduler') can be cleanly separated from the part responsible for deciding
-whether a key needs to be rebuilt (a `rebuilder'). We tackle each
-aspect separately in subsections~\S\ref{sec-dependency-orderings}
-and~\S\ref{sec-out-of-date}.
-
-\subsection{The Scheduler: Respecting the Dependency Order}
-\label{sec-dependency-orderings}
-
-Section \S\ref{sec-background} introduced three different \emph{task schedulers}
-that decide which tasks to execute and in what order; see the ``Scheduler'' column
-of Table~\ref{tab-summary} in \S\ref{sec-background-summary}.
-This subsection explores the properties of the three schedulers, and
-possible implementations.
-
-\vspace{-2mm}
-\subsubsection{Topological}\label{sec-topological}
-
-The topological scheduler pre-computes a linear order of tasks, which when followed,
-ensures the build result is correct regardless of the initial store. Given a
-task description and the output \hs{key}, you can compute the linear order by
-first finding the (acyclic) graph of the \hs{key}'s reachable dependencies, and
-then computing a topological sort. However, as we have seen in~\S\ref{sec-deps},
-we can only extract dependencies from an applicative task, which requires the
-build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic
-dependencies.
-
-\vspace{-2mm}
-\subsubsection{Restarting}\label{sec-restarting}
-
-To handle dynamic dependencies we can use the following approach: build tasks in
-an arbitrary initial order, discovering their dependencies on the fly; whenever
-a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task, and
-switch to building the dependency \hs{dep}; eventually the previously aborted
-task is restarted and makes further progress thanks to \hs{dep} now being up to
-date. This approach requires a way to abort tasks that have failed due to
-out-of-date dependencies. It is also not minimal in the sense that a task may
-start, do some meaningful work, and then abort.
-
-To reduce the number of aborts (often to zero) \Excel records the
-discovered task order in its \emph{calc chain}, and uses it as the
-starting point for the next build (\S\ref{sec-background-excel}).
-\Bazel's restarting scheduler does not store the discovered order
-between build runs; instead, it stores the most recent task dependency
-information. Since this information may become outdated, \Bazel may
-also need to abort a task if a newly discovered dependency is out of date.
-
-\vspace{-2mm}
-\subsubsection{Suspending}\label{sec-suspending}
-
-An alternative approach, utilised by the \hs{busy} build system
-(\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
-they are requested, suspending the currently running task. By combining that
-with tracking the keys that have already been built, one can obtain a minimal
-build system with dynamic dependencies.
-
-This approach requires that a task may be started and then suspended until
-another task is complete. Suspending can be done with cheap green threads and
-blocking (the original approach of \Shake) or using continuation-passing style
-\cite{claessen_continuations} (what \Shake currently does).
+\todo{Explain how to "add constructiveness" to a trace system.
+Mabye even concretely:  `constructive :: TraceSystem -> TraceSystem`.Maybe collapse verifying traces and constrctive traces, at least in table 2.
+(This would have impact on the structore of the whole trace section.)}
 
@@
 \end{itemize}
-
-\subsection{Build Systems \`a la Carte}\label{sec-design-space}
-
-\begin{table}
-\caption{Build systems \`a la carte.\label{tab-build-systems}}
-\vspace{-4mm}
-% \smaller
-\centering
-\begin{tabular}{l||c|c|c}
-\hline
- & \multicolumn{3}{c}{Scheduling algorithm} \\
-Rebuilding strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
-\hline
-Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
-Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
-Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
-Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
-\end{tabular}
-\vspace{-4mm}
-\end{table}
-
-With the information in this section we can build Table~\ref{tab-build-systems},
-which tabulates combinations of the scheduling algorithm and the rebuilding
-strategy, providing 12~possible build systems, 8~of which are inhabited
-by existing build systems (we discuss these systems in~\S\ref{sec-background} and
-\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
-build systems. The most interesting unfilled spot in the table is suspending
-constructive traces, which would provide many benefits, and which we title
-\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
-% (as we plan on extending \Shake to occupy that spot)

### 2019-04-18 papers/jfp/6-implementations.tex
Lift the contents of todo.md into the .tex and start to do the proposed splits

--- 2019-03-05 papers/icfp/5-implementations.tex
@@
 \section{Build systems, concretely}\label{sec-implementations}
+
+\subsection{Build Systems \`a la Carte}\label{sec-design-space}
+
+\begin{table}
+\caption{Build systems \`a la carte.\label{tab-build-systems}}
+\vspace{-4mm}
+% \smaller
+\centering
+\begin{tabular}{l||c|c|c}
+\hline
+ & \multicolumn{3}{c}{Scheduling algorithm} \\
+Rebuilding strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
+\hline
+Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
+Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
+Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
+Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
+\end{tabular}
+\vspace{-4mm}
+\end{table}
+
+With the information in this section we can build Table~\ref{tab-build-systems},
+which tabulates combinations of the scheduling algorithm and the rebuilding
+strategy, providing 12~possible build systems, 8~of which are inhabited
+by existing build systems (we discuss these systems in~\S\ref{sec-background} and
+\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
+build systems. The most interesting unfilled spot in the table is suspending
+constructive traces, which would provide many benefits, and which we title
+\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
+% (as we plan on extending \Shake to occupy that spot)
+
+\subsection{Introduction}
 

### 2019-04-18 papers/jfp/7-experience.tex
Lift the contents of todo.md into the .tex and start to do the proposed splits

\section{Experience}\label{sec-experience}

\todo{Write this section}

\subsection{Cloud Shake}

\begin{itemize}
\item Implemented and released
\item Absolute paths and system binaries
\item Deferred materialisation. Different invariants for (a) local builds, (b) cloud sharing (you must list all the things you produce), (c) sharing + deferred materialisation (you must declare all the things you consume). Interaction with early cut-off.
\end{itemize}

\subsection{Cloud Hadrian}

\begin{itemize}
\item  Real-life example
\item Many examples demonstrating the use of dynamic dependencies
\end{itemize}

\subsection{Experience}

\begin{itemize}
\item  Cloud stuff needed tracing infrastructure to expose dependencies.
\item  Good profiling support (here are pictures) exposes places where we are over-sequential, and over-big tasks (build systems work best with lots of small tasks).
\item Ability to compute critical paths
\item Numbers: no-op rebuild time; fresh clone rebuild time; (vs clean build).
\end{itemize}

### 2019-04-18 papers/jfp/8-engineering.tex
Lift the contents of todo.md into the .tex and start to do the proposed splits

--- 2019-04-14 papers/jfp/6-engineering.tex
@@
 \section{Engineering aspects}\label{sec-engineering}
+
+\todo{Tasks with multiple outputs:
+  https://github.com/snowleopard/build/blob/master/src/Build/Multi.hs}
+
+\todo{Typed tasks:
+  https://github.com/snowleopard/build/blob/master/src/Build/Task/Typed.hs}
+
+\todo{Handling failures, see last section here:
+  https://blogs.ncl.ac.uk/andreymokhov/the-task-abstraction/}
+
+\todo{Self-tracking (?):
+  https://github.com/snowleopard/build/blob/master/src/Build/SelfTracking.hs}
+
+\todo{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.}
 

### 2019-04-18 papers/jfp/9-related.tex
Lift the contents of todo.md into the .tex and start to do the proposed splits

--- 2018-07-07 paper/7-related.tex
@@
 \section{Related work}\label{sec-related}
+
+\todo{Relationship to SHake paper.   Fully subsumes it, and explains it much better.  Oracles no longer imoprtant.
+}
 
@@
 Most build systems, when viewed at the level we talk, can be captured with minor
-variations of the code presented in \S\ref{sec-implementations}. Below we list
+variations on the code presented in \S\ref{sec-implementations}. Below we list
 some notable examples:
@@
 The one build system we are aware of that cannot be modelled in our framework is
-\Fabricate by \citet{fabricate}. In \Fabricate a build system is a script that is
+\Fabricate by \cite{fabricate}. In \Fabricate a build system is a script that is
 run in-order, in the spirit of:
@@
 
-\subsection{Self-adjusting computation}
+\subsection{Self-adjusting Computation}
 
@@
 well studied area, and in particular the contrast between different formulations
-has been thoroughly investigated, e.g. see~\citet{acar2007selfadjusting}.
+has been thoroughly investigated, e.g. see~\cite{acar2007selfadjusting}.
 Self-adjusting computations can automatically adjust to an external change
@@
 algorithms for self-adjusting computations (with a few open-source
-implementations, e.g. \Incremental by~\citet{incremental}). We plan to
+implementations, e.g. \Incremental by~\cite{incremental}). We plan to
 investigate how these insights can be utilised by build systems as future work.

### 2019-04-18 papers/jfp/1-intro.tex
Lift the contents of todo.md into the .tex and start to do the proposed splits
Rework the intro a bit

@@
 
-In this paper we offer a general framework in which to understand and compare
+In this paper we give a general framework (refined from \cite{mokhov2018buildsystems}) in which to understand and compare
 build systems, in a way that is both abstract (omitting incidental detail)
-and yet precise (implemented as Haskell code). Specifically we make these
-contributions:
+and yet precise (implemented as Haskell code).
+With this framework we can explain confusing details from the original Shake paper \cite{mitchell2012shake} and build new features useful for the GHC build system \cite{hadrian}.
+Specifically we make these contributions:
 \begin{itemize}
@@
   any build system: \emph{the order in which tasks are
-  built}~(\S\ref{sec-dependency-orderings}) and \emph{whether or not a
-  task is rebuilt}~(\S\ref{sec-out-of-date}). These choices turn out to
+  built}~(\S\ref{sec-scheduler}) and \emph{whether or not a
+  task is rebuilt}~(\S\ref{sec-rebuilder}). These choices turn out to
   be orthogonal, which leads us to a new classification of the design
@@
 with desired properties, for example, to combine the advantages of \Shake and
-\Bazel (\S\ref{sec-implementation-cloud}).
+\Bazel (\S\ref{sec-cloud-shake}), and evaluate the result (\S\ref{sec-experience}).
 
@@
 leading to a better understanding of what they do and how they compare,
-and suggesting exploration of other (as yet unoccupied points) in the
+and making it easier to explore other points in the
 landscape.

### 2019-04-18 papers/jfp/4-schedulers.tex
Rework the intro a bit
Rework the schedulers chapter

@@
-\section{Build Systems \`a la Carte Schedulers}\label{sec-build}
-
-\todo{Elaborate (it's very terse right now) with examples}
+\section{Build System Schedulers}\label{sec-scheduler}
 
@@
 
-\vspace{1mm}
 We have emboldened two different aspects above: the part of the
@@
 (a `scheduler') can be cleanly separated from the part responsible for deciding
-whether a key needs to be rebuilt (a `rebuilder'). We tackle each
-aspect separately in subsections~\S\ref{sec-dependency-orderings}
-and~\S\ref{sec-out-of-date}.
-
-\subsection{The Scheduler: Respecting the Dependency Order}
-\label{sec-dependency-orderings}
+whether a key needs to be rebuilt (a `rebuilder'). In this section we discuss schedulers,
+leaving rebuilders for \S\ref{sec-rebuilder}.
 
@@
 
-\vspace{-2mm}
-\subsubsection{Topological}\label{sec-topological}
+\subsection{Topological Scheduler}\label{sec-topological}
 
-The topological scheduler pre-computes a linear order of tasks, which when followed,
-ensures the build result is correct regardless of the initial store. Given a
-task description and the output \hs{key}, you can compute the linear order by
-first finding the (acyclic) graph of the \hs{key}'s reachable dependencies, and
-then computing a topological sort. However, as we have seen in~\S\ref{sec-deps},
-we can only extract dependencies from an applicative task, which requires the
-build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic
-dependencies.
+The topological scheduler pre-computes a linear order of tasks, which when followed ensures
+dependencies are satisifed, then executes the required tasks in that order. Computing such a linear
+order is straightforward -- given a task description and output \hs{key} first find the (acyclic)
+graph of the \hs{key}'s dependencies, then compute a topological sort. Taking the example from
+Figure~\ref{fig-make}, we might compute the order:
 
-\vspace{-2mm}
-\subsubsection{Restarting}\label{sec-restarting}
+\begin{enumerate}
+\item \hs{util.o}
+\item \hs{main.o}
+\item \hs{main.exe}
+\end{enumerate}
 
-To handle dynamic dependencies we can use the following approach: build tasks in
-an arbitrary initial order, discovering their dependencies on the fly; whenever
-a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task, and
-switch to building the dependency \hs{dep}; eventually the previously aborted
-task is restarted and makes further progress thanks to \hs{dep} now being up to
-date. This approach requires a way to abort tasks that have failed due to
-out-of-date dependencies. It is also not minimal in the sense that a task may
-start, do some meaningful work, and then abort.
+Given the dependencies, we could have equally chosen to build \hs{main.o} first, but \hs{main.exe} \emph{must}
+come last.
 
-To reduce the number of aborts (often to zero) \Excel records the
+The advantage of this scheme is simplicity -- compute an order, then execute tasks in that order.
+In addition any missing nodes or cycles can be detected from the graph, and reported to the user
+before any work has commenced.
+
+The downside of this approach is that it requires the dependencies of each task in advance.
+As we saw in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task, which requires the
+build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic dependencies.
+
+\subsection{Restarting Scheduler}\label{sec-restarting}
+
+To handle dynamic dependencies we cannot precompute a static order -- we must
+interleave running tasks and ordering tasks. One approach is just to build tasks
+in an arbitrary order, but if a task calls \hs{fetch} on an out-of-date key \hs{dep}
+abort the task and build \hs{dep} instead. Returning to the example from
+Figure~\ref{fig-make}, we might build in the order:
+
+\begin{enumerate}
+\item \hs{main.exe} (abort because it requires \hs{util.o})
+\item \hs{util.o}
+\item \hs{main.o}
+\item \hs{main.exe}
+\end{enumerate}
+
+Here we start with \hs{main.exe} (chosen randomly), but discover it requires \hs{util.o}, so
+instead start building \hs{util.o}. Next we choose to build \hs{main.o} (again, randomly),
+before finally returning to \hs{main.exe} which now has all its dependencies available and
+completes without error.
+
+The approach works, but has a number of disadvantages. Firstly, it requires a technical mechanism
+to abort a task -- easy enough in our theoretical setting of \hs{Task} but leads to engineering concerns
+in the real world. Secondly, it is not minimal in the sense that a task may
+start, do some meaningful work, and then abort, repeating that same work when restarted.
+
+As a refinement, to reduce the number of aborts (often to zero) \Excel records the
 discovered task order in its \emph{calc chain}, and uses it as the
@@
 between build runs; instead, it stores the most recent task dependency
-information. Since this information may become outdated, \Bazel may
+information from which it can computed a linear order. Since this information may become outdated, \Bazel may
 also need to abort a task if a newly discovered dependency is out of date.
 
-\vspace{-2mm}
-\subsubsection{Suspending}\label{sec-suspending}
+\subsection{Suspending Scheduler}\label{sec-suspending}
 
@@
 (\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
-they are requested, suspending the currently running task. By combining that
-with tracking the keys that have already been built, one can obtain a minimal
-build system with dynamic dependencies.
+they are requested, suspending the currently running task. Using
+Figure~\ref{fig-make}, we would build:
 
-This approach requires that a task may be started and then suspended until
-another task is complete. Suspending can be done with cheap green threads and
-blocking (the original approach of \Shake) or using continuation-passing style
-\cite{claessen_continuations} (what \Shake currently does).
+\begin{enumerate}
+\item \hs{main.exe} (suspended)
+\begin{itemize}
+\item \hs{util.o}
+\end{itemize}
+\item \hs{main.exe} (resumsed then suspended)
+\begin{itemize}
+\item \hs{main.o}
+\end{itemize}
+\item \hs{main.exe} (completed)
+\end{enumerate}
+
+As we see, the \hs{main.exe} is started first as it is the required target. It soon discovers
+a dependency on \hs{util.o} and suspends to build \hs{util.o}, resumes a build before building \hs{main.o}
+then continues its work to finish \hs{main.exe}.
+
+This scheduler (when combined with a suitable rebuilder) provides a minimal build system
+which supports dynamic dependencies.
+
+The difficulties are all engineering in nature. To implement suspension there are two
+standard approaches:
+
+\begin{itemize}
+\item Blocking threads or processes. This approach is relatively easy, but can require
+significant resources, especially if a large number of tasks are suspended. In languages with
+cheap green threads (e.g. Haskell) the approach is more feasible, and it was the original approach
+taken by \Shake.
+\item Continuation-passing style \cite{claessen_continuations} can allow the remainder of
+a task to be captured, paused, and resumed later. Continuation passing is efficient, but requires
+the build script to be architected to allow capturing continuations. \Shake currently uses
+continuations.
+\end{itemize}
+
+While a suspending scheduler is theoretically optimal, it is only better in practice
+than a restarting scheduler if the cost of duplicate work avoided outweighs the cost of suspending tasks.

### 2019-04-23 papers/jfp/3-abstractions.tex
Add another todo to AM
Work on S3

--- 2019-03-05 papers/jfp/3-abstractions.tex
@@
 \section{Build systems, abstractly}\label{sec-abstractions}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 
 \begin{figure}
-\vspace{-1mm}
+% \vspace{-1mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \end{minted}
-\vspace{-1mm}
+% \vspace{-1mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \end{minted}
-\vspace{-1mm}
+% \vspace{-1mm}
 \begin{minted}[fontsize=\small]{haskell}
--- Build system components: a scheduler and a rebuilder (see §5)
+-- Build system components: a scheduler (see §4) and a rebuilder (see §5)
 type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
@@
 \end{minted}
-\vspace{-3.5mm}
+% \vspace{-3.5mm}
 \caption{Type signatures of key build systems abstractions.}\label{fig-types}
-\vspace{-4.5mm}
+% \vspace{-4.5mm}
 \end{figure}
@@
 present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
-and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-build}
-and~\S\ref{sec-implementations} scrutinise the abstractions further and provide
-concrete implementations for several build systems.
+and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-scheduler},
+\S\ref{sec-rebuilder} and~\S\ref{sec-implementations} scrutinise the
+abstractions further and provide concrete implementations for several build
+systems.
 
-\vspace{-1.5mm}
+% \vspace{-1.5mm}
 \subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 
@@
 \end{minted}
-\vspace{-3mm}
+% \vspace{-3mm}
 \caption{Standard library definitions.}\label{fig-stdlib}
-\vspace{-4mm}
+% \vspace{-4mm}
 \end{figure}
@@
         Nothing   -> gets (getValue k)
-        Just task -> do v <- run task fetch; modify (putValue k v); return v
+        Just task -> do v <- run task fetch
+                        modify (putValue k v)
+                        return v
 \end{minted}
@@
 \subsection{The Need for Polymorphism in Task}\label{sec-why-polymorphism}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 
@@
 
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 \subsection{Monadic Tasks}\label{sec-task-monad}
@@
 \label{def-correct}
-\vspace{2mm}
+% \vspace{2mm}
 
@@
 which in this case is obtained by reading from the standard input.
+
+\subsection{Examples of Tasks}\label{sub-task-examples}
+
+The \hs{Task} abstraction can be used to express pure functions in a way that is
+convenient for their memoization, as will be further discussed
+in~\S\ref{sec-related-memo}. Below we give examples of tasks corresponding to
+several well-known functions.
+
+The \emph{Fibonacci sequence} $F_n = F_{n-1} + F_{n-2}$ is the one of the most
+favourite examples in the functional programming community and so we start with
+it:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+fibonacci :: Tasks Applicative Integer Integer
+fibonacci n = if n < 2 then Nothing else
+    Just $ Task $ \fetch -> (+) <$> fetch (n - 1) <*> fetch (n - 2)
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Here the keys $n < 2$ are input parameters, and one can obtain the usual
+Fibonacci sequence by picking 0 and 1 for $n = 0$ and $n = 1$, respectively.
+Any minimal build system will compute the sequence with memoization, i.e.
+without recomputing the same value twice.
+
+% TODO: Add a figure with a dependency graph?
+The dependencies of Fibonacci numbers are statically known, hence we can express
+it using \hs{Tasks}~\hs{Applicative}, benefiting from static dependency
+analysis:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\ghci@ dependencies (fromJust $ fibonacci 5)
+[4,3]
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Interestingly, the \emph{Ackermann function} -- a famous example of a function
+that is not primitive recursive -- cannot be expressed as an applicative task,
+since it needs to perform an intermediate recursive call to determine one of its
+dependencies $A(m, n-1)$:
+
+\[
+A(m, n) = \left\{
+\begin{array}{ll}
+n+1 & \mbox{if } m = 0 \\
+A(m-1, 1) & \mbox{if } m > 0 \mbox{ and } n = 0 \\
+A(m-1, A(m, n-1)) & \mbox{if } m > 0 \mbox{ and } n > 0.
+\end{array}\right.
+\]
+
+We therefore use \hs{Tasks}~\hs{Monad} to express this function, binding the
+dynamic dependency to variable \hs{index}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+ackermann :: Tasks Monad (Integer, Integer) Integer
+ackermann (n, m)
+    | m < 0 || n < 0 = Nothing
+    | m == 0    = Just $ Task $ const $ pure (n + 1)
+    | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
+    | otherwise = Just $ Task $ \fetch -> do index <- fetch (m, n - 1)
+                                             fetch (m - 1, index)
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Now that we have seen examples of applicative and monadic tasks, let us finish
+with an example of a functorial task -- the \emph{Collatz sequence}, where
+given an initial value $c_0$, we calculate the next value $c_n$ from $c_{n-1}$
+either by dividing $c_{n-1}$ by 2 (if it is even) or multiplying it by 3 and
+adding 1 (if it is odd):
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+collatz :: Tasks Functor Integer Integer
+collatz n | n <= 0    = Nothing
+          | otherwise = Just $ Task $ \fetch -> f <$> fetch (n - 1)
+  where
+    f k | even k    = k `div` 2
+        | otherwise = 3 * k + 1
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Functorial tasks correspond to computations with a linear dependency chain. For
+example, computing the element $c_8$ of the Collatz sequence starting from
+$c_0=6$ leads to the following dependency chain:
+$c_0 = 6 \rightarrow 3 \rightarrow 10 \rightarrow 5 \rightarrow 16 \rightarrow 8
+\rightarrow 4 \rightarrow 2 \rightarrow 1 = c_8$.
+
+Collatz sequence is a good example of the early cutoff optimisation
+(\S\ref{sec-background-shake}): if we recompute $c_8$ starting from a different
+initial value $c_0=40$, the resulting computation will have a large overlap with
+the previous one:
+$c_0 = 40 \rightarrow 20 \rightarrow 10 \rightarrow 5 \rightarrow 16 \rightarrow
+8 \rightarrow 4 \rightarrow 2 \rightarrow 1 = c_8$. We can therefore stop the
+recomputation after just two steps, since $c_2=10$ has not changed.
+
+Note that functorial tasks allow us to statically extract even more precise
+information about their dependencies than applicative tasks. Indeed, we
+statically know that a \hs{Task}~\hs{Functor} has exactly one \hs{dependency}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+dependency :: Task Functor k v -> k
+dependency task = getConst $ run task Const
+\end{minted}
+\vspace{1mm}

### 2019-04-23 papers/jfp/5-rebuilders.tex
Rework the intro a bit
Start work on rebuilders

@@
-\section{Build Systems \`a la Carte Rebuilders}\label{sec-build}
+\section{Rebuilders}\label{sec-rebuilder}
+
+A build system can be split into a scheduler (as defined in \S\ref{sec-scheduler}) and a \emph{rebuilder}. Suppose the scheduler decides that a key should be brought up to date. The next
+question is: does any work need to be done, or is the key already up to date?
+Or, in a cloud-build system, do we have a cached copy of the value we need?
+
+While \S\ref{sec-background} explicitly listed the schedulers, the rebuilders were introduced more implicitly, primarily by the information they retain to make their decisions.
+From the examples we have looked at we see four fundamental rebuilds, each with a number of tweaks and variations within them.
 
@@
 
-\subsection{The Rebuilder: Determining Out-of-date Keys}\label{sec-out-of-date}
-
-Suppose the scheduler decides that a key should be brought up to date. The next
-question is: does any work need to be done, or is the key already up to date?
-Or, in a cloud-build system, do we have a cached copy of the value we need?
-These questions can be addressed in one of four fundamental ways, with a number
-of tweaks and variations within them.
-
-\vspace{-2mm}
-\subsubsection{A Dirty Bit}\label{sec-dirty-bit}
+\subsection{A Dirty Bit}\label{sec-dirty-bit}
 
@@
 two builds is marked dirty. If a key and all its transitive dependencies are
-clean, the key does not need to be rebuilt.
+clean, the key does not need to be rebuilt. Taking the example from
+Figure~\ref{fig-make}, if \hs{util.c} changes then that would be marked dirty, and \hs{util.o} and \hs{main.exe} would rebuild as they transitively depend on \hs{util.c}.
 
@@
 
-\vspace{-2mm}
-\subsubsection{Verifying Traces}\label{sec-verifying-traces}
+The use of a single bit of information is useful to reduce memory consumption, and
+in the case of \Make, to integrate with the file system. However, as the examples show,
+in constrained environments where a dirty bit is chosen, it's often done as part of
+a series of compromises.
+
+\subsection{Verifying Traces}\label{sec-verifying-traces}
 
@@
 key is dirty and must be rebuilt~--~in essence a \emph{trace} which we can use
-to \emph{verify} existing values. For traces, there are two essential
+to \emph{verify} existing values. Taking the example from Figure \ref{fig-make},
+we might record that \hs{util.o} dependend on \hs{util.c} (at hash 1) and \hs{util.h}
+(at hash 2) and produced \hs{util.o} (at hash 4). Next time round, if all those files
+still have the same hashes, there is nothing to do, and we can skip running the build.
+If they are different, we repeat the action, and record the new values.
+
+For traces, there are two essential
 operations~--~adding a new value to the trace store, and using the traces to
@@
 \noindent
-Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} stores
+Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} can store
 only hashes, of type \hs{Hash}~\hs{v}, of those values. Since the verifying
@@
 
+A verifying trace, and other types of traces discussed in this section, support
+dynamic dependencies and minimality; furthermore, all traces except for deep
+traces~(\S\ref{sec-deep-constructive-traces}) support early cutoff.
+
+\subsubsection{Representation Optimisations}
+
 One potential implementation would be to record all arguments passed to
@@
 
-A verifying trace, and other types of traces discussed in this section, support
-dynamic dependencies and minimality; furthermore, all traces except for deep
-traces~(\S\ref{sec-deep-constructive-traces}) support early cutoff.
+In the example above, we have used abstract types for the traces. Concretely,
+in our example code, they are all recorded as lists of:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Trace k v = Trace { key :: k, depends :: [(k, Hash v)], result :: Hash v }
+\end{minted}
+
+Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
+\hs{v} (constructive traces). A real system is highly likely to use a more
+optimised implementation. Some of the most obvious optimisations are:
+
+\begin{itemize}
+\item Any system using verifying traces is unlikely to see significant benefit
+from storing more than one \hs{Trace} per key\footnote{There is a small chance
+of a benefit if the dependencies change but the result does not, and then the
+dependencies change back to what they were before.}. Therefore, such systems can
+store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
+initial \hs{k} is the \hs{key} field of \hs{Trace}.
+
+\item Any system using \hs{Applicative} dependencies can omit the dependency
+keys from the \hs{Trace} since they can be recovered from the \hs{key} field.
+
+\item Any \hs{Applicative} build system using constructive traces, e.g.
+\CloudBuild, can index directly from the key and results to the output result~--~i.e.
+\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
+are stored on a central server, the client can compute the key and the hashes of
+its dependencies, then make a single call to the server to retrieve the result.
+
+\item Many cloud build systems store hashes of values in the trace information,
+then have a separate content-addressable cache which associates hashes with
+their actual contents.
+\vspace{-1mm}
+\end{itemize}
+
+\subsubsection{Step Verifying Traces}
+
+The \Shake build system, and the associated paper \cite{mitchell2012shake} in \S2.3.3 describes a slightly different trace structure, one that is similar to verifying traces, but stores slightly less data, and has slightly different early cutoff semantics.
+Rather than storing the \hs{Hash v} for each dependency, it instead stores a built time and changed time for each key, and for the dependencies just stores a list of \hs{k}.
+The resulting \hs{Trace} type ressembles:
+
+\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
+data Trace k v = Trace
+    {key :: k
+    ,result :: Hash v
+    ,depends :: [k]
+    ,built :: Time
+    ,changed :: Time
+    }
+\end{minted}
+
+The \hs{built} field is when the \hs{key} last rebuilt.
+The \hs{changed} field is when the \hs{result} last changed -- if the last build changed the value, it will be equal to \hs{built}, otherwise it will be older.
+The \hs{recordVT} function needs to consult the previous \hs{Trace} to know whether to use \hs{built} or the previous \hs{changed} value as the new \hs{changed} value.
+The \hs{verifyVT} needs to verify that the \hs{built} time of \hs{key} is newer than the \hs{changed} values of all its \hs{depends}.
+This scheme preserves minimality and early cutoff.
+A variant with only one \hs{Time} field would loose early cutoff.
+Furthermore, the \hs{Time} stamp only needs record which execution of the build is running, so every key built in the same run can share the same \hs{Time} value -- it just needs to be monotically increasing between runs.
+
+This optimisation is useful, at least in the case of \Shake, to save space.
+A hash takes up 32 bytes in typical programs, while a key (in \Shake only) is an \hs{Int} taking only 4 bytes.
+Furthermore, \Shake permits values to be arbitrarily large, and permits a custom equality (two values can be bit-for-bit unequal but considered equal by \Shake), meaning that a \hs{Hash} isn't a valid encoding.
+For build systems that used \hs{Applicative} dependencies, the \hs{depends} could be omitted entirely, making the size of a \hs{Trace} $O(1)$ instead of $O(n)$ where $n$ is the number of dependencies.
+
+While verifying step traces are mostly an optimisation, there are some observable differences from verifying traces.
+Considering Figure \ref{fig-make}, imagine we built \hs{main.exe} (everything gets a \hs{built} and \hs{changed} of timestamp 1).
+Next we change \hs{util.c} and rebuild \hs{util.o} (both get a \hs{built} and \hs{changed} of timestamp 2).
+Next we put \hs{util.c} back to what it was originally, and build \hs{main.exe}.
+With verifying traces, the dependencies of \hs{main.exe} would be equal, and \hs{main.exe} would not rebuild.
+With verifying step traces, the \hs{changed} field of \hs{util.o} would increase once more, and \hs{main.exe} would rebuild.
+Other than when building subsets of the targets, we are unaware of any other situation where verifying step traces are less powerful.
 

### 2019-04-24 papers/jfp/3-abstractions.tex
Finish S3.8

@@
 Given \emph{one} task description \cmd{T}, we want to explore \emph{many} build
-systems that can build \cmd{T} -- and we will do so in sections~\S\ref{sec-build}
-and~\S\ref{sec-implementations}. As we shall see, each build system will use a
-different \hs{f}, so the task description must not fix \hs{f}.
+systems that can build \cmd{T} -- and we will do so in
+section~\S\ref{sec-implementations}. As we shall see, each build system will use
+a different \hs{f}, so the task description must not fix \hs{f}.
 
@@
 
-\subsection{Computing Dependencies}\label{sec-deps}\label{secdeps}
+\subsection{Computing Dependencies}\label{sec-deps}
 
@@
 
-\subsection{Examples of Tasks}\label{sub-task-examples}
+\subsection{Examples of Tasks}\label{sec-task-examples}
 
-The \hs{Task} abstraction can be used to express pure functions in a way that is
-convenient for their memoization, as will be further discussed
-in~\S\ref{sec-related-memo}. Below we give examples of tasks corresponding to
-several well-known functions.
+In this section we give examples of tasks whose definitions involve different
+constraints on the computation context: \hs{Functor}, \hs{Applicative},
+\hs{Monad} and \hs{MonadState}~\hs{s}.
 
-The \emph{Fibonacci sequence} $F_n = F_{n-1} + F_{n-2}$ is the one of the most
-favourite examples in the functional programming community and so we start with
-it:
+We start with one of the favourite examples for functional
+programmers -- the \emph{Fibonacci sequence} $F_n = F_{n-1} + F_{n-2}$:
 
@@
 Here the keys $n < 2$ are input parameters, and one can obtain the usual
-Fibonacci sequence by picking 0 and 1 for $n = 0$ and $n = 1$, respectively.
-Any minimal build system will compute the sequence with memoization, i.e.
-without recomputing the same value twice.
+Fibonacci sequence by picking $F_0=0$ and $F_1=1$, respectively. Any minimal
+build system will compute the sequence with memoization, i.e. without
+recomputing the same value twice.
 
 % TODO: Add a figure with a dependency graph?
-The dependencies of Fibonacci numbers are statically known, hence we can express
-it using \hs{Tasks}~\hs{Applicative}, benefiting from static dependency
-analysis:
+Dependencies of elements of the Fibonacci sequence are known statically, hence
+we can express it using \hs{Tasks}~\hs{Applicative}, and benefit from static
+dependency analysis (\S\ref{sec-deps}):
 
@@
 that is not primitive recursive -- cannot be expressed as an applicative task,
-since it needs to perform an intermediate recursive call to determine one of its
-dependencies $A(m, n-1)$:
+because it needs to perform an intermediate recursive call to determine the
+value of one of its dependencies $A(m, n-1)$:
 
@@
 ackermann :: Tasks Monad (Integer, Integer) Integer
-ackermann (n, m)
+ackermann (m, n)
     | m < 0 || n < 0 = Nothing
@@
 \noindent
-Now that we have seen examples of applicative and monadic tasks, let us finish
-with an example of a functorial task -- the \emph{Collatz sequence}, where
-given an initial value $c_0$, we calculate the next value $c_n$ from $c_{n-1}$
-either by dividing $c_{n-1}$ by 2 (if it is even) or multiplying it by 3 and
-adding 1 (if it is odd):
+Functorial tasks are less common than applicative and monadic, but there is a
+classic example too -- the \emph{Collatz sequence}, where given an initial value
+$c_0$, we calculate the next value $c_n$ from $c_{n-1}$ either by dividing
+$c_{n-1}$ by 2 (if it is even) or multiplying it by 3 and adding 1 (if it is
+odd):
 
@@
 
-Note that functorial tasks allow us to statically extract even more precise
-information about their dependencies than applicative tasks. Indeed, we
-statically know that a \hs{Task}~\hs{Functor} has exactly one \hs{dependency}:
+Note that we can statically extract even more precise dependency information
+from functorial tasks compared to applicative tasks. Indeed, we statically know
+that a \hs{Task}~\hs{Functor} has \emph{exactly one} \hs{dependency}:
 
@@
 \vspace{1mm}
+
+\noindent
+The \hs{Tasks} abstraction allows us to express pure functions in a way that
+is convenient for their memoization and incremental recomputation
+(see~\S\ref{sec-related-memo} for a discussion on memoization). If we
+furthermore need to share computation results via a cloud cache, we can use
+\hs{Tasks}~\hs{(MonadState}~\hs{s)} that will
+play an important role in~\S\ref{sec-implementations}. Intuitively, by making
+a shared state of type \hs{s} available to a task, we give it the abilities to
+lookup and store computation results in a cache by using the \hs{MonadState}
+methods \hs{get} and \hs{set}. For example, below we implement a cloud version
+of the Ackermann task that uses a \hs{Cache} of type
+\hs{Map}~\hs{(Integer,}~\hs{Integer)}~\hs{Integer} for sharing results of known
+Ackermann values.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Cache = Map (Integer, Integer) Integer
+
+cloudAckermann :: Tasks (MonadState Cache) (Integer, Integer) Integer
+cloudAckermann (m, n)
+    | m < 0 || n < 0 = Nothing
+    | m == 0    = Just $ Task $ const $ pure (n + 1)
+    | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
+    | otherwise = Just $ Task $ \fetch -> do
+        cache <- get
+        case Map.lookup (m, n) cache of
+            Nothing -> do index <- fetch (m, n - 1)
+                          value <- fetch (m - 1, index)
+                          modify (Map.insert (m, n) value)
+                          return value
+            Just value -> return value
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The main case ($m>0 \wedge n>0$) starts by looking up the pair of indices
+\hs{(}\hs{m,}~\hs{n)} in the \hs{cache}. If the cache has \hs{Nothing}, we
+calculate the resulting \hs{value} as before and \hs{modify} the cache
+accordingly; otherwise, if we have a cache hit, we return the obtained value
+immediately, skipping the actual calculation. We will use this technique in our
+models of cloud build systems in~\S\ref{sec-implementations}.

### 2019-04-24 test/Examples.hs

@@ -3,6 +3,10 @@ module Examples where
 import Build.Task
 import Control.Applicative
 import Control.Monad.Fail (MonadFail)
+import Control.Monad.State.Class
+import Data.Map (Map)
+
+import qualified Data.Map as Map
 
 -- | A useful fetch for experimenting with build systems in interactive GHC.
 fetchIO :: (Show k, Read v) => k -> IO v
@@ -53,12 +57,30 @@ fibonacci n
 -- Formally, it has no inputs, but we return Nothing for negative inputs.
 -- For example, a[m, 1] = 2, 3, 5, 13, 65535, ...
 ackermann :: Tasks Monad (Integer, Integer) Integer
-ackermann (n, m)
+ackermann (m, n)
     | m < 0 || n < 0 = Nothing
     | m == 0    = Just $ Task $ const $ pure (n + 1)
     | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
     | otherwise = Just $ Task $ \fetch -> do index <- fetch (m, n - 1)
                                              fetch (m - 1, index)
+
+-- A cloud version of the Ackermann task that uses a cache to store known values
+-- of the Ackermann function.
+type Cache = Map (Integer, Integer) Integer
+
+cloudAckermann :: Tasks (MonadState Cache) (Integer, Integer) Integer
+cloudAckermann (m, n)
+    | m < 0 || n < 0 = Nothing
+    | m == 0    = Just $ Task $ const $ pure (n + 1)
+    | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
+    | otherwise = Just $ Task $ \fetch -> do
+        cache <- get
+        case Map.lookup (m, n) cache of
+            Nothing -> do index <- fetch (m, n - 1)
+                          value <- fetch (m - 1, index)
+                          modify (Map.insert (m, n) value)
+                          return value
+            Just value -> return value
 
 -- Unlike Collatz and Fibonacci computations, the Ackermann computation cannot
 -- be statically analysed for dependencies. We can only find the first dependency

### 2019-04-24 papers/jfp/3-abstractions.tex
Tweaks to S3.8

@@
 a shared state of type \hs{s} available to a task, we give it the abilities to
-lookup and store computation results in a cache by using the \hs{MonadState}
-methods \hs{get} and \hs{set}. For example, below we implement a cloud version
-of the Ackermann task that uses a \hs{Cache} of type
+lookup and update cached computation results using the \hs{MonadState} methods
+\hs{get} and \hs{modify}. For example, below we implement a cloud version of the
+Ackermann task that uses a \hs{Cache} of type
 \hs{Map}~\hs{(Integer,}~\hs{Integer)}~\hs{Integer} for sharing results of known
@@
 accordingly; otherwise, if we have a cache hit, we return the obtained value
-immediately, skipping the actual calculation. We will use this technique in our
-models of cloud build systems in~\S\ref{sec-implementations}.
+immediately, skipping the actual calculation and thus potentially saving a large
+amount of work. Indeed you do not want to recompute $A(4,2)=2^{65536}-3$
+unnecessarily; all of its 19,729 decimal digits have already been helpfully
+computed, e.g. see~\cite{ackermann42}. We will use this technique in our models
+of cloud build systems in~\S\ref{sec-implementations}.

### 2019-04-24 papers/jfp/5-rebuilders.tex
Redo the rebuilders section

@@
 From the examples we have looked at we see four fundamental rebuilds, each with a number of tweaks and variations within them.
-
-\todo{Elaborate (it's very terse right now) with examples}
-
-\todo{Add Shake's step traces: Optimisation, Correctness}
-
-\todo{Explain how to "add constructiveness" to a trace system.
-Mabye even concretely:  `constructive :: TraceSystem -> TraceSystem`.Maybe collapse verifying traces and constrctive traces, at least in table 2.
-(This would have impact on the structore of the whole trace section.)}
 
@@
 \hs{recordVT} in a list, and verify by simply checking if any list item matches
-the information passed by \hs{verifyVT}.  We discuss smarter implementations
-in~\S\ref{sec-smart-traces}.
-
-In the example above, we have used abstract types for the traces. Concretely,
-in our example code, they are all recorded as lists of:
+the information passed by \hs{verifyVT}. Concretely,
+in our implementations from \S\ref{sec-implementations}, traces are recorded as lists of:
 
@@
 
-Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
-\hs{v} (constructive traces). A real system is highly likely to use a more
-optimised implementation. Some of the most obvious optimisations are:
+\noindent Where \hs{r} is \hs{Hash}~\hs{v} for verifying traces (and \hs{v} for constructive traces, discussed later in \S\ref{sec-constructive-traces}).
+A real system is highly likely to use a more optimised implementation.
 
-\begin{itemize}
-\item Any system using verifying traces is unlikely to see significant benefit
-from storing more than one \hs{Trace} per key\footnote{There is a small chance
-of a benefit if the dependencies change but the result does not, and then the
-dependencies change back to what they were before.}. Therefore, such systems can
-store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
-initial \hs{k} is the \hs{key} field of \hs{Trace}.
-
-\item Any system using \hs{Applicative} dependencies can omit the dependency
+The first optimisation is that any system using \hs{Applicative} dependencies can omit the dependency
 keys from the \hs{Trace} since they can be recovered from the \hs{key} field.
 
-\item Any \hs{Applicative} build system using constructive traces, e.g.
-\CloudBuild, can index directly from the key and results to the output result~--~i.e.
-\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
-are stored on a central server, the client can compute the key and the hashes of
-its dependencies, then make a single call to the server to retrieve the result.
-
-\item Many cloud build systems store hashes of values in the trace information,
-then have a separate content-addressable cache which associates hashes with
-their actual contents.
-\vspace{-1mm}
-\end{itemize}
+The next optimisation is that there is only very minor benefit from storing more than one \hs{Trace} per key.
+Therefore, veryifying traces can be stored as \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
+initial \hs{k} is the \hs{key} field of \hs{Trace} -- making \hs{verifyVT} fast.
+Note that storing only one \hs{Trace} per key means that if the dependencies change but the result does not, and then the
+dependencies change back to what they were before, there will not be a valid \hs{Trace} available and the system will have to rebuild, where a complete list of all historical traces would not have had to.
+However, it also means that the number of \hs{Trace} structures is bounded by the number of distinct keys, regardless of how many builds are executed -- a useful property.
 
@@
 This scheme preserves minimality and early cutoff.
-A variant with only one \hs{Time} field would loose early cutoff.
+A variant with only one \hs{Time} field would loose early cutoff, and indeed corresponds quite closely to \Make.
 Furthermore, the \hs{Time} stamp only needs record which execution of the build is running, so every key built in the same run can share the same \hs{Time} value -- it just needs to be monotically increasing between runs.
@@
 
-\subsubsection{Constructive Traces}\label{sec-constructive-traces}
+\subsection{Constructive Traces}\label{sec-constructive-traces}
 
 A verifying trace deliberately records only small hashes, so that it can be small.
-A \emph{constructive} trace additionally stores the resulting value.
+In contrast, a \emph{constructive} trace also stores the resulting value. Concretely,
+it records a list of \hs{Trace}~\hs{k}~\hs{v}.
 Once we are storing the complete result it makes sense
@@
 providing cloud-build functionality. We can represent this additional
-information by providing the operations:
+information with the operations:
 
@@
 
-\subsubsection{Deep Constructive Traces}\label{sec-deep-constructive-traces}
+Any \hs{Applicative} build system using constructive traces, e.g.
+\CloudBuild, can index directly from the key and results to the output result~--~i.e.
+\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
+are stored on a central server, the client can compute the key and the hashes of
+its dependencies, then make a single call to the server to retrieve the result.
+
+In practice, many cloud build systems store hashes of values in the trace information,
+then have a separate content-addressable cache which associates hashes with
+their actual contents.
+
+\subsection{Deep Constructive Traces}\label{sec-deep-constructive-traces}
 
@@
 
+Concretely, taking the example from Figure \ref{fig-make}, to decide whether \hs{main.exe}
+is out of data, a \emph{constructive} trace would look at \hs{util.o} and \hs{main.o}
+(the immediate dependencies), while a \hs{deep constructive} trace would look at \hs{util.c},
+\hs{util.h} and \hs{main.c}.
+
+An advantage of deep constructive traces is that to decide if \hs{main.exe} is up to date only
+requires consulting its inputs, not even considering \hs{util.o} or \hs{main.o}.
+Such a feature is often known as a \emph{shallow build}, as discussed in~\S\ref{sec-cloud-aspects}.
+
+There are two primary disadvantages of deep constructive traces:
+
+\begin{description}
+\item[Deterministic tasks] If the tasks are not \emph{deterministic} then it is possible to
+violate correctness, as illustrated by an example in~\S\ref{sec-cloud-aspects}
+(see Fig.~\ref{fig-frankenbuild}).
+\item[No early cutoff] Such traces cannot support early cutoff
+(\S\ref{sec-background-shake}), since the results of intermediate computations are not considered.
+\end{description}
+
 Current build systems using deep constructive traces always record hashes of
@@
 special case of $n = \infty$, and constructive traces are the special case of
-$n = 1$. When $n \ge 2$, deep constructive traces require the tasks to be
-\emph{deterministic}, as otherwise it is possible to violate correctness, as
-illustrated by an example in~\S\ref{sec-cloud-aspects}
-(see Fig.~\ref{fig-frankenbuild}).
-
-A downside of deep constructive traces is that they cannot support early cutoff
-(\S\ref{sec-background-shake}), other than at $n$ levels of dependencies. On the
-other hand, these traces are particularly useful for \emph{shallow builds}, as
-discussed in~\S\ref{sec-cloud-aspects}.
-
-\subsection{Efficient Data Structures for Traces}\label{sec-smart-traces}
-
-In the examples above, we have used abstract types for the traces. Concretely,
-in our example code, they are all recorded as lists of:
-
-\begin{minted}[xleftmargin=10pt]{haskell}
-data Trace k v r = Trace { key :: k, depends :: [(k, Hash v)], result :: r }
-\end{minted}
-
-Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
-\hs{v} (constructive traces). A real system is highly likely to use a more
-optimised implementation. Some of the most obvious optimisations are:
-
-\begin{itemize}
-\item Any system using verifying traces is unlikely to see significant benefit
-from storing more than one \hs{Trace} per key\footnote{There is a small chance
-of a benefit if the dependencies change but the result does not, and then the
-dependencies change back to what they were before.}. Therefore, such systems can
-store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
-initial \hs{k} is the \hs{key} field of \hs{Trace}.
-
-\item Any system using \hs{Applicative} dependencies can omit the dependency
-keys from the \hs{Trace} since they can be recovered from the \hs{key} field.
-
-\item Any \hs{Applicative} build system using constructive traces, e.g.
-\CloudBuild, can index directly from the key and results to the output result~--~i.e.
-\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
-are stored on a central server, the client can compute the key and the hashes of
-its dependencies, then make a single call to the server to retrieve the result.
-
-\item Many cloud build systems store hashes of values in the trace information,
-then have a separate content-addressable cache which associates hashes with
-their actual contents.
-\vspace{-1mm}
-\end{itemize}
+$n = 1$. By picking values of $n$ in between we would regain some early cutoff, at the
+cost of losing such simple shallow builds and still requiring determinstic tasks.

### 2019-04-24 papers/jfp/6-implementations.tex
Clean up the intro to S6

@@
-\section{Build systems, concretely}\label{sec-implementations}
-
-\subsection{Build Systems \`a la Carte}\label{sec-design-space}
+\section{Build systems, concretely}\label{sec-implementations} \label{sec-design-space}
 
@@
 
-With the information in this section we can build Table~\ref{tab-build-systems},
+In the previous sections we discussed the types of build systems, and how they
+can be broken down into two main components: a scheduler (\S\ref{sec-scheduler})
+and a rebuilder (\S\ref{sec-rebuilder}).
+In this section we make this abstract distinction concrete, by
+implementing a number of build systems as a composition of a scheduler
+and a rebuilder. The result can be summarized in Table~\ref{tab-build-systems},
 which tabulates combinations of the scheduling algorithm and the rebuilding
@@
 
-\subsection{Introduction}
+\subsection{Concrete Implementations}
 
-In the previous sections we discussed the types of build systems, and how they
-can be broken down into two main components: a scheduler and a rebuilder.
-In this section we make this abstract distinction concrete, by
-implementing a number of build systems as a composition of a scheduler
-and a rebuilder.  Our design is expressed by these types (Fig.~\ref{fig-types}):
+We can define schedulers and rebuilders more concretely with the types (Fig.~\ref{fig-types}):
 

### 2019-04-24 papers/jfp/7-experience.tex
Rework the intro a bit
Start fleshing out S7

@@
 
-\todo{Write this section}
+As a set of authors, we've written a paper about the \Shake build system \cite{mitchell2012shake},
+applying that build system to GHC \cite{hadrian}, and a paper about the structure of build
+systems \cite{mokhov2018buildsystems}. The original purpose of that final paper was to create
+``Cloud Shake'', and use it for the GHC build system. In this section we tie these papers together,
+reflecting on our experience after all these papers.
 
-\subsection{Cloud Shake}
+\subsection{Experience from \Shake}
+
+The original design of \Shake hasn't changed since the initial paper.
+Since that paper there have been roughly 5,000 commits to the \Shake project\footnote{\url{https://github.com/ndmitchell/shake}}.
+They add concepts like resources (if two rules content from a single external resource), rewrite serialisation to be faster, and documentation including a website, and add lots of tests.
+The biggest change in that time period was moving from blocking threads to continuations for the suspending scheduler.
+The most visible change is switching \hs{*>} for \hs{\%>}\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.
+
+We consider the lack of change to suggest that \Shake is based on underlying principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}.
+
+\subsection{Experience from building GHC with \Shake}
+
+\todo{AM: Can you write this}
+
+The GHC build system, known as Hadrian, has continued development.
+The use of dynamic dependencies has made the build much easier, and it's now more maintainable.
+It's now merged into the GHC repo and tested as standard by the CI.
+It's current state is ...
+In particular, it takes N minutes to build with \Shake, vs M minutes to build with \Make.
+The \Shake version has attracted more contributors, and been easier to modify, etc.
+
+The design we originally outlined remains the one in use.
+There has been a huge amount of engineering work to make the tests pass (many of the details encoded in the \Make build system were incidental -- but some were essential).
+
+One of the benefits of using Shake is that we have access to high quality profiling information, allowing us to compute critical paths and other metrics (see \cite{mitchell2019ghcrebuildtimes} for a tour of the metrics).
+This information has shown us that GHC is slow to build (40m on an 8 CPU machine on Windows), that more CPUs would not help (on unlimited CPUs it would take at least 37m), and that a handful of steps (two Haskell compiles, some calls to \hs{configure}) are responsible for a significant amount of that time (at least 10m).
+
+\subsection{Experience from Cloud Shake}\label{sec-cloud-shake}
 
@@
 Interaction with early cut-off.
+\item  Cloud stuff needed tracing infrastructure to expose dependencies.
 \end{itemize}
 
-\subsection{Cloud Hadrian}
+\subsection{Experience from building GHC with Cloud \Shake}
 
-\begin{itemize}
-\item  Real-life example
-\item Many examples demonstrating the use of dynamic dependencies
-\end{itemize}
+Can reduce times a lot.
 
-\subsection{Experience}
+Needs to get much better dependencies.
 
-\begin{itemize}
-\item  Cloud stuff needed tracing infrastructure to expose dependencies.
-\item  Good profiling support (here are pictures) exposes places where we are over-sequential, and over-big tasks (build systems work best with lots of small tasks).
-\item Ability to compute critical paths
-\item Numbers: no-op rebuild time; fresh clone rebuild time; (vs clean build).
-\end{itemize}
+Also been deployed by Standard Chartered.

### 2019-04-25 papers/jfp/7-experience.tex
More work on S7

@@
 
-\todo{AM: Can you write this}
+\todo{AM: Can you flesh this out}
 
-The GHC build system, known as Hadrian, has continued development.
+The build system used to compile GHC, known as Hadrian, has continued development.
 The use of dynamic dependencies has made the build much easier, and it's now more maintainable.
@@
 The \Shake version has attracted more contributors, and been easier to modify, etc.
+The code is N lines.
 
@@
 
-\subsection{Experience from Cloud Shake}\label{sec-cloud-shake}
+\subsection{Experience from Cloud \Shake}\label{sec-cloud-shake}
+
+Converting \Shake into Cloud \Shake wasn't a difficult process armed with a roadmap. The key was the introduction of two new functions:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]] -> v -> [FilePath] -> IO ()
+lookupCloud :: (k -> m (Maybe (Hash v))) -> k -> Ver -> Ver -> m (Maybe (v, [[k]], IO ()))
+\end{minted}
+
+These functions are suspiciously like \hs{recordCT} and \hs{constructCT} from \S\ref{sec-constructive-traces}, with their differences perhaps the most illustrative of the changes required.
+(We have made the replacements, like replacing \hs{Key} for \hs{k}, to reduce irrelevant differences.)
 
 \begin{itemize}
-\item Implemented and released
+\item There are two \hs{Ver} arguments being passed to each operation.
+These are the versions of the build script, and the rule for this particular key.
+If either version changes then it is as though the key has changed, and nothing will match.
+These versions are important to avoid using stale build products from previous versions of the build script.
+\item The list of dependencies to \hs{addCloud} is a list of lists, rather than a simple list.
+The reason is that \Shake allows a list of dependencies to be supplied at once, so they can all be built in parallel.
+\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the files that this rule produces -- which must be declared with \hs{produces}, or the output keys from a rule.
+\item The \hs{lookupCloud} allows an explicit \hs{Nothing} when looking up a dependent key, since some keys are not buildable.
+\item The \hs{lookupCloud} returns at most one result, rather than a list. This change was made for simplicity.
+\end{itemize}
+
+The integration of these functions into \Shake is also interesting.
+We found the most expedient design was to leave \Shake with a verifying trace, but if the verifying trace doesn't work, we use then consult the constructive trace.
+By bolting constructive traces onto the side of Shake we avoid reengineering of the central database.
+We haven't found any significant downsides from the bolt-on approach thus far, so it may be a sensible route to go even if developing from scratch -- allowing an optimised verified trace implementation in many cases, and falling back to a more complex implementation only rarely.
+
+The one thing we haven't yet completed on the engineering side is a move to hosting caches over HTTP.
+At the moment all caches are on shared file systems.
+This approach can use mounted drives to mirror HTTP connections onto file systems, and reuse tools for managing file systems, share caches with rsync, and is simple.
+Unfortunately, on certain OS's (e.g.
+Windows) mounting an HTTP endpoint as a file system requires administrator privileges, so an HTTP cache is still required.
+
+\subsection{Experience from building GHC with Cloud \Shake}
+
+\begin{itemize}
 \item Absolute paths and system binaries
@@
 
-\subsection{Experience from building GHC with Cloud \Shake}
-
 Can reduce times a lot.

### 2019-04-25 papers/jfp/8-engineering.tex
Expand S8.1: Partial stores and exceptions

--- 2019-04-14 papers/jfp/6-engineering.tex
@@
 \section{Engineering aspects}\label{sec-engineering}
+
+\todo{AM}{Tasks with multiple outputs:
+  https://github.com/snowleopard/build/blob/master/src/Build/Multi.hs}
+
+\todo{AM}{Typed tasks:
+  https://github.com/snowleopard/build/blob/master/src/Build/Task/Typed.hs}
+
+\todo{AM}{Self-tracking (?):
+  https://github.com/snowleopard/build/blob/master/src/Build/SelfTracking.hs}
+
+\todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.}
 
@@
 values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
-model the errors inside the store and the \hs{Task}, but because \hs{v} is
-polymorphic for builds, it does not let the build system apply special behaviour
-for errors, e.g. early aborting.
+model the errors inside the store and the \hs{Task}, but because our build
+system models are polymorphic in \hs{v}, they cannot apply any special behaviour
+for errors, e.g. aborting the build early. There are many other ways of
+modelling errors, and in this section we show three simple examples.
+
+As mentioned above, we can include failures into the type of values \hs{v}, for
+example, to model a partial store we can use a simple like algebraic data type
+isomorphic to \hs{Maybe}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Value = FileNotFound | FileContents ByteString
+\end{minted}
+\vspace{1mm}
+
+\noindent
+This is convenient if \emph{tasks are aware of failures}. For example, a task
+may be able to cope with missing files, e.g. if \hs{fetch}~\hs{"username.txt"}
+returns \hs{FileNotFound}, the task could use the literal string \hs{"User"} as
+a default value. In this case the task will \emph{depend} on the fact that the
+file \cmd{username.txt} is missing, and will need to be rebuilt if the user
+later creates this file. In general, we can use values of type
+\hs{Either}~\hs{e}~\hs{v} when dealing with failures of type \hs{e}. To
+automatically convert a ``failure-free'' task into an equivalent task operating
+with values of type \hs{Either}~\hs{e}~\hs{v} we can use the following function:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+liftEither :: Task Monad k v -> Task Monad k (Either e v)
+liftEither task = Task $ \fetch ->
+    runExceptT $ run task (ExceptT . fetch)
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Here \hs{liftEither} wraps the result of the given
+\hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{Either}~\hs{e}~\hs{v} into \hs{EitherT}
+and then runs the task in the \hs{EitherT} monad transformer~\cite{liang1995monad}.
+
+Another approach is to include failures into the computation context \hs{f}. For
+example, we can choose \hs{f}~\hs{=}~\hs{Maybe}:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+fetchA1A2 :: String -> Maybe Integer
+fetchA1A2 k = Map.lookup k (Map.fromList [("A1", 10), ("A2", 20)])
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Since tasks are polymorphic in \hs{f}, we can directly run them with the
+callback \hs{fetchA1A2}. For example, the task description \hs{sprsh1}
+from~\S\ref{sec-task} will return \hs{Just}~\hs{(Just}~\hs{30)} for the task
+\cmd{B1 = A1 + A2}, \hs{Just}~\hs{Nothing} for the task \cmd{B2 = B1 * 2}
+because \hs{fetchA1A2} failed, and simply \hs{Nothing} for the input \cmd{A1}.
+
+This approach is convenient if \emph{tasks are not aware of failures}, e.g. we
+can model Excel formulas as pure arithmetic functions, and introduce failures
+``for free'' if/when needed by instantiating \hs{Tasks} with an appropriate
+\hs{f}.
+
+Finally, the task itself might not want to encode failures into the type of
+values \hs{v}, but instead \emph{demand that} \hs{f} \emph{has a built-in notion
+of failures}. This can be done by choosing a suitable constraint \hs{c}, such as
+\hs{Alternative}, \hs{MonadPlus} or even better something specific to failures,
+such as \hs{MonadFail}. Then both the callback and the task can reuse the same
+failure mechanism as shown below:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+class Monad m => MonadFail m where
+    fail :: String -> m a
+
+sprsh4 :: Tasks MonadFail String Integer
+sprsh4 "B1" = Just $ Task $ \fetch -> do
+    a1 <- fetch "A1"
+    a2 <- fetch "A2"
+    if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
+sprsh4 _ = Nothing
+\end{minted}
+\vspace{1mm}
+
+\noindent
+With this approach we can define a \hs{MonadFail} build system that can handle
+errors by aborting the build early.
+
+\todo{AM}{Add a MonadFail build system.}
 
@@
 We have given simple implementations assuming a single thread of execution,
-but all the build systems we address can actually build independent keys in parallel.
-While it complicates the model, the complications can be restricted exclusively
-to the scheduler:
+but all the build systems we address can actually build independent keys in
+parallel. While it complicates the model, the complications can be restricted
+exclusively to the scheduler:
 

### 2019-04-25 src/Build/Task/Monad.hs

@@ -4,7 +4,7 @@
 -- Dependencies of monadic tasks can only be discovered dynamically, i.e. during
 -- their execution.
 module Build.Task.Monad (
-    track, trackPure, isInput, computePure, compute, partial, exceptional
+    track, trackPure, isInput, computePure, compute, liftMaybe, liftEither
     ) where
 
 import Control.Monad.Trans
@@ -48,13 +48,13 @@ compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
 -- partial lookup function @k -> m (Maybe v)@. This essentially lifts the task
 -- from the type of values @v@ to @Maybe v@, where the result @Nothing@
 -- indicates that the task failed because of a missing dependency.
-partial :: Task Monad k v -> Task Monad k (Maybe v)
-partial task = Task $ \fetch -> runMaybeT $ run task (MaybeT . fetch)
+liftMaybe :: Task Monad k v -> Task Monad k (Maybe v)
+liftMaybe task = Task $ \fetch -> runMaybeT $ run task (MaybeT . fetch)
 
 -- | Convert a task with a total lookup function @k -> m v@ into a task with a
 -- lookup function that can throw exceptions @k -> m (Either e v)@. This
 -- essentially lifts the task from the type of values @v@ to @Either e v@, where
 -- the result @Left e@ indicates that the task failed because of a failed
 -- dependency lookup, and @Right v@ yeilds the value otherwise.
-exceptional :: Task Monad k v -> Task Monad k (Either e v)
-exceptional task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
+liftEither :: Task Monad k v -> Task Monad k (Either e v)
+liftEither task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)

### 2019-04-25 papers/jfp/7-experience.tex
Work on S7

@@
 As a set of authors, we've written a paper about the \Shake build system \cite{mitchell2012shake},
-applying that build system to GHC \cite{hadrian}, and a paper about the structure of build
-systems \cite{mokhov2018buildsystems}. The original purpose of that final paper was to create
+applying \Shake to building GHC \cite{hadrian}, and a paper about the categorisation of build
+systems \cite{mokhov2018buildsystems}. The original motivation of the research in that final paper was to create
 ``Cloud Shake'', and use it for the GHC build system. In this section we tie these papers together,
-reflecting on our experience after all these papers.
+reflecting on our experience since these papers.
 
@@
 Since that paper there have been roughly 5,000 commits to the \Shake project\footnote{\url{https://github.com/ndmitchell/shake}}.
-They add concepts like resources (if two rules content from a single external resource), rewrite serialisation to be faster, and documentation including a website, and add lots of tests.
+They add concepts like resources (if two rules content from a single external resource), rewrite serialisation to be faster, documentation including a website\footnote{\url{https://shakebuild.com}}, and add lots of tests.
 The biggest change in that time period was moving from blocking threads to continuations for the suspending scheduler.
@@
 
-We consider the lack of change to suggest that \Shake is based on underlying principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}.
+We consider the lack of change to suggest that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}.
 
@@
 
-Converting \Shake into Cloud \Shake wasn't a difficult process armed with a roadmap. The key was the introduction of two new functions:
+Converting \Shake into Cloud \Shake wasn't a difficult process once armed with a roadmap. The key was the introduction of two new functions:
 
@@
 These functions are suspiciously like \hs{recordCT} and \hs{constructCT} from \S\ref{sec-constructive-traces}, with their differences perhaps the most illustrative of the changes required.
-(We have made the replacements, like replacing \hs{Key} for \hs{k}, to reduce irrelevant differences.)
+(We have made some replacements from full \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant differences.)
 
@@
 
-\subsection{Experience from building GHC with Cloud \Shake}
+\subsection{Experience from using Cloud \Shake}
 
-\begin{itemize}
-\item Absolute paths and system binaries
-\item Deferred materialisation.
-Different invariants for (a) local builds, (b) cloud sharing (you must list all the things you produce), (c) sharing + deferred materialisation (you must declare all the things you consume).
-Interaction with early cut-off.
-\item  Cloud stuff needed tracing infrastructure to expose dependencies.
-\end{itemize}
+While we expected the GHC build system to be the first to take advantage of Cloud \Shake, we were actually beaten to it by Standard Chartered who report\footnote{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}:
 
-Can reduce times a lot.
+\begin{quote}
+Thanks for the symlinks release, we just finished upgrading this build system to use --share. ...
+Building from scratch with a warm cache takes around 5 seconds, saving us up to 2 hours. Not bad!
+\end{quote}
 
-Needs to get much better dependencies.
+We have also got the GHC build working with Cloud \Shake, resulting in build times of N minutes instead of M minutes.
+Due to aspects like \hs{configure} we don't see the dramatic reductions reported by Standard Chartered.
+Converting to a build suitable for sharing is not overly onerous, but nor is it trivial.
+Some things that require attention:
 
-Also been deployed by Standard Chartered.
+\begin{description}
+\item[Irrelevant differences] A common problem is that you don't get shared caching when you want it.
+To get caching, it is important that all machines are identical for the aspects that involve the build system.
+In particular, that means that all paths must be relative, and all dependencies on system binaries (e.g.
+\hs{gcc}) should either be controlled by the build system (e.g.
+using \Nix) or have tolerence to account for minor machine variations.
+\item[Insufficient produced files] A build rule must declare which files it produces, so these are stored in the cache and can be restored from the cache.
+If you don't use a remote cache then it is permissible to have files not known to the build system which change at the same time as files that are known.
+Once using a remote cache any file that is produced as a side-effect of a rule and then later used \emph{must} be declared.
+Failure to declare such files means that if some step produces a file, and a later step relies on it, that step breaks.
+In practice, most issues encountered during the move to Cloud builds for GHC were a lack of declared produced files.
+\item[Insufficient dependencies] While insufficient dependencies are often a problem, the move to a Cloud build makes them even more critical, as missing dependencies are likely to mean a result is \emph{never built locally}, rather than just build locally only once.
+Furthermore, dependencies in a Cloud setting must be more complete.
+For example, using the build system from Figure \ref{fig-make}, with a local build system without early cutoff, if the rule to produce \hs{main.exe} also made use of \hs{main.c} but didn't declare so, that isn't a problem.
+With the addition of early cutoff, if \hs{main.c} changes in an important way but \hs{main.o} doesn't update that causes issues, but is a somewhat rare occurance.
+With Cloud builds that becomes somewhat more common.
+\end{description}
+
+To help with the final two issues -- insufficient dependencies and produced files -- we have further enhanced the \Shake lint modes, coupling them to a program called \hs{FSATrace}, which detects which files are read/written by a command line execution.
+Such information has been very helpful in making the GHC build Cloud ready.

### 2019-04-25 papers/jfp/8-engineering.tex
Work on S8.1

@@
 Since tasks are polymorphic in \hs{f}, we can directly run them with the
-callback \hs{fetchA1A2}. For example, the task description \hs{sprsh1}
-from~\S\ref{sec-task} will return \hs{Just}~\hs{(Just}~\hs{30)} for the task
-\cmd{B1 = A1 + A2}, \hs{Just}~\hs{Nothing} for the task \cmd{B2 = B1 * 2}
-because \hs{fetchA1A2} failed, and simply \hs{Nothing} for the input \cmd{A1}.
+callback \hs{fetchA1A2}. For example, the task \cmd{B1 = A1 + A2} from
+\hs{sprsh1} in~\S\ref{sec-task} returns \hs{Just}~\hs{30}, whereas the task
+\cmd{B2 = B1 * 2} returns \hs{Nothing} because \hs{fetchA1A2} fails on \cmd{B1}.
 
@@
 \noindent
-With this approach we can define a \hs{MonadFail} build system that can handle
-errors by aborting the build early.
+With this approach we can implement a build system that accepts
+\hs{Tasks}~\hs{MonadFail}~\hs{k}~\hs{v} and handles errors by aborting the
+build early and returning
+\hs{Either}~\hs{String}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)} instead of just
+\hs{Store}~\hs{i}~\hs{k}~\hs{v} as in~\S\ref{sec-general-build}. One possible
+implementation is based on introducing an extra \hs{Either}~\hs{String} layer
+into the monad stack; while fairly direct it is also tedious, and hence we omit
+it.
 
-\todo{AM}{Add a MonadFail build system.}
+% \todo{AM}{Add a MonadFail build system.}
 

### 2019-04-25 papers/jfp/8-engineering.tex
Rearrange TODOs

@@
 \section{Engineering aspects}\label{sec-engineering}
-
-\todo{AM}{Tasks with multiple outputs:
-  https://github.com/snowleopard/build/blob/master/src/Build/Multi.hs}
-
-\todo{AM}{Typed tasks:
-  https://github.com/snowleopard/build/blob/master/src/Build/Task/Typed.hs}
-
-\todo{AM}{Self-tracking (?):
-  https://github.com/snowleopard/build/blob/master/src/Build/SelfTracking.hs}
-
-\todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.}
 
@@
 \hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus},
-respectively. More concretely, the following task description corresponds to
-a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}.
+respectively. Here is a task description corresponding to a spreadsheet with the
+formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}:
 
@@
 sprsh3 :: Tasks MonadPlus String Integer
-sprsh3 "B1"@\,@=@\,@Just $ Task $ \fetch -> (+) <$> fetch "A1" <*> (pure@\,@1 <|> pure@\,@2)
-sprsh3 _   @\,@=@\,@Nothing
+sprsh3 "B1" = Just $ Task $ \fetch -> (+) <$> fetch "A1"
+                                          <*> (pure 1 <|> pure 2)
+sprsh3 _    = Nothing
 \end{minted}
@@
 Such tasks can be modelled in our framework by adjusting the correctness
-definition (\S\ref{sec-build-correctness}): instead of requiring that the
+definition~(\S\ref{sec-build-correctness}), where instead of requiring that the
 \hs{result}ing value \emph{equals} the one obtained by recomputing the \hs{task}:
@@
 value (`ND' stands for `non-deterministic').
+
+\todo{AM}{Show implementation of \hs{computeND}}.
 
@@
 
+\todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.}
+
 \subsection{Self-tracking}\label{sec-tracking-aspects}
@@
 depending on itself.
+
+\todo{AM}{Self-tracking (?):
+  https://github.com/snowleopard/build/blob/master/src/Build/SelfTracking.hs}
 
@@
 polymorphic tasks.}.
+
+\todo{AM}{Tasks with multiple outputs:
+  https://github.com/snowleopard/build/blob/master/src/Build/Multi.hs}
+
+\todo{AM}{Typed tasks:
+  https://github.com/snowleopard/build/blob/master/src/Build/Task/Typed.hs}

### 2019-04-25 papers/jfp/2-background.tex
Make the tables look presentable

@@
 \centering
-\begin{tabular}{l||l|l||l|c|c|c}
+\begin{tabular}{llllccc}
 \hline
-$\!$Build system$\!$& Persistent build information & Scheduler   & Dependencies    & Minimal & Cutoff & Cloud$\!$\\\hline
-$\!$\Make       $\!$& File modification times      & Topological & Static          & Yes     & No     & No   $\!$\\
-$\!$\Excel      $\!$& Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   $\!$\\
-$\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   $\!$\\
-$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
+Build system\gap & Persistent build information\gap & Scheduler\gap   & Dependencies\gap    & Minimal\gap & Cutoff\gap & Cloud\\\hline
+\Make       & File modification times      & Topological\gap & Static          & Yes     & No     & No   \\
+\Excel      & Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   \\
+\Shake      & Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   \\
+\Bazel      & Cloud cache, command history\gap & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  \\\hline
 \end{tabular}

### 2019-04-25 papers/jfp/6-implementations.tex
Make the tables look presentable

@@
 \centering
-\begin{tabular}{l||c|c|c}
+\begin{tabular}{lccc}
 \hline
  & \multicolumn{3}{c}{Scheduling algorithm} \\
-Rebuilding strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
-\hline
-Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
-Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
-Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
+Rebuilding strategy\gap  & \gap{}Topological\gap\S\ref{sec-topological}\gap & \gap{}Restarting\gap\S\ref{sec-restarting}\gap & \gap{}Suspending\gap\S\ref{sec-suspending}\gap    \\\hline
+Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\
+Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\
+Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\
 Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline

### 2019-04-26 papers/jfp/8-engineering.tex
Work on S8

@@
 but neither is it beautiful or informative.
+
+\todo{NM}{Should we add a paragraph or two giving a bit more details about how
+parallelism is supported in \Shake?}
 
@@
 
-\vspace{0.5mm}
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
-\vspace{0.5mm}
+\vspace{1mm}
 
@@
 
-\todo{AM}{Self-tracking (?):
-  https://github.com/snowleopard/build/blob/master/src/Build/SelfTracking.hs}
+% \vspace{1mm}
+% \begin{minted}[xleftmargin=10pt]{haskell}
+% data Key k     = Key k   | KeyTask k
+% data Value v t = Value v | ValueTask t
+
+% fetchValue :: Functor f => (Key k -> f (Value v t)) -> k -> f v
+% fetchValue fetch key = extract <$> fetch (Key key)
+%   where
+%     extract (Value v) = v
+%     extract _ = error "Inconsistent fetch"
+
+% -- | Fetch a task description.
+% fetchValueTask :: Functor f => (Key k -> f (Value v t)) -> k -> f t
+% fetchValueTask fetch key = extract <$> fetch (KeyTask key)
+%   where
+%     extract (ValueTask t) = t
+%     extract _ = error "Inconsistent fetch"
+
+% -- | A model for 'Monad', works beautifully and allows storing the key on the
+% -- disk.
+% selfTrackingM :: forall k v t. (t -> Task Monad k v) -> Tasks Monad k t -> Tasks Monad (Key k) (Value v t)
+% selfTrackingM _      _     (KeyTask _) = Nothing -- Task keys are inputs
+% selfTrackingM parser tasks (Key     k) = runTask <$> tasks k
+%   where
+%     -- Fetch the task description, parse it, and then run the obtained task
+%     runTask :: Task Monad k t -> Task Monad (Key k) (Value v t)
+%     runTask act = Task $ \fetch -> do
+%         task <- parser <$> run act (fetchValueTask fetch)
+%         Value <$> run task (fetchValue fetch)
+% \end{minted}
+% \vspace{1mm}
 
@@
 values, allowing types that are stored only in the \Shake's build information.
+\Shake users have remarked that polymorphism provides a much easier expression
+of concepts, e.g. see~\cite{hadrian}.
 
@@
 build can express volatile dependencies and supports cutoff, the version number
-could equally be written to a file and used in a similar way.
+could equally be written to a file and used in a similar way, albeit at the cost
+of introducing additional correctness invariants, e.g. that the file
+\cmd{gcc-version.txt} does indeed contains two dot-separated integers.
+
+Generalised algebraic data types (GADTs)~\cite{spj2006gadts} can be used to
+extend our models to support ``typed'' build tasks. The key idea is to replace
+the monomorphic callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
+polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
+where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
+corresponding values. The idea is best explained by way of an example:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Version = Version { major :: Int, minor :: Int }
+    deriving (Eq, Ord)
+
+data Key a where
+    File       :: FilePath -> Key String
+    GccVersion :: Key Version
+\end{minted}
+\vspace{1mm}
+
+\noindent
+Here we extend the usual mapping from file paths to file contents with an
+additional key type \hs{GccVersion} mapped to a \hs{Version}. The task
+abstraction needs to be adjusted to cope with such keys (the prefix \hs{T}
+stands for ``typed''):
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Fetch k f = @\std{forall}@ v. k v -> f v
+
+newtype TTask c k v = TTask { run :: @\std{forall}@ f. c f => Fetch k f -> f v }
+
+type TTasks c k = @\std{forall}@ v. k v -> Maybe (TTask c k v)
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The changes compared to the definition in~\S\ref{sec-task} are minimal: (i) the
+\hs{TTask} now uses a typed \hs{Fetch} callback (we define a separate type
+synonym only for readability), and (ii) the type of \hs{TTasks} is now
+polymorphic in \hs{v} instead of being parameterised by a concrete \hs{v}. The
+example below demonstrates how \hs{fetch} can be used to retrieve dependencies
+of different types: the rule \cmd{release.txt} concatenates the contents of two
+\hs{File}s, while the rule \cmd{main.o} uses the numeric \hs{GccVersion} to
+determine the way the \hs{source} file should be compiled.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+example :: TTasks Monad Key
+example (File "release.txt") = Just $ TTask $ \fetch -> do
+    readme  <- fetch (File "README")
+    license <- fetch (File "LICENSE")
+    return (readme ++ license)
+example (File "main.o") = Just $ TTask $ \fetch -> do
+    let source = "main.c"
+    version <- fetch GccVersion
+    if version >= Version 8 0 then compileNew source
+                              else compileOld source
+example _ = Nothing
+\end{minted}
+\vspace{1mm}
 
@@
 
-\Shake users have remarked that polymorphism provides a much easier expression
-of concepts, e.g.~\cite{hadrian}, but it is not essential and we therefore do
-not model it in this paper\footnote{See modules \cmd{Build.Multi} and
-\cmd{Build.Task.Typed} in our framework for models of multiple-output and
-polymorphic tasks.}.
+% \vspace{1mm}
+% \begin{minted}[xleftmargin=10pt]{haskell}
+% type Partition k = k -> [k]
 
-\todo{AM}{Tasks with multiple outputs:
-  https://github.com/snowleopard/build/blob/master/src/Build/Multi.hs}
-
-\todo{AM}{Typed tasks:
-  https://github.com/snowleopard/build/blob/master/src/Build/Task/Typed.hs}
+% multi :: Eq k => Partition k -> Tasks Applicative [k] [v] -> Tasks Applicative [k] [v]
+% multi partition tasks keys
+%     | k:_ <- keys, partition k == keys = tasks keys
+%     | otherwise = Just $ Task $ \fetch ->
+%         sequenceA [ select k <$> fetch (partition k) | k <- keys ]
+%   where
+%     select k = fromMaybe (error msg) . lookup k . zip (partition k)
+%     msg = "Partition invariants violated"
+% \end{minted}
+% \vspace{1mm}

### 2019-04-26 src/Build/Task/Typed.hs

@@ -1,4 +1,4 @@
-{-# LANGUAGE CPP, ConstraintKinds, RankNTypes, GADTs, TypeFamilies #-}
+{-# LANGUAGE CPP, ConstraintKinds, RankNTypes, GADTs #-}
 #if __GLASGOW_HASKELL__ < 800
 {-# OPTIONS_GHC -Wno-unused-binds #-}
 #else
@@ -36,13 +36,43 @@ type ShowKey k = forall a. k a -> String
 dependencies :: ShowKey k -> Task Applicative k -> k a -> [String]
 dependencies showKey task = maybe [] getConst . task (\k -> Const [showKey k])
 
------------------------------------- Example -----------------------------------
+----------------------------- GCC versison example -----------------------------
+data Version = Version { major :: Int, minor :: Int }
+    deriving (Eq, Ord)
+
 data Key a where
-    Base       :: Key Int
-    Number     :: Key Int
-    SplitDigit :: Key (Int, Int)
-    LastDigit  :: Key Int
-    BaseDigits :: Key [Int]
+    File       :: FilePath -> Key String
+    GccVersion :: Key Version
+
+newtype Task1 c k v = Task1 { run :: forall f. c f => Fetch k f -> f v }
+
+type Tasks1 c k = forall a. k a -> Maybe (Task1 c k a)
+
+example :: Tasks1 Monad Key
+example (File "release.txt") = Just $ Task1 $ \fetch -> do
+    readme  <- fetch (File "README")
+    license <- fetch (File "LICENSE")
+    return (readme ++ license)
+example (File "main.o") = Just $ Task1 $ \fetch -> do
+    let source = "main.c"
+    version <- fetch GccVersion
+    if version >= Version 8 0 then compileNew source
+                              else compileOld source
+example _ = Nothing
+
+compileNew :: String -> f String
+compileNew = undefined
+
+compileOld :: String -> f String
+compileOld = undefined
+
+------------------------------------ Example -----------------------------------
+data KeyN a where
+    Base       :: KeyN Int
+    Number     :: KeyN Int
+    SplitDigit :: KeyN (Int, Int)
+    LastDigit  :: KeyN Int
+    BaseDigits :: KeyN [Int]
 
 -- | A build task for some simple typed numeric calculations. We can perform
 -- static analysis of this task using the function 'dependencies'. For example:
@@ -51,14 +81,14 @@ data Key a where
[…]

### 2019-04-26 test/Examples.hs

@@ -1,3 +1,4 @@
+{-# LANGUAGE ConstraintKinds, RankNTypes, GADTs #-}
 module Examples where
 
 import Build.Task
@@ -133,9 +134,9 @@ staticIF _ _    = Nothing
 
 -------------------------- Dynamic programming example -------------------------
 
-data Key = A Int | B Int | C Int Int deriving Eq
+data K = A Int | B Int | C Int Int deriving Eq
 
-editDistance :: Tasks Monad Key Int
+editDistance :: Tasks Monad K Int
 editDistance (C i 0) = Just $ Task $ const $ pure i
 editDistance (C 0 j) = Just $ Task $ const $ pure j
 editDistance (C i j) = Just $ Task $ \fetch -> do

### 2019-04-26 papers/jfp/8-engineering.tex
Finish S8.7 (Polymorphism and multiple-output build tasks)

@@
 
-\subsection{Polymorphism}\label{sec-polymorphism}
+\subsection{Polymorphism and Multiple-Output Build Tasks}\label{sec-polymorphism}
 
@@
 A more compelling example is build tasks that produce multiple output
-keys~--~for example, \cmd{ghc Foo.hs} produces both \cmd{Foo.hi} and \cmd{Foo.o}.
+keys~--~for example, \cmd{ghc A.hs} produces both \cmd{A.hi} and \cmd{A.o}.
 That can be represented by having a key whose value is a pair of file names, and
-whose result is a pair of file contents. From that, the task for \cmd{Foo.hi}
+whose result is a pair of file contents. From that, the task for \cmd{A.hi}
 can be the first component of the result of the pair. Again, such an operation
@@
 
-% \vspace{1mm}
-% \begin{minted}[xleftmargin=10pt]{haskell}
-% type Partition k = k -> [k]
+Note that multiple-output build tasks require one more convenience mechanism,
+which is orthogonal to polymorphism. Multiple-output task descriptions partition
+the set of keys into \emph{elementary subsets} e.g.
+$\{\cmd{A.hi},\cmd{A.o}\} \rightarrow \cmd{ghcTask}$,
+$\{\cmd{B.o}\} \rightarrow \cmd{gccTask}$, etc. It is very convenient to
+allow to fetch arbitrary sets of keys, such as $\{\cmd{A.o},\cmd{B.o}\}$,
+without manually decomposing them into appropriate elementary subsets. This
+requires a method to efficiently construct task descriptions operating on
+arbitrary sets of keys rather on elementary subsets. One way to implement such
+``multiple-output task closure'' is shown below.
 
-% multi :: Eq k => Partition k -> Tasks Applicative [k] [v] -> Tasks Applicative [k] [v]
-% multi partition tasks keys
-%     | k:_ <- keys, partition k == keys = tasks keys
-%     | otherwise = Just $ Task $ \fetch ->
-%         sequenceA [ select k <$> fetch (partition k) | k <- keys ]
-%   where
-%     select k = fromMaybe (error msg) . lookup k . zip (partition k)
-%     msg = "Partition invariants violated"
-% \end{minted}
-% \vspace{1mm}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+type Partition k = k -> [k]
+
+multi :: Eq k => Partition k -> Tasks Applicative [k] [v]
+                             -> Tasks Applicative [k] [v]
+multi partition tasks keys
+    | k:_ <- keys, partition k == keys = tasks keys
+    | otherwise = Just $ Task $ \fetch ->
+        sequenceA [ select k <$> fetch (partition k) | k <- keys ]
+  where
+    select k = fromMaybe (error msg) . lookup k . zip (partition k)
+    msg = "Partition invariants violated"
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The function \hs{multi} takes a task description \hs{tasks} that operates on
+elementary key subsets, and computes its closure supporting all possible
+combinations of keys. The first argument \hs{partition} specifies the elementary
+key subsets supported by the original \hs{tasks}. There are two cases: (i) if a
+given set of \hs{keys} is an elementary subset, we pass the query to \hs{tasks};
+(ii) otherwise, we examine every key \hs{k} in \hs{keys}, \hs{fetch} its
+corresponding elementary subset, and \hs{select} the appropriate value from the
+result. Returning to the above example, when given the query
+$\{\cmd{A.o},\cmd{B.o}\}$, this function will \hs{fetch} elementary subsets
+$\{\cmd{A.hi},\cmd{A.o}\}$ and $\{\cmd{B.o}\}$, thereby depending on the
+corresponding build tasks \cmd{ghcTask} and \cmd{gccTask}.
+
+We chose not to use polymorphic tasks in the above implementation for
+simplicity, but nothing prevents us from combining polymorphism with
+multiple-output task closure.

### 2019-04-26 papers/jfp/3-abstractions.tex
Expand S8.1: Partial stores and exceptions
Tweaks to S3

@@
 
-This section presents purely functional abstractions that allow us to express
+This section presents abstractions that allow us to express
 all the intricacies of build systems discussed in~\S\ref{sec-background},
-and design complex build systems from simple primitives. Specifically, we
+and express complex build systems from simple primitives. Specifically, we
 present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
@@
 
-We will model build systems concretely, as Haskell programs. To that end,
+We model build systems concretely, as Haskell programs. To that end,
 Fig.~\ref{fig-types} provides the type signatures for all key abstractions
 introduced in the paper. For example, \hs{Store}~\hs{i}~\hs{k}~\hs{v} is the
-type of stores, with several associate functions (\hs{getValue}, etc.). We use
+type of stores, with several associated functions (\hs{getValue}, etc.). We use
 \hs{k} as a type variable ranging over keys, \hs{v} for values, and \hs{i} for
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
+newtype Task  c k v = Task
+    { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
 type    Tasks c k v = k -> Maybe (Task c k v)
@@
 \hs{Task} to every non-input key; input keys are associated with \hs{Nothing}.
-The highly-abstracted type \hs{Task} describes how to build a value when given
+The highly-abstracted type \hs{Task} describes how to build a value given
 a way to build its dependencies, and is best explained by an example.
@@
 instance Monoid m => Applicative (Const m) where
-    pure _              = Const mempty   -- mempty is the identity of the monoid m
-    Const x <*> Const y = Const (x <> y) -- <> is the binary operation of the monoid m
+    pure _              = Const mempty   -- mempty is identity for monoid m
+    Const x <*> Const y = Const (x <> y) -- <> is the binary operation for m
 \end{minted}
@@
 sprsh1 :: Tasks Applicative String Integer
-sprsh1 "B1" = Just $ Task $ \fetch -> ((+)  <$> fetch "A1" <*> fetch "A2")
-sprsh1 "B2" = Just $ Task $ \fetch -> ((*2) <$> fetch "B1")
+sprsh1 "B1" = Just $ Task $ \fetch ->
+    ((+)  <$> fetch "A1" <*> fetch "A2")
+sprsh1 "B2" = Just $ Task $ \fetch ->
+    ((*2) <$> fetch "B1")
 sprsh1 _    = Nothing
@@
 place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
-\hs{fetch} \emph{different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.
+\hs{fetch} \emph{different keys} depending on whether or not \hs{c1}~\hs{==}~\hs{1}.
 
@@
 compute :: Task Monad k v -> Store i k v -> v
-compute task store = runIdentity $ run task (\k -> Identity@\,@(getValue k store))
+compute task store = runIdentity $ run task $
+    \k -> Identity $ getValue k store
 \end{minted}
@@
 monad: the function
-\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
+\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{\$}~\hs{getValue}~\hs{k}~\hs{store} has the type
 \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
@@
 
-The presented definition of correctness needs to be adjusted to accommodate
+The presented definition of correctness needs to be adjusted for
 non-deterministic tasks and shallow cloud builds, as will be discussed
@@
 unnecessarily; all of its 19,729 decimal digits have already been helpfully
-computed, e.g. see~\cite{ackermann42}. We will use this technique in our models
-of cloud build systems in~\S\ref{sec-implementations}.
+computed, e.g. see~\cite{ackermann42}. We will use \hs{MonadState} tasks in our
+models of cloud build systems in~\S\ref{sec-implementations}.

### 2019-04-26 papers/jfp/7-experience.tex
Link in the Shake precision in this paper

@@
 We consider the lack of change to suggest that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}.
+There are two main aspects of the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, the first is the rebuilder, which is described as verifying step traces in \S\ref{sec-step-traces}.
+The second is the power of \Shake tasks (called actions in the original paper.
+Originally these tasks were described as being equivalent to:
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Action k v = Finished v
+                | Depends v (v -> Action)
+\end{minted}
+
+In some sense this is the ``free'' version of the \hs{Task Monad} -- the two types form an isomorphism. Similary, \Make tasks were described as equivalent to
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Rule k v = Rule {depends :: [k], action :: [v] -> v}
+\end{minted}
+
+Once again, this type is isomorphism to \hs{Task Applicative}.
+By describing these types using \hs{Task} we are able to pinpoint the differences (\hs{Monad} vs \hs{Applicative}) and use existing literature to prove what is and isn't possible.
 

### 2019-04-26 src/Build/SelfTracking/Typed.hs
Add a typed variant of selfTracking

@@ -1,0 +1,29 @@
+{-# LANGUAGE ConstraintKinds, GADTs, RankNTypes, ScopedTypeVariables #-}
+
+module Build.SelfTracking.Typed (
+    Fetch, TaskT (..), TasksT, Key (..), selfTracking
+  ) where
+
+import Build.Task
+
+type Fetch k f = forall v. k v -> f v
+
+newtype TaskT c k v = TaskT { runT :: forall f. c f => Fetch k f -> f v }
+
+type TasksT c k = forall v. k v -> Maybe (TaskT c k v)
+
+-- | The type variable @s@ stands for "scripts" written in some task description
+-- language.
+data Key k v s a where
+    Script :: k -> Key k v s s -- Keys for build scripts
+    Value  :: k -> Key k v s v -- Keys for all other values
+
+selfTracking :: forall k v s. (s -> Task Monad k v) -> Tasks Monad k s -> TasksT Monad (Key k v s)
+selfTracking _     _     (Script _) = Nothing -- Scripts are inputs
+selfTracking parse tasks (Value  k) = runScript <$> tasks k
+  where
+    -- Fetch the script, parse it, and then run the obtained task
+    runScript :: Task Monad k s -> TaskT Monad (Key k v s) v
+    runScript task = TaskT $ \fetch -> do
+        script <- run task (fetch . Script)
+        run (parse script) (fetch . Value)

### 2019-04-26 src/Build/Task/Typed.hs

@@ -44,16 +44,16 @@ data Key a where
     File       :: FilePath -> Key String
     GccVersion :: Key Version
 
-newtype Task1 c k v = Task1 { run :: forall f. c f => Fetch k f -> f v }
+newtype TaskT c k v = TaskT { runT :: forall f. c f => Fetch k f -> f v }
 
-type Tasks1 c k = forall a. k a -> Maybe (Task1 c k a)
+type TasksT c k = forall a. k a -> Maybe (TaskT c k a)
 
-example :: Tasks1 Monad Key
-example (File "release.txt") = Just $ Task1 $ \fetch -> do
+example :: TasksT Monad Key
+example (File "release.txt") = Just $ TaskT $ \fetch -> do
     readme  <- fetch (File "README")
     license <- fetch (File "LICENSE")
     return (readme ++ license)
-example (File "main.o") = Just $ Task1 $ \fetch -> do
+example (File "main.o") = Just $ TaskT $ \fetch -> do
     let source = "main.c"
     version <- fetch GccVersion
     if version >= Version 8 0 then compileNew source

### 2019-04-26 papers/jfp/8-engineering.tex
Move self-tracking to S8.7 and add an implementation

@@
 
-\subsection{Self-tracking}\label{sec-tracking-aspects}
-
-Some build systems, for example \Excel and \Ninja, are capable of recomputing a
-task if either its dependencies change, \emph{or} the task itself changes. For
-example:
-
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{text}
-A1 = 20      B1 = A1 + A2
-A2 = 10
-\end{minted}
-\vspace{0.5mm}
-
-\noindent
-In \Excel the user can alter the value produced by \cmd{B1} by either editing
-the inputs of \cmd{A1} or \cmd{A2}, \emph{or} editing the formula in
-\cmd{B1}~--~e.g. to \cmd{A1 * A2}. This pattern can be captured by describing
-the rule producing \cmd{B1} as also depending on the value \cmd{B1-formula}.
-The implementation can be given very directly in a
-\hs{Tasks}~\hs{Monad}~--~concretely, first look up the formula, then interpret
-it:
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-sprsh4 "B1" = Just $ Task $ \fetch -> do
-    formula <- fetch "B1-formula"
-    evalFormula fetch formula
-\end{minted}
-\vspace{1mm}
-
-\noindent
-The build systems that have precise self-tracking are all ones which use a
-\emph{non-embedded domain specific language} to describe build tasks. Those
-which use a full programming language, e.g. \Shake, are faced with the challenge
-of implementing equality on arbitrary task functions. For such build systems,
-the pessimistic assumption that any change to the build system potentially
-changes any build task can often be used~--~the classic example being a makefile
-depending on itself.
-
-% \vspace{1mm}
-% \begin{minted}[xleftmargin=10pt]{haskell}
-% data Key k     = Key k   | KeyTask k
-% data Value v t = Value v | ValueTask t
-
-% fetchValue :: Functor f => (Key k -> f (Value v t)) -> k -> f v
-% fetchValue fetch key = extract <$> fetch (Key key)
-%   where
-%     extract (Value v) = v
-%     extract _ = error "Inconsistent fetch"
-
-% -- | Fetch a task description.
-% fetchValueTask :: Functor f => (Key k -> f (Value v t)) -> k -> f t
-% fetchValueTask fetch key = extract <$> fetch (KeyTask key)
-%   where
-%     extract (ValueTask t) = t
-%     extract _ = error "Inconsistent fetch"
-
-% -- | A model for 'Monad', works beautifully and allows storing the key on the
-% -- disk.
-% selfTrackingM :: forall k v t. (t -> Task Monad k v) -> Tasks Monad k t -> Tasks Monad (Key k) (Value v t)
-% selfTrackingM _      _     (KeyTask _) = Nothing -- Task keys are inputs
-% selfTrackingM parser tasks (Key     k) = runTask <$> tasks k
-%   where
-%     -- Fetch the task description, parse it, and then run the obtained task
-%     runTask :: Task Monad k t -> Task Monad (Key k) (Value v t)
-%     runTask act = Task $ \fetch -> do
-%         task <- parser <$> run act (fetchValueTask fetch)
-%         Value <$> run task (fetchValue fetch)
-% \end{minted}
-% \vspace{1mm}
-
 \subsection{Iterative Computations}\label{sec-iterative-compute}
@@
 additional key type \hs{GccVersion} mapped to a \hs{Version}. The task
-abstraction needs to be adjusted to cope with such keys (the prefix \hs{T}
+abstraction needs to be adjusted to cope with such keys (the suffix \hs{T}
 stands for ``typed''):
@@
 
-newtype TTask c k v = TTask { run :: @\std{forall}@ f. c f => Fetch k f -> f v }
+newtype TaskT c k v = TaskT { runT :: @\std{forall}@ f. c f => Fetch k f -> f v }
 
-type TTasks c k = @\std{forall}@ v. k v -> Maybe (TTask c k v)
+type TasksT c k = @\std{forall}@ v. k v -> Maybe (TaskT c k v)
 \end{minted}
@@
 The changes compared to the definition in~\S\ref{sec-task} are minimal: (i) the
-\hs{TTask} now uses a typed \hs{Fetch} callback (we define a separate type
-synonym only for readability), and (ii) the type of \hs{TTasks} is now
+\hs{TaskT} now uses a typed \hs{Fetch} callback (we define a separate type
+synonym only for readability), and (ii) the type of \hs{TasksT} is now
 polymorphic in \hs{v} instead of being parameterised by a concrete \hs{v}. The
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-example :: TTasks Monad Key
-example (File "release.txt") = Just $ TTask $ \fetch -> do
+example :: TasksT Monad Key
+example (File "release.txt") = Just $ TaskT $ \fetch -> do
     readme  <- fetch (File "README")
@@
     return (readme ++ license)
-example (File "main.o") = Just $ TTask $ \fetch -> do
+example (File "main.o") = Just $ TaskT $ \fetch -> do
     let source = "main.c"
@@
 multiple-output task closure.
+
+\todo{AM/NM}{We are talking about sets above, but are using lists, which is a
+bit inconsistent. We could switch from \hs{[k]} to \hs{Set k} fairly easily.
+Shall we?}
+
+\subsection{Self-tracking}\label{sec-tracking-aspects}
+
+Some build systems, for example \Excel and \Ninja, are capable of recomputing a
+task if either its dependencies change, \emph{or} the task itself changes. For
+example:
+
+\vspace{0.5mm}
+\begin{minted}[xleftmargin=10pt]{text}
+A1 = 20      B1 = A1 + A2
+A2 = 10
+\end{minted}
+\vspace{0.5mm}
+
+\noindent
+In \Excel the user can alter the value produced by \cmd{B1} by either editing
+the inputs of \cmd{A1} or \cmd{A2}, \emph{or} editing the formula in
+\cmd{B1}~--~e.g. to \cmd{A1 * A2}. This pattern can be captured by describing
+the rule producing \cmd{B1} as also depending on the value \cmd{B1-formula}.
+The implementation can be given very directly in a
+\hs{Tasks}~\hs{Monad}~--~concretely, first look up the formula, then interpret
+it:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+sprsh4 "B1" = Just $ Task $ \fetch -> do
+    formula <- fetch "B1-formula"
+    evalFormula fetch formula
+\end{minted}
+\vspace{1mm}
+
+\noindent
+The build systems that have precise self-tracking are all ones which use a
+\emph{non-embedded domain specific language} to describe build tasks. Those
+which use a full programming language, e.g. \Shake, are faced with the challenge
+of implementing equality on arbitrary task functions. For such build systems,
+the pessimistic assumption that any change to the build system potentially
+changes any build task can often be used~--~the classic example being a makefile
+depending on itself.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Key k v s a where
+    Script :: k -> Key k v s s -- Keys for build scripts
+    Value  :: k -> Key k v s v -- Keys for all other values
+
+selfTracking :: (s -> Task Monad k v) -> Tasks Monad k s
+                                      -> TasksT Monad (Key k v s)
+selfTracking _     _     (Script _) = Nothing -- Scripts are inputs
+selfTracking parse tasks (Value  k) = runScript <$> tasks k
+  where
+    -- Fetch the script, parse it, and then run the obtained task
+    runScript :: Task Monad k s -> TaskT Monad (Key k v s) v
+    runScript task = TaskT $ \fetch -> do
+        script <- run task (fetch . Script)
+        run (parse script) (fetch . Value)
+\end{minted}
+\vspace{1mm}

### 2019-04-26 papers/jfp/5-rebuilders.tex
Link in the Shake precision in this paper
Minor tweaks

@@
 \begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
-recordVT@\,@::@\,@k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
-verifyVT@\,@::@\,@(Monad m,@\,@Eq k,@\,@Eq v) => k -> Hash v -> (k -> m (Hash v)) -> VT k v -> m Bool
+recordVT :: k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
+verifyVT :: (Monad m, Eq k, Eq v)
+         => k -> Hash v -> (k -> m (Hash v)) -> VT k v -> m Bool
 \end{minted}
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Trace k v = Trace { key :: k, depends :: [(k, Hash v)], result :: Hash v }
+data Trace k v = Trace
+    {key :: k
+    ,depends :: [(k, Hash v)]
+    ,result :: Hash v
+    }
 \end{minted}
@@
 
-\subsubsection{Step Verifying Traces}
+\subsubsection{Verifying Step Traces}\label{sec-step-traces}
 
@@
 recordCT    :: k -> v -> [(k, Hash v)] -> CT k v -> CT k v
-constructCT :: (Monad m, Eq k, Eq v) => k -> (k -> m (Hash v)) -> CT k v -> m [v]
+constructCT :: (Monad m, Eq k, Eq v)
+            => k -> (k -> m (Hash v)) -> CT k v -> m [v]
 \end{minted}

### 2019-04-26 papers/jfp/7-experience.tex
Minor tweaks

@@
 data Action k v = Finished v
-                | Depends v (v -> Action)
+                | Depends v (v -> Action k v)
 \end{minted}
@@
 
-Once again, this type is isomorphism to \hs{Task Applicative}.
+Once again, this type is isomorphiv to \hs{Task Applicative}.
 By describing these types using \hs{Task} we are able to pinpoint the differences (\hs{Monad} vs \hs{Applicative}) and use existing literature to prove what is and isn't possible.

### 2019-04-27 papers/jfp/8-engineering.tex
Finish S8.7: Self-tracking

@@
 
+Below we show how to implement self-tracking in a build system that allows users
+to describe build tasks by \emph{scripts} written in a non-embedded domain
+specific language. We will denote the type of scripts by \hs{s}, and will assume
+that scripts are indexed by keys \hs{k} just like all other values \hs{v}. More
+specifically, we use the following GADT to tag keys \hs{k} with corresponding
+result types: \hs{s} for scripts, and \hs{v} for other values.
+
 \vspace{1mm}
@@
     Value  :: k -> Key k v s v -- Keys for all other values
+\end{minted}
+\vspace{1mm}
 
+\noindent
+The function \hs{selfTracking} defined below is a generalisation of the approach
+explained in the above \Excel example \hs{sprsh4}. The function takes a parser
+for scripts, of type \hs{s}~\hs{->}~\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}, and a
+description of \emph{how to build all scripts}, of type
+\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{s}. For \hs{sprsh4}, the latter would simply
+fetch \cmd{B1-formula} when given the key \cmd{B1} and return \hs{Nothing}
+otherwise, but the presented approach can cope with much more sophisticated
+scenarios where scripts themselves are derived from ``script sources'', e.g. all
+C compilation scripts can be obtained from a single pattern rule, such as
+\cmd{gcc -c [source] -o [object]}. The resulting typed task description
+\hs{TasksT}~\hs{Monad}~\hs{(Key}~\hs{k}~\hs{v}~\hs{s)} tracks both values and
+scripts that compute them, and is therefore self-tracking.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 selfTracking :: (s -> Task Monad k v) -> Tasks Monad k s
@@
 \vspace{1mm}
+
+\noindent
+It is possible to implement \hs{selfTracking} without relying on GADTs and typed
+tasks presented in~\S\ref{sec-polymorphism}, but the cost of using partial
+functions.

### 2019-04-27 papers/jfp/7-experience.tex
Clean up the section on relationship to Shake tasks

--- 2019-04-26 papers/jfp/7-experience.tex
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Action k v = Finished v
-                | Depends v (v -> Action)
+data Action k v r = Finished r
+                  | Depends v (v -> Action k r)
 \end{minted}
 
-In some sense this is the ``free'' version of the \hs{Task Monad} -- the two types form an isomorphism. Similary, \Make tasks were described as equivalent to
+In some sense this is the ``free'' version of the \hs{Task Monad} \cite{free-applicatives} -- the two types form an isomorphism.
+The originally paper uses concrete types \hs{Key} and \hs{Value}, we generalise these types to \hs{k} and \hs{v}, and also add \hs{r} so that \hs{Action k v} can be an instance of \hs{Monad}.
+We can write that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type:
 
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Rule k v = Rule {depends :: [k], action :: [v] -> v}
+instance Monad (Action k v) where
+    return = Finished
+    Finished x >>= f = f x
+    Depends ds op >>= f = Depends ds $ \v -> f =<< op v
+
+toAction :: Task Monad k v -> Action k v v
+toAction (Task run) = run $ \k -> Depends k Finished
+
+fromAction :: Action k v v -> Task Monad k v
+fromAction x = Task $ \fetch -> f fetch x
+    where f fetch (Finished v) = return v
+          f fetch (Depends d op) = f fetch . op =<< fetch d
 \end{minted}
 
-Once again, this type is isomorphism to \hs{Task Applicative}.
+Similary, \Make tasks were described as equivalent to
+
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Rule k v r = Rule {depends :: [k], action :: [v] -> r}
+\end{minted}
+
+Once again, this type is isomorphic to \hs{Task Applicative}, and we can define a similar \hs{Applicative} instance and conversion functions.
 By describing these types using \hs{Task} we are able to pinpoint the differences (\hs{Monad} vs \hs{Applicative}) and use existing literature to prove what is and isn't possible.

### 2019-04-27 papers/jfp/7-experience.tex
Use \Cloud \Shake consistently

@@
 
-\subsection{Experience from Cloud \Shake}\label{sec-cloud-shake}
+\subsection{Experience from \Cloud \Shake}\label{sec-cloud-shake}
 
-Converting \Shake into Cloud \Shake wasn't a difficult process once armed with a roadmap. The key was the introduction of two new functions:
+Converting \Shake into \Cloud \Shake wasn't a difficult process once armed with a roadmap. The key was the introduction of two new functions:
 
@@
 
-\subsection{Experience from using Cloud \Shake}
+\subsection{Experience from using \Cloud \Shake}
 
-While we expected the GHC build system to be the first to take advantage of Cloud \Shake, we were actually beaten to it by Standard Chartered who report\footnote{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}:
+While we expected the GHC build system to be the first to take advantage of \Cloud \Shake, we were actually beaten to it by Standard Chartered who report\footnote{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}:
 
@@
 
-We have also got the GHC build working with Cloud \Shake, resulting in build times of N minutes instead of M minutes.
+We have also got the GHC build working with \Cloud \Shake, resulting in build times of N minutes instead of M minutes.
 Due to aspects like \hs{configure} we don't see the dramatic reductions reported by Standard Chartered.

### 2019-04-27 papers/jfp/7-experience.tex
Proofing and tweaking

@@
 systems \cite{mokhov2018buildsystems}. The original motivation of the research in that final paper was to create
-``Cloud Shake'', and use it for the GHC build system. In this section we tie these papers together,
+``\Cloud \Shake'', and apply it to the GHC build system. In this section we tie these papers together,
 reflecting on our experience since these papers.
@@
 
-The original design of \Shake hasn't changed since the initial paper.
-Since that paper there have been roughly 5,000 commits to the \Shake project\footnote{\url{https://github.com/ndmitchell/shake}}.
-They add concepts like resources (if two rules content from a single external resource), rewrite serialisation to be faster, documentation including a website\footnote{\url{https://shakebuild.com}}, and add lots of tests.
+The original design of \Shake hasn't changed since the initial paper, but the implementation has continued to mature -- there have been roughly 5,000 subsequent commits to the \Shake project\footnote{\url{https://github.com/ndmitchell/shake}}.
+These commits add concepts like resources (if two rules contend on a single external resource), rewriting serialisation to be faster, documentation including a website\footnote{\url{https://shakebuild.com}}, and add lots of tests.
 The biggest change in that time period was moving from blocking threads to continuations for the suspending scheduler.
-The most visible change is switching \hs{*>} for \hs{\%>}\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.
+The most visible change is switching \hs{*>} to \hs{%>} for defining rules\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.
 
-We consider the lack of change to suggest that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}.
-There are two main aspects of the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, the first is the rebuilder, which is described as verifying step traces in \S\ref{sec-step-traces}.
-The second is the power of \Shake tasks (called actions in the original paper.
-Originally these tasks were described as being equivalent to:
+We consider the lack of change suggestive that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}.
+There are two main aspects to the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, Firstly, the rebuilder can now be described as verifying step traces as per \S\ref{sec-step-traces}, with a much clearer relationship to the unoptimised verifying traces in \S\ref{sec-verifying-traces}.
+The second is the power of \Shake tasks (called actions in the original paper).
+Originally these tasks were described using the data type:
 
@@
 data Action k v r = Finished r
-                  | Depends v (v -> Action k r)
+                  | Depends v (v -> Action k v r)
 \end{minted}
 
-In some sense this is the ``free'' version of the \hs{Task Monad} \cite{free-applicatives} -- the two types form an isomorphism.
-The originally paper uses concrete types \hs{Key} and \hs{Value}, we generalise these types to \hs{k} and \hs{v}, and also add \hs{r} so that \hs{Action k v} can be an instance of \hs{Monad}.
-We can write that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type:
+In some sense, this is the ``free'' version of \hs{Task Monad} \cite{free-applicatives} -- the two types form an isomorphism.
+The original paper uses concrete types \hs{Key} and \hs{Value}, we generalise these types to \hs{k} and \hs{v}, and also add \hs{r} so that \hs{Action k v} can be an instance of \hs{Monad}.
+We define that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type:
 
@@
 
-Similary, \Make tasks were described as equivalent to
+\noindent Similarly, \Make tasks were described as equivalent to
 
@@
 Once again, this type is isomorphic to \hs{Task Applicative}, and we can define a similar \hs{Applicative} instance and conversion functions.
-By describing these types using \hs{Task} we are able to pinpoint the differences (\hs{Monad} vs \hs{Applicative}) and use existing literature to prove what is and isn't possible.
+By describing these types using \hs{Task} we are able to describe the differences more concisely (\hs{Monad} vs \hs{Applicative}) and use existing literature to determine what is and isn't possible.
 
@@
 
-Converting \Shake into \Cloud \Shake wasn't a difficult process once armed with a roadmap. The key was the introduction of two new functions:
+Converting \Shake into \Cloud \Shake wasn't a difficult process once armed with a roadmap from \cite{mokhov2018buildsystems}. The key was the introduction of two new functions:
 
 \begin{minted}[xleftmargin=10pt]{haskell}
-addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]] -> v -> [FilePath] -> IO ()
-lookupCloud :: (k -> m (Maybe (Hash v))) -> k -> Ver -> Ver -> m (Maybe (v, [[k]], IO ()))
+addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]]
+         -> v -> [FilePath] -> IO ()
+lookupCloud :: (k -> m (Maybe (Hash v))) -> k -> Ver -> Ver
+            -> m (Maybe (v, [[k]], IO ()))
 \end{minted}
@@
 
-The integration of these functions into \Shake is also interesting.
-We found the most expedient design was to leave \Shake with a verifying trace, but if the verifying trace doesn't work, we use then consult the constructive trace.
+The integrate these functions into \Shake we found the most expedient design was to leave \Shake with a verifying trace, but if the verifying trace doesn't skip building, we consult the constructive trace.
 By bolting constructive traces onto the side of Shake we avoid reengineering of the central database.

### 2019-04-27 papers/jfp/1-intro.tex
Polishing S1-S3

@@
 
-In this paper we give a general framework (refined from \cite{mokhov2018buildsystems}) in which to understand and compare
-build systems, in a way that is both abstract (omitting incidental detail)
-and yet precise (implemented as Haskell code).
-With this framework we can explain confusing details from the original Shake paper \cite{mitchell2012shake} and build new features useful for the GHC build system \cite{hadrian}.
-Specifically we make these contributions:
+In this paper we give a general framework in which to understand and compare
+build systems, in a way that is both abstract (omitting incidental detail) and
+yet precise (implemented as Haskell code). Specifically we make these
+contributions:
+
 \begin{itemize}
@@
 of a variety of different real-life build systems, including \Make, \Shake,
-\Bazel, \CloudBuild, \Buck, \Nix, and \Excel\footnote{\Excel appears very
-different to the others but, seen through the lens of this paper, it is very
-close indeed.}, each by the composition of the two design choices
-(\S\ref{sec-implementations}). Doing this modelling in a single setting
-allows the differences and similarities between these huge systems to be
-brought out clearly\footnote{All our models are executable and are available on
-Hackage: \url{https://hackage.haskell.org/package/build-1.0}.}.
+\Bazel, \Buck, \Nix, and \Excel\footnote{\Excel appears very different to the
+others but, seen through the lens of this paper, it is very close.}, each by the
+composition of the two design choices~(\S\ref{sec-implementations}). Doing this
+modelling in a single setting allows the differences and similarities between
+these huge systems to be brought out clearly\footnote{All our models are
+executable and are available on Hackage as \cmd{build-1.0}.}. We use the
+abstractions to explain confusing details from the original \Shake
+paper~\cite{mitchell2012shake} and build new features useful for the GHC build
+system~\cite{hadrian}.
 
@@
 with desired properties, for example, to combine the advantages of \Shake and
-\Bazel (\S\ref{sec-cloud-shake}), and evaluate the result (\S\ref{sec-experience}).
+\Bazel, and evaluate the result~(\S\ref{sec-experience}).
 
 \end{itemize}
-In short, instead of seeing build systems as unrelated
-points in space, we now see them as locations in a connected landscape,
-leading to a better understanding of what they do and how they compare,
-and making it easier to explore other points in the
-landscape.
+
+In short, instead of seeing build systems as unrelated points in space, we now
+see them as locations in a connected landscape, leading to a better
+understanding of what they do and how they compare, and making it easier to
+explore other points in the landscape.
+
 We discuss engineering aspects in \S\ref{sec-engineering}, and related
-work in \S\ref{sec-related}.
+work in \S\ref{sec-related}. This paper is an extended version of an earlier
+conference paper~\cite{mokhov2018buildsystems}.
 

### 2019-04-27 papers/jfp/2-background.tex
Polishing S1-S3

@@
 \Make\footnote{There are numerous implementations of \Make and none comes with a
-formal specification. In this paper we therefore use a simple and sensible
-approximation to a real \Make that you might find on your machine.} was developed
+formal specification. In this paper we use a simple approximation to a real
+\Make that you might find on your machine.} was developed
 more than 40 years ago to automatically build software libraries and executable
-programs from source code. It uses \emph{makefiles} to describe tasks (often
-referred to as \emph{build rules}) and their dependencies in a simple textual form.
-For example:
+programs from source code. It uses \emph{makefiles} to describe tasks, often
+referred to as \emph{build rules}, and their dependencies in a simple textual
+form. For example:
 
@@
 \end{minted}
-\vspace{1mm}
+% \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{makefile}
@@
 executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
-the sake of simplicity. In reality, there are multiple versions of \cmd{gcc};
-the actual binary used to compile and link files, and standard libraries
+the sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and
+the actual binary used to compile and link files, as well as standard libraries
 such that \cmd{stdio.h}, are often also listed as dependencies.} the command
@@
 \end{subfigure}
+\hfill
 \begin{subfigure}[b]{0.32\linewidth}
@@
 
+\vspace{2mm}
 \textbf{Definition} (Minimality). A build system is \emph{minimal} if it
@@
 the sum of their values, producing the result in cell \cmd{B1}. If either of
-the inputs change, \Excel will recompute the result.
+the inputs change, \Excel will recompute \cmd{B1}.
 
-Unlike \Make, \Excel does not need to know all task dependencies upfront. Indeed, some
-dependencies may change \emph{dynamically} according to computation results. For
+Unlike \Make, \Excel does not need to know all task dependencies upfront.
+Indeed, some dependencies may change \emph{dynamically} during computation. For
 example:
 
-\vspace{1mm}
+\vspace{0.5mm}
 \begin{minted}[xleftmargin=10pt]{text}
@@
 formula in \cmd{B1} are determined by the value of \cmd{C1}, so it is impossible
-to compute the dependency graph before the build starts\footnote{In
-this particular example one might say that the value of
-\cmd{C1} is statically known, but imagine that it is the result of a long
-computation chain -- its value will only become available during the build.}.
+to compute the dependency graph before the build starts. In this particular
+example one might say that the value of \cmd{C1} is statically known, but
+imagine that it is the result of a long computation chain -- its value will only
+become available during the build.
 
-To support dynamic dependencies, \Excel's calc engine~\cite{excel_recalc} is
-significantly different from \Make. \Excel arranges the cells into a linear
+To support dynamic dependencies, \Excel's calculation engine~\cite{excel_recalc}
+is significantly different from \Make. \Excel arranges the cells into a linear
 sequence, called the \emph{calc chain}.  During the build, \Excel processes
@@
 calc chain respects all the dynamic dependencies of the spreadsheet. When an
-input value (or formula) is changed, \Excel uses the final calc chain from the
+input value or formula is changed, \Excel uses the final calc chain from the
 \emph{previous} build as its starting point so that, in the common case where
@@
 \begin{subfigure}[b]{0.90\linewidth}
+\vspace{1mm}
 \centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
-\caption{Since \cmd{docs.txt} was modified, we rebuild \cmd{release.txt} and
-\cmd{release.tar}, discovering a new dependency.}
+\caption{The input file \cmd{docs.txt} was modified, hence we rebuild
+\cmd{release.txt} and \cmd{release.tar}, discovering a new dependency
+\cmd{README} in the process.}
 \end{subfigure}
@@
 
+\vspace{-1mm}
 \subsection{\Shake: Dynamic Dependencies with No Remorse}
@@
     \item \cmd{LICENSE} is an input text file containing the project license.
-    \item \cmd{release.txt} is a text file listing all files that should be in the release. This file
-    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt},
-    which list all binary and documentation files of the project.
-    \item \cmd{release.tar} is the release archive built by executing the
-    command \cmd{tar} on the release files.
+    \item \cmd{release.txt} lists all release files. This file is produced by
+    concatenating input text files \cmd{bins.txt} and \cmd{docs.txt}, which list
+    binary and documentation files of the project.
+    \item \cmd{release.tar} is the archive built by executing the command
+    \cmd{tar} on the release files.
 \end{itemize}
@@
 \emph{adaptive}~\cite{acar2002adaptive}, and
-\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and
-\S\ref{sec-related}). Second, instead of aborting and deferring the execution of
-tasks whose newly discovered dependencies have not yet been built (as \Excel
-does), \Shake \emph{suspends} their execution until the dependencies are brought
-up to date. We refer to this task scheduling algorithm as \emph{suspending},
-see a concrete implementation in~\S\ref{sec-implementation-shake}.
+\emph{self-adjusting computations} -- see
+Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
+instead of aborting and deferring the execution of tasks whose newly discovered
+dependencies have not yet been built (as \Excel does), \Shake \emph{suspends}
+their execution until the dependencies are brought up to date. We refer to this
+task scheduling algorithm as \emph{suspending}, see a concrete implementation
+in~\S\ref{sec-implementation-shake}.
 
@@
 earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
-early cutoff: \Make and \Excel do not, whereas \Shake and \Bazel (introduced
-below) do.
+early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
+not.
 
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
-% \vspace{-1mm}
+\vspace{-0.5mm}
 \caption{Download sources}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
-% \vspace{-1mm}
+\vspace{-0.5mm}
 \caption{Build \cmd{main.exe}}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
-% \vspace{-1mm}
+\vspace{-0.5mm}
 \caption{Modify \cmd{util.c} and rebuild}
 \end{subfigure}
-% \vspace{-2.5mm}
+\vspace{-1mm}
 \caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
@@
 rectangles.\label{fig-bazel}}
-% \vspace{-3mm}
+\vspace{-1mm}
 \end{figure}
 
-\begin{table}[h]
+\vspace{-1.5mm}
+\subsection{Summary}\label{sec-background-summary}
+\vspace{-0.5mm}
+
+We summarise differences between four discussed build systems in
+Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
+to the information that build systems persistently store between builds:
+\begin{itemize}
+    \item \Make stores file modification times, or rather, it relies on the file
+    system to do that.
+    \item \Excel stores one dirty bit per cell and the calc chain from the
+    previous build.
+    \item \Shake stores the dependency graph discovered in the previous build,
+    annotated with file content hashes for efficient checking of file changes.
+    \item \Bazel stores the content-addressable cache and the history of all
+    previous build commands annotated with file hashes. This information is
+    shared among all users.
+\end{itemize}
+
+In this paper we elucidate which build system properties are consequences of
+specific implementation choices (stored metadata and task scheduling algorithm),
+and how one can obtain new build systems with desired properties by recombining
+parts of existing implementations. As a compelling example,
+in~\S\ref{sec-implementation-cloud} we demonstrate how to combine the advantages
+of \Shake and \Bazel.
+
+\begin{table}
 % \vspace{-2mm}
@@
 \end{table}
-
-% \vspace{-1.5mm}
-\subsection{Summary}\label{sec-background-summary}
-% \vspace{-0.5mm}
-
-We summarise differences between four discussed build systems in
-Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
-to the information that build systems persistently store between builds:
-\begin{itemize}
-    \item \Make stores file modification times, or rather, it relies on the file
-    system to do that.
-    \item \Excel stores one dirty bit per cell and the calc chain from the
-    previous build.
-    \item \Shake stores the dependency graph discovered in the previous build,
-    annotated with file content hashes for efficient checking of file changes.
-    \item \Bazel stores the content-addressable cache and the history of all
-    previous build commands annotated with file hashes. This information is
-    shared among all users of the build system.
-\end{itemize}
-
-In this paper we elucidate which build system properties are consequences of
-specific implementation choices (stored metadata and task scheduling algorithm),
-and how one can obtain new build systems with desired properties by recombining
-parts of existing implementations. As a compelling example,
-in~\S\ref{sec-implementation-cloud} we demonstrate how to combine the advantages
-of \Shake and \Bazel.

### 2019-04-28 papers/jfp/3-abstractions.tex
Polishing S1-S3
Polishing S3-S4

@@
 % \vspace{-0.5mm}
+
+This section presents abstractions that allow us to express
+all the intricacies of build systems discussed in~\S\ref{sec-background},
+and express complex build systems from simple primitives. Specifically, we
+present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
+and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-scheduler},
+\S\ref{sec-rebuilder} and~\S\ref{sec-implementations} scrutinise the
+abstractions further and provide concrete implementations for several build
+systems.
+
+\vspace{-1.5mm}
+\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
+\vspace{-0.5mm}
+
+\emph{Keys, values, and the store.} The goal of any build system is to
+bring up to date a \emph{store} that implements a mapping from \emph{keys} to
+\emph{values}. In software build systems the store is the file system, the
+keys are filenames, and the values are file contents. In \Excel, the store is
+the worksheets, the keys are cell names (such as \cmd{A1}) and the values are
+numbers, strings, etc., displayed as the cell contents. Many build systems use
+\emph{hashes} of values as compact summaries with a fast equality check.
+
+\emph{Input, output, and intermediate values.} Some values must be provided by
+the user as \emph{input}. For example, \cmd{main.c} can be edited by the user
+who relies on the build system to compile it into \cmd{main.o} and subsequently
+\cmd{main.exe}. End build products, such as \cmd{main.exe}, are \emph{output}
+values. All other values (in this case \cmd{main.o}) are \emph{intermediate};
+they are not interesting for the user but are produced in the process of turning
+inputs into outputs.
+
+\emph{Persistent build information.} As well as the key/value mapping, the
+store also contains information maintained by the build system itself, which
+persists from one invocation of the build system to the next -- its ``memory''.
+
+\emph{Task description.} Any build system requires the user to specify how
+to compute the new value for one key, using the (up to date) values of its
+dependencies. We call this specification the \emph{task description}. For
+example, in \Excel, the formulae of the spreadsheet constitute the task
+description; in \Make the rules in the makefile are the task description.
+
+\emph{Build system.} A \emph{build system} takes a task description, a
+\emph{target} key, and a store, and returns a new store in which the target key
+and all its dependencies have up to date values.
+
+We model build systems concretely, as Haskell programs. To that end,
+Fig.~\ref{fig-types} provides the type signatures for all key abstractions
+introduced in the paper. For example, \hs{Store}~\hs{i}~\hs{k}~\hs{v} is the
+type of stores, with several associated functions (\hs{getValue}, etc.). We use
+\hs{k} as a type variable ranging over keys, \hs{v} for values, and \hs{i} for
+the persistent build information. Fig.~\ref{fig-stdlib} lists standard library
+definitions.
 
@@
 
-This section presents abstractions that allow us to express
-all the intricacies of build systems discussed in~\S\ref{sec-background},
-and express complex build systems from simple primitives. Specifically, we
-present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
-and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-scheduler},
-\S\ref{sec-rebuilder} and~\S\ref{sec-implementations} scrutinise the
-abstractions further and provide concrete implementations for several build
-systems.
-
-% \vspace{-1.5mm}
-\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
-% \vspace{-0.5mm}
-
-\emph{Keys, values, and the store.} The goal of any build system is to
-bring up to date a \emph{store} that implements a mapping from \emph{keys} to
-\emph{values}. In software build systems the store is the file system, the
-keys are filenames, and the values are file contents. In \Excel, the store is
-the worksheets, the keys are cell names (such as \cmd{A1}) and the values are
-numbers, strings, etc., displayed as the cell contents. Many build systems use
-\emph{hashes} of values as compact summaries with a fast equality check.
-
-\emph{Input, output, and intermediate values.} Some values must be provided by
-the user as \emph{input}. For example, \cmd{main.c} can be edited by the user
-who relies on the build system to compile it into \cmd{main.o} and subsequently
-\cmd{main.exe}. End build products, such as \cmd{main.exe}, are \emph{output}
-values. All other values (in this case \cmd{main.o}) are \emph{intermediate};
-they are not interesting for the user but are produced in the process of turning
-inputs into outputs.
-
-\emph{Persistent build information.} As well as the key/value mapping, the
-store also contains information maintained by the build system itself, which
-persists from one invocation of the build system to the next -- its ``memory''.
-
-\emph{Task description.} Any build system requires the user to specify how
-to compute the new value for one key, using the (up to date) values of its
-dependencies. We call this specification the \emph{task description}. For
-example, in \Excel, the formulae of the spreadsheet constitute the task
-description; in \Make the rules in the makefile are the task description.
-
-\emph{Build system.} A \emph{build system} takes a task description, a
-\emph{target} key, and a store, and returns a new store in which the target key
-and all its dependencies have an up to date value.
-
-We model build systems concretely, as Haskell programs. To that end,
-Fig.~\ref{fig-types} provides the type signatures for all key abstractions
-introduced in the paper. For example, \hs{Store}~\hs{i}~\hs{k}~\hs{v} is the
-type of stores, with several associated functions (\hs{getValue}, etc.). We use
-\hs{k} as a type variable ranging over keys, \hs{v} for values, and \hs{i} for
-the persistent build information. Fig.~\ref{fig-stdlib} lists standard library
-definitions.
-
 \subsection{The Task Abstraction}\label{sec-task}
@@
 Our first main abstraction is for \emph{task descriptions}:
-\begin{minted}[xleftmargin=10pt]{haskell}
-newtype Task  c k v = Task
-    { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
+\begin{minted}[xleftmargin=5pt]{haskell}
+newtype Task  c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
 type    Tasks c k v = k -> Maybe (Task c k v)
@@
 with the following task description:
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+sprsh1 :: Tasks Applicative String Integer
+sprsh1 "B1" = Just $ Task $ \fetch -> ((+)  <$> fetch "A1"
+                                            <*> fetch "A2")
+sprsh1 "B2" = Just $ Task $ \fetch -> ((*2) <$> fetch "B1")
+sprsh1 _    = Nothing
+\end{minted}
+\vspace{1mm}
+
+\noindent
+We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
+(Real spreadsheet cells would contain a wider range of values, of course.) The
+task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
+but not the input values. It pattern-matches on the key to see if it has a task
+description (in the \Excel case, a formula) for it. If not, it returns
+\hs{Nothing}, indicating that the key is an input. If there is a formula in the
+cell, it returns the \hs{Task} to compute the value of the formula. Every
+\hs{Task} is given a \emph{callback} \hs{fetch} to find the value of any keys on
+which it depends.
 
@@
 \end{figure}
-
-\vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
-sprsh1 :: Tasks Applicative String Integer
-sprsh1 "B1" = Just $ Task $ \fetch ->
-    ((+)  <$> fetch "A1" <*> fetch "A2")
-sprsh1 "B2" = Just $ Task $ \fetch ->
-    ((*2) <$> fetch "B1")
-sprsh1 _    = Nothing
-\end{minted}
-\vspace{1mm}
-
-\noindent
-We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
-(Real spreadsheet cells would contain a wider range of values, of course.) The
-task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
-but not the input values. It pattern-matches on the key to see if it has a task
-description (in the \Excel case, a formula) for it. If not, it returns
-\hs{Nothing}, indicating that the key is an input. If there is a formula in the
-cell, it returns the \hs{Task} to compute the value of the formula. Every
-\hs{Task} is given a \emph{callback} \hs{fetch} to find the value of any keys on
-which it depends.
 
@@
 
-Nevertheless, \hs{busy} can easily handle the example \hs{sprsh1}
-from the previous subsection~\S\ref{sec-task}. In the GHCi session below we
-initialise the store with \cmd{A1} set to 10 and all other cells set to 20.
+Nevertheless, \hs{busy} can easily handle the example task description
+\hs{sprsh1} from the previous subsection~\S\ref{sec-task}. In the GHCi session
+below we initialise the store with \cmd{A1} set to~10 and all other cells set
+to~20.
 
@@
 
-The previous example shows why the \hs{Task} abstraction is polymorphic in
-\hs{f}, recall the definition:
+The previous example illustrates why the \hs{Task} abstraction is polymorphic
+in~\hs{f}; recall its definition from~\S\ref{sec-task}:
 
-\begin{minted}[xleftmargin=10pt]{haskell}
-newtype Task c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
+\begin{minted}[xleftmargin=5pt]{haskell}
+newtype Task c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
 \end{minted}
@@
 section~\S\ref{sec-implementations}. As we shall see, each build system will use
-a different \hs{f}, so the task description must not fix \hs{f}.
+a different \hs{f}, so the task description must not fix~\hs{f}.
 
@@
   \cmd{D}.
-\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
-  even use the application operator \hs{<*>}, which limits dependencies to a
-  linear chain, as e.g. in Docker containers~\cite{docker}. It is
-  interesting to note that, when run on such a task description, the \hs{busy}
-  build system will build each key at most once, thus partially fulfilling the
-  minimality requirement~\ref{def-minimal}. Alas, it still has no mechanism to
-  decide which input keys changed since the previous build.
+\item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
+  description cannot even use the application operator \hs{<*>}, which limits
+  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}. It
+  is interesting to note that, when run on such a task description, the
+  \hs{busy} build system will build each key at most once, thus partially
+  fulfilling the minimality requirement~\ref{def-minimal}. Alas, it still has no
+  mechanism to decide which input keys changed since the previous build.
+\item \hs{Task}~\hs{MonadFail} correspond to monadic tasks that may fail. For
+  example, the formula \cmd{A1/A2} may fail due to division by zero. We will
+  discuss this in~\S\ref{sec-failures}.
 \item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
@@
 \noindent
-Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel
-(which uses dynamic dependencies) is perfectly happy. We can express this
-spreadsheet using our task abstraction as:
+Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, so \Make
+would not be able to order the tasks topologically, but \Excel, which uses
+dynamic dependencies, is perfectly happy. We can express this spreadsheet using
+our task abstraction as follows:
 
@@
 
-\begin{minted}[xleftmargin=5pt]{haskell}
+\begin{minted}[xleftmargin=10pt]{haskell}
 compute :: Task Monad k v -> Store i k v -> v
-compute task store = runIdentity $ run task $
-    \k -> Identity $ getValue k store
+compute task store = runIdentity (run task fetch)
+  where
+    fetch :: k -> Identity v
+    fetch k = Identity (getValue k store)
 \end{minted}
@@
 we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
-The use of \hs{Identity} just fixes the `impedance mismatch' between the
+The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
 function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
@@
 To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
-monad: the function
-\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{\$}~\hs{getValue}~\hs{k}~\hs{store} has the type
-\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
-result has type \hs{Identity}~\hs{v}, which we unwrap with \hs{runIdentity}.
+monad and pass to the \hs{task}. The result has type \hs{Identity}~\hs{v}, which
+we unwrap with \hs{runIdentity}.
 
-% To fix the mismatch, we unwrap using the function \hs{runIdentity} of type
-% \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
-% which both forces the monad to \hs{Identity} and strips it away.
-
-% \vspace{-1mm}
+\vspace{2mm}
 \textbf{Definition}~(Correctness). Suppose \hs{build} is a build system,
@@
     \]
-    In other words, no inputs were corrupted during the
-    build.
+    In other words, no inputs were corrupted during the build.
 
@@
 A build system is \emph{correct} if it produces a correct \hs{result} for any
-given \hs{tasks}, \hs{key} and \hs{store}.
+\hs{tasks}, \hs{key} and \hs{store}.
 
 \label{def-correct}
-% \vspace{2mm}
+\vspace{2mm}
 
@@
 running its code in a different \hs{Applicative}! This is not new, for example
-see~\cite{free-applicatives}, but it is extremely cool. We will see a practical
-use for \hs{dependencies} when implementing applicative build systems,
-see~\S\ref{sec-implementation-make}.
+see~Capriotti~\&~Kaposi~\shortcite{free-applicatives}, but it is extremely cool.
+We will see a practical use for \hs{dependencies} when implementing applicative
+build systems, see~\S\ref{sec-implementation-make}.
 
@@
 on a store with concrete values, which will determine the discovered
-dependencies. Accordingly, we introduce the function \hs{track}: a combination
+dependencies. Accordingly, we introduce the function \hs{track} -- a combination
 of \hs{compute} and \hs{dependencies} that computes both the resulting value and
@@
 import Control.Monad.Writer
-\end{minted}
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
+
 track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [(k, v)])
@@
     trackingFetch :: k -> WriterT [(k, v)] m v
-    trackingFetch k = do v <- lift (fetch k); tell [(k, v)]; return v
+    trackingFetch k = do v <- lift@\,@(fetch k); tell@\,@[(k, v)]; return@\,\blk{v}@
 \end{minted}
@@
 \noindent
-This implementation uses the standard Haskell \hs{WriterT} \emph{monad
-transformer}~\cite{liang1995monad}, which allows us to record additional
-information -- a list of key/value pairs \hs{[(@@k,}~\hs{v)]} -- when computing
+This function uses the standard Haskell \hs{WriterT} \emph{monad
+transformer}~\cite{liang1995monad} for recording additional
+information~--~a list of key/value pairs \hs{[(@@k,}~\hs{v)]}~--~when executing
 a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
@@
 \end{minted}
-\vspace{1mm}
+\vspace{-2mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 spreadsheet in~\S\ref{sec-task-monad}) are determined by the value of \cmd{C1},
-which in this case is obtained by reading from the standard input.
+which in this case is obtained by reading from the standard input using
+\hs{fetchIO}.
 
@@
 dependency :: Task Functor k v -> k
-dependency task = getConst $ run task Const
+dependency task = getConst (run task Const)
 \end{minted}

### 2019-04-28 papers/jfp/4-schedulers.tex
Start work on rebuilders
Polishing S3-S4

@@
-\section{Build System Schedulers}\label{sec-scheduler}
+\section{Schedulers}\label{sec-scheduler}
 
 The focus of this paper is on a variety of implementations of
-\hs{Build}~\hs{c}~\hs{i}~\hs{k}~\hs{v}, given
-a \emph{client-supplied} implementation of \hs{Tasks}~\hs{c}~\hs{k}~\hs{v}. That
-is, we are going to take \hs{Tasks} as given from now on, and explore variants of
-\hs{Build}: first abstractly (in this section) and then concretely
+\hs{Build}~\hs{c}~\hs{i}~\hs{k}~\hs{v}, given a \emph{user-supplied}
+implementation of \hs{Tasks}~\hs{c}~\hs{k}~\hs{v}. That is, we are going to take
+\hs{Tasks} as given from now on, and explore variants of \hs{Build}: first
+abstractly (in this section and in~\S\ref{sec-rebuilder}) and then concretely
 in~\S\ref{sec-implementations}.
@@
 
-We have emboldened two different aspects above: the part of the
-build system responsible for scheduling tasks in the dependency order
-(a `scheduler') can be cleanly separated from the part responsible for deciding
-whether a key needs to be rebuilt (a `rebuilder'). In this section we discuss schedulers,
-leaving rebuilders for \S\ref{sec-rebuilder}.
+We have emboldened two different aspects above: the part of the build system
+responsible for scheduling tasks in the dependency order (a ``scheduler'') can
+be cleanly separated from the part responsible for deciding whether a key needs
+to be rebuilt (a ``rebuilder''). In this section we discuss schedulers, leaving
+rebuilders for \S\ref{sec-rebuilder}.
 
 Section \S\ref{sec-background} introduced three different \emph{task schedulers}
-that decide which tasks to execute and in what order; see the ``Scheduler'' column
-of Table~\ref{tab-summary} in \S\ref{sec-background-summary}.
-This subsection explores the properties of the three schedulers, and
-possible implementations.
+that decide which tasks to execute and in what order; see the ``Scheduler''
+column of Table~\ref{tab-summary} in \S\ref{sec-background-summary}. This
+subsection explores the properties of the three schedulers, and possible
+implementations.
 
@@
 
-The topological scheduler pre-computes a linear order of tasks, which when followed ensures
-dependencies are satisifed, then executes the required tasks in that order. Computing such a linear
-order is straightforward -- given a task description and output \hs{key} first find the (acyclic)
-graph of the \hs{key}'s dependencies, then compute a topological sort. Taking the example from
-Figure~\ref{fig-make}, we might compute the order:
+The topological scheduler pre-computes a linear order of tasks, which when
+followed ensures dependencies are satisfied, then executes the required tasks in
+that order. Computing such a linear order is straightforward -- given a task
+description and output \hs{key} first find the (acyclic) graph of the \hs{key}'s
+dependencies, then compute a topological order. Taking the \Make example from
+Fig.~\ref{fig-make}, we might compute the order:
 
 \begin{enumerate}
-\item \hs{util.o}
-\item \hs{main.o}
-\item \hs{main.exe}
+\item \cmd{util.o}
+\item \cmd{main.o}
+\item \cmd{main.exe}
 \end{enumerate}
 
-Given the dependencies, we could have equally chosen to build \hs{main.o} first, but \hs{main.exe} \emph{must}
-come last.
+\noindent
+Given the dependencies, we could have equally chosen to build \cmd{main.o}
+first, but \cmd{main.exe} \emph{must} come last.
 
-The advantage of this scheme is simplicity -- compute an order, then execute tasks in that order.
-In addition any missing nodes or cycles can be detected from the graph, and reported to the user
-before any work has commenced.
+The advantage of this scheme is simplicity -- compute an order, then execute
+tasks in that order. In addition any missing nodes or cycles can be detected
+from the graph, and reported to the user before any work has commenced.
 
-The downside of this approach is that it requires the dependencies of each task in advance.
-As we saw in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task, which requires the
-build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic dependencies.
+The downside of this approach is that it requires the dependencies of each task
+in advance. As we saw in~\S\ref{sec-deps}, we can only extract dependencies from
+an applicative task, which requires the build system to choose
+\hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic dependencies.
 
@@
 interleave running tasks and ordering tasks. One approach is just to build tasks
-in an arbitrary order, but if a task calls \hs{fetch} on an out-of-date key \hs{dep}
-abort the task and build \hs{dep} instead. Returning to the example from
-Figure~\ref{fig-make}, we might build in the order:
+in an arbitrary order, but if a task calls \hs{fetch} on an out-of-date key
+\cmd{dep}, abort the task and build \cmd{dep} instead. Returning to the example
+from Fig.~\ref{fig-make}, we might build in the order:
 
 \begin{enumerate}
-\item \hs{main.exe} (abort because it requires \hs{util.o})
-\item \hs{util.o}
-\item \hs{main.o}
-\item \hs{main.exe}
+\item \cmd{main.exe} (abort because it requires \cmd{util.o})
+\item \cmd{util.o}
+\item \cmd{main.o}
+\item \cmd{main.exe} (restart from scratch, completing successfully this time)
 \end{enumerate}
 
-Here we start with \hs{main.exe} (chosen randomly), but discover it requires \hs{util.o}, so
-instead start building \hs{util.o}. Next we choose to build \hs{main.o} (again, randomly),
-before finally returning to \hs{main.exe} which now has all its dependencies available and
-completes without error.
+\noindent
+We start with \cmd{main.exe} (chosen randomly), but discover it requires
+\cmd{util.o}, so instead start building \cmd{util.o}. Next we choose to build
+\cmd{main.o} (again, randomly), before finally returning to \cmd{main.exe},
+which now has all its dependencies available and completes without error.
 
-The approach works, but has a number of disadvantages. Firstly, it requires a technical mechanism
-to abort a task -- easy enough in our theoretical setting of \hs{Task} but leads to engineering concerns
-in the real world. Secondly, it is not minimal in the sense that a task may
-start, do some meaningful work, and then abort, repeating that same work when restarted.
+The approach works, but has a number of disadvantages. Firstly, it requires a
+technical mechanism to abort a task -- easy enough in our theoretical setting of
+\hs{Task} (see the implementation in~\S\ref{sec-implementation-excel}) but leads
+to engineering concerns in the real world. Secondly, it is not minimal in the
+sense that a task may start, do some meaningful work, and then abort, repeating
+that same work when restarted.
 
-As a refinement, to reduce the number of aborts (often to zero) \Excel records the
-discovered task order in its \emph{calc chain}, and uses it as the
-starting point for the next build (\S\ref{sec-background-excel}).
-\Bazel's restarting scheduler does not store the discovered order
-between build runs; instead, it stores the most recent task dependency
-information from which it can computed a linear order. Since this information may become outdated, \Bazel may
-also need to abort a task if a newly discovered dependency is out of date.
+As a refinement, to reduce the number of aborts (often to zero) \Excel records
+the discovered task order in its \emph{calc chain}, and uses it as the starting
+point for the next build~(\S\ref{sec-background-excel}). \Bazel's restarting
+scheduler does not store the discovered order between build runs; instead, it
+stores the most recent task dependency information from which it can compute a
+linear order. Since this information may become outdated, \Bazel may also need
+to abort a task if a newly discovered dependency is out of date.
 
@@
 (\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
-they are requested, suspending the currently running task. Using
-Figure~\ref{fig-make}, we would build:
-
-\begin{enumerate}
-\item \hs{main.exe} (suspended)
-\begin{itemize}
-\item \hs{util.o}
-\end{itemize}
-\item \hs{main.exe} (resumsed then suspended)
-\begin{itemize}
-\item \hs{main.o}
-\end{itemize}
-\item \hs{main.exe} (completed)
-\end{enumerate}
-
-As we see, the \hs{main.exe} is started first as it is the required target. It soon discovers
-a dependency on \hs{util.o} and suspends to build \hs{util.o}, resumes a build before building \hs{main.o}
-then continues its work to finish \hs{main.exe}.
-
-This scheduler (when combined with a suitable rebuilder) provides a minimal build system
-which supports dynamic dependencies.
-
-The difficulties are all engineering in nature. To implement suspension there are two
-standard approaches:
+they are requested, suspending the currently running task. Using the example
+from Fig.~\ref{fig-make}, we would build:
 
 \begin{itemize}
-\item Blocking threads or processes. This approach is relatively easy, but can require
-significant resources, especially if a large number of tasks are suspended. In languages with
-cheap green threads (e.g. Haskell) the approach is more feasible, and it was the original approach
-taken by \Shake.
-\item Continuation-passing style \cite{claessen_continuations} can allow the remainder of
-a task to be captured, paused, and resumed later. Continuation passing is efficient, but requires
-the build script to be architected to allow capturing continuations. \Shake currently uses
-continuations.
+\item \cmd{main.exe} (suspended)\\
+  $\hookrightarrow$ \cmd{util.o}
+\item \cmd{main.exe} (resumed then suspended again)\\
+  $\hookrightarrow$   \cmd{main.o}
+\item \cmd{main.exe} (completed)
 \end{itemize}
 
-While a suspending scheduler is theoretically optimal, it is only better in practice
-than a restarting scheduler if the cost of duplicate work avoided outweighs the cost of suspending tasks.
+\noindent
+As we see, the \cmd{main.exe} is started first as it is the required target. It
+soon discovers a dependency on \cmd{util.o} and suspends to build \cmd{util.o},
+resumes a build before building \cmd{main.o} then continues its work to finish
+\cmd{main.exe}.
+
+This scheduler (when combined with a suitable rebuilder) provides a minimal
+build system which supports dynamic dependencies. The difficulties are all
+engineering in nature. To implement suspension there are two standard
+approaches:
+
+\begin{itemize}
+\item Blocking threads or processes. This approach is relatively easy, but can
+require significant resources, especially if a large number of tasks are
+suspended. In languages with cheap green threads (e.g. Haskell) the approach is
+more feasible, and it was the original approach taken by \Shake.
+\item Continuation-passing style \cite{claessen_continuations} can allow the
+remainder of a task to be captured, paused, and resumed later. Continuation
+passing is efficient, but requires the build script to be architected to allow
+capturing continuations. \Shake currently uses continuations.
+\end{itemize}
+
+\noindent
+While a suspending scheduler is theoretically optimal, it is only better in
+practice than a restarting scheduler if the cost of avoided duplicate work
+outweighs the cost of suspending tasks. Note furthermore that the cost of
+duplicate work may often be just a fraction of the overall build cost.
+
+\q{Andrey}{It's unclear how our \hs{busy} build system fits into the two above
+approaches. It probably is a third approach (i.e. using stack), but it's
+unsuitable for real build systems that call external processes. Shall we mention
+this?}
+
+\q{Andrey}{A follow-up to the above: our \Shake models use neither of the two
+above approaches, which seems a bit unsatisfactory. Can we at least hint how
+our models could be adapted to use the continuation-passing style? Can we write
+\hs{busyCPS}?}

### 2019-04-28 papers/jfp/5-rebuilders.tex
Polishing S5

@@
 
-A build system can be split into a scheduler (as defined in \S\ref{sec-scheduler}) and a \emph{rebuilder}. Suppose the scheduler decides that a key should be brought up to date. The next
-question is: does any work need to be done, or is the key already up to date?
-Or, in a cloud-build system, do we have a cached copy of the value we need?
+A build system can be split into a scheduler (as defined
+in~\S\ref{sec-scheduler}) and a \emph{rebuilder}. Suppose the scheduler decides
+that a key should be brought up to date. The next question is: does any work
+need to be done, or is the key already up to date? Or, in a cloud-build system,
+do we have a cached copy of the value we need?
 
-While \S\ref{sec-background} explicitly listed the schedulers, the rebuilders were introduced more implicitly, primarily by the information they retain to make their decisions.
-From the examples we have looked at we see four fundamental rebuilds, each with a number of tweaks and variations within them.
+While~\S\ref{sec-background} explicitly listed the schedulers, the rebuilders
+were introduced more implicitly, primarily by the information they retain to
+make their decisions. From the examples we have looked at we see four
+fundamental rebuilders, each with a number of tweaks and variations within them.
 
@@
 clean, the key does not need to be rebuilt. Taking the example from
-Figure~\ref{fig-make}, if \hs{util.c} changes then that would be marked dirty, and \hs{util.o} and \hs{main.exe} would rebuild as they transitively depend on \hs{util.c}.
+Fig.~\ref{fig-make}(c), if \cmd{main.c} changes then it would be marked dirty,
+and \cmd{main.o} and \cmd{main.exe} would be rebuilt as they transitively depend
+on \cmd{main.c}.
 
@@
 than necessary, but not vice versa. The over-approximation is as follows: a cell
-is marked dirty if its formula statically refers to a dirty cell, or if the
-formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
-guessed from the formula alone. The over-approximation is clear for
-\cmd{INDIRECT}, but it is also present for \cmd{IF}, where both branches are
-followed even though dynamically only one is used.
+is marked dirty (i) if its formula statically refers to a dirty cell, or (ii) if
+the formula calls a \emph{volatile} function like \cmd{INDIRECT} whose
+dependencies cannot be guessed from the formula alone. The over-approximation is
+clear for \cmd{INDIRECT}, but it is also present for \cmd{IF}, where both
+branches are followed even though dynamically only one is used.
 
@@
 
-With a dirty bit it is possible to achieve minimality. However, to achieve early
-cutoff (\S\ref{sec-background-shake}) it would be important to reset the dirty
+With a dirty bit it is possible to achieve
+minimality~(\S\ref{sec-background-make}). However, to achieve early
+cutoff~(\S\ref{sec-background-shake}) it would be important to clear the dirty
 bit after a computation that did not change the value and make sure that cells
@@
 
-The use of a single bit of information is useful to reduce memory consumption, and
-in the case of \Make, to integrate with the file system. However, as the examples show,
-in constrained environments where a dirty bit is chosen, it's often done as part of
-a series of compromises.
+A dirty bit rebuilder is useful to reduce memory consumption, and in the case
+of \Make, to integrate with the file system. However, as the examples show, in
+constrained environments where a dirty bit is chosen, it is often done as part
+of a series of compromises.
+
+\q{Andrey}{Can we say something like: ``The authors are not aware of a build
+system that uses a dirty bit rebuilder, is minimal, and supports the early
+cutoff optimisaion''? Perhaps, we can even make a claim, that such a build
+system is impossible with a believable argument/proof sketch?}
 
@@
 values/hashes of dependencies used last time, and if something has changed, the
-key is dirty and must be rebuilt~--~in essence a \emph{trace} which we can use
-to \emph{verify} existing values. Taking the example from Figure \ref{fig-make},
-we might record that \hs{util.o} dependend on \hs{util.c} (at hash 1) and \hs{util.h}
-(at hash 2) and produced \hs{util.o} (at hash 4). Next time round, if all those files
-still have the same hashes, there is nothing to do, and we can skip running the build.
-If they are different, we repeat the action, and record the new values.
+key is dirty and must therefore be rebuilt~--~in essence, keeping a \emph{trace}
+which we can use to \emph{verify} existing values. Taking the example from
+Fig.~\ref{fig-bazel}(c), we might record that the key \cmd{util.o} (at hash 8)
+depended on the keys \cmd{util.c} (at hash 7) and \cmd{util.h} (at hash 2). Next
+time round, if the scheduler decides that it is time for \cmd{util.o} to be
+rebuilt and all keys still have the same hashes as in the recorded trace, there
+is nothing to do, and we can skip rebuilding. If any of the hashes is different,
+we rebuild \cmd{util.o}, and record a trace with the new values.
 
-For traces, there are two essential
-operations~--~adding a new value to the trace store, and using the traces to
-determine if a key needs rebuilding. Assuming a store of verifying traces
-\hs{VT}~\hs{k}~\hs{v}, the operations are:
+For traces, there are two essential operations~--~adding a new trace to the
+trace store, and using the traces to determine if a key needs rebuilding.
+Assuming a store of verifying traces \hs{VT}~\hs{k}~\hs{v}, the operations are:
 
-\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 recordVT :: k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
+\end{minted}
+\vspace{-2mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 verifyVT :: (Monad m, Eq k, Eq v)
@@
 
-More interestingly, to \emph{verify} whether a key needs rebuilding we supply
-the key, the hash of its current value, a function for obtaining the post-build
-value of any key (using a scheduling strategy as per
-\S\ref{sec-dependency-orderings}), and the existing \hs{VT} information. The
-result will be a \hs{Bool} where \hs{True} indicates that the current value is
-already up to date, and \hs{False} indicates that it should be rebuilt.
+More interestingly, to \emph{verify} whether a key needs rebuilding we use
+\hs{verifyVT}, supplying the key, the hash of its current value, a function for
+obtaining the post-build value of any key (using a scheduling strategy as
+per~\S\ref{sec-scheduler}), and the existing trace store \hs{VT}. The result
+will be a \hs{Bool} where \hs{True} indicates that the current value is already
+up to date, and \hs{False} indicates that it should be rebuilt.
+
+\q{Andrey}{The "function for obtaining the post-build value of any key" is a bit
+mysterious. It would be nice to add a clarification. Or could we simplify the
+interface and only provide a function with type
+\hs{k -> Hash v -> [(k, Hash v)] -> VT k v -> Bool}? Presumably, \hs{verifyVT}
+with the current type signature could be built on top of it?}
 
@@
 
-\subsubsection{Representation Optimisations}
+\subsubsection{Trace Representation}
 
@@
 \hs{recordVT} in a list, and verify by simply checking if any list item matches
-the information passed by \hs{verifyVT}. Concretely,
-in our implementations from \S\ref{sec-implementations}, traces are recorded as lists of:
+the information passed by \hs{verifyVT}. Concretely, in our implementations
+from~\S\ref{sec-implementations}, traces are recorded as lists of:
 
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Trace k v = Trace
-    {key :: k
-    ,depends :: [(k, Hash v)]
-    ,result :: Hash v
-    }
+data Trace k v a = Trace { key     :: k
+                         , depends :: [(k, Hash v)]
+                         , result  :: a }
 \end{minted}
+\vspace{1mm}
 
-\noindent Where \hs{r} is \hs{Hash}~\hs{v} for verifying traces (and \hs{v} for constructive traces, discussed later in \S\ref{sec-constructive-traces}).
-A real system is highly likely to use a more optimised implementation.
+\noindent
+Where \hs{a} is \hs{Hash}~\hs{v} for verifying traces (and \hs{v} for
+constructive traces, discussed later in~\S\ref{sec-constructive-traces}). A real
+system is highly likely to use a more optimised implementation.
 
-The first optimisation is that any system using \hs{Applicative} dependencies can omit the dependency
-keys from the \hs{Trace} since they can be recovered from the \hs{key} field.
+The first optimisation is that any system using \hs{Applicative} dependencies
+can omit the dependency keys from the \hs{Trace} since they can be recovered
+from the \hs{key} field~(\S\ref{sec-deps}).
 
-The next optimisation is that there is only very minor benefit from storing more than one \hs{Trace} per key.
-Therefore, veryifying traces can be stored as \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
-initial \hs{k} is the \hs{key} field of \hs{Trace} -- making \hs{verifyVT} fast.
-Note that storing only one \hs{Trace} per key means that if the dependencies change but the result does not, and then the
-dependencies change back to what they were before, there will not be a valid \hs{Trace} available and the system will have to rebuild, where a complete list of all historical traces would not have had to.
-However, it also means that the number of \hs{Trace} structures is bounded by the number of distinct keys, regardless of how many builds are executed -- a useful property.
+The next optimisation is that there is only very minor benefit from storing more
+than one \hs{Trace} per key. Therefore, verifying traces can be stored as
+\hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
+initial \hs{k} is the \hs{key} field of \hs{Trace}, thus making \hs{verifyVT}
+fast. Note that storing only one \hs{Trace} per key means that if the
+dependencies change but the result does not, and then the dependencies change
+back to what they were before, there will be no valid \hs{Trace} available and
+the build system will have to rebuild, whereas a complete list of all historical
+traces would not have had to. However, it also means that the number of
+\hs{Trace} structures is bounded by the number of distinct keys, regardless of
+how many builds are executed -- a useful property.
 

### 2019-04-28 papers/jfp/4-schedulers.tex
Answer questions in S4

@@
 This scheduler (when combined with a suitable rebuilder) provides a minimal
-build system which supports dynamic dependencies. The difficulties are all
-engineering in nature. To implement suspension there are two standard
-approaches:
+build system which supports dynamic dependencies. In our model, a suspending
+scheduler easy to write -- it makes a function call to compute each dependency. However,
+a more practical implementation is likely to build multiple dependencies in parallel,
+which then requires a more explicit suspension and resumption. To implement
+suspension there are two standard approaches:
 
@@
 duplicate work may often be just a fraction of the overall build cost.
-
-\q{Andrey}{It's unclear how our \hs{busy} build system fits into the two above
-approaches. It probably is a third approach (i.e. using stack), but it's
-unsuitable for real build systems that call external processes. Shall we mention
-this?}
-
-\q{Andrey}{A follow-up to the above: our \Shake models use neither of the two
-above approaches, which seems a bit unsatisfactory. Can we at least hint how
-our models could be adapted to use the continuation-passing style? Can we write
-\hs{busyCPS}?}

### 2019-04-28 papers/jfp/5-rebuilders.tex
Describe how to build a dirty bit builder

@@
 
-\q{Andrey}{Can we say something like: ``The authors are not aware of a build
-system that uses a dirty bit rebuilder, is minimal, and supports the early
-cutoff optimisaion''? Perhaps, we can even make a claim, that such a build
-system is impossible with a believable argument/proof sketch?}
+It is possible to write a dirty-bit builder which is minimal and supports early
+cutoff. To do so, you need to start with all inputs that have changed marked dirty,
+then a node must be rebuilt if any children are dirty, and must set that node to dirty
+only if the result changes. At the end of the build all dirty bits
+must be reset to not-dirty. This approach only works if all targets are rebuilt each time,
+since clearing dirty bits of children who aren't targets will cause them to incorrectly
+not rebuild subsequently. To avoid resetting the dirty bit,
+it is possible to use successive execution numbers, which ultimately leads to an approach
+we call verifying step traces in \S\ref{sec-step-traces}.
 

### 2019-04-28 papers/jfp/5-rebuilders.tex
Explain post-build keys

@@
 
-\q{Andrey}{The "function for obtaining the post-build value of any key" is a bit
-mysterious. It would be nice to add a clarification. Or could we simplify the
-interface and only provide a function with type
-\hs{k -> Hash v -> [(k, Hash v)] -> VT k v -> Bool}? Presumably, \hs{verifyVT}
-with the current type signature could be built on top of it?}
+The most complex argument of \hs{verifyVT} is a function to obtain the post-build
+value of any key, which we refer to as \hs{fetch}. With an \hs{Applicative} task,
+\hs{fetch} will be called on the dependencies of the \hs{k} argument. However, with a
+\hs{Monad} task, the dependencies are not known from the task alone, they are only
+recorded from previous executions and stored in \hs{VT}. If the build system has
+two traces for a given \hs{k}, they will both request the same dependency first,
+since \hs{Task} is determinstic. However, depending on that first result,
+they may then request different successive dependencies from \hs{fetch}. A curious
+result is that for suspending schedulers (\S\ref{sec-suspending}) in many cases
+the actual build steps are performed as a consequence of checking if a key needs rebuilding!
 

### 2019-04-28 papers/jfp/1-intro.tex
Proof reading and tweaking

@@
 these huge systems to be brought out clearly\footnote{All our models are
-executable and are available on Hackage as \cmd{build-1.0}.}. We use the
-abstractions to explain confusing details from the original \Shake
-paper~\cite{mitchell2012shake} and build new features useful for the GHC build
-system~\cite{hadrian}.
+executable and are available on Hackage as \cmd{build-1.0}.}.
 
@@
 \Bazel, and evaluate the result~(\S\ref{sec-experience}).
+
+\item We can use these new
+abstractions to more clearly explain details from the original \Shake
+paper~\cite{mitchell2012shake} (\S\ref{sec-step-traces}, \S\ref{sec-experience-shake})
+and build new features useful for the GHC build system~\cite{hadrian} (\S\ref{sec-using-cloud-shake}).
 

### 2019-04-28 papers/jfp/6-implementations.tex
Proofing and tweaking
Proof reading and tweaking

@@
 type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
-type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
+type Rebuilder c   ir k v =
+    k -> v -> Task c k v -> Task (MonadState ir) k v
 \end{minted}
@@
 
+As discussed in \S\ref{sec-step-traces}, \Shake actually use verifying step traces, but we
+choose to focus on the more explicit verifying traces. We have implemented
+verifying step traces in our framework, and they compose with schedulers as
+you would hope.
+
 \subsection{Cloud Build Systems: \Bazel, \CloudBuild, \Cloud \Shake, \Buck and \Nix}
@@
 rules in the language available to users. Instead, users have to rely on a
-(rich) collection of predefined built-in rules, which cover many
+collection of predefined built-in rules, which cover many
 important instances of dynamic dependencies.

### 2019-04-28 papers/jfp/7-experience.tex
Proof reading and tweaking

@@
 
-\subsection{Experience from \Shake}
+\subsection{Experience from \Shake}\label{sec-experience-shake}
 
@@
 The biggest change in that time period was moving from blocking threads to continuations for the suspending scheduler.
-The most visible change is switching \hs{*>} to \hs{%>} for defining rules\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.
+The most visible change is switching from \hs{*>} to \hs{%>} for defining rules\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.
 
 We consider the lack of change suggestive that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}.
-There are two main aspects to the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, Firstly, the rebuilder can now be described as verifying step traces as per \S\ref{sec-step-traces}, with a much clearer relationship to the unoptimised verifying traces in \S\ref{sec-verifying-traces}.
+There are two main aspects to the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, Firstly, the rebuilder can now be described as verifying step traces as per \S\ref{sec-step-traces}, with a much clearer relationship to the unoptimised verifying traces of \S\ref{sec-verifying-traces}.
 The second is the power of \Shake tasks (called actions in the original paper).
@@
 
-In some sense, this is the ``free'' version of \hs{Task Monad} \cite{free-applicatives} -- the two types form an isomorphism.
+This type is equivalent to \hs{Task Monad} -- the two types form an isomorphism.
 The original paper uses concrete types \hs{Key} and \hs{Value}, we generalise these types to \hs{k} and \hs{v}, and also add \hs{r} so that \hs{Action k v} can be an instance of \hs{Monad}.
-We define that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type:
+We define that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type, witnessing the isomorhpism:
 
@@
 These functions are suspiciously like \hs{recordCT} and \hs{constructCT} from \S\ref{sec-constructive-traces}, with their differences perhaps the most illustrative of the changes required.
-(We have made some replacements from full \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant differences.)
+(We have made some minor changes from actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant differences.)
 
 \begin{itemize}
-\item There are two \hs{Ver} arguments being passed to each operation.
+\item There are two \hs{Ver} arguments being passed to each function.
 These are the versions of the build script, and the rule for this particular key.
@@
 The reason is that \Shake allows a list of dependencies to be supplied at once, so they can all be built in parallel.
-\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the files that this rule produces -- which must be declared with \hs{produces}, or the output keys from a rule.
-\item The \hs{lookupCloud} allows an explicit \hs{Nothing} when looking up a dependent key, since some keys are not buildable.
-\item The \hs{lookupCloud} returns at most one result, rather than a list. This change was made for simplicity.
+\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the files that this rule produces.
+These produced files include those which are output keys from a rule and those declared with \hs{produces}.
+\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking up a dependent key, since some keys are not buildable.
+\item The \hs{lookupCloud} function returns at most one result, rather than a list. This change was made for simplicity.
 \end{itemize}
 
-The integrate these functions into \Shake we found the most expedient design was to leave \Shake with a verifying trace, but if the verifying trace doesn't skip building, we consult the constructive trace.
+To integrate these functions into \Shake we found the most expedient route was to leave \Shake with verifying traces, but if the verifying trace doesn't match, we consult the constructive trace.
 By bolting constructive traces onto the side of Shake we avoid reengineering of the central database.
-We haven't found any significant downsides from the bolt-on approach thus far, so it may be a sensible route to go even if developing from scratch -- allowing an optimised verified trace implementation in many cases, and falling back to a more complex implementation only rarely.
+We haven't found any significant downsides from the bolt-on approach thus far, so it may be a sensible route to go even if developing from scratch -- allowing an optimised verified trace implementation in many cases, and falling back to a more complex implementation (requiring consulting remote servers) only rarely.
 
@@
 Unfortunately, on certain OS's (e.g.
-Windows) mounting an HTTP endpoint as a file system requires administrator privileges, so an HTTP cache is still required.
+Windows) mounting an HTTP endpoint as a file system requires administrator privileges, so an HTTP cache is still desirable.
 
-\subsection{Experience from using \Cloud \Shake}
+\subsection{Experience from using \Cloud \Shake}\label{sec-using-cloud-shake}
 

### 2019-04-28 papers/jfp/8-engineering.tex
Polishing S3-S4
More work tidying up S8

@@
 
-\subsection{Partial Stores and Exceptions}
+\subsection{Partial Stores and Exceptions}\label{sec-failures}
 
@@
 because they are out of date but because of races with earlier tasks that had
-not yet finished. As a consequence, over successive runs, potentially racey
-dependencies will be separated, giving better parallelism over time.
+not yet finished. As a consequence, if a the calcuation order is persisted over
+successive runs (as \Excel does), potentially racey dependencies will be separated,
+giving better parallelism over time.
 
-\item The \hs{suspending} scheduler can be made parallel by starting static
-dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
-resolved, using the strategy described by~\cite{marlow2014haxl}.
+\item The \hs{suspending} scheduler can be made parallel by starting multiple
+dependencies simulataneously. One approach is to make the request for dependencies
+take a list of keys, as implemented by \Shake. Another approach is to treat the
+\hs{Applicative} dependencies of a \hs{Task Monad} in parallel, using the strategy
+described by~\cite{marlow2014haxl}.
 \end{enumerate}
+
+Once sufficient parallelism is available the next challenge is preventing excess
+parallelism and machine resource starvation, which is usually acheived with a thread pool or thread limit.
 
@@
 but neither is it beautiful or informative.
-
-\todo{NM}{Should we add a paragraph or two giving a bit more details about how
-parallelism is supported in \Shake?}
 
@@
 
-\item[Non-determinism] Some tasks are \emph{non-deterministic}, producing a
-result from a possible set. As an example, GHC when compiled using parallelism
+\item[Non-determinism] Some tasks are \emph{non-deterministic}, producing any
+result from a possible set. As an example, GHC programs compiled using parallelism
 can change the order in which unique variables are obtained from the supply,
@@
 
-Overall we choose not to deal with cyclic dependencies, a choice that most build
+Overall we choose not to deal with cyclic dependencies -- a choice that most build
 systems also follow. There are computation frameworks that support tasks with
@@
 
-\subsection{Polymorphism and Multiple-Output Build Tasks}\label{sec-polymorphism}
+\subsection{Polymorphic values}\label{sec-polymorphism}
 
@@
 Generalised algebraic data types (GADTs)~\cite{spj2006gadts} can be used to
-extend our models to support ``typed'' build tasks. The key idea is to replace
-the monomorphic callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
-polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
+extend our models to support ``typed'' build tasks. The idea is to replace
+the callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
+more polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
 where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
@@
 
-A more compelling example is build tasks that produce multiple output
+\subsection{Multiple-Output Build Tasks}\label{sec-multiple-outputs}
+
+We can use the extra polymorphism from \S\ref{sec-polymorphism} for build tasks that produce multiple output
 keys~--~for example, \cmd{ghc A.hs} produces both \cmd{A.hi} and \cmd{A.o}.
@@
 whose result is a pair of file contents. From that, the task for \cmd{A.hi}
-can be the first component of the result of the pair. Again, such an operation
+can be the first component of the result of the pair. Such an operation
 can be encoded without polymorphic keys provided the pair of files (or a dummy
 file representing the pair) is marked as changed if either of the contained
-files change. Once again, polymorphic dependencies provide convenience rather
-than power.
+files change, but it is more elegant and type-safe with the extra polymorphism.
 
-Note that multiple-output build tasks require one more convenience mechanism,
+Multiple-output build tasks require one more convenience mechanism,
 which is orthogonal to polymorphism. Multiple-output task descriptions partition

### 2019-04-28 papers/jfp/10-conclusions.tex
More tidying up of S8

--- 2019-03-05 papers/jfp/8-conclusions.tex
@@
 By first decomposing the pieces, we show how to recompose the pieces to find new points in the design space.
-In particular, a simple recombination leads to a design for a monadic suspending cloud build system.
-Armed with that blueprint we hope to actually implement such a system as future work.
+In particular, a simple recombination leads to a design for a monadic suspending cloud build system, which we have implemented and use in our day-to-day development.

### 2019-04-28 papers/jfp/8-engineering.tex
More tidying up of S8

@@
 It is possible to implement \hs{selfTracking} without relying on GADTs and typed
-tasks presented in~\S\ref{sec-polymorphism}, but the cost of using partial
+tasks presented in~\S\ref{sec-polymorphism}, at the cost of using partial
 functions.

### 2019-04-28 papers/jfp/9-related.tex
Work on S3
More tidying up of S8

@@
 \section{Related work}\label{sec-related}
-
-\todo{Relationship to SHake paper.   Fully subsumes it, and explains it much better.  Oracles no longer imoprtant.
-}
 
@@
 
-\item \Redo~\cite{redo-idea}\cite{grosskurth2007redo}\cite{redo} almost exactly
+\item \Redo~\cite{redo-idea,grosskurth2007redo,redo} almost exactly
 matches \Shake at the level of detail given here, differing only on aspects like

### 2019-04-28 papers/jfp/8-engineering.tex
Add notes on some of the TODO's

@@
 
-\todo{AM}{Show implementation of \hs{computeND}}.
+\todo{AM}{Show implementation of \hs{computeND}. Neil says: not convinced it's that interesting.}.
 
@@
 
-\todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.}
+\todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.
+Neil says: I think this is a stretch.
+It's a lot more nuanced as we saw from Bazel discussions so I'd probably skip it.}
 
@@
 bit inconsistent. We could switch from \hs{[k]} to \hs{Set k} fairly easily.
-Shall we?}
+Shall we? Neil says: Haskell code dealing with sets is quite a lot more complex. People are used
+to being quite fuzzy around the difference - I'd leave it as is.}
 

### 2019-04-28 papers/jfp/3-abstractions.tex
Make sure S3 doesn't overflow the margins

@@
 \begin{minted}[fontsize=\small]{haskell}
--- An abstract store containing a key/value map and persistent build information
+-- Abstract store containing a key/value map and persistent build information
 data Store i k v -- i = info, k = key, v = value
@@
   | m < 0 || n < 0 = Nothing
-  | m == 0    = Just $ Task $ const $ pure (n + 1)
-  | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
-  | otherwise = Just $ Task $ \fetch -> do index <- fetch (m, n - 1)
-                                           fetch (m - 1, index)
+  | m == 0    = Just $ Task $ const $ pure (n+1)
+  | n == 0    = Just $ Task $ \fetch -> fetch (m-1, 1)
+  | otherwise = Just $ Task $ \fetch -> do index <- fetch (m, n-1)
+                                           fetch (m-1, index)
 \end{minted}
@@
 
-cloudAckermann :: Tasks (MonadState Cache) (Integer, Integer) Integer
-cloudAckermann (m, n)
+cloudAck :: Tasks (MonadState Cache) (Integer, Integer) Integer
+cloudAck (m, n)
     | m < 0 || n < 0 = Nothing
-    | m == 0    = Just $ Task $ const $ pure (n + 1)
-    | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
+    | m == 0    = Just $ Task $ const $ pure (n+1)
+    | n == 0    = Just $ Task $ \fetch -> fetch (m-1, 1)
     | otherwise = Just $ Task $ \fetch -> do
@@
         case Map.lookup (m, n) cache of
-            Nothing -> do index <- fetch (m, n - 1)
-                          value <- fetch (m - 1, index)
+            Nothing -> do index <- fetch (m, n-1)
+                          value <- fetch (m-1, index)
                           modify (Map.insert (m, n) value)

### 2019-04-28 papers/jfp/4-schedulers.tex
Polish S4-S5

@@
 \noindent
-As we see, the \cmd{main.exe} is started first as it is the required target. It
-soon discovers a dependency on \cmd{util.o} and suspends to build \cmd{util.o},
+As we see, \cmd{main.exe} is started first as it is the required target. It soon
+discovers a dependency on \cmd{util.o} and suspends to build \cmd{util.o},
 resumes a build before building \cmd{main.o} then continues its work to finish
@@
 build system which supports dynamic dependencies. In our model, a suspending
-scheduler easy to write -- it makes a function call to compute each dependency. However,
-a more practical implementation is likely to build multiple dependencies in parallel,
-which then requires a more explicit suspension and resumption. To implement
-suspension there are two standard approaches:
+scheduler is easy to write -- it makes a function call to compute each
+dependency. However, a more practical implementation is likely to build multiple
+dependencies in parallel, which then requires a more explicit task suspension
+and resumption. To implement suspension there are two standard approaches:
 
@@
 more feasible, and it was the original approach taken by \Shake.
-\item Continuation-passing style \cite{claessen_continuations} can allow the
+\item Continuation-passing style~\cite{claessen_continuations} can allow the
 remainder of a task to be captured, paused, and resumed later. Continuation

### 2019-04-28 papers/jfp/5-rebuilders.tex
Polish S4-S5

--- 2019-04-28 papers/jfp/5-rebuilders.tex
@@
 that a key should be brought up to date. The next question is: does any work
-need to be done, or is the key already up to date? Or, in a cloud-build system,
+need to be done, or is the key already up to date? Or, in a cloud build system,
 do we have a cached copy of the value we need?
@@
 \hs{verifyVT}, supplying the key, the hash of its current value, a function for
-obtaining the post-build value of any key (using a scheduling strategy as
-per~\S\ref{sec-scheduler}), and the existing trace store \hs{VT}. The result
-will be a \hs{Bool} where \hs{True} indicates that the current value is already
-up to date, and \hs{False} indicates that it should be rebuilt.
+obtaining the hash of the post-build value of any key (using a scheduling
+strategy as per~\S\ref{sec-scheduler}), and the existing trace store \hs{VT}.
+The result will be a \hs{Bool} where \hs{True} indicates that the current value
+is already up to date, and \hs{False} indicates that it should be rebuilt.
 
-\q{Andrey}{The "function for obtaining the post-build value of any key" is a bit
-mysterious. It would be nice to add a clarification. Or could we simplify the
-interface and only provide a function with type
-\hs{k -> Hash v -> [(k, Hash v)] -> VT k v -> Bool}? Presumably, \hs{verifyVT}
-with the current type signature could be built on top of it?}
+The most complex argument of \hs{verifyVT} is a function
+\hs{fetchHash}~\hs{::}~\hs{k}~\hs{->}~\hs{m}~\hs{(Hash}~\hs{v)} to obtain the
+hash of the post-build value of any key. With an applicative task,
+\hs{fetchHash} will be called on the statically known task dependencies.
+However, with a monadic task, the dependencies are not known from the task
+alone, they are only recorded from previous executions and stored in \hs{VT}. If
+the build system has two traces for a given key~\hs{k}, they will both request
+the same dependency first, since \hs{Task}~\hs{Monad} is deterministic. However,
+depending on that first result, they may then request different successive
+dependencies using \hs{fetchHash}. A curious result is that for suspending
+schedulers~(\S\ref{sec-suspending}) in many cases the actual build steps are
+performed as a consequence of checking if a key needs rebuilding!
 
@@
 dynamic dependencies and minimality; furthermore, all traces except for deep
-traces~(\S\ref{sec-deep-constructive-traces}) support early cutoff.
+traces~(\S\ref{sec-deep-constructive-traces}) support the early cutoff
+optimisation~(\S\ref{sec-background-shake}).
 
@@
 initial \hs{k} is the \hs{key} field of \hs{Trace}, thus making \hs{verifyVT}
-fast. Note that storing only one \hs{Trace} per key means that if the
-dependencies change but the result does not, and then the dependencies change
-back to what they were before, there will be no valid \hs{Trace} available and
-the build system will have to rebuild, whereas a complete list of all historical
-traces would not have had to. However, it also means that the number of
-\hs{Trace} structures is bounded by the number of distinct keys, regardless of
-how many builds are executed -- a useful property.
+much faster. Note that storing only one \hs{Trace} per key means that if the
+dependencies of a key change but the resulting value does not, and then the
+dependencies change back to what they were before, there will be no valid
+\hs{Trace} available and the key will therefore have to be rebuilt, whereas a
+complete list of all historical traces would allow the rebuilding to be skipped.
+On the other hand, bounding the number of \hs{Trace} structures by the number of
+distinct keys, regardless of how many builds are executed, is a useful property.
 
@@
 
-The \Shake build system, and the associated paper \cite{mitchell2012shake} in \S2.3.3 describes a slightly different trace structure, one that is similar to verifying traces, but stores slightly less data, and has slightly different early cutoff semantics.
-Rather than storing the \hs{Hash v} for each dependency, it instead stores a built time and changed time for each key, and for the dependencies just stores a list of \hs{k}.
-The resulting \hs{Trace} type ressembles:
+The \Shake build system and the associated paper -- see~\S2.3.3
+in~Mitchell~\shortcite{mitchell2012shake} -- use a slightly different trace
+structure, called \emph{step traces}, one that is similar to verifying traces,
+but stores less data, and has different early cutoff semantics. Rather than
+storing the \hs{Hash}~\hs{v} for each dependency, it instead stores \hs{built}
+time and \hs{changed} time for the \hs{key}, and a list of dependency keys
+(without the hashes). The resulting \hs{StepTrace} type resembles:
 
-\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
-data Trace k v = Trace
-    {key :: k
-    ,result :: Hash v
-    ,depends :: [k]
-    ,built :: Time
-    ,changed :: Time
-    }
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+data StepTrace k v = StepTrace { key     :: k
+                               , result  :: Hash v
+                               , built   :: Time
+                               , changed :: Time
+                               , depends :: [k] }
 \end{minted}
+\vspace{1mm}
 
-The \hs{built} field is when the \hs{key} last rebuilt.
-The \hs{changed} field is when the \hs{result} last changed -- if the last build changed the value, it will be equal to \hs{built}, otherwise it will be older.
-The \hs{recordVT} function needs to consult the previous \hs{Trace} to know whether to use \hs{built} or the previous \hs{changed} value as the new \hs{changed} value.
-The \hs{verifyVT} needs to verify that the \hs{built} time of \hs{key} is newer than the \hs{changed} values of all its \hs{depends}.
-This scheme preserves minimality and early cutoff.
-A variant with only one \hs{Time} field would loose early cutoff, and indeed corresponds quite closely to \Make.
-Furthermore, the \hs{Time} stamp only needs record which execution of the build is running, so every key built in the same run can share the same \hs{Time} value -- it just needs to be monotically increasing between runs.
+\noindent
+The \hs{built} field is when the \hs{key} last rebuilt. The \hs{changed} field
+is when the \hs{result} last changed -- if the last build changed the value, it
+will be equal to \hs{built}, otherwise it will be older. The function
+\hs{recordVT} consults the previous step traces to know whether to keep the
+previous \hs{changed} value or change it to \hs{built}. The function
+\hs{verifyVT} is a bit more subtle; given a key \hs{k} and the hash of its
+current value \hs{h}, it performs the following steps:
 
-This optimisation is useful, at least in the case of \Shake, to save space.
-A hash takes up 32 bytes in typical programs, while a key (in \Shake only) is an \hs{Int} taking only 4 bytes.
-Furthermore, \Shake permits values to be arbitrarily large, and permits a custom equality (two values can be bit-for-bit unequal but considered equal by \Shake), meaning that a \hs{Hash} isn't a valid encoding.
-For build systems that used \hs{Applicative} dependencies, the \hs{depends} could be omitted entirely, making the size of a \hs{Trace} $O(1)$ instead of $O(n)$ where $n$ is the number of dependencies.
+\begin{itemize}
+\item Find the latest (with respect to the field \hs{built}) step trace matching
+  \hs{k}. If it does not exist, return \hs{False}: \hs{k} was never built
+  before and cannot be verified.
+\item If \hs{h} does not equal the \hs{result} field of the trace, return
+  \hs{False}: the current \hs{k}'s value was written externally and needs
+  rebuilding.
+\item For each key \hs{d} in \hs{depends}:\\
+  $\circ$ Make sure \hs{d} is up-to-date, suspending the current task if needed;\\
+  $\circ$ If \hs{d}'s latest \hs{changed} time is greater than \hs{k}'s
+    \hs{built} time, return \hs{False}.
+\item Return \hs{True}: the current \hs{k}'s value is up-to-date.
+\end{itemize}
 
-While verifying step traces are mostly an optimisation, there are some observable differences from verifying traces.
-Considering Figure \ref{fig-make}, imagine we built \hs{main.exe} (everything gets a \hs{built} and \hs{changed} of timestamp 1).
-Next we change \hs{util.c} and rebuild \hs{util.o} (both get a \hs{built} and \hs{changed} of timestamp 2).
-Next we put \hs{util.c} back to what it was originally, and build \hs{main.exe}.
-With verifying traces, the dependencies of \hs{main.exe} would be equal, and \hs{main.exe} would not rebuild.
-With verifying step traces, the \hs{changed} field of \hs{util.o} would increase once more, and \hs{main.exe} would rebuild.
-Other than when building subsets of the targets, we are unaware of any other situation where verifying step traces are less powerful.
+\noindent
+This approach preserves minimality and early cutoff. A variant with only one
+\hs{Time} field would lose early cutoff, and indeed corresponds quite closely to
+\Make. Furthermore, the \hs{Time} stamp only needs to record which execution of
+the build is running, so every key built in the same run can share the same
+\hs{Time} value -- it just needs to be monotonically increasing between runs.
+
+This optimisation is useful, at least in the case of \Shake, to save space. A
+typical cryptographic hash takes up 32 bytes, while a key (in \Shake only) is an
+\hs{Int} taking only~4~bytes. Furthermore, \Shake permits values to be
+arbitrarily large, and permits a custom equality (two values can be bit-for-bit
+unequal but considered equal by \Shake), meaning that a \hs{Hash} is not a valid
+encoding. For build systems that use \hs{Applicative} dependencies, the
+\hs{depends} could be omitted entirely, making the size of a \hs{StepTrace}
+$O(1)$ instead of $O(n)$, where $n$ is the number of dependencies.
+
+\begin{figure}%[h]
+\begin{subfigure}[b]{0.32\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/step-example-step1.pdf}}
+\caption{Initial full build}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{0.32\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/step-example-step2.pdf}}
+\caption{Change \cmd{util.c}, build \cmd{util.o}}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{0.33\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/step-example-step3.pdf}}
+\caption{Restore \cmd{util.c}, build \cmd{main.exe}}
+\end{subfigure}
+% \vspace{-2mm}
+\caption{An example of verifying step traces. The small rectangles show the
+\hs{changed} (left) and \hs{built} (right) timestamps of each non-input key
+in the trace store.
+\label{fig-step}}
+% \vspace{-2mm}
+\end{figure}
+
+While verifying step traces are mostly an optimisation, there are some
+observable differences from verifying traces, as demonstrated by an example in
+Fig.~\ref{fig-step}. We first make the full build: all keys get a \hs{built} and
+\hs{changed} of timestamp 1. Next we change \cmd{util.c} and build \cmd{util.o};
+the latter is changed as a result and hence both \hs{built} and \hs{changed} are
+increased to 2. Finally, we change \cmd{util.c} back to what it was originally,
+and build \cmd{main.exe}. With verifying traces, the hashes of the dependencies
+of \cmd{main.exe} would be equal to the initial build, and \cmd{main.exe} would
+not need rebuilding. With verifying step traces, the \hs{changed} field of
+\cmd{util.o} would increase once more, and \cmd{main.exe} would therefore be
+rebuilt. As shown in Fig.~\ref{fig-step}(c), the \hs{changed} field of
+\cmd{main.exe} remains 1, since the actual value is unchanged. Other than when
+building subsets of the targets, we are unaware of any other situation where
+verifying step traces are less powerful.
 

### 2019-04-28 papers/jfp/5-rebuilders.tex
Minor tweaks to S5

@@
 However, with a monadic task, the dependencies are not known from the task
-alone, they are only recorded from previous executions and stored in \hs{VT}. If
+alone, they are only recorded from previous executions stored in \hs{VT}. If
 the build system has two traces for a given key~\hs{k}, they will both request
 the same dependency first, since \hs{Task}~\hs{Monad} is deterministic. However,
-depending on that first result, they may then request different successive
+based on that first result, they may then request different subsequent
 dependencies using \hs{fetchHash}. A curious result is that for suspending
@@
 in~Mitchell~\shortcite{mitchell2012shake} -- use a slightly different trace
-structure, called \emph{step traces}, one that is similar to verifying traces,
-but stores less data, and has different early cutoff semantics. Rather than
+structure, called \emph{verifying step traces},
+which stores less data than veryifying traces, and has slightly different early cutoff semantics. Rather than
 storing the \hs{Hash}~\hs{v} for each dependency, it instead stores \hs{built}
-time and \hs{changed} time for the \hs{key}, and a list of dependency keys
+time and \hs{changed} time for each \hs{key}, and a list of dependency keys
 (without the hashes). The resulting \hs{StepTrace} type resembles:
@@
 This optimisation is useful, at least in the case of \Shake, to save space. A
-typical cryptographic hash takes up 32 bytes, while a key (in \Shake only) is an
+typical cryptographic hash takes up 32 bytes, while a key (in \Shake) is an
 \hs{Int} taking only~4~bytes. Furthermore, \Shake permits values to be

### 2019-04-29 papers/jfp/5-rebuilders.tex
Polish S5-S6

--- 2019-04-28 papers/jfp/5-rebuilders.tex
@@
 
-\subsubsection{Trace Representation}
+\subsubsection{Trace Representation}\label{sec-trace-representation}
 
@@
 
-A verifying trace deliberately records only small hashes, so that it can be small.
-In contrast, a \emph{constructive} trace also stores the resulting value. Concretely,
-it records a list of \hs{Trace}~\hs{k}~\hs{v}.
-Once we are storing the complete result it makes sense
-to record many constructive traces per key, and to share them with other users,
-providing cloud-build functionality. We can represent this additional
-information with the operations:
+A verifying trace deliberately records only hashes, so that it can be small. In
+contrast, a \emph{constructive} trace also stores the resulting value.
+Concretely, it records a list of \hs{Trace}~\hs{k}~\hs{v}~\hs{v}, where
+\hs{Trace} is as defined in~\S\ref{sec-trace-representation}. Once we are
+storing the complete result it makes sense to record many constructive traces
+per key, and to share them with other users, providing cloud-build
+functionality. We can represent this additional information with the operations:
 
-\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 recordCT    :: k -> v -> [(k, Hash v)] -> CT k v -> CT k v
+\end{minted}
+\vspace{-2mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 constructCT :: (Monad m, Eq k, Eq v)
@@
 has been replaced with \hs{constructCT}, which instead of taking the hash of the
-current value as \emph{input}, returns a list of suitable values as \emph{output}.
-If the current value in the store matches one of the possible values, the build
-can skip this key. If the resulting list is empty, the key must be rebuilt.
-However, if the current value does not match the store, and there is a possible
-value, we can use any value from the constructive list \emph{without} doing any
-work to build it, and copy it into the store.
+current value as \emph{input}, returns a list of possible values as
+\emph{output} -- there may be more than one, because some build tools are
+non-deterministic, see~\S\ref{sec-non-determinism}.
+
+Regardless of the chosen scheduler~(\S\ref{sec-scheduler}), there are three
+cases to consider when using a rebuilder based on constructive traces:
+
+\begin{itemize}
+  \item If \hs{constructCT} returns the empty list of possible values, the key
+        must be rebuilt.
+  \item If the current value in the store matches one of the possible values,
+        the build system can skip this key. Here a constructive trace is used
+        for verifying the current value.
+  \item If the current value in the store does not match any possible value, we
+        can use any of the possible values \emph{without} doing any work to
+        build it, and copy it into the store.
+\end{itemize}
 
 Any \hs{Applicative} build system using constructive traces, e.g.
-\CloudBuild, can index directly from the key and results to the output result~--~i.e.
+\CloudBuild~\S\ref{sec-implementation-cloud}, can index directly from the
+key and the hashes of its dependencies to the output result~--~i.e.
 \hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
 are stored on a central server, the client can compute the key and the hashes of
-its dependencies, then make a single call to the server to retrieve the result.
+its dependencies locally, and then make a single call to the server to retrieve
+the result.
 
-In practice, many cloud build systems store hashes of values in the trace information,
-then have a separate content-addressable cache which associates hashes with
-their actual contents.
+In practice, many cloud build systems store hashes of values in the trace store,
+i.e. use \hs{Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v)} trace entries just like the
+verifying traces, and have a separate content-addressable cache which associates
+hashes with their actual contents.
 
@@
 
-Concretely, taking the example from Figure \ref{fig-make}, to decide whether \hs{main.exe}
-is out of data, a \emph{constructive} trace would look at \hs{util.o} and \hs{main.o}
-(the immediate dependencies), while a \hs{deep constructive} trace would look at \hs{util.c},
-\hs{util.h} and \hs{main.c}.
+Concretely, taking the example from Fig.~\ref{fig-make}, to decide whether
+\cmd{main.exe} is out of date, a \emph{constructive} trace would look at
+\cmd{util.o} and \cmd{main.o} (the immediate dependencies), whereas a
+\emph{deep constructive} trace would look at \cmd{util.c}, \cmd{util.h} and
+\cmd{main.c}.
 
-An advantage of deep constructive traces is that to decide if \hs{main.exe} is up to date only
-requires consulting its inputs, not even considering \hs{util.o} or \hs{main.o}.
-Such a feature is often known as a \emph{shallow build}, as discussed in~\S\ref{sec-cloud-aspects}.
+An advantage of deep constructive traces is that to decide if \cmd{main.exe} is
+up to date only requires consulting its inputs, not even considering
+\cmd{util.o} or \cmd{main.o}. Such a feature is often known as a
+\emph{shallow build}, as discussed in~\S\ref{sec-cloud-aspects}.
 
@@
 
-\begin{description}
-\item[Deterministic tasks] If the tasks are not \emph{deterministic} then it is possible to
-violate correctness, as illustrated by an example in~\S\ref{sec-cloud-aspects}
-(see Fig.~\ref{fig-frankenbuild}).
-\item[No early cutoff] Such traces cannot support early cutoff
-(\S\ref{sec-background-shake}), since the results of intermediate computations are not considered.
-\end{description}
+\begin{itemize}
+  \item \textbf{Tasks must be deterministic}: If the tasks are not
+        \emph{deterministic} then it is possible to violate correctness, as
+        illustrated by an example in~\S\ref{sec-cloud-aspects}, see
+        Fig.~\ref{fig-frankenbuild}.
+  \item \textbf{No early cutoff}: Deep constructive traces cannot support early
+        cutoff (\S\ref{sec-background-shake}), since the results of intermediate
+        computations are not considered.
+\end{itemize}
 

### 2019-04-29 papers/jfp/6-implementations.tex
Polish S5-S6

@@
-\section{Build systems, concretely}\label{sec-implementations} \label{sec-design-space}
+\section{Build systems, concretely}\label{sec-implementations}\label{sec-design-space}
 
@@
 \caption{Build systems \`a la carte.\label{tab-build-systems}}
-\vspace{-4mm}
+\vspace{-2mm}
 % \smaller
@@
 \hline
- & \multicolumn{3}{c}{Scheduling algorithm} \\
-Rebuilding strategy\gap  & \gap{}Topological\gap\S\ref{sec-topological}\gap & \gap{}Restarting\gap\S\ref{sec-restarting}\gap & \gap{}Suspending\gap\S\ref{sec-suspending}\gap    \\\hline
+ & \multicolumn{3}{c}{\textbf{Scheduling algorithm}}\vspace{1.5mm} \\
+\textbf{Rebuilding strategy}\gap  & \gap{}Topological\gap\S\ref{sec-topological}\gap & \gap{}Restarting\gap\S\ref{sec-restarting}\gap & \gap{}Suspending\gap\S\ref{sec-suspending}\gap    \\\hline
 Dirty bit\hfill\S\ref{sec-dirty-bit}                                           & \Make       & \Excel & -              \\
@@
 \end{tabular}
-\vspace{-4mm}
+\vspace{-2mm}
 \end{table}
@@
 
-We can define schedulers and rebuilders more concretely with the types (Fig.~\ref{fig-types}):
+We can define schedulers and rebuilders more concretely with the types
+(Fig.~\ref{fig-types}):
 
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
+\vspace{1mm}
+\begin{minted}[fontsize=\small,xleftmargin=5pt]{haskell}
 type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
-type Rebuilder c   ir k v =
-    k -> v -> Task c k v -> Task (MonadState ir) k v
+type Rebuilder c   ir k v = k -> v -> Task c k v -> Task@\,@(MonadState ir) k v
 \end{minted}
-\vspace{0.5mm}
+\vspace{1mm}
 
@@
 \end{minted}
-\vspace{-3mm}
+% \vspace{-3mm}
 \caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
-\vspace{-5mm}
+% \vspace{-5mm}
 \end{figure}
@@
 \hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
-the new modification time of the \hs{key}\footnote{The real \Make
-relies on the file system to track file modification times, but we prefer
-to make it explicit here.}; otherwise we can just return the current \hs{value}.
+the new modification time of the \hs{key}\footnote{The real \Make relies on the
+file system to track file modification times, but we prefer to make this
+explicit in our model.}; otherwise we can just return the current \hs{value}.
 
@@
 implementation of textbook graph algorithms \hs{reachable} and \hs{topSort},
-e.g. see~\cite{cormen2001introduction}.
+e.g. see~Cormen~\etal~\shortcite{cormen2001introduction}.
 
@@
     chain    <- gets (snd . getInfo)
-    newChain <- liftChain $ go Set.empty $ chain ++ [target | target `notElem` chain]
+    newChain <- liftChain $ go Set.empty
+                          $ chain ++ [target | target `notElem` chain]
     modify $ mapInfo $ \(ir, _) -> (ir, newChain)
@@
                     | otherwise           = return $ Left k
-            result <- liftStore (run newTask fetch) -- liftStore is defined in Fig. 7
+        result <- liftStore (run newTask fetch) -- liftStore is in Fig. 8
         case result of
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Expand the scope of visibility of a stateful computation (implementation omitted)
+-- Expand the scope of visibility of a stateful computation (omitted)
 liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
 \end{minted}
-\vspace{-2.5mm}
+% \vspace{-2.5mm}
 \caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
-\vspace{-2.5mm}
+% \vspace{-2.5mm}
 \end{figure}
@@
 
-Our model of \Shake is shown in Fig.~\ref{fig-shake-implementation}. It stores
-verifying traces \hs{VT}~\hs{k}~\hs{v} defined in~\S\ref{sec-verifying-traces}
-as persistent build information and is composed of the \hs{suspending} scheduler
+Our model of \Shake (Fig.~\ref{fig-shake-implementation}) stores verifying
+traces \hs{VT}~\hs{k}~\hs{v} defined in~\S\ref{sec-verifying-traces} as
+persistent build information and is composed of the \hs{suspending} scheduler
 and the \hs{vtRebuilder}.
@@
 suspending :: Ord k => Scheduler Monad i i k v
-suspending rebuilder tasks target store = fst $ execState@\,@(fetch target)@\,@(store,@\,@Set.empty)
+suspending rebuilder tasks target store =
+    fst $ execState (fetch target) (store, Set.empty)
   where
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Run a task using a callback that operates on a larger state (implementation omitted)
+-- Run a task using a callback that operates on a larger state (omitted)
 liftRun :: Task (MonadState i) k v
@@
 \end{minted}
-\vspace{-1mm}
+% \vspace{-1mm}
 \caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
-\vspace{-3mm}
+% \vspace{-3mm}
 \end{figure}
 
-The rebuilder executes the verification query \hs{verifyVT} to determine if the
-\hs{key} is \hs{upToDate}. If it is, the rebuilder simply returns the \hs{key}'s
-current \hs{value}. Otherwise it executes the \hs{task}, obtaining both a
-\hs{newValue} and the \hs{key}'s dynamic dependencies \hs{deps} (see the
-definition of \hs{track} in~\S\ref{sec-deps}), which are subsequently recorded
-in a new verification trace using \hs{recordVT}.
+The rebuilder performs the \hs{verifyVT} query to determine if the \hs{key} is
+\hs{upToDate}. If it is, the rebuilder simply returns the \hs{key}'s current
+\hs{value}. Otherwise it executes the \hs{task}, obtaining both a \hs{newValue}
+and the \hs{key}'s dynamic dependencies \hs{deps} (see the definition of
+\hs{track} in~\S\ref{sec-deps}), which are subsequently recorded in the trace
+store using \hs{recordVT}.
 
@@
 
+Fig.~\ref{fig-cloud-implementations} shows our models of several cloud build
+systems. \Bazel, \CloudBuild and \Cloud \Shake are based on constructive traces
+(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
+variant (\S\ref{sec-deep-constructive-traces}).
+
+The implementation of \hs{ctRebuilder} is analogous to that of \hs{vtRebuilder}
+in Fig.~\ref{fig-shake-implementation}, but the \hs{verifyVT} query is replaced
+with a more powerful query to \hs{constructCT} that returns a list of suitable
+\hs{cachedValues} by looking them up the cloud cache. If the current \hs{value}
+is in the list, we can use it as is. Otherwise, if the list is non-empty, we can
+use an arbitrary \hs{cachedValue}. Finally, if the cache has no suitable values,
+we fall back to executing the \hs{task}. The obtained \hs{newValue} and the
+\hs{task}'s dependencies are recorded as a new constructive trace for future
+use.
+
 \begin{figure}
@@
 bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
-bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
+bazel = restartingQ ctRebuilder
+\end{minted}
+\vspace{0mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- A restarting scheduler based on a build queue, omitted (22 lines)
+restartingQ :: (Hashable v, Eq k) => Scheduler Monad (CT k v) (CT k v) k v
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
+-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 10
 cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
@@
 \begin{minted}[fontsize=\small]{haskell}
--- CloudBuild build system, implementation of 'topological' is given in Fig. 7
+-- CloudBuild build system, implementation of 'topological' is given in Fig. 8
 cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Buck build system, implementation of 'topological' is given in Fig. 7
+-- Buck build system, implementation of 'topological' is given in Fig. 8
 buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
@@
 \begin{minted}[fontsize=\small]{haskell}
--- Nix build system, implementation of 'suspending' is given in Fig. 9
+-- Nix build system, implementation of 'suspending' is given in Fig. 10
 nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
@@
 \end{minted}
-\vspace{-2mm}
-\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
-our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
-(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
-chain.}
+% \vspace{-2mm}
+\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix in our framework.}
 \label{fig-cloud-implementations}
-\vspace{-2mm}
+% \vspace{-2mm}
 \end{figure}
-
-Fig.~\ref{fig-cloud-implementations} shows our models of several cloud build
-systems. \Bazel, \CloudBuild and \Cloud \Shake are based on constructive traces
-(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
-variant (\S\ref{sec-deep-constructive-traces}).
-
-The implementation of \hs{ctRebuilder} is analogous to that of \hs{vtRebuilder}
-in Fig.~\ref{fig-shake-implementation}, but the \hs{verifyVT} query is replaced
-with a more powerful query to \hs{constructCT} that returns a list of suitable
-\hs{cachedValues} by looking them up the cloud cache. If the current \hs{value}
-is in the list, we can use it as is. Otherwise, if the list is non-empty, we can
-use an arbitrary \hs{cachedValue}. Finally, if the cache has no suitable values,
-we fall back to executing the \hs{task}. The obtained \hs{newValue} and the
-\hs{task}'s dependencies are recorded as a new constructive trace for future
-use.
 
@@
 
-Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the
-\hs{dctRebuilder}, a rebuilder based on deep constructive traces
-(\S\ref{sec-deep-constructive-traces}), whose implementation we omit since it
-is very similar to that of \hs{ctRebuilder}. \Buck uses the \hs{topological}
-scheduler and is therefore an applicative build system, whereas \Nix uses the
-\hs{suspending} scheduler and is monadic.
+Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the rebuilder
+based on deep constructive traces (\S\ref{sec-deep-constructive-traces}), called
+\hs{dctRebuilder}, whose implementation we omit since it is very similar to that
+of \hs{ctRebuilder}. \Buck uses the \hs{topological} scheduler and is therefore
+an applicative build system, whereas \Nix uses the \hs{suspending} scheduler and
+is monadic.
+
+\todo{AM}{Add implementation of \hs{dctRebuilder}?}
 

### 2019-04-29 papers/jfp/7-experience.tex
More work tidying up S8
Drop some TODOs
Polish S7-S9

@@
 
-As a set of authors, we've written a paper about the \Shake build system \cite{mitchell2012shake},
-applying \Shake to building GHC \cite{hadrian}, and a paper about the categorisation of build
-systems \cite{mokhov2018buildsystems}. The original motivation of the research in that final paper was to create
-``\Cloud \Shake'', and apply it to the GHC build system. In this section we tie these papers together,
-reflecting on our experience since these papers.
+As a set of authors, we've written a paper about the \Shake build
+system~\cite{mitchell2012shake}, applying \Shake to building GHC~\cite{hadrian},
+and a paper about the categorisation of build
+systems~\cite{mokhov2018buildsystems}. The original motivation of the research
+in that final paper was to create ``\Cloud \Shake'', and apply it to the GHC
+build system. In this section we tie these papers together, reflecting on our
+experience since these papers.
 
@@
 
-The original design of \Shake hasn't changed since the initial paper, but the implementation has continued to mature -- there have been roughly 5,000 subsequent commits to the \Shake project\footnote{\url{https://github.com/ndmitchell/shake}}.
-These commits add concepts like resources (if two rules contend on a single external resource), rewriting serialisation to be faster, documentation including a website\footnote{\url{https://shakebuild.com}}, and add lots of tests.
-The biggest change in that time period was moving from blocking threads to continuations for the suspending scheduler.
-The most visible change is switching from \hs{*>} to \hs{%>} for defining rules\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.
+The original design of \Shake has not changed since the initial paper, but the
+implementation has continued to mature -- there have been roughly 5,000
+subsequent commits to the \Shake
+project\footnote{\url{https://github.com/ndmitchell/shake}}. These commits add
+concepts like \emph{resources} (for handling situations when two build tasks
+contend on a single external resource), rewriting serialisation to be faster,
+documentation including a website\footnote{\url{https://shakebuild.com}}, and
+add a lot of tests. The biggest change in that time period was moving from
+blocking threads to continuations for the suspending scheduler. The most visible
+change is switching from \hs{*>} to \hs{@\%@>} for defining rules\footnote{A
+conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost
+all external and internal details remain the same.
 
-We consider the lack of change suggestive that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}.
-There are two main aspects to the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, Firstly, the rebuilder can now be described as verifying step traces as per \S\ref{sec-step-traces}, with a much clearer relationship to the unoptimised verifying traces of \S\ref{sec-verifying-traces}.
-The second is the power of \Shake tasks (called actions in the original paper).
-Originally these tasks were described using the data type:
+We consider the lack of change suggestive that \Shake is based on fundamental
+principles -- principles we can now name and describe as a consequence
+of the categorisation paper~\cite{mokhov2018buildsystems}. There are two main
+aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
+more clearly in this paper. Firstly, the rebuilder can now be described using
+verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
+to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. The second
+aspect is the power of \Shake tasks, or ``actions'' in the original paper,
+which were originally described using this data type:
 
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Action k v r = Finished r
-                  | Depends v (v -> Action k v r)
+data Action k v a = Finished a
+                  | Depends k (v -> Action k v a)
 \end{minted}
+\vspace{1mm}
 
-This type is equivalent to \hs{Task Monad} -- the two types form an isomorphism.
-The original paper uses concrete types \hs{Key} and \hs{Value}, we generalise these types to \hs{k} and \hs{v}, and also add \hs{r} so that \hs{Action k v} can be an instance of \hs{Monad}.
-We define that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type, witnessing the isomorhpism:
+\noindent
+This type is equivalent to \hs{Task}~\hs{Monad} -- the two types form an
+isomorphism. The original paper uses concrete types \hs{Key} and \hs{Value}, we
+generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
+\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}. We define that
+\hs{Monad} instance, and use it to write conversion functions to/from our
+\hs{Task} type, witnessing the isomorphism:
 
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
     Finished x    >>= f = f x
-    Depends ds op >>= f = Depends ds $ \v -> f =<< op v
+    Depends ds op >>= f = Depends ds $ \v -> op v >>= f
 
@@
 fromAction x = Task $ \fetch -> f fetch x
-    where f fetch (Finished v) = return v
-          f fetch (Depends d op) = f fetch . op =<< fetch d
+  where
+    f _     (Finished v  ) = return v
+    f fetch (Depends d op) = fetch d >>= f fetch . op
 \end{minted}
+\vspace{1mm}
 
-\noindent Similarly, \Make tasks were described as equivalent to
+\noindent
+Similarly, \Make tasks were described as equivalent to
 
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Rule k v r = Rule {depends :: [k], action :: [v] -> r}
+data Rule k v a = Rule { depends :: [k], action :: [v] -> a }
 \end{minted}
+\vspace{1mm}
 
-Once again, this type is isomorphic to \hs{Task Applicative}, and we can define a similar \hs{Applicative} instance and conversion functions.
-By describing these types using \hs{Task} we are able to describe the differences more concisely (\hs{Monad} vs \hs{Applicative}) and use existing literature to determine what is and isn't possible.
+Assuming the lengths of the lists \hs{[@@k]} and \hs{[@@v]} always match, this
+type is isomorphic to \hs{Task}~\hs{Applicative}, and we can define a similar
+\hs{Applicative} instance and conversion functions. By describing these types
+using \hs{Task} we are able to describe the differences more concisely
+(\hs{Monad} vs \hs{Applicative}), use existing literature to determine what is
+and isn't possible, and explore other constraints beyond just \hs{Monad} and
+\hs{Applicative}.
 
@@
 
-\todo{AM: Can you flesh this out}
+\todo{AM}{Finish this section.}
 
-The build system used to compile GHC, known as Hadrian, has continued development.
-The use of dynamic dependencies has made the build much easier, and it's now more maintainable.
-It's now merged into the GHC repo and tested as standard by the CI.
-It's current state is ...
-In particular, it takes N minutes to build with \Shake, vs M minutes to build with \Make.
-The \Shake version has attracted more contributors, and been easier to modify, etc.
-The code is N lines.
+The build system used to compile GHC, known as Hadrian, has continued
+development. The use of dynamic dependencies has made the build much easier, and
+it's now more maintainable. It's now merged into the GHC repo and tested as
+standard by the CI. It's current state is ... In particular, it takes N minutes
+to build with \Shake, vs M minutes to build with \Make. The \Shake version has
+attracted more contributors, and been easier to modify, etc. The code is N
+lines.
 
-The design we originally outlined remains the one in use.
-There has been a huge amount of engineering work to make the tests pass (many of the details encoded in the \Make build system were incidental -- but some were essential).
+The design we originally outlined remains the one in use. There has been a huge
+amount of engineering work to make the tests pass (many of the details encoded
+in the \Make build system were incidental -- but some were essential).
 
-One of the benefits of using Shake is that we have access to high quality profiling information, allowing us to compute critical paths and other metrics (see \cite{mitchell2019ghcrebuildtimes} for a tour of the metrics).
-This information has shown us that GHC is slow to build (40m on an 8 CPU machine on Windows), that more CPUs would not help (on unlimited CPUs it would take at least 37m), and that a handful of steps (two Haskell compiles, some calls to \hs{configure}) are responsible for a significant amount of that time (at least 10m).
+One of the benefits of using Shake is that we have access to high quality
+profiling information, allowing us to compute critical paths and other metrics,
+see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for a tour of the metrics.
+This information has shown us that GHC is slow to build (40m on an 8 CPU machine
+on Windows), that more CPUs would not help (on unlimited CPUs it would take at
+least 37m), and that a handful of steps (two Haskell compiles, some calls to
+\cmd{configure}) are responsible for a significant amount of that time (at least
+10m).
 
@@
 
-Converting \Shake into \Cloud \Shake wasn't a difficult process once armed with a roadmap from \cite{mokhov2018buildsystems}. The key was the introduction of two new functions:
+Converting \Shake into \Cloud \Shake was not a difficult process once armed with
+a roadmap~\cite{mokhov2018buildsystems}. The key was the introduction of two new
+functions:
 
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 \end{minted}
+\vspace{1mm}
 
-These functions are suspiciously like \hs{recordCT} and \hs{constructCT} from \S\ref{sec-constructive-traces}, with their differences perhaps the most illustrative of the changes required.
-(We have made some minor changes from actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant differences.)
+\noindent
+These functions are suspiciously like \hs{recordCT} and \hs{constructCT}
+from~\S\ref{sec-constructive-traces}, with their differences perhaps the most
+illustrative of the changes required. (We have made some minor changes from
+actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant
+differences.)
 
 \begin{itemize}
-\item There are two \hs{Ver} arguments being passed to each function.
-These are the versions of the build script, and the rule for this particular key.
-If either version changes then it is as though the key has changed, and nothing will match.
-These versions are important to avoid using stale build products from previous versions of the build script.
-\item The list of dependencies to \hs{addCloud} is a list of lists, rather than a simple list.
-The reason is that \Shake allows a list of dependencies to be supplied at once, so they can all be built in parallel.
-\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the files that this rule produces.
-These produced files include those which are output keys from a rule and those declared with \hs{produces}.
-\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking up a dependent key, since some keys are not buildable.
-\item The \hs{lookupCloud} function returns at most one result, rather than a list. This change was made for simplicity.
+\item There are two \hs{Ver} arguments being passed to each function. These are
+      the versions of the build script, and the rule for this particular key. If
+      either version changes then it is as though the key has changed, and
+      nothing will match. These versions are important to avoid using stale
+      build products from previous versions of the build script.
+\item The list of dependencies to \hs{addCloud} is a list of lists, rather than
+      a simple list. The reason is that \Shake allows a list of dependencies to
+      be supplied at once, so they can all be built in parallel.
+\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
+      files that this rule produces. These produced files include those which
+      are output keys from a rule and those declared with the function
+      \hs{produces}.
+\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking
+      up a dependent key, since some keys are not buildable.
+\item The \hs{lookupCloud} function returns at most one result, rather than a
+      list. This change was made for simplicity.
 \end{itemize}
 
-To integrate these functions into \Shake we found the most expedient route was to leave \Shake with verifying traces, but if the verifying trace doesn't match, we consult the constructive trace.
-By bolting constructive traces onto the side of Shake we avoid reengineering of the central database.
-We haven't found any significant downsides from the bolt-on approach thus far, so it may be a sensible route to go even if developing from scratch -- allowing an optimised verified trace implementation in many cases, and falling back to a more complex implementation (requiring consulting remote servers) only rarely.
+To integrate these functions into \Shake we found the most expedient route was
+to leave \Shake with verifying traces, but if the verifying trace does not
+match, we consult the constructive trace. By bolting constructive traces onto
+the side of Shake we avoid re-engineering of the central database. We have not
+found any significant downsides from the bolt-on approach thus far, so it may be
+a sensible route to go even if developing from scratch -- allowing an optimised
+verified trace implementation in many cases, and falling back to a more complex
+implementation (requiring consulting remote servers) only rarely.
 
-The one thing we haven't yet completed on the engineering side is a move to hosting caches over HTTP.
-At the moment all caches are on shared file systems.
-This approach can use mounted drives to mirror HTTP connections onto file systems, and reuse tools for managing file systems, share caches with rsync, and is simple.
-Unfortunately, on certain OS's (e.g.
-Windows) mounting an HTTP endpoint as a file system requires administrator privileges, so an HTTP cache is still desirable.
+The one thing we have not yet completed on the engineering side is a move to
+hosting caches over HTTP. At the moment all caches are on shared file systems.
+This approach can use mounted drives to mirror HTTP connections onto file
+systems, and reuse tools for managing file systems, share caches with
+\cmd{rsync}, and is simple. Unfortunately, on certain operating systems (e.g.
+Windows) mounting an HTTP endpoint as a file system requires administrator
+privileges, so an HTTP cache is still desirable.
 
@@
 
-While we expected the GHC build system to be the first to take advantage of \Cloud \Shake, we were actually beaten to it by Standard Chartered who report\footnote{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}:
+While we expected the GHC build system to be the first to take advantage of
+\Cloud \Shake, we were actually beaten to it by Standard Chartered who
+report\footnote{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}:
 
-\begin{quote}
-Thanks for the symlinks release, we just finished upgrading this build system to use --share. ...
-Building from scratch with a warm cache takes around 5 seconds, saving us up to 2 hours. Not bad!
-\end{quote}
+\vspace{1mm}
+\begin{center}
+\parbox{0.8 \textwidth}{\emph{Thanks for the symlinks release, we just finished upgrading this build
+system to use \cmd{-}\cmd{-share}. ... Building from scratch with a warm cache
+takes around 5 seconds, saving us up to 2 hours. Not bad!}}
+\end{center}
+\vspace{1mm}
 
-We have also got the GHC build working with \Cloud \Shake, resulting in build times of N minutes instead of M minutes.
-Due to aspects like \hs{configure} we don't see the dramatic reductions reported by Standard Chartered.
-Converting to a build suitable for sharing is not overly onerous, but nor is it trivial.
-Some things that require attention:
+We have also got the GHC build working with \Cloud \Shake, resulting in build
+times of 3m and 30s instead of 1h and 8m (on Windows). Due to aspects like slow
+and fully sequential \cmd{configure} we do not see the dramatic reductions
+reported by Standard Chartered. Converting to a build suitable for sharing is
+not overly onerous, but nor is it trivial. Some things that require attention:
 
-\begin{description}
-\item[Irrelevant differences] A common problem is that you don't get shared caching when you want it.
-To get caching, it is important that all machines are identical for the aspects that involve the build system.
-In particular, that means that all paths must be relative, and all dependencies on system binaries (e.g.
-\hs{gcc}) should either be controlled by the build system (e.g.
-using \Nix) or have tolerence to account for minor machine variations.
-\item[Insufficient produced files] A build rule must declare which files it produces, so these are stored in the cache and can be restored from the cache.
-If you don't use a remote cache then it is permissible to have files not known to the build system which change at the same time as files that are known.
-Once using a remote cache any file that is produced as a side-effect of a rule and then later used \emph{must} be declared.
-Failure to declare such files means that if some step produces a file, and a later step relies on it, that step breaks.
-In practice, most issues encountered during the move to Cloud builds for GHC were a lack of declared produced files.
-\item[Insufficient dependencies] While insufficient dependencies are often a problem, the move to a Cloud build makes them even more critical, as missing dependencies are likely to mean a result is \emph{never built locally}, rather than just build locally only once.
-Furthermore, dependencies in a Cloud setting must be more complete.
-For example, using the build system from Figure \ref{fig-make}, with a local build system without early cutoff, if the rule to produce \hs{main.exe} also made use of \hs{main.c} but didn't declare so, that isn't a problem.
-With the addition of early cutoff, if \hs{main.c} changes in an important way but \hs{main.o} doesn't update that causes issues, but is a somewhat rare occurance.
-With Cloud builds that becomes somewhat more common.
-\end{description}
+\begin{itemize}
+\item \textbf{Irrelevant differences}: A common problem is that you do not get
+      shared caching when you want it. To get caching, it is important that all
+      machines are identical for the aspects that involve the build system. In
+      particular, that means that all paths must be relative, and all
+      dependencies on system binaries (e.g. \cmd{gcc}) should either be
+      controlled by the build system (e.g. using \Nix) or have tolerance to
+      account for minor machine variations.
+\item \textbf{Insufficient produced files}: A build rule must declare all files
+      it produces, so these are stored in the cache and can be restored from the
+      cache. If you do not use a remote cache then it is permissible to have
+      files not known to the build system which change at the same time as files
+      that are known. Once using a remote cache any file that is produced as a
+      side-effect of a rule and then later used \emph{must} be declared. Failure
+      to declare such files means that if some step produces a file, and a later
+      step relies on it, that step breaks. In practice, most issues encountered
+      during the move to cloud builds for GHC were a lack of declared produced
+      files.
+\item \textbf{Insufficient dependencies}: While insufficient dependencies are
+      often a problem, the move to a cloud build makes them even more critical,
+      as missing dependencies are likely to mean a result is \emph{never built
+      locally}, rather than just build locally only once. Furthermore,
+      dependencies in a cloud setting must be more complete. For example, using
+      the build system from Fig.~\ref{fig-make}, with a local build system
+      without early cutoff, if the rule to produce \cmd{main.exe} also made use
+      of \cmd{main.c} but did not declare so, that is not a problem. With the
+      addition of early cutoff, if \cmd{main.c} changes in an important way but
+      \cmd{main.o} is not rebuilt that causes issues, but is a somewhat rare
+      occurrence. With cloud builds that becomes somewhat more common.
+\end{itemize}
 
-To help with the final two issues -- insufficient dependencies and produced files -- we have further enhanced the \Shake lint modes, coupling them to a program called \hs{FSATrace}, which detects which files are read/written by a command line execution.
-Such information has been very helpful in making the GHC build Cloud ready.
+To help with the final two issues -- insufficient dependencies and produced
+files -- we have further enhanced the \Shake lint modes, coupling them to a
+program called \cmd{FSATrace}, which detects which files are read/written by a
+command line execution. Such information has been very helpful in making the GHC
+build cloud ready.

### 2019-04-29 papers/jfp/8-engineering.tex
Drop some TODOs
Polish S7-S9

--- 2019-04-28 papers/jfp/8-engineering.tex
@@
 because they are out of date but because of races with earlier tasks that had
-not yet finished. As a consequence, if a the calcuation order is persisted over
-successive runs (as \Excel does), potentially racey dependencies will be separated,
-giving better parallelism over time.
+not yet finished. As a consequence, if a the calculation order is persisted over
+successive runs (as \Excel does), potentially racey dependencies will be
+separated, giving better parallelism over time.
 
 \item The \hs{suspending} scheduler can be made parallel by starting multiple
-dependencies simulataneously. One approach is to make the request for dependencies
+dependencies in parallel. One approach is to make the request for dependencies
 take a list of keys, as implemented by \Shake. Another approach is to treat the
-\hs{Applicative} dependencies of a \hs{Task Monad} in parallel, using the strategy
-described by~\cite{marlow2014haxl}.
+\hs{Applicative} dependencies of a \hs{Task}~\hs{Monad} in parallel, as
+described by~Marlow~\etal~\shortcite{marlow2014haxl}.
 \end{enumerate}
@@
 Once sufficient parallelism is available the next challenge is preventing excess
-parallelism and machine resource starvation, which is usually acheived with a thread pool or thread limit.
+parallelism and machine resource starvation, which is usually achieved with a
+thread pool or thread limit.
 
@@
 
-\begin{description}
-\item[Untracked dependencies] Some tasks depend on untracked values -- for example
-C compilation will explicitly list the \cmd{source.c} file as a dependency,
-but it may not record that the version of \cmd{gcc} is also a dependency.
+\begin{itemize}
+\item \textbf{Untracked dependencies}: Some tasks depend on untracked values --
+      for example C compilation will explicitly list the \cmd{source.c} file as
+      a dependency, but it may not record that the version of \cmd{gcc} is also
+      a dependency.
 
-\item[Non-determinism] Some tasks are \emph{non-deterministic}, producing any
-result from a possible set. As an example, GHC programs compiled using parallelism
-can change the order in which unique variables are obtained from the supply,
-producing different but semantically identical results.
+\item \textbf{Non-determinism}: Some tasks are \emph{non-deterministic},
+      producing any result from a possible set. As an example, GHC programs
+      compiled using parallelism can change the order in which unique variables
+      are obtained from the supply, producing different but semantically
+      identical results.
 
-\item[Volatility] Some tasks are defined to change in every build, e.g. \Excel
-provides a~``function'' \cmd{RANDBETWEEN} producing a fresh random number in a
-specified range on each recalculation. Similarly, build systems like \Make and
-\Shake provide volatile \emph{phony rules}. The main difference from
-non-deterministic tasks is that volatile tasks cannot be cached. They are best
-modelled as depending on a special key \cmd{RealWorld} whose value is changed
-in every build.
-\end{description}
+\item \textbf{Volatility}: Some tasks are defined to change in every build, e.g.
+      \Excel provides a~``function'' \cmd{RANDBETWEEN} producing a fresh random
+      number in a specified range on each recalculation. Similarly, build
+      systems like \Make and \Shake provide volatile \emph{phony rules}. The
+      main difference from non-deterministic tasks is that volatile tasks cannot
+      be cached. They are best modelled as depending on a special key
+      \cmd{RealWorld} whose value is changed in every build.
+\end{itemize}
 
@@
 
-\todo{AM}{Show implementation of \hs{computeND}}.
+% \todo{AM}{Show implementation of \hs{computeND}.
+% Neil says: not convinced it's that interesting.}.
 
@@
 
-\begin{description}
-\item[Communication] When traces or contents are stored in the cloud,
+\begin{itemize}
+\item \textbf{Communication}: When traces or contents are stored in the cloud,
 communication can become a bottleneck, so it is important to send only the
@@
 
-\item[Offloading] Once the cloud is storing build products and traces, it is
-possible for the cloud to also contain dedicated workers that can execute tasks
-remotely~--~offloading some of the computation and potentially running vastly
-more commands in parallel.
+\item \textbf{Offloading}: Once the cloud is storing build products and traces,
+it is possible for the cloud to also contain dedicated workers that can execute
+tasks remotely~--~offloading some of the computation and potentially running
+vastly more commands in parallel.
 
-\item[Eviction] The cloud storage, as modelled
+\item \textbf{Eviction}: The cloud storage, as modelled
 in~\S\ref{sec-constructive-traces}, grows indefinitely, but often resource
@@
 
-\item[Shallow builds] Building the end result, e.g. an installer package, often
-involves many intermediate tasks. The intermediate results may be large, so some
-cloud build systems are designed to build end products \emph{without
-materialising} intermediate results, producing a so-called \emph{shallow build}
-(see an example in~\S\ref{sec-background-bazel}). Some build systems go even
-further, integrating with the file system to only materialise the file when the
-user accesses it \cite{gvfs}.
-\end{description}
+\item \textbf{Shallow builds}: Building the end result, e.g. an installer
+package, often involves many intermediate tasks. The intermediate results may be
+large, so some cloud build systems are designed to build end products
+\emph{without materialising} intermediate results, producing a so-called
+\emph{shallow build} (see an example in~\S\ref{sec-background-bazel}). Some
+build systems go even further, integrating with the file system to only
+materialise the file when the user accesses it~\cite{gvfs}.
+\end{itemize}
 
@@
 \begin{figure}
-\begin{subfigure}[b]{0.32\linewidth}
+\begin{subfigure}[b]{0.31\linewidth}
 \centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-build.pdf}}
@@
 \end{subfigure}
-\begin{subfigure}[b]{0.32\linewidth}
+\begin{subfigure}[b]{0.31\linewidth}
 \centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-clean.pdf}}
@@
 \end{subfigure}
-\begin{subfigure}[b]{0.34\linewidth}
+\begin{subfigure}[b]{0.36\linewidth}
 \centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-rebuild.pdf}}
 % \vspace{-1.5mm}
-\caption{Build \cmd{main.prof} \& \cmd{report.txt}}
+\caption{Build \cmd{main.prof} and \cmd{report.txt}}
 \end{subfigure}
@@
 
-\todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.}
+% \todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote
+% execution service. Neil says: I think this is a stretch. It's a lot more nuanced
+% as we saw from Bazel discussions so I'd probably skip it.}
 
@@
 For examples like \LaTeX~we consider the proper encoding to not be circular
-tasks, but a series of iterative steps, as described
-by~\cite{shake-fixed-point}. It is important that the number of executions is
-bounded, otherwise the build system may not terminate (a legitimate concern
-with \LaTeX, which can be put into a situation where it is bistable or diverging
-over multiple executions). The examples in \Excel tend to encode either mutable
-state, or recurrence relations. The former is only required because \Excel
-inherently lacks the ability to write mutable state, and the latter is probably
-better solved using explicit recurrence formulae.
+tasks, but a series of iterative steps, as described by
+Mitchell~\shortcite{shake-fixed-point}. It is important that the number of
+executions is bounded, otherwise the build system may not terminate (a
+legitimate concern with \LaTeX, which can be put into a situation where it is
+bistable or diverging over multiple executions). The examples in \Excel tend to
+encode either mutable state, or recurrence relations. The former is only
+required because \Excel inherently lacks the ability to write mutable state, and
+the latter is probably better solved using explicit recurrence formulae.
 
@@
 cyclic dependencies under the assumption that tasks are \emph{monotonic} in a
-certain sense, e.g. see~\cite{pottier2009lazy} and~\cite{radul2009propagation}.
+certain sense~\cite{pottier2009lazy,radul2009propagation}.
 
@@
 \Shake users have remarked that polymorphism provides a much easier expression
-of concepts, e.g. see~\cite{hadrian}.
+of concepts, e.g. see~Mokhov~\etal~\shortcite{hadrian}.
 
@@
 \cmd{gcc}~--~for many builds it should be a dependency. In \Shake it is possible
-to define an \emph{oracle} rule as per~\cite{mitchell2012shake} whose value is
-the result of running \cmd{gcc -}\cmd{-version} and which is volatile, making the
-\cmd{gcc} version something that can be depended upon. Of course, provided the
-build can express volatile dependencies and supports cutoff, the version number
-could equally be written to a file and used in a similar way, albeit at the cost
-of introducing additional correctness invariants, e.g. that the file
-\cmd{gcc-version.txt} does indeed contains two dot-separated integers.
+to define an \emph{oracle} rule as per Mitchell~\shortcite{mitchell2012shake}
+whose value is the result of running \cmd{gcc -}\cmd{-version} and which is
+volatile, making the \cmd{gcc} version something that can be depended upon. Of
+course, provided the build can express volatile dependencies and supports
+cutoff, the version number could equally be written to a file and used in a
+similar way, albeit at the cost of introducing additional correctness
+invariants, e.g. that the file \cmd{gcc-version.txt} does indeed contains two
+dot-separated integers.
 
-Generalised algebraic data types (GADTs)~\cite{spj2006gadts} can be used to
+Generalised algebraic data types, or GADTs~\cite{spj2006gadts}, can be used to
 extend our models to support ``typed'' build tasks. The idea is to replace
@@
 
-\todo{AM/NM}{We are talking about sets above, but are using lists, which is a
-bit inconsistent. We could switch from \hs{[k]} to \hs{Set k} fairly easily.
-Shall we?}
+% \todo{AM/NM}{We are talking about sets above, but are using lists, which is a
+% bit inconsistent. We could switch from \hs{[k]} to \hs{Set k} fairly easily.
+% Shall we? Neil says: Haskell code dealing with sets is quite a lot more complex.
+% People are used to being quite fuzzy around the difference - I'd leave it as is.
+% Andrey says: Also, using \hs{Set v} doesn't seem possible, because we won't be
+% able to relate \hs{v}s with corresponding \hs{k}s}.
 

### 2019-04-29 papers/jfp/9-related.tex
Polish S7-S9

@@
 The one build system we are aware of that cannot be modelled in our framework is
-\Fabricate by \cite{fabricate}. In \Fabricate a build system is a script that is
-run in-order, in the spirit of:
-% I'm not sure the footnote adds much, let's save space?
-% \footnote{\Fabricate requires scripts to be
-% written in Python, but those details are not fundamental to what makes
-% \Fabricate special.}
+\Fabricate by~Hoyt~\etal~\shortcite{fabricate}. In \Fabricate a build system is
+a script that is run in-order, in the spirit of:
 
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{bash}
@@
 monadic \hs{Task} function\footnote{\Shake provides support for
-\Fabricate{}-like build systems~--~see \cmd{Development.Shake.Forward}
-in the \Shake library.}. However, in our pure model the mapping is not perfect
-as \cmd{gcc} writes to arbitrary files whose locations are not known in advance.
+\Fabricate{}-like build systems~--~see \cmd{Development.Shake.Forward}.}.
+However, in our pure model the mapping is not perfect as \cmd{gcc} writes to
+arbitrary files whose locations are not known in advance.
 
@@
 well studied area, and in particular the contrast between different formulations
-has been thoroughly investigated, e.g. see~\cite{acar2007selfadjusting}.
-Self-adjusting computations can automatically adjust to an external change
-to their inputs. A classic example is a self-adjusting sorting algorithm, which
-can efficiently (in $O(\log{n})$ time where $n$ is the length of the input)
-recalculate the result given an incremental change of the input. While very
-close to build systems in spirit, self-adjusting computations are mostly used
-for in-memory computation and rely on the ability to dynamically allocate new
-keys in the store for sharing intermediate computations~--~an intriguing feature
-rarely seen in build systems (\Shake's oracles~\S\ref{sec-polymorphism} can be
-used to model this feature to a limited degree).
+has been thoroughly investigated, e.g.
+see~Acar~\etal~\shortcite{acar2007selfadjusting}. Self-adjusting computations
+can automatically adjust to an external change to their inputs. A classic
+example is a self-adjusting sorting algorithm, which can efficiently (in
+$O(\log{n})$ time where $n$ is the length of the input) recalculate the result
+given an incremental change of the input. While very close to build systems in
+spirit, self-adjusting computations are mostly used for in-memory computation
+and rely on the ability to dynamically allocate new keys in the store for
+sharing intermediate computations~--~an intriguing feature rarely seen in build
+systems (\Shake's oracles~\S\ref{sec-polymorphism} can be used to model this
+feature to a limited degree).
 
@@
 algorithms for self-adjusting computations (with a few open-source
-implementations, e.g. \Incremental by~\cite{incremental}). We plan to
-investigate how these insights can be utilised by build systems as future work.
+implementations, e.g. \Incremental by Jane Street~\shortcite{incremental}. We
+plan to investigate how these insights can be utilised by build systems as
+future work.
 
@@
 \emph{replacing} a symbol. The dynamic programming solution of this problem is
-so widely known (e.g., see~\cite{cormen2001introduction}) that we provide its
-encoding in our \hs{Tasks} abstraction without further explanation. We address
-elements of strings $a_i$ and $b_i$ by keys \hs{A}~$i$ and \hs{B}~$i$,
-respectively, while the cost of a subproblem $c_{ij}$ is identified by
-\hs{C}~$i$~$j$.
+so widely known, e.g., see~Cormen~\etal~\shortcite{cormen2001introduction}, that
+we provide its encoding in our \hs{Tasks} abstraction without further
+explanation. We address elements of strings $a_i$ and $b_i$ by keys \hs{A}~$i$
+and \hs{B}~$i$, respectively, while the cost of a subproblem $c_{ij}$ is
+identified by \hs{C}~$i$~$j$.
 
-\vspace{0.5mm}
-\begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
 data Key = A Int | B Int | C Int Int deriving Eq
@@
 \vspace{0mm}
-\begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
+\begin{minted}[xleftmargin=10pt]{haskell}
 editDistance :: Tasks Monad Key Int

### 2019-04-29 src/Build/Task/Free.hs

@@ -2,52 +2,50 @@
 {-# OPTIONS_GHC -Wno-incomplete-uni-patterns #-}
 
 -- | The free description of tasks.
-module Build.Task.Free(
-    Depend(..), toDepend, fromDepend,
-    Depends(..), toDepends, fromDepends
-    ) where
+module Build.Task.Free (
+    Rule (..), toRule, fromRule, Action (..), toAction, fromAction
+  ) where
 
 import Build.Task
+import Control.Monad
 
-
-data Depend k v r = Depend [k] ([v] -> r)
+------------------------- Isomorphism with Make's Rule -------------------------
+data Rule k v r = Rule [k] ([v] -> r)
     deriving Functor
 
-instance Applicative (Depend k v) where
-    pure v = Depend [] (\[] -> v)
-    Depend d1 f1 <*> Depend d2 f2 = Depend (d1++d2) $ \vs -> let (v1,v2) = splitAt (length d1) vs in f1 v1 $ f2 v2
+instance Applicative (Rule k v) where
+    pure v = Rule [] (\[] -> v)
+    Rule d1 f1 <*> Rule d2 f2 = Rule (d1++d2) $ \vs ->
+        let (v1,v2) = splitAt (length d1) vs in f1 v1 $ f2 v2
 
+getRule :: k -> Rule k v v
+getRule k = Rule [k] $ \[v] -> v
 
-data Depends k v r = Depends k (v -> Depends k v r)
-                   | Done r
+toRule :: Task Applicative k v -> Rule k v v
+toRule (Task f) = f getRule
+
+fromRule :: Rule k v v -> Task Applicative k v
+fromRule (Rule ds f) = Task $ \fetch -> f <$> traverse fetch ds
+
+------------------------ Isomorphism with Shake's Action -----------------------
+data Action k v a = Finished a
+                  | Depends k (v -> Action k v a)
     deriving Functor
 
-instance Applicative (Depends k v) where
-    pure = return
-    f1 <*> f2 = f2 >>= \v -> ($ v) <$> f1
+instance Applicative (Action k v) where
+    pure  = Finished
+    (<*>) = ap
 
-instance Monad (Depends k v) where
-    return = Done
-    Done x >>= f = f x
-    Depends ds op >>= f = Depends ds $ \vs -> f =<< op vs
+instance Monad (Action k v) where
+    return = Finished
+    Finished x    >>= f = f x
[…]

### 2019-04-29 papers/jfp/3-abstractions.tex
Fix spacing, add a TODO

@@
 
-cloudAck :: Tasks (MonadState Cache) (Integer, Integer) Integer
-cloudAck (m, n)
+cloudAckermann :: Tasks (MonadState@\,@Cache)@\,@(Integer,@\,@Integer)@\,@Integer
+cloudAckermann (m, n)
     | m < 0 || n < 0 = Nothing

### 2019-04-30 papers/jfp/7-experience.tex
Fix spacing, add a TODO
Add a reference to Jaskelioff and O’Connor

@@
 
-Assuming the lengths of the lists \hs{[@@k]} and \hs{[@@v]} always match, this
-type is isomorphic to \hs{Task}~\hs{Applicative}, and we can define a similar
+Assuming the lengths of the lists \hs{[@@k]} and \hs{[@@v]} always match, the
+data type \hs{Rule}~\hs{k}~\hs{v}~\hs{v} is isomorphic to
+\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v}, and we can define a similar
 \hs{Applicative} instance and conversion functions. By describing these types
@@
 \hs{Applicative}.
+
+These and other isomorphisms for \emph{second-order functionals}, i.e.
+functions of the form
+
+\vspace{1mm}
+\begin{center}
+\begin{minipage}{0.5\textwidth}
+\begin{minted}[xleftmargin=10pt]{haskell}
+@\std{forall}@ f. c f => (k -> f v) -> f a
+\end{minted}
+\end{minipage}
+\end{center}
+\vspace{1mm}
+
+for various choices of~\hs{c}, are studied in depth by~Jaskelioff and
+O’Connor~\shortcite{jaskelioff2015representation}.
 

### 2019-04-30 papers/jfp/7-experience.tex
Work on S7

@@
 
+\Hadrian is a build system for the Glasgow Haskell Compiler~\cite{ghc}. It was
+developed to replace a \Make-based build system to solve multiple scalability
+and maintainability challenges. As discussed in detail
+by~Mokhov~\etal~\shortcite{hadrian}, most of these challenges were consequences
+of two key \Make's shortcomings: (i)~poor abstraction facilities of makefiles,
+i.e. the need to program in a single namespace of mutable string variables, and
+(ii)~the lack of dynamic dependencies~(\S\ref{sec-background-shake}).
+\Hadrian takes advantage both from \Shake's features and from the host language
+Haskell, making the GHC build system more maintainable. Interestingly, although
+\Shake is not a self-tracking build system~(\S\ref{sec-tracking-aspects}),
+\Hadrian implements a small domain-specific language for constructing build
+command lines, and then tracks command lines by treating them as a type of
+values -- a kind of \emph{partial self-tracking} made possible by \Shake's
+support for value polymorphism~(\S\ref{sec-polymorphism}).
+
 \todo{AM}{Finish this section.}
 
-The build system used to compile GHC, known as Hadrian, has continued
+Benchmarking:
+
+\begin{itemize}
+  \item Building \Hadrian: 55s. For comparison, building the Cabal library: 4m.
+
+  \item Initial Windows tarball download: 45s.
+
+  \item Initial \cmd{boot} and \cmd{configure} on Windows: 2m 25s.
+
+  \item Full GHC build using \Hadrian with default settings:
+        \cmd{hadrian/build.bat -j8}: 57m. Subsequent zero build: 5s.
+        Default build includes GHC, 29 bundled libraries (each in vanilla and
+        profiling build way), and 6 bundled executables (such as Haddock).
+
+  \item Full GHC build using \Hadrian with default settings in the cloud mode:
+        \cmd{hadrian/build.bat -j8 -}\cmd{-share=cache}: 58m.
+        Subsequent zero build: 5s.
+        Subsequent full rebuild after deleting the build directory: 5m 27s.
+
+  \item Full GHC build using \Make with default settings: \cmd{make -j8}: 60m.
+        Subsequent zero build: 18s.
+\end{itemize}
+
+The build system used to compile GHC, known as \Hadrian, has continued
 development. The use of dynamic dependencies has made the build much easier, and

### 2019-04-30 papers/jfp/7-experience.tex
Finish S7.2: Hadrian

--- 2019-04-30 papers/jfp/7-experience.tex
@@
 
-\todo{AM}{Finish this section.}
+\Hadrian is a build system for the Glasgow Haskell Compiler~\cite{ghc}. It was
+developed to replace a \Make-based build system and solve multiple scalability
+and maintainability challenges. As discussed in detail
+by~Mokhov~\etal~\shortcite{hadrian}, most of these challenges were consequences
+of two key \Make's shortcomings: (i)~poor abstraction facilities of makefiles,
+i.e. the need to program in a single namespace of mutable string variables, and
+(ii)~the lack of dynamic dependencies~(\S\ref{sec-background-shake}).
+\Hadrian benefits both from \Shake's features and from the host language
+Haskell, making the new GHC build system easier to understand and maintain.
+Interestingly, although \Shake is not a self-tracking build
+system~(\S\ref{sec-tracking-aspects}), \Hadrian implements a little
+domain-specific language for constructing build command lines, and then tracks
+command lines by treating them as a type of values -- an example of
+\emph{partial self-tracking} made possible by \Shake's support for value
+polymorphism~(\S\ref{sec-polymorphism}).
 
-The build system used to compile GHC, known as Hadrian, has continued
-development. The use of dynamic dependencies has made the build much easier, and
-it's now more maintainable. It's now merged into the GHC repo and tested as
-standard by the CI. It's current state is ... In particular, it takes N minutes
-to build with \Shake, vs M minutes to build with \Make. The \Shake version has
-attracted more contributors, and been easier to modify, etc. The code is N
-lines.
+The development of \Cloud \Shake allows GHC developers to take advantage of
+sharing build results between builds, which is particularly promising in the
+context of GHC's continuous integration (CI) infrastructure responsible for
+building and testing every GHC commit. Building GHC~8.8 from scratch with
+default settings takes around 1 hour on Windows using both \Hadrian and the
+original \Make-based build system. This includes building the compiler itself,
+29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
+6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
+only a small part of the whole codebase, making the cloud-build functionality
+highly desirable. In particular, if a commit does not touch any GHC source files
+at all (e.g. only modifies the documentation), \Hadrian can build GHC in under
+3~minutes, by simply creating symbolic links to the previously built results
+stored in the \Cloud~\Shake cache. Note that a small number of GHC build rules
+cannot be cached since they mutate the bundled package database (e.g. to
+register a new library); \Cloud~\Shake provides a way to manually label such
+build rules, thereby excluding them from caching.
 
-The design we originally outlined remains the one in use. There has been a huge
-amount of engineering work to make the tests pass (many of the details encoded
-in the \Make build system were incidental -- but some were essential).
+One of the benefits of using \Shake is that we have access to high quality build
+profiling information, allowing us to compute critical paths and other metrics,
+see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for an overview of \Shake's
+profiling features. This information has shown us that GHC is slow to build
+(1~hour on an 4~CPU Windows machine), that more CPUs would not help (on
+unlimited CPUs the speed up would be less than 10\%), and that a handful of
+build tasks (two anomalously slow Haskell compilations, and some calls to slow
+single-threaded \cmd{configure}) are responsible for a significant amount of
+build time (at least 10~minutes).
 
-One of the benefits of using Shake is that we have access to high quality
-profiling information, allowing us to compute critical paths and other metrics,
-see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for a tour of the metrics.
-This information has shown us that GHC is slow to build (40m on an 8 CPU machine
-on Windows), that more CPUs would not help (on unlimited CPUs it would take at
-least 37m), and that a handful of steps (two Haskell compiles, some calls to
-\cmd{configure}) are responsible for a significant amount of that time (at least
-10m).
+% Benchmarking:
+% \begin{itemize}
+%   \item Building \Hadrian: 55s. For comparison, building the Cabal library: 4m.
+%   \item Initial Windows tarball download: 45s.
+%   \item Initial \cmd{boot} and \cmd{configure} on Windows: 2m 25s.
+%   \item Full GHC build using \Hadrian with default settings:
+%         \cmd{hadrian/build.bat -j8}: 57m. Subsequent zero build: 5s.
+%   \item Full GHC build using \Hadrian with default settings in the cloud mode:
+%         \cmd{hadrian/build.bat -j8 -}\cmd{-share=cache}: 58m.
+%         Subsequent zero build: 5s.
+%         Subsequent full rebuild after deleting the build directory: 5m 27s.
+%   \item Full GHC build using \Make with default settings: \cmd{make -j8}: 60m.
+%         Subsequent zero build: 18s.
+% \end{itemize}
+
+% The build system used to compile GHC, known as \Hadrian, has continued
+% development. The use of dynamic dependencies has made the build much easier, and
+% it's now more maintainable. It's now merged into the GHC repo and tested as
+% standard by the CI. It's current state is ... In particular, it takes N minutes
+% to build with \Shake, vs M minutes to build with \Make. The \Shake version has
+% attracted more contributors, and been easier to modify, etc. The code is N
+% lines.
+
+% The design we originally outlined remains the one in use. There has been a huge
+% amount of engineering work to make the tests pass (many of the details encoded
+% in the \Make build system were incidental -- but some were essential).
 

### 2019-04-30 papers/jfp/5-rebuilders.tex
Fix two typos, drop repeated text

@@
 structure, called \emph{verifying step traces},
-which stores less data than veryifying traces, and has slightly different early cutoff semantics. Rather than
+which stores less data than verifying traces, and has slightly different early cutoff semantics. Rather than
 storing the \hs{Hash}~\hs{v} for each dependency, it instead stores \hs{built}
@@
 $n = 1$. By picking values of $n$ in between we would regain some early cutoff, at the
-cost of losing such simple shallow builds and still requiring determinstic tasks.
+cost of losing such simple shallow builds and still requiring deterministic tasks.

### 2019-05-01 papers/jfp/2-background.tex
Work on S7
Drafting changes in sections 1-3

@@
 
-Build systems automate the execution of repeatable tasks for individual users
-and large organisations. In this section we explore the design space of build
+Build systems automate the execution of repeatable tasks, at a scale from individual users
+up to large organisations. In this section we explore the design space of build
 systems, using four concrete examples: \Make~\cite{feldman1979make},
@@
 more than 40 years ago to automatically build software libraries and executable
-programs from source code. It uses \emph{makefiles} to describe tasks, often
-referred to as \emph{build rules}, and their dependencies in a simple textual
+programs from source code. It uses \emph{makefiles} to describe \emph{tasks} --- often
+referred to as \emph{build rules} --- and their \emph{dependencies}, in a simple textual
 form. For example:
@@
 executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
-the sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and
-the actual binary used to compile and link files, as well as standard libraries
-such that \cmd{stdio.h}, are often also listed as dependencies.} the command
+  the sake of simplicity. In reality, there are multiple versions of \cmd{gcc}.
+  To account for this, the actual binary for \cmd{gcc} is often listed as a dependency, along
+  with any supporting binaries, or standard libraries (such as \cmd{stdio.h}),
+  that are used by \cmd{gcc}.
+} the command
 \cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
@@
 If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
-will build~\cmd{util.o} and~\cmd{main.o}, in any order since these tasks are
+will build~\cmd{util.o} and~\cmd{main.o}, in any order (or even in parallel) since these tasks are
 independent, and then \cmd{main.exe}. If the user modifies \cmd{util.h} and runs
@@
 graph is \emph{acyclic} then each task needs to be executed at most once. Cyclic
-task dependencies are typically not allowed in build systems but there are rare
+task dependencies are typically not allowed in build systems, although there are rare
 exceptions, see~\S\ref{sec-iterative-compute}.
@@
 \textbf{Definition} (Minimality). A build system is \emph{minimal} if it
-executes tasks at most once per build and only if they transitively depend on
+executes tasks at most once per build, and only if they transitively depend on
 inputs that changed since the previous build.\label{def-minimal}
@@
 to compute the dependency graph before the build starts. In this particular
-example one might say that the value of \cmd{C1} is statically known, but
-imagine that it is the result of a long computation chain -- its value will only
+example the value of \cmd{C1} is a constant, but
+it might instead be the result of a long computation chain --
+so its value will only
 become available during the build.
@@
 before the build. Makefiles cannot express such dependencies, requiring problematic
-workarounds such as \emph{build phases} \cite{hadrian}.
+workarounds such as \emph{build phases}~\cite{hadrian}.
 In \Shake we can express the rule for \cmd{release.tar} as:
@@
 \Shake's implementation is different from both \Make and \Excel in two aspects.
-First, it uses the dependency graph from the previous build to decide which
-files need to be rebuilt. This idea has a long history, going back to
+First, to decide which files need to be rebuilt,
+it uses the dependency graph that is dynamically constructed during
+the previous build . This idea has a long history, going back to
 \emph{incremental}~\cite{demers1981incremental},
@@
 
-Consider an example in Fig.~\ref{fig-bazel}. The user starts by downloading the
+Consider the example in Fig.~\ref{fig-bazel}. The user starts by downloading the
 sources, whose content hashes are (for simplicity)~1,~2 and~3, and requests to
@@
 To support cloud builds, \Bazel maintains (i) a \emph{content-addressable cache}
-that can be used to download a previously built file given the hash of its
-content, and (ii) the history of all executed build commands annotated with
-observed file hashes. The latter allows the build engine to bypass the execution
+that maps the hash of a file's content to the actual content of that file;
+(ii) a memo table that records all executed build commands
+with their input file hashes and output file value.
+The memo table allows the build engine to bypass the execution
 of a task, by predicting the hash of the result from the hashes of its
-dependencies, and subsequently download the result from the cache. A concrete
+dependencies; then the content-addressable cache allows the engine
+to download the result (if needed) based on the result hash. A concrete
 implementation is provided in~\S\ref{sec-implementation-cloud}.

### 2019-05-01 papers/jfp/3-abstractions.tex
Drafting changes in sections 1-3

@@
 
-This section presents abstractions that allow us to express
+We have introduced a number of components and
+characteristics of build systems:
+tasks, dependencies, early cutoff, minimality, etc.
+It is easy to get confused.
+To make all this more concrete, this section presents executable
+abstractions that can express
 all the intricacies of build systems discussed in~\S\ref{sec-background},
-and express complex build systems from simple primitives. Specifically, we
+and allow us to construct complex build systems from simple primitives. Specifically, we
 present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
@@
 \emph{Build system.} A \emph{build system} takes a task description, a
-\emph{target} key, and a store, and returns a new store in which the target key
+\emph{target key}, and a store, and returns a new store in which the target key
 and all its dependencies have up to date values.
 
-We model build systems concretely, as Haskell programs. To that end,
+We model a build system concretely, as a Haskell program. To that end,
 Fig.~\ref{fig-types} provides the type signatures for all key abstractions
@@
 A \hs{Task} describes a single build task, while \hs{Tasks} associates a
-\hs{Task} to every non-input key; input keys are associated with \hs{Nothing}.
+\hs{Task} with every non-input key; input keys are associated with \hs{Nothing}.
 The highly-abstracted type \hs{Task} describes how to build a value given
@@
 given in Fig.~\ref{fig-stdlib}. We will explain why in~\S\ref{sec-general-build}.
-
 For now, we content ourselves with observing that a task description, of type
@@
 compilers, calc chains, file systems, caches, and all other complexities of real
-build systems. It just computes a single output, in a side-effect-free way,
-using a callback (\hs{fetch}) to find the values of its dependencies.
+build systems. It just computes a single output,
+using a callback (\hs{fetch}) to find the values of its dependencies,
+and limiting side effects to those described by \hs{c}.
 
@@
 As we can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
-right order (if it had built \cmd{B2} before building \cmd{B1}, the result would
+correct order (if it had built \cmd{B2} before building \cmd{B1}, the result would
 have been $20 * 2 = 40$ instead of $(10 + 20) * 2 = 60$). As an example showing
@@
 \hs{fetch}~\hs{"B1"} twice -- once per occurrence of \cmd{B1} in the
-formula.
+formula -- and each call would recompute the formula in \cmd{B1}.
 
@@
 
-But nor can the task description work for \emph{any} \hs{f}; most task
+But the task description cannot possibly work for \emph{any} \hs{f} whatsoever; most task
 descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
@@
 expressing that \hs{f} can only be instantiated by types that satisfy the
-constraint \hs{c}. So the type \hs{Task} emerges naturally, almost inevitably.
+constraint \hs{c} and, in exchange, the task has access to the operations of class \hs{c}.
+So the type \hs{Task} emerges naturally, almost inevitably.
 But now that it \emph{has} emerged, we find that constraints \hs{c} classify
-task descriptions in a very interesting way:
+task descriptions in a very interesting, and practically useful, way:
 \begin{itemize}
@@
 computations. In our framework, such task descriptions correspond to the type
-\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{v}. Consider this spreadsheet example:
+\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Consider this spreadsheet example:
 
@@
 we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
+Here is our first use of polymorpism in \hs{f}, discussed in \S\ref{sec-why-polymorphism}.
 The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
@@
 of \Excel) looking at the syntax tree of the formula.  But a task description
-has no such syntax tree. Yet, remarkably, we can use the polymorphism of a
+has no such syntax tree: as you can see in the definition of \hs{Task} in Fig.~\ref{fig-types},
+a task is just a function, so all we can do is call it.
 \hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of

### 2019-05-01 papers/jfp/1-intro.tex
Drafting changes in sections 1-3
Changes from SPJ

@@
   support for early cutoff; self-tracking build systems; and the type of
-  persistent build information. In~\S\ref{sec-background} we identify some key
+  persistent build information. In~\S\ref{sec-background} we identify some of these key
   properties, illustrated by four carefully-chosen build systems.
@@
 with desired properties, for example, to combine the advantages of \Shake and
-\Bazel, and evaluate the result~(\S\ref{sec-experience}).
+\Bazel. Writing this paper gave us the insights to combine dynamic dependencies and
+cloud build in a principled way; we evaluate the result~(\S\ref{sec-experience}).
 

### 2019-05-02 papers/jfp/2-background.tex
Changes from SPJ
Minor edits in S2 and S3

@@
 it uses the dependency graph that is dynamically constructed during
-the previous build . This idea has a long history, going back to
+the previous build. This idea has a long history, going back to
 \emph{incremental}~\cite{demers1981incremental},
@@
 early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
-not.
+not (see \S\ref{sec-verifying-traces} for an explanation of why).
 
@@
 
-To support cloud builds, \Bazel maintains (i) a \emph{content-addressable cache}
+To support cloud builds, \Bazel maintains (i)~a \emph{content-addressable cache}
 that maps the hash of a file's content to the actual content of that file;
-(ii) a memo table that records all executed build commands
-with their input file hashes and output file value.
+(ii)~a memo table that records all executed build commands
+with their input and output file hashes.
 The memo table allows the build engine to bypass the execution

### 2019-05-02 papers/jfp/3-abstractions.tex
Minor edits in S2 and S3

@@
 \end{minted}
+\vspace{1mm}
+\begin{minted}[fontsize=\small]{haskell}
+-- Standard types from Control.Monad.Trans.Writer
+newtype WriterT w m a = WriterT { runWriterT :: m (a, w) }
+tell :: Monad m => w   -> WriterT w m () -- write a value to the log
+lift :: Monad m => m a -> WriterT w m a  -- lift an action into WriterT
+\end{minted}
+
 % \vspace{-3mm}
@@
 a task is just a function, so all we can do is call it.
+Yet, remarkably, we can use the polymorphism of a
 \hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
@@
 the list of its dependencies (key/value pairs) in an arbitrary monadic
-context~\hs{m}:
+context~\hs{m}. We need this function to be polymorphic over~\hs{m}, because
+each build system will execute tasks in its own monad, as we shall see
+in~\S\ref{sec-implementations}.
 
@@
 information~--~a list of key/value pairs \hs{[(@@k,}~\hs{v)]}~--~when executing
-a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
+a task in an arbitrary monad~\hs{m} (see the types in Fig.~\ref{fig-stdlib}).
+We substitute the given \hs{fetch} with a
 \hs{trackingFetch} that, in addition to fetching a value, tracks the
@@
 with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
-example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
+example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}, by defining
+a corresponding \hs{fetchIO} of type \hs{String}~\hs{->}~\hs{IO}~\hs{Integer},
+which allows us to demonstrate the dynamic nature of monadic dependencies in
+GHCi.
 

### 2019-05-02 papers/jfp/7-experience.tex
Fix two typos, drop repeated text
Add examples to S7.4

@@
 
-We have also got the GHC build working with \Cloud \Shake, resulting in build
-times of 3m and 30s instead of 1h and 8m (on Windows). Due to aspects like slow
-and fully sequential \cmd{configure} we do not see the dramatic reductions
-reported by Standard Chartered. Converting to a build suitable for sharing is
+% We have also got the GHC build working with \Cloud \Shake, resulting in build
+% times of 3m and 30s instead of 1h and 8m (on Windows). Due to aspects like slow
+% and fully sequential \cmd{configure} we do not see the dramatic reductions
+% reported by Standard Chartered.
+
+Converting to a build suitable for sharing is
 not overly onerous, but nor is it trivial. Some things that require attention:
@@
 \item \textbf{Irrelevant differences}: A common problem is that you do not get
-      shared caching when you want it. To get caching, it is important that all
-      machines are identical for the aspects that involve the build system. In
-      particular, that means that all paths must be relative, and all
-      dependencies on system binaries (e.g. \cmd{gcc}) should either be
-      controlled by the build system (e.g. using \Nix) or have tolerance to
-      account for minor machine variations.
+      shared caching when you want it. As one example, imagine two users install
+      \cmd{gcc} on different paths (say \cmd{/usr/bin} and \cmd{/usr/local/bin})
+      -- if these paths are recorded by the build system the users won't share
+      cache entries. As another example, consider a compiler that embeds the current time
+      in the output -- any users who build that file locally won't get any shared
+      caching of subsequent outputs. Possible solutions include using relative paths,
+      depending only on version numbers for system binaries (e.g. \cmd{gcc}),
+      controlling the environment closely (e.g. using \Nix) and extra flags to
+      encourage compilers to be more deterministic.
 \item \textbf{Insufficient produced files}: A build rule must declare all files
-      it produces, so these are stored in the cache and can be restored from the
-      cache. If you do not use a remote cache then it is permissible to have
-      files not known to the build system which change at the same time as files
-      that are known. Once using a remote cache any file that is produced as a
-      side-effect of a rule and then later used \emph{must} be declared. Failure
-      to declare such files means that if some step produces a file, and a later
-      step relies on it, that step breaks. In practice, most issues encountered
-      during the move to cloud builds for GHC were a lack of declared produced
+      it produces, so these can be included in the cache.
+      As an example using Haskell, compilation of \cmd{Foo.hs} produces
+      \cmd{Foo.hi} and \cmd{Foo.o}. If you declare the rule as producing \cmd{Foo.hi},
+      and only ever depend on \cmd{Foo.hi}, but \emph{use} \cmd{Foo.o} after
+      depending on \cmd{Foo.hi}, a local build will probably work. However,
+      if \cmd{Foo.hi} is downloaded from a remote cache, \cmd{Foo.o} will not be present,
+      and subsequent commands may fail (e.g. linking). In practice, most issues encountered
+      during the move to cloud builds for GHC were caused by failing to declare produced
       files.
 \item \textbf{Insufficient dependencies}: While insufficient dependencies are
-      often a problem, the move to a cloud build makes them even more critical,
-      as missing dependencies are likely to mean a result is \emph{never built
-      locally}, rather than just build locally only once. Furthermore,
-      dependencies in a cloud setting must be more complete. For example, using
-      the build system from Fig.~\ref{fig-make}, with a local build system
-      without early cutoff, if the rule to produce \cmd{main.exe} also made use
-      of \cmd{main.c} but did not declare so, that is not a problem. With the
-      addition of early cutoff, if \cmd{main.c} changes in an important way but
-      \cmd{main.o} is not rebuilt that causes issues, but is a somewhat rare
-      occurrence. With cloud builds that becomes somewhat more common.
+      always a problem, the move to a cloud build makes them even more serious.
+      If you omit a dependency on a file that changes rarely, each developer will
+      have built it locally at least once, and more regularly if they ever do clean
+      builds. In contrast, with a cloud build, the file might only be built
+      \emph{just once}, and reused forevermore. If things go wrong, a clean build
+      is no longer a refuge of last resort.
 \end{itemize}

### 2019-05-02 papers/jfp/7-experience.tex
Word smithing in Section 7

@@
 
-As a set of authors, we've written a paper about the \Shake build
-system~\cite{mitchell2012shake}, applying \Shake to building GHC~\cite{hadrian},
-and a paper about the categorisation of build
-systems~\cite{mokhov2018buildsystems}. The original motivation of the research
-in that final paper was to create ``\Cloud \Shake'', and apply it to the GHC
-build system. In this section we tie these papers together, reflecting on our
+We have described a framework that can describe, and indeed execute in
+prototype form, a wide spectrum of build systems.  But our ultimate
+goal is a practical one: to use these insights to construct a better
+build system.  Our earlier work on the \Shake build
+system~\cite{mitchell2012shake}, and applying \Shake to
+building GHC~\cite{hadrian}, makes progress in that direction.
+
+Based on the thinking in this paper we have extended \Shake to
+become \Cloud \Shake, the first cloud-capable build system to support
+both early cutoff and monadic dependencies (\S{sec-implementation-cloud}),
+and deployed it to GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
+In this section we tie these papers together, reflecting on our
 experience since these papers.
@@
 documentation including a website\footnote{\url{https://shakebuild.com}}, and
-add a lot of tests. The biggest change in that time period was moving from
-blocking threads to continuations for the suspending scheduler. The most visible
-change is switching from \hs{*>} to \hs{@\%@>} for defining rules\footnote{A
-conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost
-all external and internal details remain the same.
-
+add a lot of tests. The biggest change in that time period was an implementation
+change: moving from
+blocking threads to continuations for the suspending scheduler.
+But otherwise, almost all external and internal details remain the same\footnote{
+  The most visible change is purely notational: switching from \hs{*>} to \hs{@\%@>}
+  for defining rules, because a conflicting \hs{*>} operator was added to the
+  Haskell \hs{Prelude}.}.
 We consider the lack of change suggestive that \Shake is based on fundamental
 principles -- principles we can now name and describe as a consequence
-of the categorisation paper~\cite{mokhov2018buildsystems}. There are two main
+of this paper.
+
+There are two main
 aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
@@
 verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
-to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. The second
-aspect is the power of \Shake tasks, or ``actions'' in the original paper,
-which were originally described using this data type:
-
+to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. Secondly,
+in the original paper the tasks (there called ``actions'')
+were described in continuation-passing style using this data type\footnote{
+The original paper uses concrete types \hs{Key} and \hs{Value}. Here we
+generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
+\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}}.
 \vspace{1mm}
@@
 \vspace{1mm}
-
 \noindent
-This type is equivalent to \hs{Task}~\hs{Monad} -- the two types form an
-isomorphism. The original paper uses concrete types \hs{Key} and \hs{Value}, we
-generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
-\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}. We define that
-\hs{Monad} instance, and use it to write conversion functions to/from our
-\hs{Task} type, witnessing the isomorphism:
-
+In this paper we describe tasks more directly, in a monadic (or applicative or functorial)
+style. But in fact the two are equivalent: \hs{Task Monad k v} is
+isomporphic to \hs{Action k v v}.
+To be concrete, the functions \hs{toAction} and \hs{fromAction} witness the
+isomorphism in both directions.
 \vspace{1mm}
@@
 \noindent
-Similarly, \Make tasks were described as equivalent to
+Similarly, in the original paper \Make tasks were described as equivalent to
 
@@
 \vspace{1mm}
-
+\noindent
 Assuming the lengths of the lists \hs{[@@k]} and \hs{[@@v]} always match, the
@@
 \hs{Task}~\hs{Applicative}~\hs{k}~\hs{v}, and we can define a similar
-\hs{Applicative} instance and conversion functions. By describing these types
+\hs{Applicative} instance and conversion functions.
+
+By describing these types
 using \hs{Task} we are able to describe the differences more concisely
@@
 \hs{Applicative}.
-
 These and other isomorphisms for \emph{second-order functionals}, i.e.
@@
 
-\subsection{Experience from building GHC with \Shake}
-
-\Hadrian is a build system for the Glasgow Haskell Compiler~\cite{ghc}. It was
-developed to replace a \Make-based build system and solve multiple scalability
-and maintainability challenges. As discussed in detail
-by~Mokhov~\etal~\shortcite{hadrian}, most of these challenges were consequences
-of two key \Make's shortcomings: (i)~poor abstraction facilities of makefiles,
-i.e. the need to program in a single namespace of mutable string variables, and
-(ii)~the lack of dynamic dependencies~(\S\ref{sec-background-shake}).
-\Hadrian benefits both from \Shake's features and from the host language
-Haskell, making the new GHC build system easier to understand and maintain.
-Interestingly, although \Shake is not a self-tracking build
-system~(\S\ref{sec-tracking-aspects}), \Hadrian implements a little
-domain-specific language for constructing build command lines, and then tracks
-command lines by treating them as a type of values -- an example of
-\emph{partial self-tracking} made possible by \Shake's support for value
-polymorphism~(\S\ref{sec-polymorphism}).
-
-The development of \Cloud \Shake allows GHC developers to take advantage of
-sharing build results between builds, which is particularly promising in the
-context of GHC's continuous integration (CI) infrastructure responsible for
-building and testing every GHC commit. Building GHC~8.8 from scratch with
-default settings takes around 1 hour on Windows using both \Hadrian and the
-original \Make-based build system. This includes building the compiler itself,
-29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
-6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
-only a small part of the whole codebase, making the cloud-build functionality
-highly desirable. In particular, if a commit does not touch any GHC source files
-at all (e.g. only modifies the documentation), \Hadrian can build GHC in under
-3~minutes, by simply creating symbolic links to the previously built results
-stored in the \Cloud~\Shake cache. Note that a small number of GHC build rules
-cannot be cached since they mutate the bundled package database (e.g. to
-register a new library); \Cloud~\Shake provides a way to manually label such
-build rules, thereby excluding them from caching.
-
-One of the benefits of using \Shake is that we have access to high quality build
-profiling information, allowing us to compute critical paths and other metrics,
-see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for an overview of \Shake's
-profiling features. This information has shown us that GHC is slow to build
-(1~hour on an 4~CPU Windows machine), that more CPUs would not help (on
-unlimited CPUs the speed up would be less than 10\%), and that a handful of
-build tasks (two anomalously slow Haskell compilations, and some calls to slow
-single-threaded \cmd{configure}) are responsible for a significant amount of
-build time (at least 10~minutes).
-
-% Benchmarking:
-% \begin{itemize}
-%   \item Building \Hadrian: 55s. For comparison, building the Cabal library: 4m.
-%   \item Initial Windows tarball download: 45s.
-%   \item Initial \cmd{boot} and \cmd{configure} on Windows: 2m 25s.
-%   \item Full GHC build using \Hadrian with default settings:
-%         \cmd{hadrian/build.bat -j8}: 57m. Subsequent zero build: 5s.
-%   \item Full GHC build using \Hadrian with default settings in the cloud mode:
-%         \cmd{hadrian/build.bat -j8 -}\cmd{-share=cache}: 58m.
-%         Subsequent zero build: 5s.
-%         Subsequent full rebuild after deleting the build directory: 5m 27s.
-%   \item Full GHC build using \Make with default settings: \cmd{make -j8}: 60m.
-%         Subsequent zero build: 18s.
-% \end{itemize}
-
-% The build system used to compile GHC, known as \Hadrian, has continued
-% development. The use of dynamic dependencies has made the build much easier, and
-% it's now more maintainable. It's now merged into the GHC repo and tested as
-% standard by the CI. It's current state is ... In particular, it takes N minutes
-% to build with \Shake, vs M minutes to build with \Make. The \Shake version has
-% attracted more contributors, and been easier to modify, etc. The code is N
-% lines.
-
-% The design we originally outlined remains the one in use. There has been a huge
-% amount of engineering work to make the tests pass (many of the details encoded
-% in the \Make build system were incidental -- but some were essential).
-
 \subsection{Experience from \Cloud \Shake}\label{sec-cloud-shake}
@@
 Converting \Shake into \Cloud \Shake was not a difficult process once armed with
-a roadmap~\cite{mokhov2018buildsystems}. The key was the introduction of two new
+the roadmap in this paper. The key was the introduction of two new
 functions:
@@
 from~\S\ref{sec-constructive-traces}, with their differences perhaps the most
-illustrative of the changes required. (We have made some minor changes from
+illustrative of the changes required\footnote{
+  We have made some minor changes from
 actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant
-differences.)
+differences.}.
 
 \begin{itemize}
-\item There are two \hs{Ver} arguments being passed to each function. These are
+\item Two \hs{Ver} arguments are passed to each function. These are
       the versions of the build script, and the rule for this particular key. If
@@
       a simple list. The reason is that \Shake allows a list of dependencies to
-      be supplied at once, so they can all be built in parallel.
+      be specfied simultaneously, so they can all be built in parallel.
 \item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
@@
       shared caching when you want it. As one example, imagine two users install
-      \cmd{gcc} on different paths (say \cmd{/usr/bin} and \cmd{/usr/local/bin})
-      -- if these paths are recorded by the build system the users won't share
+      \cmd{gcc} on different paths (say \cmd{/usr/bin/gcc} and \cmd{/usr/local/bin/gcc}).
+      If these paths are recorded by the build system, the users won't share
       cache entries. As another example, consider a compiler that embeds the current time
-      in the output -- any users who build that file locally won't get any shared
-      caching of subsequent outputs. Possible solutions include using relative paths,
-      depending only on version numbers for system binaries (e.g. \cmd{gcc}),
-      controlling the environment closely (e.g. using \Nix) and extra flags to
+      in the output: any users who build that file locally won't get any shared
+      caching of subsequent outputs. Possible solutions include using relative paths;
+      depending only on version numbers for system binaries (e.g. \cmd{gcc});
+      controlling the environment closely (e.g. using \Nix); and extra flags to
       encourage compilers to be more deterministic.
@@
       \cmd{Foo.hi} and \cmd{Foo.o}. If you declare the rule as producing \cmd{Foo.hi},
-      and only ever depend on \cmd{Foo.hi}, but \emph{use} \cmd{Foo.o} after
+      and other rules \emph{depend on} \cmd{Foo.hi}, but \emph{also use} \cmd{Foo.o} after
       depending on \cmd{Foo.hi}, a local build will probably work. However,
@@
       files.
-\item \textbf{Insufficient dependencies}: While insufficient dependencies are
+\item \textbf{Missing dependencies}: While missing dependencies are
       always a problem, the move to a cloud build makes them even more serious.
-      If you omit a dependency on a file that changes rarely, each developer will
-      have built it locally at least once, and more regularly if they ever do clean
+      Suppose you omit a dependency on a file \cmd{foo} that is produced by a rule,
+      but that changes rarely (e.g. perhaps \cmd{foo} is derived from
+      a configuration input file).
+      With purely local builds, each developer will build \cmd{foo} at least once (when
+      starting from scratch), and more regularly if they ever do clean
       builds. In contrast, with a cloud build, the file might only be built
@@
 build cloud ready.
+
+\subsection{Experience from building GHC with \Shake}
+
+\Hadrian is a build system for the Glasgow Haskell Compiler~\cite{ghc}. It was
+developed to replace a \Make-based build system and solve multiple scalability
+and maintainability challenges. As discussed in detail
+by~Mokhov~\etal~\shortcite{hadrian}, most of these challenges were consequences
+of two key shortcomings of \Make: (i)~poor abstraction facilities of makefiles,
+notably the need to program in a single namespace of mutable string variables, and
+(ii)~the lack of dynamic dependencies~(\S\ref{sec-background-shake}).
+\Hadrian benefits both from \Shake's features and from the host language
+Haskell, making the new GHC build system easier to understand and maintain.
+
+Interestingly, although \Shake is not a self-tracking build
+system~(\S\ref{sec-tracking-aspects}), \Hadrian implements a little
+domain-specific language for constructing build command lines, and then tracks
+command lines by treating them as a type of values -- an example of
+\emph{partial self-tracking} made possible by \Shake's support for value
+polymorphism~(\S\ref{sec-polymorphism}).
+
+The subsequent development of \Cloud \Shake allows GHC developers to take advantage of
+sharing build results between builds, which is particularly promising in the
+context of GHC's continuous integration (CI) infrastructure responsible for
+building and testing every GHC commit. Building GHC~8.8 from scratch with
+default settings takes around 1 hour on Windows using both \Hadrian and the
+original \Make-based build system. This includes building the compiler itself,
+29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
+6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
+only a small part of the whole codebase, making the cloud-build functionality
+highly desirable. For example, if a commit does not touch any GHC source files
+(e.g. only modifies the documentation), \Hadrian can build GHC from scratch in under
+3~minutes, by simply creating symbolic links to the previously build results
+stored in the \Cloud~\Shake cache.
+
+A small number of GHC build rules
+cannot be cached since they mutate the bundled package database (e.g. to
+register a new library); \Cloud~\Shake provides a way to manually label such
+build rules, thereby excluding them from caching.
+
+One of the benefits of using \Shake is that we have access to high quality build
+profiling information, allowing us to compute critical paths and other metrics;
+see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for an overview of \Shake's
+profiling features. This information has shown us, for example, that more CPUs would not help (on
+unlimited CPUs the speed up would be less than 10\%), and that a handful of
+build tasks (two anomalously slow Haskell compilations, and some calls to slow
+single-threaded \cmd{configure}) are responsible for a significant fraction of
+build time (at least 15\%).
+
+% Benchmarking:
+% \begin{itemize}
+%   \item Building \Hadrian: 55s. For comparison, building the Cabal library: 4m.
+%   \item Initial Windows tarball download: 45s.
+%   \item Initial \cmd{boot} and \cmd{configure} on Windows: 2m 25s.
+%   \item Full GHC build using \Hadrian with default settings:
+%         \cmd{hadrian/build.bat -j8}: 57m. Subsequent zero build: 5s.
+%   \item Full GHC build using \Hadrian with default settings in the cloud mode:
+%         \cmd{hadrian/build.bat -j8 -}\cmd{-share=cache}: 58m.
+%         Subsequent zero build: 5s.
+%         Subsequent full rebuild after deleting the build directory: 5m 27s.
+%   \item Full GHC build using \Make with default settings: \cmd{make -j8}: 60m.
+%         Subsequent zero build: 18s.
+% \end{itemize}
+
+% The build system used to compile GHC, known as \Hadrian, has continued
+% development. The use of dynamic dependencies has made the build much easier, and
+% it's now more maintainable. It's now merged into the GHC repo and tested as
+% standard by the CI. It's current state is ... In particular, it takes N minutes
+% to build with \Shake, vs M minutes to build with \Make. The \Shake version has
+% attracted more contributors, and been easier to modify, etc. The code is N
+% lines.
+
+% The design we originally outlined remains the one in use. There has been a huge
+% amount of engineering work to make the tests pass (many of the details encoded
+% in the \Make build system were incidental -- but some were essential).

### 2019-05-02 papers/jfp/7-experience.tex
Redrafting in section 8

@@
 become \Cloud \Shake, the first cloud-capable build system to support
-both early cutoff and monadic dependencies (\S{sec-implementation-cloud}),
-and deployed it to GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
+both early cutoff and monadic dependencies (\S\ref{sec-implementation-cloud}),
+and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
 In this section we tie these papers together, reflecting on our
-experience since these papers.
+experience of turning theory into practice.
 

### 2019-05-02 papers/jfp/8-engineering.tex
Redrafting in section 8

@@
       compiled using parallelism can change the order in which unique variables
-      are obtained from the supply, producing different but semantically
+      are obtained from the name supply, producing different but semantically
       identical results.
@@
       main difference from non-deterministic tasks is that volatile tasks cannot
-      be cached. They are best modelled as depending on a special key
+      be cached. They are readily modelled as depending on a special key
       \cmd{RealWorld} whose value is changed in every build.
@@
 dependencies can be marked as volatile, a technique \Excel takes with the
-\cmd{INDIRECT} function, removing the untracked dependency at the cost of
+\cmd{INDIRECT} function, fixing the untracked dependency at the cost of
 minimality.
@@
 \noindent
-The build systems that have precise self-tracking are all ones which use a
-\emph{non-embedded domain specific language} to describe build tasks. Those
-which use a full programming language, e.g. \Shake, are faced with the challenge
-of implementing equality on arbitrary task functions. For such build systems,
-the pessimistic assumption that any change to the build system potentially
-changes any build task can often be used~--~the classic example being a makefile
-depending on itself.
+The build systems that have precise self-tracking are all ones which
+use a \emph{non-embedded domain specific language} to describe build
+tasks; that is, a task is described by a data structure (or syntax
+tree), and tasks can be compared for equality by comparing those data
+structures. Those which use a full programming language, e.g. \Shake,
+are faced with the challenge of implementing equality on arbitrary
+tasks -- and a task is just a function.  For such build systems, the
+only safe approach is to assume (pessimistically) that any
+change to the build system potentially changes any build task ~--~ the
+classic example being a makefile depending on itself.
 

### 2019-05-02 papers/jfp/7-experience.tex
Improve things that were confusing to SPJ in S7

@@
 
-for various choices of~\hs{c}, are studied in depth by~Jaskelioff and
+\noindent for various choices of~\hs{c}, are studied in depth by~Jaskelioff and
 O’Connor~\shortcite{jaskelioff2015representation}.
@@
 
-Converting to a build suitable for sharing is
-not overly onerous, but nor is it trivial. Some things that require attention:
+Converting to a build suitable for sharing is not overly onerous, but nor is it trivial.
+In particular, a cloud build is less forgiving about untracked operations -- things that
+are wrong but usually harmless in local builds often cause serious problems in a cloud setting.
+Some things that require attention in moving to a cloud build:
 
@@
       and other rules \emph{depend on} \cmd{Foo.hi}, but \emph{also use} \cmd{Foo.o} after
-      depending on \cmd{Foo.hi}, a local build will probably work. However,
+      depending on \cmd{Foo.hi}, a local build will probably work (although treating \cmd{Foo.o}
+      as a proper dependency would definitely be preferable). However,
       if \cmd{Foo.hi} is downloaded from a remote cache, \cmd{Foo.o} will not be present,
-      and subsequent commands may fail (e.g. linking). In practice, most issues encountered
+      and subsequent commands may fail (e.g. linking).
+      In practice, most issues encountered
       during the move to cloud builds for GHC were caused by failing to declare produced
@@
 \item \textbf{Missing dependencies}: While missing dependencies are
-      always a problem, the move to a cloud build makes them even more serious.
-      Suppose you omit a dependency on a file \cmd{foo} that is produced by a rule,
-      but that changes rarely (e.g. perhaps \cmd{foo} is derived from
-      a configuration input file).
-      With purely local builds, each developer will build \cmd{foo} at least once (when
-      starting from scratch), and more regularly if they ever do clean
-      builds. In contrast, with a cloud build, the file might only be built
-      \emph{just once}, and reused forevermore. If things go wrong, a clean build
-      is no longer a refuge of last resort.
+      always a problem, the move to a cloud build makes them more serious.
+      With local builds outputs will be built at least once per user, but with a
+      cloud build they might only be built \emph{once ever}.
 \end{itemize}
@@
 
-A small number of GHC build rules
-cannot be cached since they mutate the bundled package database (e.g. to
-register a new library); \Cloud~\Shake provides a way to manually label such
+A small number of GHC build rules cannot be cached. These rules register new libraries
+in the GHC package database, and rather than producing a output file, they mutate
+a shared file. \Cloud~\Shake provides a way to manually label such
 build rules, thereby excluding them from caching.

### 2019-05-02 papers/jfp/8-engineering.tex
Address Simon's feedback on S8.1

@@
 \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{Either}~\hs{e}~\hs{v} into \hs{EitherT}
-and then runs the task in the \hs{EitherT} monad transformer~\cite{liang1995monad}.
+and then runs the task in the \hs{EitherT} monad
+transformer~\cite{liang1995monad}.
 
-Another approach is to include failures into the computation context \hs{f}. For
-example, we can choose \hs{f}~\hs{=}~\hs{Maybe}:
+Another approach is to include failures into the computation context \hs{f}.
+Recall from~\S\ref{sec-why-polymorphism} that we require tasks to be polymorphic
+in~\hs{f}, and we can therefore choose to execute a \hs{Task} in an~\hs{f} with
+failures, the simplest example being \hs{f}~\hs{=}~\hs{Maybe}. Below we define
+a callback \hs{fetchA1A2} that returns \hs{Just}~\hs{value} for keys \cmd{A1}
+and \cmd{A2} but fails with \hs{Nothing} on all other keys:
 
@@
 \noindent
-Since tasks are polymorphic in \hs{f}, we can directly run them with the
-callback \hs{fetchA1A2}. For example, the task \cmd{B1 = A1 + A2} from
-\hs{sprsh1} in~\S\ref{sec-task} returns \hs{Just}~\hs{30}, whereas the task
-\cmd{B2 = B1 * 2} returns \hs{Nothing} because \hs{fetchA1A2} fails on \cmd{B1}.
+We can directly run any \hs{Task} with this callback. For example, the task
+\cmd{B1 = A1 + A2} from \hs{sprsh1} in~\S\ref{sec-task} returns
+\hs{Just}~\hs{30}, whereas the task \cmd{B2 = B1 * 2} returns \hs{Nothing}
+because \hs{fetchA1A2} fails on \cmd{B1}.
 
 This approach is convenient if \emph{tasks are not aware of failures}, e.g. we
-can model Excel formulas as pure arithmetic functions, and introduce failures
+can model \Excel formulas as pure arithmetic functions, and introduce failures
 ``for free'' if/when needed by instantiating \hs{Tasks} with an appropriate
-\hs{f}.
+\hs{f}. In a real system this \hs{f} would be more complex than just \hs{Maybe},
+for example \hs{MaybeT}~\hs{(State}~\hs{Spreadsheet)}, thus allowing us to
+combine failures with access to \Excel's spreadsheet state by using the
+\hs{MaybeT} monad transformer~\cite{liang1995monad}.
 
@@
 \hs{Either}~\hs{String}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)} instead of just
-\hs{Store}~\hs{i}~\hs{k}~\hs{v} as in~\S\ref{sec-general-build}. One possible
-implementation is based on introducing an extra \hs{Either}~\hs{String} layer
-into the monad stack; while fairly direct it is also tedious, and hence we omit
-it.
+\hs{Store}~\hs{i}~\hs{k}~\hs{v} as in our \hs{Build}
+abstraction~\S\ref{sec-general-build}. One possible implementation of such a
+failure-handling build system is based on adding an extra
+\hs{Either}~\hs{String} layer into the monad stack, e.g. augmenting the monad
+\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)} used by the schedulers
+in~\S\ref{sec-implementations} with exceptions. We omit the actual
+implementation: while fairly direct it is also tedious due to additional
+wrapping and unwrapping.
 

### 2019-05-02 papers/jfp/8-engineering.tex
Drop repeated text in S8.1

@@
 valid value. In the real world, build systems frequently deal with errors, e.g.
-``file not found'' or ``compilation failed''. We can model such failure
-conditions by instantiating \hs{v} to either \hs{Maybe}~\hs{v} (for missing
-values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
-model the errors inside the store and the \hs{Task}, but because our build
-system models are polymorphic in \hs{v}, they cannot apply any special behaviour
-for errors, e.g. aborting the build early. There are many other ways of
-modelling errors, and in this section we show three simple examples.
+``file not found'' or ``compilation failed''. There are many ways of modelling
+errors, and in this section we give three simple examples.
 
-As mentioned above, we can include failures into the type of values \hs{v}, for
-example, to model a partial store we can use a simple like algebraic data type
-isomorphic to \hs{Maybe}:
+% We can model such failure
+% conditions by instantiating \hs{v} to either \hs{Maybe}~\hs{v} (for missing
+% values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
+% model the errors inside the store and the \hs{Task}, but because our build
+% system models are polymorphic in \hs{v}, they cannot apply any special behaviour
+% for errors, e.g. aborting the build early.
+
+One simple approach is to include failures into the type of values \hs{v}, for
+example, to model a partial store we can use an algebraic data type isomorphic
+to \hs{Maybe}:
 
@@
 Here \hs{liftEither} wraps the result of the given
-\hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{Either}~\hs{e}~\hs{v} into \hs{EitherT}
-and then runs the task in the \hs{EitherT} monad
+\hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{Either}~\hs{e}~\hs{v} into \hs{ExceptT}
+and then runs the task in the \hs{ExceptT} monad
 transformer~\cite{liang1995monad}.

### 2019-05-02 papers/jfp/2-background.tex
Fix typo

@@
     need files
-    system "tar" $ ["-cf", "release.tar"] ++ files
+    cmd "tar" $ ["-cf", "release.tar"] ++ files
 \end{minted}
@@
 We first declare the static dependency on \cmd{release.txt}, then read its
-content (a list of files) and depend on each listed file, dynamically. Finally, we specify the
+content (a list of files) and depend on each listed file, dynamically. Finally,
+we specify the
 command to produce the resulting archive. Crucially, the archive will only be

### 2019-05-02 papers/jfp/8-engineering.tex

@@
 
-To legitimise shallow builds, we need to relax the correctness
-Definition~\ref{def-correct} as follows. Let the \hs{shallow} store correspond
-to the result of a shallow build. Then \hs{shallow} is correct, if \emph{there
-exists} a \hs{result} which satisfies all requirements of the
-Definition~\ref{def-correct}, \emph{such that} \hs{shallow} agrees with the
-\hs{result} on all the input keys \hs{k}~$\in$~$I$:
-\[
-\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result} % \text{,}
-\]
-and on the target \hs{key}:
-\[
-\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result} % \text{.}
-\]
-
-\noindent
-This relaxes the requirements on shallow builds by dropping the constraints on
-the \hs{shallow} store for all intermediate keys \hs{k}~$\in$~$O\setminus \{\hs{key}\}$.
+Shallow builds have a slightly weaker correctness property than
+Definition~\ref{def-correct}. Instead of demanding that \emph{all} keys reachable
+from the target match, we only require matches for the target itself and the \emph{input
+keys} reachable from the target.
 
@@
 
-Deep constructive traces (\S\ref{sec-deep-constructive-traces}) combined with
-task non-determinism (\S\ref{sec-non-determinism}) can lead to very subtle bugs,
-especially in the cloud setting. Fig.~\ref{fig-frankenbuild} shows a
+As described in \S\ref{sec-using-cloud-shake}, non-determinism  (\S\ref{sec-non-determinism})
+is harmful for cloud builds, reducing the number of cache hits.
+However, for deep constructive traces (\S\ref{sec-deep-constructive-traces}) it
+is much worse, even leading to \emph{incorrect results}.
+Fig.~\ref{fig-frankenbuild} shows a
 \emph{Frankenbuild}~\cite{esfahani2016cloudbuild} example, where the target

### 2019-05-02 papers/jfp/8-engineering.tex
Switch to MonadRandom

@@
 
-One tempting way of modelling non-determinism is to enrich \hs{Tasks} from
-\hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus},
-respectively. Here is a task description corresponding to a spreadsheet with the
-formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}:
+One way of modelling non-determinism is to \emph{demand that}~\hs{f}~\emph{has a
+built-in source of non-determinism}, for example, by enriching
+\hs{Tasks}~\hs{Monad} to \hs{Tasks}~\hs{MonadRandom} that has access to an
+additional method
+\hs{getRandom}~\hs{::}~\hs{(Integer,}~\hs{Integer)}~\hs{->}~\hs{m}~\hs{Integer},
+i.e. a source of random numbers in a specified range. Here is a task description
+corresponding to a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}:
 
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh3 :: Tasks MonadPlus String Integer
-sprsh3 "B1" = Just $ Task $ \fetch -> (+) <$> fetch "A1"
-                                          <*> (pure 1 <|> pure 2)
+sprsh3 :: Tasks MonadRandom String Integer
+sprsh3 "B1" = Just $ Task $ \fetch -> do a1 <- fetch "A1"
+                                         r  <- getRandom (1,2)
+                                         return (a1 + r)
 sprsh3 _    = Nothing

### 2019-05-02 papers/jfp/8-engineering.tex
Rewrite S8.7

--- 2019-05-02 papers/jfp/8-engineering.tex
@@
 
-We can use the extra polymorphism from \S\ref{sec-polymorphism} for build tasks that produce multiple output
-keys~--~for example, \cmd{ghc A.hs} produces both \cmd{A.hi} and \cmd{A.o}.
-That can be represented by having a key whose value is a pair of file names, and
-whose result is a pair of file contents. From that, the task for \cmd{A.hi}
-can be the first component of the result of the pair. Such an operation
-can be encoded without polymorphic keys provided the pair of files (or a dummy
-file representing the pair) is marked as changed if either of the contained
-files change, but it is more elegant and type-safe with the extra polymorphism.
+Some build tasks produce multiple output keys -- for example \cmd{ghc A.hs} produces both
+\cmd{A.hi} and \cmd{A.o}. This pattern can be encoded by having a key \cmd{A.hi+A.o} which produces
+both results, then each of \cmd{A.hi} and \cmd{A.o} can extract the relevant result
+from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety from
+\S\ref{sec-polymorphism}, using either pairs or lists for both the key and value:
 
-Multiple-output build tasks require one more convenience mechanism,
-which is orthogonal to polymorphism. Multiple-output task descriptions partition
-the set of keys into \emph{elementary subsets} e.g.
-$\{\cmd{A.hi},\cmd{A.o}\} \rightarrow \cmd{ghcTask}$,
-$\{\cmd{B.o}\} \rightarrow \cmd{gccTask}$, etc. It is very convenient to
-allow to fetch arbitrary sets of keys, such as $\{\cmd{A.o},\cmd{B.o}\}$,
-without manually decomposing them into appropriate elementary subsets. This
-requires a method to efficiently construct task descriptions operating on
-arbitrary sets of keys rather on elementary subsets. One way to implement such
-``multiple-output task closure'' is shown below.
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Key a where
+    File  :: FilePath -> Key String
+    Files :: [FilePath] -> Key [String]
+
+task :: TasksT Monad Key
+task (File "A.hi") = Just $ TaskT $ \fetch -> do
+    [hi, o] <- fetch (Files ["A.hi","A.o"])
+    pure hi
+task (File "A.o") = Just $ TaskT $ \fetch -> do
+    [hi, o] <- fetch (Files ["A.hi","A.o"])
+    pure o
+task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
+    -- run ghc here
+\end{minted}
+\vspace{1mm}
+
+The only awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for exactly the same
+\cmd{Files} key. If one \hs{File} key swapped the order of the list to \hs{Files} then it would likely
+run \cmd{ghc} twice. To help construct a well-formed multiple-output build task
+it is convenient to partition the set of keys into \emph{elementary subsets}.
+We can encode such a partition with a function from any key to all the members of its
+subset (in a consistent order).
 
@@
 type Partition k = k -> [k]
-
-multi :: Eq k => Partition k -> Tasks Applicative [k] [v]
-                             -> Tasks Applicative [k] [v]
-multi partition tasks keys
-    | k:_ <- keys, partition k == keys = tasks keys
-    | otherwise = Just $ Task $ \fetch ->
-        sequenceA [ select k <$> fetch (partition k) | k <- keys ]
-  where
-    select k = fromMaybe (error msg) . lookup k . zip (partition k)
-    msg = "Partition invariants violated"
 \end{minted}
@@
 
-\noindent
-The function \hs{multi} takes a task description \hs{tasks} that operates on
-elementary key subsets, and computes its closure supporting all possible
-combinations of keys. The first argument \hs{partition} specifies the elementary
-key subsets supported by the original \hs{tasks}. There are two cases: (i) if a
-given set of \hs{keys} is an elementary subset, we pass the query to \hs{tasks};
-(ii) otherwise, we examine every key \hs{k} in \hs{keys}, \hs{fetch} its
-corresponding elementary subset, and \hs{select} the appropriate value from the
-result. Returning to the above example, when given the query
-$\{\cmd{A.o},\cmd{B.o}\}$, this function will \hs{fetch} elementary subsets
-$\{\cmd{A.hi},\cmd{A.o}\}$ and $\{\cmd{B.o}\}$, thereby depending on the
-corresponding build tasks \cmd{ghcTask} and \cmd{gccTask}.
+\noindent In our example the partition function on either of the output names \hs{"A.hi"} or \hs{"A.o"} would
+return \hs{["A.hi", "A.o"]}.
 
-We chose not to use polymorphic tasks in the above implementation for
-simplicity, but nothing prevents us from combining polymorphism with
-multiple-output task closure.
+With a suitable \hs{Partition} it is possible to create a mapping that resolves \hs{File}
+keys automatically into the correct \hs{Files} key.
+
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+task :: Partition FilePath -> TasksT Monad Key
+task partition (File k) = Just $ TaskT $ \fetch -> do
+    let ks = partition k
+    let Just i = elemIndex k ks
+    vs <- fetch (Files ks)
+    pure (vs !! i)
+-- patterns matching Files
+\end{minted}
+\vspace{1mm}
+
+In the above function we compute the partition, find the index of this key within that
+partition (using \hs{elemIndex}), run the \hs{fetch} on every key in the partition,
+then extract out our specific result (using the indexing operation \hs{!!}).
 

### 2019-05-03 papers/jfp/3-abstractions.tex
Work on S8.3

@@
   discuss this in~\S\ref{sec-failures}.
-\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
+\item \hs{Task}~\hs{MonadPlus}, \hs{Task}~\hs{MonadRandon} and their
   variants can be used for describing tasks with a certain type of

### 2019-05-03 papers/jfp/8-engineering.tex
Work on S8.3

@@
 
-Most of the implementations in \S\ref{sec-implementations} can deal with
-non-determinism, apart from \Buck, which relies on deterministic tasks, and
-in turn can optimise the number of roundtrips required to the server.
+Most of the implementations in~\S\ref{sec-implementations} can deal with
+non-determinism, apart from \Buck, which uses the assumption of task determinism
+to reduce the number of cloud lookups: if all tasks are deterministic, one needs
+just a single cloud lookup for obtaining the value of the target key from the
+hash of its terminal inputs. This is highly desirable not only for shallow
+builds, but for non-shallow builds too, because under the assumption of task
+determinism \emph{all} intermediate values are fully determined by the terminal
+inputs, and can therefore be requested in a single batch query to the server.
 
@@
 Such tasks can be modelled in our framework by adjusting the correctness
-definition~(\S\ref{sec-build-correctness}), where instead of requiring that the
-\hs{result}ing value \emph{equals} the one obtained by recomputing the \hs{task}:
-\[
-\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result} %\text{,}
-\]
-\noindent
-we now require that \hs{result} contains \emph{one possible result of
-recomputing} the \hs{task}:
-\[
-\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result} %\text{,}
-\]
-where
-\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{[@@v]}
-returns the list of all possible results of the \hs{task} instead of just one
-value (`ND' stands for `non-deterministic').
+definition~(\S\ref{sec-build-correctness}): instead of requiring that the
+resulting value \emph{equals the result} of recomputing the task, we now require
+that resulting value \emph{belongs to the set of possible results} of
+recomputing the task, i.e. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
+example.
 
@@
 
-Note that \hs{Task}~\hs{MonadPlus} is powerful enough to model dependency-level
-non-determinism, for example, \cmd{INDIRECT("A" \& RANDBETWEEN(1,2))}, whereas
-most build tasks in real-life build systems only experience a value-level
-non-determinism. \Excel handles this example simply by marking the cell
-volatile~--~an approach that can be readily adopted by any of our
-implementations.
+Note that \hs{Task}~\hs{MonadRandom} is powerful enough to model
+dependency-level non-determinism, for example,
+\cmd{INDIRECT("A" \& RANDBETWEEN(1,2))}, whereas most build tasks in real-life
+build systems only experience a value-level non-determinism. \Excel handles this
+example simply by marking the cell volatile~--~an approach that can be readily
+adopted by any of our implementations.
 

### 2019-05-03 papers/jfp/8-engineering.tex
Rework polymorphic values

@@
 
-\subsection{Polymorphic values}\label{sec-polymorphism}
+\subsection{Key-dependent value types}\label{sec-polymorphism}
 
-Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build
-system that works directly on \hs{k} and \hs{v} values. However, some build
-systems provide greater flexibility, e.g. \Shake permits polymorphic keys and
-values, allowing types that are stored only in the \Shake's build information.
-\Shake users have remarked that polymorphism provides a much easier expression
-of concepts, e.g. see~Mokhov~\etal~\shortcite{hadrian}.
+Key-dependent value types allows a single build system to work with multiple different
+types of values, where the type of any particular value is determined by the key. As an example
+of why this might be useful, consider a build system that can depend on files
+(whose contents are strings) or system executables (represented by their version number) --
+using a single type for both values reduces type safety.
+\Shake permits such key-dependent value types, for example the \emph{oracle} rule from
+Mitchell~\shortcite{mitchell2012shake}, and users have remarked that this additional type safety
+provides a much easier expression of concepts, e.g. see~Mokhov~\etal~\shortcite{hadrian}.
 
-As one example of richer key/value types, consider the version of
-\cmd{gcc}~--~for many builds it should be a dependency. In \Shake it is possible
-to define an \emph{oracle} rule as per Mitchell~\shortcite{mitchell2012shake}
-whose value is the result of running \cmd{gcc -}\cmd{-version} and which is
-volatile, making the \cmd{gcc} version something that can be depended upon. Of
-course, provided the build can express volatile dependencies and supports
-cutoff, the version number could equally be written to a file and used in a
-similar way, albeit at the cost of introducing additional correctness
-invariants, e.g. that the file \cmd{gcc-version.txt} does indeed contains two
-dot-separated integers.
-
-Generalised algebraic data types, or GADTs~\cite{spj2006gadts}, can be used to
-extend our models to support ``typed'' build tasks. The idea is to replace
+We can encode key-dependent value types using generalised algebraic data types,
+or GADTs~\cite{spj2006gadts}. The idea is to replace
 the callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
@@
 where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
-corresponding values. The idea is best explained by way of an example:
+corresponding values. The variable \hs{k} has changed from kind \hs{*} (a type),
+to \hs{* -> *} (a type function), permitting the key to constrain the type of the value.
+The idea is best explained by way of an example:
 
@@
     File    :: FilePath -> Key String
-    GccVersion :: Key Version
+    Program :: String -> Key Version
 \end{minted}
@@
 Here we extend the usual mapping from file paths to file contents with an
-additional key type \hs{GccVersion} mapped to a \hs{Version}. The task
+additional key type \hs{Program} which maps the name of a program to its installed \hs{Version}. The task
 abstraction needs to be adjusted to cope with such keys (the suffix \hs{T}
@@
 of different types: the rule \cmd{release.txt} concatenates the contents of two
-\hs{File}s, while the rule \cmd{main.o} uses the numeric \hs{GccVersion} to
-determine the way the \hs{source} file should be compiled.
+\hs{File}s, while the rule \cmd{main.o} uses the numeric \hs{Program "gcc"} to
+determine how the \hs{source} file should be compiled.
 
@@
     let source = "main.c"
-    version <- fetch GccVersion
+    version <- fetch (Program "gcc")
     if version >= Version 8 0 then compileNew source
@@
 \vspace{1mm}
+
+Note that like all key-dependent tasks, this example could be expressed
+without key-dependence, at the cost of some type safety. As we'll see in
+\S\ref{sec-multiple-outputs} and \S\ref{sec-tracking-aspects}, using key-dependent
+value types can make it easier to work with more complicated tasks.
 

### 2019-05-03 papers/jfp/9-related.tex
Rework polymorphic values

@@
 combination. \Ninja~is also capable of modelling build rules that produce
-multiple results, a limited form of polymorphism \S\ref{sec-polymorphism}.
+multiple results, a limited form of multiple value types \S\ref{sec-polymorphism}.
 
@@
 matches \Shake at the level of detail given here, differing only on aspects like
-polymorphic dependencies~\S\ref{sec-polymorphism}.
+rules producing multiple files~\S\ref{sec-polymorphism}.
 

### 2019-05-03 papers/jfp/5-rebuilders.tex
Minor rewording

@@
 values/hashes of dependencies used last time, and if something has changed, the
-key is dirty and must therefore be rebuilt~--~in essence, keeping a \emph{trace}
+key is dirty and must be rebuilt~--~in essence, keeping a \emph{trace}
 which we can use to \emph{verify} existing values. Taking the example from
@@
 For traces, there are two essential operations~--~adding a new trace to the
-trace store, and using the traces to determine if a key needs rebuilding.
+trace store, and using the trace store to determine if a key needs rebuilding.
 Assuming a store of verifying traces \hs{VT}~\hs{k}~\hs{v}, the operations are:
@@
 Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} can store
-only hashes, of type \hs{Hash}~\hs{v}, of those values. Since the verifying
+only hashes of those values, with type \hs{Hash}~\hs{v}. Since the verifying
 trace persists from one build to the next -- it constitutes the build system's
@@
 \item If \hs{h} does not equal the \hs{result} field of the trace, return
-  \hs{False}: the current \hs{k}'s value was written externally and needs
+  \hs{False}: the current \hs{k}'s value was changed externally and thus needs
   rebuilding.
@@
 The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
-the hash of the resulting value, we require the actual value. The \hs{verifyVT}
+the hash of the resulting value, we pass the actual value. The \hs{verifyVT}
 has been replaced with \hs{constructCT}, which instead of taking the hash of the
@@
 In practice, many cloud build systems store hashes of values in the trace store,
-i.e. use \hs{Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v)} trace entries just like the
+i.e. use \hs{Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v)} entries just like
 verifying traces, and have a separate content-addressable cache which associates
@@
         \emph{deterministic} then it is possible to violate correctness, as
-        illustrated by an example in~\S\ref{sec-cloud-aspects}, see
+        illustrated by the example in~\S\ref{sec-cloud-aspects}, see
         Fig.~\ref{fig-frankenbuild}.
@@
 $n = 1$. By picking values of $n$ in between we would regain some early cutoff, at the
-cost of losing such simple shallow builds and still requiring deterministic tasks.
+cost of losing such simple shallow builds, while still requiring deterministic tasks.

### 2019-05-03 papers/jfp/6-implementations.tex
Update TODOs
Tweaks to S6

@@
 \hs{dirtyBitRebuilder}, see Fig.~\ref{fig-excel-implementation}. The persistent
-build information \hs{ExcelInfo}~\hs{k} is a pair: (i) a map
+build information \hs{ExcelInfo}~\hs{k} is a pair of: (i) a map
 \hs{k}~\hs{->}~\hs{Bool} associating a dirty bit with every key, and (ii) a
-calc chain of type \hs{[@@k]} recorded in the previous build
+calc chain of type \hs{[@@k]} recorded from the previous build
 (\S\ref{sec-background-excel}).
@@
 systems. \Bazel, \CloudBuild and \Cloud \Shake are based on constructive traces
-(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
-variant (\S\ref{sec-deep-constructive-traces}).
+(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use deep
+constructive traces (\S\ref{sec-deep-constructive-traces}).
 
@@
 collection of predefined built-in rules, which cover many
-important instances of dynamic dependencies.
+common instances of dynamic dependencies.
 
@@
 Microsoft's \CloudBuild~-- an applicative build system that combines
-conventional scheduling of statically known directed acyclic graphs of
-dependencies with constructive traces~\cite{esfahani2016cloudbuild}. Note that
+conventional scheduling of statically known directed acyclic graphs
+with constructive traces~\cite{esfahani2016cloudbuild}. Note that
 we need to convert a monadic \hs{ctRebuilder} into an applicative one by
@@
 
-\todo{AM}{Add implementation of \hs{dctRebuilder}?}
+% \todo{AM}{Add implementation of \hs{dctRebuilder}?}
 

### 2019-05-03 papers/jfp/7-experience.tex
Rework polymorphic values
Tweaks to S7

@@
 in the original paper the tasks (there called ``actions'')
-were described in continuation-passing style using this data type\footnote{
+were described in continuation-passing style using the data type:\footnote{
 The original paper uses concrete types \hs{Key} and \hs{Value}. Here we
 generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
-\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}}.
+\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}}
 \vspace{1mm}
@@
 \noindent
-Similarly, in the original paper \Make tasks were described as equivalent to
+Similarly, in the original paper \Make tasks were described as:
 
@@
 from~\S\ref{sec-constructive-traces}, with their differences perhaps the most
-illustrative of the changes required\footnote{
-  We have made some minor changes from
+illustrative of the changes required\footnote{We have made some minor changes from
 actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant
@@
 command lines by treating them as a type of values -- an example of
-\emph{partial self-tracking} made possible by \Shake's support for value
-polymorphism~(\S\ref{sec-polymorphism}).
+\emph{partial self-tracking} made possible by \Shake's support for key-dependent
+value types~(\S\ref{sec-polymorphism}).
 
@@
 building and testing every GHC commit. Building GHC~8.8 from scratch with
-default settings takes around 1 hour on Windows using both \Hadrian and the
-original \Make-based build system. This includes building the compiler itself,
+default settings takes around 1 hour on Windows using either \Hadrian or the
+original \Make-based build system. This time includes building the compiler itself,
 29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
@@
 a shared file. \Cloud~\Shake provides a way to manually label such
-build rules, thereby excluding them from caching.
+build rules to exclude them from caching.
 

### 2019-05-03 papers/jfp/8-engineering.tex
Make the multiple-output thing clearer
Tweaks to S8

@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Value = FileNotFound | FileContents ByteString
+data Value = FileNotFound | FileContents String
 \end{minted}
@@
 but all the build systems we address can actually build independent keys in
-parallel. While it complicates the model, the complications can be restricted
-exclusively to the scheduler:
+parallel. While it complicates the model, the complications are restricted
+to the scheduler:
 
@@
     pure (vs !! i)
--- patterns matching Files
+task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
+    -- run ghc here
+-- more rules on elementary subsets as required
 \end{minted}

### 2019-05-03 papers/jfp/3-abstractions.tex
Clarify the purpose of S3.8
Fix typos and formatting

@@
 we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
-Here is our first use of polymorpism in \hs{f}, discussed in \S\ref{sec-why-polymorphism}.
+Here is our first use of polymorphism in \hs{f}, discussed in~\S\ref{sec-why-polymorphism}.
 The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
@@
 constraints on the computation context: \hs{Functor}, \hs{Applicative},
-\hs{Monad} and \hs{MonadState}~\hs{s}.
+\hs{Monad} and \hs{MonadState}~\hs{s}. The purpose of these examples is to
+continue building the intuition behind the \hs{Task} abstraction, and prepare
+the reader for richer types of tasks that will appear
+in~\S\ref{sec-implementations} and~\S\ref{sec-engineering}.
 

### 2019-05-03 papers/jfp/5-rebuilders.tex
Fix typos and formatting

@@
 \noindent
-The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
+The function \hs{recordCT} is similar to \hs{recordVT}
+from~\S\ref{sec-verifying-traces}, but instead of just passing
 the hash of the resulting value, we pass the actual value. The \hs{verifyVT}

### 2019-05-03 papers/jfp/7-experience.tex
Fix typos and formatting

@@
 
-Based on the thinking in this paper we have extended \Shake to
+Based on the theory developed in this paper we have extended \Shake to
 become \Cloud \Shake, the first cloud-capable build system to support
-both early cutoff and monadic dependencies (\S\ref{sec-implementation-cloud}),
+both early cutoff and monadic dependencies~(\S\ref{sec-implementation-cloud}),
 and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
@@
 \vspace{1mm}
+
 \noindent
 In this paper we describe tasks more directly, in a monadic (or applicative or functorial)
-style. But in fact the two are equivalent: \hs{Task Monad k v} is
-isomporphic to \hs{Action k v v}.
-To be concrete, the functions \hs{toAction} and \hs{fromAction} witness the
-isomorphism in both directions.
+style. But in fact the two are equivalent: \hs{Task}~\hs{Monad}~\hs{k}~\hs{v} is
+isomorphic to \hs{Action}~\hs{k}~\hs{v}~\hs{v}.
+To be concrete, the functions \hs{toAction} and \hs{fromAction} defined below
+witness the isomorphism in both directions.
+
 \vspace{1mm}
@@
 \noindent for various choices of~\hs{c}, are studied in depth by~Jaskelioff and
-O’Connor~\shortcite{jaskelioff2015representation}.
+O'Connor~\shortcite{jaskelioff2015representation}.
 
@@
       a simple list. The reason is that \Shake allows a list of dependencies to
-      be specfied simultaneously, so they can all be built in parallel.
+      be specified simultaneously, so they can all be built in parallel.
 \item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
@@
 \begin{center}
-\parbox{0.8 \textwidth}{\emph{Thanks for the symlinks release, we just finished upgrading this build
-system to use \cmd{-}\cmd{-share}. ... Building from scratch with a warm cache
-takes around 5 seconds, saving us up to 2 hours. Not bad!}}
+\parbox{0.8 \textwidth}{\emph{Thanks for the symlinks release, we just finished
+upgrading this build system to use \textsf{-}\textsf{-share}. ... Building from
+scratch with a warm cache takes around 5 seconds, saving us up to 2 hours. Not
+bad!}}
 \end{center}

### 2019-05-03 papers/jfp/8-engineering.tex
Fix typos and formatting

@@
 \vspace{1mm}
-\begin{minted}[xleftmargin=10pt]{haskell}
+\begin{minted}[xleftmargin=5pt]{haskell}
 type Fetch k f = @\std{forall}@ v. k v -> f v
 
-newtype TaskT c k v = TaskT { runT :: @\std{forall}@ f. c f => Fetch k f -> f v }
+newtype TaskT c k v = TaskT@\,@{@\,@runT :: @\std{forall}@ f. c f => Fetch k f -> f v@\,@}
 
@@
 both results, then each of \cmd{A.hi} and \cmd{A.o} can extract the relevant result
-from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety from
-\S\ref{sec-polymorphism}, using either pairs or lists for both the key and value:
+from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety
+from~\S\ref{sec-polymorphism}, using either pairs or lists for both the key and value:
 
@@
 task (File "A.hi") = Just $ TaskT $ \fetch -> do
-    [hi, o] <- fetch (Files ["A.hi","A.o"])
-    pure hi
+    [hi,o] <- fetch (Files ["A.hi","A.o"])
+    return hi
 task (File "A.o") = Just $ TaskT $ \fetch -> do
-    [hi, o] <- fetch (Files ["A.hi","A.o"])
-    pure o
+    [hi,o] <- fetch (Files ["A.hi","A.o"])
+    return o
 task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
-    -- run ghc here
+    compileWithGHC "A.hs"
 \end{minted}
@@
 
-The only awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for exactly the same
-\cmd{Files} key. If one \hs{File} key swapped the order of the list to \hs{Files} then it would likely
-run \cmd{ghc} twice. To help construct a well-formed multiple-output build task
-it is convenient to partition the set of keys into \emph{elementary subsets}.
-We can encode such a partition with a function from any key to all the members of its
-subset (in a consistent order).
+The only awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for
+exactly the same \hs{Files} key. If one \hs{File} key swapped the order of the
+list to \hs{Files} then it would likely run GHC twice. To help construct a
+well-formed multiple-output build task it is convenient to partition the set of
+keys into \emph{elementary subsets}. We can encode such a partition with a
+function from any key to all the members of its subset (in a consistent order).
 
@@
 
-\noindent In our example the partition function on either of the output names \hs{"A.hi"} or \hs{"A.o"} would
-return \hs{["A.hi", "A.o"]}.
+\noindent
+In our example the partition function on either of the output names \hs{"A.hi"}
+or \hs{"A.o"} would return \hs{["A.hi","A.o"]}.
 
-With a suitable \hs{Partition} it is possible to create a mapping that resolves \hs{File}
-keys automatically into the correct \hs{Files} key.
+With a suitable \hs{Partition} it is possible to create a mapping that resolves
+\hs{File} keys automatically into the correct \hs{Files} key.
 
@@
     vs <- fetch (Files ks)
-    pure (vs !! i)
+    return (vs !! i)
 task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
-    -- run ghc here
--- more rules on elementary subsets as required
+    compileWithGHC "A.hs"
+... -- more tasks for elementary subsets as required
 \end{minted}
@@
 
-\vspace{0.5mm}
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{text}

### 2019-05-03 papers/jfp/10-conclusions.tex
Restore acknowledgements

@@
 
-We have investigated multiple build systems, showing how their properties are consequences of two implementation choices: what order you build in and how you decide whether to rebuild.
-By first decomposing the pieces, we show how to recompose the pieces to find new points in the design space.
-In particular, a simple recombination leads to a design for a monadic suspending cloud build system, which we have implemented and use in our day-to-day development.
+We have investigated multiple build systems, showing how their properties are
+consequences of two implementation choices: what order you build in and how you
+decide whether to rebuild. By first decomposing the pieces, we show how to
+recompose the pieces to find new points in the design space. In particular, a
+simple recombination leads to a design for a monadic suspending cloud build
+system, which we have implemented and use in our day-to-day development.
+
+\section*{Acknowledgements}
+
+Thanks to anonymous reviewers and everyone else who provided us with feedback
+on earlier drafts: Ulf Adams, Arseniy Alekseyev, Dan Bentley, Martin
+Br\"{u}stel, Ulan Degenbaev, Jeremie Dimino, Andrew Fitzgibbon, Georgy
+Lukyanov, Simon Marlow, Evan Martin, Yaron Minsky, Guillaume Maudoux, Philip
+Patsch, Michael Peyton Jones, Andrew Phillips, Fran\c{c}ois Pottier, Rohit
+Ramesh, Irakli Safareli, Zhen Zhang. Your contributions were incredibly
+valuable.
+
+Andrey Mokhov is funded by a Royal Society Industry Fellowship \cmd{IF160117}
+on the topic ``Towards Cloud Build Systems with Dynamic Dependency Graphs''.

### 2019-05-25 papers/jfp/8-engineering.tex
Fix sprshN numbering

@@
 
-sprsh4 :: Tasks MonadFail String Integer
-sprsh4 "B1" = Just $ Task $ \fetch -> do
+sprsh3 :: Tasks MonadFail String Integer
+sprsh3 "B1" = Just $ Task $ \fetch -> do
     a1 <- fetch "A1"
@@
     if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
-sprsh4 _ = Nothing
+sprsh3 _ = Nothing
 \end{minted}
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh3 :: Tasks MonadRandom String Integer
-sprsh3 "B1" = Just $ Task $ \fetch -> do a1 <- fetch "A1"
+sprsh4 :: Tasks MonadRandom String Integer
+sprsh4 "B1" = Just $ Task $ \fetch -> do a1 <- fetch "A1"
                                          r  <- getRandom (1,2)
                                          return (a1 + r)
-sprsh3 _    = Nothing
+sprsh4 _    = Nothing
 \end{minted}
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh4 "B1" = Just $ Task $ \fetch -> do
+sprsh5 "B1" = Just $ Task $ \fetch -> do
     formula <- fetch "B1-formula"
@@
 The function \hs{selfTracking} defined below is a generalisation of the approach
-explained in the above \Excel example \hs{sprsh4}. The function takes a parser
+explained in the above \Excel example \hs{sprsh5}. The function takes a parser
 for scripts, of type \hs{s}~\hs{->}~\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}, and a
 description of \emph{how to build all scripts}, of type
-\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{s}. For \hs{sprsh4}, the latter would simply
+\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{s}. For \hs{sprsh5}, the latter would simply
 fetch \cmd{B1-formula} when given the key \cmd{B1} and return \hs{Nothing}

### 2019-10-13 papers/jfp/1-intro.tex
Start working on a revision

@@
   dependencies; local vs cloud; deterministic vs non-deterministic build tasks;
-  support for early cutoff; self-tracking build systems; and the type of
-  persistent build information. In~\S\ref{sec-background} we identify some of these key
+  early cutoff; self-tracking build systems; and the type of persistently stored
+  build information. In~\S\ref{sec-background} we identify some of these key
   properties, illustrated by four carefully-chosen build systems.
@@
   crisply encapsulate what a build system is (\S\ref{sec-abstractions}),
-  allowing us, for example, to speak about what it means for a build system to be correct.
+  allowing us, for example, to speak about what it means for a build system to
+  be correct.
 
@@
 
-\item We can use these new
-abstractions to more clearly explain details from the original \Shake
-paper~\cite{mitchell2012shake} (\S\ref{sec-step-traces}, \S\ref{sec-experience-shake})
-and build new features useful for the GHC build system~\cite{hadrian} (\S\ref{sec-using-cloud-shake}).
+\item We can use these new abstractions to more clearly explain details from the
+original \Shake paper~\cite{mitchell2012shake} (\S\ref{sec-step-traces},
+\S\ref{sec-experience-shake}) and build new features useful for the GHC build
+system~\cite{hadrian} (\S\ref{sec-using-cloud-shake}).
 
@@
 understanding of what they do and how they compare, and making it easier to
-explore other points in the landscape.
+explore other points in the landscape. While we steer clear of many engineering
+aspects of real build systems, in~\S\ref{sec-engineering} we show how they can
+be incorporated into the presented abstractions. The related work is discussed
+in~\S\ref{sec-related}.
 
-We discuss engineering aspects in \S\ref{sec-engineering}, and related
-work in \S\ref{sec-related}. This paper is an extended version of an earlier
-conference paper~\cite{mokhov2018buildsystems}.
-
-% Papers about ``frameworks'' are often fuzzy.  This one is not: all our
-% abstractions are defined in Haskell, and we have (freely-available)
-% executable models of all the build systems we describe.
+This paper is an extended version of an earlier conference
+paper~\cite{mokhov2018buildsystems}. The key changes compared to the earlier
+version are: (i) we added further clarifications and examples
+into~\S\ref{sec-abstractions}, in particular,~\S\ref{sec-task-examples} is
+entirely new; (ii)~\S\ref{sec-scheduler} and~\S\ref{sec-rebuilder} are based on
+the material from the conference paper but have been substantially expanded to
+include further details and examples, as well as completely new material such
+as~\S\ref{sec-step-traces}; (iii)~\S\ref{sec-experience} is completely new; (iv)
+\S\ref{sec-failures} and \S\S\ref{sec-polymorphism}-\ref{sec-tracking-aspects}
+are almost entirely new, and \S\ref{sec-non-determinism} has been revised.
+The new material focuses on our experience and various important practical
+considerations, hence justifying the "and Practice" part of the paper title.

### 2019-10-13 papers/jfp/3-abstractions.tex
Start working on a revision

@@
 -- Build tasks (see §3.2)
-newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
+data Task  c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
 type Tasks c k v = k -> Maybe (Task c k v)
+\end{minted}
+% \vspace{-1mm}
+\begin{minted}[fontsize=\small]{haskell}
+run :: c f => Task c k v -> (k -> f v) -> f v
+run (Task task) = task
 \end{minted}
@@
 Our first main abstraction is for \emph{task descriptions}:
-\begin{minted}[xleftmargin=5pt]{haskell}
-newtype Task  c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Task  c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
 type Tasks c k v = k -> Maybe (Task c k v)
+\end{minted}
+\begin{minted}[xleftmargin=10pt]{haskell}
+run :: c f => Task c k v -> (k -> f v) -> f v
+run (Task task) = task
 \end{minted}

### 2019-10-14 papers/jfp/1-intro.tex
Add a TODO

@@
 The new material focuses on our experience and various important practical
-considerations, hence justifying the "and Practice" part of the paper title.
+considerations, hence justifying the ``and Practice'' part of the paper title.

### 2019-10-14 papers/jfp/8-engineering.tex
Start working on a revision
Add a TODO

@@
-\section{Engineering aspects}\label{sec-engineering}
+\section{Engineering Aspects}\label{sec-engineering}
 
@@
 
-\subsection{Key-dependent value types}\label{sec-polymorphism}
+\subsection{Key-dependent Value Types}\label{sec-polymorphism}
 
@@
 functions.
+
+\subsection{Integration with IDEs}\label{sec-integration-with-ides}
+
+\textbf{TODO:} Neil, could you add 1-2 paragraphs, perhaps with a code snippet
+illustrating keys that are pairs of a file and a compilation stage? Perhaps,
+you could link to the DAML IDE for an implementation.

### 2019-10-16 papers/jfp/7-experience.tex
Add 7.1 on Haskell as design language

@@
 and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
-In this section we tie these papers together, reflecting on our
-experience of turning theory into practice.
+In this section we reflect on our experience of turning theory into practice.
+
+\subsection{Haskell as a design language}\label{sec-design-lang}
+
+Build systems are surprisingly tricky.  It is easy to waffle, and
+remarkably hard to be precise. As this paper exemplifies, it is
+possible to use Haskell as a \emph{design language}, to express quite
+abstract ideas in a precise form -- indeed, precise enough to be executed.
+
+Moreover, doing so is extremely beneficial.
+The discipline of writing executable prototypes in
+Haskell had a profound effect on our thinking. It forced misconceptions
+to to surface early.  It required us to be explicit about side effects.
+It gave us a huge incentive to design abstractions that had simple types
+and an explicable purpose.
+
+Consider, for example, our \hs{Task} type:
+\begin{minted}[xleftmargin=10pt]{haskell}
+data Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
+\end{minted}
+We started off with a much more concrete type, and explored
+multiple variants.  During those iterations, this single line of
+code gave us a tangible, concrete basis for productive conversations,
+much more so than general conversations about ``tasks''.
+
+It is also worth noting that we needed a rather expressive language,
+to faithfully express the abstractions that seen natural in this
+setting.  In the case of \hs{Task} we use a data constructor with a
+polymorphic field; a higher-kinded type variable \hs{f :: * -> *}; an
+even more abstracted kind for \hs{c :: (* -> *) -> Constraint}; and,
+of course, type classes.
+
+Moreover, when thinking about the type constructors over which \hs{f} might
+usefully range, it turned out that we could adopt the \emph{existing} abstractions
+of \hs{Functor}, \hs{Applicative}, \hs{Monad} and so on.
+That in turn led us to think in a new way about the taxonomy
+of build systems -- see \S~\ref{sec-why-polymorphism}.
+
+The effect of using a concrete design language went well beyond
+merely \emph{expressing} our ideas: it \emph{directly influenced}
+our thinking.  For example, here are definitions for a scheduler and rebuilder,
+from \S~\ref{sec-concrete-impl}:
+\begin{minted}[fontsize=\small,xleftmargin=5pt]{haskell}
+type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
+type Rebuilder c   ir k v = k -> v -> Task c k v -> Task@\,@(MonadState ir) k v
+\end{minted}
+These powerful and modular abstractions, which ultimately formed part
+of the conceptual structure of the paper, emerged fairly late in the
+project, as we repeatedly reviewed, redesigned, and refactored our
+executable prototype.  It is hard to believe that we could have
+developed them without the support of Haskell as a design language.
 

### 2019-10-16 papers/jfp/8-engineering.tex
Add file watching and polling section

--- 2019-05-25 papers/jfp/8-engineering.tex
@@
 functions.
+
+\subsection{File watching vs polling}\label{sec-integration-with-ides}
+
+Most build systems use files for their inputs.
+When running a build, to check if a file is up-to-date, one option is to take the modification time or compute the content hash of each input file.
+However, for the largest projects, the mere act of checking the modification time for all the input files can become prohibitive.
+To overcome that problem, build systems targeted at large scale, e.g.
+\Bazel and \Buck, make use of file watching API's that detect when a file has changed.
+Using that information the build can avoid performing many file accesses.

### 2019-10-16 papers/jfp/8-engineering.tex
Expand S8.9

@@
 
-Most build systems use files for their inputs.
-When running a build, to check if a file is up-to-date, one option is to take the modification time or compute the content hash of each input file.
-However, for the largest projects, the mere act of checking the modification time for all the input files can become prohibitive.
-To overcome that problem, build systems targeted at large scale, e.g.
-\Bazel and \Buck, make use of file watching API's that detect when a file has changed.
-Using that information the build can avoid performing many file accesses.
+Most build systems use files for their inputs. When running a build, to check if
+a file is up-to-date, one option is to take the modification time or compute the
+content hash of each input file. However, for the largest projects, the mere act
+of checking the modification time for all the input files can become
+prohibitive. To overcome that problem, build systems targeted at large scale,
+e.g. \Bazel and \Buck, make use of file watching API's that detect when a file
+has changed. Using that information the build can avoid performing many file
+accesses.
+
+File-watching build systems face new engineering challenges. Firstly, a running
+task may become \emph{obsolete} due to a concurrent modification of its direct
+or transitive dependencies. To minimise the amount of unnecessary work, a
+scheduler can \emph{cancel} the obsolete task and restart it once its
+dependencies are up-to-date. Note that topological
+schedulers~(\S\ref{sec-topological}) cannot be easily adapted to support such
+situations, since a new topological sort may be required when file dependencies
+change. A further complication is that spurious dependency cycles may form
+during a series of file modifications, and a file-watching build system
+integrated with an IDE should be able to cope with such spurious cycles
+gracefully instead of terminating with an error.

### 2019-10-17 papers/jfp/1-intro.tex
Minor fixes and text tweaks, mostly for nicer formatting

@@
 
-\item We can use these new abstractions to more clearly explain details from the
-original \Shake paper~\cite{mitchell2012shake} (\S\ref{sec-step-traces},
-\S\ref{sec-experience-shake}) and build new features useful for the GHC build
-system~\cite{hadrian} (\S\ref{sec-using-cloud-shake}).
+\item We can use the presented abstractions to more clearly explain details from
+the original \Shake paper~(\S\ref{sec-step-traces}, \S\ref{sec-experience-shake})
+and develop new cloud build features, which are already in use in industry and in
+the GHC build system~(\S\S\ref{sec-using-cloud-shake}-\ref{sec-hadrian}).
 
@@
 explore other points in the landscape. While we steer clear of many engineering
-aspects of real build systems, in~\S\ref{sec-engineering} we show how they can
-be incorporated into the presented abstractions. The related work is discussed
+aspects of real build systems, in~\S\ref{sec-engineering} we discuss these
+aspects in the context of the presented abstractions. The related work is covered
 in~\S\ref{sec-related}.
@@
 version are: (i) we added further clarifications and examples
-into~\S\ref{sec-abstractions}, in particular,~\S\ref{sec-task-examples} is
+to~\S\ref{sec-abstractions}, in particular,~\S\ref{sec-task-examples} is
 entirely new; (ii)~\S\ref{sec-scheduler} and~\S\ref{sec-rebuilder} are based on
@@
 as~\S\ref{sec-step-traces}; (iii)~\S\ref{sec-experience} is completely new; (iv)
-\S\ref{sec-failures} and \S\S\ref{sec-polymorphism}-\ref{sec-tracking-aspects}
+\S\ref{sec-failures} and \S\S\ref{sec-polymorphism}-\ref{sec-file-watching}
 are almost entirely new, and \S\ref{sec-non-determinism} has been revised.

### 2019-10-17 papers/jfp/10-conclusions.tex
Minor fixes and text tweaks, mostly for nicer formatting

@@
 
-Andrey Mokhov is funded by a Royal Society Industry Fellowship \cmd{IF160117}
-on the topic ``Towards Cloud Build Systems with Dynamic Dependency Graphs''.
+Andrey Mokhov's research was funded by a Royal Society Industry Fellowship
+\cmd{IF160117} on the topic ``Towards Cloud Build Systems with Dynamic
+Dependency Graphs''.

### 2019-10-17 papers/jfp/2-background.tex
Minor fixes and text tweaks, mostly for nicer formatting

@@
 
-Build systems automate the execution of repeatable tasks, at a scale from individual users
-up to large organisations. In this section we explore the design space of build
-systems, using four concrete examples: \Make~\cite{feldman1979make},
-\Shake~\cite{mitchell2012shake}, \Bazel~\cite{bazel}, and
+Build systems automate the execution of repeatable tasks, at a scale from individual
+users up to large organisations. In this section we explore the design space of build
+systems, using four examples: \Make~\cite{feldman1979make},
+\Shake~\cite{mitchell2012shake}, \mbox{\Bazel~\cite{bazel}}, and
 \Excel~\cite{advanced_excel}.
@@
 tasks in a \emph{topological order}. For a more concrete description
-see~\S\ref{sec-implementation-make}.
+see~\S\ref{sec-topological} and~\S\ref{sec-implementation-make}.
 
@@
 We refer to this algorithm as \emph{restarting}, and discuss it
-in more detail in~\S\ref{sec-implementation-excel}.
+in more detail in~\S\ref{sec-restarting} and~\S\ref{sec-implementation-excel}.
 
@@
 determined by the content of \cmd{release.txt}, which might not even exist
-before the build. Makefiles cannot express such dependencies, requiring problematic
-workarounds such as \emph{build phases}~\cite{hadrian}.
-In \Shake we can express the rule for \cmd{release.tar} as:
+before the build. Makefiles cannot express such dependencies, requiring
+workarounds such as \emph{build phases}, which are known to be
+problematic~\cite{hadrian}. In \Shake we can express the rule for
+\cmd{release.tar} as:
 
@@
 \emph{incremental}~\cite{demers1981incremental},
-\emph{adaptive}~\cite{acar2002adaptive}, and
+\emph{adaptive}~\cite{acar2002adaptive}, as well as
 \emph{self-adjusting computations} -- see
@@
 their execution until the dependencies are brought up to date. We refer to this
-task scheduling algorithm as \emph{suspending}, see a concrete implementation
+task scheduling algorithm as \emph{suspending}, see a further discussion
+in~\S\ref{sec-suspending} and a concrete implementation
 in~\S\ref{sec-implementation-shake}.
@@
 early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
-not (see \S\ref{sec-verifying-traces} for an explanation of why).
+not; see~\S\ref{sec-dirty-bit} for an explanation of why.
 
@@
 intermediates in the cloud.
+
+\begin{figure}
+% \vspace{-2mm}
+\begin{subfigure}[b]{0.25\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
+\vspace{-0.5mm}
+\caption{Download sources}
+\end{subfigure}
+\begin{subfigure}[b]{0.40\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
+\vspace{-0.5mm}
+\caption{Build \cmd{main.exe}}
+\end{subfigure}
+\begin{subfigure}[b]{0.31\linewidth}
+\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
+\vspace{-0.5mm}
+\caption{Modify \cmd{util.c} and rebuild}
+\end{subfigure}
+\vspace{-1mm}
+\caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
+by downloading it from the cloud and skipping intermediate files (only their
+hashes are needed), (c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which
+requires building \cmd{util.o} (nobody has compiled \cmd{util.c} before) and
+downloading \cmd{main.o} (it is needed for linking \cmd{main.exe}). File hashes
+are shown in circles, and non-materialised intermediates in dashed rounded
+rectangles.\label{fig-bazel}}
+\vspace{-1mm}
+\end{figure}
 
@@
 
-\begin{figure}
-% \vspace{-2mm}
-\begin{subfigure}[b]{0.25\linewidth}
-\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
-\vspace{-0.5mm}
-\caption{Download sources}
-\end{subfigure}
-\begin{subfigure}[b]{0.40\linewidth}
-\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
-\vspace{-0.5mm}
-\caption{Build \cmd{main.exe}}
-\end{subfigure}
-\begin{subfigure}[b]{0.31\linewidth}
-\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
-\vspace{-0.5mm}
-\caption{Modify \cmd{util.c} and rebuild}
-\end{subfigure}
-\vspace{-1mm}
-\caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
-by downloading it from the cloud and skipping intermediate files (only their
-hashes are needed), (c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which
-requires building \cmd{util.o} (nobody has compiled \cmd{util.c} before) and
-downloading \cmd{main.o} (it is needed for linking \cmd{main.exe}). File hashes
-are shown in circles, and non-materialised intermediates in dashed rounded
-rectangles.\label{fig-bazel}}
-\vspace{-1mm}
-\end{figure}
-
-\vspace{-1.5mm}
+% \vspace{-1.5mm}
 \subsection{Summary}\label{sec-background-summary}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 

### 2019-10-17 papers/jfp/3-abstractions.tex
Add 7.1 on Haskell as design language
Note that newer Docker builds aren't merely Functor anymore
Minor fixes and text tweaks, mostly for nicer formatting

--- 2019-05-03 papers/jfp/3-abstractions.tex
@@
 
-\vspace{-1.5mm}
+% \vspace{-1.5mm}
 \subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 
@@
 -- Build tasks (see §3.2)
-newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
+newtype Task  c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
 type    Tasks c k v = k -> Maybe (Task c k v)
+\end{minted}
+% \vspace{-1mm}
+\begin{minted}[fontsize=\small]{haskell}
+run :: c f => Task c k v -> (k -> f v) -> f v
+run (Task task) fetch = task fetch
 \end{minted}
@@
 Our first main abstraction is for \emph{task descriptions}:
-\begin{minted}[xleftmargin=5pt]{haskell}
-newtype Task  c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
+\begin{minted}[xleftmargin=10pt]{haskell}
+newtype Task  c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
 type    Tasks c k v = k -> Maybe (Task c k v)
@@
 \begin{minted}[xleftmargin=5pt]{haskell}
-newtype Task c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
+newtype Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
 \end{minted}
@@
   description cannot even use the application operator \hs{<*>}, which limits
-  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}. It
+  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}
+  (ignoring the recent multi-stage builds). It
   is interesting to note that, when run on such a task description, the

### 2019-10-17 papers/jfp/5-rebuilders.tex
Minor fixes and text tweaks, mostly for nicer formatting

@@
 cutoff~(\S\ref{sec-background-shake}) it would be important to clear the dirty
-bit after a computation that did not change the value and make sure that cells
+bit after a computation that did not change the value and make sure that keys
 that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
@@
 \Make, it is impossible to mark a file clean and at the same time not mark the
-keys that depend on it dirty. \Make can approximate early cutoff by not
+files that depend on it dirty. \Make can approximate early cutoff by not
 modifying the result file, and not marking it clean, but then it will be rebuilt
@@
 
-A dirty bit rebuilder is useful to reduce memory consumption, and in the case
+A dirty-bit rebuilder is useful to reduce memory consumption, and in the case
 of \Make, to integrate with the file system. However, as the examples show, in
@@
 
-It is possible to write a dirty-bit builder which is minimal and supports early
-cutoff. To do so, you need to start with all inputs that have changed marked dirty,
-then a node must be rebuilt if any children are dirty, and must set that node to dirty
-only if the result changes. At the end of the build all dirty bits
-must be reset to not-dirty. This approach only works if all targets are rebuilt each time,
-since clearing dirty bits of children who aren't targets will cause them to incorrectly
-not rebuild subsequently. To avoid resetting the dirty bit,
-it is possible to use successive execution numbers, which ultimately leads to an approach
-we call verifying step traces in \S\ref{sec-step-traces}.
+It is possible to implement a dirty-bit rebuilder that is minimal and supports early
+cutoff. To do so, the build system should start with all inputs that have changed
+marked dirty, then a key must be rebuilt if any of its direct dependencies are dirty,
+marking the key dirty only if the result has changed. At the end of the build all
+dirty bits must be cleared. This approach only works if all targets are rebuilt each
+time because clearing dirty bits of keys that are not transitive dependencies of
+current targets will cause them to incorrectly not rebuild subsequently. To avoid
+resetting the dirty bit, it is possible to use successive execution numbers, which
+ultimately leads to an approach we call verifying step traces in~\S\ref{sec-step-traces}.
 
@@
 Any \hs{Applicative} build system using constructive traces, e.g.
-\CloudBuild~\S\ref{sec-implementation-cloud}, can index directly from the
-key and the hashes of its dependencies to the output result~--~i.e.
+\CloudBuild~(\S\ref{sec-implementation-cloud}), can index directly from the
+key and the hashes of its dependencies to the output result, i.e.
 \hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces

### 2019-10-17 papers/jfp/6-implementations.tex
Add 7.1 on Haskell as design language
Minor fixes and text tweaks, mostly for nicer formatting

@@
 
-\subsection{Concrete Implementations}
+\subsection{Concrete Implementations} \label{sec-concrete-impl}
 
@@
 conventional scheduling of statically known directed acyclic graphs
-with constructive traces~\cite{esfahani2016cloudbuild}. Note that
-we need to convert a monadic \hs{ctRebuilder} into an applicative one by
-applying an adapter function \hs{adaptRebuilder}, which unwraps a given
-\hs{Task}~\hs{Applicative} and wraps it into \hs{Task}~\hs{Monad}.
+with constructive traces~\cite{esfahani2016cloudbuild}. We convert a
+monadic \hs{ctRebuilder} into an applicative one by applying an adapter
+\hs{adaptRebuilder}, which unwraps a given \hs{Task}~\hs{Applicative}
+and wraps it into \hs{Task}~\hs{Monad}.
 

### 2019-10-17 papers/jfp/7-experience.tex
Typo
Minor fixes and text tweaks, mostly for nicer formatting

@@
 Haskell had a profound effect on our thinking. It forced misconceptions
-to to surface early.  It required us to be explicit about side effects.
+to surface early.  It required us to be explicit about side effects.
 It gave us a huge incentive to design abstractions that had simple types
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-data Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
+newtype Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
 \end{minted}
@@
 It is also worth noting that we needed a rather expressive language,
-to faithfully express the abstractions that seen natural in this
+to faithfully express the abstractions that seem natural in this
 setting.  In the case of \hs{Task} we use a data constructor with a
-polymorphic field; a higher-kinded type variable \hs{f :: * -> *}; an
-even more abstracted kind for \hs{c :: (* -> *) -> Constraint}; and,
+polymorphic field; a higher-kinded type variable \hs{f}~\hs{::}~\hs{*}~\hs{->}~\hs{*}; an
+even more abstracted kind for \hs{c}~\hs{::}~\hs{(*}~\hs{->}~\hs{*)}~\hs{->}~\hs{Constraint}; and,
 of course, type classes.
@@
 
-\subsection{Experience from building GHC with \Shake}
+\subsection{Experience from building GHC with \Shake}\label{sec-hadrian}
 

### 2019-10-17 papers/jfp/8-engineering.tex
Minor fixes and text tweaks, mostly for nicer formatting

@@
 
-Overall we choose not to deal with cyclic dependencies -- a choice that most build
-systems also follow. There are computation frameworks that support tasks with
-cyclic dependencies under the assumption that tasks are \emph{monotonic} in a
-certain sense~\cite{pottier2009lazy,radul2009propagation}.
+We choose not to deal with cyclic dependencies -- a choice that most build
+systems also follow. There are computation frameworks that support dependency
+cycles under the assumption that tasks are \emph{monotonic} in a certain
+sense~\cite{pottier2009lazy,radul2009propagation}.
 
@@
 
-newtype TaskT c k v = TaskT@\,@{@\,@runT :: @\std{forall}@ f. c f => Fetch k f -> f v@\,@}
+newtype TaskT c k v = TaskT (@\std{forall}@ f. c f => Fetch k f -> f v)
 
@@
 
-\subsection{File watching vs polling}\label{sec-integration-with-ides}
+\subsection{File watching vs polling}\label{sec-file-watching}
 

### 2019-10-19 papers/jfp/2-background.tex
Work on S1-3

@@
   the sake of simplicity. In reality, there are multiple versions of \cmd{gcc}.
-  To account for this, the actual binary for \cmd{gcc} is often listed as a dependency, along
-  with any supporting binaries, or standard libraries (such as \cmd{stdio.h}),
-  that are used by \cmd{gcc}.
+  To account for this, the actual binary for \cmd{gcc} is often also listed as a
+  dependency, along with any supporting binaries, or standard libraries (such as
+  \cmd{stdio.h}), that are used by \cmd{gcc}.
 } the command
@@
 graph is \emph{acyclic} then each task needs to be executed at most once. Cyclic
-task dependencies are typically not allowed in build systems, although there are rare
-exceptions, see~\S\ref{sec-iterative-compute}.
+task dependencies are typically not allowed in build systems, although there are
+rare exceptions, see~\S\ref{sec-iterative-compute}.
 
-\noindent
-The following property is essential for build systems; indeed, it is their
-raison d'\^etre:
+The fewer tasks are executed in a partial rebuild, the better. To be more
+specific, consider the following property, which is essential for build systems;
+indeed, it is their raison d'\^etre:
 
@@
 \vspace{2mm}
+
+\noindent
+This property is tightly linked to build system \emph{correctness}, which we
+will be ready to define in~\S\ref{sec-build-correctness}; for now, we will use
+minimality as the guiding principle when exploring the design space of build
+systems.
 
@@
 
-\vspace{-1mm}
 \subsection{\Shake: Dynamic Dependencies with No Remorse}
@@
 dependencies~\cite{mitchell2012shake} without sacrificing
-the minimality requirement. Building on the \Make example
+the minimality requirement.
+
+Building on the \Make example
 from~\S\ref{sec-background-make}, we add the following files whose
@@
 \Shake's implementation is different from both \Make and \Excel in two aspects.
-First, to decide which files need to be rebuilt,
-it uses the dependency graph that is dynamically constructed during
-the previous build. This idea has a long history, going back to
-\emph{incremental}~\cite{demers1981incremental},
-\emph{adaptive}~\cite{acar2002adaptive}, as well as
-\emph{self-adjusting computations} -- see
+First, to decide which files need to be rebuilt, it stores the \emph{dependency
+graph} that is constructed during the previous build (\mbox{instead of
+just file modification times or a linear chain}). This idea has a long history,
+going back to \emph{incremental}~\cite{demers1981incremental},
+\emph{adaptive}~\cite{acar2002adaptive}, and
+\emph{self-adjusting computations}~--~see
 Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
@@
 
-\Shake also supports the \emph{early cutoff optimisation}. When it
-executes a task and the result is unchanged from the previous build, it is
-unnecessary to execute the dependent tasks, and hence \Shake can stop a build
-earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
-early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
-not; see~\S\ref{sec-dirty-bit} for an explanation of why.
+\Shake also supports the \emph{early cutoff optimisation}, which is illustrated
+in Fig.~\ref{fig-cutoff}. When it executes a task and the result is unchanged
+from the previous build, it is unnecessary to execute the dependent tasks, and
+hence \Shake can stop a build earlier. Not all build systems support early
+cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do not;
+see~\S\ref{sec-dirty-bit} for an explanation of why.
 
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 \caption{Download sources}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 \caption{Build \cmd{main.exe}}
@@
 \centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
-\vspace{-0.5mm}
+% \vspace{-0.5mm}
 \caption{Modify \cmd{util.c} and rebuild}
 \end{subfigure}
-\vspace{-1mm}
+% \vspace{-1mm}
 \caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
@@
 rectangles.\label{fig-bazel}}
-\vspace{-1mm}
+% \vspace{-1mm}
 \end{figure}
@@
 task multiple times as new dependencies are discovered and rebuilt, but it
-supports the early cutoff optimisation.
+supports the early cutoff optimisation. Note that in practice the cost of
+duplicate work due to the use of a restarting scheduler may often be just a
+small fraction of the overall build cost~(\S\ref{sec-suspending}).
 
@@
 dependencies; then the content-addressable cache allows the engine
-to download the result (if needed) based on the result hash. A concrete
-implementation is provided in~\S\ref{sec-implementation-cloud}.
+to download the result (if needed) based on the result hash. Further details
+and a concrete implementation will be provided
+in~\S\ref{sec-constructive-traces} and~\S\ref{sec-implementation-cloud}.
 
@@
 \begin{table}
-% \vspace{-2mm}
+\vspace{2mm}
 \caption{Build system differences.\label{tab-summary}}
-% \vspace{-4mm}
+\vspace{-1mm}
 % \smaller
@@
 $^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
-% \vspace{-8mm}
+\vspace{1mm}
 \end{table}

### 2019-10-19 papers/jfp/3-abstractions.tex
Work on S1-3

@@
 
+\vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{text}
@@
 cell, it returns the \hs{Task} to compute the value of the formula. Every
-\hs{Task} is given a \emph{callback} \hs{fetch} to find the value of any keys on
-which it depends.
+task is given a \emph{callback} \hs{fetch} to find the value of any keys on
+which it depends. To run a \hs{Task}, we simply apply the function it holds to a
+callback (see the definition of the function \hs{run} in Fig.~\ref{fig-types}).
 
@@
 \end{minted}
-\vspace{1mm}
+\vspace{0.5mm}
 \begin{minted}[fontsize=\small]{haskell}
@@
 \end{minted}
-
 % \vspace{-3mm}
@@
 target key is up to date. What exactly does ``up to date'' mean?  We answer
-that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:
+that precisely in \S\ref{sec-build-correctness}.
+
+\noindent
+Here is a simple build system:
 

### 2019-10-19 papers/jfp/3-abstractions.tex
Polish on S3-4

@@
 which it depends. To run a \hs{Task}, we simply apply the function it holds to a
-callback (see the definition of the function \hs{run} in Fig.~\ref{fig-types}).
+suitable callback (see the definition of the function \hs{run} in
+Fig.~\ref{fig-types}).
 
@@
 \noindent
-The \hs{busy} build system defines the callback \hs{fetch} so that, when given a
-key, it brings the key up to date in the store, and returns its value.
-The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
-Fig.~\ref{fig-stdlib} -- initialised with the incoming \hs{store} by \hs{execState}.
-To bring a key up to date, \hs{fetch} asks the task description \hs{tasks} how
-to compute the value of \hs{k}. If \hs{tasks} returns \hs{Nothing} the key is an
+The \hs{busy} build system defines the callback \hs{fetch} that, when given a
+key, brings the key up to date in the store, and returns its value.
+The function \hs{fetch} runs in the standard~Haskell \hs{State} monad
+(Fig.~\ref{fig-stdlib}) initialised with the incoming \hs{store} by \hs{execState}.
+To bring a key~\hs{k} up to date, \hs{fetch} asks the task description \hs{tasks}
+how to compute its value. If \hs{tasks} returns \hs{Nothing} the key is an
 input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
@@
 key/value mapping in the store, and returns \hs{v}. Notice that \hs{fetch}
-passes itself to \hs{task} as an argument, so that the latter can use \hs{fetch}
+passes itself to \hs{task} as an argument, so the latter can use \hs{fetch}
 to recursively find the values of \hs{k}'s dependencies.
@@
 As we can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
-correct order (if it had built \cmd{B2} before building \cmd{B1}, the result would
-have been $20 * 2 = 40$ instead of $(10 + 20) * 2 = 60$). As an example showing
-that \hs{busy} is not minimal, imagine that the formula in cell \cmd{B2} was
-\cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
+correct order (if it had built \cmd{B2} before building \cmd{B1}, the result
+would have been $20 * 2 = 40$ instead of $(10 + 20) * 2 = 60$). As an example
+showing that \hs{busy} is not minimal, imagine that the formula in cell \cmd{B2}
+was \cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
 \hs{fetch}~\hs{"B1"} twice -- once per occurrence of \cmd{B1} in the
@@
 
+To avoid the recomputation, \hs{busy} can keep the set of processed keys in the
+state monad (in addition to the \hs{store}), treat processed keys as inputs in
+the \hs{getValue} branch of the \hs{fetch} callback, and include the key~\hs{k}
+into the set of processed keys in the \hs{putValue} branch. This eliminates
+unnecessary work \emph{within a single build}, but the next build needs to
+recursively recompute all target's dependencies again even if no inputs changed.
+To save work \emph{between builds}, it is necessary to store some build
+information~\hs{i} persistently.
+
 \subsection{The Need for Polymorphism in Task}\label{sec-why-polymorphism}
-% \vspace{-0.5mm}
 
 The previous example illustrates why the \hs{Task} abstraction is polymorphic
-in~\hs{f}; recall its definition from~\S\ref{sec-task}:
+in~\hs{f}. Recall its definition from~\S\ref{sec-task}:
 
-\begin{minted}[xleftmargin=5pt]{haskell}
+\begin{minted}[xleftmargin=10pt]{haskell}
 newtype Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
@@
 
-We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
-Given \emph{one} task description \cmd{T}, we want to explore \emph{many} build
+We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}. Given
+\emph{one} task description \cmd{T}, we want to explore \emph{many} build
 systems that can build \cmd{T} -- and we will do so in
@@
 task descriptions in a very interesting, and practically useful, way:
+
 \begin{itemize}
-\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
+\item \hs{Task}~\hs{Applicative}: In \hs{sprsh1} we needed only \hs{Applicative}
   operations, expressing the fact that the dependencies between cells can be
@@
   ``computing'' them (see \S\ref{sec-deps}).
-\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
+\item \hs{Task}~\hs{Monad}: As we shall see in \S\ref{sec-task-monad}, a monadic
   task allows \emph{dynamic} dependencies, in which a formula may depend on cell
   \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
-  \cmd{D}.
+  \cmd{D}. A simple example of a task with dynamic dependencies is \Excel
+  formula \cmd{INDIRECT("A" \& C1)} from~\S\ref{sec-background-excel}.
 \item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
@@
   mechanism to decide which input keys changed since the previous build.
-\item \hs{Task}~\hs{MonadFail} correspond to monadic tasks that may fail. For
+\item \hs{Task}~\hs{Selective} corresponds to task descriptions with
+  \emph{conditional statements}, e.g. \Excel formula \cmd{IF(C1=1,B2,A2)},
+  where it is possible to statically \emph{over-approximate} the set of task
+  dependencies. Here \hs{Selective} is a type class of \emph{selective
+  applicative functors}~\cite{mokhov_selective_2019}, which allows us to model
+  build systems like \Dune~\cite{dune} using the presented framework.
+\item \hs{Task}~\hs{MonadFail} corresponds to monadic tasks that may fail. For
   example, the formula \cmd{A1/A2} may fail due to division by zero. We will
   discuss this in~\S\ref{sec-failures}.
-\item \hs{Task}~\hs{MonadPlus}, \hs{Task}~\hs{MonadRandon} and their
+\item \hs{Task}~\hs{MonadPlus}, \hs{Task}~\hs{MonadRandom} and their
   variants can be used for describing tasks with a certain type of
@@
 place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
-\hs{fetch} \emph{different keys} depending on whether or not \hs{c1}~\hs{==}~\hs{1}.
+\hs{fetch} \emph{different keys} depending on whether or not
+\hs{c1}~\hs{==}~\hs{1}. Note that in this example one can statically determine
+the sets of possible dependencies of the formulae \cmd{IF(C1=1,B2,A2)} and
+\cmd{IF(C1=1,A1,B1)} but this cannot be done in general -- recall the
+spreadsheet with the formula \cmd{INDIRECT("A" \& C1)}
+from~\S\ref{sec-background-excel}, where the argument of the \cmd{INDIRECT}
+function is a string computed dynamically during the build. Such tasks can also
+be captured using \hs{Tasks}~\hs{Monad}:
 
+\vspace{1mm}
+\begin{minted}[xleftmargin=10pt]{haskell}
+sprsh3 :: Tasks Monad String Integer
+sprsh3 "B1" = Just $ Task $ \fetch -> do
+    c1 <- fetch "C1"
+    fetch ("A" ++ show c1)
+sprsh3 _ = Nothing
+\end{minted}
+\vspace{1mm}
+
+\noindent
 Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
@@
 we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
-Here is our first use of polymorphism in \hs{f}, discussed in~\S\ref{sec-why-polymorphism}.
+This is another use of polymorphism in \hs{f}, discussed in~\S\ref{sec-why-polymorphism}.
 The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
@@
 
-The presented definition of correctness needs to be adjusted for
-non-deterministic tasks and shallow cloud builds, as will be discussed
-in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects},
+The presented definition of correctness needs to be adjusted for build systems
+that support non-deterministic tasks and shallow cloud builds, as will be
+discussed in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects},
 respectively.
@@
 of \Excel) looking at the syntax tree of the formula.  But a task description
-has no such syntax tree: as you can see in the definition of \hs{Task} in Fig.~\ref{fig-types},
-a task is just a function, so all we can do is call it.
+has no such syntax tree: as you can see in the definition of \hs{Task} in
+Fig.~\ref{fig-types}, a task is just a function, so all we can do is call it.
 Yet, remarkably, we can use the polymorphism of a
@@
 \hs{dependencies} does:
+
 \vspace{1mm}
@@
 
+Here is an implementation of \hs{track} based on the standard Haskell
+\hs{WriterT} \emph{monad transformer}~\cite{liang1995monad}, whose main
+types are listed in Fig.~\ref{fig-stdlib}:
+
 \vspace{1mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
-import Control.Monad.Writer
-
 track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [(k, v)])
@@
 \noindent
-This function uses the standard Haskell \hs{WriterT} \emph{monad
-transformer}~\cite{liang1995monad} for recording additional
+This function uses the \hs{WriterT} transformer for recording additional
 information~--~a list of key/value pairs \hs{[(@@k,}~\hs{v)]}~--~when executing
-a task in an arbitrary monad~\hs{m} (see the types in Fig.~\ref{fig-stdlib}).
-We substitute the given \hs{fetch} with a
+a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
 \hs{trackingFetch} that, in addition to fetching a value, tracks the
@@
 \hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
-with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
-example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}, by defining
-a corresponding \hs{fetchIO} of type \hs{String}~\hs{->}~\hs{IO}~\hs{Integer},
-which allows us to demonstrate the dynamic nature of monadic dependencies in
-GHCi.
+with dynamic dependencies, see~\S\ref{sec-implementation-shake}.
+
+Here we show an example of \hs{track}ing monadic tasks when
+\hs{m}~\hs{=}~\hs{IO}, by defining a corresponding \hs{fetchIO} of type
+\hs{String}~\hs{->}~\hs{IO}~\hs{Integer}, which allows us to demonstrate the
+dynamic nature of monadic dependencies in GHCi.
 
@@
 
-\noindent
 Here the keys $n < 2$ are input parameters, and one can obtain the usual
@@
 unnecessarily; all of its 19,729 decimal digits have already been helpfully
-computed, e.g. see~\cite{ackermann42}. We will use \hs{MonadState} tasks in our
-models of cloud build systems in~\S\ref{sec-implementations}.
+computed, e.g. see~Kosara~\shortcite{ackermann42}. We will use \hs{MonadState}
+tasks in our models of cloud build systems in~\S\ref{sec-implementations}.

### 2019-10-19 papers/jfp/4-schedulers.tex
Polish on S3-4

@@
 that decide which tasks to execute and in what order; see the ``Scheduler''
-column of Table~\ref{tab-summary} in \S\ref{sec-background-summary}. This
-subsection explores the properties of the three schedulers, and possible
-implementations.
+column of Table~\ref{tab-summary} in \S\ref{sec-background-summary}. The
+following subsections explore the properties of the three schedulers, and
+possible implementations.
 
@@
 that order. Computing such a linear order is straightforward -- given a task
-description and output \hs{key} first find the (acyclic) graph of the \hs{key}'s
-dependencies, then compute a topological order. Taking the \Make example from
-Fig.~\ref{fig-make}, we might compute the order:
+description and a target \hs{key}, first find the (acyclic) graph of the
+\hs{key}'s dependencies, then compute a topological order. Taking the \Make
+example from Fig.~\ref{fig-make}, we might compute the order:
 
 \begin{enumerate}
+\item \cmd{main.o}
 \item \cmd{util.o}
-\item \cmd{main.o}
 \item \cmd{main.exe}
@@
 \noindent
-Given the dependencies, we could have equally chosen to build \cmd{main.o}
+Given the dependencies, we could have equally chosen to build \cmd{util.o}
 first, but \cmd{main.exe} \emph{must} come last.
@@
 The advantage of this scheme is simplicity -- compute an order, then execute
-tasks in that order. In addition any missing nodes or cycles can be detected
-from the graph, and reported to the user before any work has commenced.
+tasks in that order. In addition, any missing keys or dependency cycles can be
+detected from the graph, and reported to the user before any work has commenced.
 
@@
 interleave running tasks and ordering tasks. One approach is just to build tasks
-in an arbitrary order, but if a task calls \hs{fetch} on an out-of-date key
+in an arbitrary order, and if a task calls \hs{fetch} on an out-of-date key
 \cmd{dep}, abort the task and build \cmd{dep} instead. Returning to the example
-from Fig.~\ref{fig-make}, we might build in the order:
+from Fig.~\ref{fig-make}, we might build the tasks in the order:
 
 \begin{enumerate}
-\item \cmd{main.exe} (abort because it requires \cmd{util.o})
+\item \cmd{main.exe} (abort because it depends on \cmd{util.o} which is out of
+      date)
+\item \cmd{main.o}
 \item \cmd{util.o}
-\item \cmd{main.o}
 \item \cmd{main.exe} (restart from scratch, completing successfully this time)
@@
 \noindent
-We start with \cmd{main.exe} (chosen randomly), but discover it requires
-\cmd{util.o}, so instead start building \cmd{util.o}. Next we choose to build
-\cmd{main.o} (again, randomly), before finally returning to \cmd{main.exe},
-which now has all its dependencies available and completes without error.
+We start with \cmd{main.exe} (chosen randomly), but discover it depends on
+\cmd{main.o}, so instead start building \cmd{main.o}. Next we choose to build
+\cmd{util.o} (again, randomly), before finally returning to \cmd{main.exe} that
+now has all its dependencies available and completes successfully.
 
 The approach works, but has a number of disadvantages. Firstly, it requires a
-technical mechanism to abort a task -- easy enough in our theoretical setting of
-\hs{Task} (see the implementation in~\S\ref{sec-implementation-excel}) but leads
-to engineering concerns in the real world. Secondly, it is not minimal in the
-sense that a task may start, do some meaningful work, and then abort, repeating
-that same work when restarted.
+technical mechanism to abort a task, which is easy in our theoretical setting
+with \hs{Task} (see an implementation in~\S\ref{sec-implementation-excel}) but
+leads to engineering concerns in the real world. Secondly, it is not minimal in
+the sense that a task may start, do some meaningful work, and then abort,
+repeating that same work when restarted.
 
@@
 (\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
-they are requested, suspending the currently running task. Using the example
-from Fig.~\ref{fig-make}, we would build:
+they are requested, suspending the currently running task when needed. Using the
+example from Fig.~\ref{fig-make}, we would build:
 
@@
 \item \cmd{main.exe} (suspended)\\
+  $\hookrightarrow$ \cmd{main.o}
+\item \cmd{main.exe} (resumed then suspended again)\\
   $\hookrightarrow$   \cmd{util.o}
-\item \cmd{main.exe} (resumed then suspended again)\\
-  $\hookrightarrow$   \cmd{main.o}
 \item \cmd{main.exe} (completed)
@@
 \noindent
-As we see, \cmd{main.exe} is started first as it is the required target. It soon
-discovers a dependency on \cmd{util.o} and suspends to build \cmd{util.o},
-resumes a build before building \cmd{main.o} then continues its work to finish
-\cmd{main.exe}.
+We start building \cmd{main.exe} first as it is the required target. We soon
+discover a dependency on \cmd{main.o} and suspend the current task
+\cmd{main.exe} to build \cmd{main.o}, then resume and suspend again to build
+\cmd{util.o}, and finally complete the target \cmd{main.exe}.
 
 This scheduler (when combined with a suitable rebuilder) provides a minimal
-build system which supports dynamic dependencies. In our model, a suspending
+build system that supports dynamic dependencies. In our model, a suspending
 scheduler is easy to write -- it makes a function call to compute each
@@
 passing is efficient, but requires the build script to be architected to allow
-capturing continuations. \Shake currently uses continuations.
+capturing continuations. \Shake currently uses this approach.
 \end{itemize}
@@
 \noindent
-While a suspending scheduler is theoretically optimal, it is only better in
-practice than a restarting scheduler if the cost of avoided duplicate work
+While a suspending scheduler is theoretically optimal, in practice it is better
+than a restarting scheduler only if the cost of avoided duplicate work
 outweighs the cost of suspending tasks. Note furthermore that the cost of

### 2019-10-19 papers/jfp/8-engineering.tex
Polish on S3-4

@@
 This approach is convenient if \emph{tasks are not aware of failures}, e.g. we
-can model \Excel formulas as pure arithmetic functions, and introduce failures
+can model \Excel formulae as pure arithmetic functions, and introduce failures
 ``for free'' if/when needed by instantiating \hs{Tasks} with an appropriate
@@
 
-sprsh3 :: Tasks MonadFail String Integer
-sprsh3 "B1" = Just $ Task $ \fetch -> do
+sprsh4 :: Tasks MonadFail String Integer
+sprsh4 "B1" = Just $ Task $ \fetch -> do
     a1 <- fetch "A1"
@@
     if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
-sprsh3 _ = Nothing
+sprsh4 _ = Nothing
 \end{minted}
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh4 :: Tasks MonadRandom String Integer
-sprsh4 "B1" = Just $ Task $ \fetch -> do a1 <- fetch "A1"
+sprsh5 :: Tasks MonadRandom String Integer
+sprsh5 "B1" = Just $ Task $ \fetch -> do a1 <- fetch "A1"
                                          r  <- getRandom (1,2)
                                          return (a1 + r)
-sprsh4 _    = Nothing
+sprsh5 _    = Nothing
 \end{minted}
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-sprsh5 "B1" = Just $ Task $ \fetch -> do
+sprsh6 "B1" = Just $ Task $ \fetch -> do
     formula <- fetch "B1-formula"
@@
 The function \hs{selfTracking} defined below is a generalisation of the approach
-explained in the above \Excel example \hs{sprsh5}. The function takes a parser
+explained in the above \Excel example \hs{sprsh6}. The function takes a parser
 for scripts, of type \hs{s}~\hs{->}~\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}, and a
 description of \emph{how to build all scripts}, of type
-\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{s}. For \hs{sprsh5}, the latter would simply
+\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{s}. For \hs{sprsh6}, the latter would simply
 fetch \cmd{B1-formula} when given the key \cmd{B1} and return \hs{Nothing}

### 2019-10-20 papers/jfp/5-rebuilders.tex
Polish S5-7.

@@
 
+\vspace{-1mm}
 \subsubsection{Trace Representation}\label{sec-trace-representation}
@@
 
+\vspace{-1mm}
 \subsubsection{Verifying Step Traces}\label{sec-step-traces}
@@
 The \Shake build system and the associated paper -- see~\S2.3.3
-in~Mitchell~\shortcite{mitchell2012shake} -- use a slightly different trace
-structure, called \emph{verifying step traces},
-which stores less data than verifying traces, and has slightly different early cutoff semantics. Rather than
+in~Mitchell~\shortcite{mitchell2012shake} -- use a different trace structure,
+called \emph{verifying step traces}, which stores less data than verifying
+traces, and has slightly different early cutoff semantics. Rather than
 storing the \hs{Hash}~\hs{v} for each dependency, it instead stores \hs{built}
@@
 \hs{Int} taking only~4~bytes. Furthermore, \Shake permits values to be
-arbitrarily large, and permits a custom equality (two values can be bit-for-bit
-unequal but considered equal by \Shake), meaning that a \hs{Hash} is not a valid
-encoding. For build systems that use \hs{Applicative} dependencies, the
-\hs{depends} could be omitted entirely, making the size of a \hs{StepTrace}
-$O(1)$ instead of $O(n)$, where $n$ is the number of dependencies.
+arbitrarily large, and supports a custom value equality (two values can be
+bit-for-bit unequal but considered equal by \Shake), hence \hs{Hash}~\hs{v} is
+not a valid encoding. For applicative tasks, \hs{depends} can be omitted, making
+the size of a \hs{StepTrace} $O(1)$ instead of $O(n)$, where $n$ is the number
+of dependencies.
 
@@
 
-A verifying trace deliberately records only hashes, so that it can be small. In
-contrast, a \emph{constructive} trace also stores the resulting value.
+A verifying trace records only hashes or time stamps, so that it can be small.
+In contrast, a \emph{constructive} trace also stores the resulting value.
 Concretely, it records a list of \hs{Trace}~\hs{k}~\hs{v}~\hs{v}, where
@@
 per key, and to share them with other users, providing cloud-build
-functionality. We can represent this additional information with the operations:
+functionality. We can access this additional information with these operations:
 
@@
 The function \hs{recordCT} is similar to \hs{recordVT}
-from~\S\ref{sec-verifying-traces}, but instead of just passing
-the hash of the resulting value, we pass the actual value. The \hs{verifyVT}
+from~\S\ref{sec-verifying-traces}, but instead of just passing the hash of the
+resulting value, we pass the actual value. The function \hs{verifyVT}
 has been replaced with \hs{constructCT}, which instead of taking the hash of the
@@
 \CloudBuild~(\S\ref{sec-implementation-cloud}), can index directly from the
-key and the hashes of its dependencies to the output result, i.e.
+key and the hashes of its dependencies to the resulting value, e.g. using a
 \hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
@@
 Current build systems using deep constructive traces always record hashes of
-terminal \emph{input keys}, but the technique works equally well if we skip any
-number of dependency levels (say $n$ levels). The input-only approach is the
+terminal input keys, but the technique also works if we skip any number
+of dependency levels (say $n$ levels). The input-only approach is the
 special case of $n = \infty$, and constructive traces are the special case of
-$n = 1$. By picking values of $n$ in between we would regain some early cutoff, at the
-cost of losing such simple shallow builds, while still requiring deterministic tasks.
+$n = 1$. By picking values of $n$ in between we would regain some early cutoff,
+at the cost of losing such simple shallow builds, while still requiring
+determinism.

### 2019-10-20 papers/jfp/6-implementations.tex
Polish S5-7.

@@
-\section{Build systems, concretely}\label{sec-implementations}\label{sec-design-space}
+\section{Build Systems, Concretely}\label{sec-implementations}\label{sec-design-space}
 
 \begin{table}
+\vspace{2mm}
 \caption{Build systems \`a la carte.\label{tab-build-systems}}
-\vspace{-2mm}
+\vspace{-3mm}
 % \smaller
@@
 and a rebuilder. The result can be summarized in Table~\ref{tab-build-systems},
-which tabulates combinations of the scheduling algorithm and the rebuilding
-strategy, providing 12~possible build systems, 8~of which are inhabited
-by existing build systems (we discuss these systems in~\S\ref{sec-background} and
-\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
-build systems. The most interesting unfilled spot in the table is suspending
-constructive traces, which would provide many benefits, and which we title
-\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
-% (as we plan on extending \Shake to occupy that spot)
+which tabulates 12 possible combinations, 8~of which are inhabited
+by existing build systems (we discuss these systems in~\S\ref{sec-background}
+and~\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
+build systems. The most interesting unfilled spot in the table corresponds to
+a suspending scheduler composed with a constructive trace rebuilder. Such a
+build system would provide many benefits; we title it \Cloud \Shake and explore
+further in~\S\ref{sec-implementation-cloud}.
 
+\vspace{-1mm}
 \subsection{Concrete Implementations} \label{sec-concrete-impl}
@@
 
-The scheduler calls the rebuilder, but passes it a \hs{fetch} function
-that the latter calls when it needs the value of a dependent key.  This
-callback returns control to the scheduler, which may in turn call the
-rebuilder to bring the dependent key up to date, and so on.
+The scheduler runs the \hs{Task} returned by the rebuilder passing it a
+\hs{fetch}~callback~that the task uses to find values of its dependencies.
+The callback returns control to \mbox{the scheduler}, which may in turn call
+the rebuilder again to bring another key up to date, and so on.
 
@@
 tasks, which restricts \Make to static dependencies, as reflected in the
-type~\hs{Build}~\hs{Applicative}. Moreover, any other build system that uses
-the \hs{topological} scheduler will also inherit the same restriction.
+type~\hs{Build}~\hs{Applicative}. Any other build system that uses
+the \hs{topological} scheduler will inherit the same restriction.
 
@@
             fetch :: k -> State ir (Either k v)
-            fetch k | k `Set.member` done = return $ Right (getValue k store)
+            fetch k | Set.member k done = return $ Right (getValue k store)
                     | otherwise         = return $ Left k
@@
 
-After the \hs{newTask} is executed (with the help of \hs{liftStore}) there are
-two cases to consider:
+After the \hs{newTask} is executed (using \hs{liftStore}), there are two cases
+to consider:
 
@@
 Note that the task returned by the \hs{rebuilder} expects a total callback
-function and cannot be directly executed with the partial callback \hs{fetch}.
-We fix the mismatch with the function \hs{try} that relies on the standard
-monad transformer \hs{ExceptT} from the \cmd{transformers} library. We also
-need the helper \hs{liftChain}, whose implementation we omit since it
-is analogous to \hs{liftStore} in Fig.~\ref{fig-make-implementation}.
+function and cannot be executed with the partial callback \hs{fetch}. We fix the
+mismatch with the function \hs{try} that relies on the standard monad
+transformer \hs{ExceptT}~\cite{liang1995monad}. The helper \hs{liftChain} is
+analogous to \hs{liftStore} in Fig.~\ref{fig-make-implementation}, so we omit
+its implementation.
 
@@
         case tasks key of
-            Just task | key `Set.notMember` done -> do
+            Just task | Set.notMember key done -> do
                 value <- gets (getValue key . fst)
@@
 \end{minted}
-% \vspace{-1mm}
+\vspace{-1mm}
 \caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
-% \vspace{-3mm}
+\vspace{-3mm}
 \end{figure}
@@
 
-As discussed in \S\ref{sec-step-traces}, \Shake actually use verifying step traces, but we
-choose to focus on the more explicit verifying traces. We have implemented
-verifying step traces in our framework, and they compose with schedulers as
-you would hope.
+As discussed in \S\ref{sec-step-traces}, \Shake actually use verifying step
+traces, but here we choose to focus on the more explicit verifying traces. We
+have implemented verifying step traces in our framework, and they compose with
+schedulers as you would hope.
 
@@
 \hs{dctRebuilder}, whose implementation we omit since it is very similar to that
-of \hs{ctRebuilder}. \Buck uses the \hs{topological} scheduler and is therefore
-an applicative build system, whereas \Nix uses the \hs{suspending} scheduler and
-is monadic.
+of \hs{ctRebuilder}. \Buck uses the \hs{topological} scheduler and is an
+applicative build system, whereas \Nix uses the \hs{suspending} scheduler and
+is therefore monadic.
 

### 2019-10-20 papers/jfp/7-experience.tex
Polish S5-7.

@@
 
-We have described a framework that can describe, and indeed execute in
+We have presented a framework that can describe, and indeed execute in
 prototype form, a wide spectrum of build systems.  But our ultimate
-goal is a practical one: to use these insights to construct a better
-build system.  Our earlier work on the \Shake build
-system~\cite{mitchell2012shake}, and applying \Shake to
-building GHC~\cite{hadrian}, makes progress in that direction.
+goal is a practical one: to use these insights to develop better
+build systems. Our earlier work on \Shake~\cite{mitchell2012shake}, and applying
+\Shake to building GHC~\cite{hadrian}, makes progress in that direction.
 
@@
 both early cutoff and monadic dependencies~(\S\ref{sec-implementation-cloud}),
-and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
-In this section we reflect on our experience of turning theory into practice.
+and used it to implement GHC's (very substantial) build system,
+\Hadrian~\cite{hadrian}. In this section we reflect on our experience of turning
+theory into practice.
 
-\subsection{Haskell as a design language}\label{sec-design-lang}
+\vspace{-1mm}
+\subsection{Haskell as a Design Language}\label{sec-design-lang}
 
@@
 even more abstracted kind for \hs{c}~\hs{::}~\hs{(*}~\hs{->}~\hs{*)}~\hs{->}~\hs{Constraint}; and,
-of course, type classes.
+of course, type classes. Our models have been translated to
+Rust~\cite{translation_rust} and Kotlin~\cite{translation_kotlin}, and in both
+cases there was a loss of precision due to language-specific limitations.
 
-Moreover, when thinking about the type constructors over which \hs{f} might
-usefully range, it turned out that we could adopt the \emph{existing} abstractions
+When thinking about the type constructors over which \hs{f} might usefully
+range, it turned out that we could adopt the \emph{existing abstractions}
 of \hs{Functor}, \hs{Applicative}, \hs{Monad} and so on.
 That in turn led us to think in a new way about the taxonomy
-of build systems -- see \S~\ref{sec-why-polymorphism}.
+of build systems -- see~\S\ref{sec-why-polymorphism}. In the other direction,
+trying to express an \emph{existing build system} \Dune in our models led us to
+finding a new abstraction -- the \hs{Selective} \mbox{type
+class~\cite{mokhov_selective_2019}}, which turned out to be useful outside the
+build systems domain.
 
@@
 from \S~\ref{sec-concrete-impl}:
+
+\vspace{0.5mm}
 \begin{minted}[fontsize=\small,xleftmargin=5pt]{haskell}
@@
 \end{minted}
+\vspace{0.5mm}
+
 These powerful and modular abstractions, which ultimately formed part
@@
 project, as we repeatedly reviewed, redesigned, and refactored our
-executable prototype.  It is hard to believe that we could have
+executable prototypes. It is hard to believe that we could have
 developed them without the support of Haskell as a design language.
@@
 subsequent commits to the \Shake
-project\footnote{\url{https://github.com/ndmitchell/shake}}. These commits add
-concepts like \emph{resources} (for handling situations when two build tasks
+project\footnote{See \url{https://github.com/ndmitchell/shake}.}. These commits
+add concepts like \emph{resources} (for handling situations when two build tasks
 contend on a single external resource), rewriting serialisation to be faster,
-documentation including a website\footnote{\url{https://shakebuild.com}}, and
-add a lot of tests. The biggest change in that time period was an implementation
-change: moving from
+documentation including a website\footnote{See \url{https://shakebuild.com}.},
+and add a lot of tests. The biggest change in that time period was an
+implementation change: moving from
 blocking threads to continuations for the suspending scheduler.
-But otherwise, almost all external and internal details remain the same\footnote{
-  The most visible change is purely notational: switching from \hs{*>} to \hs{@\%@>}
-  for defining rules, because a conflicting \hs{*>} operator was added to the
-  Haskell \hs{Prelude}.}.
+But otherwise, almost all external and internal details remain the
+  same\footnote{The most visible change is purely notational: switching from
+  \hs{*>} to \hs{@\%@>} for defining rules, because a conflicting \hs{*>}
+  operator was added to the Haskell \hs{Prelude}.}.
 We consider the lack of change suggestive that \Shake is based on fundamental
@@
 in the original paper the tasks (there called ``actions'')
-were described in continuation-passing style using the data type:\footnote{
-The original paper uses concrete types \hs{Key} and \hs{Value}. Here we
-generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
-\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}}
-\vspace{1mm}
+were described in continuation-passing style using the data
+type\footnote{The original paper uses concrete types \hs{Key} and \hs{Value}.
+Here we generalise these types to~\hs{k} and~\hs{v}, and also add~\hs{a} so that
+\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}.}:
+\vspace{0.5mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
     Depends ds op >>= f = Depends ds $ \v -> op v >>= f
-
+\end{minted}
+\begin{minted}[xleftmargin=10pt]{haskell}
 toAction :: Task Monad k v -> Action k v v
-toAction (Task run) = run $ \k -> Depends k Finished
-
+toAction (Task run) = run (\k -> Depends k Finished)
+\end{minted}
+\begin{minted}[xleftmargin=10pt]{haskell}
 fromAction :: Action k v v -> Task Monad k v
-fromAction x = Task $ \fetch -> f fetch x
+fromAction x = Task (\fetch -> f fetch x)
   where
@@
 
-By describing these types
+By expressing these types
 using \hs{Task} we are able to describe the differences more concisely
@@
 \begin{minted}[xleftmargin=10pt]{haskell}
-addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]]
-         -> v -> [FilePath] -> IO ()
+addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]] -> v -> [k] -> IO ()
 lookupCloud :: (k -> m (Maybe (Hash v))) -> k -> Ver -> Ver
@@
       be specified simultaneously, so they can all be built in parallel.
-\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
+\item The \hs{addCloud} function also takes a list of keys \hs{[k]}, being the
       files that this rule produces. These produced files include those which
@@
 match, we consult the constructive trace. By bolting constructive traces onto
-the side of Shake we avoid re-engineering of the central database. We have not
+the side of \Shake we avoid re-engineering of the central database. We have not
 found any significant downsides from the bolt-on approach thus far, so it may be
@@
 \Cloud \Shake, we were actually beaten to it by Standard Chartered who
-report\footnote{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}:
+report\footnote{\mbox{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}}:
 
@@
 command line execution. Such information has been very helpful in making the GHC
-build cloud ready.
+build cloud ready~\cite{cloud_hadrian}.
 
@@
 
-A small number of GHC build rules cannot be cached. These rules register new libraries
-in the GHC package database, and rather than producing a output file, they mutate
-a shared file. \Cloud~\Shake provides a way to manually label such
+A small number of GHC build rules cannot be cached. These rules register
+libraries in the package database, and rather than producing files, they
+mutate a shared file. \Cloud~\Shake provides a way to manually label such
 build rules to exclude them from caching.
@@
 unlimited CPUs the speed up would be less than 10\%), and that a handful of
-build tasks (two anomalously slow Haskell compilations, and some calls to slow
-single-threaded \cmd{configure}) are responsible for a significant fraction of
-build time (at least 15\%).
+build tasks (two anomalously slow Haskell compilations, and calls to slow
+single-threaded \cmd{configure}) take up a significant fraction of build time
+(at least 15\%).
 

### 2019-10-20 papers/jfp/8-engineering.tex
Final tweaks

@@
 Recall from~\S\ref{sec-why-polymorphism} that we require tasks to be polymorphic
-in~\hs{f}, and we can therefore choose to execute a \hs{Task} in an~\hs{f} with
+in~\hs{f}, and can therefore choose to execute a \hs{Task} in an~\hs{f} with
 failures, the simplest example being \hs{f}~\hs{=}~\hs{Maybe}. Below we define
-a callback \hs{fetchA1A2} that returns \hs{Just}~\hs{value} for keys \cmd{A1}
-and \cmd{A2} but fails with \hs{Nothing} on all other keys:
+a callback that returns \hs{Just}~\hs{value} for keys \cmd{A1} and \cmd{A2} but
+fails with \hs{Nothing} on all other keys:
 
@@
     fail :: String -> m a
-
+\end{minted}
+\begin{minted}[xleftmargin=10pt]{haskell}
 sprsh4 :: Tasks MonadFail String Integer
@@
 because they are out of date but because of races with earlier tasks that had
-not yet finished. As a consequence, if a the calculation order is persisted over
-successive runs (as \Excel does), potentially racey dependencies will be
+not yet finished. As a consequence, if the build order is persisted over
+successive runs (as in \Excel), potentially racey dependencies will be
 separated, giving better parallelism over time.
@@
 parallelism and machine resource starvation, which is usually achieved with a
-thread pool or thread limit.
+thread pool/limit.
 
-The actual implementation of the parallel schedulers is not overly onerous,
-but neither is it beautiful or informative.
+The actual implementation of the parallel schedulers is not overly onerous, but
+neither is it beautiful or informative, so we omit it.
 
@@
 
-Interestingly, there is significant interplay between all three sources of
+There is significant interplay between all three sources of
 impurity. Systems like \Bazel use various sandboxing techniques to guard against
 missing dependencies, but none are likely to capture all dependencies right down
-to the CPU model and microcode version. Tasks that do have untracked
-dependencies can be marked as volatile, a technique \Excel takes with the
-\cmd{INDIRECT} function, fixing the untracked dependency at the cost of
-minimality.
+to the CPU model and microcode version. Tasks with untracked dependencies can be
+marked volatile, a technique \Excel takes with the \cmd{INDIRECT} function,
+fixing the untracked dependency at the cost of minimality.
 
@@
 definition~(\S\ref{sec-build-correctness}): instead of requiring that the
-resulting value \emph{equals the result} of recomputing the task, we now require
-that resulting value \emph{belongs to the set of possible results} of
+produced value \emph{equals the result} of recomputing the task, we now require
+that produced value \emph{belongs to the set of possible results} of
 recomputing the task, i.e. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
@@
 
-Note that \hs{Task}~\hs{MonadRandom} is powerful enough to model
+Interestingly, \hs{Task}~\hs{MonadRandom} is powerful enough to express
 dependency-level non-determinism, for example,
 \cmd{INDIRECT("A" \& RANDBETWEEN(1,2))}, whereas most build tasks in real-life
-build systems only experience a value-level non-determinism. \Excel handles this
-example simply by marking the cell volatile~--~an approach that can be readily
-adopted by any of our implementations.
+build systems only experience a value-level non-determinism. \mbox{\Excel}
+handles this example simply by marking the cell volatile~--~an approach that can
+be readily adopted by any of our implementations.
 
+\vspace{-1mm}
 \subsection{Cloud Implementation}\label{sec-cloud-aspects}
@@
 
-\item \textbf{Shallow builds}: Building the end result, e.g. an installer
-package, often involves many intermediate tasks. The intermediate results may be
-large, so some cloud build systems are designed to build end products
-\emph{without materialising} intermediate results, producing a so-called
-\emph{shallow build} (see an example in~\S\ref{sec-background-bazel}). Some
-build systems go even further, integrating with the file system to only
-materialise the file when the user accesses it~\cite{gvfs}.
+\item \textbf{Shallow builds}: Building the end target, e.g. an installer
+package, often involves many intermediate tasks. The values produced by these
+intermediate tasks may be large, so some cloud build systems are designed to
+build end targets \emph{without materialising} any intermediate values,
+producing a so-called \emph{shallow build} -- see an example
+in~\S\ref{sec-background-bazel}. Some build systems go even further, integrating
+with the file system to only materialise the file when the user accesses
+it~\cite{gvfs}.
 \end{itemize}
 
-Shallow builds have a slightly weaker correctness property than
+Shallow builds have a slightly weaker correctness property than in the
 Definition~\ref{def-correct}. Instead of demanding that \emph{all} keys reachable
-from the target match, we only require matches for the target itself and the \emph{input
-keys} reachable from the target.
+from the target match, we only require matches for the target itself and the
+\emph{input keys} reachable from the target.
 
@@
 \begin{subfigure}[b]{0.31\linewidth}
-\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-build.pdf}}
+\centerline{\includegraphics[scale=0.27]{fig/frankenbuild-example-build.pdf}}
 % \vspace{-1.5mm}
@@
 \begin{subfigure}[b]{0.31\linewidth}
-\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-clean.pdf}}
+\centerline{\includegraphics[scale=0.27]{fig/frankenbuild-example-clean.pdf}}
 % \vspace{-1.5mm}
@@
 \begin{subfigure}[b]{0.36\linewidth}
-\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-rebuild.pdf}}
+\centerline{\includegraphics[scale=0.27]{fig/frankenbuild-example-rebuild.pdf}}
 % \vspace{-1.5mm}
@@
 \end{subfigure}
-% \vspace{-1.5mm}
-\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
+\vspace{-1mm}
+\caption{Frankenbuild example: (a)~build a human-readable profiling report for
 \cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
-tool, then record deep constructive traces in the cloud, (b)~remove built files
-locally and evict \cmd{main.prof} from the cloud storage, (c)~build
-\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
-the new hash value), then build \cmd{report.txt} by downloading it from the
-matching deep constructive trace in the cloud, resulting in a Frankenbuild
-because \cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted
-cloud storage entries are highlighted; file hashes are shown in circles.
+tool, saving deep constructive traces in the cloud, (b)~remove built files
+locally and evict \cmd{main.prof} from the cloud storage, (c)~rebuild
+\cmd{main.prof} (profiling is non-deterministic, hence a new hash value), then
+build \cmd{report.txt} by downloading it from the matching deep constructive
+trace in the cloud. The result is a Frankenbuild because \cmd{main.prof} and
+\cmd{report.txt} are inconsistent. New and evicted cloud storage entries are
+highlighted; file hashes are shown in circles.
 \label{fig-frankenbuild}}
@@
 
-As described in \S\ref{sec-using-cloud-shake}, non-determinism  (\S\ref{sec-non-determinism})
-is harmful for cloud builds, reducing the number of cache hits.
-However, for deep constructive traces (\S\ref{sec-deep-constructive-traces}) it
-is much worse, even leading to \emph{incorrect results}.
-Fig.~\ref{fig-frankenbuild} shows a
+As described in \S\ref{sec-using-cloud-shake}, non-determinism
+(\S\ref{sec-non-determinism}) is harmful for cloud builds, reducing the number
+of cache hits. However, for deep constructive traces
+(\S\ref{sec-deep-constructive-traces}) it is much worse, even leading to
+\emph{incorrect results}. Fig.~\ref{fig-frankenbuild} shows a
 \emph{Frankenbuild}~\cite{esfahani2016cloudbuild} example, where the target
@@
 immediate dependency \cmd{main.prof}. This inconsistency is caused by two factors:
-(i)~inherent non-determinism of profiling: running a profiling tool on the very
+(i)~inherent non-determinism of profiling -- running a profiling tool on the very
 same \cmd{main.exe} will produce different \cmd{main.prof} results every time;
-and (ii)~relying on deep constructive traces, which cache build results based
-only on the hashes of terminal task inputs (in this case \cmd{main.exe}). Note
-that the resulting store is incorrect according to all three definitions of
-correctness: the main definition~(\S\ref{sec-build-correctness}), the variant
-for non-deterministic tasks~(\S\ref{sec-non-determinism}) and the variant for
-shallow builds~(this section).
+and (ii)~relying on deep constructive traces that cache build results based
+on the hashes of terminal task inputs (in this case \cmd{main.exe}). The result
+violates all three definitions of correctness: the main
+definition~(\S\ref{sec-build-correctness}), the variant for non-deterministic
+tasks~(\S\ref{sec-non-determinism}) and the variant for shallow
+builds~(this section).
 
@@
 
+\vspace{-1mm}
 \subsection{Iterative Computations}\label{sec-iterative-compute}
@@
 
+\vspace{-1mm}
 \subsection{Key-dependent Value Types}\label{sec-polymorphism}
 
-Key-dependent value types allows a single build system to work with multiple different
-types of values, where the type of any particular value is determined by the key. As an example
-of why this might be useful, consider a build system that can depend on files
-(whose contents are strings) or system executables (represented by their version number) --
-using a single type for both values reduces type safety.
-\Shake permits such key-dependent value types, for example the \emph{oracle} rule from
-Mitchell~\shortcite{mitchell2012shake}, and users have remarked that this additional type safety
-provides a much easier expression of concepts, e.g. see~Mokhov~\etal~\shortcite{hadrian}.
+Key-dependent value types allow a build system to work with multiple different
+types of values, where the type of any particular value is determined by the
+key. As an example of why this might be useful, consider a build system where
+keys can be files (whose contents are strings) or system executables
+(represented by their version number) -- using a single type for both values
+reduces type safety. \Shake permits such key-dependent value types, e.g. see
+the \emph{oracle} rule in Mitchell~\shortcite{mitchell2012shake}, and users
+have remarked that this additional type safety provides a much easier expression
+of concepts~\cite{hadrian}.
 
@@
 from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety
-from~\S\ref{sec-polymorphism}, using either pairs or lists for both the key and value:
+from~\S\ref{sec-polymorphism}:
 
@@
 
-The only awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for
+One awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for
 exactly the same \hs{Files} key. If one \hs{File} key swapped the order of the
@@
 
-In the above function we compute the partition, find the index of this key within that
-partition (using \hs{elemIndex}), run the \hs{fetch} on every key in the partition,
-then extract out our specific result (using the indexing operation \hs{!!}).
+In the above function we compute the elementary subset \hs{ks} of the given key
+\hs{k}, find the index of \hs{k} within the subset (using \hs{elemIndex}), run
+the \hs{fetch} to build every key in the subset, then extract out the value
+corresponding to \hs{k} (using the indexing operation \hs{!!}).
 
@@
 tree), and tasks can be compared for equality by comparing those data
-structures. Those which use a full programming language, e.g. \Shake,
+structures. Build systems that use a full programming language, e.g. \Shake,
 are faced with the challenge of implementing equality on arbitrary
@@
 change to the build system potentially changes any build task ~--~ the
-classic example being a makefile depending on itself.
+classic example being build tasks depending on the makefile itself.
 
@@
                                       -> TasksT Monad (Key k v s)
-selfTracking _     _     (Script _) = Nothing -- Scripts are inputs
-selfTracking parse tasks (Value  k) = runScript <$> tasks k
+selfTracking parse tasks key = case key of
+    Script k -> getScript <$> tasks k
+    Value  k -> runScript <$> tasks k
   where
-    -- Fetch the script, parse it, and then run the obtained task
+    -- Build the script and return it
+    getScript :: Task Monad k s -> TaskT Monad (Key k v s) s
+    getScript task = TaskT $ \fetch -> run task (fetch . Script)
+    -- Build the script, parse it, and then run the obtained task
     runScript :: Task Monad k s -> TaskT Monad (Key k v s) v
@@
 prohibitive. To overcome that problem, build systems targeted at large scale,
-e.g. \Bazel and \Buck, make use of file watching API's that detect when a file
-has changed. Using that information the build can avoid performing many file
+e.g. \Bazel and \Buck, rely on file watching API's to detect when a file has
+changed. Using that information the build can avoid performing many file
 accesses.

### 2019-10-20 papers/jfp/9-related.tex
Add a TODO
Say more about Redo vs Shake
Final tweaks

@@
-\section{Related work}\label{sec-related}
+\section{Related Work}\label{sec-related}
 
@@
 \item \Dune~\cite{dune} is a build system designed for OCaml/Reason projects.
-Its distinguishing feature is that it uses
-\emph{arrows}~\cite{hughes2000generalising} rather than monads to model
-dynamic dependencies, which simplifies static dependency approximation.
+Its original implementation used \emph{arrows}~\cite{hughes2000generalising}
+rather than monads to model dynamic dependencies, which simplified static
+dependency approximation. \Dune was later redesigned to use a flavour of
+selective functors~\cite{mokhov_selective_2019}, making it a closer fit to our
+abstractions.
 
@@
 of precision, but a fully general resolution handler extends the \hs{Task}
-abstraction with additional features.
+abstraction with new features.
 
@@
 matches \Shake at the level of detail given here, differing only on aspects like
-rules producing multiple files~\S\ref{sec-polymorphism}.
+rules producing multiple files~\S\ref{sec-polymorphism}. While \Redo predates
+\Shake, they were developed independently; we use \Shake as a prototypical
+example of a monadic build system because its implementation presents a closer
+mapping to our \hs{Task} abstraction.
 
@@
 \cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
-skipping any commands where no inputs have changed. The key difference from our
-\hs{Tasks} abstraction is that instead of supplying a mapping from outputs to
-tasks, \Fabricate supplies a list of statements, in an order, without declaring
-what each line produces. There is no need to schedule the statements, and not
-enough information to do so.
+skipping any commands where no inputs have changed. The main difference from our
+\hs{Tasks} abstraction is that instead of supplying a mapping from keys to
+tasks, a \Fabricate script supplies a list of build statements, in a
+user-scheduled order, without declaring what each statement reads or write.
 
@@
 However, in our pure model the mapping is not perfect as \cmd{gcc} writes to
-arbitrary files whose locations are not known in advance.
+arbitrary files whose locations are not known in advance. One way of capturing
+arbitrary writes in our model is to switch from one callback \hs{fetch} to
+\emph{two callbacks}, say \hs{read} and \hs{write}, allowing us to track both
+reads and writes separately.
 
@@
 systems (\Shake's oracles~\S\ref{sec-polymorphism} can be used to model this
-feature to a limited degree).
+feature to a limited degree). Another important optimisation that self-adjusting
+computation engines often support is the incremental processing of
+\emph{deltas}, where instead of marking a value as ``changed to 8'', one can
+mark it as ``changed by $+1$'', assuming it was equal to 7 before. When a delta
+is small, it can often be propagated to the output more efficiently than by
+recomputing the output value from scratch.
 
 A lot of research has been dedicated to finding efficient data structures and
-algorithms for self-adjusting computations (with a few open-source
+algorithms for self-adjusting computations, with a few open-source
 implementations, e.g. \Incremental by Jane Street~\shortcite{incremental}. We
@@
 function instead of recomputing them each time the function is called. Minimal
-build systems (see the Definition~\ref{def-minimal}) certainly perform
+build systems (\S\ref{sec-background-make}) certainly perform
 memoization: they \emph{store values instead of recomputing them each time}.
@@
 \end{minted}
-\vspace{0mm}
 \begin{minted}[xleftmargin=10pt]{haskell}
@@
 When asked to build \hs{C}~$n$~$m$, a minimal build system will calculate the
-result using memoization. Furthermore, when an input symbol $a_i$ is changed,
-only necessary, incremental recomputation will be performed~--~an optimisation
-that cannot be achieved just with memoization.
+result using memoization. Furthermore, when an input $a_i$ is changed, only
+necessary, incremental recomputation will be performed~--~an optimisation that
+cannot be achieved just with memoization.
 

### 2019-10-21 papers/jfp/4-schedulers.tex
Proofreading

@@
 
-As per the definition of minimality~\ref{def-minimal}, a minimal build
+As per the definition of minimality~(\S\ref{def-minimal}), a minimal build
 system must \textbf{rebuild only out-of-date keys} and at most once. The only
@@
 \hs{key}'s dependencies, then compute a topological order. Taking the \Make
-example from Fig.~\ref{fig-make}, we might compute the order:
+example from Fig.~\ref{fig-make}, we might compute the following order:
 
@@
 \cmd{dep}, abort the task and build \cmd{dep} instead. Returning to the example
-from Fig.~\ref{fig-make}, we might build the tasks in the order:
+from Fig.~\ref{fig-make}, we might build the tasks as follows:
 
@@
 \noindent
-We start with \cmd{main.exe} (chosen randomly), but discover it depends on
+We start with \cmd{main.exe} (an arbitrary choice), but discover it depends on
 \cmd{main.o}, so instead start building \cmd{main.o}. Next we choose to build
-\cmd{util.o} (again, randomly), before finally returning to \cmd{main.exe} that
-now has all its dependencies available and completes successfully.
+\cmd{util.o} (again, arbitrarily), before finally returning to \cmd{main.exe}
+that now has all its dependencies available and completes successfully.
 
-The approach works, but has a number of disadvantages. Firstly, it requires a
+This approach works, but has a number of disadvantages. Firstly, it requires a
 technical mechanism to abort a task, which is easy in our theoretical setting

### 2019-10-21 papers/jfp/6-implementations.tex
Proofreading

@@
 
-We can define schedulers and rebuilders more concretely with the types
+We can define schedulers and rebuilders more concretely with the following types
 (Fig.~\ref{fig-types}):
@@
 
-These two abstractions are the key to modularity: \emph{we can combine any
+These two abstractions are the key to modularity: \emph{we can compose any
 scheduler with any rebuilder, and obtain a correct build system}.
 In this section we will write a scheduler for each column of
-Table~\ref{tab-build-systems}, and a rebuilder for each row; then combine them
+Table~\ref{tab-build-systems}, and a rebuilder for each row; then compose them
 to obtain the build systems in the table's body.
@@
 
-As discussed in \S\ref{sec-step-traces}, \Shake actually use verifying step
+As discussed in \S\ref{sec-step-traces}, \Shake actually uses verifying step
 traces, but here we choose to focus on the more explicit verifying traces. We
@@
 specified by the persistently stored calc chain, \Bazel uses a \emph{build
-queue}. The build starts with the queue containing all dirty keys. Similar to
+queue}. The build starts with the queue containing all dirty keys. Similarly to
 \Excel, the rebuilding of a key extracted from the queue may fail because one of
@@
 
-Using the abstractions built thus far, we have shown how to combine schedulers
+Using the abstractions built thus far, we have shown how to compose schedulers
 with rebuilders to reproduce existing build systems. To us, the most interesting
-build system as yet unavailable would combine a suspending scheduler with
+build system as yet unavailable would compose a suspending scheduler with
 constructive traces~--~providing a cloud-capable build system that is minimal,
@@
 is possible to define and test such a system, which we call \Cloud \Shake. All
-we need to do is combine \hs{suspending} with \hs{ctRebuilder},
+we need to do is compose \hs{suspending} with \hs{ctRebuilder},
 as shown in Fig.~\ref{fig-cloud-implementations}.

### 2019-10-21 papers/jfp/7-experience.tex
Proofreading

@@
 our thinking.  For example, here are definitions for a scheduler and rebuilder,
-from \S~\ref{sec-concrete-impl}:
+from~\S\ref{sec-concrete-impl}:
 
@@
 project\footnote{See \url{https://github.com/ndmitchell/shake}.}. These commits
-add concepts like \emph{resources} (for handling situations when two build tasks
+added concepts like \emph{resources} (for handling situations when two build tasks
 contend on a single external resource), rewriting serialisation to be faster,
@@
 in the original paper the tasks (there called ``actions'')
-were described in continuation-passing style using the data
+were described in continuation-passing style using the following data
 type\footnote{The original paper uses concrete types \hs{Key} and \hs{Value}.
@@
 
-\subsection{Experience from using \Cloud \Shake}\label{sec-using-cloud-shake}
+\subsection{Experience from Using \Cloud \Shake}\label{sec-using-cloud-shake}
 
@@
 \Cloud \Shake, we were actually beaten to it by Standard Chartered who
-report\footnote{\mbox{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}}:
+reported\footnote{\mbox{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}}:
 
@@
       always a problem, the move to a cloud build makes them more serious.
-      With local builds outputs will be built at least once per user, but with a
+      With local builds, outputs will be built at least once per user, but with a
       cloud build they might only be built \emph{once ever}.
@@
 files -- we have further enhanced the \Shake lint modes, coupling them to a
-program called \cmd{FSATrace}, which detects which files are read/written by a
+utility called \cmd{FSATrace}, which detects which files are read/written by a
 command line execution. Such information has been very helpful in making the GHC
@@
 
-\subsection{Experience from building GHC with \Shake}\label{sec-hadrian}
+\subsection{Experience from Building GHC with \Shake}\label{sec-hadrian}
 
@@
 (e.g. only modifies the documentation), \Hadrian can build GHC from scratch in under
-3~minutes, by simply creating symbolic links to the previously build results
+3~minutes, by simply creating symbolic links to the previously built results
 stored in the \Cloud~\Shake cache.

### 2019-10-21 papers/jfp/8-engineering.tex
Proofreading

@@
 that produced value \emph{belongs to the set of possible results} of
-recomputing the task, i.e. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
+recomputing the task, e.g. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
 example.
@@
 Note that like all key-dependent tasks, this example could be expressed
-without key-dependence, at the cost of some type safety. As we'll see in
+without key-dependence, at the cost of some type safety. As we will see in
 \S\ref{sec-multiple-outputs} and \S\ref{sec-tracking-aspects}, using key-dependent
@@
 
-\subsection{File watching vs polling}\label{sec-file-watching}
+\subsection{File Watching vs Polling}\label{sec-file-watching}
 

### 2019-10-21 papers/jfp/9-related.tex
Proofreading

@@
 \item \Redo~\cite{redo-idea,grosskurth2007redo,redo} almost exactly
-matches \Shake at the level of detail given here, differing only on aspects like
+matches \Shake at the level of detail given here, differing only in aspects like
 rules producing multiple files~\S\ref{sec-polymorphism}. While \Redo predates
@@
 tasks, a \Fabricate script supplies a list of build statements, in a
-user-scheduled order, without declaring what each statement reads or write.
+\emph{user-scheduled order}, without declaring what each statement reads or write.
 

### 2019-10-21 papers/jfp/7-experience.tex
Wibbles

@@
 code gave us a tangible, concrete basis for productive conversations,
-much more so than general conversations about ``tasks''.
+much more so than general debate about ``tasks''.
 
-It is also worth noting that we needed a rather expressive language,
+It is also worth noting that we needed a rather expressive language
 to faithfully express the abstractions that seem natural in this
-setting.  In the case of \hs{Task} we use a data constructor with a
+setting.  In the case of \hs{Task} we needed: a data constructor with a
 polymorphic field; a higher-kinded type variable \hs{f}~\hs{::}~\hs{*}~\hs{->}~\hs{*}; an
 even more abstracted kind for \hs{c}~\hs{::}~\hs{(*}~\hs{->}~\hs{*)}~\hs{->}~\hs{Constraint}; and,
-of course, type classes. Our models have been translated to
+of course, type classes. Our models have since been translated to
 Rust~\cite{translation_rust} and Kotlin~\cite{translation_kotlin}, and in both
@@
 When thinking about the type constructors over which \hs{f} might usefully
-range, it turned out that we could adopt the \emph{existing abstractions}
+range, it turned out that we could adopt the \emph{existing} abstractions
 of \hs{Functor}, \hs{Applicative}, \hs{Monad} and so on.
-That in turn led us to think in a new way about the taxonomy
+That in turn led us to a new taxonomy
 of build systems -- see~\S\ref{sec-why-polymorphism}. In the other direction,
@@
 of the conceptual structure of the paper, emerged fairly late in the
-project, as we repeatedly reviewed, redesigned, and refactored our
+project as we repeatedly reviewed, redesigned, and refactored our
 executable prototypes. It is hard to believe that we could have

### 2019-10-21 papers/jfp/7-experience.tex
Update reference

@@
 
+\noindent
 These powerful and modular abstractions, which ultimately formed part

### 2019-10-22 papers/jfp/7-experience.tex
Reword the paragraph on cloud Hadrian

@@
 
-The subsequent development of \Cloud \Shake allows GHC developers to take advantage of
-sharing build results between builds, which is particularly promising in the
-context of GHC's continuous integration (CI) infrastructure responsible for
-building and testing every GHC commit. Building GHC~8.8 from scratch with
-default settings takes around 1 hour on Windows using either \Hadrian or the
-original \Make-based build system. This time includes building the compiler itself,
-29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
-6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
-only a small part of the whole codebase, making the cloud-build functionality
-highly desirable. For example, if a commit does not touch any GHC source files
-(e.g. only modifies the documentation), \Hadrian can build GHC from scratch in under
-3~minutes, by simply creating symbolic links to the previously built results
-stored in the \Cloud~\Shake cache.
+The development of \Cloud \Shake allows GHC developers to benefit from caching
+build results between builds. Building GHC~8.8 from scratch takes $\sim$1~hour
+on Windows using \Hadrian or the original \Make-based build system. This time
+includes building the compiler itself, 29~bundled libraries, such as \cmd{base}
+(each in vanilla and profiled way), and 6~bundled executables, such as
+\cmd{Haddock}. One would hope that the build cache would be particularly useful
+for GHC's continuous integration system that builds and tests every commit but
+our experience has been mixed. In an ideal case, when a commit does not affect
+the resulting GHC binaries, e.g. only modifies tests, \Hadrian can build GHC
+from scratch in just 3~minutes, by simply creating symbolic links to the
+previously built results stored in the cache. However, if a commit modifies the
+``Stage 1 GHC'' executable -- an intermediate compiler built as part of the GHC
+bootstrapping -- any further cache hits become unlikely, thus limiting
+benefits of the build cache to Stage 1 only~\cite{cloud_hadrian}.
 

### 2019-10-25 papers/jfp/8-engineering.tex
Use \Latex rather than \LaTeX (it's not clear we have copyright for the Latex logo)

@@
 Some computations are best described not by a chain of acyclic dependencies,
-but by a loop. For example, \LaTeX~requires repeated rebuilding until it
+but by a loop. For example, \Latex~requires repeated rebuilding until it
 reaches a fixed point, which can be directly expressed in build systems, such as
@@
 
-For examples like \LaTeX~we consider the proper encoding to not be circular
+For examples like \Latex~we consider the proper encoding to not be circular
 tasks, but a series of iterative steps, as described by
@@
 executions is bounded, otherwise the build system may not terminate (a
-legitimate concern with \LaTeX, which can be put into a situation where it is
+legitimate concern with \Latex, which can be put into a situation where it is
 bistable or diverging over multiple executions). The examples in \Excel tend to

### 2019-11-01 papers/jfp/8-engineering.tex
Add a paragraph on concurrent output modification

@@
 gracefully instead of terminating with an error.
+
+Build systems that run continuously are also more likely to encounter errors
+caused by concurrent modification of \emph{build outputs}. For example, if an
+output file is checked into the source repository then downloading a new
+version of the file can interfere with the build task producing it, resulting
+in a corrupted output that the build system will be unable to detect. This
+problem can be solved by ensuring that tasks have an \emph{exclusive access}
+to their outputs, e.g. by sandboxing the tasks whose outputs can be modified
+externally.

### 2019-11-08 papers/jfp/1-intro.tex
Minor tweaks

@@
 with desired properties, for example, to combine the advantages of \Shake and
-\Bazel. Writing this paper gave us the insights to combine dynamic dependencies and
-cloud build in a principled way; we evaluate the result~(\S\ref{sec-experience}).
+\Bazel. Writing this paper gave us the insights to combine dynamic dependencies
+and cloud build systems in a principled way; we evaluate the result
+in~\S\ref{sec-experience}.
 

### 2019-11-08 papers/jfp/2-background.tex
Wibbles
Minor tweaks

@@
 will be ready to define in~\S\ref{sec-build-correctness}; for now, we will use
-minimality as the guiding principle when exploring the design space of build
+minimality as a guiding principle when exploring the design space of build
 systems.
@@
 
-\subsection{\Shake: Dynamic Dependencies with No Remorse}
+\subsection{\Shake: Dynamic Dependencies without Remorse}
 \label{sec-background-shake}
